### 2022-09-30 paper/paper.qmd
init

---
title: "Demo arXiv template"
format:
  arxiv-pdf:
    keep-tex: true  
  arxiv-html: default
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, VIC
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@monash.edu
    url: https://huizezhangsh.netlify.app/
  - name: Collaborators
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, VIC
        country: Australia
abstract: |
  - The paper describes a framework for constructing indices 
  - use drought indices as examples, but appliable in general to environmental indices constructed from multivariate spatio-temporal data
keywords: 
  - template
  - demo
bibliography: bibliography.bib  
---

### 2022-09-30 paper/paper.tex
init

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
  \setmathfont[]{Latin Modern Math}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Demo arXiv template},
  pdfauthor={H. Sherry Zhang; Collaborators},
  pdfkeywords={template, demo},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Demo arXiv template}
\author{
\textbf{H. Sherry Zhang}~\orcidlink{0000-0002-7122-1463}\\Department of
Econometrics and Business Statistics\\Monash University\\Melbourne,
VIC\\\href{mailto:huize.zhang@monash.edu}{huize.zhang@monash.edu}\\\\\\
\textbf{Collaborators}\\Department of Econometrics and Business
Statistics\\Monash University\\Melbourne, VIC\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
This document is only a demo explaining how to use the template.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
template \sep 
demo

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi

\end{document}

### 2022-09-30 paper/paper.qmd
some structure and some content

@@
 keywords: 
-  - template
-  - demo
+  - spatio-temporal data
+  - indices
+  - data pipeline
 bibliography: bibliography.bib  
 ---
+
+# Introduction
+
+Index construction is a way to summarise complicated information (used in environmental data). 
+The complexity of these information may involve a spatial distribution, a temporal scale that defines the frequency of the data, and a multivariate perspective that collects different climate variables.
+
+Numerous indices have been proposed by researchers and practitioners to monitor natural hazard, for example, @alahacoon_comprehensive_2022 reviews 111 drought indices derived from traditional and remote sensing data; [review of index construction in other area, climate indices, many other reviews in drought].
+
+<!-- These indices are customised to distinct local weather condition and historical data availability. -->
+
+Various methods are proposed to extract multivariate information on the spatial extent, across time.
+
+Each individual index follows its own data pipeline and it can be difficult to evaluate how an index can be affected by tweaking parameters in a certain step, rearranging the order of steps, using to different method.
+This paper proposes a data pipeline for constructing indices using multivariate spatio-temporal data.
+The steps involved in the pipeline are general in nature and flexible to be adopted to most index construction for environmental data.
+
+# Natural hazard indices
+
+## Climate indices
+
+## Drought indices
+
+# Data pipeline
+
+Constructing a pipeline that divides a complex procedure into steps that can be concatenated has been adopted widely in the computational statistics community. 
+One particular example is a machine learning pipeline, tidymodel [@tidymodel]. 
+[more details on pipeline]
+
+**Statistical pipeline in interactive graphics**
+
+The data pipeline in interactive graphics is a set of steps that transform the raw data to the plots displayed on the screen. 
+The initial pipeline proposed by @buja_elements_1988 involves the following steps: non-linear transformation, variables standardization, randomization, projection engine, and viewporting.
+Another example in the early work of pipeline by @sutherland_orca_2000 describes a three-step pipeline: variable standardization, dimension reduction, and scaling data into the viewing window. 
+This pipeline also includes the transformation on spatial and temporal variables, i.e. computing time lag on temporal variables.
+@wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program and breaking down graphic rendering into steps is also beneficial for understand the implementation and compare between different graphic systems.
+
+**Statistical pipeline in other domains**
+
+TODO: read papers on other pipeline in R
+
+# A pipeline for natural hazard indices
+
+uniform workflow to work with index construction.
+
+  - illustration 
+  - math notation
+  - benefit of the pipeline approach 
+    - index diagnostic
+    - uncertainty
+
+# Examples
+
+## Constructing Standardised Precipitation Index (SPI)
+
+  - a basic workflow and congruence with results in the `SPEI` pkg
+  - allow multiple distribution fit 
+  - allow bootstrap uncertainty
+
+# Reference

### 2022-09-30 paper/paper.tex
some structure and some content

@@
 \makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
   pdfauthor={H. Sherry Zhang; Collaborators},
-  pdfkeywords={template, demo},
+  pdfkeywords={spatio-temporal data, indices, data pipeline},
   colorlinks=true,
@@
 \begin{abstract}
-This document is only a demo explaining how to use the template.
+\begin{itemize}
+\tightlist
+\item
+  The paper describes a framework for constructing indices
+\item
+  use drought indices as examples, but appliable in general to
+  environmental indices constructed from multivariate spatio-temporal
+  data
+\end{itemize}
 \end{abstract}
@@
 \def\sep{\textbullet\ }
-template \sep 
-demo
+spatio-temporal data \sep indices \sep 
+data pipeline
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+Index construction is a way to summarise complicated information (used
+in environmental data). The complexity of these information may involve
+a spatial distribution, a temporal scale that defines the frequency of
+the data, and a multivariate perspective that collects different climate
+variables.
+
+Numerous indices have been proposed by researchers and practitioners to
+monitor natural hazard, for example, Alahacoon and Edirisinghe (2022)
+reviews 111 drought indices derived from traditional and remote sensing
+data; {[}review of index construction in other area, climate indices,
+many other reviews in drought{]}.
+
+Various methods are proposed to extract multivariate information on the
+spatial extent, across time.
+
+Each individual index follows its own data pipeline and it can be
+difficult to evaluate how an index can be affected by tweaking
+parameters in a certain step, rearranging the order of steps, using to
+different method. This paper proposes a data pipeline for constructing
+indices using multivariate spatio-temporal data. The steps involved in
+the pipeline are general in nature and flexible to be adopted to most
+index construction for environmental data.
+
+\hypertarget{natural-hazard-indices}{%
+\section{Natural hazard indices}\label{natural-hazard-indices}}
+
+\hypertarget{climate-indices}{%
+\subsection{Climate indices}\label{climate-indices}}
+
+\hypertarget{drought-indices}{%
+\subsection{Drought indices}\label{drought-indices}}
+
+\hypertarget{data-pipeline}{%
+\section{Data pipeline}\label{data-pipeline}}
+
+Constructing a pipeline that divides a complex procedure into steps that
+can be concatenated has been adopted widely in the computational
+statistics community. One particular example is a machine learning
+pipeline, tidymodel (Kuhn and Wickham 2020). {[}more details on
+pipeline{]}
+
+\textbf{Statistical pipeline in interactive graphics}
+
+The data pipeline in interactive graphics is a set of steps that
+transform the raw data to the plots displayed on the screen. The initial
+pipeline proposed by Buja et al. (1988) involves the following steps:
+non-linear transformation, variables standardization, randomization,
+projection engine, and viewporting. Another example in the early work of
+pipeline by Sutherland et al. (2000) describes a three-step pipeline:
+variable standardization, dimension reduction, and scaling data into the
+viewing window. This pipeline also includes the transformation on
+spatial and temporal variables, i.e.~computing time lag on temporal
+variables. Wickham et al. (2009) argues that whether made explicit or
+not, pipeline has to be presented in every graphics program and breaking
+down graphic rendering into steps is also beneficial for understand the
+implementation and compare between different graphic systems.
+
+\textbf{Statistical pipeline in other domains}
+
+TODO: read papers on other pipeline in R
+
+\hypertarget{a-pipeline-for-natural-hazard-indices}{%
+\section{A pipeline for natural hazard
+indices}\label{a-pipeline-for-natural-hazard-indices}}
+
+uniform workflow to work with index construction.
+
+\begin{itemize}
+\tightlist
+\item
+  illustration
+\item
+  math notation
+\item
+  benefit of the pipeline approach
+
+  \begin{itemize}
+  \tightlist
+  \item
+    index diagnostic
+  \item
+    uncertainty
+  \end{itemize}
+\end{itemize}
+
+\hypertarget{examples}{%
+\section{Examples}\label{examples}}
+
+\hypertarget{constructing-standardised-precipitation-index-spi}{%
+\subsection{Constructing Standardised Precipitation Index
+(SPI)}\label{constructing-standardised-precipitation-index-spi}}
+
+\begin{itemize}
+\tightlist
+\item
+  a basic workflow and congruence with results in the \texttt{SPEI} pkg
+\item
+  allow multiple distribution fit
+\item
+  allow bootstrap uncertainty
+\end{itemize}
+
+\hypertarget{reference}{%
+\section*{Reference}\label{reference}}
+\addcontentsline{toc}{section}{Reference}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-alahacoon_comprehensive_2022}{}}%
+Alahacoon, Niranga, and Mahesh Edirisinghe. 2022. {``A Comprehensive
+Assessment of Remote Sensing and Traditional Based Drought Monitoring
+Indices at Global and Regional Scale.''} \emph{Geomatics, Natural
+Hazards and Risk} 13 (December): 762--99.
+\url{https://doi.org/10.1080/19475705.2022.2044394}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-buja_elements_1988}{}}%
+Buja, A, D Asimov, C Hurley, and JA McDonald. 1988. {``Elements of a
+Viewing Pipeline for Data Analysis.''} In \emph{Dynamic Graphics for
+Statistics}, 277--308. Wadsworth, Belmont.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidymodel}{}}%
+Kuhn, Max, and Hadley Wickham. 2020. \emph{Tidymodels: A Collection of
+Packages for Modeling and Machine Learning Using Tidyverse Principles.}
+\url{https://www.tidymodels.org}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sutherland_orca_2000}{}}%
+Sutherland, Peter, Anthony Rossini, Thomas Lumley, Nicholas Lewin-Koh,
+Julie Dickerson, Zach Cox, and Dianne Cook. 2000. {``Orca: {A}
+{Visualization} {Toolkit} for {High}-{Dimensional} {Data}.''}
+\emph{Journal of Computational and Graphical Statistics} 9 (3): 509--29.
+\url{https://www.jstor.org/stable/1390943}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%
+Wickham, Hadley, Michael Lawrence, Dianne Cook, Andreas Buja, Heike
+Hofmann, and Deborah F. Swayne. 2009. {``The Plumbing of Interactive
+Graphics.''} \emph{Computational Statistics} 24 (2): 207--15.
+\url{https://doi.org/10.1007/s00180-008-0116-x}.
+
+\end{CSLReferences}
 

### 2022-10-02 paper/paper.qmd
build up more content

@@
 
-Numerous indices have been proposed by researchers and practitioners to monitor natural hazard, for example, @alahacoon_comprehensive_2022 reviews 111 drought indices derived from traditional and remote sensing data; [review of index construction in other area, climate indices, many other reviews in drought].
-
 <!-- These indices are customised to distinct local weather condition and historical data availability. -->
@@
 
-# Data pipeline
+# Data pipeline in R
 
-Constructing a pipeline that divides a complex procedure into steps that can be concatenated has been adopted widely in the computational statistics community. 
-One particular example is a machine learning pipeline, tidymodel [@tidymodel]. 
-[more details on pipeline]
-
-**Statistical pipeline in interactive graphics**
+Constructing a pipeline that divides a complex procedure into steps that can be concatenated has been adopted widely in the R community.
 
@@
 The initial pipeline proposed by @buja_elements_1988 involves the following steps: non-linear transformation, variables standardization, randomization, projection engine, and viewporting.
+The initial pipeline proposed by @buja_elements_1988 involves the following steps: non-linear transformation, variables standardization, randomization, projection engine, and viewporting. 
 Another example in the early work of pipeline by @sutherland_orca_2000 describes a three-step pipeline: variable standardization, dimension reduction, and scaling data into the viewing window. 
+This pipeline also includes the transformation on spatial and temporal variables, i.e. computing time lag on temporal variables.
 This pipeline also includes the transformation on spatial and temporal variables, i.e. computing time lag on temporal variables. 
@@
 
-**Statistical pipeline in other domains**
+The data pipeline concept is further enhanced by the pipe operator (`%>%`) in R where a set of operations, or steps, can be chained together to form a set of instructions.
 
-TODO: read papers on other pipeline in R
+A more recent data pipeline is tidymodels [@tidymodels], a set of packages for machine learning models following the tidyverse principles [@tidyverse]. \[expand on tidymodels\]
 
-# A pipeline for natural hazard indices
+# A pipeline for building natural hazard indices
+
+The construction of natural hazard indices also follows a set of steps, which is usually illustrated using a flowchart in the paper. 
+However, every researcher follows a certain design philosophy and steps taken in the index constructed by different researchers are not aligned. 
+This discourages experiment with multiple indices. Initiate a new workflow when computing a new index.
+
+## Raw data
+
+The data used to construct the natural hazard index usually have three dimensions, one for location, one for time, and one for multivariate. 
+Mathematically, it can be written as $X_{j, s, t}$, where $j = 1, 2, \cdots, J$ for variable, $s = 1, 2, \cdots, S$ for location, and $t = 1, 2, \cdots, T$ for time.
+
+The location $s$ can refer to vector points or areas characterised by longitude-latitude coordinates, or raster cells obtained from satellite images.
+
+The time dimension $t$ can be daily, weekly, biweekly (14-16 days), monthly, or even quarterly
+
+Variables
+
+This multidimensional array structure is commonly used in geospatial analysis
+
+Given the variety of data sources at different spatial resolution and temporal granularity, the raw data may first come in multiple pieces. 
+Sometimes, even a considerable amount of work is needed to align the spatial and temporal extent of multivariate data.
+
+A notation for different variables have different spatial and temporal granularity $X_{j_1, s_1, t_1}$???
+
+## Spatial aggregation
+
+## Variable transformation
+
+Take one or multiple variables to create a new variable, linearly? 
+sometimes called feature extraction in the machine learning community 
+With drought indices, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.]
+
+## Temporal processing
+
+## Dimension reduction
+
+## Normalising
+
+only for distribution fit, convert from an fitted distribution to normal
+
+## Benchmarking
+
+## Simplification
 
@@
     -   uncertainty
+
+# Extending the pipeline
 

### 2022-10-02 paper/paper.tex
build up more content

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{data-pipeline}{%
-\section{Data pipeline}\label{data-pipeline}}
+\hypertarget{data-pipeline-in-r}{%
+\section{Data pipeline in R}\label{data-pipeline-in-r}}
 
 Constructing a pipeline that divides a complex procedure into steps that
-can be concatenated has been adopted widely in the computational
-statistics community. One particular example is a machine learning
-pipeline, tidymodel (Kuhn and Wickham 2020). {[}more details on
-pipeline{]}
-
-\textbf{Statistical pipeline in interactive graphics}
+can be concatenated has been adopted widely in the R community.
 
@@
 
-\textbf{Statistical pipeline in other domains}
+The data pipeline concept is further enhanced by the pipe operator
+(\texttt{\%\textgreater{}\%}) in R where a set of operations, or steps,
+can be chained together to form a set of instructions.
 
-TODO: read papers on other pipeline in R
+A more recent data pipeline is tidymodels (Kuhn and Wickham 2020), a set
+of packages for machine learning models following the tidyverse
+principles (Wickham et al. 2019). {[}expand on tidymodels{]}
 
-\hypertarget{a-pipeline-for-natural-hazard-indices}{%
-\section{A pipeline for natural hazard
-indices}\label{a-pipeline-for-natural-hazard-indices}}
+\hypertarget{a-pipeline-for-building-natural-hazard-indices}{%
+\section{A pipeline for building natural hazard
+indices}\label{a-pipeline-for-building-natural-hazard-indices}}
+
+The construction of natural hazard indices also follows a set of steps,
+which is usually illustrated using a flowchart in the paper. However,
+every researcher follows a certain design philosophy and steps taken in
+the index constructed by different researchers are not aligned. This
+discourages experiment with multiple indices. Initiate a new workflow
+when computing a new index.
+
+\hypertarget{raw-data}{%
+\subsection{Raw data}\label{raw-data}}
+
+The data used to construct the natural hazard index usually have three
+dimensions, one for location, one for time, and one for multivariate.
+Mathematically, it can be written as \(X_{j, s, t}\), where
+\(j = 1, 2, \cdots, J\) for variable, \(s = 1, 2, \cdots, S\) for
+location, and \(t = 1, 2, \cdots, T\) for time.
+
+The location \(s\) can refer to vector points or areas characterised by
+longitude-latitude coordinates, or raster cells obtained from satellite
+images.
+
+The time dimension \(t\) can be daily, weekly, biweekly (14-16 days),
+monthly, or even quarterly
+
+Variables
+
+This multidimensional array structure is commonly used in geospatial
+analysis
+
+Given the variety of data sources at different spatial resolution and
+temporal granularity, the raw data may first come in multiple pieces.
+Sometimes, even a considerable amount of work is needed to align the
+spatial and temporal extent of multivariate data.
+
+A notation for different variables have different spatial and temporal
+granularity \(X_{j_1, s_1, t_1}\)???
+
+\hypertarget{spatial-aggregation}{%
+\subsection{Spatial aggregation}\label{spatial-aggregation}}
+
+\hypertarget{variable-transformation}{%
+\subsection{Variable transformation}\label{variable-transformation}}
+
+\hypertarget{temporal-processing}{%
+\subsection{Temporal processing}\label{temporal-processing}}
+
+\hypertarget{dimension-reduction}{%
+\subsection{Dimension reduction}\label{dimension-reduction}}
+
+\hypertarget{normalising}{%
+\subsection{Normalising}\label{normalising}}
+
+\hypertarget{benchmarking}{%
+\subsection{Benchmarking}\label{benchmarking}}
+
+\hypertarget{simplification}{%
+\subsection{Simplification}\label{simplification}}
 
@@
 \end{itemize}
+
+\hypertarget{extending-the-pipeline}{%
+\section{Extending the pipeline}\label{extending-the-pipeline}}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tidymodel}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-tidymodels}{}}%
 Kuhn, Max, and Hadley Wickham. 2020. \emph{Tidymodels: A Collection of
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
+Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
+D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
+{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
+4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%

### 2022-10-04 paper/paper.qmd
add more content

@@
 abstract: |
-  - The paper describes a framework for constructing indices 
-  - use drought indices as examples, but appliable in general to environmental indices constructed from multivariate spatio-temporal data
+  - Indices, useful, quantify severity, early monitoring, 
+  - A huge number of indices have been proposed by domain experts, however, a large majority of them are not being adopted and reused in research or in practice. 
+  - One of the most commonly used drought index, SPI and SPEI, are available through the `SPEI` package and its readily avaiability to be applied on user's data 
+  - However, even if most of the indices have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember
+  - The paper describes a general pipeline framework to construct indices from spatio-temporal data, so that indices can be constructed through [a set of general steps].
+  - The benefit of such a framework is that ...
+  - The framework will be demonstrated with drought indices as examples, but appliable in general to environmental indices constructed from multivariate spatio-temporal data
 keywords: 
@@
 
-Various methods are proposed to extract multivariate information on the spatial extent, across time.
+Various methods are proposed to extract multivariate information on the spatial coverage, across time.
+
+Indices with readily available code to implement have more potential to be computed by researchers, i.e.
+the package `SPEI` [@SPEI] has implemented the SPI and SPEI index for calculating drought indices.
+[Another package in a different domain]
 
@@
 
+## Economics/ social indices
+
 # Data pipeline in R
+
+## Tidy data
+
+Before the concept of tidy data [@wickham_tidy_2014], tabular data arrive at data analysts in all different ways. 
+Different analysts would write customised scripts for analysing the specific data. 
+These scripts can be extended to other data analysed by the same people or group but this is not generalizable directly to another dataset. 
+When the tidy data concept comes, variables and values are arranged so that 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
+With this specific layout, wrangling on tabular data can be standardised into a grammar of data manipulation in `dplyr` [@dplyr].
+
+A similar issue happens with index construction where researchers construct their own indices in their own ways and there has not yet been a tidy principle on index construction.
+Also, this tidy principle on index construction is more complex than those in tidy data and the `dplyr` package.
+It has to encompass the workflow of transformation from the raw data towards the final index series.
+
+## Data pipeline
 
@@
 This discourages experiment with multiple indices. Initiate a new workflow when computing a new index.
+
+The most popular indices (i.e. SPI, SPEI, etc) have existing software implementation (`SPEI`) to be applied to a different set of data.
+
+constructing time series index should also be encapsulated in my framework
+
+reusable
 
@@
 
-only for distribution fit, convert from an fitted distribution to normal
+specifically for converting from a fitted distribution to normal score via reverse CDF function, non-parametric formula, or empirical approximation, a common step in many index: SPI, SSI, Z score.
+The purpose of normalising is to convert the index into a standardised series after all the steps for the ease of comparison.
 
@@
 ## Simplification
+
+Discretise the continuous index into a few labelled categories. For communicating the severity of natural hazard to general public.
 

### 2022-10-04 paper/paper.tex
add more content

@@
 \item
-  The paper describes a framework for constructing indices
+  Indices, useful, quantify severity, early monitoring,
 \item
-  use drought indices as examples, but appliable in general to
-  environmental indices constructed from multivariate spatio-temporal
-  data
+  A huge number of indices have been proposed by domain experts,
+  however, a large majority of them are not being adopted and reused in
+  research or in practice.
+\item
+  One of the most commonly used drought index, SPI and SPEI, are
+  available through the \texttt{SPEI} package and its readily
+  avaiability to be applied on user's data
+\item
+  However, even if most of the indices have available existing
+  implementation available to users, there would be hundreds of them and
+  it will be burdensome to remember
+\item
+  The paper describes a general pipeline framework to construct indices
+  from spatio-temporal data, so that indices can be constructed through
+  {[}a set of general steps{]}.
+\item
+  The benefit of such a framework is that \ldots{}
+\item
+  The framework will be demonstrated with drought indices as examples,
+  but appliable in general to environmental indices constructed from
+  multivariate spatio-temporal data
 \end{itemize}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-Numerous indices have been proposed by researchers and practitioners to
-monitor natural hazard, for example, Alahacoon and Edirisinghe (2022)
-reviews 111 drought indices derived from traditional and remote sensing
-data; {[}review of index construction in other area, climate indices,
-many other reviews in drought{]}.
+Various methods are proposed to extract multivariate information on the
+spatial coverage, across time.
 
-Various methods are proposed to extract multivariate information on the
-spatial extent, across time.
+Indices with readily available code to implement have more potential to
+be computed by researchers, i.e.~the package \texttt{SPEI} (Beguería and
+Vicente-Serrano 2017) has implemented the SPI and SPEI index for
+calculating drought indices. {[}Another package in a different domain{]}
 
@@
 
+\hypertarget{economics-social-indices}{%
+\subsection{Economics/ social indices}\label{economics-social-indices}}
+
 \hypertarget{data-pipeline-in-r}{%
 \section{Data pipeline in R}\label{data-pipeline-in-r}}
+
+\hypertarget{tidy-data}{%
+\subsection{Tidy data}\label{tidy-data}}
+
+Before the concept of tidy data (Wickham 2014), tabular data arrive at
+data analysts in all different ways. Different analysts would write
+customised scripts for analysing the specific data. These scripts can be
+extended to other data analysed by the same people or group but this is
+not generalizable directly to another dataset. When the tidy data
+concept comes, variables and values are arranged so that 1) Each
+variable forms a column, 2) Each observation forms a row, and; 3) Each
+type of observational unit forms a table. With this specific layout,
+wrangling on tabular data can be standardised into a grammar of data
+manipulation in \texttt{dplyr} (Wickham et al. 2022).
+
+A similar issue happens with index construction where researchers
+construct their own indices in their own ways and there has not yet been
+a tidy principle on index construction. Also, this tidy principle on
+index construction is more complex than those in tidy data and the
+\texttt{dplyr} package. It has to encompass the workflow of
+transformation from the raw data towards the final index series.
+
+\hypertarget{data-pipeline}{%
+\subsection{Data pipeline}\label{data-pipeline}}
 
@@
 non-linear transformation, variables standardization, randomization,
-projection engine, and viewporting. Another example in the early work of
-pipeline by Sutherland et al. (2000) describes a three-step pipeline:
-variable standardization, dimension reduction, and scaling data into the
-viewing window. This pipeline also includes the transformation on
-spatial and temporal variables, i.e.~computing time lag on temporal
-variables. Wickham et al. (2009) argues that whether made explicit or
-not, pipeline has to be presented in every graphics program and breaking
-down graphic rendering into steps is also beneficial for understand the
-implementation and compare between different graphic systems.
+projection engine, and viewporting. The initial pipeline proposed by
+Buja et al. (1988) involves the following steps: non-linear
+transformation, variables standardization, randomization, projection
+engine, and viewporting. Another example in the early work of pipeline
+by Sutherland et al. (2000) describes a three-step pipeline: variable
+standardization, dimension reduction, and scaling data into the viewing
+window. This pipeline also includes the transformation on spatial and
+temporal variables, i.e.~computing time lag on temporal variables. This
+pipeline also includes the transformation on spatial and temporal
+variables, i.e.~computing time lag on temporal variables. Wickham et al.
+(2009) argues that whether made explicit or not, pipeline has to be
+presented in every graphics program and breaking down graphic rendering
+into steps is also beneficial for understand the implementation and
+compare between different graphic systems.
 
@@
 when computing a new index.
+
+The most popular indices (i.e.~SPI, SPEI, etc) have existing software
+implementation (\texttt{SPEI}) to be applied to a different set of data.
+
+constructing time series index should also be encapsulated in my
+framework
+
+reusable
 
@@
 
+Take one or multiple variables to create a new variable, linearly?
+sometimes called feature extraction in the machine learning community
+With drought indices, the extraction of meaningful variables from the
+original data is usually supported by the water balance model, for
+example, in SPEI, the step that create \(d\) out of precipitation and
+potential evapotranspiration (PET) has theoretical backup from {[}see
+paper.{]}
+
 \hypertarget{temporal-processing}{%
@@
 
+specifically for converting from a fitted distribution to normal score
+via reverse CDF function, non-parametric formula, or empirical
+approximation, a common step in many index: SPI, SSI, Z score. The
+purpose of normalising is to convert the index into a standardised
+series after all the steps for the ease of comparison.
+
 \hypertarget{benchmarking}{%
@@
 \subsection{Simplification}\label{simplification}}
+
+Discretise the continuous index into a few labelled categories. For
+communicating the severity of natural hazard to general public.
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alahacoon_comprehensive_2022}{}}%
-Alahacoon, Niranga, and Mahesh Edirisinghe. 2022. {``A Comprehensive
-Assessment of Remote Sensing and Traditional Based Drought Monitoring
-Indices at Global and Regional Scale.''} \emph{Geomatics, Natural
-Hazards and Risk} 13 (December): 762--99.
-\url{https://doi.org/10.1080/19475705.2022.2044394}.
+\leavevmode\vadjust pre{\hypertarget{ref-SPEI}{}}%
+Beguería, Santiago, and Sergio M. Vicente-Serrano. 2017. \emph{SPEI:
+Calculation of the Standardised Precipitation-Evapotranspiration Index}.
+\url{https://CRAN.R-project.org/package=SPEI}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%
+Wickham, Hadley. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
+Software} 59 (September): 1--23.
+\url{https://doi.org/10.18637/jss.v059.i10}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
@@
 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2022.
+\emph{Dplyr: A Grammar of Data Manipulation}.
 

### 2022-10-05 paper/paper.qmd
update on abstract and intro

@@
   - Indices, useful, quantify severity, early monitoring, 
-  - A huge number of indices have been proposed by domain experts, however, a large majority of them are not being adopted and reused in research or in practice. 
-  - One of the most commonly used drought index, SPI and SPEI, are available through the `SPEI` package and its readily avaiability to be applied on user's data 
-  - However, even if most of the indices have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember
-  - The paper describes a general pipeline framework to construct indices from spatio-temporal data, so that indices can be constructed through [a set of general steps].
-  - The benefit of such a framework is that ...
+  - A huge number of indices have been proposed by domain experts, however, a large majority of them are not being adopted, reused, and compared in research or in practice. 
+  - One of the reasons for this is the plenty of indices are quite complex and there is no obvious easy-to-use implementation to apply them to user's data. 
+  - The paper describes a general pipeline framework to construct indices from spatio-temporal data, 
+  - This allows all the indices to be constructed through a uniform data pipeline and different indices to vary on the details of each step in the data pipeline and their orders. 
+  - The pipeline proposed aim to smooth the workflow of index construction through breaking down the complicated steps proposed by various indices into small building blocks shared by most of the indices.
   - The framework will be demonstrated with drought indices as examples, but appliable in general to environmental indices constructed from multivariate spatio-temporal data
@@
 Various methods are proposed to extract multivariate information on the spatial coverage, across time.
+Each individual index follows its own data pipeline and it can be difficult to evaluate how an index can be affected by tweaking parameters in a certain step, rearranging the order of steps, using to different method.
 
@@
 the package `SPEI` [@SPEI] has implemented the SPI and SPEI index for calculating drought indices.
-[Another package in a different domain]
+[Another package in a different domain].  
+However, even if most of the indices have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember, classify, and compare between.
+There is a need for a framework to generally describe how indices can be constructed from spatio-temporal data.
 
-Each individual index follows its own data pipeline and it can be difficult to evaluate how an index can be affected by tweaking parameters in a certain step, rearranging the order of steps, using to different method.
-This paper proposes a data pipeline for constructing indices using multivariate spatio-temporal data.
+This paper proposes such data pipeline for constructing indices.
 The steps involved in the pipeline are general in nature and flexible to be adopted to most index construction for environmental data.

### 2022-10-05 paper/paper.tex
update on abstract and intro

@@
   A huge number of indices have been proposed by domain experts,
-  however, a large majority of them are not being adopted and reused in
-  research or in practice.
+  however, a large majority of them are not being adopted, reused, and
+  compared in research or in practice.
 \item
-  One of the most commonly used drought index, SPI and SPEI, are
-  available through the \texttt{SPEI} package and its readily
-  avaiability to be applied on user's data
-\item
-  However, even if most of the indices have available existing
-  implementation available to users, there would be hundreds of them and
-  it will be burdensome to remember
+  One of the reasons for this is the plenty of indices are quite complex
+  and there is no obvious easy-to-use implementation to apply them to
+  user's data.
 \item
   The paper describes a general pipeline framework to construct indices
-  from spatio-temporal data, so that indices can be constructed through
-  {[}a set of general steps{]}.
+  from spatio-temporal data,
 \item
-  The benefit of such a framework is that \ldots{}
+  This allows all the indices to be constructed through a uniform data
+  pipeline and different indices to vary on the details of each step in
+  the data pipeline and their orders.
+\item
+  The pipeline proposed aim to smooth the workflow of index construction
+  through breaking down the complicated steps proposed by various
+  indices into small building blocks shared by most of the indices.
 \item
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 Various methods are proposed to extract multivariate information on the
-spatial coverage, across time.
+spatial coverage, across time. Each individual index follows its own
+data pipeline and it can be difficult to evaluate how an index can be
+affected by tweaking parameters in a certain step, rearranging the order
+of steps, using to different method.
 
@@
 Vicente-Serrano 2017) has implemented the SPI and SPEI index for
-calculating drought indices. {[}Another package in a different domain{]}
+calculating drought indices. {[}Another package in a different
+domain{]}.\\
+However, even if most of the indices have available existing
+implementation available to users, there would be hundreds of them and
+it will be burdensome to remember, classify, and compare between. There
+is a need for a framework to generally describe how indices can be
+constructed from spatio-temporal data.
 
-Each individual index follows its own data pipeline and it can be
-difficult to evaluate how an index can be affected by tweaking
-parameters in a certain step, rearranging the order of steps, using to
-different method. This paper proposes a data pipeline for constructing
-indices using multivariate spatio-temporal data. The steps involved in
-the pipeline are general in nature and flexible to be adopted to most
-index construction for environmental data.
+This paper proposes such data pipeline for constructing indices. The
+steps involved in the pipeline are general in nature and flexible to be
+adopted to most index construction for environmental data.
 

### 2022-10-05 paper/paper.qmd
notes after the meeting

@@
 
+* indices, why do we need them, decision making
+* define what is an index, what is not even if they are called index
+* only very brief on indices in the introduction
+* potential audience + what do you think they should learn from this work.
+
 # Introduction
 
-Index construction is a way to summarise complicated information (used in environmental data). 
+<!-- Index construction is a way to summarise complicated information (used in environmental data). **it basically says nothing** -->
 The complexity of these information may involve a spatial distribution, a temporal scale that defines the frequency of the data, and a multivariate perspective that collects different climate variables.
@@
 
-# Natural hazard indices
+**Why index is useful, why people care about indices**
 
-## Climate indices
+**review indices, climate, drought, economic, UN and other indices**
 
-## Drought indices
+**What is the problem with current workflow on index construction**
 
-## Economics/ social indices
+**who would benefit from this paper: 1) people proposing new indices; 2) people analysing index variation, compare between indices, 3) those making decisions from indices, 4) other statistician, how to modularise steps for anlaysis in your domain.**
 
@@
 
+**Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem.
+After processing them and align them together they become the "raw data"**
+
 The data used to construct the natural hazard index usually have three dimensions, one for location, one for time, and one for multivariate. 
@@
 
+mostly happen with raster data
+
+## Scaling
+
+standardising
+close to variable transformation step
+
 ## Variable transformation
 
-Take one or multiple variables to create a new variable, linearly? 
-sometimes called feature extraction in the machine learning community 
-With drought indices, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.]
+<!-- Take one or multiple variables to create a new variable,  -->
+Restrict it to single variable, square root, log etc could be linearly, also non-linear
+
+GAM, can you do additive model pairwise/ three-way
 
@@
 ## Dimension reduction
+
+sometimes called feature extraction in the machine learning community 
+With drought indices, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.]
 
@@
 
-# Extending the pipeline
+# Incorporating new methods into the pipeline
+
+<!-- # [Extending the pipeline] a different name here: alternate a module with a different (method) -->
 
@@
 
+## Calculating SPEI with raster data
+
 # Reference

### 2022-10-05 paper/paper.tex
notes after the meeting

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, boxrule=0pt, sharp corners, breakable]}{\end{tcolorbox}}\fi
+
+\begin{itemize}
+\tightlist
+\item
+  indices, why do we need them, decision making
+\item
+  define what is an index, what is not even if they are called index
+\item
+  only very brief on indices in the introduction
+\item
+  potential audience + what do you think they should learn from this
+  work.
+\end{itemize}
 
@@
 
-Index construction is a way to summarise complicated information (used
-in environmental data). The complexity of these information may involve
-a spatial distribution, a temporal scale that defines the frequency of
-the data, and a multivariate perspective that collects different climate
-variables.
+The complexity of these information may involve a spatial distribution,
+a temporal scale that defines the frequency of the data, and a
+multivariate perspective that collects different climate variables.
 
@@
 
-\hypertarget{natural-hazard-indices}{%
-\section{Natural hazard indices}\label{natural-hazard-indices}}
+\textbf{Why index is useful, why people care about indices}
 
-\hypertarget{climate-indices}{%
-\subsection{Climate indices}\label{climate-indices}}
+\textbf{review indices, climate, drought, economic, UN and other
+indices}
 
-\hypertarget{drought-indices}{%
-\subsection{Drought indices}\label{drought-indices}}
+\textbf{What is the problem with current workflow on index construction}
 
-\hypertarget{economics-social-indices}{%
-\subsection{Economics/ social indices}\label{economics-social-indices}}
+\textbf{who would benefit from this paper: 1) people proposing new
+indices; 2) people analysing index variation, compare between indices,
+3) those making decisions from indices, 4) other statistician, how to
+modularise steps for anlaysis in your domain.}
 
@@
 
+\textbf{Another section on original data directly downloaded, can have
+different spatial resolution, temporal granularity, data quality
+problem. After processing them and align them together they become the
+``raw data''}
+
 The data used to construct the natural hazard index usually have three
@@
 
+mostly happen with raster data
+
+\hypertarget{scaling}{%
+\subsection{Scaling}\label{scaling}}
+
+standardising close to variable transformation step
+
 \hypertarget{variable-transformation}{%
@@
 
-Take one or multiple variables to create a new variable, linearly?
+Restrict it to single variable, square root, log etc could be linearly,
+also non-linear
+
+GAM, can you do additive model pairwise/ three-way
+
+\hypertarget{temporal-processing}{%
+\subsection{Temporal processing}\label{temporal-processing}}
+
+\hypertarget{dimension-reduction}{%
+\subsection{Dimension reduction}\label{dimension-reduction}}
+
 sometimes called feature extraction in the machine learning community
@@
 paper.{]}
-
-\hypertarget{temporal-processing}{%
-\subsection{Temporal processing}\label{temporal-processing}}
-
-\hypertarget{dimension-reduction}{%
-\subsection{Dimension reduction}\label{dimension-reduction}}
 
@@
 
-\hypertarget{extending-the-pipeline}{%
-\section{Extending the pipeline}\label{extending-the-pipeline}}
+\hypertarget{incorporating-new-methods-into-the-pipeline}{%
+\section{Incorporating new methods into the
+pipeline}\label{incorporating-new-methods-into-the-pipeline}}
 
@@
 \end{itemize}
+
+\hypertarget{calculating-spei-with-raster-data}{%
+\subsection{Calculating SPEI with raster
+data}\label{calculating-spei-with-raster-data}}
 

### 2022-10-08 paper/paper.qmd
build up on what is/ not an index

@@
 <!-- Index construction is a way to summarise complicated information (used in environmental data). **it basically says nothing** -->
-The complexity of these information may involve a spatial distribution, a temporal scale that defines the frequency of the data, and a multivariate perspective that collects different climate variables.
+<!-- The complexity of these information may involve a spatial distribution, a temporal scale that defines the frequency of the data, and a multivariate perspective that collects different climate variables.
+-->
 
@@
 
-Various methods are proposed to extract multivariate information on the spatial coverage, across time.
-Each individual index follows its own data pipeline and it can be difficult to evaluate how an index can be affected by tweaking parameters in a certain step, rearranging the order of steps, using to different method.
+<!-- Various methods are proposed to extract multivariate information on the spatial coverage, across time. -->
+<!-- Each individual index follows its own data pipeline and it can be difficult to evaluate how an index can be affected by tweaking parameters in a certain step, rearranging the order of steps, using to different method.
+-->
 
-Indices with readily available code to implement have more potential to be computed by researchers, i.e.
+<!-- Indices with readily available code to implement have more potential to be computed by researchers, i.e.
 the package `SPEI` [@SPEI] has implemented the SPI and SPEI index for calculating drought indices.
-[Another package in a different domain].  
-However, even if most of the indices have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember, classify, and compare between.
-There is a need for a framework to generally describe how indices can be constructed from spatio-temporal data.
+-->
+<!-- [Another package in a different domain].   -->
+<!-- However, even if most of the indices have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember, classify, and compare between.
+-->
+<!-- There is a need for a framework to generally describe how indices can be constructed from spatio-temporal data. -->
 
-This paper proposes such data pipeline for constructing indices.
-The steps involved in the pipeline are general in nature and flexible to be adopted to most index construction for environmental data.
+<!-- This paper proposes such data pipeline for constructing indices. -->
+<!-- The steps involved in the pipeline are general in nature and flexible to be adopted to most index construction for environmental data. -->
 
@@
 
-**review indices, climate, drought, economic, UN and other indices**
+  - decision making: need to consider multi-facet for allocating resources, [...]
+  - to quantify concept without direct measure.
+1) almost impossible to include all the items in the market, indices created from a basket of representative items are useful to characterise the market condition; 2) the concept of interest can't be directly measured, i.e.
+livability, human developement etc
+
+Integrate multiple information to produce a single number summary (across time).
+
+**Define what is an index, what is not**
+
+Indices quantify a concept of interest that is hard to measure using [...]. 
+In finance and economic, the behavior of stock market, indices like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite uses a representative set of companies.
+Natural hazard are also indices difficult to define because ....
+Various drought, flood indices have been developed to characterise the severity and monitor from the perspective of meteorology, agricultural, hydrology, and also socio-economy. 
+Examples of indices are ubiquitous: economic indices for quantifying productibility (GDP?), inflation (CPI), [...]; Social development indices like Human Development Index and Global Liveability Index; diversity indices to characterise the richness, evenness and dominance in ecology
+
+Despite many measurements are called indices, they are actually not indices. 
+A direct rate or transformation on a measured variable: birth & death rate computes the total number of new-borns and death per 1000 people. 
+Many remote sensing measures are not indices, i.e.
+Normalized Difference Vegetation Index (NDVI), despite being named as an index, is a mere statistical transformation: the difference of near-infrared (NIR) and red visible channel over its sum.
+
+  <!-- - what is not:  -->
+  <!--   - variable transformation in remote sensing, i.e. NDVI is a variable transformation of near-infrared region (NIR) and red visible band.  -->
+  <!--   - Another set of measures that are not indices (climate index): Precipitation of driest month, Precipitation of warmest month. These are features -->
+
+- need think through: 
+  - is cash rate/ interest rate indices? employment rate? eocnomic indices?
+
+<!-- **review indices, climate, drought, economic, UN and other indices** -->
+
+<!--   - financial and economics index: stock market index (Dow, S&P 500, and Nasdaq), inflation index (CPI, PCE, PPI)  -->
+<!--   - indices in newly emerging policy areas such as industrial competitiveness, sustainable development, globalization and innovation. -->
+<!--   - Social development indices: Human Development Index (HDI): life expectancy + education + gini -->
 
@@
 
-**who would benefit from this paper: 1) people proposing new indices; 2) people analysing index variation, compare between indices, 3) those making decisions from indices, 4) other statistician, how to modularise steps for anlaysis in your domain.**
+  - most indices are presented as calculated numerical or categorical values without open source code accompanied.
+
+  - This causes the problem that if analysts are not agreed with a certain step in the index construction, there is no room for experiment with other possibility.
+
+**who would benefit from this paper**
+
+  - domain scientists who propose new indices and [use multiple indices to compare]
+  - decision makers, towards understand the 
+  - statisticians, so as to [draw focus/ advocate] on statistical workflow
 
@@
 
-standardising
+A specific transformation on the scale of the data
+
+z-score standardising, min-max standardisation into [0, 1] or [0, 100], percentage change on the baseline
 close to variable transformation step
@@
 
+Also include weighting
+
 ## Normalising
+
+The purpose of normalising is for cross-comparison.
+This step can get criticism from analysts for ...
 

### 2022-10-08 paper/paper.tex
build up on what is/ not an index

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, boxrule=0pt, sharp corners, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, interior hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-The complexity of these information may involve a spatial distribution,
-a temporal scale that defines the frequency of the data, and a
-multivariate perspective that collects different climate variables.
-
-Various methods are proposed to extract multivariate information on the
-spatial coverage, across time. Each individual index follows its own
-data pipeline and it can be difficult to evaluate how an index can be
-affected by tweaking parameters in a certain step, rearranging the order
-of steps, using to different method.
-
-Indices with readily available code to implement have more potential to
-be computed by researchers, i.e.~the package \texttt{SPEI} (Beguería and
-Vicente-Serrano 2017) has implemented the SPI and SPEI index for
-calculating drought indices. {[}Another package in a different
-domain{]}.\\
-However, even if most of the indices have available existing
-implementation available to users, there would be hundreds of them and
-it will be burdensome to remember, classify, and compare between. There
-is a need for a framework to generally describe how indices can be
-constructed from spatio-temporal data.
-
-This paper proposes such data pipeline for constructing indices. The
-steps involved in the pipeline are general in nature and flexible to be
-adopted to most index construction for environmental data.
-
 \textbf{Why index is useful, why people care about indices}
 
-\textbf{review indices, climate, drought, economic, UN and other
-indices}
+\begin{itemize}
+\tightlist
+\item
+  decision making: need to consider multi-facet for allocating
+  resources, {[}\ldots{]}
+\item
+  to quantify concept without direct measure. 1) almost impossible to
+  include all the items in the market, indices created from a basket of
+  representative items are useful to characterise the market condition;
+  2) the concept of interest can't be directly measured,
+  i.e.~livability, human developement etc
+\end{itemize}
+
+Integrate multiple information to produce a single number summary
+(across time).
+
+\textbf{Define what is an index, what is not}
+
+Indices quantify a concept of interest that is hard to measure using
+{[}\ldots{]}. In finance and economic, the behavior of stock market,
+indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
+Composite uses a representative set of companies. Natural hazard are
+also indices difficult to define because \ldots. Various drought, flood
+indices have been developed to characterise the severity and monitor
+from the perspective of meteorology, agricultural, hydrology, and also
+socio-economy. Examples of indices are ubiquitous: economic indices for
+quantifying productibility (GDP?), inflation (CPI), {[}\ldots{]}; Social
+development indices like Human Development Index and Global Liveability
+Index; diversity indices to characterise the richness, evenness and
+dominance in ecology
+
+Despite many measurements are called indices, they are actually not
+indices. A direct rate or transformation on a measured variable: birth
+\& death rate computes the total number of new-borns and death per 1000
+people. Many remote sensing measures are not indices, i.e.~Normalized
+Difference Vegetation Index (NDVI), despite being named as an index, is
+a mere statistical transformation: the difference of near-infrared (NIR)
+and red visible channel over its sum.
+
+\begin{itemize}
+\tightlist
+\item
+  need think through:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    is cash rate/ interest rate indices? employment rate? eocnomic
+    indices?
+  \end{itemize}
+\end{itemize}
 
@@
 
-\textbf{who would benefit from this paper: 1) people proposing new
-indices; 2) people analysing index variation, compare between indices,
-3) those making decisions from indices, 4) other statistician, how to
-modularise steps for anlaysis in your domain.}
+\begin{itemize}
+\item
+  most indices are presented as calculated numerical or categorical
+  values without open source code accompanied.
+\item
+  This causes the problem that if analysts are not agreed with a certain
+  step in the index construction, there is no room for experiment with
+  other possibility.
+\end{itemize}
+
+\textbf{who would benefit from this paper}
+
+\begin{itemize}
+\tightlist
+\item
+  domain scientists who propose new indices and {[}use multiple indices
+  to compare{]}
+\item
+  decision makers, towards understand the
+\item
+  statisticians, so as to {[}draw focus/ advocate{]} on statistical
+  workflow
+\end{itemize}
 
@@
 
-standardising close to variable transformation step
+A specific transformation on the scale of the data
+
+z-score standardising, min-max standardisation into {[}0, 1{]} or {[}0,
+100{]}, percentage change on the baseline close to variable
+transformation step
 
@@
 
+Also include weighting
+
 \hypertarget{normalising}{%
 \subsection{Normalising}\label{normalising}}
+
+The purpose of normalising is for cross-comparison. This step can get
+criticism from analysts for \ldots{}
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-SPEI}{}}%
-Beguería, Santiago, and Sergio M. Vicente-Serrano. 2017. \emph{SPEI:
-Calculation of the Standardised Precipitation-Evapotranspiration Index}.
-\url{https://CRAN.R-project.org/package=SPEI}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-buja_elements_1988}{}}%

### 2022-10-10 paper/paper.qmd
more build on introduction

@@
 Indices quantify a concept of interest that is hard to measure using [...].
+over time.
 In finance and economic, the behavior of stock market, indices like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite uses a representative set of companies.
@@
 
-Despite many measurements are called indices, they are actually not indices. 
+Despite many measurements are called indices, they are actually not indices, but metrics. 
 A direct rate or transformation on a measured variable: birth & death rate computes the total number of new-borns and death per 1000 people. 
@@
 Normalized Difference Vegetation Index (NDVI), despite being named as an index, is a mere statistical transformation: the difference of near-infrared (NIR) and red visible channel over its sum.
+
+fire index
 
@@
 
+There has not yet been a standardised workflow on index construction. 
+For simple indices, users may be able to work out the calculation from the equations or workflow provided, while for more complicated construction, software is not even available for open-source consumption.
+Even when the software implementation is available, most indices are computed through a single index command, for example, the package `SPEI` provides two drought indices: SPI and SPEI through function `spi()` and `spei()`.
+This software design focuses on the name of the index and includes all the steps in a single index function. 
+With this design, there is little room for modification. For example, fit a new distribution that is not provided in the initial code. 
+Also, when scaled to a large number of indices, it increases the cognitive load of remembering hundreds of index names. 
+A better approach is to modularise these steps and construct a pipeline for users to build indices through choosing different building blocks from each module.
+With this design, users focus on the operations involved in the index construction.
+If an operation from another index is useful, users can borrow it to modify an existing index.
+
 **who would benefit from this paper**
 
-  - domain scientists who propose new indices and [use multiple indices to compare]
+The article contributes a pipeline framework to construct general indices.
+
+  - domain scientists who propose new indices and directly use indices in analysis. [use multiple indices to compare]
   - decision makers, towards understand the 
-  - statisticians, so as to [draw focus/ advocate] on statistical workflow
+  - statisticians, so as to [draw focus/ advocate] on statistical workflow and software design.
 
-# Data pipeline in R
+The rest of the paper is structured as follows: the concept of data pipeline in R is reviewed in @sec-data-pipeline-in-r.
+@sec-a-pipeline-for-building-statistical-indices presents the data pipeline for index construction.
+@sec-incorporating-new-buliding-blocks-into-the-pipeline explains how to include a new building block in each pipeline module.
+Examples are given in @sec-examples to demonstrate the index construction with the pipeline built.
 
-## Tidy data
+# Data pipeline in R {#sec-data-pipeline-in-r}
+
+## Tidy data {#sec-tidy-data}
 
@@
 
-## Data pipeline
+## Data pipeline {#sec-data-pipeline}
 
@@
 
-A more recent data pipeline is tidymodels [@tidymodels], a set of packages for machine learning models following the tidyverse principles [@tidyverse]. \[expand on tidymodels\]
+A more recent data pipeline is tidymodels [@tidymodels], a set of packages for machine learning models following the tidyverse principles [@tidyverse]. 
+Steps: Exploratory data analysis (EDA), feature engineering, model tuning and selection, and model evaluation.
 
-# A pipeline for building natural hazard indices
+# A pipeline for building statistical indices {#sec-a-pipeline-for-building-statistical-indices}
 
@@
 
-# Incorporating new methods into the pipeline
+# Incorporating new buliding blocks into the pipeline {#sec-incorporating-new-buliding-blocks-into-the-pipeline}
 
@@
 
-# Examples
+# Examples {#sec-examples}
 

### 2022-10-10 paper/paper.tex
more build on introduction

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, interior hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 Indices quantify a concept of interest that is hard to measure using
-{[}\ldots{]}. In finance and economic, the behavior of stock market,
-indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
+{[}\ldots{]}. over time. In finance and economic, the behavior of stock
+market, indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
 Composite uses a representative set of companies. Natural hazard are
@@
 Despite many measurements are called indices, they are actually not
-indices. A direct rate or transformation on a measured variable: birth
-\& death rate computes the total number of new-borns and death per 1000
-people. Many remote sensing measures are not indices, i.e.~Normalized
-Difference Vegetation Index (NDVI), despite being named as an index, is
-a mere statistical transformation: the difference of near-infrared (NIR)
-and red visible channel over its sum.
+indices, but metrics. A direct rate or transformation on a measured
+variable: birth \& death rate computes the total number of new-borns and
+death per 1000 people. Many remote sensing measures are not indices,
+i.e.~Normalized Difference Vegetation Index (NDVI), despite being named
+as an index, is a mere statistical transformation: the difference of
+near-infrared (NIR) and red visible channel over its sum.
+
+fire index
 
@@
 
+There has not yet been a standardised workflow on index construction.
+For simple indices, users may be able to work out the calculation from
+the equations or workflow provided, while for more complicated
+construction, software is not even available for open-source
+consumption. Even when the software implementation is available, most
+indices are computed through a single index command, for example, the
+package \texttt{SPEI} provides two drought indices: SPI and SPEI through
+function \texttt{spi()} and \texttt{spei()}. This software design
+focuses on the name of the index and includes all the steps in a single
+index function. With this design, there is little room for modification.
+For example, fit a new distribution that is not provided in the initial
+code. Also, when scaled to a large number of indices, it increases the
+cognitive load of remembering hundreds of index names. A better approach
+is to modularise these steps and construct a pipeline for users to build
+indices through choosing different building blocks from each module.
+With this design, users focus on the operations involved in the index
+construction. If an operation from another index is useful, users can
+borrow it to modify an existing index.
+
 \textbf{who would benefit from this paper}
+
+The article contributes a pipeline framework to construct general
+indices.
 
@@
 \item
-  domain scientists who propose new indices and {[}use multiple indices
-  to compare{]}
+  domain scientists who propose new indices and directly use indices in
+  analysis. {[}use multiple indices to compare{]}
 \item
@@
   statisticians, so as to {[}draw focus/ advocate{]} on statistical
-  workflow
+  workflow and software design.
 \end{itemize}
 
-\hypertarget{data-pipeline-in-r}{%
-\section{Data pipeline in R}\label{data-pipeline-in-r}}
+The rest of the paper is structured as follows: the concept of data
+pipeline in R is reviewed in Section~\ref{sec-data-pipeline-in-r}.
+Section~\ref{sec-a-pipeline-for-building-statistical-indices} presents
+the data pipeline for index construction.
+Section~\ref{sec-incorporating-new-buliding-blocks-into-the-pipeline}
+explains how to include a new building block in each pipeline module.
+Examples are given in Section~\ref{sec-examples} to demonstrate the
+index construction with the pipeline built.
 
-\hypertarget{tidy-data}{%
-\subsection{Tidy data}\label{tidy-data}}
+\hypertarget{sec-data-pipeline-in-r}{%
+\section{Data pipeline in R}\label{sec-data-pipeline-in-r}}
+
+\hypertarget{sec-tidy-data}{%
+\subsection{Tidy data}\label{sec-tidy-data}}
 
@@
 
-\hypertarget{data-pipeline}{%
-\subsection{Data pipeline}\label{data-pipeline}}
+\hypertarget{sec-data-pipeline}{%
+\subsection{Data pipeline}\label{sec-data-pipeline}}
 
@@
 of packages for machine learning models following the tidyverse
-principles (Wickham et al. 2019). {[}expand on tidymodels{]}
+principles (Wickham et al. 2019). Steps: Exploratory data analysis
+(EDA), feature engineering, model tuning and selection, and model
+evaluation.
 
-\hypertarget{a-pipeline-for-building-natural-hazard-indices}{%
-\section{A pipeline for building natural hazard
-indices}\label{a-pipeline-for-building-natural-hazard-indices}}
+\hypertarget{sec-a-pipeline-for-building-statistical-indices}{%
+\section{A pipeline for building statistical
+indices}\label{sec-a-pipeline-for-building-statistical-indices}}
 
@@
 
-\hypertarget{incorporating-new-methods-into-the-pipeline}{%
-\section{Incorporating new methods into the
-pipeline}\label{incorporating-new-methods-into-the-pipeline}}
+\hypertarget{sec-incorporating-new-buliding-blocks-into-the-pipeline}{%
+\section{Incorporating new buliding blocks into the
+pipeline}\label{sec-incorporating-new-buliding-blocks-into-the-pipeline}}
 
-\hypertarget{examples}{%
-\section{Examples}\label{examples}}
+\hypertarget{sec-examples}{%
+\section{Examples}\label{sec-examples}}
 

### 2022-10-10 paper/paper.qmd
write up on index definition and what's not subsection

@@
   - decision making: need to consider multi-facet for allocating resources, [...]
-  - to quantify concept without direct measure.
-1) almost impossible to include all the items in the market, indices created from a basket of representative items are useful to characterise the market condition; 2) the concept of interest can't be directly measured, i.e.
-livability, human developement etc
+  - to include or constructed to quantify a concept of interest: some papers on constructing tourism based climate indices etc
 
@@
 
-Indices quantify a concept of interest that is hard to measure using [...].
-over time.
-In finance and economic, the behavior of stock market, indices like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite uses a representative set of companies.
-Natural hazard are also indices difficult to define because ....
-Various drought, flood indices have been developed to characterise the severity and monitor from the perspective of meteorology, agricultural, hydrology, and also socio-economy. 
-Examples of indices are ubiquitous: economic indices for quantifying productibility (GDP?), inflation (CPI), [...]; Social development indices like Human Development Index and Global Liveability Index; diversity indices to characterise the richness, evenness and dominance in ecology
+Indices quantify a concept of interest with no direct measure across time.
+The concept of interest doesn't have direct measure can because it is impractical to measure the concept of interest at the population level. 
+For example, it would be nearly impossible to include all the available stocks in the market to characterise stock market behavior, indices like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite select a representative set of stocks to measure the overall market behavior.
+Also belong to this category is the economic indices like Consumer Price Index, where price changes of a basket of items are weighted to measure inflation. 
+The lack of direct measure could also due to the absence of a universal definition on the concept of interest. 
+Many natural and social concepts are fundamentally multi-facets and some indices combines measures from different domains to characterise the concept of interest. 
+This includes drought indices constructed from meteorological, agricultural, hydrological, and social-economic variables [give some names?]. 
+Social development indices like Human Development Index [ref] and Global Liveability Index [ref] measure various aspects on the quality of human capital and urban life.
 
-Despite many measurements are called indices, they are actually not indices, but metrics. 
-A direct rate or transformation on a measured variable: birth & death rate computes the total number of new-borns and death per 1000 people. 
-Many remote sensing measures are not indices, i.e.
-Normalized Difference Vegetation Index (NDVI), despite being named as an index, is a mere statistical transformation: the difference of near-infrared (NIR) and red visible channel over its sum.
+Despite many measurements have index in its name, they cannot be technically classified as indices according to the definition given above. 
+A major reason for these measures to lose its index membership is that the underlying concept of interest is straight forward to measure once defined. 
+Sometime variables like precipitation of driest month or percentage of days when maximum temperature is below a certain percentile are classified as climate indices, however, precipitation and temperature can be directly measured by rain gauge and thermometer.
+These variables are transformation on precipitation and temperature. 
+They can be used to construct indices but are not indices themselves. 
+What is also not indices are many remote sensing based drought measures as they are transformation on one or more remote sensor bands. 
+This includes Normalized Difference Vegetation Index (NDVI) [ref], a transformation difference over sum on near-infrared (NIR) and red visible channel, and Vegetation Condition Index [ref], a rescale of NDVI into [0, 1], among others.
 
-fire index
-
-  <!-- - what is not:  -->
-  <!--   - variable transformation in remote sensing, i.e. NDVI is a variable transformation of near-infrared region (NIR) and red visible band.  -->
-  <!--   - Another set of measures that are not indices (climate index): Precipitation of driest month, Precipitation of warmest month. These are features -->
-
-- need think through: 
-  - is cash rate/ interest rate indices? employment rate? eocnomic indices?
-
-<!-- **review indices, climate, drought, economic, UN and other indices** -->
-
-<!--   - financial and economics index: stock market index (Dow, S&P 500, and Nasdaq), inflation index (CPI, PCE, PPI)  -->
-<!--   - indices in newly emerging policy areas such as industrial competitiveness, sustainable development, globalization and innovation. -->
-<!--   - Social development indices: Human Development Index (HDI): life expectancy + education + gini -->
+TODO: confirm the actual climate indices are indices by our definition
+Questions: need reference for stock market indices, CPI?
 
 **What is the problem with current workflow on index construction**
-
-  - most indices are presented as calculated numerical or categorical values without open source code accompanied.
-
-  - This causes the problem that if analysts are not agreed with a certain step in the index construction, there is no room for experiment with other possibility.
 
@@
   - domain scientists who propose new indices and directly use indices in analysis. [use multiple indices to compare]
+  - general scientists who uses indices as part of analysis, modelling process.
+Official indices are usually published in a public table with values calculated.
+Although methods of calculation are provided by the relevant index provider, raw data and scripts are usually not available for public consumption.
+The pipeline proposed in the paper opens up the possibility for general scientists to construct indices customised to their research purpose.
   - decision makers, towards understand the 

### 2022-10-10 paper/paper.tex
write up on index definition and what's not subsection

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 \item
-  to quantify concept without direct measure. 1) almost impossible to
-  include all the items in the market, indices created from a basket of
-  representative items are useful to characterise the market condition;
-  2) the concept of interest can't be directly measured,
-  i.e.~livability, human developement etc
+  to include or constructed to quantify a concept of interest
 \end{itemize}
@@
 
-Indices quantify a concept of interest that is hard to measure using
-{[}\ldots{]}. over time. In finance and economic, the behavior of stock
-market, indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
-Composite uses a representative set of companies. Natural hazard are
-also indices difficult to define because \ldots. Various drought, flood
-indices have been developed to characterise the severity and monitor
-from the perspective of meteorology, agricultural, hydrology, and also
-socio-economy. Examples of indices are ubiquitous: economic indices for
-quantifying productibility (GDP?), inflation (CPI), {[}\ldots{]}; Social
-development indices like Human Development Index and Global Liveability
-Index; diversity indices to characterise the richness, evenness and
-dominance in ecology
+Indices quantify a concept of interest with no direct measure across
+time. The concept of interest doesn't have direct measure can because it
+is impractical to measure the concept of interest at the population
+level. For example, it would be nearly impossible to include all the
+available stocks in the market to characterise stock market behavior,
+indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
+Composite select a representative set of stocks to measure the overall
+market behavior. Also belong to this category is the economic indices
+like Consumer Price Index, where price changes of a basket of items are
+weighted to measure inflation. The lack of direct measure could also due
+to the absence of a universal definition on the concept of interest.
+Many natural and social concepts are fundamentally multi-facets and some
+indices combines measures from different domains to characterise the
+concept of interest. This includes drought indices constructed from
+meteorological, agricultural, hydrological, and social-economic
+variables {[}give some names?{]}. Social development indices like Human
+Development Index {[}ref{]} and Global Liveability Index {[}ref{]}
+measure various aspects on the quality of human capital and urban life.
 
-Despite many measurements are called indices, they are actually not
-indices, but metrics. A direct rate or transformation on a measured
-variable: birth \& death rate computes the total number of new-borns and
-death per 1000 people. Many remote sensing measures are not indices,
-i.e.~Normalized Difference Vegetation Index (NDVI), despite being named
-as an index, is a mere statistical transformation: the difference of
-near-infrared (NIR) and red visible channel over its sum.
+Despite many measurements have index in its name, they cannot be
+technically classified as indices according to the definition given
+above. A major reason for these measures to lose its index membership is
+that the underlying concept of interest is straight forward to measure
+once defined. Sometime variables like precipitation of driest month or
+percentage of days when maximum temperature is below a certain
+percentile are classified as climate indices, however, precipitation and
+temperature can be directly measured by rain gauge and thermometer.
+These variables are transformation on precipitation and temperature.
+They can be used to construct indices but are not indices themselves.
+What is also not indices are many remote sensing based drought measures
+as they are transformation on one or more remote sensor bands. This
+includes Normalized Difference Vegetation Index (NDVI) {[}ref{]}, a
+transformation difference over sum on near-infrared (NIR) and red
+visible channel, and Vegetation Condition Index {[}ref{]}, a rescale of
+NDVI into {[}0, 1{]}, among others.
 
-fire index
-
-\begin{itemize}
-\tightlist
-\item
-  need think through:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    is cash rate/ interest rate indices? employment rate? eocnomic
-    indices?
-  \end{itemize}
-\end{itemize}
+TODO: confirm the actual climate indices are indices by our definition
+Questions: need reference for stock market indices, CPI?
 
 \textbf{What is the problem with current workflow on index construction}
-
-\begin{itemize}
-\item
-  most indices are presented as calculated numerical or categorical
-  values without open source code accompanied.
-\item
-  This causes the problem that if analysts are not agreed with a certain
-  step in the index construction, there is no room for experiment with
-  other possibility.
-\end{itemize}
 
@@
   analysis. {[}use multiple indices to compare{]}
+\item
+  general scientists who uses indices as part of analysis, modelling
+  process. Official indices are usually published in a public table with
+  values calculated. Although methods of calculation are provided by the
+  relevant index provider, raw data and scripts are usually not
+  available for public consumption. The pipeline proposed in the paper
+  opens up the possibility for general scientists to construct indices
+  customised to their research purpose.
 \item

### 2022-10-11 paper/paper.qmd
A draft on why people should care about indices in intro

@@
 
-* indices, why do we need them, decision making
-* define what is an index, what is not even if they are called index
-* only very brief on indices in the introduction
-* potential audience + what do you think they should learn from this work.
+\newpage
 
@@
 
-  - decision making: need to consider multi-facet for allocating resources, [...]
-  - to include or constructed to quantify a concept of interest: some papers on constructing tourism based climate indices etc
+Quantities that can't be directly measured is ubiquitous in our society and they can often be relevant for policy making, resource allocation, early prevention and operational purpose.
+For example, fire authority would be interested to quantify fire risk since bushfire can have a huge impact on monetary loss, health, and local ecosystem among others.
+Climatologists would be interested in monitoring the change in global condition since variability in atmospheric and oceanic condition has a direct impact on the global weather and climate.
+Usually more than one variable is related to the concept of interest and they often needs to be integrated to make decisions.
+<!-- From a modelling perspective, these unobserved quantities are called latent variables and existing methods can incorporate these latent variables in the model structure.  -->
+A commonly used approach here is to construct an index with the observable variables.
+This allows analysts to compare the quantity of interest across countries and also cross time.
 
-Integrate multiple information to produce a single number summary (across time).
+<!--   - decision making: need to consider multi-facet for allocating resources, [...] -->
+<!--   - to include or constructed to quantify a concept of interest: some papers on constructing tourism based climate indices etc -->
+
+<!-- Integrate multiple information to produce a single number summary (across time). -->
 
@@
 
-Indices quantify a concept of interest with no direct measure across time.
+The meaning of indices varies across disciplines and in this article, an index is defined as a tool to quantify a concept of interest that does not have direct measure.
 The concept of interest doesn't have direct measure can because it is impractical to measure the concept of interest at the population level. 
@@
 
-reusable
+Here we assume a concept of interest is determined, relevant variables/ indicators are identified and available to construct indices.
 

### 2022-10-11 paper/paper.tex
A draft on why people should care about indices in intro

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
-\begin{itemize}
-\tightlist
-\item
-  indices, why do we need them, decision making
-\item
-  define what is an index, what is not even if they are called index
-\item
-  only very brief on indices in the introduction
-\item
-  potential audience + what do you think they should learn from this
-  work.
-\end{itemize}
+\newpage
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  decision making: need to consider multi-facet for allocating
-  resources, {[}\ldots{]}
-\item
-  to include or constructed to quantify a concept of interest
-\end{itemize}
-
-Integrate multiple information to produce a single number summary
-(across time).
+Quantities that can't be directly measured is ubiquitous in our society
+and they can often be relevant for policy making, resource allocation,
+early prevention and operational purpose. For example, fire authority
+would be interested to quantify fire risk since bushfire can have a huge
+impact on monetary loss, health, and local ecosystem among others.
+Climatologists would be interested in monitoring the change in global
+condition since variability in atmospheric and oceanic condition has a
+direct impact on the global weather and climate. Usually more than one
+variable is related to the concept of interest and they often needs to
+be integrated to make decisions. A commonly used approach here is to
+construct an index with the observable variables. This allows analysts
+to compare the quantity of interest across countries and also cross
+time.
 
@@
 
-Indices quantify a concept of interest with no direct measure across
-time. The concept of interest doesn't have direct measure can because it
-is impractical to measure the concept of interest at the population
-level. For example, it would be nearly impossible to include all the
-available stocks in the market to characterise stock market behavior,
-indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
-Composite select a representative set of stocks to measure the overall
-market behavior. Also belong to this category is the economic indices
-like Consumer Price Index, where price changes of a basket of items are
-weighted to measure inflation. The lack of direct measure could also due
-to the absence of a universal definition on the concept of interest.
-Many natural and social concepts are fundamentally multi-facets and some
-indices combines measures from different domains to characterise the
-concept of interest. This includes drought indices constructed from
-meteorological, agricultural, hydrological, and social-economic
-variables {[}give some names?{]}. Social development indices like Human
-Development Index {[}ref{]} and Global Liveability Index {[}ref{]}
-measure various aspects on the quality of human capital and urban life.
+The meaning of indices varies across disciplines and in this article, an
+index is defined as a tool to quantify a concept of interest that does
+not have direct measure. The concept of interest doesn't have direct
+measure can because it is impractical to measure the concept of interest
+at the population level. For example, it would be nearly impossible to
+include all the available stocks in the market to characterise stock
+market behavior, indices like Dow Jones Industrial Average, S\&P 500,
+and Nasdaq Composite select a representative set of stocks to measure
+the overall market behavior. Also belong to this category is the
+economic indices like Consumer Price Index, where price changes of a
+basket of items are weighted to measure inflation. The lack of direct
+measure could also due to the absence of a universal definition on the
+concept of interest. Many natural and social concepts are fundamentally
+multi-facets and some indices combines measures from different domains
+to characterise the concept of interest. This includes drought indices
+constructed from meteorological, agricultural, hydrological, and
+social-economic variables {[}give some names?{]}. Social development
+indices like Human Development Index {[}ref{]} and Global Liveability
+Index {[}ref{]} measure various aspects on the quality of human capital
+and urban life.
 
@@
 
-reusable
+Here we assume a concept of interest is determined, relevant variables/
+indicators are identified and available to construct indices.
 

### 2022-10-11 paper/paper.qmd
refine on why NDVI is not an index + text editing

@@
 
-Quantities that can't be directly measured is ubiquitous in our society and they can often be relevant for policy making, resource allocation, early prevention and operational purpose.
-For example, fire authority would be interested to quantify fire risk since bushfire can have a huge impact on monetary loss, health, and local ecosystem among others.
-Climatologists would be interested in monitoring the change in global condition since variability in atmospheric and oceanic condition has a direct impact on the global weather and climate.
-Usually more than one variable is related to the concept of interest and they often needs to be integrated to make decisions.
+Many variables relevant to decision making cannot be directly measured, however, they are crucial for resource allocation, early prevention, and other operational purpose.
+For example, fire authorities would be interested to quantify fire risk since bushfires can have a huge impact on monetary loss, health, and the local ecosystem among others.
+Climatologists would be interested in monitoring the change in global conditions since variability in atmospheric and oceanic conditions has a direct impact on global weather and climate.
+Usually, more than one variable is related to the concept of interest and they often need to be integrated to make decisions.
 <!-- From a modelling perspective, these unobserved quantities are called latent variables and existing methods can incorporate these latent variables in the model structure.  -->
-A commonly used approach here is to construct an index with the observable variables.
-This allows analysts to compare the quantity of interest across countries and also cross time.
+In many domains, constructing an index is a common approach to characterise these unobservable quantities.
+The index approach allows researchers to compare the quantity of interest across countries and also cross time.
 
@@
 
-The meaning of indices varies across disciplines and in this article, an index is defined as a tool to quantify a concept of interest that does not have direct measure.
-The concept of interest doesn't have direct measure can because it is impractical to measure the concept of interest at the population level. 
+The meaning of indices varies across disciplines and in this article, an index is defined as a tool to quantify a concept of interest that does not have a direct measure.
+The concept of interest doesn't have a direct measure can because it is impractical to measure the concept of interest at the population level. 
 For example, it would be nearly impossible to include all the available stocks in the market to characterise stock market behavior, indices like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite select a representative set of stocks to measure the overall market behavior.
-Also belong to this category is the economic indices like Consumer Price Index, where price changes of a basket of items are weighted to measure inflation. 
-The lack of direct measure could also due to the absence of a universal definition on the concept of interest. 
-Many natural and social concepts are fundamentally multi-facets and some indices combines measures from different domains to characterise the concept of interest. 
+Also belonging to this category are the economic indices like the Consumer Price Index, where price changes of a basket of items are weighted to measure inflation. 
+The lack of direct measure could also be due to the absence of a universal definition on the concept of interest. 
+Many natural and social concepts are fundamentally multi-facets and some indices combine measures from different domains to characterise the concept of interest. 
 This includes drought indices constructed from meteorological, agricultural, hydrological, and social-economic variables [give some names?]. 
-Social development indices like Human Development Index [ref] and Global Liveability Index [ref] measure various aspects on the quality of human capital and urban life.
+Social development indices like Human Development Index [@hdi] and Global Liveability Index [@gli] measure various aspects of the quality of human capital and urban life.
 
-Despite many measurements have index in its name, they cannot be technically classified as indices according to the definition given above. 
-A major reason for these measures to lose its index membership is that the underlying concept of interest is straight forward to measure once defined. 
-Sometime variables like precipitation of driest month or percentage of days when maximum temperature is below a certain percentile are classified as climate indices, however, precipitation and temperature can be directly measured by rain gauge and thermometer.
+Despite many measurements having the term *index* in their name, they cannot be technically classified as indices according to the definition given above. 
+A major reason for these measures to lose their index memberships is that the underlying concept of interest is straightforward to measure once defined. 
+Sometimes variables like precipitation of the driest month or percentage of days when maximum temperature is below a certain percentile are classified as climate indices, however, precipitation and temperature can be directly measured by rain gauge and thermometer.
 These variables are transformation on precipitation and temperature. 
 They can be used to construct indices but are not indices themselves. 
-What is also not indices are many remote sensing based drought measures as they are transformation on one or more remote sensor bands. 
-This includes Normalized Difference Vegetation Index (NDVI) [ref], a transformation difference over sum on near-infrared (NIR) and red visible channel, and Vegetation Condition Index [ref], a rescale of NDVI into [0, 1], among others.
+Similarly, a set of remote sensing indices should not be classified as indices, but variables, since they are measures of electromagnetic wave reflectance.
+This includes Normalized Difference Vegetation Index (NDVI) [@tucker_red_1979], derived from the ratio of difference over sum on two segments in the spectrum, also called band: near-infrared (NIR) and red.
+So are the "indices" derived from NDVI, e.g. Vegetation Condition Index [@kogan_application_1995]. 
+Notice that this does not exclude all the construction derived from remotes sensor variables to be valid indices.
+For example, Vegetation Drought Response Index [@brown_vegetation_2008] is a valid index since it integrates climate, satellite, and biophysical variables to quantify vegetation stress.
 
-TODO: confirm the actual climate indices are indices by our definition
-Questions: need reference for stock market indices, CPI?
+TODO: clean up terminology:
+
+  - "index" only for valid indices
+  - "variable" the most generic dump????
+  - "measure" for item quantified by instruments or devices in standard units
+
+FYI:
+
+  - Vegetation Condition Index  = (NDVI - NDVI_min)/ (NDVI_max - NDVI_min)
+  - A measure is a variable, the variable itself is not an index. 
+  - Not all multivariate "indices" are indices (i.e.
+those remote sensor transformations), but generally they are.
+It is possible that an index is constructed from a single variable (i.e.
+SPI), but be careful on whether it is variable or index.
 
@@
 
+A conversion of non-normal variables into (approximately) normal.
+
 GAM, can you do additive model pairwise/ three-way

### 2022-10-11 paper/paper.tex
refine on why NDVI is not an index + text editing

--- 2022-10-10 paper/paper.tex
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable, boxrule=0pt, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
-\begin{itemize}
-\tightlist
-\item
-  indices, why do we need them, decision making
-\item
-  define what is an index, what is not even if they are called index
-\item
-  only very brief on indices in the introduction
-\item
-  potential audience + what do you think they should learn from this
-  work.
-\end{itemize}
+\newpage
 
@@
 
+Many variables relevant to decision making cannot be directly measured,
+however, they are crucial for resource allocation, early prevention, and
+other operational purpose. For example, fire authorities would be
+interested to quantify fire risk since bushfires can have a huge impact
+on monetary loss, health, and the local ecosystem among others.
+Climatologists would be interested in monitoring the change in global
+conditions since variability in atmospheric and oceanic conditions has a
+direct impact on global weather and climate. Usually, more than one
+variable is related to the concept of interest and they often need to be
+integrated to make decisions. In many domains, constructing an index is
+a common approach to characterise these unobservable quantities. The
+index approach allows researchers to compare the quantity of interest
+across countries and also cross time.
+
+\textbf{Define what is an index, what is not}
+
+The meaning of indices varies across disciplines and in this article, an
+index is defined as a tool to quantify a concept of interest that does
+not have a direct measure. The concept of interest doesn't have a direct
+measure can because it is impractical to measure the concept of interest
+at the population level. For example, it would be nearly impossible to
+include all the available stocks in the market to characterise stock
+market behavior, indices like Dow Jones Industrial Average, S\&P 500,
+and Nasdaq Composite select a representative set of stocks to measure
+the overall market behavior. Also belonging to this category are the
+economic indices like the Consumer Price Index, where price changes of a
+basket of items are weighted to measure inflation. The lack of direct
+measure could also be due to the absence of a universal definition on
+the concept of interest. Many natural and social concepts are
+fundamentally multi-facets and some indices combine measures from
+different domains to characterise the concept of interest. This includes
+drought indices constructed from meteorological, agricultural,
+hydrological, and social-economic variables {[}give some names?{]}.
+Social development indices like Human Development Index (United Nations
+Development Programme 2022) and Global Liveability Index (Economist
+Intelligence Unit 2019) measure various aspects of the quality of human
+capital and urban life.
+
+Despite many measurements having the term \emph{index} in their name,
+they cannot be technically classified as indices according to the
+definition given above. A major reason for these measures to lose their
+index memberships is that the underlying concept of interest is
+straightforward to measure once defined. Sometimes variables like
+precipitation of the driest month or percentage of days when maximum
+temperature is below a certain percentile are classified as climate
+indices, however, precipitation and temperature can be directly measured
+by rain gauge and thermometer. These variables are transformation on
+precipitation and temperature. They can be used to construct indices but
+are not indices themselves. Similarly, a set of remote sensing indices
+should not be classified as indices, but variables, since they are
+measures of electromagnetic wave reflectance. This includes Normalized
+Difference Vegetation Index (NDVI) (Tucker 1979), derived from the ratio
+of difference over sum on two segments in the spectrum, also called
+band: near-infrared (NIR) and red. So are the ``indices'' derived from
+NDVI, e.g.~Vegetation Condition Index (Kogan 1995). Notice that this
+does not exclude all the construction derived from remotes sensor
+variables to be valid indices. For example, Vegetation Drought Response
+Index (Brown et al. 2008) is a valid index since it integrates climate,
+satellite, and biophysical variables to quantify vegetation stress.
+
+TODO: clean up terminology:
+
 \begin{itemize}
@@
 \item
-  decision making: need to consider multi-facet for allocating
-  resources, {[}\ldots{]}
+  ``index'' only for valid indices
 \item
-  to include or constructed to quantify a concept of interest
+  ``variable'' the most generic dump????
+\item
+  ``measure'' for item quantified by instruments or devices in standard
+  units
 \end{itemize}
 
-Integrate multiple information to produce a single number summary
-(across time).
+FYI:
 
-\textbf{Define what is an index, what is not}
-
-Indices quantify a concept of interest with no direct measure across
-time. The concept of interest doesn't have direct measure can because it
-is impractical to measure the concept of interest at the population
-level. For example, it would be nearly impossible to include all the
-available stocks in the market to characterise stock market behavior,
-indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
-Composite select a representative set of stocks to measure the overall
-market behavior. Also belong to this category is the economic indices
-like Consumer Price Index, where price changes of a basket of items are
-weighted to measure inflation. The lack of direct measure could also due
-to the absence of a universal definition on the concept of interest.
-Many natural and social concepts are fundamentally multi-facets and some
-indices combines measures from different domains to characterise the
-concept of interest. This includes drought indices constructed from
-meteorological, agricultural, hydrological, and social-economic
-variables {[}give some names?{]}. Social development indices like Human
-Development Index {[}ref{]} and Global Liveability Index {[}ref{]}
-measure various aspects on the quality of human capital and urban life.
-
-Despite many measurements have index in its name, they cannot be
-technically classified as indices according to the definition given
-above. A major reason for these measures to lose its index membership is
-that the underlying concept of interest is straight forward to measure
-once defined. Sometime variables like precipitation of driest month or
-percentage of days when maximum temperature is below a certain
-percentile are classified as climate indices, however, precipitation and
-temperature can be directly measured by rain gauge and thermometer.
-These variables are transformation on precipitation and temperature.
-They can be used to construct indices but are not indices themselves.
-What is also not indices are many remote sensing based drought measures
-as they are transformation on one or more remote sensor bands. This
-includes Normalized Difference Vegetation Index (NDVI) {[}ref{]}, a
-transformation difference over sum on near-infrared (NIR) and red
-visible channel, and Vegetation Condition Index {[}ref{]}, a rescale of
-NDVI into {[}0, 1{]}, among others.
-
-TODO: confirm the actual climate indices are indices by our definition
-Questions: need reference for stock market indices, CPI?
+\begin{itemize}
+\tightlist
+\item
+  Vegetation Condition Index = (NDVI - NDVI\_min)/ (NDVI\_max -
+  NDVI\_min)
+\item
+  A measure is a variable, the variable itself is not an index.
+\item
+  Not all multivariate ``indices'' are indices (i.e.~those remote sensor
+  transformations), but generally they are. It is possible that an index
+  is constructed from a single variable (i.e.~SPI), but be careful on
+  whether it is variable or index.
+\end{itemize}
 
@@
 also non-linear
+
+A conversion of non-normal variables into (approximately) normal.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-brown_vegetation_2008}{}}%
+Brown, Jesslyn F., Brian D. Wardlow, Tsegaye Tadesse, Michael J. Hayes,
+and Bradley C. Reed. 2008. {``The {Vegetation} {Drought} {Response}
+{Index} ({VegDRI}): {A} {New} {Integrated} {Approach} for {Monitoring}
+{Drought} {Stress} in {Vegetation}.''} \emph{GIScience \& Remote
+Sensing} 45 (1): 16--46.
+\url{https://doi.org/10.2747/1548-1603.45.1.16}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-buja_elements_1988}{}}%
@@
 Statistics}, 277--308. Wadsworth, Belmont.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gli}{}}%
+Economist Intelligence Unit. 2019. {``The Global Liveability Index
+2019.''} The Economist.
+\url{https://www.cbeinternational.ca/pdf/Liveability-Free-report-2019.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kogan_application_1995}{}}%
+Kogan, F. N. 1995. {``Application of Vegetation Index and Brightness
+Temperature for Drought Detection.''} \emph{Advances in Space Research},
+Natural {Hazards}: {Monitoring} and {Assessment} {Using} {Remote}
+{Sensing} {Technique}, 15 (11): 91--100.
+\url{https://doi.org/10.1016/0273-1177(95)00079-T}.
 
@@
 \url{https://www.jstor.org/stable/1390943}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tucker_red_1979}{}}%
+Tucker, Compton J. 1979. {``Red and Photographic Infrared Linear
+Combinations for Monitoring Vegetation.''} \emph{Remote Sensing of
+Environment} 8 (2): 127--50.
+\url{https://doi.org/10.1016/0034-4257(79)90013-0}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
+United Nations Development Programme. 2022. {``Human Development Report
+2021-22.''} New York. \url{http://report.hdr.undp.org}.
 

### 2022-10-12 paper/paper.qmd
more refine

@@
 
-Many variables relevant to decision making cannot be directly measured, however, they are crucial for resource allocation, early prevention, and other operational purpose.
-For example, fire authorities would be interested to quantify fire risk since bushfires can have a huge impact on monetary loss, health, and the local ecosystem among others.
-Climatologists would be interested in monitoring the change in global conditions since variability in atmospheric and oceanic conditions has a direct impact on global weather and climate.
-Usually, more than one variable is related to the concept of interest and they often need to be integrated to make decisions.
+Many concepts relevant to decision making cannot be directly measured, however, they are crucial for resource allocation, early prevention, and other operational purpose.
+For example, fire authorities would be interested to quantify fire risk since bushfires can have a huge impact on monetary loss, health, and the local ecosystem.
+Climatologists would be interested in monitoring the change in global climate since variability in atmospheric and oceanic conditions has a direct impact on global weather and climate.
+Usually this concept of interest is associated with more than one variables and these variables need to be integrated to make decisions on the subject matter.
 <!-- From a modelling perspective, these unobserved quantities are called latent variables and existing methods can incorporate these latent variables in the model structure.  -->
-In many domains, constructing an index is a common approach to characterise these unobservable quantities.
-The index approach allows researchers to compare the quantity of interest across countries and also cross time.
+A common approach to quantify concepts like these is to construct an index using these relevant variables. 
+This allows researchers to compare the quantity of interest across entities (i.e. countries, regions) and also cross time.
 
@@
 The meaning of indices varies across disciplines and in this article, an index is defined as a tool to quantify a concept of interest that does not have a direct measure.
-The concept of interest doesn't have a direct measure can because it is impractical to measure the concept of interest at the population level. 
-For example, it would be nearly impossible to include all the available stocks in the market to characterise stock market behavior, indices like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite select a representative set of stocks to measure the overall market behavior.
+The concept of interest doesn't have a direct measure can because it is impractical to measure at the population level. 
+For example, it would be nearly impossible to include all the available stocks in the market to characterise stock market behavior, so indices like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite select a representative set of stocks to measure the overall market behavior.
 Also belonging to this category are the economic indices like the Consumer Price Index, where price changes of a basket of items are weighted to measure inflation. 
-The lack of direct measure could also be due to the absence of a universal definition on the concept of interest. 
-Many natural and social concepts are fundamentally multi-facets and some indices combine measures from different domains to characterise the concept of interest. 
-This includes drought indices constructed from meteorological, agricultural, hydrological, and social-economic variables [give some names?]. 
+The lack of direct measure could also because the concept itself is an unobservable human construction, rather than a physical quantity that can be measured.
+Many natural hazard and social concepts falls into this category. 
+This includes drought indices constructed from meteorological, agricultural, hydrological, and social-economic variables, e.g.
+Standardised Precipitation Index (SPI) [@mckee1993relationship] and Aggregated Drought Index (ADI) [@keyantash_aggregate_2004] among others.
 Social development indices like Human Development Index [@hdi] and Global Liveability Index [@gli] measure various aspects of the quality of human capital and urban life.
 
-Despite many measurements having the term *index* in their name, they cannot be technically classified as indices according to the definition given above. 
-A major reason for these measures to lose their index memberships is that the underlying concept of interest is straightforward to measure once defined. 
-Sometimes variables like precipitation of the driest month or percentage of days when maximum temperature is below a certain percentile are classified as climate indices, however, precipitation and temperature can be directly measured by rain gauge and thermometer.
-These variables are transformation on precipitation and temperature. 
-They can be used to construct indices but are not indices themselves. 
-Similarly, a set of remote sensing indices should not be classified as indices, but variables, since they are measures of electromagnetic wave reflectance.
+Despite many quantity having the term *index* in their name, they cannot be technically classified as indices according to the definition given above. 
+The reason for these quantities to lose their index memberships is that they are variables can be accurately measured given the instrument precision. 
+This includes quantities like precipitation of the driest month or percentage of days when maximum temperature is below 10th percentile. 
+They are measures of precipitation and percentage of days under specific conditions (dries month, maximum temperature below 10th percentile).
+They are variables, or indicators, that can be used to construct indices but are not indices themselves. 
+Similarly, a set of remote sensing indices are not indices, since they are measures of electromagnetic wave reflectance.
 This includes Normalized Difference Vegetation Index (NDVI) [@tucker_red_1979], derived from the ratio of difference over sum on two segments in the spectrum, also called band: near-infrared (NIR) and red.
@@
 
-TODO: clean up terminology:
-
-  - "index" only for valid indices
-  - "variable" the most generic dump????
-  - "measure" for item quantified by instruments or devices in standard units
-
 FYI:
@@
   - Vegetation Condition Index  = (NDVI - NDVI_min)/ (NDVI_max - NDVI_min)
-  - A measure is a variable, the variable itself is not an index. 
-  - Not all multivariate "indices" are indices (i.e.
-those remote sensor transformations), but generally they are.
-It is possible that an index is constructed from a single variable (i.e.
-SPI), but be careful on whether it is variable or index.
+  - A measure is a variable, but is not an index. 
+  - Not all multivariate quantities are indices (i.e. those remote sensor measures). It is possible that an index is constructed from a single variable (i.e. SPI).
+  - Indices is not a single/ set of measures
 

### 2022-10-12 paper/paper.tex
more refine

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-Many variables relevant to decision making cannot be directly measured,
+Many concepts relevant to decision making cannot be directly measured,
 however, they are crucial for resource allocation, early prevention, and
@@
 interested to quantify fire risk since bushfires can have a huge impact
-on monetary loss, health, and the local ecosystem among others.
-Climatologists would be interested in monitoring the change in global
-conditions since variability in atmospheric and oceanic conditions has a
-direct impact on global weather and climate. Usually, more than one
-variable is related to the concept of interest and they often need to be
-integrated to make decisions. In many domains, constructing an index is
-a common approach to characterise these unobservable quantities. The
-index approach allows researchers to compare the quantity of interest
-across countries and also cross time.
+on monetary loss, health, and the local ecosystem. Climatologists would
+be interested in monitoring the change in global climate since
+variability in atmospheric and oceanic conditions has a direct impact on
+global weather and climate. Usually this concept of interest is
+associated with more than one variables and these variables need to be
+integrated to make decisions on the subject matter. A common approach to
+quantify concepts like these is to construct an index using these
+relevant variables. This allows researchers to compare the quantity of
+interest across entities (i.e.~countries, regions) and also cross time.
 
@@
 not have a direct measure. The concept of interest doesn't have a direct
-measure can because it is impractical to measure the concept of interest
-at the population level. For example, it would be nearly impossible to
-include all the available stocks in the market to characterise stock
-market behavior, indices like Dow Jones Industrial Average, S\&P 500,
-and Nasdaq Composite select a representative set of stocks to measure
-the overall market behavior. Also belonging to this category are the
-economic indices like the Consumer Price Index, where price changes of a
-basket of items are weighted to measure inflation. The lack of direct
-measure could also be due to the absence of a universal definition on
-the concept of interest. Many natural and social concepts are
-fundamentally multi-facets and some indices combine measures from
-different domains to characterise the concept of interest. This includes
-drought indices constructed from meteorological, agricultural,
-hydrological, and social-economic variables {[}give some names?{]}.
-Social development indices like Human Development Index (United Nations
-Development Programme 2022) and Global Liveability Index (Economist
-Intelligence Unit 2019) measure various aspects of the quality of human
-capital and urban life.
+measure can because it is impractical to measure at the population
+level. For example, it would be nearly impossible to include all the
+available stocks in the market to characterise stock market behavior, so
+indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
+Composite select a representative set of stocks to measure the overall
+market behavior. Also belonging to this category are the economic
+indices like the Consumer Price Index, where price changes of a basket
+of items are weighted to measure inflation. The lack of direct measure
+could also because the concept itself is an unobservable human
+construction, rather than a physical quantity that can be measured. Many
+natural hazard and social concepts falls into this category. This
+includes drought indices constructed from meteorological, agricultural,
+hydrological, and social-economic variables, e.g.~Standardised
+Precipitation Index (SPI) (McKee et al. 1993) and Aggregated Drought
+Index (ADI) (Keyantash and Dracup 2004) among others. Social development
+indices like Human Development Index (United Nations Development
+Programme 2022) and Global Liveability Index (Economist Intelligence
+Unit 2019) measure various aspects of the quality of human capital and
+urban life.
 
-Despite many measurements having the term \emph{index} in their name,
-they cannot be technically classified as indices according to the
-definition given above. A major reason for these measures to lose their
-index memberships is that the underlying concept of interest is
-straightforward to measure once defined. Sometimes variables like
-precipitation of the driest month or percentage of days when maximum
-temperature is below a certain percentile are classified as climate
-indices, however, precipitation and temperature can be directly measured
-by rain gauge and thermometer. These variables are transformation on
-precipitation and temperature. They can be used to construct indices but
-are not indices themselves. Similarly, a set of remote sensing indices
-should not be classified as indices, but variables, since they are
-measures of electromagnetic wave reflectance. This includes Normalized
-Difference Vegetation Index (NDVI) (Tucker 1979), derived from the ratio
-of difference over sum on two segments in the spectrum, also called
-band: near-infrared (NIR) and red. So are the ``indices'' derived from
-NDVI, e.g.~Vegetation Condition Index (Kogan 1995). Notice that this
-does not exclude all the construction derived from remotes sensor
-variables to be valid indices. For example, Vegetation Drought Response
-Index (Brown et al. 2008) is a valid index since it integrates climate,
-satellite, and biophysical variables to quantify vegetation stress.
-
-TODO: clean up terminology:
-
-\begin{itemize}
-\tightlist
-\item
-  ``index'' only for valid indices
-\item
-  ``variable'' the most generic dump????
-\item
-  ``measure'' for item quantified by instruments or devices in standard
-  units
-\end{itemize}
+Despite many quantity having the term \emph{index} in their name, they
+cannot be technically classified as indices according to the definition
+given above. The reason for these quantities to lose their index
+memberships is that they are variables can be accurately measured given
+the instrument precision. This includes quantities like precipitation of
+the driest month or percentage of days when maximum temperature is below
+10th percentile. They are measures of precipitation and percentage of
+days under specific conditions (dries month, maximum temperature below
+10th percentile). They are variables, or indicators, that can be used to
+construct indices but are not indices themselves. Similarly, a set of
+remote sensing indices are not indices, since they are measures of
+electromagnetic wave reflectance. This includes Normalized Difference
+Vegetation Index (NDVI) (Tucker 1979), derived from the ratio of
+difference over sum on two segments in the spectrum, also called band:
+near-infrared (NIR) and red. So are the ``indices'' derived from NDVI,
+e.g.~Vegetation Condition Index (Kogan 1995). Notice that this does not
+exclude all the construction derived from remotes sensor variables to be
+valid indices. For example, Vegetation Drought Response Index (Brown et
+al. 2008) is a valid index since it integrates climate, satellite, and
+biophysical variables to quantify vegetation stress.
 
@@
 \item
-  A measure is a variable, the variable itself is not an index.
+  A measure is a variable, but is not an index.
 \item
-  Not all multivariate ``indices'' are indices (i.e.~those remote sensor
-  transformations), but generally they are. It is possible that an index
-  is constructed from a single variable (i.e.~SPI), but be careful on
-  whether it is variable or index.
+  Not all multivariate quantities are indices (i.e.~those remote sensor
+  measures). It is possible that an index is constructed from a single
+  variable (i.e.~SPI).
+\item
+  Indices is not a single/ set of measures
 \end{itemize}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-keyantash_aggregate_2004}{}}%
+Keyantash, John A., and John A. Dracup. 2004. {``An Aggregate Drought
+Index: {Assessing} Drought Severity Based on Fluctuations in the
+Hydrologic Cycle and Surface Water Storage.''} \emph{Water Resources
+Research} 40 (9). \url{https://doi.org/10.1029/2003WR002610}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-kogan_application_1995}{}}%
@@
 \url{https://www.tidymodels.org}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mckee1993relationship}{}}%
+McKee, Thomas B, Nolan J Doesken, John Kleist, et al. 1993. {``The
+Relationship of Drought Frequency and Duration to Time Scales.''} In
+\emph{Proceedings of the 8th Conference on Applied Climatology},
+17:179--83. 22. Boston, MA, USA.
 

### 2022-10-13 paper/paper.qmd
add changes/ todos after meeting

@@
 
-<!-- Index construction is a way to summarise complicated information (used in environmental data). **it basically says nothing** -->
-<!-- The complexity of these information may involve a spatial distribution, a temporal scale that defines the frequency of the data, and a multivariate perspective that collects different climate variables.
--->
-
-<!-- These indices are customised to distinct local weather condition and historical data availability. -->
-
-<!-- Various methods are proposed to extract multivariate information on the spatial coverage, across time. -->
 <!-- Each individual index follows its own data pipeline and it can be difficult to evaluate how an index can be affected by tweaking parameters in a certain step, rearranging the order of steps, using to different method.
@@
 
-<!-- Indices with readily available code to implement have more potential to be computed by researchers, i.e.
-the package `SPEI` [@SPEI] has implemented the SPI and SPEI index for calculating drought indices.
--->
-<!-- [Another package in a different domain].   -->
 <!-- However, even if most of the indices have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember, classify, and compare between.
 -->
-<!-- There is a need for a framework to generally describe how indices can be constructed from spatio-temporal data. -->
-
-<!-- This paper proposes such data pipeline for constructing indices. -->
-<!-- The steps involved in the pipeline are general in nature and flexible to be adopted to most index construction for environmental data. -->
 
 **Why index is useful, why people care about indices**
+
+*incorporate the following in why using index: multiple pieces of information (variables) that need to be taken into account*
 
@@
 
-<!--   - decision making: need to consider multi-facet for allocating resources, [...] -->
-<!--   - to include or constructed to quantify a concept of interest: some papers on constructing tourism based climate indices etc -->
-
-<!-- Integrate multiple information to produce a single number summary (across time). -->
-
 **Define what is an index, what is not**
 
-The meaning of indices varies across disciplines and in this article, an index is defined as a tool to quantify a concept of interest that does not have a direct measure.
+In this article, an index is defined as a tool to quantify a concept of interest that does not have a direct measure.
 The concept of interest doesn't have a direct measure can because it is impractical to measure at the population level. 
@@
 Social development indices like Human Development Index [@hdi] and Global Liveability Index [@gli] measure various aspects of the quality of human capital and urban life.
+
+*still need to tweak the tone a bit: "they are called index, they are not the index we will talk about"*
 
@@
 
-FYI:
+<!-- FYI:  -->
 
-  - Vegetation Condition Index  = (NDVI - NDVI_min)/ (NDVI_max - NDVI_min)
-  - A measure is a variable, but is not an index. 
-  - Not all multivariate quantities are indices (i.e. those remote sensor measures). It is possible that an index is constructed from a single variable (i.e. SPI).
-  - Indices is not a single/ set of measures
+<!--   - A measure is a variable, but is not an index.  -->
+<!--   - Not all multivariate quantities are indices (i.e. those remote sensor measures). It is possible that an index is constructed from a single variable (i.e. SPI). -->
+<!--   - Indices is not a single/ set of measures -->
 
-**What is the problem with current workflow on index construction**
+\newpage
+
+**What is the problem with current index construction**
+
+*indices constructed separately, but there are lots of are same [!], but hard to see from the current construction.
+"messy data are messy in there own way".
+Describe how they are same, so we can modular them*
 
@@
 For simple indices, users may be able to work out the calculation from the equations or workflow provided, while for more complicated construction, software is not even available for open-source consumption.
-Even when the software implementation is available, most indices are computed through a single index command, for example, the package `SPEI` provides two drought indices: SPI and SPEI through function `spi()` and `spei()`.
+Even when the software implementation is available, most indices are computed through a single index command, [given examples but not too domain specific]. 
 This software design focuses on the name of the index and includes all the steps in a single index function. 
@@
 
-The article contributes a pipeline framework to construct general indices.
+*this work could be interested to scientists actively developing new indices*
 
@@
   - general scientists who uses indices as part of analysis, modelling process.
-Official indices are usually published in a public table with values calculated.
-Although methods of calculation are provided by the relevant index provider, raw data and scripts are usually not available for public consumption.
 The pipeline proposed in the paper opens up the possibility for general scientists to construct indices customised to their research purpose.
-  - decision makers, towards understand the 
   - statisticians, so as to [draw focus/ advocate] on statistical workflow and software design.
+
+**what can be done if people adopt this pipeline/ why it is beneficial?**
+
+- transparency: official published, can decompose the indices, 
+- allow for calculation on uncertainty with randomization, 
+- compare indices,
+- tweak/ swap building blocks, 
+- create new indices based on combinations of existing steps.
 
@@
 Steps: Exploratory data analysis (EDA), feature engineering, model tuning and selection, and model evaluation.
+
+- easy to operate properly: should be designed that users know what is approriate to do
+- promote good scientific methodology: should protect uses from doing stupid things,  doing right thing
+
+good statistical practice
 
@@
 
+## Normalising
+
+The purpose of normalising is for cross-comparison.
+This step can get criticism from analysts for ...
+
+specifically for converting from a fitted distribution to normal score via reverse CDF function, non-parametric formula, or empirical approximation, a common step in many index: SPI, SSI, Z score.
+The purpose of normalising is to convert the index into a standardised series after all the steps for the ease of comparison.
+
+Normalising is usually the last step
+
 ## Variable transformation
@@
 
-A conversion of non-normal variables into (approximately) normal.
+change the shape of the variable
+
+<!-- A conversion of non-normal variables into (approximately) normal. wrong! -->
 
@@
 Also include weighting
-
-## Normalising
-
-The purpose of normalising is for cross-comparison.
-This step can get criticism from analysts for ...
-
-specifically for converting from a fitted distribution to normal score via reverse CDF function, non-parametric formula, or empirical approximation, a common step in many index: SPI, SSI, Z score.
-The purpose of normalising is to convert the index into a standardised series after all the steps for the ease of comparison.
 

### 2022-10-13 paper/paper.tex
add changes/ todos after meeting

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 \textbf{Why index is useful, why people care about indices}
+
+\emph{incorporate the following in why using index: multiple pieces of
+information (variables) that need to be taken into account}
 
@@
 
-The meaning of indices varies across disciplines and in this article, an
-index is defined as a tool to quantify a concept of interest that does
-not have a direct measure. The concept of interest doesn't have a direct
-measure can because it is impractical to measure at the population
-level. For example, it would be nearly impossible to include all the
-available stocks in the market to characterise stock market behavior, so
-indices like Dow Jones Industrial Average, S\&P 500, and Nasdaq
-Composite select a representative set of stocks to measure the overall
-market behavior. Also belonging to this category are the economic
-indices like the Consumer Price Index, where price changes of a basket
-of items are weighted to measure inflation. The lack of direct measure
-could also because the concept itself is an unobservable human
+In this article, an index is defined as a tool to quantify a concept of
+interest that does not have a direct measure. The concept of interest
+doesn't have a direct measure can because it is impractical to measure
+at the population level. For example, it would be nearly impossible to
+include all the available stocks in the market to characterise stock
+market behavior, so indices like Dow Jones Industrial Average, S\&P 500,
+and Nasdaq Composite select a representative set of stocks to measure
+the overall market behavior. Also belonging to this category are the
+economic indices like the Consumer Price Index, where price changes of a
+basket of items are weighted to measure inflation. The lack of direct
+measure could also because the concept itself is an unobservable human
 construction, rather than a physical quantity that can be measured. Many
@@
 urban life.
+
+\emph{still need to tweak the tone a bit: ``they are called index, they
+are not the index we will talk about''}
 
@@
 
-FYI:
+\newpage
 
-\begin{itemize}
-\tightlist
-\item
-  Vegetation Condition Index = (NDVI - NDVI\_min)/ (NDVI\_max -
-  NDVI\_min)
-\item
-  A measure is a variable, but is not an index.
-\item
-  Not all multivariate quantities are indices (i.e.~those remote sensor
-  measures). It is possible that an index is constructed from a single
-  variable (i.e.~SPI).
-\item
-  Indices is not a single/ set of measures
-\end{itemize}
+\textbf{What is the problem with current index construction}
 
-\textbf{What is the problem with current workflow on index construction}
+\emph{indices constructed separately, but there are lots of are same
+{[}!{]}, but hard to see from the current construction. ``messy data are
+messy in there own way''. Describe how they are same, so we can modular
+them}
 
@@
 consumption. Even when the software implementation is available, most
-indices are computed through a single index command, for example, the
-package \texttt{SPEI} provides two drought indices: SPI and SPEI through
-function \texttt{spi()} and \texttt{spei()}. This software design
-focuses on the name of the index and includes all the steps in a single
-index function. With this design, there is little room for modification.
-For example, fit a new distribution that is not provided in the initial
-code. Also, when scaled to a large number of indices, it increases the
-cognitive load of remembering hundreds of index names. A better approach
-is to modularise these steps and construct a pipeline for users to build
-indices through choosing different building blocks from each module.
-With this design, users focus on the operations involved in the index
-construction. If an operation from another index is useful, users can
-borrow it to modify an existing index.
+indices are computed through a single index command, {[}given examples
+but not too domain specific{]}. This software design focuses on the name
+of the index and includes all the steps in a single index function. With
+this design, there is little room for modification. For example, fit a
+new distribution that is not provided in the initial code. Also, when
+scaled to a large number of indices, it increases the cognitive load of
+remembering hundreds of index names. A better approach is to modularise
+these steps and construct a pipeline for users to build indices through
+choosing different building blocks from each module. With this design,
+users focus on the operations involved in the index construction. If an
+operation from another index is useful, users can borrow it to modify an
+existing index.
 
@@
 
-The article contributes a pipeline framework to construct general
-indices.
+\emph{this work could be interested to scientists actively developing
+new indices}
 
@@
   general scientists who uses indices as part of analysis, modelling
-  process. Official indices are usually published in a public table with
-  values calculated. Although methods of calculation are provided by the
-  relevant index provider, raw data and scripts are usually not
-  available for public consumption. The pipeline proposed in the paper
-  opens up the possibility for general scientists to construct indices
-  customised to their research purpose.
-\item
-  decision makers, towards understand the
+  process. The pipeline proposed in the paper opens up the possibility
+  for general scientists to construct indices customised to their
+  research purpose.
 \item
@@
   workflow and software design.
+\end{itemize}
+
+\textbf{what can be done if people adopt this pipeline/ why it is
+beneficial?}
+
+\begin{itemize}
+\tightlist
+\item
+  transparency: official published, can decompose the indices,
+\item
+  allow for calculation on uncertainty with randomization,
+\item
+  compare indices,
+\item
+  tweak/ swap building blocks,
+\item
+  create new indices based on combinations of existing steps.
 \end{itemize}
@@
 
+\begin{itemize}
+\tightlist
+\item
+  easy to operate properly: should be designed that users know what is
+  approriate to do
+\item
+  promote good scientific methodology: should protect uses from doing
+  stupid things, doing right thing
+\end{itemize}
+
+good statistical practice
+
 \hypertarget{sec-a-pipeline-for-building-statistical-indices}{%
@@
 
+\hypertarget{normalising}{%
+\subsection{Normalising}\label{normalising}}
+
+The purpose of normalising is for cross-comparison. This step can get
+criticism from analysts for \ldots{}
+
+specifically for converting from a fitted distribution to normal score
+via reverse CDF function, non-parametric formula, or empirical
+approximation, a common step in many index: SPI, SSI, Z score. The
+purpose of normalising is to convert the index into a standardised
+series after all the steps for the ease of comparison.
+
+Normalising is usually the last step
+
 \hypertarget{variable-transformation}{%
@@
 
-A conversion of non-normal variables into (approximately) normal.
+change the shape of the variable
 
@@
 Also include weighting
-
-\hypertarget{normalising}{%
-\subsection{Normalising}\label{normalising}}
-
-The purpose of normalising is for cross-comparison. This step can get
-criticism from analysts for \ldots{}
-
-specifically for converting from a fitted distribution to normal score
-via reverse CDF function, non-parametric formula, or empirical
-approximation, a common step in many index: SPI, SSI, Z score. The
-purpose of normalising is to convert the index into a standardised
-series after all the steps for the ease of comparison.
 

### 2022-10-18 paper/paper.qmd
a draft on section what's the problem with current index construction

@@
 
-*indices constructed separately, but there are lots of are same [!], but hard to see from the current construction.
+*From last week: indices constructed separately, but there are lots of are same [!], but hard to see from the current construction.
 "messy data are messy in there own way".
@@
 
-There has not yet been a standardised workflow on index construction. 
-For simple indices, users may be able to work out the calculation from the equations or workflow provided, while for more complicated construction, software is not even available for open-source consumption.
-Even when the software implementation is available, most indices are computed through a single index command, [given examples but not too domain specific]. 
-This software design focuses on the name of the index and includes all the steps in a single index function. 
-With this design, there is little room for modification. For example, fit a new distribution that is not provided in the initial code. 
-Also, when scaled to a large number of indices, it increases the cognitive load of remembering hundreds of index names. 
-A better approach is to modularise these steps and construct a pipeline for users to build indices through choosing different building blocks from each module.
-With this design, users focus on the operations involved in the index construction.
-If an operation from another index is useful, users can borrow it to modify an existing index.
+Currently, index construction lacks structure. 
+It is often up to the researchers to decide whether to provide open source code on the new indices, what would be the best user interface for the new indices, and how easily the new indices can be used by end users, along with other existing indices.
+This causes the problem that indices are often designed separately and many indices simply lump all the steps into a [single command function]. 
+While this simplifies the amount of code needed to derive index values from a set of inputs, many problems arise when the code is adopted outside the research group that wrote the index. 
+First of all, indices created by a single command provide little room for modification.
+For example, many indices involve a step to fit variables into a particular distribution.
+Index researchers may provide options for different fits, but it is generally hard to experiment with options outside the provided distributions.
+Also, many indices share similar or identical steps in part of their construction, a [single command function] eliminates the possibility to borrow codes written by others and forces researchers to rebuild the wheels when creating new indices.
+Lastly, users may wish to inspect intermediate results during the index construction to validate the output or to compare results with other indices, but [single command functions] do not support this demand unless the users can customise the source code as they need.
+
+<!-- indices are constructed separately in this design. -->
+<!-- Many steps are more or less similar when constructing indices, but this would be hard to observe from the current construction.  -->
+<!-- For example, many indices needs standardisation in its workflow, it can be used to rescale the input variables or the index series. -->
+<!-- But the current construction does not make it clear through its user interface but sweep it, along with all the other steps in the index construction, under a single function call.  -->
+
+<!-- There has not yet been a standardised workflow on index construction.  -->
+<!-- For simple indices, users may be able to work out the calculation from the equations or workflow provided, while for more complicated construction, software is not even available for open-source consumption.
+-->
+<!-- Even when the software implementation is available, most indices are computed through a single index command, [given examples but not too domain specific].  -->
+<!-- This software design focuses on the name of the index and includes all the steps in a single index function.  -->
+<!-- With this design, there is little room for modification. For example, fit a new distribution that is not provided in the initial code.  -->
+<!-- Also, when scaled to a large number of indices, it increases the cognitive load of remembering hundreds of index names.  -->
+<!-- A better approach is to modularise these steps and construct a pipeline for users to build indices through choosing different building blocks from each module. -->
+<!-- With this design, users focus on the operations involved in the index construction. -->
+<!-- If an operation from another index is useful, users can borrow it to modify an existing index. -->
+
+**what can be done if people adopt this pipeline/ why it is beneficial?**
+
+Ideally each command corresponds to a user action
+
+While jumbled indices jumble in their own ways, perspicuous indices are all alike. 
+A clear pipeline for indices construction brings transparency to the algorithm. 
+Many often, official websites like National Oceanic and Atmospheric Administration (NOAA) will regularly updates the raw data and the indices calculated for users' consumption and provides description of the methodology.
+However, the source code and the scripts is usually not publicly available and there is no tool, or users need to be its own tool, to verify the calculation. 
+Once a data pipeline is built, users can follow the steps in the index construction to validate the numbers published by the authorities. 
+Once a data pipeline is constructed, it opens possibility to do statistics on indices.
+For example, bootstrap and other methods could be applied into the pipeline to calculate uncertainty on the index value with randomisation. 
+Properties of existing indices can be examined through swapping the steps involved in the construction. 
+For example, some indices prefer to standardise the input variables in the beginning of the workflow while other prefer to move the standardisation towards the end and directly apply it on the index value.
+A flexible pipline approach gives the freedom to construct the index in both ways by placing the standardisation step in different stages of the pipeline. 
+The importance of the position of a single steps can be compared through indices constructed by placing the step at different stages of the pipeline. 
+Also, the pipeline approach motivates new indices through combining existing steps from different indices.
+
+<!-- - transparency: official published, can decompose the indices,  -->
+<!-- - allow for calculation on uncertainty with randomization,  -->
+<!-- - compare indices, -->
+<!-- - tweak/ swap building blocks,  -->
+<!-- - create new indices based on combinations of existing steps. -->
 
@@
   - statisticians, so as to [draw focus/ advocate] on statistical workflow and software design.
-
-**what can be done if people adopt this pipeline/ why it is beneficial?**
-
-- transparency: official published, can decompose the indices, 
-- allow for calculation on uncertainty with randomization, 
-- compare indices,
-- tweak/ swap building blocks, 
-- create new indices based on combinations of existing steps.
 

### 2022-10-18 paper/paper.tex
a draft on section what's the problem with current index construction

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-\emph{indices constructed separately, but there are lots of are same
-{[}!{]}, but hard to see from the current construction. ``messy data are
-messy in there own way''. Describe how they are same, so we can modular
-them}
+\emph{From last week: indices constructed separately, but there are lots
+of are same {[}!{]}, but hard to see from the current construction.
+``messy data are messy in there own way''. Describe how they are same,
+so we can modular them}
 
-There has not yet been a standardised workflow on index construction.
-For simple indices, users may be able to work out the calculation from
-the equations or workflow provided, while for more complicated
-construction, software is not even available for open-source
-consumption. Even when the software implementation is available, most
-indices are computed through a single index command, {[}given examples
-but not too domain specific{]}. This software design focuses on the name
-of the index and includes all the steps in a single index function. With
-this design, there is little room for modification. For example, fit a
-new distribution that is not provided in the initial code. Also, when
-scaled to a large number of indices, it increases the cognitive load of
-remembering hundreds of index names. A better approach is to modularise
-these steps and construct a pipeline for users to build indices through
-choosing different building blocks from each module. With this design,
-users focus on the operations involved in the index construction. If an
-operation from another index is useful, users can borrow it to modify an
-existing index.
+Currently, index construction lacks structure. It is often up to the
+researchers to decide whether to provide open source code on the new
+indices, what would be the best user interface for the new indices, and
+how easily the new indices can be used by end users, along with other
+existing indices. This causes the problem that indices are often
+designed separately and many indices simply lump all the steps into a
+{[}single command function{]}. While this simplifies the amount of code
+needed to derive index values from a set of inputs, many problems arise
+when the code is adopted outside the research group that wrote the
+index. First of all, indices created by a single command provide little
+room for modification. For example, many indices involve a step to fit
+variables into a particular distribution. Index researchers may provide
+options for different fits, but it is generally hard to experiment with
+options outside the provided distributions. Also, many indices share
+similar or identical steps in part of their construction, a {[}single
+command function{]} eliminates the possibility to borrow codes written
+by others and forces researchers to rebuild the wheels when creating new
+indices. Lastly, users may wish to inspect intermediate results during
+the index construction to validate the output or to compare results with
+other indices, but {[}single command functions{]} do not support this
+demand unless the users can customise the source code as they need.
+
+\textbf{what can be done if people adopt this pipeline/ why it is
+beneficial?}
+
+Ideally each command corresponds to a user action
+
+While jumbled indices jumble in their own ways, perspicuous indices are
+all alike. A clear pipeline for indices construction brings transparency
+to the algorithm. Many often, official websites like National Oceanic
+and Atmospheric Administration (NOAA) will regularly updates the raw
+data and the indices calculated for users' consumption and provides
+description of the methodology. However, the source code and the scripts
+is usually not publicly available and there is no tool, or users need to
+be its own tool, to verify the calculation. Once a data pipeline is
+built, users can follow the steps in the index construction to validate
+the numbers published by the authorities. Once a data pipeline is
+constructed, it opens possibility to do statistics on indices. For
+example, bootstrap and other methods could be applied into the pipeline
+to calculate uncertainty on the index value with randomisation.
+Properties of existing indices can be examined through swapping the
+steps involved in the construction. For example, some indices prefer to
+standardise the input variables in the beginning of the workflow while
+other prefer to move the standardisation towards the end and directly
+apply it on the index value. A flexible pipline approach gives the
+freedom to construct the index in both ways by placing the
+standardisation step in different stages of the pipeline. The importance
+of the position of a single steps can be compared through indices
+constructed by placing the step at different stages of the pipeline.
+Also, the pipeline approach motivates new indices through combining
+existing steps from different indices.
 
@@
   workflow and software design.
-\end{itemize}
-
-\textbf{what can be done if people adopt this pipeline/ why it is
-beneficial?}
-
-\begin{itemize}
-\tightlist
-\item
-  transparency: official published, can decompose the indices,
-\item
-  allow for calculation on uncertainty with randomization,
-\item
-  compare indices,
-\item
-  tweak/ swap building blocks,
-\item
-  create new indices based on combinations of existing steps.
 \end{itemize}

### 2022-10-18 paper/paper.qmd
more tweak

@@
 Currently, index construction lacks structure. 
-It is often up to the researchers to decide whether to provide open source code on the new indices, what would be the best user interface for the new indices, and how easily the new indices can be used by end users, along with other existing indices.
-This causes the problem that indices are often designed separately and many indices simply lump all the steps into a [single command function]. 
-While this simplifies the amount of code needed to derive index values from a set of inputs, many problems arise when the code is adopted outside the research group that wrote the index. 
-First of all, indices created by a single command provide little room for modification.
-For example, many indices involve a step to fit variables into a particular distribution.
-Index researchers may provide options for different fits, but it is generally hard to experiment with options outside the provided distributions.
-Also, many indices share similar or identical steps in part of their construction, a [single command function] eliminates the possibility to borrow codes written by others and forces researchers to rebuild the wheels when creating new indices.
-Lastly, users may wish to inspect intermediate results during the index construction to validate the output or to compare results with other indices, but [single command functions] do not support this demand unless the users can customise the source code as they need.
+It is often up to the researchers or research institutions to decide whether to provide open source code on the new indices, what would be the best user interface for the new indices, and how easily the new indices can be used by end users, along with other existing indices.
+Institutes like National Oceanic and Atmospheric Administration (NOAA) and United States Drought Monitor (USDM) moderate the indices in their domains.
+They provide public information to describe the methodology used in the calculation, publish the index values for public consumption while the source code used for computing is often not available for public consumption.
+For researchers, the level of varies. Some research groups provide open sourced code available for [explain situation with individual researchers]
+This causes the problem that computation on indices are not transparent and code design on indices are done separately.
+Researchers who wish to validate the calculation from large institutes need to reinvent the wheels that have already been built.
+Open source codes provide narrow margin for adjustment. [more here]
+<!-- For example, many indices involve a step to fit variables into a particular distribution. -->
+<!-- Index researchers may provide options for different fits, but it is generally hard to experiment with options outside the provided distributions. -->
+Similar steps used by different indices are difficult to spot since they all hide under commands for calculating different indices. 
+Lastly, these functions does not support inspecting intermediate results if users wish to check the output of a certain step.
+
+<!-- users may wish to inspect intermediate results during the index construction to validate the output or to compare results with other indices, but [single command functions] do not support this demand unless the users can customise the source code as they need.
+-->
 
@@
 
-Ideally each command corresponds to a user action
+*This week: need to extend each point of why it is beneficial further with more information?*
 
 While jumbled indices jumble in their own ways, perspicuous indices are all alike. 
-A clear pipeline for indices construction brings transparency to the algorithm. 
-Many often, official websites like National Oceanic and Atmospheric Administration (NOAA) will regularly updates the raw data and the indices calculated for users' consumption and provides description of the methodology.
-However, the source code and the scripts is usually not publicly available and there is no tool, or users need to be its own tool, to verify the calculation. 
-Once a data pipeline is built, users can follow the steps in the index construction to validate the numbers published by the authorities. 
-Once a data pipeline is constructed, it opens possibility to do statistics on indices.
-For example, bootstrap and other methods could be applied into the pipeline to calculate uncertainty on the index value with randomisation. 
-Properties of existing indices can be examined through swapping the steps involved in the construction. 
-For example, some indices prefer to standardise the input variables in the beginning of the workflow while other prefer to move the standardisation towards the end and directly apply it on the index value.
-A flexible pipline approach gives the freedom to construct the index in both ways by placing the standardisation step in different stages of the pipeline. 
-The importance of the position of a single steps can be compared through indices constructed by placing the step at different stages of the pipeline. 
-Also, the pipeline approach motivates new indices through combining existing steps from different indices.
+This paper proposes a data pipeline for index construction. [more on what we bring onto the table: construct indices through pipeline steps where each step corresponds to a single building block???]
 
-<!-- - transparency: official published, can decompose the indices,  -->
-<!-- - allow for calculation on uncertainty with randomization,  -->
-<!-- - compare indices, -->
-<!-- - tweak/ swap building blocks,  -->
-<!-- - create new indices based on combinations of existing steps. -->
+The pipeline framework provides an open source tool to construct indices in general. This is beneficial because it opens up the possibility to create, study, and use indices in a more principled way. 
+More specifically, it enables researchers to 1) validate the indices calculated from external organisations, 2) unify various indices under the same framework for comparison, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty  on indices, and finally, 5) inspire proposal of new indices from existing ones.
+
+<!-- Once a data pipeline is constructed, it opens possibility to do statistics on indices. -->
+<!-- For example, bootstrap and other methods could be applied into the pipeline to calculate uncertainty on the index value with randomisation.  -->
+<!-- Properties of existing indices can be examined through swapping the steps involved in the construction.  -->
+<!-- For example, some indices prefer to standardise the input variables in the beginning of the workflow while other prefer to move the standardisation towards the end and directly apply it on the index value.
+-->
+<!-- A flexible pipline approach gives the freedom to construct the index in both ways by placing the standardisation step in different stages of the pipeline.  -->
+<!-- The importance of the position of a single steps can be compared through indices constructed by placing the step at different stages of the pipeline.  -->
+<!-- Also, the pipeline approach motivates new indices through combining existing steps from different indices.  -->
 

### 2022-10-18 paper/paper.tex
more tweak

--- 2022-10-13 paper/paper.tex
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, interior hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\emph{indices constructed separately, but there are lots of are same
-{[}!{]}, but hard to see from the current construction. ``messy data are
-messy in there own way''. Describe how they are same, so we can modular
-them}
+\emph{From last week: indices constructed separately, but there are lots
+of are same {[}!{]}, but hard to see from the current construction.
+``messy data are messy in there own way''. Describe how they are same,
+so we can modular them}
 
-There has not yet been a standardised workflow on index construction.
-For simple indices, users may be able to work out the calculation from
-the equations or workflow provided, while for more complicated
-construction, software is not even available for open-source
-consumption. Even when the software implementation is available, most
-indices are computed through a single index command, {[}given examples
-but not too domain specific{]}. This software design focuses on the name
-of the index and includes all the steps in a single index function. With
-this design, there is little room for modification. For example, fit a
-new distribution that is not provided in the initial code. Also, when
-scaled to a large number of indices, it increases the cognitive load of
-remembering hundreds of index names. A better approach is to modularise
-these steps and construct a pipeline for users to build indices through
-choosing different building blocks from each module. With this design,
-users focus on the operations involved in the index construction. If an
-operation from another index is useful, users can borrow it to modify an
-existing index.
+Currently, index construction lacks structure. It is often up to the
+researchers or research institutions to decide whether to provide open
+source code on the new indices, what would be the best user interface
+for the new indices, and how easily the new indices can be used by end
+users, along with other existing indices. Institutes like National
+Oceanic and Atmospheric Administration (NOAA) and United States Drought
+Monitor (USDM) moderate the indices in their domains. They provide
+public information to describe the methodology used in the calculation,
+publish the index values for public consumption while the source code
+used for computing is often not available for public consumption. For
+researchers, the level of varies. Some research groups provide open
+sourced code available for {[}explain situation with individual
+researchers{]} This causes the problem that computation on indices are
+not transparent and code design on indices are done separately.
+Researchers who wish to validate the calculation from large institutes
+need to reinvent the wheels that have already been built. Open source
+codes provide narrow margin for adjustment. {[}more here{]} Similar
+steps used by different indices are difficult to spot since they all
+hide under commands for calculating different indices. Lastly, these
+functions does not support inspecting intermediate results if users wish
+to check the output of a certain step.
+
+\textbf{what can be done if people adopt this pipeline/ why it is
+beneficial?}
+
+\emph{This week: need to extend each point of why it is beneficial
+further with more information?}
+
+While jumbled indices jumble in their own ways, perspicuous indices are
+all alike. This paper proposes a data pipeline for index construction.
+{[}more on what we bring onto the table: construct indices through
+pipeline steps where each step corresponds to a single building
+block???{]}
+
+The pipeline framework provides an open source tool to construct indices
+in general. This is beneficial because it opens up the possibility to
+create, study, and use indices in a more principled way. More
+specifically, it enables researchers to 1) validate the indices
+calculated from external organisations, 2) unify various indices under
+the same framework for comparison, 3) swap or adjust individual steps in
+the index construction to study their contribution, 4) calculate
+uncertainty on indices, and finally, 5) inspire proposal of new indices
+from existing ones.
 

### 2022-10-19 paper/paper.qmd
a draft on the next three paragrpahs in introduction

--- 2022-10-13 paper/paper.qmd
@@
 
-*indices constructed separately, but there are lots of are same [!], but hard to see from the current construction.
-"messy data are messy in there own way".
-Describe how they are same, so we can modular them*
+Currently, index construction lacks standardised workflow. 
+It is often up to researchers or research institutions to decide whether to provide open source code on the new indices, what would be the best user interface for other researchers to use the new indices, and how easily the new indices can be compared with other existing indices.
+This makes the computation lack transparency and indices cumbersome to experiment with.
+Researchers who wish to validate the indices calculated from large institutes need to reinvent the wheels themselves since the source code used for computing is often not available for public consumption.
+Open source codes provided by research groups has a narrow margin for exploring other options outside the provided.
+Similar steps used by different indices are difficult to spot since the design of user interface for indices often include all the steps under a single function call.
+And lastly, it is generally hard to inspect intermediate results during the index construction if users wish to check the output of a certain step.
 
-There has not yet been a standardised workflow on index construction. 
-For simple indices, users may be able to work out the calculation from the equations or workflow provided, while for more complicated construction, software is not even available for open-source consumption.
-Even when the software implementation is available, most indices are computed through a single index command, [given examples but not too domain specific]. 
-This software design focuses on the name of the index and includes all the steps in a single index function. 
-With this design, there is little room for modification. For example, fit a new distribution that is not provided in the initial code. 
-Also, when scaled to a large number of indices, it increases the cognitive load of remembering hundreds of index names. 
-A better approach is to modularise these steps and construct a pipeline for users to build indices through choosing different building blocks from each module.
-With this design, users focus on the operations involved in the index construction.
-If an operation from another index is useful, users can borrow it to modify an existing index.
+<!-- Institutes like National Oceanic and Atmospheric Administration (NOAA) and United States Drought Monitor (USDM) moderate indices in their domains. -->
+<!-- They provide public information to describe the methodology used in the calculation, publish the index values for public consumption while the source code used for computing is often not available for public consumption.
+-->
+
+<!-- For example, many indices involve a step to fit variables into a particular distribution. -->
+<!-- Index researchers may provide options for different fits, but it is generally hard to experiment with options outside the provided distributions. -->
+
+<!-- For example, many indices needs standardisation in its workflow, it can be used to rescale the input variables or the index series. -->
+<!-- But the current construction does not make it clear through its user interface but sweep it, along with all the other steps in the index construction, under a single function call.  -->
+
+**what can be done if people adopt this pipeline/ why it is beneficial?**
+
+*This week: need to extend each point of why it is beneficial further with more information?*
+
+While every jumbled indices jumble in their own ways, perspicuous indices are all alike. 
+This paper proposes a data pipeline for index construction. 
+By recognising the common steps shared by many indices, we develop a pipeline that breaks down index construction into multiple modules and allow operations in various modules to be combined like building blocks to construct indices.
+The pipeline approach is general in nature while adaptable to most index construction. 
+It allows indices to be created, studied and compared in a structured tidy form and enables analysis on indices to be performed easily: 
+More specifically, it enables researchers to 1) validate the indices calculated from external organisations, 2) unify various indices under the same framework for comparison, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty  on indices through bootstrap or others, and finally, 5) inspire proposal of new indices from combining different steps in existing indices.
 
@@
 
-*this work could be interested to scientists actively developing new indices*
+This work could be interested to researchers actively developing new indices to delivery new indices in an easy-to-reproduce design. 
+It would also provide analysts who wish to compute a range of indices in their analysis a uniform interface to build relevant indices from raw data. 
+For statistician and software developing engineers, this work frames the workflow of index construction and could motivates similar work in other workflow used in practice.
 
-  - domain scientists who propose new indices and directly use indices in analysis. [use multiple indices to compare]
-  - general scientists who uses indices as part of analysis, modelling process.
-The pipeline proposed in the paper opens up the possibility for general scientists to construct indices customised to their research purpose.
-  - statisticians, so as to [draw focus/ advocate] on statistical workflow and software design.
-
-**what can be done if people adopt this pipeline/ why it is beneficial?**
-
-- transparency: official published, can decompose the indices, 
-- allow for calculation on uncertainty with randomization, 
-- compare indices,
-- tweak/ swap building blocks, 
-- create new indices based on combinations of existing steps.
-
-The rest of the paper is structured as follows: the concept of data pipeline in R is reviewed in @sec-data-pipeline-in-r.
-@sec-a-pipeline-for-building-statistical-indices presents the data pipeline for index construction.
+The rest of the paper is structured as follows: @sec-data-pipeline-in-r reviews the concept of data pipeline in R. 
+The pipeline framework for index construction is presented in @sec-a-pipeline-for-building-statistical-indices.
 @sec-incorporating-new-buliding-blocks-into-the-pipeline explains how to include a new building block in each pipeline module.

### 2022-10-19 paper/paper.tex
a draft on the next three paragrpahs in introduction

--- 2022-10-13 paper/paper.tex
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, frame hidden, interior hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-\emph{indices constructed separately, but there are lots of are same
-{[}!{]}, but hard to see from the current construction. ``messy data are
-messy in there own way''. Describe how they are same, so we can modular
-them}
-
-There has not yet been a standardised workflow on index construction.
-For simple indices, users may be able to work out the calculation from
-the equations or workflow provided, while for more complicated
-construction, software is not even available for open-source
-consumption. Even when the software implementation is available, most
-indices are computed through a single index command, {[}given examples
-but not too domain specific{]}. This software design focuses on the name
-of the index and includes all the steps in a single index function. With
-this design, there is little room for modification. For example, fit a
-new distribution that is not provided in the initial code. Also, when
-scaled to a large number of indices, it increases the cognitive load of
-remembering hundreds of index names. A better approach is to modularise
-these steps and construct a pipeline for users to build indices through
-choosing different building blocks from each module. With this design,
-users focus on the operations involved in the index construction. If an
-operation from another index is useful, users can borrow it to modify an
-existing index.
-
-\textbf{who would benefit from this paper}
-
-\emph{this work could be interested to scientists actively developing
-new indices}
-
-\begin{itemize}
-\tightlist
-\item
-  domain scientists who propose new indices and directly use indices in
-  analysis. {[}use multiple indices to compare{]}
-\item
-  general scientists who uses indices as part of analysis, modelling
-  process. The pipeline proposed in the paper opens up the possibility
-  for general scientists to construct indices customised to their
-  research purpose.
-\item
-  statisticians, so as to {[}draw focus/ advocate{]} on statistical
-  workflow and software design.
-\end{itemize}
+Currently, index construction lacks standardised workflow. It is often
+up to researchers or research institutions to decide whether to provide
+open source code on the new indices, what would be the best user
+interface for other researchers to use the new indices, and how easily
+the new indices can be compared with other existing indices. This makes
+the computation lack transparency and indices cumbersome to experiment
+with. Researchers who wish to validate the indices calculated from large
+institutes need to reinvent the wheels themselves since the source code
+used for computing is often not available for public consumption. Open
+source codes provided by research groups has a narrow margin for
+exploring other options outside the provided. Similar steps used by
+different indices are difficult to spot since the design of user
+interface for indices often include all the steps under a single
+function call. And lastly, it is generally hard to inspect intermediate
+results during the index construction if users wish to check the output
+of a certain step.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  transparency: official published, can decompose the indices,
-\item
-  allow for calculation on uncertainty with randomization,
-\item
-  compare indices,
-\item
-  tweak/ swap building blocks,
-\item
-  create new indices based on combinations of existing steps.
-\end{itemize}
+\emph{This week: need to extend each point of why it is beneficial
+further with more information?}
 
-The rest of the paper is structured as follows: the concept of data
-pipeline in R is reviewed in Section~\ref{sec-data-pipeline-in-r}.
-Section~\ref{sec-a-pipeline-for-building-statistical-indices} presents
-the data pipeline for index construction.
+While every jumbled indices jumble in their own ways, perspicuous
+indices are all alike. This paper proposes a data pipeline for index
+construction. By recognising the common steps shared by many indices, we
+develop a pipeline that breaks down index construction into multiple
+modules and allow operations in various modules to be combined like
+building blocks to construct indices. The pipeline approach is general
+in nature while adaptable to most index construction. It allows indices
+to be created, studied and compared in a structured tidy form and
+enables analysis on indices to be performed easily: More specifically,
+it enables researchers to 1) validate the indices calculated from
+external organisations, 2) unify various indices under the same
+framework for comparison, 3) swap or adjust individual steps in the
+index construction to study their contribution, 4) calculate uncertainty
+on indices through bootstrap or others, and finally, 5) inspire proposal
+of new indices from combining different steps in existing indices.
+
+\textbf{who would benefit from this paper}
+
+This work could be interested to researchers actively developing new
+indices to delivery new indices in an easy-to-reproduce design. It would
+also provide analysts who wish to compute a range of indices in their
+analysis a uniform interface to build relevant indices from raw data.
+For statistician and software developing engineers, this work frames the
+workflow of index construction and could motivates similar work in other
+workflow used in practice.
+
+The rest of the paper is structured as follows:
+Section~\ref{sec-data-pipeline-in-r} reviews the concept of data
+pipeline in R. The pipeline framework for index construction is
+presented in
+Section~\ref{sec-a-pipeline-for-building-statistical-indices}.
 Section~\ref{sec-incorporating-new-buliding-blocks-into-the-pipeline}

### 2022-10-19 paper/paper.qmd
spelling and grammar

@@
 
-Currently, index construction lacks standardised workflow. 
+Currently, index construction lacks a standardised workflow. 
 It is often up to researchers or research institutions to decide whether to provide open source code on the new indices, what would be the best user interface for other researchers to use the new indices, and how easily the new indices can be compared with other existing indices.
@@
 Researchers who wish to validate the indices calculated from large institutes need to reinvent the wheels themselves since the source code used for computing is often not available for public consumption.
-Open source codes provided by research groups has a narrow margin for exploring other options outside the provided.
-Similar steps used by different indices are difficult to spot since the design of user interface for indices often include all the steps under a single function call.
+Open-source code provided by research groups has a narrow margin for exploring other options outside the provided.
+Similar steps used by different indices are difficult to spot since the design of the user interface for indices often includes all the steps under a single function call.
 And lastly, it is generally hard to inspect intermediate results during the index construction if users wish to check the output of a certain step.
@@
 
-While every jumbled indices jumble in their own ways, perspicuous indices are all alike. 
+While every jumbled index jumbles in its own way, perspicuous indices are all alike. 
 This paper proposes a data pipeline for index construction. 
 By recognising the common steps shared by many indices, we develop a pipeline that breaks down index construction into multiple modules and allow operations in various modules to be combined like building blocks to construct indices.
-The pipeline approach is general in nature while adaptable to most index construction. 
-It allows indices to be created, studied and compared in a structured tidy form and enables analysis on indices to be performed easily: 
+The pipeline approach is general while adaptable to most index construction. 
+It allows indices to be created, studied, and compared in a structured tidy form and enables statistical analysis of indices to be performed easily: 
 More specifically, it enables researchers to 1) validate the indices calculated from external organisations, 2) unify various indices under the same framework for comparison, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty on indices through bootstrap or others, and finally, 5) inspire proposal of new indices from combining different steps in existing indices.
@@
 
-This work could be interested to researchers actively developing new indices to delivery new indices in an easy-to-reproduce design. 
+This work could interests researchers actively developing new indices to deliver new indices in an easy-to-reproduce design. 
 It would also provide analysts who wish to compute a range of indices in their analysis a uniform interface to build relevant indices from raw data. 
-For statistician and software developing engineers, this work frames the workflow of index construction and could motivates similar work in other workflow used in practice.
+For statisticians and software developing engineers, this work frames the workflow of index construction and could motivate similar work in other workflow used in practice.
 

### 2022-10-19 paper/paper.tex
spelling and grammar

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, frame hidden, interior hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-Currently, index construction lacks standardised workflow. It is often
+Currently, index construction lacks a standardised workflow. It is often
 up to researchers or research institutions to decide whether to provide
@@
 institutes need to reinvent the wheels themselves since the source code
-used for computing is often not available for public consumption. Open
-source codes provided by research groups has a narrow margin for
+used for computing is often not available for public consumption.
+Open-source code provided by research groups has a narrow margin for
 exploring other options outside the provided. Similar steps used by
-different indices are difficult to spot since the design of user
-interface for indices often include all the steps under a single
+different indices are difficult to spot since the design of the user
+interface for indices often includes all the steps under a single
 function call. And lastly, it is generally hard to inspect intermediate
@@
 
-While every jumbled indices jumble in their own ways, perspicuous
-indices are all alike. This paper proposes a data pipeline for index
+While every jumbled index jumbles in its own way, perspicuous indices
+are all alike. This paper proposes a data pipeline for index
 construction. By recognising the common steps shared by many indices, we
@@
 building blocks to construct indices. The pipeline approach is general
-in nature while adaptable to most index construction. It allows indices
-to be created, studied and compared in a structured tidy form and
-enables analysis on indices to be performed easily: More specifically,
-it enables researchers to 1) validate the indices calculated from
-external organisations, 2) unify various indices under the same
-framework for comparison, 3) swap or adjust individual steps in the
-index construction to study their contribution, 4) calculate uncertainty
-on indices through bootstrap or others, and finally, 5) inspire proposal
-of new indices from combining different steps in existing indices.
+while adaptable to most index construction. It allows indices to be
+created, studied, and compared in a structured tidy form and enables
+statistical analysis of indices to be performed easily: More
+specifically, it enables researchers to 1) validate the indices
+calculated from external organisations, 2) unify various indices under
+the same framework for comparison, 3) swap or adjust individual steps in
+the index construction to study their contribution, 4) calculate
+uncertainty on indices through bootstrap or others, and finally, 5)
+inspire proposal of new indices from combining different steps in
+existing indices.
 
@@
 
-This work could be interested to researchers actively developing new
-indices to delivery new indices in an easy-to-reproduce design. It would
-also provide analysts who wish to compute a range of indices in their
+This work could interests researchers actively developing new indices to
+deliver new indices in an easy-to-reproduce design. It would also
+provide analysts who wish to compute a range of indices in their
 analysis a uniform interface to build relevant indices from raw data.
-For statistician and software developing engineers, this work frames the
-workflow of index construction and could motivates similar work in other
-workflow used in practice.
+For statisticians and software developing engineers, this work frames
+the workflow of index construction and could motivate similar work in
+other workflow used in practice.
 

### 2022-10-19 paper/paper.qmd
tweaks and todo after today's meeting

@@
 
-**What is the problem with current index construction**
+**What is the challenges with current index construction**
 
@@
 It is often up to researchers or research institutions to decide whether to provide open source code on the new indices, what would be the best user interface for other researchers to use the new indices, and how easily the new indices can be compared with other existing indices.
-This makes the computation lack transparency and indices cumbersome to experiment with.
-Researchers who wish to validate the indices calculated from large institutes need to reinvent the wheels themselves since the source code used for computing is often not available for public consumption.
-Open-source code provided by research groups has a narrow margin for exploring other options outside the provided.
-Similar steps used by different indices are difficult to spot since the design of the user interface for indices often includes all the steps under a single function call.
-And lastly, it is generally hard to inspect intermediate results during the index construction if users wish to check the output of a certain step.
+This makes the computation lack transparency and indices cumbersome to experiment with:
+
+  - Researchers who wish to validate the indices calculated from large institutes need to reinvent the wheels themselves since the source code used for computing is often not available for public consumption;
+  - Open-source code provided by research groups has a narrow margin for exploring other options outside the provided;
+  - Similar steps used by different indices are difficult to spot since the design of the user interface for indices often includes all the steps under a single function call; and
+  - It is generally hard to inspect intermediate results during the index construction if users wish to check the output of a certain step.
+
+TODO: see if there is any paper describing this type of pains
 
@@
 -->
+
+TODO: useful to reference tidy data and tidy model that makes the workflow on modelling tidy somewhere in introduction
 
@@
 
-*This week: need to extend each point of why it is beneficial further with more information?*
-
-While every jumbled index jumbles in its own way, perspicuous indices are all alike. 
+~~While every jumbled index jumbles in its own way, [perspicuous] clearly-laied out indices are all alike.~~
 This paper proposes a data pipeline for index construction. 
@@
 It allows indices to be created, studied, and compared in a structured tidy form and enables statistical analysis of indices to be performed easily: 
-More specifically, it enables researchers to 1) validate the indices calculated from external organisations, 2) unify various indices under the same framework for comparison, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty on indices through bootstrap or others, and finally, 5) inspire proposal of new indices from combining different steps in existing indices.
+More specifically, it enables researchers to 1) validate the indices calculated from external organisations, 2) unify various indices under the same framework for computing, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty on indices through bootstrap or others, 5) enhance existing indices through comparing and studying their statistical properties, and finally, 6) propose new indices from combining different steps in existing indices.
+
+TODO: first sentence: tailor to their indices without thinking about the big picture
 
@@
 
-This work could interests researchers actively developing new indices to deliver new indices in an easy-to-reproduce design. 
+This work is of interest to researchers actively developing new indices since it encourages new indices to be delivered in an easy-to-reproduce design. 
 It would also provide analysts who wish to compute a range of indices in their analysis a uniform interface to build relevant indices from raw data. 
-For statisticians and software developing engineers, this work frames the workflow of index construction and could motivate similar work in other workflow used in practice.
+For statisticians and software developing engineers, this work frames the process of index construction in a more user-oriented workflow and could motivate similar research for other process in scientific computing.
 

### 2022-10-19 paper/paper.tex
tweaks and todo after today's meeting

@@
 \usepackage{xcolor}
+\usepackage[normalem]{ulem}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, frame hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-\textbf{What is the problem with current index construction}
+\textbf{What is the challenges with current index construction}
 
@@
 the computation lack transparency and indices cumbersome to experiment
-with. Researchers who wish to validate the indices calculated from large
-institutes need to reinvent the wheels themselves since the source code
-used for computing is often not available for public consumption.
+with:
+
+\begin{itemize}
+\tightlist
+\item
+  Researchers who wish to validate the indices calculated from large
+  institutes need to reinvent the wheels themselves since the source
+  code used for computing is often not available for public consumption;
+\item
   Open-source code provided by research groups has a narrow margin for
-exploring other options outside the provided. Similar steps used by
-different indices are difficult to spot since the design of the user
-interface for indices often includes all the steps under a single
-function call. And lastly, it is generally hard to inspect intermediate
-results during the index construction if users wish to check the output
-of a certain step.
+  exploring other options outside the provided;
+\item
+  Similar steps used by different indices are difficult to spot since
+  the design of the user interface for indices often includes all the
+  steps under a single function call; and
+\item
+  It is generally hard to inspect intermediate results during the index
+  construction if users wish to check the output of a certain step.
+\end{itemize}
+
+TODO: see if there is any paper describing this type of pains
+
+TODO: useful to reference tidy data and tidy model that makes the
+workflow on modelling tidy somewhere in introduction
 
@@
 
-\emph{This week: need to extend each point of why it is beneficial
-further with more information?}
+\sout{While every jumbled index jumbles in its own way,
+{[}perspicuous{]} clearly-laied out indices are all alike.} This paper
+proposes a data pipeline for index construction. By recognising the
+common steps shared by many indices, we develop a pipeline that breaks
+down index construction into multiple modules and allow operations in
+various modules to be combined like building blocks to construct
+indices. The pipeline approach is general while adaptable to most index
+construction. It allows indices to be created, studied, and compared in
+a structured tidy form and enables statistical analysis of indices to be
+performed easily: More specifically, it enables researchers to 1)
+validate the indices calculated from external organisations, 2) unify
+various indices under the same framework for computing, 3) swap or
+adjust individual steps in the index construction to study their
+contribution, 4) calculate uncertainty on indices through bootstrap or
+others, 5) enhance existing indices through comparing and studying their
+statistical properties, and finally, 6) propose new indices from
+combining different steps in existing indices.
 
-While every jumbled index jumbles in its own way, perspicuous indices
-are all alike. This paper proposes a data pipeline for index
-construction. By recognising the common steps shared by many indices, we
-develop a pipeline that breaks down index construction into multiple
-modules and allow operations in various modules to be combined like
-building blocks to construct indices. The pipeline approach is general
-while adaptable to most index construction. It allows indices to be
-created, studied, and compared in a structured tidy form and enables
-statistical analysis of indices to be performed easily: More
-specifically, it enables researchers to 1) validate the indices
-calculated from external organisations, 2) unify various indices under
-the same framework for comparison, 3) swap or adjust individual steps in
-the index construction to study their contribution, 4) calculate
-uncertainty on indices through bootstrap or others, and finally, 5)
-inspire proposal of new indices from combining different steps in
-existing indices.
+TODO: first sentence: tailor to their indices without thinking about the
+big picture
 
@@
 
-This work could interests researchers actively developing new indices to
-deliver new indices in an easy-to-reproduce design. It would also
-provide analysts who wish to compute a range of indices in their
-analysis a uniform interface to build relevant indices from raw data.
-For statisticians and software developing engineers, this work frames
-the workflow of index construction and could motivate similar work in
-other workflow used in practice.
+This work is of interest to researchers actively developing new indices
+since it encourages new indices to be delivered in an easy-to-reproduce
+design. It would also provide analysts who wish to compute a range of
+indices in their analysis a uniform interface to build relevant indices
+from raw data. For statisticians and software developing engineers, this
+work frames the process of index construction in a more user-oriented
+workflow and could motivate similar research for other process in
+scientific computing.
 

### 2022-10-24 paper/paper.qmd
a draft on tidy data- need more work

@@
 
-\newpage
+**What is the challenges with current index construction**
 
-**What is the challenges with current index construction**
+*see if there is any paper describing this type of pains*
+
+*useful to reference tidy data and tidy model that makes the workflow on modelling tidy somewhere in introduction*
 
@@
 
-TODO: see if there is any paper describing this type of pains
-
 <!-- Institutes like National Oceanic and Atmospheric Administration (NOAA) and United States Drought Monitor (USDM) moderate indices in their domains. -->
@@
 -->
-
-TODO: useful to reference tidy data and tidy model that makes the workflow on modelling tidy somewhere in introduction
 
@@
 
+*first sentence: tailor to their indices without thinking about the big picture*
+
 ~~While every jumbled index jumbles in its own way, [perspicuous] clearly-laied out indices are all alike.~~
@@
 More specifically, it enables researchers to 1) validate the indices calculated from external organisations, 2) unify various indices under the same framework for computing, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty on indices through bootstrap or others, 5) enhance existing indices through comparing and studying their statistical properties, and finally, 6) propose new indices from combining different steps in existing indices.
-
-TODO: first sentence: tailor to their indices without thinking about the big picture
 
@@
 
+\newpage
+
 # Data pipeline in R {#sec-data-pipeline-in-r}
@@
 
-Before the concept of tidy data [@wickham_tidy_2014], tabular data arrive at data analysts in all different ways. 
-Different analysts would write customised scripts for analysing the specific data. 
-These scripts can be extended to other data analysed by the same people or group but this is not generalizable directly to another dataset. 
-When the tidy data concept comes, variables and values are arranged so that 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
-With this specific layout, wrangling on tabular data can be standardised into a grammar of data manipulation in `dplyr` [@dplyr].
+As a programming language for data analysis, R provides a scripted environment for users to experiment with data without the complex abstraction a typical programming language requires. 
+The flexible and intuitive workflow is paramount for data analysis since in a data analysis project, the full workflow of how data should be processed from the raw data to the final results is usually only consolidated till the end of the project and analysts need the freedom to quickly experiment with different ideas when working with data.
+This approach to data analysis is enhanced by the tidyverse suite where a collection of R packages are built for data science under the concept of tidy data. 
+The tidy data principle [@wickham_tidy_2014] prescribes three simple rules for structuring tabular data: 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
+The packages built in the tidyverse suite provide operations that are direct actions on the data and users, even new users, can easily read from the code on what has been performed and what to expect after each operation.
+Researchers and analysts have a similar need for an intuitive and flexible process to construct indices since a similar workflow occurs where researchers experiment with many different indices, discover issues with the current ones, and experiment with potential ideas before finally arrive at a solution.
 
-A similar issue happens with index construction where researchers construct their own indices in their own ways and there has not yet been a tidy principle on index construction.
-Also, this tidy principle on index construction is more complex than those in tidy data and the `dplyr` package.
-It has to encompass the workflow of transformation from the raw data towards the final index series.
+<!-- Researchers working on index construction should consider adopt a workflow where operations on data can be clearly reflected by the code while flexible to experiment  -->
+
+<!-- Before the concept of tidy data [@wickham_tidy_2014], tabular data arrive at data analysts in all different ways.  -->
+<!-- Different analysts would write customised scripts for analysing the specific data.  -->
+<!-- These scripts can be extended to other data analysed by the same people or group but this is not generalizable directly to another dataset.  -->
+<!-- When the tidy data concept comes, variables and values are arranged so that 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
+-->
+<!-- With this specific layout, wrangling on tabular data can be standardised into a grammar of data manipulation in `dplyr` [@dplyr].  -->
+
+<!-- A similar issue happens with index construction where researchers construct their own indices in their own ways and there has not yet been a tidy principle on index construction. -->
+<!-- Also, this tidy principle on index construction is more complex than those in tidy data and the `dplyr` package. -->
+<!-- It has to encompass the workflow of transformation from the raw data towards the final index series. -->
 

### 2022-10-24 paper/paper.tex
a draft on tidy data- need more work

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, frame hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-\newpage
+\textbf{What is the challenges with current index construction}
 
-\textbf{What is the challenges with current index construction}
+\emph{see if there is any paper describing this type of pains}
+
+\emph{useful to reference tidy data and tidy model that makes the
+workflow on modelling tidy somewhere in introduction}
 
@@
 
-TODO: see if there is any paper describing this type of pains
-
-TODO: useful to reference tidy data and tidy model that makes the
-workflow on modelling tidy somewhere in introduction
-
 \textbf{what can be done if people adopt this pipeline/ why it is
 beneficial?}
+
+\emph{first sentence: tailor to their indices without thinking about the
+big picture}
 
@@
 combining different steps in existing indices.
-
-TODO: first sentence: tailor to their indices without thinking about the
-big picture
 
@@
 
+\newpage
+
 \hypertarget{sec-data-pipeline-in-r}{%
@@
 
-Before the concept of tidy data (Wickham 2014), tabular data arrive at
-data analysts in all different ways. Different analysts would write
-customised scripts for analysing the specific data. These scripts can be
-extended to other data analysed by the same people or group but this is
-not generalizable directly to another dataset. When the tidy data
-concept comes, variables and values are arranged so that 1) Each
-variable forms a column, 2) Each observation forms a row, and; 3) Each
-type of observational unit forms a table. With this specific layout,
-wrangling on tabular data can be standardised into a grammar of data
-manipulation in \texttt{dplyr} (Wickham et al. 2022).
-
-A similar issue happens with index construction where researchers
-construct their own indices in their own ways and there has not yet been
-a tidy principle on index construction. Also, this tidy principle on
-index construction is more complex than those in tidy data and the
-\texttt{dplyr} package. It has to encompass the workflow of
-transformation from the raw data towards the final index series.
+As a programming language for data analysis, R provides a scripted
+environment for users to experiment with data without the complex
+abstraction a typical programming language requires. The flexible and
+intuitive workflow is paramount for data analysis since in a data
+analysis project, the full workflow of how data should be processed from
+the raw data to the final results is usually only consolidated till the
+end of the project and analysts need the freedom to quickly experiment
+with different ideas when working with data. This approach to data
+analysis is enhanced by the tidyverse suite where a collection of R
+packages are built for data science under the concept of tidy data. The
+tidy data principle (Wickham 2014) prescribes three simple rules for
+structuring tabular data: 1) Each variable forms a column, 2) Each
+observation forms a row, and; 3) Each type of observational unit forms a
+table. The packages built in the tidyverse suite provide operations that
+are direct actions on the data and users, even new users, can easily
+read from the code on what has been performed and what to expect after
+each operation. Researchers and analysts have a similar need for an
+intuitive and flexible process to construct indices since a similar
+workflow occurs where researchers experiment with many different
+indices, discover issues with the current ones, and experiment with
+potential ideas before finally arrive at a solution.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2022.
-\emph{Dplyr: A Grammar of Data Manipulation}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%

### 2022-10-24 paper/paper.qmd
more on the tidy data paragraph

@@
 
-As a programming language for data analysis, R provides a scripted environment for users to experiment with data without the complex abstraction a typical programming language requires. 
-The flexible and intuitive workflow is paramount for data analysis since in a data analysis project, the full workflow of how data should be processed from the raw data to the final results is usually only consolidated till the end of the project and analysts need the freedom to quickly experiment with different ideas when working with data.
-This approach to data analysis is enhanced by the tidyverse suite where a collection of R packages are built for data science under the concept of tidy data. 
-The tidy data principle [@wickham_tidy_2014] prescribes three simple rules for structuring tabular data: 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
-The packages built in the tidyverse suite provide operations that are direct actions on the data and users, even new users, can easily read from the code on what has been performed and what to expect after each operation.
-Researchers and analysts have a similar need for an intuitive and flexible process to construct indices since a similar workflow occurs where researchers experiment with many different indices, discover issues with the current ones, and experiment with potential ideas before finally arrive at a solution.
+As a scripted language for data analysis, R provides a interactive environment for users to code with data without the abstraction a typical programming language requires. 
+This flexible environment is paramount for data analysis since in practice, data analysis involves trials and errors on different steps to solve the problems.
+Analysts need the freedom to quickly experiment with different ideas when working with data.
+The tidyverse suite provides a collection of R package for conducting data analysis.
+These tools directly operate on data, and users, even new users, can easily read from the code on what operation is performed on the data and what to expect as the output.
+All of these are built upon a simple prior knowledge of the tidy data principle [@wickham_tidy_2014], which prescribes three rules for structuring tabular data: 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
 
-<!-- Researchers working on index construction should consider adopt a workflow where operations on data can be clearly reflected by the code while flexible to experiment  -->
+Constructing indices also involves this working pattern of trails and errors on different options before finding the final solution.
+A tidy workflow will give researchers more flexibility to tune the steps as they wish and while maintain a readable and intuitive code base.
+
+<!-- While data can be arranged in various different ways, maintaining a fixed representation gives a predictable output of operations on data.  -->
+
+<!-- Only when each step in the workflow has been mapped out, do data analysts need to consider those abstract concept of object oriented programming and functional programming to help them design the workflow.
+-->
+<!-- how raw data should be processed to the final results can only be consolidated when  -->
 

### 2022-10-24 paper/paper.tex
more on the tidy data paragraph

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-As a programming language for data analysis, R provides a scripted
-environment for users to experiment with data without the complex
-abstraction a typical programming language requires. The flexible and
-intuitive workflow is paramount for data analysis since in a data
-analysis project, the full workflow of how data should be processed from
-the raw data to the final results is usually only consolidated till the
-end of the project and analysts need the freedom to quickly experiment
-with different ideas when working with data. This approach to data
-analysis is enhanced by the tidyverse suite where a collection of R
-packages are built for data science under the concept of tidy data. The
-tidy data principle (Wickham 2014) prescribes three simple rules for
-structuring tabular data: 1) Each variable forms a column, 2) Each
-observation forms a row, and; 3) Each type of observational unit forms a
-table. The packages built in the tidyverse suite provide operations that
-are direct actions on the data and users, even new users, can easily
-read from the code on what has been performed and what to expect after
-each operation. Researchers and analysts have a similar need for an
-intuitive and flexible process to construct indices since a similar
-workflow occurs where researchers experiment with many different
-indices, discover issues with the current ones, and experiment with
-potential ideas before finally arrive at a solution.
+As a scripted language for data analysis, R provides a interactive
+environment for users to code with data without the abstraction a
+typical programming language requires. This flexible environment is
+paramount for data analysis since in practice, data analysis involves
+trials and errors on different steps to solve the problems. Analysts
+need the freedom to quickly experiment with different ideas when working
+with data. The tidyverse suite provides a collection of R package for
+conducting data analysis. These tools directly operate on data, and
+users, even new users, can easily read from the code on what operation
+is performed on the data and what to expect as the output. All of these
+are built upon a simple prior knowledge of the tidy data principle
+(Wickham 2014), which prescribes three rules for structuring tabular
+data: 1) Each variable forms a column, 2) Each observation forms a row,
+and; 3) Each type of observational unit forms a table.
+
+Constructing indices also involves this working pattern of trails and
+errors on different options before finding the final solution. A tidy
+workflow will give researchers more flexibility to tune the steps as
+they wish and while maintain a readable and intuitive code base.
 

### 2022-10-25 paper/paper.qmd
a draft on data pipeline

@@
 As a scripted language for data analysis, R provides a interactive environment for users to code with data without the abstraction a typical programming language requires. 
-This flexible environment is paramount for data analysis since in practice, data analysis involves trials and errors on different steps to solve the problems.
+This flexible environment is paramount for data analysis since in practice, data analysis is an iterative process that involves trials and errors to solve the problems.
 Analysts need the freedom to quickly experiment with different ideas when working with data.
@@
 These tools directly operate on data, and users, even new users, can easily read from the code on what operation is performed on the data and what to expect as the output.
-All of these are built upon a simple prior knowledge of the tidy data principle [@wickham_tidy_2014], which prescribes three rules for structuring tabular data: 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
+All of these are built upon a simple prior knowledge of the tidy data principle [@wickham_tidy_2014], which prescribes three rules to structure tabular data: 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
 
-Constructing indices also involves this working pattern of trails and errors on different options before finding the final solution.
+Constructing indices also involves this working pattern of trials and errors on different options before finding the final solution.
 A tidy workflow will give researchers more flexibility to tune the steps as they wish and while maintain a readable and intuitive code base.
@@
 
-Constructing a pipeline that divides a complex procedure into steps that can be concatenated has been adopted widely in the R community.
+The concept of pipeline (and pipe) is not new concept to computing. It refers to a set of data processing elements connected in series, where the output of one element is the input of the next one.
+@wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program.
+The paper also argues that breaking down graphic rendering into steps is beneficial for understand the implementation and compare between different graphic systems. 
+Both arguments are applicable to index construction.
+Papers that proposing new indices always present a workflow diagram to illustrate how data is transformed into the final indices, but not much implementation has explicitly present this this step-by-step transformation in the code design.
 
-The data pipeline in interactive graphics is a set of steps that transform the raw data to the plots displayed on the screen. 
-The initial pipeline proposed by @buja_elements_1988 involves the following steps: non-linear transformation, variables standardization, randomization, projection engine, and viewporting.
-The initial pipeline proposed by @buja_elements_1988 involves the following steps: non-linear transformation, variables standardization, randomization, projection engine, and viewporting. 
-Another example in the early work of pipeline by @sutherland_orca_2000 describes a three-step pipeline: variable standardization, dimension reduction, and scaling data into the viewing window. 
-This pipeline also includes the transformation on spatial and temporal variables, i.e. computing time lag on temporal variables.
-This pipeline also includes the transformation on spatial and temporal variables, i.e. computing time lag on temporal variables. 
-@wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program and breaking down graphic rendering into steps is also beneficial for understand the implementation and compare between different graphic systems.
-
-The data pipeline concept is further enhanced by the pipe operator (`%>%`) in R where a set of operations, or steps, can be chained together to form a set of instructions.
-
-A more recent data pipeline is tidymodels [@tidymodels], a set of packages for machine learning models following the tidyverse principles [@tidyverse]. 
-Steps: Exploratory data analysis (EDA), feature engineering, model tuning and selection, and model evaluation.
-
-- easy to operate properly: should be designed that users know what is approriate to do
-- promote good scientific methodology: should protect uses from doing stupid things,  doing right thing
-
-good statistical practice
+To better understand the pipeline approach in statistic computing, examples in interactive graphics and machine learning models are given.
+In interactive graphics, the data pipeline is used to describe how the data is processed to be rendered on the screen. 
+Examples of early work on building pipeline for interactive graphics includes @buja_elements_1988, @sutherland_orca_2000, and @xie_reactive_2014. 
+Steps in these pipelines include non-linear transformation, variable standardization, randomization and dimension reduction.
+A more familiar example from tidymodels [@tidymodels] presents a set of packages for machine learning models following the tidyverse principles [@tidyverse]. 
+The machine learning pipeline constitutes data resampling, feature engineering, model fitting, model tuning, and model evaluation each being responsible by a separate package. 
+Before tidymodel, analysts who wish to fit different models need to learn the workflow of each individual algorithm and need to rewrite most of the model fitting, tuning, and evaluation code when switching from one model to another.
+The tidymodel approach unifies a wide range of models under a single pipeline where users can substitute different models while maintain the same pipeline.
+This allows users to easily experiment with different machine learning models a lot easier.
 

### 2022-10-25 paper/paper.tex
a draft on data pipeline

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 typical programming language requires. This flexible environment is
-paramount for data analysis since in practice, data analysis involves
-trials and errors on different steps to solve the problems. Analysts
-need the freedom to quickly experiment with different ideas when working
-with data. The tidyverse suite provides a collection of R package for
-conducting data analysis. These tools directly operate on data, and
-users, even new users, can easily read from the code on what operation
-is performed on the data and what to expect as the output. All of these
-are built upon a simple prior knowledge of the tidy data principle
-(Wickham 2014), which prescribes three rules for structuring tabular
-data: 1) Each variable forms a column, 2) Each observation forms a row,
-and; 3) Each type of observational unit forms a table.
+paramount for data analysis since in practice, data analysis is an
+iterative process that involves trials and errors to solve the problems.
+Analysts need the freedom to quickly experiment with different ideas
+when working with data. The tidyverse suite provides a collection of R
+package for conducting data analysis. These tools directly operate on
+data, and users, even new users, can easily read from the code on what
+operation is performed on the data and what to expect as the output. All
+of these are built upon a simple prior knowledge of the tidy data
+principle (Wickham 2014), which prescribes three rules to structure
+tabular data: 1) Each variable forms a column, 2) Each observation forms
+a row, and; 3) Each type of observational unit forms a table.
 
-Constructing indices also involves this working pattern of trails and
+Constructing indices also involves this working pattern of trials and
 errors on different options before finding the final solution. A tidy
@@
 
-Constructing a pipeline that divides a complex procedure into steps that
-can be concatenated has been adopted widely in the R community.
+The concept of pipeline (and pipe) is not new concept to computing. It
+refers to a set of data processing elements connected in series, where
+the output of one element is the input of the next one. Wickham et al.
+(2009) argues that whether made explicit or not, pipeline has to be
+presented in every graphics program. The paper also argues that breaking
+down graphic rendering into steps is beneficial for understand the
+implementation and compare between different graphic systems. Both
+arguments are applicable to index construction. Papers that proposing
+new indices always present a workflow diagram to illustrate how data is
+transformed into the final indices, but not much implementation has
+explicitly present this this step-by-step transformation in the code
+design.
 
-The data pipeline in interactive graphics is a set of steps that
-transform the raw data to the plots displayed on the screen. The initial
-pipeline proposed by Buja et al. (1988) involves the following steps:
-non-linear transformation, variables standardization, randomization,
-projection engine, and viewporting. The initial pipeline proposed by
-Buja et al. (1988) involves the following steps: non-linear
-transformation, variables standardization, randomization, projection
-engine, and viewporting. Another example in the early work of pipeline
-by Sutherland et al. (2000) describes a three-step pipeline: variable
-standardization, dimension reduction, and scaling data into the viewing
-window. This pipeline also includes the transformation on spatial and
-temporal variables, i.e.~computing time lag on temporal variables. This
-pipeline also includes the transformation on spatial and temporal
-variables, i.e.~computing time lag on temporal variables. Wickham et al.
-(2009) argues that whether made explicit or not, pipeline has to be
-presented in every graphics program and breaking down graphic rendering
-into steps is also beneficial for understand the implementation and
-compare between different graphic systems.
-
-The data pipeline concept is further enhanced by the pipe operator
-(\texttt{\%\textgreater{}\%}) in R where a set of operations, or steps,
-can be chained together to form a set of instructions.
-
-A more recent data pipeline is tidymodels (Kuhn and Wickham 2020), a set
-of packages for machine learning models following the tidyverse
-principles (Wickham et al. 2019). Steps: Exploratory data analysis
-(EDA), feature engineering, model tuning and selection, and model
-evaluation.
-
-\begin{itemize}
-\tightlist
-\item
-  easy to operate properly: should be designed that users know what is
-  approriate to do
-\item
-  promote good scientific methodology: should protect uses from doing
-  stupid things, doing right thing
-\end{itemize}
-
-good statistical practice
+To better understand the pipeline approach in statistic computing,
+examples in interactive graphics and machine learning models are given.
+In interactive graphics, the data pipeline is used to describe how the
+data is processed to be rendered on the screen. Examples of early work
+on building pipeline for interactive graphics includes Buja et al.
+(1988), Sutherland et al. (2000), and Xie, Hofmann, and Cheng (2014).
+Steps in these pipelines include non-linear transformation, variable
+standardization, randomization and dimension reduction. A more familiar
+example from tidymodels (Kuhn and Wickham 2020) presents a set of
+packages for machine learning models following the tidyverse principles
+(Wickham et al. 2019). The machine learning pipeline constitutes data
+resampling, feature engineering, model fitting, model tuning, and model
+evaluation each being responsible by a separate package. Before
+tidymodel, analysts who wish to fit different models need to learn the
+workflow of each individual algorithm and need to rewrite most of the
+model fitting, tuning, and evaluation code when switching from one model
+to another. The tidymodel approach unifies a wide range of models under
+a single pipeline where users can substitute different models while
+maintain the same pipeline. This allows users to easily experiment with
+different machine learning models a lot easier.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-xie_reactive_2014}{}}%
+Xie, Yihui, Heike Hofmann, and Xiaoyue Cheng. 2014. {``Reactive
+{Programming} for {Interactive} {Graphics}.''} \emph{Statistical
+Science} 29 (2): 201--13.
+\url{https://www.jstor.org/stable/43288470?seq=1}.
+
 \end{CSLReferences}

### 2022-10-26 paper/paper.qmd
changes after meeting + a subsection on a toy example

@@
 bibliography: bibliography.bib  
+editor_options: 
+  chunk_output_type: inline
 ---
@@
 
-## Tidy data {#sec-tidy-data}
+The concept of pipeline (and pipe) is not a new concept to computing. It refers to a set of data processing elements connected in series, where the output of one element is the input of the next one.
+@wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program.
+The paper also argues that breaking down graphic rendering into steps is beneficial for understand the implementation and compare between different graphic systems.
 
-As a scripted language for data analysis, R provides a interactive environment for users to code with data without the abstraction a typical programming language requires. 
-This flexible environment is paramount for data analysis since in practice, data analysis is an iterative process that involves trials and errors to solve the problems.
-Analysts need the freedom to quickly experiment with different ideas when working with data.
+*The pipeline system would work partially because of an agreed data structure (tidy data) and [same input and output format] - why the pipeline approach would work.*
+
 The tidyverse suite provides a collection of R package for conducting data analysis.
@@
 
-<!-- While data can be arranged in various different ways, maintaining a fixed representation gives a predictable output of operations on data.  -->
-
-<!-- Only when each step in the workflow has been mapped out, do data analysts need to consider those abstract concept of object oriented programming and functional programming to help them design the workflow.
--->
-<!-- how raw data should be processed to the final results can only be consolidated when  -->
-
-<!-- Before the concept of tidy data [@wickham_tidy_2014], tabular data arrive at data analysts in all different ways.  -->
-<!-- Different analysts would write customised scripts for analysing the specific data.  -->
-<!-- These scripts can be extended to other data analysed by the same people or group but this is not generalizable directly to another dataset.  -->
-<!-- When the tidy data concept comes, variables and values are arranged so that 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
--->
-<!-- With this specific layout, wrangling on tabular data can be standardised into a grammar of data manipulation in `dplyr` [@dplyr].  -->
-
-<!-- A similar issue happens with index construction where researchers construct their own indices in their own ways and there has not yet been a tidy principle on index construction. -->
-<!-- Also, this tidy principle on index construction is more complex than those in tidy data and the `dplyr` package. -->
-<!-- It has to encompass the workflow of transformation from the raw data towards the final index series. -->
-
-## Data pipeline {#sec-data-pipeline}
-
-The concept of pipeline (and pipe) is not new concept to computing. It refers to a set of data processing elements connected in series, where the output of one element is the input of the next one.
-@wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program.
-The paper also argues that breaking down graphic rendering into steps is beneficial for understand the implementation and compare between different graphic systems. 
-Both arguments are applicable to index construction.
-Papers that proposing new indices always present a workflow diagram to illustrate how data is transformed into the final indices, but not much implementation has explicitly present this this step-by-step transformation in the code design.
-
 To better understand the pipeline approach in statistic computing, examples in interactive graphics and machine learning models are given.
@@
 Steps in these pipelines include non-linear transformation, variable standardization, randomization and dimension reduction.
+
 A more familiar example from tidymodels [@tidymodels] presents a set of packages for machine learning models following the tidyverse principles [@tidyverse]. 
@@
 
+Both arguments are applicable to index construction.
+Papers that proposing new indices always present a workflow diagram to illustrate how data is transformed into the final indices, but not much implementation has explicitly present this this step-by-step transformation in the code design.
+
+\newpage
+
 # A pipeline for building statistical indices {#sec-a-pipeline-for-building-statistical-indices}
+
+## How does the pipeline constructin of an index look like?
+
+Consider a commonly used drought index: Standardised Precipitation Index (SPI) [@spi].
+Its construction involves the following steps: 1) aggregate monthly summed precipitation into a time scale $k$, 2) fit the aggregated series into a distribution to get the density values, and 3) convert the density values to normal quantiles.
+Using the pipeline approach, the index will be constructed as:
+
+```{r eval = FALSE, echo = TRUE}
+DATA %>% 
+  ts_aggregate(scale = 12, col = prcp) %>% 
+  normalising(dist = "gamma", col = prcp, fit_method = "lmom") %>% 
+  augment(col = "prcp_agg")
+```
+
+*The exact code needs to be further turned: 1) use verb: ts_aggreate -> aggregate_ts?, normalising -> normalise, 2) using `var` instead of `col` since it may be confused with "color", 3) normalising -> fit?
+since technically this step is no longer to normalise the fitted density value into normal quantiles but a distribution fitting step.*
+
+Here the column `prcp` is first temporally *aggregated* with a time scale of 12.
+The output is then normalised with a gamma distribution, fitted using the L-moment.
+The `augment` step, following the verb in the `broom` package, accepts the fitted distributions and adds information about each month in the dataset to get the final index.
+
+```{r eval = FALSE, echo = FALSE}
+res <- w %>% 
+  ts_aggregate(scale = 12, date = ym, id = id, col = prcp) %>% 
+  normalising(dist = "gamma", date = ym, id = id, col = prcp,
+              method = "lmom") %>% 
+  augment(id = id, date = ym,  col = "prcp_agg")
+
+res <- res %>% 
+  rename(
+         pipeline_d = prcp_agg,
+         pipeline_idx = .index)
+
+data(wichita)
+wichita <- ts(wichita[,-c(1,2)], end=c(2011,10), frequency=12)
+spi12 <- spi(wichita[,'PRCP'], 12)
+from_SPI <- tibble(
+  ym = yearmonth(zoo::yearmon(zoo::index(wichita))), 
+  SPI_idx = as.vector(spi12$fitted)
+)
+
+compare <- res %>% left_join(from_SPI, by = "ym") 
+compare %>% 
+  ggplot(aes(x = SPI_idx, y = pipeline_idx)) + 
+  geom_point() + 
+  theme_bw() + 
+  theme(legend.position = "bottom", 
+        aspect.ratio = 1) +
+  ggtitle("Index values of Standardised Precipitation Index (SPI)", 
+          subtitle = "using wichita data from the SPEI package") + 
+  xlab("Calculated from a single command") + 
+  ylab("Calculated from pipeline")
+
+```
+
+```{r fig-toy-example, eval = TRUE, echo = FALSE}
+#| fig-align: center
+#| fig-cap: "this is the caption"
+#| out-height: 50%
+#| out-width: 50%
+knitr::include_graphics(here::here("figures/toy-example-spi.png"))
+```
+
+While the calculated index value from the pipeline approach is the same as those calculated using a single line command like `spi(DATA, ARGUMENTS)`, as shown in @fig-toy-example, the pipeline approach gives more flexibility to perform diagnostics on the intermediate steps and to extend on the initial index calculation.
+For example, model goodness of fit test can be checked after `normalising()` through:
+
+```{r eval = FALSE, echo = TRUE}
+normalising(dist = "gamma", col = d, fit_method = "lmom") %>% 
+  gof_test()
+```
+
+and bootstrap uncertainty on the index, with 100 boostrap samples, can be calculated using `n_boot = 100` in the `normalising()` as:
+
+```{r eval = FALSE, echo = TRUE}
+normalising(dist = "gamma", col = d, fit_method = "lmom", n_boot = 100) 
+```
+
+## Pipeline steps for constructing indies
+
+*put a diagram here to summarise the pipeline*
 
@@
 
-## Raw data
+each step as a point rather than subsection
 
-**Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem.
-After processing them and align them together they become the "raw data"**
+**Raw data**
+
+*Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem.
+After processing them and align them together they become the "raw data"*
 
@@
 
-## Spatial aggregation
+**Spatial aggregation**
 
@@
 
-## Scaling
+**Scaling**
 
@@
 
-## Normalising
+**Normalising**
 
@@
 
-## Variable transformation
+**Variable transformation**
 
@@
 
-## Temporal processing
+**Temporal processing**
 
-## Dimension reduction
+**Dimension reduction**
 
@@
 
-## Benchmarking
+**Benchmarking**
 
-## Simplification
+**Simplification**
 

### 2022-10-26 paper/paper.tex
changes after meeting + a subsection on a toy example

@@
 \fi
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{sec-tidy-data}{%
-\subsection{Tidy data}\label{sec-tidy-data}}
+The concept of pipeline (and pipe) is not a new concept to computing. It
+refers to a set of data processing elements connected in series, where
+the output of one element is the input of the next one. Wickham et al.
+(2009) argues that whether made explicit or not, pipeline has to be
+presented in every graphics program. The paper also argues that breaking
+down graphic rendering into steps is beneficial for understand the
+implementation and compare between different graphic systems.
 
-As a scripted language for data analysis, R provides a interactive
-environment for users to code with data without the abstraction a
-typical programming language requires. This flexible environment is
-paramount for data analysis since in practice, data analysis is an
-iterative process that involves trials and errors to solve the problems.
-Analysts need the freedom to quickly experiment with different ideas
-when working with data. The tidyverse suite provides a collection of R
-package for conducting data analysis. These tools directly operate on
-data, and users, even new users, can easily read from the code on what
-operation is performed on the data and what to expect as the output. All
-of these are built upon a simple prior knowledge of the tidy data
-principle (Wickham 2014), which prescribes three rules to structure
-tabular data: 1) Each variable forms a column, 2) Each observation forms
-a row, and; 3) Each type of observational unit forms a table.
+\emph{The pipeline system would work partially because of an agreed data
+structure (tidy data) and {[}same input and output format{]} - why the
+pipeline approach would work.}
+
+The tidyverse suite provides a collection of R package for conducting
+data analysis. These tools directly operate on data, and users, even new
+users, can easily read from the code on what operation is performed on
+the data and what to expect as the output. All of these are built upon a
+simple prior knowledge of the tidy data principle (Wickham 2014), which
+prescribes three rules to structure tabular data: 1) Each variable forms
+a column, 2) Each observation forms a row, and; 3) Each type of
+observational unit forms a table.
 
@@
 they wish and while maintain a readable and intuitive code base.
-
-\hypertarget{sec-data-pipeline}{%
-\subsection{Data pipeline}\label{sec-data-pipeline}}
-
-The concept of pipeline (and pipe) is not new concept to computing. It
-refers to a set of data processing elements connected in series, where
-the output of one element is the input of the next one. Wickham et al.
-(2009) argues that whether made explicit or not, pipeline has to be
-presented in every graphics program. The paper also argues that breaking
-down graphic rendering into steps is beneficial for understand the
-implementation and compare between different graphic systems. Both
-arguments are applicable to index construction. Papers that proposing
-new indices always present a workflow diagram to illustrate how data is
-transformed into the final indices, but not much implementation has
-explicitly present this this step-by-step transformation in the code
-design.
 
@@
 Steps in these pipelines include non-linear transformation, variable
-standardization, randomization and dimension reduction. A more familiar
-example from tidymodels (Kuhn and Wickham 2020) presents a set of
-packages for machine learning models following the tidyverse principles
-(Wickham et al. 2019). The machine learning pipeline constitutes data
-resampling, feature engineering, model fitting, model tuning, and model
-evaluation each being responsible by a separate package. Before
-tidymodel, analysts who wish to fit different models need to learn the
-workflow of each individual algorithm and need to rewrite most of the
-model fitting, tuning, and evaluation code when switching from one model
-to another. The tidymodel approach unifies a wide range of models under
-a single pipeline where users can substitute different models while
-maintain the same pipeline. This allows users to easily experiment with
-different machine learning models a lot easier.
+standardization, randomization and dimension reduction.
+
+A more familiar example from tidymodels (Kuhn and Wickham 2020) presents
+a set of packages for machine learning models following the tidyverse
+principles (Wickham et al. 2019). The machine learning pipeline
+constitutes data resampling, feature engineering, model fitting, model
+tuning, and model evaluation each being responsible by a separate
+package. Before tidymodel, analysts who wish to fit different models
+need to learn the workflow of each individual algorithm and need to
+rewrite most of the model fitting, tuning, and evaluation code when
+switching from one model to another. The tidymodel approach unifies a
+wide range of models under a single pipeline where users can substitute
+different models while maintain the same pipeline. This allows users to
+easily experiment with different machine learning models a lot easier.
+
+Both arguments are applicable to index construction. Papers that
+proposing new indices always present a workflow diagram to illustrate
+how data is transformed into the final indices, but not much
+implementation has explicitly present this this step-by-step
+transformation in the code design.
+
+\newpage
 
@@
 indices}\label{sec-a-pipeline-for-building-statistical-indices}}
+
+\hypertarget{how-does-the-pipeline-constructin-of-an-index-look-like}{%
+\subsection{How does the pipeline constructin of an index look
+like?}\label{how-does-the-pipeline-constructin-of-an-index-look-like}}
+
+Consider a commonly used drought index: Standardised Precipitation Index
+(SPI) (McKee, Doesken, and Kleist 1993). Its construction involves the
+following steps: 1) aggregate monthly summed precipitation into a time
+scale \(k\), 2) fit the aggregated series into a distribution to get the
+density values, and 3) convert the density values to normal quantiles.
+Using the pipeline approach, the index will be constructed as:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{ts\_aggregate}\NormalTok{(}\AttributeTok{scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{col =}\NormalTok{ prcp) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{normalising}\NormalTok{(}\AttributeTok{dist =} \StringTok{"gamma"}\NormalTok{, }\AttributeTok{col =}\NormalTok{ prcp, }\AttributeTok{fit\_method =} \StringTok{"lmom"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{augment}\NormalTok{(}\AttributeTok{col =} \StringTok{"prcp\_agg"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\emph{The exact code needs to be further turned: 1) use verb:
+ts\_aggreate -\textgreater{} aggregate\_ts?, normalising -\textgreater{}
+normalise, 2) using \texttt{var} instead of \texttt{col} since it may be
+confused with ``color'', 3) normalising -\textgreater{} fit? since
+technically this step is no longer to normalise the fitted density value
+into normal quantiles but a distribution fitting step.}
+
+Here the column \texttt{prcp} is first temporally \emph{aggregated} with
+a time scale of 12. The output is then normalised with a gamma
+distribution, fitted using the L-moment. The \texttt{augment} step,
+following the verb in the \texttt{broom} package, accepts the fitted
+distributions and adds information about each month in the dataset to
+get the final index.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.5\textwidth,height=0.5\textheight]{../figures/toy-example-spi.png}
+
+}
+
+\caption{\label{fig-toy-example}this is the caption}
+
+\end{figure}
+
+While the calculated index value from the pipeline approach is the same
+as those calculated using a single line command like
+\texttt{spi(DATA,\ ARGUMENTS)}, as shown in
+Figure~\ref{fig-toy-example}, the pipeline approach gives more
+flexibility to perform diagnostics on the intermediate steps and to
+extend on the initial index calculation. For example, model goodness of
+fit test can be checked after \texttt{normalising()} through:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{normalising}\NormalTok{(}\AttributeTok{dist =} \StringTok{"gamma"}\NormalTok{, }\AttributeTok{col =}\NormalTok{ d, }\AttributeTok{fit\_method =} \StringTok{"lmom"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{gof\_test}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+and bootstrap uncertainty on the index, with 100 boostrap samples, can
+be calculated using \texttt{n\_boot\ =\ 100} in the
+\texttt{normalising()} as:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{normalising}\NormalTok{(}\AttributeTok{dist =} \StringTok{"gamma"}\NormalTok{, }\AttributeTok{col =}\NormalTok{ d, }\AttributeTok{fit\_method =} \StringTok{"lmom"}\NormalTok{, }\AttributeTok{n\_boot =} \DecValTok{100}\NormalTok{) }
+\end{Highlighting}
+\end{Shaded}
+
+\hypertarget{pipeline-steps-for-constructing-indies}{%
+\subsection{Pipeline steps for constructing
+indies}\label{pipeline-steps-for-constructing-indies}}
+
+\emph{put a diagram here to summarise the pipeline}
 
@@
 
-\hypertarget{raw-data}{%
-\subsection{Raw data}\label{raw-data}}
+each step as a point rather than subsection
 
-\textbf{Another section on original data directly downloaded, can have
+\textbf{Raw data}
+
+\emph{Another section on original data directly downloaded, can have
 different spatial resolution, temporal granularity, data quality
@@
 
-\hypertarget{spatial-aggregation}{%
-\subsection{Spatial aggregation}\label{spatial-aggregation}}
+\textbf{Spatial aggregation}
 
@@
 
-\hypertarget{scaling}{%
-\subsection{Scaling}\label{scaling}}
+\textbf{Scaling}
 
@@
 
-\hypertarget{normalising}{%
-\subsection{Normalising}\label{normalising}}
+\textbf{Normalising}
 
@@
 
-\hypertarget{variable-transformation}{%
-\subsection{Variable transformation}\label{variable-transformation}}
+\textbf{Variable transformation}
 
@@
 
-\hypertarget{temporal-processing}{%
-\subsection{Temporal processing}\label{temporal-processing}}
+\textbf{Temporal processing}
 
-\hypertarget{dimension-reduction}{%
-\subsection{Dimension reduction}\label{dimension-reduction}}
+\textbf{Dimension reduction}
 
@@
 
-\hypertarget{benchmarking}{%
-\subsection{Benchmarking}\label{benchmarking}}
+\textbf{Benchmarking}
 
-\hypertarget{simplification}{%
-\subsection{Simplification}\label{simplification}}
+\textbf{Simplification}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-spi}{}}%
+McKee, Thomas B, Nolan J Doesken, and John Kleist. 1993. {``The
+Relationship of Drought Frequency and Duration to Time Scales.''} In
+\emph{Proceedings of the 8th Conference of Applied Climatology},
+179--84. Anaheim, CA: American Meteorological Society.
+
 \leavevmode\vadjust pre{\hypertarget{ref-sutherland_orca_2000}{}}%

### 2022-10-27 paper/paper.qmd
restructure section 2 data pipeline in R

@@
 keywords:
-  - spatio-temporal data
   - indices
   - data pipeline
+  - software design
 bibliography: bibliography.bib  
@@
 
+**Why you should care about pipeline**
+
 The concept of pipeline (and pipe) is not a new concept to computing. It refers to a set of data processing elements connected in series, where the output of one element is the input of the next one.
@@
 The paper also argues that breaking down graphic rendering into steps is beneficial for understand the implementation and compare between different graphic systems. 
+The discussion on pipeline construction is well documented in early interactive graphics software: @buja_elements_1988, @sutherland_orca_2000, and @xie_reactive_2014 and their pipeline steps include non-linear transformation, variable standardization, randomization and dimension reduction.
 
-*The pipeline system would work partially because of an agreed data structure (tidy data) and [same input and output format] - why the pipeline approach would work.*
+**What is pipeline, its underlying software design philosophy, and how these are reflected in R**
 
-The tidyverse suite provides a collection of R package for conducting data analysis.
-These tools directly operate on data, and users, even new users, can easily read from the code on what operation is performed on the data and what to expect as the output.
-All of these are built upon a simple prior knowledge of the tidy data principle [@wickham_tidy_2014], which prescribes three rules to structure tabular data: 1) Each variable forms a column, 2) Each observation forms a row, and; 3) Each type of observational unit forms a table.
+One of the most commonly known pipeline example is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program. 
+The pipeline design avoid writing a complicated program to solve a complex problem but building simple programs that do one thing well and work well together.
+This design philosophy is also embraced in the R community by the tidyverse principles. 
+In tidyverse, the flow from the previous program to the next is ensured through reusing the tidy data format [@wickham_tidy_2014], which prescribes three rules on how to layout tabular data.
+To solve a complicated data problem using tidyverse, one typically builds the solution using a collection of functions and these functions have a strong human centered design with intuitive naming and easy to reference through autocomplete.
 
-Constructing indices also involves this working pattern of trials and errors on different options before finding the final solution.
-A tidy workflow will give researchers more flexibility to tune the steps as they wish and while maintain a readable and intuitive code base.
+**A proper pipeline example in R: tidymodel**
 
-To better understand the pipeline approach in statistic computing, examples in interactive graphics and machine learning models are given.
-In interactive graphics, the data pipeline is used to describe how the data is processed to be rendered on the screen. 
-Examples of early work on building pipeline for interactive graphics includes @buja_elements_1988, @sutherland_orca_2000, and @xie_reactive_2014. 
-Steps in these pipelines include non-linear transformation, variable standardization, randomization and dimension reduction.
-
-A more familiar example from tidymodels [@tidymodels] presents a set of packages for machine learning models following the tidyverse principles [@tidyverse]. 
+A pipeline example in R that adopts the tidyverse principle is the tidymodel suite [@tidymodels] for building machine learning models. 
 The machine learning pipeline constitutes data resampling, feature engineering, model fitting, model tuning, and model evaluation each being responsible by a separate package. 
@@
 
+**Constructing indices would also benefit from pipeline and embracing the aforementioned design philosophy.**
+
+*more work here*
+
 Both arguments are applicable to index construction.
 Papers that proposing new indices always present a workflow diagram to illustrate how data is transformed into the final indices, but not much implementation has explicitly present this this step-by-step transformation in the code design.
+Constructing indices also involves this working pattern of trials and errors on different options before finding the final solution.
+A tidy workflow will give researchers more flexibility to tune the steps as they wish and while maintain a readable and intuitive code base.
 

### 2022-10-27 paper/paper.tex
restructure section 2 data pipeline in R

@@
   pdfauthor={H. Sherry Zhang; Collaborators},
-  pdfkeywords={spatio-temporal data, indices, data pipeline},
+  pdfkeywords={indices, data pipeline, software design},
   colorlinks=true,
@@
 \def\sep{\textbullet\ }
-spatio-temporal data \sep indices \sep 
-data pipeline
+indices \sep data pipeline \sep 
+software design
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
+\textbf{Why you should care about pipeline}
+
 The concept of pipeline (and pipe) is not a new concept to computing. It
@@
 down graphic rendering into steps is beneficial for understand the
-implementation and compare between different graphic systems.
+implementation and compare between different graphic systems. The
+discussion on pipeline construction is well documented in early
+interactive graphics software: Buja et al. (1988), Sutherland et al.
+(2000), and Xie, Hofmann, and Cheng (2014) and their pipeline steps
+include non-linear transformation, variable standardization,
+randomization and dimension reduction.
 
-\emph{The pipeline system would work partially because of an agreed data
-structure (tidy data) and {[}same input and output format{]} - why the
-pipeline approach would work.}
+\textbf{What is pipeline, its underlying software design philosophy, and
+how these are reflected in R}
 
-The tidyverse suite provides a collection of R package for conducting
-data analysis. These tools directly operate on data, and users, even new
-users, can easily read from the code on what operation is performed on
-the data and what to expect as the output. All of these are built upon a
-simple prior knowledge of the tidy data principle (Wickham 2014), which
-prescribes three rules to structure tabular data: 1) Each variable forms
-a column, 2) Each observation forms a row, and; 3) Each type of
-observational unit forms a table.
+One of the most commonly known pipeline example is perhaps the Unix
+pipeline where programs can be concatenated with \texttt{\textbar{}} to
+flow the output from the last program into the next program. The
+pipeline design avoid writing a complicated program to solve a complex
+problem but building simple programs that do one thing well and work
+well together. This design philosophy is also embraced in the R
+community by the tidyverse principles. In tidyverse, the flow from the
+previous program to the next is ensured through reusing the tidy data
+format (Wickham 2014), which prescribes three rules on how to layout
+tabular data. To solve a complicated data problem using tidyverse, one
+typically builds the solution using a collection of functions and these
+functions have a strong human centered design with intuitive naming and
+easy to reference through autocomplete.
 
-Constructing indices also involves this working pattern of trials and
-errors on different options before finding the final solution. A tidy
-workflow will give researchers more flexibility to tune the steps as
-they wish and while maintain a readable and intuitive code base.
+\textbf{A proper pipeline example in R: tidymodel}
 
-To better understand the pipeline approach in statistic computing,
-examples in interactive graphics and machine learning models are given.
-In interactive graphics, the data pipeline is used to describe how the
-data is processed to be rendered on the screen. Examples of early work
-on building pipeline for interactive graphics includes Buja et al.
-(1988), Sutherland et al. (2000), and Xie, Hofmann, and Cheng (2014).
-Steps in these pipelines include non-linear transformation, variable
-standardization, randomization and dimension reduction.
+A pipeline example in R that adopts the tidyverse principle is the
+tidymodel suite (Kuhn and Wickham 2020) for building machine learning
+models. The machine learning pipeline constitutes data resampling,
+feature engineering, model fitting, model tuning, and model evaluation
+each being responsible by a separate package. Before tidymodel, analysts
+who wish to fit different models need to learn the workflow of each
+individual algorithm and need to rewrite most of the model fitting,
+tuning, and evaluation code when switching from one model to another.
+The tidymodel approach unifies a wide range of models under a single
+pipeline where users can substitute different models while maintain the
+same pipeline. This allows users to easily experiment with different
+machine learning models a lot easier.
 
-A more familiar example from tidymodels (Kuhn and Wickham 2020) presents
-a set of packages for machine learning models following the tidyverse
-principles (Wickham et al. 2019). The machine learning pipeline
-constitutes data resampling, feature engineering, model fitting, model
-tuning, and model evaluation each being responsible by a separate
-package. Before tidymodel, analysts who wish to fit different models
-need to learn the workflow of each individual algorithm and need to
-rewrite most of the model fitting, tuning, and evaluation code when
-switching from one model to another. The tidymodel approach unifies a
-wide range of models under a single pipeline where users can substitute
-different models while maintain the same pipeline. This allows users to
-easily experiment with different machine learning models a lot easier.
+\textbf{Constructing indices would also benefit from pipeline and
+embracing the aforementioned design philosophy.}
+
+\emph{more work here}
 
@@
 implementation has explicitly present this this step-by-step
-transformation in the code design.
+transformation in the code design. Constructing indices also involves
+this working pattern of trials and errors on different options before
+finding the final solution. A tidy workflow will give researchers more
+flexibility to tune the steps as they wish and while maintain a readable
+and intuitive code base.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
-Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
-D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
-{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
-4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%

### 2022-10-31 paper/paper.qmd
section 2 draft

@@
 
-The concept of pipeline (and pipe) is not a new concept to computing. It refers to a set of data processing elements connected in series, where the output of one element is the input of the next one.
-@wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program.
-The paper also argues that breaking down graphic rendering into steps is beneficial for understand the implementation and compare between different graphic systems. 
+Data pipeline is not a new concept to computing. It refers to a set of data processing elements connected in series, where the output of one element is the input of the next one.
+@wickham_plumbing_2009 argues that whether made explicit or not, the pipeline has to be presented in every graphics program.
+The paper also argues that breaking down graphic rendering into steps is beneficial for understanding the implementation and comparing between different graphic systems. 
 The discussion on pipeline construction is well documented in early interactive graphics software: @buja_elements_1988, @sutherland_orca_2000, and @xie_reactive_2014 and their pipeline steps include non-linear transformation, variable standardization, randomization and dimension reduction.
@@
 
-One of the most commonly known pipeline example is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program. 
-The pipeline design avoid writing a complicated program to solve a complex problem but building simple programs that do one thing well and work well together.
-This design philosophy is also embraced in the R community by the tidyverse principles. 
-In tidyverse, the flow from the previous program to the next is ensured through reusing the tidy data format [@wickham_tidy_2014], which prescribes three rules on how to layout tabular data.
-To solve a complicated data problem using tidyverse, one typically builds the solution using a collection of functions and these functions have a strong human centered design with intuitive naming and easy to reference through autocomplete.
+One of the most commonly known pipeline examples is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program, i.e.
 
-**A proper pipeline example in R: tidymodel**
+````
+command 1 | command 2 | command 3 | ...
+````
 
-A pipeline example in R that adopts the tidyverse principle is the tidymodel suite [@tidymodels] for building machine learning models. 
-The machine learning pipeline constitutes data resampling, feature engineering, model fitting, model tuning, and model evaluation each being responsible by a separate package. 
-Before tidymodel, analysts who wish to fit different models need to learn the workflow of each individual algorithm and need to rewrite most of the model fitting, tuning, and evaluation code when switching from one model to another.
-The tidymodel approach unifies a wide range of models under a single pipeline where users can substitute different models while maintain the same pipeline.
-This allows users to easily experiment with different machine learning models a lot easier.
+To solve a complex problem, the Unix system builds simple programs that do one thing well and work well together.
+This design is also reflected in the tidyverse ecosystem in R.
+To solve a complicated data problem using tidyverse, analysts typically build the solution using a collection of tools from the tidyverse toolbox. 
+The data object can flow smoothly from one command to the next, safeguarded by the tidy data format [@wickham_tidy_2014], which prescribes three rules on how to lay out tabular data.
+The tidyverse tools also embrace a strong human-centered design where function names are intuitive and easy to reference through autocomplete.
+With the tidyverse design principle in mind, the tidymodel suite enables analysts to build machine learning models through the data pipeline. 
+It includes typical tasks required in machine learning like data resampling, feature engineering, model fitting, model tuning, and model evaluation.
+An advantage of tidymodel pipeline over separate software for individual models is that analysts no longer need to write model-specific syntax to work with each model, but pipeline-specific syntax that is applicable to all the models implemented in tidymodel.
+This allows users to easily experiment with a collection of machine learning models.
 
@@
 
-*more work here*
-
-Both arguments are applicable to index construction.
-Papers that proposing new indices always present a workflow diagram to illustrate how data is transformed into the final indices, but not much implementation has explicitly present this this step-by-step transformation in the code design.
-Constructing indices also involves this working pattern of trials and errors on different options before finding the final solution.
-A tidy workflow will give researchers more flexibility to tune the steps as they wish and while maintain a readable and intuitive code base.
+In index construction, data pipeline is often presented in a workflow diagram in the research paper to illustrate how the raw data is transformed into the final indices. 
+This agrees with Wickham's argument on the presence of the data pipeline, however, more often than not, the pipeline is not made explicit in the software.
+Often the time, all the steps are lumped into a single wrapper function, rather than being split into smaller, modulated functions. 
+This increases the cost of maintaining and understanding the code base, gives analysts little freedom to customise the indices for specific needs, and hinders reusing existing code for building new indices.
+A pipeline approach unites a range of indices under a single data pipeline and analysts can compose indices from pipeline steps like building Legos from individual bricks.
+In this workflow, analysts are not limited by indices that have been already proposed and can easily combine pipeline steps to compose novel indices.
+Analysis of the indices (i.e.
+calculation of uncertainty) is also feasible by adding external code into the pipeline.
 

### 2022-10-31 paper/paper.tex
section 2 draft

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, frame hidden, boxrule=0pt, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-The concept of pipeline (and pipe) is not a new concept to computing. It
-refers to a set of data processing elements connected in series, where
-the output of one element is the input of the next one. Wickham et al.
-(2009) argues that whether made explicit or not, pipeline has to be
-presented in every graphics program. The paper also argues that breaking
-down graphic rendering into steps is beneficial for understand the
-implementation and compare between different graphic systems. The
-discussion on pipeline construction is well documented in early
-interactive graphics software: Buja et al. (1988), Sutherland et al.
-(2000), and Xie, Hofmann, and Cheng (2014) and their pipeline steps
-include non-linear transformation, variable standardization,
-randomization and dimension reduction.
+Data pipeline is not a new concept to computing. It refers to a set of
+data processing elements connected in series, where the output of one
+element is the input of the next one. Wickham et al. (2009) argues that
+whether made explicit or not, the pipeline has to be presented in every
+graphics program. The paper also argues that breaking down graphic
+rendering into steps is beneficial for understanding the implementation
+and comparing between different graphic systems. The discussion on
+pipeline construction is well documented in early interactive graphics
+software: Buja et al. (1988), Sutherland et al. (2000), and Xie,
+Hofmann, and Cheng (2014) and their pipeline steps include non-linear
+transformation, variable standardization, randomization and dimension
+reduction.
 
@@
 
-One of the most commonly known pipeline example is perhaps the Unix
+One of the most commonly known pipeline examples is perhaps the Unix
 pipeline where programs can be concatenated with \texttt{\textbar{}} to
-flow the output from the last program into the next program. The
-pipeline design avoid writing a complicated program to solve a complex
-problem but building simple programs that do one thing well and work
-well together. This design philosophy is also embraced in the R
-community by the tidyverse principles. In tidyverse, the flow from the
-previous program to the next is ensured through reusing the tidy data
-format (Wickham 2014), which prescribes three rules on how to layout
-tabular data. To solve a complicated data problem using tidyverse, one
-typically builds the solution using a collection of functions and these
-functions have a strong human centered design with intuitive naming and
-easy to reference through autocomplete.
+flow the output from the last program into the next program, i.e.~
 
-\textbf{A proper pipeline example in R: tidymodel}
+\begin{verbatim}
+command 1 | command 2 | command 3 | ...
+\end{verbatim}
 
-A pipeline example in R that adopts the tidyverse principle is the
-tidymodel suite (Kuhn and Wickham 2020) for building machine learning
-models. The machine learning pipeline constitutes data resampling,
-feature engineering, model fitting, model tuning, and model evaluation
-each being responsible by a separate package. Before tidymodel, analysts
-who wish to fit different models need to learn the workflow of each
-individual algorithm and need to rewrite most of the model fitting,
-tuning, and evaluation code when switching from one model to another.
-The tidymodel approach unifies a wide range of models under a single
-pipeline where users can substitute different models while maintain the
-same pipeline. This allows users to easily experiment with different
-machine learning models a lot easier.
+To solve a complex problem, the Unix system builds simple programs that
+do one thing well and work well together. This design is also reflected
+in the tidyverse ecosystem in R. To solve a complicated data problem
+using tidyverse, analysts typically build the solution using a
+collection of tools from the tidyverse toolbox. The data object can flow
+smoothly from one command to the next, safeguarded by the tidy data
+format (Wickham 2014), which prescribes three rules on how to lay out
+tabular data. The tidyverse tools also embrace a strong human-centered
+design where function names are intuitive and easy to reference through
+autocomplete. With the tidyverse design principle in mind, the tidymodel
+suite enables analysts to build machine learning models through the data
+pipeline. It includes typical tasks required in machine learning like
+data resampling, feature engineering, model fitting, model tuning, and
+model evaluation. An advantage of tidymodel pipeline over separate
+software for individual models is that analysts no longer need to write
+model-specific syntax to work with each model, but pipeline-specific
+syntax that is applicable to all the models implemented in tidymodel.
+This allows users to easily experiment with a collection of machine
+learning models.
 
@@
 
-\emph{more work here}
-
-Both arguments are applicable to index construction. Papers that
-proposing new indices always present a workflow diagram to illustrate
-how data is transformed into the final indices, but not much
-implementation has explicitly present this this step-by-step
-transformation in the code design. Constructing indices also involves
-this working pattern of trials and errors on different options before
-finding the final solution. A tidy workflow will give researchers more
-flexibility to tune the steps as they wish and while maintain a readable
-and intuitive code base.
+In index construction, data pipeline is often presented in a workflow
+diagram in the research paper to illustrate how the raw data is
+transformed into the final indices. This agrees with Wickham's argument
+on the presence of the data pipeline, however, more often than not, the
+pipeline is not made explicit in the software. Often the time, all the
+steps are lumped into a single wrapper function, rather than being split
+into smaller, modulated functions. This increases the cost of
+maintaining and understanding the code base, gives analysts little
+freedom to customise the indices for specific needs, and hinders reusing
+existing code for building new indices. A pipeline approach unites a
+range of indices under a single data pipeline and analysts can compose
+indices from pipeline steps like building Legos from individual bricks.
+In this workflow, analysts are not limited by indices that have been
+already proposed and can easily combine pipeline steps to compose novel
+indices. Analysis of the indices (i.e.~calculation of uncertainty) is
+also feasible by adding external code into the pipeline.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tidymodels}{}}%
-Kuhn, Max, and Hadley Wickham. 2020. \emph{Tidymodels: A Collection of
-Packages for Modeling and Machine Learning Using Tidyverse Principles.}
-\url{https://www.tidymodels.org}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-mckee1993relationship}{}}%

### 2022-11-02 paper/paper.qmd
polish the toy example on constructing SPI in Sect 3

@@
 Consider a commonly used drought index: Standardised Precipitation Index (SPI) [@spi].
-Its construction involves the following steps: 1) aggregate monthly summed precipitation into a time scale $k$, 2) fit the aggregated series into a distribution to get the density values, and 3) convert the density values to normal quantiles.
-Using the pipeline approach, the index will be constructed as:
+Its construction involves: 1) aggregating monthly precipitation (`prcp`) with a time scale, 2) fitting the aggregated series with a chosen distribution to get density values, and 3) converting the density values to normal quantiles.
+Under the pipeline approach, SPI will be constructed as:
 
@@
 DATA %>% 
-  ts_aggregate(scale = 12, col = prcp) %>% 
-  normalising(dist = "gamma", col = prcp, fit_method = "lmom") %>% 
-  augment(col = "prcp_agg")
+  aggregate(scale = 12, var = prcp) %>% 
+  normalise(dist = gamma, var = prcp, fit_method = lmom) %>% 
+  augment(var = .agg_prcp)
 ```
 
-*The exact code needs to be further turned: 1) use verb: ts_aggreate -> aggregate_ts?, normalising -> normalise, 2) using `var` instead of `col` since it may be confused with "color", 3) normalising -> fit?
-since technically this step is no longer to normalise the fitted density value into normal quantiles but a distribution fitting step.*
+<!-- *FWI:*  -->
 
-Here the column `prcp` is first temporally *aggregated* with a time scale of 12.
-The output is then normalised with a gamma distribution, fitted using the L-moment.
-The `augment` step, following the verb in the `broom` package, accepts the fitted distributions and adds information about each month in the dataset to get the final index.
+<!--   - *verbs as name for each step: `aggregate`, `normalise`* -->
+<!--   - *use NSE for variable name: `prcp`, `c(tavg, prcp)`* -->
+<!--   - *created variables are prefixed with dot: `.agg_prcp`* -->
+<!--   - *avoid string and use function to avoid "gamma" vs. "gam": `gamma()`, `lmom()`* -->
+
+Here each command corresponds to one step described above with additional arguments specific to the step. 
+The pipeline construction produces the same index value as command like `spi(...)`, as shown in @fig-toy-example, but with additional benefit:
+
+  - Multiple scales and multiple distributions can be fitted using the `c(...)` syntax to compare index values constructed from different parameterisations; 
+  - Intermediate results can be checked after each step; and
+  - Additional steps and analysis can be wired into the workflow for index diagnostics and customised user need.
 
@@
   ggplot(aes(x = SPI_idx, y = pipeline_idx)) + 
+  geom_abline(intercept = 0, slope = 1, linetype = "dashed") + 
   geom_point() + 
@@
         aspect.ratio = 1) +
-  ggtitle("Index values of Standardised Precipitation Index (SPI)", 
-          subtitle = "using wichita data from the SPEI package") + 
-  xlab("Calculated from a single command") + 
-  ylab("Calculated from pipeline")
-
+  xlab("Index values from `spi()` in package `SPEI`") + 
+  ylab("Index values from pipeline") + 
+  theme(panel.grid.minor = element_blank())
 ```
@@
 #| fig-align: center
-#| fig-cap: "this is the caption"
-#| out-height: 50%
-#| out-width: 50%
+#| fig-cap: "Standardised Precipitation Index (SPI) calculated using two approaches: the pipeline construction (y-axis) and the function `spi()` from the `SPEI` package (x-axis), using the `wichita` data from the package `SPEI`.
+All the points align diagnally, indicating the same numerical result from both approaches."
+#| out-height: 70%
+#| out-width: 70%
 knitr::include_graphics(here::here("figures/toy-example-spi.png"))
-```
-
-While the calculated index value from the pipeline approach is the same as those calculated using a single line command like `spi(DATA, ARGUMENTS)`, as shown in @fig-toy-example, the pipeline approach gives more flexibility to perform diagnostics on the intermediate steps and to extend on the initial index calculation.
-For example, model goodness of fit test can be checked after `normalising()` through:
-
-```{r eval = FALSE, echo = TRUE}
-normalising(dist = "gamma", col = d, fit_method = "lmom") %>% 
-  gof_test()
-```
-
-and bootstrap uncertainty on the index, with 100 boostrap samples, can be calculated using `n_boot = 100` in the `normalising()` as:
-
-```{r eval = FALSE, echo = TRUE}
-normalising(dist = "gamma", col = d, fit_method = "lmom", n_boot = 100) 
 ```

### 2022-11-02 paper/paper.tex
polish the toy example on constructing SPI in Sect 3

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, frame hidden, boxrule=0pt, sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 Consider a commonly used drought index: Standardised Precipitation Index
-(SPI) (McKee, Doesken, and Kleist 1993). Its construction involves the
-following steps: 1) aggregate monthly summed precipitation into a time
-scale \(k\), 2) fit the aggregated series into a distribution to get the
-density values, and 3) convert the density values to normal quantiles.
-Using the pipeline approach, the index will be constructed as:
+(SPI) (McKee, Doesken, and Kleist 1993). Its construction involves: 1)
+aggregating monthly precipitation (\texttt{prcp}) with a time scale, 2)
+fitting the aggregated series with a chosen distribution to get density
+values, and 3) converting the density values to normal quantiles. Under
+the pipeline approach, SPI will be constructed as:
 
@@
 \NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{ts\_aggregate}\NormalTok{(}\AttributeTok{scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{col =}\NormalTok{ prcp) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{normalising}\NormalTok{(}\AttributeTok{dist =} \StringTok{"gamma"}\NormalTok{, }\AttributeTok{col =}\NormalTok{ prcp, }\AttributeTok{fit\_method =} \StringTok{"lmom"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{augment}\NormalTok{(}\AttributeTok{col =} \StringTok{"prcp\_agg"}\NormalTok{)}
+  \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{var =}\NormalTok{ prcp) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{normalise}\NormalTok{(}\AttributeTok{dist =}\NormalTok{ gamma, }\AttributeTok{var =}\NormalTok{ prcp, }\AttributeTok{fit\_method =}\NormalTok{ lmom) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{augment}\NormalTok{(}\AttributeTok{var =}\NormalTok{ .agg\_prcp)}
 \end{Highlighting}
@@
 
-\emph{The exact code needs to be further turned: 1) use verb:
-ts\_aggreate -\textgreater{} aggregate\_ts?, normalising -\textgreater{}
-normalise, 2) using \texttt{var} instead of \texttt{col} since it may be
-confused with ``color'', 3) normalising -\textgreater{} fit? since
-technically this step is no longer to normalise the fitted density value
-into normal quantiles but a distribution fitting step.}
+Here each command corresponds to one step described above with
+additional arguments specific to the step. The pipeline construction
+produces the same index value as command like \texttt{spi(...)}, as
+shown in Figure~\ref{fig-toy-example}, but with additional benefit:
 
-Here the column \texttt{prcp} is first temporally \emph{aggregated} with
-a time scale of 12. The output is then normalised with a gamma
-distribution, fitted using the L-moment. The \texttt{augment} step,
-following the verb in the \texttt{broom} package, accepts the fitted
-distributions and adds information about each month in the dataset to
-get the final index.
+\begin{itemize}
+\tightlist
+\item
+  Multiple scales and multiple distributions can be fitted using the
+  \texttt{c(...)} syntax to compare index values constructed from
+  different parameterisations;
+\item
+  Intermediate results can be checked after each step; and
+\item
+  Additional steps and analysis can be wired into the workflow for index
+  diagnostics and customised user need.
+\end{itemize}
 
@@
 
-{\centering \includegraphics[width=0.5\textwidth,height=0.5\textheight]{../figures/toy-example-spi.png}
+{\centering \includegraphics[width=0.7\textwidth,height=0.7\textheight]{../figures/toy-example-spi.png}
 
@@
 
-\caption{\label{fig-toy-example}this is the caption}
+\caption{\label{fig-toy-example}Standardised Precipitation Index (SPI)
+calculated using two approaches: the pipeline construction (y-axis) and
+the function \texttt{spi()} from the \texttt{SPEI} package (x-axis),
+using the \texttt{wichita} data from the package \texttt{SPEI}. All the
+points align diagnally, indicating the same numerical result from both
+approaches.}
 
 \end{figure}
-
-While the calculated index value from the pipeline approach is the same
-as those calculated using a single line command like
-\texttt{spi(DATA,\ ARGUMENTS)}, as shown in
-Figure~\ref{fig-toy-example}, the pipeline approach gives more
-flexibility to perform diagnostics on the intermediate steps and to
-extend on the initial index calculation. For example, model goodness of
-fit test can be checked after \texttt{normalising()} through:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{normalising}\NormalTok{(}\AttributeTok{dist =} \StringTok{"gamma"}\NormalTok{, }\AttributeTok{col =}\NormalTok{ d, }\AttributeTok{fit\_method =} \StringTok{"lmom"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{gof\_test}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-and bootstrap uncertainty on the index, with 100 boostrap samples, can
-be calculated using \texttt{n\_boot\ =\ 100} in the
-\texttt{normalising()} as:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{normalising}\NormalTok{(}\AttributeTok{dist =} \StringTok{"gamma"}\NormalTok{, }\AttributeTok{col =}\NormalTok{ d, }\AttributeTok{fit\_method =} \StringTok{"lmom"}\NormalTok{, }\AttributeTok{n\_boot =} \DecValTok{100}\NormalTok{) }
-\end{Highlighting}
-\end{Shaded}
 

### 2022-11-03 paper/paper.qmd
change the small example to compare between two distributions

@@
 ---
-
-\newpage
 
@@
 
-*first sentence: tailor to their indices without thinking about the big picture*
-
-~~While every jumbled index jumbles in its own way, [perspicuous] clearly-laied out indices are all alike.~~
 This paper proposes a data pipeline for index construction. 
@@
 
-\newpage
+# Data pipeline {#sec-data-pipeline-in-r}
 
-# Data pipeline in R {#sec-data-pipeline-in-r}
+*Think about if there is another word for data pipeline*
 
@@
 
-Consider a commonly used drought index: Standardised Precipitation Index (SPI) [@spi].
-Its construction involves: 1) aggregating monthly precipitation (`prcp`) with a time scale, 2) fitting the aggregated series with a chosen distribution to get density values, and 3) converting the density values to normal quantiles.
-Under the pipeline approach, SPI will be constructed as:
+Consider a commonly used drought index: Standardized Precipitation-Evapotranspiration Index (SPEI)  [@spei].
+Its construction involves: 1) calculating potential evapotranspiration (PET) from average temperature, `tavg`, and its difference, `d`, with precipitation, `prcp`, 2) aggregating difference series with a time scale, 3) fitting the aggregated series with a chosen distribution to get density values, and 4) converting the density values to normal quantiles.
+Under the pipeline approach, SPEI will be constructed as:
 
@@
 DATA %>% 
-  aggregate(scale = 12, var = prcp) %>% 
-  normalise(dist = gamma, var = prcp, fit_method = lmom) %>% 
+  trans_pet(method = thornthwaite, col = tavg) %>% 
+  aggregate(scale = 12, var = d) %>% 
+  normalise(dist = gamma, fit_method = lmom) %>% 
   augment(var = .agg_prcp)
@@
 
+*Change the examples to two scales values, compare how indices look like*
+
 Here each command corresponds to one step described above with additional arguments specific to the step. 
-The pipeline construction produces the same index value as command like `spi(...)`, as shown in @fig-toy-example, but with additional benefit:
+The pipeline construction produces the same index value as command like `spei(...)`, as shown in @fig-toy-example, but with additional benefit:
 
@@
 
-```{r eval = FALSE, echo = FALSE}
-res <- w %>% 
-  ts_aggregate(scale = 12, date = ym, id = id, col = prcp) %>% 
-  normalising(dist = "gamma", date = ym, id = id, col = prcp,
-              method = "lmom") %>% 
-  augment(id = id, date = ym,  col = "prcp_agg")
+  *everyone is providing single command function, this is also the same for spi*
 
-res <- res %>% 
-  rename(
-         pipeline_d = prcp_agg,
-         pipeline_idx = .index)
+  *may not worthwhile if only one index, as long as start changing*
 
-data(wichita)
-wichita <- ts(wichita[,-c(1,2)], end=c(2011,10), frequency=12)
-spi12 <- spi(wichita[,'PRCP'], 12)
-from_SPI <- tibble(
-  ym = yearmonth(zoo::yearmon(zoo::index(wichita))), 
-  SPI_idx = as.vector(spi12$fitted)
-)
-
-compare <- res %>% left_join(from_SPI, by = "ym") 
-compare %>% 
-  ggplot(aes(x = SPI_idx, y = pipeline_idx)) + 
-  geom_abline(intercept = 0, slope = 1, linetype = "dashed") + 
-  geom_point() + 
-  theme_bw() + 
-  theme(legend.position = "bottom", 
-        aspect.ratio = 1) +
-  xlab("Index values from `spi()` in package `SPEI`") + 
-  ylab("Index values from pipeline") + 
-  theme(panel.grid.minor = element_blank())
-```
+  *just want to calculate index, single function is fine*
 
@@
 #| fig-align: center
-#| fig-cap: "Standardised Precipitation Index (SPI) calculated using two approaches: the pipeline construction (y-axis) and the function `spi()` from the `SPEI` package (x-axis), using the `wichita` data from the package `SPEI`.
-All the points align diagnally, indicating the same numerical result from both approaches."
-#| out-height: 70%
-#| out-width: 70%
-knitr::include_graphics(here::here("figures/toy-example-spi.png"))
+#| fig-cap: "Standardised Precipitation Evapotranspiration Index (SPEI) calculated with two distribution fits in Step 3 described above: generalised logistic (`glogist`) and generalised extreme value (`gev`) distribution, using the `wichita` data from the package `SPEI`."
+#| out-height: 30%
+#| out-width: 100%
+knitr::include_graphics(here::here("figures/toy-example-spei.png"))
 ```
 
+\newpage
+
 ## Pipeline steps for constructing indies
-
-*put a diagram here to summarise the pipeline*
 
@@
 
-each step as a point rather than subsection
+```{r}
+#| fig-align: center
+#| fig-cap: "Diagram of pipeline steps for index construction."
+#| out-height: 90%
+#| out-width: 90%
+knitr::include_graphics(here::here("figures/pipeline-steps.png"))
+```
 

### 2022-11-03 paper/paper.tex
change the small example to compare between two distributions

@@
 \usepackage{xcolor}
-\usepackage[normalem]{ulem}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
-
-\newpage
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-\emph{first sentence: tailor to their indices without thinking about the
-big picture}
-
-\sout{While every jumbled index jumbles in its own way,
-{[}perspicuous{]} clearly-laied out indices are all alike.} This paper
-proposes a data pipeline for index construction. By recognising the
-common steps shared by many indices, we develop a pipeline that breaks
-down index construction into multiple modules and allow operations in
-various modules to be combined like building blocks to construct
-indices. The pipeline approach is general while adaptable to most index
-construction. It allows indices to be created, studied, and compared in
-a structured tidy form and enables statistical analysis of indices to be
-performed easily: More specifically, it enables researchers to 1)
-validate the indices calculated from external organisations, 2) unify
-various indices under the same framework for computing, 3) swap or
-adjust individual steps in the index construction to study their
-contribution, 4) calculate uncertainty on indices through bootstrap or
-others, 5) enhance existing indices through comparing and studying their
-statistical properties, and finally, 6) propose new indices from
-combining different steps in existing indices.
+This paper proposes a data pipeline for index construction. By
+recognising the common steps shared by many indices, we develop a
+pipeline that breaks down index construction into multiple modules and
+allow operations in various modules to be combined like building blocks
+to construct indices. The pipeline approach is general while adaptable
+to most index construction. It allows indices to be created, studied,
+and compared in a structured tidy form and enables statistical analysis
+of indices to be performed easily: More specifically, it enables
+researchers to 1) validate the indices calculated from external
+organisations, 2) unify various indices under the same framework for
+computing, 3) swap or adjust individual steps in the index construction
+to study their contribution, 4) calculate uncertainty on indices through
+bootstrap or others, 5) enhance existing indices through comparing and
+studying their statistical properties, and finally, 6) propose new
+indices from combining different steps in existing indices.
 
@@
 
-\newpage
+\hypertarget{sec-data-pipeline-in-r}{%
+\section{Data pipeline}\label{sec-data-pipeline-in-r}}
 
-\hypertarget{sec-data-pipeline-in-r}{%
-\section{Data pipeline in R}\label{sec-data-pipeline-in-r}}
+\emph{Think about if there is another word for data pipeline}
 
@@
 
-Consider a commonly used drought index: Standardised Precipitation Index
-(SPI) (McKee, Doesken, and Kleist 1993). Its construction involves: 1)
-aggregating monthly precipitation (\texttt{prcp}) with a time scale, 2)
+Consider a commonly used drought index: Standardized
+Precipitation-Evapotranspiration Index (SPEI) (Vicente-Serrano,
+Beguería, and López-Moreno 2010). Its construction involves: 1)
+calculating potential evapotranspiration (PET) from average temperature,
+\texttt{tavg}, and its difference, \texttt{d}, with precipitation,
+\texttt{prcp}, 2) aggregating difference series with a time scale, 3)
 fitting the aggregated series with a chosen distribution to get density
-values, and 3) converting the density values to normal quantiles. Under
-the pipeline approach, SPI will be constructed as:
+values, and 4) converting the density values to normal quantiles. Under
+the pipeline approach, SPEI will be constructed as:
 
@@
 \NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{var =}\NormalTok{ prcp) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{normalise}\NormalTok{(}\AttributeTok{dist =}\NormalTok{ gamma, }\AttributeTok{var =}\NormalTok{ prcp, }\AttributeTok{fit\_method =}\NormalTok{ lmom) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{trans\_pet}\NormalTok{(}\AttributeTok{method =}\NormalTok{ thornthwaite, }\AttributeTok{col =}\NormalTok{ tavg) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{var =}\NormalTok{ d) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{normalise}\NormalTok{(}\AttributeTok{dist =}\NormalTok{ gamma, }\AttributeTok{fit\_method =}\NormalTok{ lmom) }\SpecialCharTok{\%\textgreater{}\%} 
   \FunctionTok{augment}\NormalTok{(}\AttributeTok{var =}\NormalTok{ .agg\_prcp)}
@@
 
+\emph{Change the examples to two scales values, compare how indices look
+like}
+
 Here each command corresponds to one step described above with
 additional arguments specific to the step. The pipeline construction
-produces the same index value as command like \texttt{spi(...)}, as
+produces the same index value as command like \texttt{spei(...)}, as
 shown in Figure~\ref{fig-toy-example}, but with additional benefit:
@@
 
+\emph{everyone is providing single command function, this is also the
+same for spi}
+
+\emph{may not worthwhile if only one index, as long as start changing}
+
+\emph{just want to calculate index, single function is fine}
+
 \begin{figure}
 
-{\centering \includegraphics[width=0.7\textwidth,height=0.7\textheight]{../figures/toy-example-spi.png}
+{\centering \includegraphics[width=1\textwidth,height=0.3\textheight]{../figures/toy-example-spei.png}
 
@@
 
-\caption{\label{fig-toy-example}Standardised Precipitation Index (SPI)
-calculated using two approaches: the pipeline construction (y-axis) and
-the function \texttt{spi()} from the \texttt{SPEI} package (x-axis),
-using the \texttt{wichita} data from the package \texttt{SPEI}. All the
-points align diagnally, indicating the same numerical result from both
-approaches.}
+\caption{\label{fig-toy-example}Standardised Precipitation
+Evapotranspiration Index (SPEI) calculated with two distribution fits in
+Step 3 described above: generalised logistic (\texttt{glogist}) and
+generalised extreme value (\texttt{gev}) distribution , using the
+\texttt{wichita} data from the package \texttt{SPEI}.}
 
 \end{figure}
+
+\newpage
 
@@
 indies}\label{pipeline-steps-for-constructing-indies}}
-
-\emph{put a diagram here to summarise the pipeline}
 
@@
 
-each step as a point rather than subsection
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\textwidth,height=0.9\textheight]{../figures/pipeline-steps.png}
+
+}
+
+\caption{Diagram of pipeline steps for index construction.}
+
+\end{figure}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-spi}{}}%
-McKee, Thomas B, Nolan J Doesken, and John Kleist. 1993. {``The
-Relationship of Drought Frequency and Duration to Time Scales.''} In
-\emph{Proceedings of the 8th Conference of Applied Climatology},
-179--84. Anaheim, CA: American Meteorological Society.
-
 \leavevmode\vadjust pre{\hypertarget{ref-sutherland_orca_2000}{}}%
@@
 2021-22.''} New York. \url{http://report.hdr.undp.org}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%
+Vicente-Serrano, Sergio M., Santiago Beguería, and Juan I. López-Moreno.
+2010. {``A {Multiscalar} {Drought} {Index} {Sensitive} to {Global}
+{Warming}: {The} {Standardized} {Precipitation} {Evapotranspiration}
+{Index}.''} \emph{Journal of Climate} 23 (7): 1696--1718.
+\url{https://journals.ametsoc.org/view/journals/clim/23/7/2009jcli2909.1.xml}.
 

### 2022-11-10 paper/paper.qmd
update pipeline diagram

@@
 DATA %>% 
-  trans_pet(method = thornthwaite, col = tavg) %>% 
+  trans_pet(method = thornthwaite, var = tavg) %>% 
   aggregate(scale = 12, var = d) %>% 
@@
 #| out-height: 90%
-#| out-width: 90%
+#| out-width: 100%
 knitr::include_graphics(here::here("figures/pipeline-steps.png"))

### 2022-11-10 paper/paper.tex
update pipeline diagram

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{trans\_pet}\NormalTok{(}\AttributeTok{method =}\NormalTok{ thornthwaite, }\AttributeTok{col =}\NormalTok{ tavg) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{trans\_pet}\NormalTok{(}\AttributeTok{method =}\NormalTok{ thornthwaite, }\AttributeTok{var =}\NormalTok{ tavg) }\SpecialCharTok{\%\textgreater{}\%} 
   \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{var =}\NormalTok{ d) }\SpecialCharTok{\%\textgreater{}\%} 
@@
 Step 3 described above: generalised logistic (\texttt{glogist}) and
-generalised extreme value (\texttt{gev}) distribution , using the
+generalised extreme value (\texttt{gev}) distribution, using the
 \texttt{wichita} data from the package \texttt{SPEI}.}
@@
 
-{\centering \includegraphics[width=0.9\textwidth,height=0.9\textheight]{../figures/pipeline-steps.png}
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{../figures/pipeline-steps.png}
 

### 2022-11-21 paper/paper.qmd
indices -> indexes

@@
 abstract: |
-  - Indices, useful, quantify severity, early monitoring, 
-  - A huge number of indices have been proposed by domain experts, however, a large majority of them are not being adopted, reused, and compared in research or in practice. 
-  - One of the reasons for this is the plenty of indices are quite complex and there is no obvious easy-to-use implementation to apply them to user's data. 
-  - The paper describes a general pipeline framework to construct indices from spatio-temporal data, 
-  - This allows all the indices to be constructed through a uniform data pipeline and different indices to vary on the details of each step in the data pipeline and their orders. 
-  - The pipeline proposed aim to smooth the workflow of index construction through breaking down the complicated steps proposed by various indices into small building blocks shared by most of the indices.
-  - The framework will be demonstrated with drought indices as examples, but appliable in general to environmental indices constructed from multivariate spatio-temporal data
+  - indexes, useful, quantify severity, early monitoring, 
+  - A huge number of indexes have been proposed by domain experts, however, a large majority of them are not being adopted, reused, and compared in research or in practice. 
+  - One of the reasons for this is the plenty of indexes are quite complex and there is no obvious easy-to-use implementation to apply them to user's data. 
+  - The paper describes a general pipeline framework to construct indexes from spatio-temporal data, 
+  - This allows all the indexes to be constructed through a uniform data pipeline and different indexes to vary on the details of each step in the data pipeline and their orders. 
+  - The pipeline proposed aim to smooth the workflow of index construction through breaking down the complicated steps proposed by various indexes into small building blocks shared by most of the indexes.
+  - The framework will be demonstrated with drought indexes as examples, but appliable in general to environmental indexes constructed from multivariate spatio-temporal data
 keywords:
-  - indices
+  - indexes
   - data pipeline
@@
 
-<!-- However, even if most of the indices have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember, classify, and compare between.
+<!-- However, even if most of the indexes have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember, classify, and compare between.
 -->
 
-**Why index is useful, why people care about indices**
+**Why index is useful, why people care about indexes**
 
@@
 The concept of interest doesn't have a direct measure can because it is impractical to measure at the population level. 
-For example, it would be nearly impossible to include all the available stocks in the market to characterise stock market behavior, so indices like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite select a representative set of stocks to measure the overall market behavior.
-Also belonging to this category are the economic indices like the Consumer Price Index, where price changes of a basket of items are weighted to measure inflation. 
+For example, it would be nearly impossible to include all the available stocks in the market to characterise stock market behavior, so indexes like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite select a representative set of stocks to measure the overall market behavior.
+Also belonging to this category are the economic indexes like the Consumer Price Index, where price changes of a basket of items are weighted to measure inflation. 
 The lack of direct measure could also because the concept itself is an unobservable human construction, rather than a physical quantity that can be measured.
 Many natural hazard and social concepts falls into this category. 
-This includes drought indices constructed from meteorological, agricultural, hydrological, and social-economic variables, e.g.
+This includes drought indexes constructed from meteorological, agricultural, hydrological, and social-economic variables, e.g.
 Standardised Precipitation Index (SPI) [@mckee1993relationship] and Aggregated Drought Index (ADI) [@keyantash_aggregate_2004] among others.
-Social development indices like Human Development Index [@hdi] and Global Liveability Index [@gli] measure various aspects of the quality of human capital and urban life.
+Social development indexes like Human Development Index [@hdi] and Global Liveability Index [@gli] measure various aspects of the quality of human capital and urban life.
 
@@
 
-Despite many quantity having the term *index* in their name, they cannot be technically classified as indices according to the definition given above. 
+Despite many quantity having the term *index* in their name, they cannot be technically classified as indexes according to the definition given above. 
 The reason for these quantities to lose their index memberships is that they are variables can be accurately measured given the instrument precision. 
@@
 They are measures of precipitation and percentage of days under specific conditions (dries month, maximum temperature below 10th percentile).
-They are variables, or indicators, that can be used to construct indices but are not indices themselves. 
-Similarly, a set of remote sensing indices are not indices, since they are measures of electromagnetic wave reflectance.
+They are variables, or indicators, that can be used to construct indexes but are not indexes themselves. 
+Similarly, a set of remote sensing indexes are not indexes, since they are measures of electromagnetic wave reflectance.
 This includes Normalized Difference Vegetation Index (NDVI) [@tucker_red_1979], derived from the ratio of difference over sum on two segments in the spectrum, also called band: near-infrared (NIR) and red.
-So are the "indices" derived from NDVI, e.g. Vegetation Condition Index [@kogan_application_1995]. 
-Notice that this does not exclude all the construction derived from remotes sensor variables to be valid indices.
+So are the "indexes" derived from NDVI, e.g. Vegetation Condition Index [@kogan_application_1995]. 
+Notice that this does not exclude all the construction derived from remotes sensor variables to be valid indexes.
 For example, Vegetation Drought Response Index [@brown_vegetation_2008] is a valid index since it integrates climate, satellite, and biophysical variables to quantify vegetation stress.
@@
 <!--   - A measure is a variable, but is not an index.  -->
-<!--   - Not all multivariate quantities are indices (i.e. those remote sensor measures). It is possible that an index is constructed from a single variable (i.e. SPI). -->
-<!--   - Indices is not a single/ set of measures -->
+<!--   - Not all multivariate quantities are indexes (i.e. those remote sensor measures). It is possible that an index is constructed from a single variable (i.e. SPI). -->
+<!--   - indexes is not a single/ set of measures -->
 
@@
 Currently, index construction lacks a standardised workflow. 
-It is often up to researchers or research institutions to decide whether to provide open source code on the new indices, what would be the best user interface for other researchers to use the new indices, and how easily the new indices can be compared with other existing indices.
-This makes the computation lack transparency and indices cumbersome to experiment with:
+It is often up to researchers or research institutions to decide whether to provide open source code on the new indexes, what would be the best user interface for other researchers to use the new indexes, and how easily the new indexes can be compared with other existing indexes.
+This makes the computation lack transparency and indexes cumbersome to experiment with:
 
-  - Researchers who wish to validate the indices calculated from large institutes need to reinvent the wheels themselves since the source code used for computing is often not available for public consumption;
+  - Researchers who wish to validate the indexes calculated from large institutes need to reinvent the wheels themselves since the source code used for computing is often not available for public consumption;
   - Open-source code provided by research groups has a narrow margin for exploring other options outside the provided;
-  - Similar steps used by different indices are difficult to spot since the design of the user interface for indices often includes all the steps under a single function call; and
+  - Similar steps used by different indexes are difficult to spot since the design of the user interface for indexes often includes all the steps under a single function call; and
   - It is generally hard to inspect intermediate results during the index construction if users wish to check the output of a certain step.
 
-<!-- Institutes like National Oceanic and Atmospheric Administration (NOAA) and United States Drought Monitor (USDM) moderate indices in their domains. -->
+<!-- Institutes like National Oceanic and Atmospheric Administration (NOAA) and United States Drought Monitor (USDM) moderate indexes in their domains. -->
 <!-- They provide public information to describe the methodology used in the calculation, publish the index values for public consumption while the source code used for computing is often not available for public consumption.
@@
 
-<!-- For example, many indices involve a step to fit variables into a particular distribution. -->
+<!-- For example, many indexes involve a step to fit variables into a particular distribution. -->
 <!-- Index researchers may provide options for different fits, but it is generally hard to experiment with options outside the provided distributions. -->
 
-<!-- For example, many indices needs standardisation in its workflow, it can be used to rescale the input variables or the index series. -->
+<!-- For example, many indexes needs standardisation in its workflow, it can be used to rescale the input variables or the index series. -->
 <!-- But the current construction does not make it clear through its user interface but sweep it, along with all the other steps in the index construction, under a single function call.  -->
@@
 This paper proposes a data pipeline for index construction. 
-By recognising the common steps shared by many indices, we develop a pipeline that breaks down index construction into multiple modules and allow operations in various modules to be combined like building blocks to construct indices.
+By recognising the common steps shared by many indexes, we develop a pipeline that breaks down index construction into multiple modules and allow operations in various modules to be combined like building blocks to construct indexes.
 The pipeline approach is general while adaptable to most index construction. 
-It allows indices to be created, studied, and compared in a structured tidy form and enables statistical analysis of indices to be performed easily: 
-More specifically, it enables researchers to 1) validate the indices calculated from external organisations, 2) unify various indices under the same framework for computing, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty on indices through bootstrap or others, 5) enhance existing indices through comparing and studying their statistical properties, and finally, 6) propose new indices from combining different steps in existing indices.
+It allows indexes to be created, studied, and compared in a structured tidy form and enables statistical analysis of indexes to be performed easily: 
+More specifically, it enables researchers to 1) validate the indexes calculated from external organisations, 2) unify various indexes under the same framework for computing, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty on indexes through bootstrap or others, 5) enhance existing indexes through comparing and studying their statistical properties, and finally, 6) propose new indexes from combining different steps in existing indexes.
 
@@
 
-This work is of interest to researchers actively developing new indices since it encourages new indices to be delivered in an easy-to-reproduce design. 
-It would also provide analysts who wish to compute a range of indices in their analysis a uniform interface to build relevant indices from raw data. 
+This work is of interest to researchers actively developing new indexes since it encourages new indexes to be delivered in an easy-to-reproduce design. 
+It would also provide analysts who wish to compute a range of indexes in their analysis a uniform interface to build relevant indexes from raw data. 
 For statisticians and software developing engineers, this work frames the process of index construction in a more user-oriented workflow and could motivate similar research for other process in scientific computing.
@@
 The rest of the paper is structured as follows: @sec-data-pipeline-in-r reviews the concept of data pipeline in R. 
-The pipeline framework for index construction is presented in @sec-a-pipeline-for-building-statistical-indices.
+The pipeline framework for index construction is presented in @sec-a-pipeline-for-building-statistical-indexes.
 @sec-incorporating-new-buliding-blocks-into-the-pipeline explains how to include a new building block in each pipeline module.
@@
 
-**Constructing indices would also benefit from pipeline and embracing the aforementioned design philosophy.**
+**Constructing indexes would also benefit from pipeline and embracing the aforementioned design philosophy.**
 
-In index construction, data pipeline is often presented in a workflow diagram in the research paper to illustrate how the raw data is transformed into the final indices. 
+In index construction, data pipeline is often presented in a workflow diagram in the research paper to illustrate how the raw data is transformed into the final indexes. 
 This agrees with Wickham's argument on the presence of the data pipeline, however, more often than not, the pipeline is not made explicit in the software.
 Often the time, all the steps are lumped into a single wrapper function, rather than being split into smaller, modulated functions. 
-This increases the cost of maintaining and understanding the code base, gives analysts little freedom to customise the indices for specific needs, and hinders reusing existing code for building new indices.
-A pipeline approach unites a range of indices under a single data pipeline and analysts can compose indices from pipeline steps like building Legos from individual bricks.
-In this workflow, analysts are not limited by indices that have been already proposed and can easily combine pipeline steps to compose novel indices.
-Analysis of the indices (i.e.
+This increases the cost of maintaining and understanding the code base, gives analysts little freedom to customise the indexes for specific needs, and hinders reusing existing code for building new indexes.
+A pipeline approach unites a range of indexes under a single data pipeline and analysts can compose indexes from pipeline steps like building Legos from individual bricks.
+In this workflow, analysts are not limited by indexes that have been already proposed and can easily combine pipeline steps to compose novel indexes.
+Analysis of the indexes (i.e.
 calculation of uncertainty) is also feasible by adding external code into the pipeline.
@@
 
-# A pipeline for building statistical indices {#sec-a-pipeline-for-building-statistical-indices}
+# A pipeline for building statistical indexes {#sec-a-pipeline-for-building-statistical-indexes}
 
@@
 
-*Change the examples to two scales values, compare how indices look like*
+*Change the examples to two scales values, compare how indexes look like*
 
@@
 
-The construction of natural hazard indices also follows a set of steps, which is usually illustrated using a flowchart in the paper. 
-However, every researcher follows a certain design philosophy and steps taken in the index constructed by different researchers are not aligned. 
-This discourages experiment with multiple indices. Initiate a new workflow when computing a new index.
+The construction of natural hazard indexes also follows a set of steps, which is usually illustrated using a flowchart in the paper.
 
-The most popular indices (i.e. SPI, SPEI, etc) have existing software implementation (`SPEI`) to be applied to a different set of data.
+* constructing time series index should also be encapsulated in the framework
 
-constructing time series index should also be encapsulated in my framework
+Here we assume a concept of interest is determined, relevant variables/ indicators are identified and available to construct indexes.
 
-Here we assume a concept of interest is determined, relevant variables/ indicators are identified and available to construct indices.
+Before the mathmetical construction of indexes
 
@@
 sometimes called feature extraction in the machine learning community 
-With drought indices, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.]
+With drought indexes, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.]
 

### 2022-11-21 paper/paper.tex
indices -> indexes

@@
   pdfauthor={H. Sherry Zhang; Collaborators},
-  pdfkeywords={indices, data pipeline, software design},
+  pdfkeywords={indexes, data pipeline, software design},
   colorlinks=true,
@@
 \item
-  Indices, useful, quantify severity, early monitoring,
+  indexes, useful, quantify severity, early monitoring,
 \item
-  A huge number of indices have been proposed by domain experts,
+  A huge number of indexes have been proposed by domain experts,
   however, a large majority of them are not being adopted, reused, and
@@
 \item
-  One of the reasons for this is the plenty of indices are quite complex
+  One of the reasons for this is the plenty of indexes are quite complex
   and there is no obvious easy-to-use implementation to apply them to
@@
 \item
-  The paper describes a general pipeline framework to construct indices
+  The paper describes a general pipeline framework to construct indexes
   from spatio-temporal data,
 \item
-  This allows all the indices to be constructed through a uniform data
-  pipeline and different indices to vary on the details of each step in
+  This allows all the indexes to be constructed through a uniform data
+  pipeline and different indexes to vary on the details of each step in
   the data pipeline and their orders.
@@
   through breaking down the complicated steps proposed by various
-  indices into small building blocks shared by most of the indices.
+  indexes into small building blocks shared by most of the indexes.
 \item
-  The framework will be demonstrated with drought indices as examples,
-  but appliable in general to environmental indices constructed from
+  The framework will be demonstrated with drought indexes as examples,
+  but appliable in general to environmental indexes constructed from
   multivariate spatio-temporal data
@@
 \def\sep{\textbullet\ }
-indices \sep data pipeline \sep 
+indexes \sep data pipeline \sep 
 software design
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden, enhanced, boxrule=0pt, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\textbf{Why index is useful, why people care about indices}
+\textbf{Why index is useful, why people care about indexes}
 
@@
 include all the available stocks in the market to characterise stock
-market behavior, so indices like Dow Jones Industrial Average, S\&P 500,
+market behavior, so indexes like Dow Jones Industrial Average, S\&P 500,
 and Nasdaq Composite select a representative set of stocks to measure
 the overall market behavior. Also belonging to this category are the
-economic indices like the Consumer Price Index, where price changes of a
+economic indexes like the Consumer Price Index, where price changes of a
 basket of items are weighted to measure inflation. The lack of direct
@@
 natural hazard and social concepts falls into this category. This
-includes drought indices constructed from meteorological, agricultural,
+includes drought indexes constructed from meteorological, agricultural,
 hydrological, and social-economic variables, e.g.~Standardised
@@
 Index (ADI) (Keyantash and Dracup 2004) among others. Social development
-indices like Human Development Index (United Nations Development
+indexes like Human Development Index (United Nations Development
 Programme 2022) and Global Liveability Index (Economist Intelligence
@@
 Despite many quantity having the term \emph{index} in their name, they
-cannot be technically classified as indices according to the definition
+cannot be technically classified as indexes according to the definition
 given above. The reason for these quantities to lose their index
@@
 10th percentile). They are variables, or indicators, that can be used to
-construct indices but are not indices themselves. Similarly, a set of
-remote sensing indices are not indices, since they are measures of
+construct indexes but are not indexes themselves. Similarly, a set of
+remote sensing indexes are not indexes, since they are measures of
 electromagnetic wave reflectance. This includes Normalized Difference
@@
 difference over sum on two segments in the spectrum, also called band:
-near-infrared (NIR) and red. So are the ``indices'' derived from NDVI,
+near-infrared (NIR) and red. So are the ``indexes'' derived from NDVI,
 e.g.~Vegetation Condition Index (Kogan 1995). Notice that this does not
 exclude all the construction derived from remotes sensor variables to be
-valid indices. For example, Vegetation Drought Response Index (Brown et
+valid indexes. For example, Vegetation Drought Response Index (Brown et
 al. 2008) is a valid index since it integrates climate, satellite, and
@@
 up to researchers or research institutions to decide whether to provide
-open source code on the new indices, what would be the best user
-interface for other researchers to use the new indices, and how easily
-the new indices can be compared with other existing indices. This makes
-the computation lack transparency and indices cumbersome to experiment
+open source code on the new indexes, what would be the best user
+interface for other researchers to use the new indexes, and how easily
+the new indexes can be compared with other existing indexes. This makes
+the computation lack transparency and indexes cumbersome to experiment
 with:
@@
 \item
-  Researchers who wish to validate the indices calculated from large
+  Researchers who wish to validate the indexes calculated from large
   institutes need to reinvent the wheels themselves since the source
@@
 \item
-  Similar steps used by different indices are difficult to spot since
-  the design of the user interface for indices often includes all the
+  Similar steps used by different indexes are difficult to spot since
+  the design of the user interface for indexes often includes all the
   steps under a single function call; and
@@
 This paper proposes a data pipeline for index construction. By
-recognising the common steps shared by many indices, we develop a
+recognising the common steps shared by many indexes, we develop a
 pipeline that breaks down index construction into multiple modules and
 allow operations in various modules to be combined like building blocks
-to construct indices. The pipeline approach is general while adaptable
-to most index construction. It allows indices to be created, studied,
+to construct indexes. The pipeline approach is general while adaptable
+to most index construction. It allows indexes to be created, studied,
 and compared in a structured tidy form and enables statistical analysis
-of indices to be performed easily: More specifically, it enables
-researchers to 1) validate the indices calculated from external
-organisations, 2) unify various indices under the same framework for
+of indexes to be performed easily: More specifically, it enables
+researchers to 1) validate the indexes calculated from external
+organisations, 2) unify various indexes under the same framework for
 computing, 3) swap or adjust individual steps in the index construction
-to study their contribution, 4) calculate uncertainty on indices through
-bootstrap or others, 5) enhance existing indices through comparing and
+to study their contribution, 4) calculate uncertainty on indexes through
+bootstrap or others, 5) enhance existing indexes through comparing and
 studying their statistical properties, and finally, 6) propose new
-indices from combining different steps in existing indices.
+indexes from combining different steps in existing indexes.
 
@@
 
-This work is of interest to researchers actively developing new indices
-since it encourages new indices to be delivered in an easy-to-reproduce
+This work is of interest to researchers actively developing new indexes
+since it encourages new indexes to be delivered in an easy-to-reproduce
 design. It would also provide analysts who wish to compute a range of
-indices in their analysis a uniform interface to build relevant indices
+indexes in their analysis a uniform interface to build relevant indexes
 from raw data. For statisticians and software developing engineers, this
@@
 presented in
-Section~\ref{sec-a-pipeline-for-building-statistical-indices}.
+Section~\ref{sec-a-pipeline-for-building-statistical-indexes}.
 Section~\ref{sec-incorporating-new-buliding-blocks-into-the-pipeline}
@@
 
-\textbf{Constructing indices would also benefit from pipeline and
+\textbf{Constructing indexes would also benefit from pipeline and
 embracing the aforementioned design philosophy.}
@@
 diagram in the research paper to illustrate how the raw data is
-transformed into the final indices. This agrees with Wickham's argument
+transformed into the final indexes. This agrees with Wickham's argument
 on the presence of the data pipeline, however, more often than not, the
@@
 maintaining and understanding the code base, gives analysts little
-freedom to customise the indices for specific needs, and hinders reusing
-existing code for building new indices. A pipeline approach unites a
-range of indices under a single data pipeline and analysts can compose
-indices from pipeline steps like building Legos from individual bricks.
-In this workflow, analysts are not limited by indices that have been
+freedom to customise the indexes for specific needs, and hinders reusing
+existing code for building new indexes. A pipeline approach unites a
+range of indexes under a single data pipeline and analysts can compose
+indexes from pipeline steps like building Legos from individual bricks.
+In this workflow, analysts are not limited by indexes that have been
 already proposed and can easily combine pipeline steps to compose novel
-indices. Analysis of the indices (i.e.~calculation of uncertainty) is
+indexes. Analysis of the indexes (i.e.~calculation of uncertainty) is
 also feasible by adding external code into the pipeline.
@@
 
-\hypertarget{sec-a-pipeline-for-building-statistical-indices}{%
+\hypertarget{sec-a-pipeline-for-building-statistical-indexes}{%
 \section{A pipeline for building statistical
-indices}\label{sec-a-pipeline-for-building-statistical-indices}}
+indexes}\label{sec-a-pipeline-for-building-statistical-indexes}}
 
@@
 
-\emph{Change the examples to two scales values, compare how indices look
+\emph{Change the examples to two scales values, compare how indexes look
 like}
@@
 
-The construction of natural hazard indices also follows a set of steps,
-which is usually illustrated using a flowchart in the paper. However,
-every researcher follows a certain design philosophy and steps taken in
-the index constructed by different researchers are not aligned. This
-discourages experiment with multiple indices. Initiate a new workflow
-when computing a new index.
+The construction of natural hazard indexes also follows a set of steps,
+which is usually illustrated using a flowchart in the paper.
 
-The most popular indices (i.e.~SPI, SPEI, etc) have existing software
-implementation (\texttt{SPEI}) to be applied to a different set of data.
-
-constructing time series index should also be encapsulated in my
+\begin{itemize}
+\tightlist
+\item
+  constructing time series index should also be encapsulated in the
   framework
+\end{itemize}
 
 Here we assume a concept of interest is determined, relevant variables/
-indicators are identified and available to construct indices.
+indicators are identified and available to construct indexes.
+
+Before the mathmetical construction of indexes
 
@@
 sometimes called feature extraction in the machine learning community
-With drought indices, the extraction of meaningful variables from the
+With drought indexes, the extraction of meaningful variables from the
 original data is usually supported by the water balance model, for

### 2022-11-21 paper/paper.qmd
first attempt to write notation into pipeline steps

@@
 format:
-  arxiv-pdf:
+  pdf:
     keep-tex: true  
-  arxiv-html: default
+    include-in-header: 
+      text: |
+        \usepackage{amsmath}
 author:
@@
 
-\newpage
-
 # A pipeline for building statistical indexes {#sec-a-pipeline-for-building-statistical-indexes}
@@
 
-The construction of natural hazard indexes also follows a set of steps, which is usually illustrated using a flowchart in the paper.
+*constructing time series index should also be encapsulated in the framework*
 
-* constructing time series index should also be encapsulated in the framework
+@fig-pipeline-steps presents the index construction pipeline from the raw data to the final indexes. 
+The available intermediate steps include temporal processing, spatial aggregation, variable transformation, scaling, dimension reduction, normalising, benchmarking, and simplification. 
+These steps do not have to be arranged in sequence and not all the steps are required to present in every index construction. 
+Spatial aggregation and temporal processing are operations of the data in space and time 
+Variable transformation, scaling, and dimension reduction concerns operations on the multivariate aspect of the data.
+Normalisation, benchmarking, and simplification can be viewed as the "post-processing" of indexes, which addresses the cross-comparison across different indexes and time, and index communication.
 
-Here we assume a concept of interest is determined, relevant variables/ indicators are identified and available to construct indexes.
+More steps are involved before the mathematical construction of indexes. 
+This includes 1) defining a concept of interest, being useful for decision making and 2) finding the relevant and measurable variables to characterise the concept. 
+These choices define the scope of the indexes and the data availability for constructing an index but not the construction itself, hence not included in the pipeline framework. 
+The framework proposed in this paper assumes the objective of an index has been defined and relevant variables have been collected and quality controlled.
 
-Before the mathmetical construction of indexes
-
-```{r}
+```{r fig-pipeline-steps}
 #| fig-align: center
@@
 
-*Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem.
+*Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\mathbf{s}$ and $\mathbf{t}$ to depend on $p$*
+
+*Initially my thought was to raw data only after aligning the resolution: Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem.
 After processing them and align them together they become the "raw data"*
 
-The data used to construct the natural hazard index usually have three dimensions, one for location, one for time, and one for multivariate. 
-Mathematically, it can be written as $X_{j, s, t}$, where $j = 1, 2, \cdots, J$ for variable, $s = 1, 2, \cdots, S$ for location, and $t = 1, 2, \cdots, T$ for time.
+Once variables are collected, they can have different spatial resolution and temporal granularity.
+Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data that contain the spatial, temporal, and multivariate aspect: the spatial component $\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal component $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
+Expanding on the multivariate aspect, the data can be written as  $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$
 
-The location $s$ can refer to vector points or areas characterised by longitude-latitude coordinates, or raster cells obtained from satellite images.
+<!-- The location $s$ can refer to vector points or areas characterised by longitude-latitude coordinates, or raster cells obtained from satellite images. -->
 
-The time dimension $t$ can be daily, weekly, biweekly (14-16 days), monthly, or even quarterly
+<!-- The time dimension $t$ can be daily, weekly, biweekly (14-16 days), monthly, or even quarterly -->
 
-Variables
+<!-- Variables -->
 
-This multidimensional array structure is commonly used in geospatial analysis
+<!-- This multidimensional array structure is commonly used in geospatial analysis -->
 
-Given the variety of data sources at different spatial resolution and temporal granularity, the raw data may first come in multiple pieces. 
-Sometimes, even a considerable amount of work is needed to align the spatial and temporal extent of multivariate data.
+<!-- Given the variety of data sources at different spatial resolution and temporal granularity, the raw data may first come in multiple pieces.  -->
+<!-- Sometimes, even a considerable amount of work is needed to align the spatial and temporal extent of multivariate data. -->
 
-A notation for different variables have different spatial and temporal granularity $X_{j_1, s_1, t_1}$???
+**Temporal processing**
+
+Temporal processing concerns the operations on the temporal aspect of the data.
+It can be represented by $f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))$, where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
+
+- purpose is to aggregate across time, smooth?
+
+An example of temporal processing is to aggregate the data, $x(s_i; t_j)$, into $k$-month sum.
+This can be used to reflect how the accumulated values change across time.
+In the drought index SPI, precipitation can be aggregated at different scales to reflect short-term, mid-term, and long-term drought severity.
+With $\psi = k$, the aggregated series can be written as $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)$ where $j^\prime$ is the time index after the aggregation.
 
@@
 
-mostly happen with raster data
+Sometimes, it can be beneficial to borrow information from nearby spatial locations or there is the need to combine raster data from satellite images with vector ground measures.
+Spatial aggregation can be written as $g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))$, where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.
+
+ - $\theta$ can be a constant or a derived from the data
+
+ - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level.
+
+**Variable transformation**
+
+A variable transformation changes the shape of the variable.
+It can be written as $h_{\tau}(x(\mathbf{s};\mathbf{t}))$, where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the potential parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
+Variable transformation involve square root transformation, log transformation, and also include those non-linear transformation.
+
+<!-- Take one or multiple variables to create a new variable,  -->
+<!-- Restrict it to single variable, square root, log etc could be linearly, also non-linear -->
+
+<!-- change the shape of the variable -->
+
+<!-- <!-- A conversion of non-normal variables into (approximately) normal. wrong! -->
+
+<!-- GAM, can you do additive model pairwise/ three-way -->
 
@@
 
-A specific transformation on the scale of the data
+Scaling is a specific variable transformation on the scale of the data. 
+The purpose to treat scaling as a separate step from variable transformation is to make the step explicit. 
+Scaling is usually in the form of $[x(s_i;t_j) - \alpha]/\gamma$.
+For example, $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$ is used in a z-score standardisation to standardise the data by the its mean and standard deviation across all time and all space units and $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$ is used to standardise the variable into a unit interval.
 
-z-score standardising, min-max standardisation into [0, 1] or [0, 100], percentage change on the baseline
-close to variable transformation step
+<!-- z-score standardising, min-max standardisation into [0, 1] or [0, 100], percentage change on the baseline -->
+<!-- close to variable transformation step -->
+
+**Dimension reduction**
+
+When multiple variables are collected in space and time to construct indices, they are often summarised with dimension reduction.
+Dimension reduction $x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})$,  where $p^* = 1, 2, \cdots, P^*$,  $p = 1, 2, \cdots, P$, and $P^* < P$.
+The most commonly used dimension reduction techniquePprincipal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF), can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the variance of the data given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
+
+<!-- sometimes called feature extraction in the machine learning community  -->
+<!-- With drought indexes, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.] -->
 
@@
 
-The purpose of normalising is for cross-comparison.
+The purpose of normalising is for cross-comparison using the normal score.
 This step can get criticism from analysts for ...
 
-specifically for converting from a fitted distribution to normal score via reverse CDF function, non-parametric formula, or empirical approximation, a common step in many index: SPI, SSI, Z score.
+This step is usually in the end of the pipeline when the data have been processed into a univariate series. 
+When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]$, where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
+Sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.
+
+- In some cases, the distibution and inverse normal can be multivariate, i.e. use copula.
+
+<!-- specifically for converting from a fitted distribution to normal score via reverse CDF function, non-parametric formula, or empirical approximation, a common step in many index: SPI, SSI, Z score.
 The purpose of normalising is to convert the index into a standardised series after all the steps for the ease of comparison.
-
-Normalising is usually the last step
-
-**Variable transformation**
-
-<!-- Take one or multiple variables to create a new variable,  -->
-Restrict it to single variable, square root, log etc could be linearly, also non-linear
-
-change the shape of the variable
-
-<!-- A conversion of non-normal variables into (approximately) normal. wrong! -->
-
-GAM, can you do additive model pairwise/ three-way
-
-**Temporal processing**
-
-**Dimension reduction**
-
-sometimes called feature extraction in the machine learning community 
-With drought indexes, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.]
-
-Also include weighting
+-->
 
@@
 
+Benchmarking sets a constant value to allow the constructed index to be compared across time.
+Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed, could be a constant or a function of the data, i.e. mean.
+
 **Simplification**
+
+In public communication, indexes are often delivered in categorical grades, along with its underlying numerical values.
+The simplification step prescribes how the continuous index values are converted into the discrete grades.
+This process can be written with the piece-wise function:
+
+$$
+\begin{cases}
+C_0 & c_1 \leq (s_i; t_j) < c_0 \\
+C_1 & c_2 \leq x(s_i; t_j) < c_1 \\
+C_2 & c_3 \leq x(s_i; t_j) < c_2 \\
+\cdots \\
+C_z & c_z \leq x(s_i; t_j)
+\end{cases}
+$$
+
+where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the threshold value in each category.
+In SPI, drought are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$.
+In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.
 

### 2022-11-21 paper/paper.tex
first attempt to write notation into pipeline steps

@@
 \documentclass[
-]{article}
+  letterpaper,
+  DIV=11,
+  numbers=noendperiod]{scrartcl}
 
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-  \setmathfont[]{Latin Modern Math}
 \fi
@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
@@
 
-\usepackage{arxiv}
-\usepackage{orcidlink}
 \usepackage{amsmath}
-\usepackage[T1]{fontenc}
+\KOMAoption{captions}{tableheading}
 \makeatletter
@@
 \title{Demo arXiv template}
-\author{
-\textbf{H. Sherry Zhang}~\orcidlink{0000-0002-7122-1463}\\Department of
-Econometrics and Business Statistics\\Monash University\\Melbourne,
-VIC\\\href{mailto:huize.zhang@monash.edu}{huize.zhang@monash.edu}\\\\\\
-\textbf{Collaborators}\\Department of Econometrics and Business
-Statistics\\Monash University\\Melbourne, VIC\\}
+\author{H. Sherry Zhang \and Collaborators}
 \date{}
+
 \begin{document}
@@
 \end{abstract}
-{\bfseries \emph Keywords}
-\def\sep{\textbullet\ }
-indexes \sep data pipeline \sep 
-software design
-
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden, enhanced, boxrule=0pt, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\newpage
-
 \hypertarget{sec-a-pipeline-for-building-statistical-indexes}{%
@@
 
-The construction of natural hazard indexes also follows a set of steps,
-which is usually illustrated using a flowchart in the paper.
+\emph{constructing time series index should also be encapsulated in the
+framework}
 
-\begin{itemize}
-\tightlist
-\item
-  constructing time series index should also be encapsulated in the
-  framework
-\end{itemize}
+Figure~\ref{fig-pipeline-steps} presents the index construction pipeline
+from the raw data to the final indexes. The available intermediate steps
+include temporal processing, spatial aggregation, variable
+transformation, scaling, dimension reduction, normalising, benchmarking,
+and simplification. These steps do not have to be arranged in sequence
+and not all the steps are required to present in every index
+construction. Spatial aggregation and temporal processing are operations
+of the data in space and time Variable transformation, scaling, and
+dimension reduction concerns operations on the multivariate aspect of
+the data. Normalisation, benchmarking, and simplification can be viewed
+as the ``post-processing'' of indexes, which addresses the
+cross-comparison across different indexes and time, and index
+communication.
 
-Here we assume a concept of interest is determined, relevant variables/
-indicators are identified and available to construct indexes.
+More steps are involved before the mathematical construction of indexes.
+This includes 1) defining a concept of interest, being useful for
+decision making and 2) finding the relevant and measurable variables to
+characterise the concept. These choices define the scope of the indexes
+and the data availability for constructing an index but not the
+construction itself, hence not included in the pipeline framework. The
+framework proposed in this paper assumes the objective of an index has
+been defined and relevant variables have been collected and quality
+controlled.
 
-Before the mathmetical construction of indexes
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{include\_graphics}\NormalTok{(here}\SpecialCharTok{::}\FunctionTok{here}\NormalTok{(}\StringTok{"figures/pipeline{-}steps.png"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
 
-\begin{figure}
+\begin{figure}[H]
 
@@
 
-\caption{Diagram of pipeline steps for index construction.}
+\caption{\label{fig-pipeline-steps}Diagram of pipeline steps for index
+construction.}
 
@@
 
-\emph{Another section on original data directly downloaded, can have
-different spatial resolution, temporal granularity, data quality
+\emph{Discussion: whether we should allow raw data to have different
+spatial/ temporal resolution: This would require the notation
+\(\mathbf{s}\) and \(\mathbf{t}\) to depend on \(p\)}
+
+\emph{Initially my thought was to raw data only after aligning the
+resolution: Another section on original data directly downloaded, can
+have different spatial resolution, temporal granularity, data quality
 problem. After processing them and align them together they become the
@@
 
-The data used to construct the natural hazard index usually have three
-dimensions, one for location, one for time, and one for multivariate.
-Mathematically, it can be written as \(X_{j, s, t}\), where
-\(j = 1, 2, \cdots, J\) for variable, \(s = 1, 2, \cdots, S\) for
-location, and \(t = 1, 2, \cdots, T\) for time.
+Once variables are collected, they can have different spatial resolution
+and temporal granularity. Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\)
+denote the raw data that contain the spatial, temporal, and multivariate
+aspect: the spatial component
+\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
+space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
+temporal component \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
+defined in the 1D space:
+\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). Expanding on the
+multivariate aspect, the data can be written as
+\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\)
 
-The location \(s\) can refer to vector points or areas characterised by
-longitude-latitude coordinates, or raster cells obtained from satellite
-images.
+\textbf{Temporal processing}
 
-The time dimension \(t\) can be daily, weekly, biweekly (14-16 days),
-monthly, or even quarterly
+Temporal processing concerns the operations on the temporal aspect of
+the data. It can be represented by
+\(f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))\), where
+\(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the parameters
+associated with the temporal operation and \(d_{\psi}\) is the number of
+parameter of \(\psi\).
 
-Variables
+\begin{itemize}
+\tightlist
+\item
+  purpose is to aggregate across time, smooth?
+\end{itemize}
 
-This multidimensional array structure is commonly used in geospatial
-analysis
-
-Given the variety of data sources at different spatial resolution and
-temporal granularity, the raw data may first come in multiple pieces.
-Sometimes, even a considerable amount of work is needed to align the
-spatial and temporal extent of multivariate data.
-
-A notation for different variables have different spatial and temporal
-granularity \(X_{j_1, s_1, t_1}\)???
+An example of temporal processing is to aggregate the data,
+\(x(s_i; t_j)\), into \(k\)-month sum. This can be used to reflect how
+the accumulated values change across time. In the drought index SPI,
+precipitation can be aggregated at different scales to reflect
+short-term, mid-term, and long-term drought severity. With \(\psi = k\),
+the aggregated series can be written as
+\(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)\)
+where \(j^\prime\) is the time index after the aggregation.
 
@@
 
-mostly happen with raster data
+Sometimes, it can be beneficial to borrow information from nearby
+spatial locations or there is the need to combine raster data from
+satellite images with vector ground measures. Spatial aggregation can be
+written as \(g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))\), where
+\(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the spatial
+operation parameters and \(d_{\theta}\) is the number of parameter of
+\(\theta\).
+
+\begin{itemize}
+\item
+  \(\theta\) can be a constant or a derived from the data
+\item
+  3 by 3 aggregation, another purpose: the spatial grid is too fine,
+  aggregate to higher level.
+\end{itemize}
+
+\textbf{Variable transformation}
+
+A variable transformation changes the shape of the variable. It can be
+written as \(h_{\tau}(x(\mathbf{s};\mathbf{t}))\), where
+\(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the potential
+parameter in the transformation, if any, and \(d_{\tau}\) is the number
+of parameter of \(\tau\). Variable transformation involve square root
+transformation, log transformation, and also include those non-linear
+transformation.
 
@@
 
-A specific transformation on the scale of the data
+Scaling is a specific variable transformation on the scale of the data.
+The purpose to treat scaling as a separate step from variable
+transformation is to make the step explicit. Scaling is usually in the
+form of \([x(s_i;t_j) - \alpha]/\gamma\). For example,
+\(\alpha = \bar{x}(s; t)\) and \(\gamma = \sigma(s; t)\) is used in a
+z-score standardisation to standardise the data by the its mean and
+standard deviation across all time and all space units and
+\(\alpha = \min[x(s_i, t_j)]\) and
+\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\) is used to
+standardise the variable into a unit interval.
 
-z-score standardising, min-max standardisation into {[}0, 1{]} or {[}0,
-100{]}, percentage change on the baseline close to variable
-transformation step
+\textbf{Dimension reduction}
+
+When multiple variables are collected in space and time to construct
+indices, they are often summarised with dimension reduction. Dimension
+reduction
+\(x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})\),
+where \(p^* = 1, 2, \cdots, P^*\), \(p = 1, 2, \cdots, P\), and
+\(P^* < P\). The most commonly used dimension reduction
+techniquePprincipal Component Analysis (PCA), also called Empirical
+Orthogonal Function (EOF), can be written as
+\(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
+where \(\lambda_p\) is the loading of the PC1, derived from maximising
+the variance of the data given the constraint
+\(\sum_{p=1}^P\lambda_p^2 = 1\).
 
@@
 
-The purpose of normalising is for cross-comparison. This step can get
-criticism from analysts for \ldots{}
+The purpose of normalising is for cross-comparison using the normal
+score. This step can get criticism from analysts for \ldots{}
 
-specifically for converting from a fitted distribution to normal score
-via reverse CDF function, non-parametric formula, or empirical
-approximation, a common step in many index: SPI, SSI, Z score. The
-purpose of normalising is to convert the index into a standardised
-series after all the steps for the ease of comparison.
+This step is usually in the end of the pipeline when the data have been
+processed into a univariate series. When the data are not distributed
+normally, a distributed is usually fitted before converting to the
+normal score via the normal reverse CDF function: \(\Phi^{-1}(.)\):
+\(\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]\), where
+\(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
+parameter and \(d_{\eta}\) is the number of parameter of \(\eta\).
+Sometimes, to account for the monthly variation in the data, the fitting
+is done separately for each month and the noramlising step can be
+written as \(\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]\), where
+\(j^*\) is all the indexes that satisfy \(j^* \mod 12 = m\) for each
+\(m = 0, 1,\cdots, 11\).
 
-Normalising is usually the last step
-
-\textbf{Variable transformation}
-
-Restrict it to single variable, square root, log etc could be linearly,
-also non-linear
-
-change the shape of the variable
-
-GAM, can you do additive model pairwise/ three-way
-
-\textbf{Temporal processing}
-
-\textbf{Dimension reduction}
-
-sometimes called feature extraction in the machine learning community
-With drought indexes, the extraction of meaningful variables from the
-original data is usually supported by the water balance model, for
-example, in SPEI, the step that create \(d\) out of precipitation and
-potential evapotranspiration (PET) has theoretical backup from {[}see
-paper.{]}
-
-Also include weighting
+\begin{itemize}
+\tightlist
+\item
+  In some cases, the distibution and inverse normal can be multivariate,
+  i.e.~use copula.
+\end{itemize}
 
@@
 
+Benchmarking sets a constant value to allow the constructed index to be
+compared across time. Here we denote it with \(u[x(s_i, t_j)]\) where
+\(u\) is a scalar of interest in the index constructed, could be a
+constant or a function of the data, i.e.~mean.
+
 \textbf{Simplification}
+
+In public communication, indexes are often delivered in categorical
+grades, along with its underlying numerical values. The simplification
+step prescribes how the continuous index values are converted into the
+discrete grades. This process can be written with the piece-wise
+function:
+
+\[
+\begin{cases}
+C_0 & c_1 \leq (s_i; t_j) < c_0 \\
+C_1 & c_2 \leq x(s_i; t_j) < c_1 \\
+C_2 & c_3 \leq x(s_i; t_j) < c_2 \\
+\cdots \\
+C_z & c_z \leq x(s_i; t_j)
+\end{cases}
+\]
+
+where \(C_0, C_1,\cdots ,C_z\) are the categories and
+\(c_0, c_1, \cdots, c_z\) are the threshold value in each category. In
+SPI, drought are sorted into four categories: mild drought:
+\([-0.99, 0]\); moderate drought: \([-1.49, -1]\); severe drought:
+\([-1.99, -1.5]\), and extreme drought: \([-\infty, -2]\). In this case,
+\(C_0, C_1, C_2, C_3\) are the drought categories: mild, moderate,
+severe, and extreme drought (\(z = 3\)) and
+\(c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2\) are the cutoff value for each
+class.
 

### 2022-11-22 paper/paper.qmd
\mathbf -> \symbf

@@
 format:
-  pdf:
+  arxiv-pdf:
     keep-tex: true  
-    include-in-header: 
-      text: |
-        \usepackage{amsmath}
 author:
@@
 
-*Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\mathbf{s}$ and $\mathbf{t}$ to depend on $p$*
+*Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\symbf{s}$ and $\symbf{t}$ to depend on $p$*
 
@@
 Once variables are collected, they can have different spatial resolution and temporal granularity.
-Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data that contain the spatial, temporal, and multivariate aspect: the spatial component $\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal component $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
-Expanding on the multivariate aspect, the data can be written as  $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$
+Let $\symbf{x}(\symbf{s};\symbf{t})$ denote the raw data that contain the spatial, temporal, and multivariate aspect: the spatial component $\symbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\symbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal component $\symbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\symbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
+Expanding on the multivariate aspect, the data can be written as  $\symbf{x}(\symbf{s}; \symbf{t}) = (x_1(\symbf{s}; \symbf{t}), x_2(\symbf{s}; \symbf{t}), \cdots, x_P(\symbf{s}; \symbf{t}))^\prime$
 
@@
 Temporal processing concerns the operations on the temporal aspect of the data.
-It can be represented by $f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))$, where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
+It can be represented by $f_{\mathcal{\psi}}(x(\symbf{s};\symbf{t}))$, where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
 
@@
 Sometimes, it can be beneficial to borrow information from nearby spatial locations or there is the need to combine raster data from satellite images with vector ground measures.
-Spatial aggregation can be written as $g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))$, where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.
+Spatial aggregation can be written as $g_{\mathcal{\theta}}(x(\symbf{s};\symbf{t}))$, where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.
 
@@
 A variable transformation changes the shape of the variable.
-It can be written as $h_{\tau}(x(\mathbf{s};\mathbf{t}))$, where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the potential parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
+It can be written as $h_{\tau}(x(\symbf{s};\symbf{t}))$, where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the potential parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
 Variable transformation involve square root transformation, log transformation, and also include those non-linear transformation.
@@
 When multiple variables are collected in space and time to construct indices, they are often summarised with dimension reduction.
-Dimension reduction $x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})$,  where $p^* = 1, 2, \cdots, P^*$,  $p = 1, 2, \cdots, P$, and $P^* < P$.
-The most commonly used dimension reduction techniquePprincipal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF), can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the variance of the data given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
+Dimension reduction $x_{p^*}(\symbf{s}; \symbf{t}) \rightarrow x_p(\symbf{s}; \symbf{t})$,  where $p^* = 1, 2, \cdots, P^*$,  $p = 1, 2, \cdots, P$, and $P^* < P$.
+The most commonly used dimension reduction techniquePprincipal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF), can be written as $x_{p^*}(\symbf{s}; \symbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\symbf{s};\symbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the variance of the data given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
 
@@
 This step is usually in the end of the pipeline when the data have been processed into a univariate series. 
-When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]$, where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
-Sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.
+When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $\Phi^{-1}[F_{\eta}(x(\symbf{s}; \symbf{t}))]$, where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
+Sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\symbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.
 

### 2022-11-22 paper/paper.tex
\mathbf -> \symbf

@@
 \documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+]{article}
 
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+  \setmathfont[]{Latin Modern Math}
 \fi
@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
+\usepackage{arxiv}
+\usepackage{orcidlink}
 \usepackage{amsmath}
-\KOMAoption{captions}{tableheading}
+\usepackage[T1]{fontenc}
+\usepackage{fontspec}
 \makeatletter
@@
 \title{Demo arXiv template}
-\author{H. Sherry Zhang \and Collaborators}
+\author{
+\textbf{H. Sherry Zhang}~\orcidlink{0000-0002-7122-1463}\\Department of
+Econometrics and Business Statistics\\Monash University\\Melbourne,
+VIC\\\href{mailto:huize.zhang@monash.edu}{huize.zhang@monash.edu}\\\\\\
+\textbf{Collaborators}\\Department of Econometrics and Business
+Statistics\\Monash University\\Melbourne, VIC\\}
 \date{}
-
 \begin{document}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi
+{\bfseries \emph Keywords}
+\def\sep{\textbullet\ }
+indexes \sep data pipeline \sep 
+software design
+
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, breakable, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{knitr}\SpecialCharTok{::}\FunctionTok{include\_graphics}\NormalTok{(here}\SpecialCharTok{::}\FunctionTok{here}\NormalTok{(}\StringTok{"figures/pipeline{-}steps.png"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{figure}[H]
+\begin{figure}
 
@@
 spatial/ temporal resolution: This would require the notation
-\(\mathbf{s}\) and \(\mathbf{t}\) to depend on \(p\)}
+\(\symbf{s}\) and \(\symbf{t}\) to depend on \(p\)}
 
@@
 
+This is what rendered by \texttt{\textbackslash{}symbf\{x\}}:
+\(uvwxyz, \boldsymbol{uvwxyz}, \pmb{uvwxyz}, \symbf{uvwxyz}\)
+
 Once variables are collected, they can have different spatial resolution
-and temporal granularity. Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\)
-denote the raw data that contain the spatial, temporal, and multivariate
+and temporal granularity. Let \(\symbf{x}(\symbf{s};\symbf{t})\) denote
+the raw data that contain the spatial, temporal, and multivariate
 aspect: the spatial component
-\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
-space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
-temporal component \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
+\(\symbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
+space: \(\symbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
+temporal component \(\symbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
 defined in the 1D space:
-\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). Expanding on the
+\(\symbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). Expanding on the
 multivariate aspect, the data can be written as
-\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\)
+\(\symbf{x}(\symbf{s}; \symbf{t}) = (x_1(\symbf{s}; \symbf{t}), x_2(\symbf{s}; \symbf{t}), \cdots, x_P(\symbf{s}; \symbf{t}))^\prime\)
 
@@
 the data. It can be represented by
-\(f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))\), where
+\(f_{\mathcal{\psi}}(x(\symbf{s};\symbf{t}))\), where
 \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the parameters
@@
 satellite images with vector ground measures. Spatial aggregation can be
-written as \(g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))\), where
+written as \(g_{\mathcal{\theta}}(x(\symbf{s};\symbf{t}))\), where
 \(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the spatial
@@
 A variable transformation changes the shape of the variable. It can be
-written as \(h_{\tau}(x(\mathbf{s};\mathbf{t}))\), where
+written as \(h_{\tau}(x(\symbf{s};\symbf{t}))\), where
 \(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the potential
@@
 reduction
-\(x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})\),
+\(x_{p^*}(\symbf{s}; \symbf{t}) \rightarrow x_p(\symbf{s}; \symbf{t})\),
 where \(p^* = 1, 2, \cdots, P^*\), \(p = 1, 2, \cdots, P\), and
@@
 Orthogonal Function (EOF), can be written as
-\(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
+\(x_{p^*}(\symbf{s}; \symbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\symbf{s};\symbf{t})\)
 where \(\lambda_p\) is the loading of the PC1, derived from maximising
@@
 normal score via the normal reverse CDF function: \(\Phi^{-1}(.)\):
-\(\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]\), where
+\(\Phi^{-1}[F_{\eta}(x(\symbf{s}; \symbf{t}))]\), where
 \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
@@
 is done separately for each month and the noramlising step can be
-written as \(\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]\), where
+written as \(\Phi^{-1}[F_{\eta}^m(x(\symbf{s};t_{j^*}))]\), where
 \(j^*\) is all the indexes that satisfy \(j^* \mod 12 = m\) for each

### 2022-11-22 paper/paper.qmd
more on scaling

@@
 
+  - divide into pre and post-processing
+
 More steps are involved before the mathematical construction of indexes. 
@@
 
+Variable transformation can be necessary for different reasons: 1) to change the shape of the variable, 2) to stablise the variance.
+
+mention modelling? -> satisfy model assumption
+
 A variable transformation changes the shape of the variable.
@@
 
-Scaling is a specific variable transformation on the scale of the data. 
-The purpose to treat scaling as a separate step from variable transformation is to make the step explicit. 
-Scaling is usually in the form of $[x(s_i;t_j) - \alpha]/\gamma$.
-For example, $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$ is used in a z-score standardisation to standardise the data by the its mean and standard deviation across all time and all space units and $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$ is used to standardise the variable into a unit interval.
+Sometimes, scaling, as a specific type of variable transformation, is merged into dimension reduction as a pre-step, i.e.
+transform the covariance matrix into correlation matrix before applying the Principal Component Analysis (PCA), but it worth to be separated into its own step to make it explicit.
+Its main different from variable transformation is that it can be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$ 
+For example, In a z-score standardisation, $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$ are used to standardise the data by the its mean and standard deviation across all time and all space units
+Another example is to scale the variable into a unit interval, which can be written as $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
+While being a common technique in data pre-processing, scaling has an impact on the outliers of the data and analysts need to be cautious on checking its effect.
 
@@
 The purpose of normalising is for cross-comparison using the normal score.
-This step can get criticism from analysts for ...
-
+This step can get criticism from analysts for being unnecessarily forcing the scale to the normal scale.
 This step is usually in the end of the pipeline when the data have been processed into a univariate series. 
@@
 In public communication, indexes are often delivered in categorical grades, along with its underlying numerical values.
-The simplification step prescribes how the continuous index values are converted into the discrete grades.
+The simplification step, sometimes can also be called discretisation, prescribes how the continuous index values are converted into the discrete grades.
 This process can be written with the piece-wise function:

### 2022-11-22 paper/paper.tex
more on scaling

@@
 \usepackage[T1]{fontenc}
-\usepackage{fontspec}
 \makeatletter
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, breakable, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
+\begin{itemize}
+\tightlist
+\item
+  divide into pre and post-processing
+\end{itemize}
+
 More steps are involved before the mathematical construction of indexes.
@@
 ``raw data''}
-
-This is what rendered by \texttt{\textbackslash{}symbf\{x\}}:
-\(uvwxyz, \boldsymbol{uvwxyz}, \pmb{uvwxyz}, \symbf{uvwxyz}\)
 
@@
 
+Variable transformation can be necessary for different reasons: 1) to
+change the shape of the variable, 2) to stablise the variance.
+
+mention modelling? -\textgreater{} satisfy model assumption
+
 A variable transformation changes the shape of the variable. It can be
@@
 
-Scaling is a specific variable transformation on the scale of the data.
-The purpose to treat scaling as a separate step from variable
-transformation is to make the step explicit. Scaling is usually in the
-form of \([x(s_i;t_j) - \alpha]/\gamma\). For example,
-\(\alpha = \bar{x}(s; t)\) and \(\gamma = \sigma(s; t)\) is used in a
-z-score standardisation to standardise the data by the its mean and
-standard deviation across all time and all space units and
-\(\alpha = \min[x(s_i, t_j)]\) and
-\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\) is used to
-standardise the variable into a unit interval.
+Sometimes, scaling, as a specific type of variable transformation, is
+merged into dimension reduction as a pre-step, i.e.~transform the
+covariance matrix into correlation matrix before applying the Principal
+Component Analysis (PCA), but it worth to be separated into its own step
+to make it explicit. Its main different from variable transformation is
+that it can be written in the form of \[[x(s_i;t_j) - \alpha]/\gamma.\]
+For example, In a z-score standardisation, \(\alpha = \bar{x}(s; t)\)
+and \(\gamma = \sigma(s; t)\) are used to standardise the data by the
+its mean and standard deviation across all time and all space units
+Another example is to scale the variable into a unit interval, which can
+be written as \(\alpha = \min[x(s_i, t_j)]\) and
+\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\). While being a common
+technique in data pre-processing, scaling has an impact on the outliers
+of the data and analysts need to be cautious on checking its effect.
 
@@
 The purpose of normalising is for cross-comparison using the normal
-score. This step can get criticism from analysts for \ldots{}
-
-This step is usually in the end of the pipeline when the data have been
-processed into a univariate series. When the data are not distributed
-normally, a distributed is usually fitted before converting to the
-normal score via the normal reverse CDF function: \(\Phi^{-1}(.)\):
+score. This step can get criticism from analysts for being unnecessarily
+forcing the scale to the normal scale. This step is usually in the end
+of the pipeline when the data have been processed into a univariate
+series. When the data are not distributed normally, a distributed is
+usually fitted before converting to the normal score via the normal
+reverse CDF function: \(\Phi^{-1}(.)\):
 \(\Phi^{-1}[F_{\eta}(x(\symbf{s}; \symbf{t}))]\), where
@@
 grades, along with its underlying numerical values. The simplification
-step prescribes how the continuous index values are converted into the
-discrete grades. This process can be written with the piece-wise
-function:
+step, sometimes can also be called discretisation, prescribes how the
+continuous index values are converted into the discrete grades. This
+process can be written with the piece-wise function:
 

### 2022-11-23 paper/paper.qmd
pdflatex is happy with \mathbf

--- 2022-11-21 paper/paper.qmd
@@
 format:
-  pdf:
+  arxiv-pdf:
     keep-tex: true  
-    include-in-header: 
-      text: |
-        \usepackage{amsmath}
+pdf-engine: pdflatex
 author:

### 2022-11-23 paper/paper.tex
pdflatex is happy with \mathbf

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
 
@@
 spatial/ temporal resolution: This would require the notation
-\(\symbf{s}\) and \(\symbf{t}\) to depend on \(p\)}
+\(\mathbf{s}\) and \(\mathbf{t}\) to depend on \(p\)}
 
@@
 Once variables are collected, they can have different spatial resolution
-and temporal granularity. Let \(\symbf{x}(\symbf{s};\symbf{t})\) denote
-the raw data that contain the spatial, temporal, and multivariate
+and temporal granularity. Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\)
+denote the raw data that contain the spatial, temporal, and multivariate
 aspect: the spatial component
-\(\symbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
-space: \(\symbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
-temporal component \(\symbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
+\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
+space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
+temporal component \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
 defined in the 1D space:
-\(\symbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). Expanding on the
+\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). Expanding on the
 multivariate aspect, the data can be written as
-\(\symbf{x}(\symbf{s}; \symbf{t}) = (x_1(\symbf{s}; \symbf{t}), x_2(\symbf{s}; \symbf{t}), \cdots, x_P(\symbf{s}; \symbf{t}))^\prime\)
+\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\)
 
@@
 the data. It can be represented by
-\(f_{\mathcal{\psi}}(x(\symbf{s};\symbf{t}))\), where
+\(f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))\), where
 \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the parameters
@@
 satellite images with vector ground measures. Spatial aggregation can be
-written as \(g_{\mathcal{\theta}}(x(\symbf{s};\symbf{t}))\), where
+written as \(g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))\), where
 \(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the spatial
@@
 A variable transformation changes the shape of the variable. It can be
-written as \(h_{\tau}(x(\symbf{s};\symbf{t}))\), where
+written as \(h_{\tau}(x(\mathbf{s};\mathbf{t}))\), where
 \(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the potential
@@
 reduction
-\(x_{p^*}(\symbf{s}; \symbf{t}) \rightarrow x_p(\symbf{s}; \symbf{t})\),
+\(x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})\),
 where \(p^* = 1, 2, \cdots, P^*\), \(p = 1, 2, \cdots, P\), and
@@
 Orthogonal Function (EOF), can be written as
-\(x_{p^*}(\symbf{s}; \symbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\symbf{s};\symbf{t})\)
+\(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
 where \(\lambda_p\) is the loading of the PC1, derived from maximising
@@
 reverse CDF function: \(\Phi^{-1}(.)\):
-\(\Phi^{-1}[F_{\eta}(x(\symbf{s}; \symbf{t}))]\), where
+\(\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]\), where
 \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
@@
 is done separately for each month and the noramlising step can be
-written as \(\Phi^{-1}[F_{\eta}^m(x(\symbf{s};t_{j^*}))]\), where
+written as \(\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]\), where
 \(j^*\) is all the indexes that satisfy \(j^* \mod 12 = m\) for each

### 2022-11-23 paper/paper.qmd
add a plot to distinguish variable transformation and scaling

@@
 editor_options: 
-  chunk_output_type: inline
+  chunk_output_type: console
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)
+
+```
 
@@
 These steps do not have to be arranged in sequence and not all the steps are required to present in every index construction. 
-Spatial aggregation and temporal processing are operations of the data in space and time 
+Spatial aggregation and temporal processing are operations of the data in space and time. 
 Variable transformation, scaling, and dimension reduction concerns operations on the multivariate aspect of the data.
-Normalisation, benchmarking, and simplification can be viewed as the "post-processing" of indexes, which addresses the cross-comparison across different indexes and time, and index communication.
+Normalisation, benchmarking, and simplification can be viewed as the post-processing of indexes, which addresses the cross-comparison across different indexes and time, and index communication.
 
@@
 transform the covariance matrix into correlation matrix before applying the Principal Component Analysis (PCA), but it worth to be separated into its own step to make it explicit.
-Its main different from variable transformation is that it can be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$ 
+Its main different from variable transformation is that scaling does not change the shape of the data and it can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
+@fig-scale-var-trans-compare shows a range of operations on a simulated variable. 
 For example, In a z-score standardisation, $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$ are used to standardise the data by the its mean and standard deviation across all time and all space units
@@
 <!-- close to variable transformation step -->
+
+```{r fig-scale-var-trans-compare}
+#| fig-cap: "Comparison of scaling and variable transformation."
+#| fig-height: 8
+#| fig-width: 12
+library(tidyverse)
+library(latex2exp)
+set.seed(123)
+x <- rgamma(n = 1000, shape = 1, scale = 2)
+
+trans_df <- tibble(origin = x, 
+       `z-score` = (x - mean(x)/ sd(x)), 
+       qudratic = x^2,
+       log = log(x),
+       `square root` = sqrt(x),
+       `cubic root` = x^(1/3),
+       minmax = (x - min(x))/ (max(x) - min(x)),
+       boxcox = (x^0.5 - 1)/0.5,
+       centering = x - mean(x)
+       ) %>% 
+  mutate(id = row_number(),
+         ) %>% 
+  pivot_longer(-id, names_to = "var", values_to = "value") %>% 
+  mutate(category = ifelse(
+    var %in% c("origin", "z-score", "minmax", "centering"),
+    "Scaling", "Var. Trans.")
+    )
+
+latex_df <- tibble(
+  var = c("origin", "z-score", "qudratic", "log", "square root", "cubic root",
+          "minmax", "boxcox", "centering"),
+  latex = c(
+           r"($x$)",
+           r"($\frac{x - \bar{x}}{sd(x)}$)",
+           r"($x^2$)",
+           r"($log(x)$)",
+           r"($\sqrt{x}$)",
+           r"($x^{1/3}$)",
+           r"($\frac{x - min(x)}{max(x) - min(x)}$)",
+           r"($\frac{x^{0.5} - 1}{0.5}$)",
+           r"($x - \bar{x}$)"
+         )
+) %>% 
+  left_join(trans_df %>% group_by(var) %>% summarise(max = max(value)))
+
+trans_df %>% 
+  ggplot() +
+  geom_density(aes(x = value, color = category, y = after_stat(scaled))) + 
+  geom_label(data = latex_df, 
+             aes(x = 0.8 * max, y = 0.8, label = TeX(latex, output = "character")), 
+             parse = TRUE) + 
+  facet_wrap(vars(var), scales = "free") + 
+  scale_color_brewer(palette = "Dark2") + 
+  theme_bw() + 
+  theme(legend.position = "bottom")
+```
 

### 2022-11-23 paper/paper.tex
add a plot to distinguish variable transformation and scaling

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, boxrule=0pt, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 to make it explicit. Its main different from variable transformation is
-that it can be written in the form of \[[x(s_i;t_j) - \alpha]/\gamma.\]
-For example, In a z-score standardisation, \(\alpha = \bar{x}(s; t)\)
-and \(\gamma = \sigma(s; t)\) are used to standardise the data by the
-its mean and standard deviation across all time and all space units
-Another example is to scale the variable into a unit interval, which can
-be written as \(\alpha = \min[x(s_i, t_j)]\) and
-\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\). While being a common
-technique in data pre-processing, scaling has an impact on the outliers
-of the data and analysts need to be cautious on checking its effect.
+that scaling does not change the shape of the data and it can usually be
+written in the form of \[[x(s_i;t_j) - \alpha]/\gamma.\]
+Figure~\ref{fig-scale-var-trans-compare} shows a range of operations on
+a simulated variable. For example, In a z-score standardisation,
+\(\alpha = \bar{x}(s; t)\) and \(\gamma = \sigma(s; t)\) are used to
+standardise the data by the its mean and standard deviation across all
+time and all space units Another example is to scale the variable into a
+unit interval, which can be written as \(\alpha = \min[x(s_i, t_j)]\)
+and \(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\). While being a
+common technique in data pre-processing, scaling has an impact on the
+outliers of the data and analysts need to be cautious on checking its
+effect.
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
+
+}
+
+\caption{\label{fig-scale-var-trans-compare}Comparison of scaling and
+variable transformation.}
+
+\end{figure}
 

### 2022-11-23 paper/paper.qmd
more drafting on the pipeline steps

@@
 
-*constructing time series index should also be encapsulated in the framework*
+<!-- *constructing time series index should also be encapsulated in the framework* -->
 
-@fig-pipeline-steps presents the index construction pipeline from the raw data to the final indexes. 
+An overview of the pipeline is given in @fig-pipeline-steps to illustrate the construction from raw data to the final indexes.
 The available intermediate steps include temporal processing, spatial aggregation, variable transformation, scaling, dimension reduction, normalising, benchmarking, and simplification. 
-These steps do not have to be arranged in sequence and not all the steps are required to present in every index construction. 
-Spatial aggregation and temporal processing are operations of the data in space and time. 
-Variable transformation, scaling, and dimension reduction concerns operations on the multivariate aspect of the data.
-Normalisation, benchmarking, and simplification can be viewed as the post-processing of indexes, which addresses the cross-comparison across different indexes and time, and index communication.
-
-  - divide into pre and post-processing
-
-More steps are involved before the mathematical construction of indexes. 
-This includes 1) defining a concept of interest, being useful for decision making and 2) finding the relevant and measurable variables to characterise the concept. 
-These choices define the scope of the indexes and the data availability for constructing an index but not the construction itself, hence not included in the pipeline framework. 
+These steps are not required to be arranged in a fixed order and not all the steps have to appear when constructing each index. 
+The pipeline includes modules for operations in the spatial, temporal, and multivariate aspect of the data, as well as modules for comparing and communicating indexes (benchmarking and simplification).
+While the pipeline starts with the raw data, steps prior to this, i.e.
+define a useful concept to construct the index, collect high quality data relevant for measuring the concept, are also cricial to the sucess of an index.
 The framework proposed in this paper assumes the objective of an index has been defined and relevant variables have been collected and quality controlled.
@@
 
-*Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\mathbf{s}$ and $\mathbf{t}$ to depend on $p$*
+<!-- *Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\mathbf{s}$ and $\mathbf{t}$ to depend on $p$* -->
 
-*Initially my thought was to raw data only after aligning the resolution: Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem.
-After processing them and align them together they become the "raw data"*
+<!-- *Initially my thought was to raw data only after aligning the resolution: Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem.
+After processing them and align them together they become the "raw data"* -->
 
-Once variables are collected, they can have different spatial resolution and temporal granularity.
-Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data that contain the spatial, temporal, and multivariate aspect: the spatial component $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal component $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
-Expanding on the multivariate aspect, the data can be written as  $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$
+<!-- Once variables are collected, they can have different spatial resolution and temporal granularity.  -->
+
+Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspect: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
+When more than one variable are invovled, the multivariate data can also be written as: $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$.
 
@@
 
-Temporal processing concerns the operations on the temporal aspect of the data.
-It can be represented by $f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))$, where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
-
-- purpose is to aggregate across time, smooth?
-
-An example of temporal processing is to aggregate the data, $x(s_i; t_j)$, into $k$-month sum.
-This can be used to reflect how the accumulated values change across time.
-In the drought index SPI, precipitation can be aggregated at different scales to reflect short-term, mid-term, and long-term drought severity.
-With $\psi = k$, the aggregated series can be written as $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)$ where $j^\prime$ is the time index after the aggregation.
+While many variables are collected separately at each time period, the underlying phenomenon, characterised by the index, is continuous. 
+For example, precipitation can be measured daily or monthly, a meteorological drought is characterised by a prelonged lack of precipitation. 
+Temporal proocessing is a step in the pipeline that includes operations in the temporal dimension and it can be generally represented by $$f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),$$ where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
+The purpose of temporal aggregation can be to aggregate variable across time or to borrow information from nearby time periods to smooth or adjust the current value.
+The drought index SPI is constructed from the aggregated precipitation with a time scale parameter $\psi = k$.
+The aggregation step in SPI can be written as $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)$ where $j^\prime$ is the time dimension after the aggregation.
 
@@
 
-Sometimes, it can be beneficial to borrow information from nearby spatial locations or there is the need to combine raster data from satellite images with vector ground measures.
-Spatial aggregation can be written as $g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))$, where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.
+Sometimes, information from neighbouring spatial locations can also be borrowed when constructing indexes. 
+This involves interpolating data from observed locations to the unobserved and aggregating data into different levels.
+In the pipeline, spatial aggregation can be written as $$g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),$$ where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.
+[GIVEN AN EXAMPLE]
 
- - $\theta$ can be a constant or a derived from the data
+<!-- $x(s_i^\prime;t_j) = \sum_{i \in \Omega}x(s_i; t_j)$ -->
 
- - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level.
+<!-- Another common use of spatial processing is to integrate satellite imagaries with ground measures, also merge satellite imagaries with different resolutions.  -->
 
-**Variable transformation**
+ <!-- - $\theta$ can be a constant or a derived from the data -->
 
-Variable transformation can be necessary for different reasons: 1) to change the shape of the variable, 2) to stablise the variance.
+ <!-- - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level. -->
 
-mention modelling? -> satisfy model assumption
+\newpage
 
-A variable transformation changes the shape of the variable.
-It can be written as $h_{\tau}(x(\mathbf{s};\mathbf{t}))$, where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the potential parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
-Variable transformation involve square root transformation, log transformation, and also include those non-linear transformation.
+**Variable transformation **
+
+Variable transformation and scaling, in the next section, are both variable pre-processing steps to prepare the data with desire statistical property to work with.
+This can be a more stablised variance, a normal distribution that satisfies modelling assumptions, or a data range that required by the algorithms.
+Variable tranformation is a general notion of a functional transformation on the variable: $$h_{\tau}(x(\mathbf{s};\mathbf{t})),$$ where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
+Typical examples of variable transformation are log, quadratic, square root, and box-cox transformation.
+
+**Scaling**
+
+While scaling can be seen as a specific type of variable tranformation, it is separated into own step to make the scaling operation explicit in the pipeline.
+The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
+For example, A z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
+@fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step.
+All the scaling operations in green results in the same shape as the original data.
 
@@
 
-**Scaling**
-
-Sometimes, scaling, as a specific type of variable transformation, is merged into dimension reduction as a pre-step, i.e.
-transform the covariance matrix into correlation matrix before applying the Principal Component Analysis (PCA), but it worth to be separated into its own step to make it explicit.
-Its main different from variable transformation is that scaling does not change the shape of the data and it can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
-@fig-scale-var-trans-compare shows a range of operations on a simulated variable. 
-For example, In a z-score standardisation, $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$ are used to standardise the data by the its mean and standard deviation across all time and all space units
-Another example is to scale the variable into a unit interval, which can be written as $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
-While being a common technique in data pre-processing, scaling has an impact on the outliers of the data and analysts need to be cautious on checking its effect.
-
-<!-- z-score standardising, min-max standardisation into [0, 1] or [0, 100], percentage change on the baseline -->
-<!-- close to variable transformation step -->
+<!-- While being a common technique in data pre-processing, scaling has an impact on the outliers of the data and analysts need to be cautious on checking its effect. -->
 
@@
 
+\newpage
+
 **Dimension reduction**
 
-When multiple variables are collected in space and time to construct indices, they are often summarised with dimension reduction.
-Dimension reduction $x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})$,  where $p^* = 1, 2, \cdots, P^*$,  $p = 1, 2, \cdots, P$, and $P^* < P$.
-The most commonly used dimension reduction techniquePprincipal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF), can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the variance of the data given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
+While multivariate information can be collected on a concept, they need to be summarised into a single number for the final index.
+Dimension reduction step $$x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),$$ where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$, reduces the variable dimension from $P$ to $P^*$.
+The most commonly used dimension reduction technique Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science, can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the data variance given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
 
@@
 This step is usually in the end of the pipeline when the data have been processed into a univariate series. 
-When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]$, where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
+When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $$\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))],$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
 Sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.

### 2022-11-23 paper/paper.tex
more drafting on the pipeline steps

--- 2022-11-23 paper/paper.tex
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\emph{constructing time series index should also be encapsulated in the
-framework}
-
-Figure~\ref{fig-pipeline-steps} presents the index construction pipeline
-from the raw data to the final indexes. The available intermediate steps
-include temporal processing, spatial aggregation, variable
-transformation, scaling, dimension reduction, normalising, benchmarking,
-and simplification. These steps do not have to be arranged in sequence
-and not all the steps are required to present in every index
-construction. Spatial aggregation and temporal processing are operations
-of the data in space and time Variable transformation, scaling, and
-dimension reduction concerns operations on the multivariate aspect of
-the data. Normalisation, benchmarking, and simplification can be viewed
-as the ``post-processing'' of indexes, which addresses the
-cross-comparison across different indexes and time, and index
-communication.
-
-\begin{itemize}
-\tightlist
-\item
-  divide into pre and post-processing
-\end{itemize}
-
-More steps are involved before the mathematical construction of indexes.
-This includes 1) defining a concept of interest, being useful for
-decision making and 2) finding the relevant and measurable variables to
-characterise the concept. These choices define the scope of the indexes
-and the data availability for constructing an index but not the
-construction itself, hence not included in the pipeline framework. The
-framework proposed in this paper assumes the objective of an index has
-been defined and relevant variables have been collected and quality
-controlled.
+An overview of the pipeline is given in Figure~\ref{fig-pipeline-steps}
+to illustrate the construction from raw data to the final indexes. The
+available intermediate steps include temporal processing, spatial
+aggregation, variable transformation, scaling, dimension reduction,
+normalising, benchmarking, and simplification. These steps are not
+required to be arranged in a fixed order and not all the steps have to
+appear when constructing each index. The pipeline includes modules for
+operations in the spatial, temporal, and multivariate aspect of the
+data, as well as modules for comparing and communicating indexes
+(benchmarking and simplification). While the pipeline starts with the
+raw data, steps prior to this, i.e.~define a useful concept to construct
+the index, collect high quality data relevant for measuring the concept,
+are also cricial to the sucess of an index. The framework proposed in
+this paper assumes the objective of an index has been defined and
+relevant variables have been collected and quality controlled.
 
@@
 
-\emph{Discussion: whether we should allow raw data to have different
-spatial/ temporal resolution: This would require the notation
-\(\mathbf{s}\) and \(\mathbf{t}\) to depend on \(p\)}
-
-\emph{Initially my thought was to raw data only after aligning the
-resolution: Another section on original data directly downloaded, can
-have different spatial resolution, temporal granularity, data quality
-problem. After processing them and align them together they become the
-``raw data''}
-
-Once variables are collected, they can have different spatial resolution
-and temporal granularity. Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\)
-denote the raw data that contain the spatial, temporal, and multivariate
-aspect: the spatial component
+Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with
+spatial, temporal, and multivariate aspect: the spatial dimension
 \(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
 space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
-temporal component \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
+temporal dimension \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
 defined in the 1D space:
-\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). Expanding on the
-multivariate aspect, the data can be written as
-\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\)
+\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). When more than
+one variable are invovled, the multivariate data can also be written as:
+\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\).
 
@@
 
-Temporal processing concerns the operations on the temporal aspect of
-the data. It can be represented by
-\(f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))\), where
+While many variables are collected separately at each time period, the
+underlying phenomenon, characterised by the index, is continuous. For
+example, precipitation can be measured daily or monthly, a
+meteorological drought is characterised by a prelonged lack of
+precipitation. Temporal proocessing is a step in the pipeline that
+includes operations in the temporal dimension and it can be generally
+represented by \[f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),\] where
 \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the parameters
 associated with the temporal operation and \(d_{\psi}\) is the number of
-parameter of \(\psi\).
-
-\begin{itemize}
-\tightlist
-\item
-  purpose is to aggregate across time, smooth?
-\end{itemize}
-
-An example of temporal processing is to aggregate the data,
-\(x(s_i; t_j)\), into \(k\)-month sum. This can be used to reflect how
-the accumulated values change across time. In the drought index SPI,
-precipitation can be aggregated at different scales to reflect
-short-term, mid-term, and long-term drought severity. With \(\psi = k\),
-the aggregated series can be written as
+parameter of \(\psi\). The purpose of temporal aggregation can be to
+aggregate variable across time or to borrow information from nearby time
+periods to smooth or adjust the current value. The drought index SPI is
+constructed from the aggregated precipitation with a time scale
+parameter \(\psi = k\). The aggregation step in SPI can be written as
 \(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)\)
-where \(j^\prime\) is the time index after the aggregation.
+where \(j^\prime\) is the time dimension after the aggregation.
 
@@
 
-Sometimes, it can be beneficial to borrow information from nearby
-spatial locations or there is the need to combine raster data from
-satellite images with vector ground measures. Spatial aggregation can be
-written as \(g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))\), where
+Sometimes, information from neighbouring spatial locations can also be
+borrowed when constructing indexes. This involves interpolating data
+from observed locations to the unobserved and aggregating data into
+different levels. In the pipeline, spatial aggregation can be written as
+\[g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),\] where
 \(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the spatial
 operation parameters and \(d_{\theta}\) is the number of parameter of
-\(\theta\).
+\(\theta\). {[}GIVEN AN EXAMPLE{]}
 
-\begin{itemize}
-\item
-  \(\theta\) can be a constant or a derived from the data
-\item
-  3 by 3 aggregation, another purpose: the spatial grid is too fine,
-  aggregate to higher level.
-\end{itemize}
+\newpage
 
-\textbf{Variable transformation}
+\textbf{Variable transformation }
 
-Variable transformation can be necessary for different reasons: 1) to
-change the shape of the variable, 2) to stablise the variance.
-
-mention modelling? -\textgreater{} satisfy model assumption
-
-A variable transformation changes the shape of the variable. It can be
-written as \(h_{\tau}(x(\mathbf{s};\mathbf{t}))\), where
-\(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the potential
-parameter in the transformation, if any, and \(d_{\tau}\) is the number
-of parameter of \(\tau\). Variable transformation involve square root
-transformation, log transformation, and also include those non-linear
-transformation.
+Variable transformation and scaling, in the next section, are both
+variable pre-processing steps to prepare the data with desire
+statistical property to work with. This can be a more stablised
+variance, a normal distribution that satisfies modelling assumptions, or
+a data range that required by the algorithms. Variable tranformation is
+a general notion of a functional transformation on the variable:
+\[h_{\tau}(x(\mathbf{s};\mathbf{t})),\] where
+\(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in the
+transformation, if any, and \(d_{\tau}\) is the number of parameter of
+\(\tau\). Typical examples of variable transformation are log,
+quadratic, square root, and box-cox transformation.
 
@@
 
-Sometimes, scaling, as a specific type of variable transformation, is
-merged into dimension reduction as a pre-step, i.e.~transform the
-covariance matrix into correlation matrix before applying the Principal
-Component Analysis (PCA), but it worth to be separated into its own step
-to make it explicit. Its main different from variable transformation is
-that it can be written in the form of \[[x(s_i;t_j) - \alpha]/\gamma.\]
-For example, In a z-score standardisation, \(\alpha = \bar{x}(s; t)\)
-and \(\gamma = \sigma(s; t)\) are used to standardise the data by the
-its mean and standard deviation across all time and all space units
-Another example is to scale the variable into a unit interval, which can
-be written as \(\alpha = \min[x(s_i, t_j)]\) and
-\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\). While being a common
-technique in data pre-processing, scaling has an impact on the outliers
-of the data and analysts need to be cautious on checking its effect.
+While scaling can be seen as a specific type of variable tranformation,
+it is separated into own step to make the scaling operation explicit in
+the pipeline. The key difference between the two steps is that variable
+transformation typically changes the shape of the data while scaling
+only changes the data scale and can usually be written in the form of
+\[[x(s_i;t_j) - \alpha]/\gamma.\] For example, A z-score standardisation
+can be written in the above form with \(\alpha = \bar{x}(s; t)\) and
+\(\gamma = \sigma(s; t)\), a min-max standardisation uses
+\(\alpha = \min[x(s_i, t_j)]\) and
+\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\).
+Figure~\ref{fig-scale-var-trans-compare} shows a collection of variable
+pre-processing operations and uses color to differentiate whether the
+operation is a variable transformation or a scaling step. All the
+scaling operations in green results in the same shape as the original
+data.
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
+
+}
+
+\caption{\label{fig-scale-var-trans-compare}Comparison of scaling and
+variable transformation.}
+
+\end{figure}
+
+\newpage
 
@@
 
-When multiple variables are collected in space and time to construct
-indices, they are often summarised with dimension reduction. Dimension
-reduction
-\(x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})\),
-where \(p^* = 1, 2, \cdots, P^*\), \(p = 1, 2, \cdots, P\), and
-\(P^* < P\). The most commonly used dimension reduction
-techniquePprincipal Component Analysis (PCA), also called Empirical
-Orthogonal Function (EOF), can be written as
+While multivariate information can be collected on a concept, they need
+to be summarised into a single number for the final index. Dimension
+reduction step
+\[x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),\]
+where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\), reduces
+the variable dimension from \(P\) to \(P^*\). The most commonly used
+dimension reduction technique Principal Component Analysis (PCA), also
+called Empirical Orthogonal Function (EOF) in earth science, can be
+written as
 \(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
 where \(\lambda_p\) is the loading of the PC1, derived from maximising
-the variance of the data given the constraint
-\(\sum_{p=1}^P\lambda_p^2 = 1\).
+the data variance given the constraint \(\sum_{p=1}^P\lambda_p^2 = 1\).
 
@@
 reverse CDF function: \(\Phi^{-1}(.)\):
-\(\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]\), where
+\[\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))],\] where
 \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution

### 2022-11-23 paper/paper.qmd
updates after meeting

@@
 
-**Raw data**
+**Notation**
 
@@
 
-<!-- The location $s$ can refer to vector points or areas characterised by longitude-latitude coordinates, or raster cells obtained from satellite images. -->
-
-<!-- The time dimension $t$ can be daily, weekly, biweekly (14-16 days), monthly, or even quarterly -->
-
-<!-- Variables -->
-
-<!-- This multidimensional array structure is commonly used in geospatial analysis -->
-
-<!-- Given the variety of data sources at different spatial resolution and temporal granularity, the raw data may first come in multiple pieces.  -->
-<!-- Sometimes, even a considerable amount of work is needed to align the spatial and temporal extent of multivariate data. -->
+  <!-- * all the different choice in each step will affect the final index series ot make decision, i.e.
+$k$ in temporal aggregation, to assess drought in different scale: short-term, mid-term, long-term.
+-->
 
@@
 
- <!-- - $\theta$ can be a constant or a derived from the data -->
-
  <!-- - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level. -->
@@
 The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
-For example, A z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
+For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
 @fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step.
@@
 
+**Distribution fit**
+
+$$F_{\eta}(x(\mathbf{s}; \mathbf{t}))$$
+
 **Normalising**
@@
 - In some cases, the distibution and inverse normal can be multivariate, i.e. use copula.
-
-<!-- specifically for converting from a fitted distribution to normal score via reverse CDF function, non-parametric formula, or empirical approximation, a common step in many index: SPI, SSI, Z score.
-The purpose of normalising is to convert the index into a standardised series after all the steps for the ease of comparison.
--->
 

### 2022-11-23 paper/paper.tex
updates after meeting

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\textbf{Raw data}
+\textbf{Notation}
 
@@
 only changes the data scale and can usually be written in the form of
-\[[x(s_i;t_j) - \alpha]/\gamma.\] For example, A z-score standardisation
+\[[x(s_i;t_j) - \alpha]/\gamma.\] For example, a z-score standardisation
 can be written in the above form with \(\alpha = \bar{x}(s; t)\) and
@@
 the data variance given the constraint \(\sum_{p=1}^P\lambda_p^2 = 1\).
+
+\textbf{Distribution fit}
+
+\[F_{\eta}(x(\mathbf{s}; \mathbf{t}))\]
 

### 2022-11-27 paper/paper.qmd
more content in temporal processing

@@
 An overview of the pipeline is given in @fig-pipeline-steps to illustrate the construction from raw data to the final indexes.
-The available intermediate steps include temporal processing, spatial aggregation, variable transformation, scaling, dimension reduction, normalising, benchmarking, and simplification. 
+The pipeline includes modules for operations in the spatial, temporal, and multivariate aspect of the data: temporal processing, spatial aggregation, variable transformation, scaling, dimension reduction, and normalising, as well as modules for comparing and communicating indexes (benchmarking and simplification).
 These steps are not required to be arranged in a fixed order and not all the steps have to appear when constructing each index. 
-The pipeline includes modules for operations in the spatial, temporal, and multivariate aspect of the data, as well as modules for comparing and communicating indexes (benchmarking and simplification).
 While the pipeline starts with the raw data, steps prior to this, i.e.
@@
 Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspect: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
-When more than one variable are invovled, the multivariate data can also be written as: $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$.
+When more than one variable are involved, the multivariate data can also be written as: $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$.
 
@@
 
-While many variables are collected separately at each time period, the underlying phenomenon, characterised by the index, is continuous. 
-For example, precipitation can be measured daily or monthly, a meteorological drought is characterised by a prelonged lack of precipitation. 
-Temporal proocessing is a step in the pipeline that includes operations in the temporal dimension and it can be generally represented by $$f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),$$ where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
-The purpose of temporal aggregation can be to aggregate variable across time or to borrow information from nearby time periods to smooth or adjust the current value.
-The drought index SPI is constructed from the aggregated precipitation with a time scale parameter $\psi = k$.
-The aggregation step in SPI can be written as $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)$ where $j^\prime$ is the time dimension after the aggregation.
+The construction of an index sometimes needs to consider information from neighbouring time periods.
+Temporal processing is a general operator on the time dimension of the data in the form of  $$f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),$$ where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
+An typical example of temporal processing is aggregation in the drought index SPI to measure the lack of precipitation for meteorological drought.
+In SPI, monthly precipitation is aggregated by a time scale parameter $k$: $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),$ where $j^\prime$ is the new time index after the aggregation.
+In this notation, each spatial location is seaprately aggregated and precipitation is ummed from $k$ month back, $j^\prime - k + 1$, to the current, $j^\prime$ period to create the aggregated series, indexed by $j^\prime$.
+
+The choice of time scales parameter $k$ can result in variation in the calculated index values: a small $k$ of 3 or 6 month produces the index more sensitive to individual months, while a large $k$ of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives dryness information relative to the long term condition.
+As will be shown in section [SECTION EXAMPLE], this variation may even lead to conflicting conclusions on the dry/wet condition of the area, highlighting the importance to account for index uncertainty when interpreting index values for decision making.
 

### 2022-11-27 paper/paper.tex
more content in temporal processing

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 to illustrate the construction from raw data to the final indexes. The
-available intermediate steps include temporal processing, spatial
-aggregation, variable transformation, scaling, dimension reduction,
-normalising, benchmarking, and simplification. These steps are not
-required to be arranged in a fixed order and not all the steps have to
-appear when constructing each index. The pipeline includes modules for
-operations in the spatial, temporal, and multivariate aspect of the
-data, as well as modules for comparing and communicating indexes
-(benchmarking and simplification). While the pipeline starts with the
-raw data, steps prior to this, i.e.~define a useful concept to construct
-the index, collect high quality data relevant for measuring the concept,
-are also cricial to the sucess of an index. The framework proposed in
-this paper assumes the objective of an index has been defined and
-relevant variables have been collected and quality controlled.
+pipeline includes modules for operations in the spatial, temporal, and
+multivariate aspect of the data: temporal processing, spatial
+aggregation, variable transformation, scaling, dimension reduction, and
+normalising, as well as modules for comparing and communicating indexes
+(benchmarking and simplification). These steps are not required to be
+arranged in a fixed order and not all the steps have to appear when
+constructing each index. While the pipeline starts with the raw data,
+steps prior to this, i.e.~define a useful concept to construct the
+index, collect high quality data relevant for measuring the concept, are
+also cricial to the sucess of an index. The framework proposed in this
+paper assumes the objective of an index has been defined and relevant
+variables have been collected and quality controlled.
 
@@
 \(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). When more than
-one variable are invovled, the multivariate data can also be written as:
+one variable are involved, the multivariate data can also be written as:
 \(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\).
@@
 
-While many variables are collected separately at each time period, the
-underlying phenomenon, characterised by the index, is continuous. For
-example, precipitation can be measured daily or monthly, a
-meteorological drought is characterised by a prelonged lack of
-precipitation. Temporal proocessing is a step in the pipeline that
-includes operations in the temporal dimension and it can be generally
-represented by \[f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),\] where
+The construction of an index sometimes needs to consider information
+from neighbouring time periods. Temporal processing is a general
+operator on the time dimension of the data in the form of
+\[f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),\] where
 \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the parameters
 associated with the temporal operation and \(d_{\psi}\) is the number of
-parameter of \(\psi\). The purpose of temporal aggregation can be to
-aggregate variable across time or to borrow information from nearby time
-periods to smooth or adjust the current value. The drought index SPI is
-constructed from the aggregated precipitation with a time scale
-parameter \(\psi = k\). The aggregation step in SPI can be written as
-\(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)\)
-where \(j^\prime\) is the time dimension after the aggregation.
+parameter of \(\psi\). An typical example of temporal processing is
+aggregation in the drought index SPI to measure the lack of
+precipitation for meteorological drought. In SPI, monthly precipitation
+is aggregated by a time scale parameter \(k\):
+\(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),\)
+where \(j^\prime\) is the new time index after the aggregation. In this
+notation, each spatial location is seaprately aggregated and
+precipitation is ummed from \(k\) month back, \(j^\prime - k + 1\), to
+the current, \(j^\prime\) period to create the aggregated series,
+indexed by \(j^\prime\).
+
+The choice of time scales parameter \(k\) can result in variation in the
+calculated index values: a small \(k\) of 3 or 6 month produces the
+index more sensitive to individual months, while a large \(k\) of 24 or
+36, an equivalent to a 2- or 3-year aggregation, gives dryness
+information relative to the long term condition. As will be shown in
+section {[}SECTION EXAMPLE{]}, this variation may even lead to
+conflicting conclusions on the dry/wet condition of the area,
+highlighting the importance to account for index uncertainty when
+interpreting index values for decision making.
 

### 2022-11-27 paper/paper.qmd
more on the normalising step

@@
 
+collect ground and satellite image -> merge
 <!-- $x(s_i^\prime;t_j) = \sum_{i \in \Omega}x(s_i; t_j)$ -->
@@
 
+weighting
+
 <!-- sometimes called feature extraction in the machine learning community  -->
@@
 
-$$F_{\eta}(x(\mathbf{s}; \mathbf{t}))$$
+*modell fit? *
+
+$$F_{\eta}(x(\mathbf{s}; \mathbf{t})),$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
+
+Considerations need to be given on selecting the best distribution to fit, 
+Various discussion on the fitting methods and goodness of fit. 
+Potentially, more care about how much difference each combination give in terms of the index value.
+
+sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.
+
+- In some cases, the distibution and inverse normal can be multivariate, i.e. use copula.
 
@@
 
-The purpose of normalising is for cross-comparison using the normal score.
-This step can get criticism from analysts for being unnecessarily forcing the scale to the normal scale.
-This step is usually in the end of the pipeline when the data have been processed into a univariate series. 
-When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $$\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))],$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
-Sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.
+This step maps the univariate series into a different scale, typically for the ease of comparison.
+For example, a normal scale, [0, 1], [0, 100] may be favored to report certain indexes.
+In drought indexes, i.e.
+SPI or SPEI, the quantiles from the fitted distribution is converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$.
 
-- In some cases, the distibution and inverse normal can be multivariate, i.e. use copula.
+Normalising is usually used in the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. 
+While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers.
+Also, the use of normal scale needs to be interpreted with caution.
+@fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
+This is concerning, especially at the extreme values, where a normal density of -5 and -4 has magnitudes of difference into its probability of occurrence (1 in 3,000,000 and 1 in 31,000, respectively).
+
+```{r fig-normalising}
+#| fig-cap: "scatterplot of quantile vs. its density values."
+library(tidyverse)
+all <- tibble(x = seq(-5, 5, 0.01), y = pnorm(x)) 
+dt <- tibble(x = c(-5, -4), y = pnorm(x)) 
+label <- tibble(x = c(-5, -4), y = pnorm(x), 
+                label = c("a chance \nof 1 in \n 3,488,556", "a chance \nof 1 in \n 31,574"))
+
+dt %>% 
+  ggplot() + 
+  geom_point(aes(x= x, y =y)) + 
+  geom_line(data = all, aes(x = x, y =y)) + 
+  ggrepel::geom_label_repel(
+    data = label, aes(x = x, y = y, label =label), 
+    nudge_y = 0.2, arrow = arrow(length = unit(0.08, "inches")), min.segment.length = 0) + 
+  scale_x_continuous(breaks = seq(-5, 5, 1)) + 
+  xlab("Density") + 
+  ylab("Quantile") + 
+  theme_bw() + 
+  theme(panel.grid.minor = element_blank())
+```
 

### 2022-11-27 paper/paper.tex
more on the normalising step

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners, enhanced, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
+collect ground and satellite image -\textgreater{} merge
+
 \newpage
@@
 
+weighting
+
 \textbf{Distribution fit}
 
-\[F_{\eta}(x(\mathbf{s}; \mathbf{t}))\]
+\emph{modell fit? }
 
-\textbf{Normalising}
-
-The purpose of normalising is for cross-comparison using the normal
-score. This step can get criticism from analysts for being unnecessarily
-forcing the scale to the normal scale. This step is usually in the end
-of the pipeline when the data have been processed into a univariate
-series. When the data are not distributed normally, a distributed is
-usually fitted before converting to the normal score via the normal
-reverse CDF function: \(\Phi^{-1}(.)\):
-\[\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))],\] where
+\[F_{\eta}(x(\mathbf{s}; \mathbf{t})),\] where
 \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
 parameter and \(d_{\eta}\) is the number of parameter of \(\eta\).
-Sometimes, to account for the monthly variation in the data, the fitting
+
+Considerations need to be given on selecting the best distribution to
+fit, Various discussion on the fitting methods and goodness of fit.
+Potentially, more care about how much difference each combination give
+in terms of the index value.
+
+sometimes, to account for the monthly variation in the data, the fitting
 is done separately for each month and the noramlising step can be
@@
 \end{itemize}
+
+\textbf{Normalising}
+
+This step maps the univariate series into a different scale, typically
+for the ease of comparison. For example, a normal scale, {[}0, 1{]},
+{[}0, 100{]} may be favored to report certain indexes. In drought
+indexes, i.e.~SPI or SPEI, the quantiles from the fitted distribution is
+converted into the normal scale via the normal reverse CDF function:
+\(\Phi^{-1}(.)\).
+
+Normalising is usually used in the end of the pipeline and its main
+difference from the scaling step is that here the change of scale also
+changes the distribution of the variable. While being commonly used,
+this step can get criticism from analysts for forcing the data into the
+decided scale, which can be either unnecessary or inaccurately
+exaggerate or downplay the outliers. Also, the use of normal scale needs
+to be interpreted with caution. Figure~\ref{fig-normalising} illustrates
+the normal density not being directly proportional to its probability of
+occurrence. This is concerning, especially at the extreme values, where
+a normal density of -5 and -4 has magnitudes of difference into its
+probability of occurrence (1 in 3,000,000 and 1 in 31,000,
+respectively).
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/fig-normalising-1.pdf}
+
+}
+
+\caption{\label{fig-normalising}scatterplot of quantile vs.~its density
+values.}
+
+\end{figure}
 

### 2022-11-28 paper/paper.qmd
section on distribution fit and tweak in normalising plot

@@
 
+Distribution fit can be seen as model fitting in its simplest term. It can be represented by
+
+<!-- With a probability model imposed,  -->
+
 $$F_{\eta}(x(\mathbf{s}; \mathbf{t})),$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
-
-Considerations need to be given on selecting the best distribution to fit, 
-Various discussion on the fitting methods and goodness of fit. 
-Potentially, more care about how much difference each combination give in terms of the index value.
-
-sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.
-
-- In some cases, the distibution and inverse normal can be multivariate, i.e. use copula.
+A distribution fit typically aims at finding the distribution that best fit the data.
+Analysts may start from a pool of candidate distributions with a chosen fitting method and goodness of fit measure. 
+While it is useful to find the ultimate best distribution to fits the data, from a probabilistic perspective, the fitting procedure itself has uncertainty associated with the data fed and parameter chosen.
+A reasonable alternative is to understand how much the index values can vary given different distributions, fitting methods, and goodness of fit tests, and whether these variations are negligible in a given application.
 
@@
 
-This step maps the univariate series into a different scale, typically for the ease of comparison.
-For example, a normal scale, [0, 1], [0, 100] may be favored to report certain indexes.
+This step maps the univariate series into a different scale, typically for the ease of comparison across regions.
+For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes.
 In drought indexes, i.e.
@@
 @fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
-This is concerning, especially at the extreme values, where a normal density of -5 and -4 has magnitudes of difference into its probability of occurrence (1 in 3,000,000 and 1 in 31,000, respectively).
+This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.
 
 ```{r fig-normalising}
-#| fig-cap: "scatterplot of quantile vs. its density values."
+#| fig-cap: "Scatterplot of normal quantiles against their density values.
+THree tail density values are highlighted with its probability of occurence labelled.
+Probability is calculated assuming monthly data: with a density of -2, the probability of occurrence is 1/pnorm(-2)/12 = 4 years.
+The non-linear relationship between the two quantities suggests normalised indexes need to be interpreted with caution since a slight change in the tail distribution can result in magnitudes of difference in its probability of occurrence."
+#| fig-width: 9
 library(tidyverse)
-all <- tibble(x = seq(-5, 5, 0.01), y = pnorm(x)) 
-dt <- tibble(x = c(-5, -4), y = pnorm(x)) 
-label <- tibble(x = c(-5, -4), y = pnorm(x), 
-                label = c("a chance \nof 1 in \n 3,488,556", "a chance \nof 1 in \n 31,574"))
+all <- tibble(x = seq(-3, 3, 0.01), y = pnorm(x)) 
+dt <- tibble(x = c(-3, -2.5, -2), y = pnorm(x)) 
+label <- tibble(x = c(-3, -2.5, -2), y = pnorm(x), 
+                label = c("once in 62 years", "once in 13 years", "once in 4 years"))
 
@@
   ggplot() + 
-  geom_point(aes(x= x, y =y)) + 
-  geom_line(data = all, aes(x = x, y =y)) + 
+  geom_point(aes(x= x, y = y)) + 
+  geom_line(data = all, aes(x = x, y = y)) + 
   ggrepel::geom_label_repel(
     data = label, aes(x = x, y = y, label =label),
-    nudge_y = 0.2, arrow = arrow(length = unit(0.08, "inches")), min.segment.length = 0) + 
-  scale_x_continuous(breaks = seq(-5, 5, 1)) + 
+    nudge_y = 0.2, nudge_x = 0.1,
+    arrow = arrow(length = unit(0.08, "inches")), min.segment.length = 0) +
+  scale_x_continuous(breaks = seq(-3, 3, 1)) + 
   xlab("Density") + 

### 2022-11-28 paper/paper.tex
section on distribution fit and tweak in normalising plot

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners, enhanced, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 
+Distribution fit can be seen as model fitting in its simplest term. It
+can be represented by
+
 \[F_{\eta}(x(\mathbf{s}; \mathbf{t})),\] where
 \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
-parameter and \(d_{\eta}\) is the number of parameter of \(\eta\).
-
-Considerations need to be given on selecting the best distribution to
-fit, Various discussion on the fitting methods and goodness of fit.
-Potentially, more care about how much difference each combination give
-in terms of the index value.
-
-sometimes, to account for the monthly variation in the data, the fitting
-is done separately for each month and the noramlising step can be
-written as \(\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]\), where
-\(j^*\) is all the indexes that satisfy \(j^* \mod 12 = m\) for each
-\(m = 0, 1,\cdots, 11\).
-
-\begin{itemize}
-\tightlist
-\item
-  In some cases, the distibution and inverse normal can be multivariate,
-  i.e.~use copula.
-\end{itemize}
+parameter and \(d_{\eta}\) is the number of parameter of \(\eta\). A
+distribution fit typically aims at finding the distribution that best
+fit the data. Analysts may start from a pool of candidate distributions
+with a chosen fitting method and goodness of fit measure. While it is
+useful to find the ultimate best distribution to fits the data, from a
+probabilistic perspective, the fitting procedure itself has uncertainty
+associated with the data fed and parameter chosen. A reasonable
+alternative is to understand how much the index values can vary given
+different distributions, fitting methods, and goodness of fit tests, and
+whether these variations are negligible in a given application.
 
@@
 This step maps the univariate series into a different scale, typically
-for the ease of comparison. For example, a normal scale, {[}0, 1{]},
-{[}0, 100{]} may be favored to report certain indexes. In drought
-indexes, i.e.~SPI or SPEI, the quantiles from the fitted distribution is
-converted into the normal scale via the normal reverse CDF function:
-\(\Phi^{-1}(.)\).
+for the ease of comparison across regions. For example, a normal scale,
+{[}0, 1{]}, or {[}0, 100{]} may be favored for reporting certain
+indexes. In drought indexes, i.e.~SPI or SPEI, the quantiles from the
+fitted distribution is converted into the normal scale via the normal
+reverse CDF function: \(\Phi^{-1}(.)\).
 
@@
 the normal density not being directly proportional to its probability of
-occurrence. This is concerning, especially at the extreme values, where
-a normal density of -5 and -4 has magnitudes of difference into its
-probability of occurrence (1 in 3,000,000 and 1 in 31,000,
-respectively).
+occurrence. This is concerning, especially at the extreme values, since
+a small difference in the tail density can have magnitudes of difference
+in its probability of occurrence.
 
@@
 
-\caption{\label{fig-normalising}scatterplot of quantile vs.~its density
-values.}
+\caption{\label{fig-normalising}Scatterplot of normal quantiles against
+their density values. THree tail density values are highlighted with its
+probability of occurence labelled. Probability is calculated assuming
+monthly data: with a density of -2, the probability of occurrence is
+1/pnorm(-2)/12 = 4 years. The non-linear relationship between the two
+quantities suggests normalised indexes need to be interpreted with
+caution since a slight change in the tail distribution can result in
+magnitudes of difference in its probability of occurrence.}
 

### 2022-11-28 paper/paper.qmd
small tweak on simplification and mark TODOs

@@
 
-**Spatial aggregation**
+**Spatial aggregation** [TODO]
 
@@
 
-**Variable transformation **
+**Variable transformation ** [TODO]
 
-Variable transformation and scaling, in the next section, are both variable pre-processing steps to prepare the data with desire statistical property to work with.
-This can be a more stablised variance, a normal distribution that satisfies modelling assumptions, or a data range that required by the algorithms.
+Variable transformation, and scaling in the next section, are both pre-processing steps that prepare the data with desire statistical property to work with.
+Some desirable property to work with can be a stable variance, a normal distribution that satisfies modelling assumptions, or a data range that required by the algorithms.
 Variable tranformation is a general notion of a functional transformation on the variable: $$h_{\tau}(x(\mathbf{s};\mathbf{t})),$$ where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
@@
 
-**Scaling**
+**Scaling** [TODO]
 
-While scaling can be seen as a specific type of variable tranformation, it is separated into own step to make the scaling operation explicit in the pipeline.
-The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
+While scaling can be seen as a specific type of variable tranformation, it is separated into own step to make it explicit in the pipeline.
+The key difference between the two steps is that variable transformation typically changes the shape of the data, while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
 For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
@@
 
-**Dimension reduction**
+**Dimension reduction** [TODO]
 
@@
 
-**Benchmarking**
+**Benchmarking** [TODO]
 
@@
 
-In public communication, indexes are often delivered in categorical grades, along with its underlying numerical values.
-The simplification step, sometimes can also be called discretisation, prescribes how the continuous index values are converted into the discrete grades.
-This process can be written with the piece-wise function:
+In public communication, the index values are usually accompanied by a categorical grade.
+The categorised grades is an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes.
+The mapping from continuous index values to the discrete grades is called simplification in the pipeline and it can be written as a piece-wise function:
 
@@
 
-where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the threshold value in each category.
+where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the thresholds for each category.
 In SPI, drought are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$.
 In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.
-
-Discretise the continuous index into a few labelled categories. For communicating the severity of natural hazard to general public.
-
-uniform workflow to work with index construction.
-
--   illustration
--   math notation
--   benefit of the pipeline approach
-    -   index diagnostic
-    -   uncertainty
 

### 2022-11-28 paper/paper.tex
small tweak on simplification and mark TODOs

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\textbf{Spatial aggregation}
+\textbf{Spatial aggregation} {[}TODO{]}
 
@@
 
-\textbf{Variable transformation }
+\textbf{Variable transformation } {[}TODO{]}
 
-Variable transformation and scaling, in the next section, are both
-variable pre-processing steps to prepare the data with desire
-statistical property to work with. This can be a more stablised
-variance, a normal distribution that satisfies modelling assumptions, or
-a data range that required by the algorithms. Variable tranformation is
-a general notion of a functional transformation on the variable:
-\[h_{\tau}(x(\mathbf{s};\mathbf{t})),\] where
+Variable transformation, and scaling in the next section, are both
+pre-processing steps that prepare the data with desire statistical
+property to work with. Some desirable property to work with can be a
+stable variance, a normal distribution that satisfies modelling
+assumptions, or a data range that required by the algorithms. Variable
+tranformation is a general notion of a functional transformation on the
+variable: \[h_{\tau}(x(\mathbf{s};\mathbf{t})),\] where
 \(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in the
@@
 
-\textbf{Scaling}
+\textbf{Scaling} {[}TODO{]}
 
 While scaling can be seen as a specific type of variable tranformation,
-it is separated into own step to make the scaling operation explicit in
-the pipeline. The key difference between the two steps is that variable
-transformation typically changes the shape of the data while scaling
-only changes the data scale and can usually be written in the form of
+it is separated into own step to make it explicit in the pipeline. The
+key difference between the two steps is that variable transformation
+typically changes the shape of the data, while scaling only changes the
+data scale and can usually be written in the form of
 \[[x(s_i;t_j) - \alpha]/\gamma.\] For example, a z-score standardisation
@@
 
-\textbf{Dimension reduction}
+\textbf{Dimension reduction} {[}TODO{]}
 
@@
 
-\textbf{Benchmarking}
+\textbf{Benchmarking} {[}TODO{]}
 
@@
 
-In public communication, indexes are often delivered in categorical
-grades, along with its underlying numerical values. The simplification
-step, sometimes can also be called discretisation, prescribes how the
-continuous index values are converted into the discrete grades. This
-process can be written with the piece-wise function:
+In public communication, the index values are usually accompanied by a
+categorical grade. The categorised grades is an ordered set of
+descriptive words or colors to communicate the severity or guide the
+comprehension of the indexes. The mapping from continuous index values
+to the discrete grades is called simplification in the pipeline and it
+can be written as a piece-wise function:
 
@@
 where \(C_0, C_1,\cdots ,C_z\) are the categories and
-\(c_0, c_1, \cdots, c_z\) are the threshold value in each category. In
-SPI, drought are sorted into four categories: mild drought:
-\([-0.99, 0]\); moderate drought: \([-1.49, -1]\); severe drought:
-\([-1.99, -1.5]\), and extreme drought: \([-\infty, -2]\). In this case,
+\(c_0, c_1, \cdots, c_z\) are the thresholds for each category. In SPI,
+drought are sorted into four categories: mild drought: \([-0.99, 0]\);
+moderate drought: \([-1.49, -1]\); severe drought: \([-1.99, -1.5]\),
+and extreme drought: \([-\infty, -2]\). In this case,
 \(C_0, C_1, C_2, C_3\) are the drought categories: mild, moderate,
@@
 class.
-
-Discretise the continuous index into a few labelled categories. For
-communicating the severity of natural hazard to general public.
-
-uniform workflow to work with index construction.
-
-\begin{itemize}
-\tightlist
-\item
-  illustration
-\item
-  math notation
-\item
-  benefit of the pipeline approach
-
-  \begin{itemize}
-  \tightlist
-  \item
-    index diagnostic
-  \item
-    uncertainty
-  \end{itemize}
-\end{itemize}
 

### 2022-11-28 paper/paper.qmd
expand dimension reduction section

@@
 
-Sometimes, information from neighbouring spatial locations can also be borrowed when constructing indexes. 
+Sometimes
+
+Information from neighbouring spatial locations can also be borrowed when constructing indexes. 
 This involves interpolating data from observed locations to the unobserved and aggregating data into different levels.
@@
 
-collect ground and satellite image -> merge
+While many indexes are calculated independently on spatial locations
+
+satellite images with different resolutions
+matching ground and satellite images data
+
 <!-- $x(s_i^\prime;t_j) = \sum_{i \in \Omega}x(s_i; t_j)$ -->
@@
  <!-- - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level. -->
-
-\newpage
 
@@
 
-\newpage
+**Dimension reduction**
 
-**Dimension reduction** [TODO]
+When constructing indexes from multivariate information, dimension reduction methods combine those information into a univariate series.
+In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate): $$x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),$$ where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$.
+The most commonly used dimension reduction technique is Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science.
+It can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the data variance given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
 
-While multivariate information can be collected on a concept, they need to be summarised into a single number for the final index.
-Dimension reduction step $$x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),$$ where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$, reduces the variable dimension from $P$ to $P^*$.
-The most commonly used dimension reduction technique Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science, can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the data variance given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
-
-weighting
+The PCA is a special case of weighting, where individual variables are summed up in a linear combination with potential restrictions imposed on the weight coefficients.
+These coefficients may be simple mathematical constructions that give equal weight to each variable or can be derived with specialised knowledge by experts in the field.
+While suggested weights can indicate norms adored by practitioners, analysts should be given the flexibility to experiment with different combinations when constructing indexes. 
+This could help understand index behavior from its sensitivity to the variables and suggest alternative weights that better suit specific tasks.
 
@@
 
-*modell fit? *
+*model fit? *
 
 Distribution fit can be seen as model fitting in its simplest term. It can be represented by 
-
 <!-- With a probability model imposed,  -->
@@
 SPI or SPEI, the quantiles from the fitted distribution is converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$.
-
 Normalising is usually used in the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable.

### 2022-11-28 paper/paper.tex
expand dimension reduction section

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-Sometimes, information from neighbouring spatial locations can also be
-borrowed when constructing indexes. This involves interpolating data
-from observed locations to the unobserved and aggregating data into
-different levels. In the pipeline, spatial aggregation can be written as
+Sometimes
+
+Information from neighbouring spatial locations can also be borrowed
+when constructing indexes. This involves interpolating data from
+observed locations to the unobserved and aggregating data into different
+levels. In the pipeline, spatial aggregation can be written as
 \[g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),\] where
@@
 
-collect ground and satellite image -\textgreater{} merge
+While many indexes are calculated independently on spatial locations
 
-\newpage
+satellite images with different resolutions matching ground and
+satellite images data
 
@@
 
-\newpage
+\textbf{Dimension reduction}
 
-\textbf{Dimension reduction} {[}TODO{]}
-
-While multivariate information can be collected on a concept, they need
-to be summarised into a single number for the final index. Dimension
-reduction step
+When constructing indexes from multivariate information, dimension
+reduction methods combine those information into a univariate series. In
+the pipeline, dimension reduction includes methods that take
+multivariate inputs and output the data in a lower dimension (often
+univariate):
 \[x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),\]
-where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\), reduces
-the variable dimension from \(P\) to \(P^*\). The most commonly used
-dimension reduction technique Principal Component Analysis (PCA), also
-called Empirical Orthogonal Function (EOF) in earth science, can be
-written as
+where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\) reduce the
+variable dimension from \(P\) to \(P^*\). The most commonly used
+dimension reduction technique is Principal Component Analysis (PCA),
+also called Empirical Orthogonal Function (EOF) in earth science. It can
+be written as
 \(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
@@
 
-weighting
+The PCA is a special case of weighting, where individual variables are
+summed up in a linear combination with potential restrictions imposed on
+the weight coefficients. These coefficients may be simple mathematical
+constructions that give equal weight to each variable or can be derived
+with specialised knowledge by experts in the field. While suggested
+weights can indicate norms adored by practitioners, analysts should be
+given the flexibility to experiment with different combinations when
+constructing indexes. This could help understand index behavior from its
+sensitivity to the variables and suggest alternative weights that better
+suit specific tasks.
 
@@
 
-\emph{modell fit? }
+\emph{model fit? }
 
@@
 fitted distribution is converted into the normal scale via the normal
-reverse CDF function: \(\Phi^{-1}(.)\).
-
-Normalising is usually used in the end of the pipeline and its main
-difference from the scaling step is that here the change of scale also
-changes the distribution of the variable. While being commonly used,
-this step can get criticism from analysts for forcing the data into the
-decided scale, which can be either unnecessary or inaccurately
-exaggerate or downplay the outliers. Also, the use of normal scale needs
-to be interpreted with caution. Figure~\ref{fig-normalising} illustrates
-the normal density not being directly proportional to its probability of
-occurrence. This is concerning, especially at the extreme values, since
-a small difference in the tail density can have magnitudes of difference
-in its probability of occurrence.
+reverse CDF function: \(\Phi^{-1}(.)\). Normalising is usually used in
+the end of the pipeline and its main difference from the scaling step is
+that here the change of scale also changes the distribution of the
+variable. While being commonly used, this step can get criticism from
+analysts for forcing the data into the decided scale, which can be
+either unnecessary or inaccurately exaggerate or downplay the outliers.
+Also, the use of normal scale needs to be interpreted with caution.
+Figure~\ref{fig-normalising} illustrates the normal density not being
+directly proportional to its probability of occurrence. This is
+concerning, especially at the extreme values, since a small difference
+in the tail density can have magnitudes of difference in its probability
+of occurrence.
 

### 2022-11-28 paper/paper.qmd
more on the spatial aggregation section

@@
 knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)
-
+library(tidyverse)
+library(latex2exp)
+library(patchwork)
 ```
@@
 In SPI, monthly precipitation is aggregated by a time scale parameter $k$: $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),$ where $j^\prime$ is the new time index after the aggregation.
-In this notation, each spatial location is seaprately aggregated and precipitation is ummed from $k$ month back, $j^\prime - k + 1$, to the current, $j^\prime$ period to create the aggregated series, indexed by $j^\prime$.
+In this notation, each spatial location is seaprately aggregated and precipitation is ummed from $k$ month back, $j^\prime - k + 1$, to the current period, $j^\prime$, to create the aggregated series, indexed by $j^\prime$.
 
@@
 
-**Spatial aggregation** [TODO]
+**Spatial aggregation**
 
-Sometimes
+While many indexes calculate independently on spatial locations, borrowing spatial information can happen at the unobservable locations.
+Also, variables can be collected in different formats: some from satellite imageries as raster data, potentially at different resolutions, some from ground stations as vector data.
+These spatial data need to be merged into a uniform unit to construct indexes.
+In the pipeline, spatial aggregation includes operations on the spatial dimension of the data and can be written as $$g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),$$ where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.
+When one of the variables collected is only available in a coarser resolution, $i^\prime$, analysts may choose to reduce other variables with finer resolution, $i$, to match the coarser resolution, as in @fig-spatial-agg.
+This is a spatial aggregation step and aggregating using the mean can be written as: $x(s_{i^\prime};t_{j}) = \bar{x}(s_{i \in i^\prime}; t_j),$, where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid.
+*(Feel like the figure here is more of a visual illustration of the equation without enough further implication =.=)*
 
-Information from neighbouring spatial locations can also be borrowed when constructing indexes. 
-This involves interpolating data from observed locations to the unobserved and aggregating data into different levels.
-In the pipeline, spatial aggregation can be written as $$g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),$$ where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.
-[GIVEN AN EXAMPLE]
+```{r fig-spatial-agg}
+#| fig-cap: "Illustration of aggregating raster data with finer resolution (right) into coarser resolution (left)." 
+#| fig-width: 10
+#| fig-height: 3
+dt <- crossing(y = 1:2, x = 1:3) %>% 
+  arrange(-y) %>% 
+  mutate(label = as.factor(1:6))
+okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#CC79A7")
 
-While many indexes are calculated independently on spatial locations
+p1 <- dt %>% 
+  ggplot(aes(x = x, y = y)) + 
+  geom_tile(aes(fill = label), color = "black") + 
+  geom_text(aes(label = label)) + 
+  scale_fill_manual(values = okabe) + 
+  theme_void() + 
+  theme(aspect.ratio = 2/3)
 
-satellite images with different resolutions
-matching ground and satellite images data
+dt2 <- crossing(y = 1:6, x = 1:9) %>% 
+  arrange(-y) %>% 
+  mutate(x1 = (x-1) %/%3 + 1, y1 = (y-1) %/%3 + 1) %>% 
+  left_join(dt, by = c("x1" = "x", "y1" = "y")) %>% 
+  mutate(id = row_number())
 
-<!-- $x(s_i^\prime;t_j) = \sum_{i \in \Omega}x(s_i; t_j)$ -->
+p2 <- dt2 %>% 
+  ggplot(aes(x = x, y = y)) + 
+  geom_tile(aes(fill = label), color = "black") + 
+  geom_text(aes(label = id)) + 
+  scale_fill_manual(values = okabe) + 
+  theme_void() + 
+  theme(aspect.ratio = 2/3)
 
-<!-- Another common use of spatial processing is to integrate satellite imagaries with ground measures, also merge satellite imagaries with different resolutions.  -->
-
- <!-- - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level. -->
+(p1 | p2) & theme(legend.position = "none")
+```
 
@@
 #| fig-width: 12
-library(tidyverse)
-library(latex2exp)
 set.seed(123)
@@
 
-Distribution fit can be seen as model fitting in its simplest term. It can be represented by 
 <!-- With a probability model imposed,  -->
-
-$$F_{\eta}(x(\mathbf{s}; \mathbf{t})),$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
+Distribution fit can be seen as model fitting in its simplest term.
+It can be represented by $$F_{\eta}(x(\mathbf{s}; \mathbf{t})),$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
 A distribution fit typically aims at finding the distribution that best fit the data.

### 2022-11-28 paper/paper.tex
more on the spatial aggregation section

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 precipitation is ummed from \(k\) month back, \(j^\prime - k + 1\), to
-the current, \(j^\prime\) period to create the aggregated series,
+the current period, \(j^\prime\), to create the aggregated series,
 indexed by \(j^\prime\).
@@
 
-\textbf{Spatial aggregation} {[}TODO{]}
+\textbf{Spatial aggregation}
 
-Sometimes
-
-Information from neighbouring spatial locations can also be borrowed
-when constructing indexes. This involves interpolating data from
-observed locations to the unobserved and aggregating data into different
-levels. In the pipeline, spatial aggregation can be written as
+While many indexes calculate independently on spatial locations,
+borrowing spatial information can happen at the unobservable locations.
+Also, variables can be collected in different formats: some from
+satellite imageries as raster data, potentially at different
+resolutions, some from ground stations as vector data. These spatial
+data need to be merged into a uniform unit to construct indexes. In the
+pipeline, spatial aggregation includes operations on the spatial
+dimension of the data and can be written as
 \[g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),\] where
@@
 operation parameters and \(d_{\theta}\) is the number of parameter of
-\(\theta\). {[}GIVEN AN EXAMPLE{]}
+\(\theta\). When one of the variables collected is only available in a
+coarser resolution, \(i^\prime\), analysts may choose to reduce other
+variables with finer resolution, \(i\), to match the coarser resolution,
+as in Figure~\ref{fig-spatial-agg}. This is a spatial aggregation step
+and aggregating using the mean can be written as:
+\(x(s_{i^\prime};t_{j}) = \bar{x}(s_{i \in i^\prime}; t_j),\), where
+\(i \in i^\prime\) includes all the cells from the finer resolution in
+the coarser grid. \emph{(Feel like the figure here is more of a visual
+illustration of the equation without enough further implication =.=)}
 
-While many indexes are calculated independently on spatial locations
+\begin{figure}
 
-satellite images with different resolutions matching ground and
-satellite images data
+{\centering \includegraphics{paper_files/figure-pdf/fig-spatial-agg-1.pdf}
+
+}
+
+\caption{\label{fig-spatial-agg}Illustration of aggregating raster data
+with finer resolution (right) into coarser resolution (left).}
+
+\end{figure}
 
@@
 Distribution fit can be seen as model fitting in its simplest term. It
-can be represented by
-
-\[F_{\eta}(x(\mathbf{s}; \mathbf{t})),\] where
+can be represented by \[F_{\eta}(x(\mathbf{s}; \mathbf{t})),\] where
 \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution

### 2022-11-28 paper/paper.qmd
variable tranformation & scaling

@@
 
-**Variable transformation ** [TODO]
+**Variable transformation **
 
-Variable transformation, and scaling in the next section, are both pre-processing steps that prepare the data with desire statistical property to work with.
-Some desirable property to work with can be a stable variance, a normal distribution that satisfies modelling assumptions, or a data range that required by the algorithms.
+Variable transformation, and scaling in the next section, are both pre-processing steps to create variables with good statistical properties.
+These properties include a stable variance, normal distribution, or a certain scale required by some algorithm further down the pipeline.
 Variable tranformation is a general notion of a functional transformation on the variable: $$h_{\tau}(x(\mathbf{s};\mathbf{t})),$$ where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
-Typical examples of variable transformation are log, quadratic, square root, and box-cox transformation.
+Transformation is needed for data that are highly skewed and some common transformations includes log, quadratic, and square root transformation.
 
-**Scaling** [TODO]
+**Scaling**
 
-While scaling can be seen as a specific type of variable tranformation, it is separated into own step to make it explicit in the pipeline.
+While scaling can be seen as a specific type of variable tranformation, it is separated into own step to make the step explicit in the pipeline.
 The key difference between the two steps is that variable transformation typically changes the shape of the data, while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
@@
 @fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step.
-All the scaling operations in green results in the same shape as the original data.
+All the scaling operations in green results in the same distribution as the original data.
 
@@
 ```{r fig-scale-var-trans-compare}
-#| fig-cap: "Comparison of scaling and variable transformation."
+#| fig-cap: "Comparison of operations in scaling (green) and variable transformation (orange) steps in free scale.
+Variables after the scaling operations have the same distribution as the origin, while the distribution changes after variable transformation.
+"
 #| fig-height: 8
@@
 Benchmarking sets a constant value to allow the constructed index to be compared across time.
-Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed, could be a constant or a function of the data, i.e. mean.
+Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed. A benchmark value could be a constant or a function of the data, i.e. mean.
 

### 2022-11-28 paper/paper.tex
variable tranformation & scaling

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\textbf{Variable transformation } {[}TODO{]}
+\textbf{Variable transformation }
 
 Variable transformation, and scaling in the next section, are both
-pre-processing steps that prepare the data with desire statistical
-property to work with. Some desirable property to work with can be a
-stable variance, a normal distribution that satisfies modelling
-assumptions, or a data range that required by the algorithms. Variable
-tranformation is a general notion of a functional transformation on the
-variable: \[h_{\tau}(x(\mathbf{s};\mathbf{t})),\] where
-\(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in the
-transformation, if any, and \(d_{\tau}\) is the number of parameter of
-\(\tau\). Typical examples of variable transformation are log,
-quadratic, square root, and box-cox transformation.
+pre-processing steps to create variables with good statistical
+properties. These properties include a stable variance, normal
+distribution, or a certain scale required by some algorithm further down
+the pipeline. Variable tranformation is a general notion of a functional
+transformation on the variable: \[h_{\tau}(x(\mathbf{s};\mathbf{t})),\]
+where \(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in
+the transformation, if any, and \(d_{\tau}\) is the number of parameter
+of \(\tau\). Transformation is needed for data that are highly skewed
+and some common transformations includes log, quadratic, and square root
+transformation.
 
-\textbf{Scaling} {[}TODO{]}
+\textbf{Scaling}
 
 While scaling can be seen as a specific type of variable tranformation,
-it is separated into own step to make it explicit in the pipeline. The
-key difference between the two steps is that variable transformation
+it is separated into own step to make the step explicit in the pipeline.
+The key difference between the two steps is that variable transformation
 typically changes the shape of the data, while scaling only changes the
@@
 operation is a variable transformation or a scaling step. All the
-scaling operations in green results in the same shape as the original
-data.
+scaling operations in green results in the same distribution as the
+original data.
 
@@
 
-\caption{\label{fig-scale-var-trans-compare}Comparison of scaling and
-variable transformation.}
+\caption{\label{fig-scale-var-trans-compare}Comparison of operations in
+scaling (green) and variable transformation (orange) steps in free
+scale. Variables after the scaling operations have the same distribution
+as the origin, while the distribution changes after variable
+transformation.}
 

### 2022-11-30 paper/paper.qmd
gramma check

@@
 An overview of the pipeline is given in @fig-pipeline-steps to illustrate the construction from raw data to the final indexes.
-The pipeline includes modules for operations in the spatial, temporal, and multivariate aspect of the data: temporal processing, spatial aggregation, variable transformation, scaling, dimension reduction, and normalising, as well as modules for comparing and communicating indexes (benchmarking and simplification).
-These steps are not required to be arranged in a fixed order and not all the steps have to appear when constructing each index. 
-While the pipeline starts with the raw data, steps prior to this, i.e.
-define a useful concept to construct the index, collect high quality data relevant for measuring the concept, are also cricial to the sucess of an index.
-The framework proposed in this paper assumes the objective of an index has been defined and relevant variables have been collected and quality controlled.
+The pipeline includes modules for operations in the spatial, temporal, and multivariate aspects of the data as well as modules for comparing and communicating indexes (benchmarking and simplification).
+Analysts are free to select the modules they need and arrange them in the order they see fit to construct indexes.
+While the starting point of the pipeline is raw data, there are steps prior to this that are crucial to the success of an index. 
+For example, the defined index needs to be useful for measuring the concept of interest and variables need to be collected from reliable sources with proper quality control.
+
+The notation used in this section will be first introduced before elaborating on each pipeline module.
+Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspects: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
+When more than one variable is involved, the multivariate data can also be written as: $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$.
 
@@
 #| fig-align: center
-#| fig-cap: "Diagram of pipeline steps for index construction."
+#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
 #| out-height: 90%
@@
 
-**Notation**
-
-<!-- *Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\mathbf{s}$ and $\mathbf{t}$ to depend on $p$* -->
-
-<!-- *Initially my thought was to raw data only after aligning the resolution: Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem.
-After processing them and align them together they become the "raw data"* -->
-
-<!-- Once variables are collected, they can have different spatial resolution and temporal granularity.  -->
-
-Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspect: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
-When more than one variable are involved, the multivariate data can also be written as: $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$.
-
-  <!-- * all the different choice in each step will affect the final index series ot make decision, i.e.
-$k$ in temporal aggregation, to assess drought in different scale: short-term, mid-term, long-term.
--->
-
 **Temporal processing**
@@
 The construction of an index sometimes needs to consider information from neighbouring time periods.
-Temporal processing is a general operator on the time dimension of the data in the form of  $$f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),$$ where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
-An typical example of temporal processing is aggregation in the drought index SPI to measure the lack of precipitation for meteorological drought.
+The temporal processing is a general operator on the time dimension of the data in the form of  $$f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),$$ where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
+A typical example of temporal processing is aggregation in the drought index SPI to measure the lack of precipitation for meteorological drought.
 In SPI, monthly precipitation is aggregated by a time scale parameter $k$: $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),$ where $j^\prime$ is the new time index after the aggregation.
-In this notation, each spatial location is seaprately aggregated and precipitation is ummed from $k$ month back, $j^\prime - k + 1$, to the current period, $j^\prime$, to create the aggregated series, indexed by $j^\prime$.
+In this notation, each spatial location is separately aggregated and precipitation is ummed from $k$ month back, $j^\prime - k + 1$, to the current period, $j^\prime$, to create the aggregated series, indexed by $j^\prime$.
 
-The choice of time scales parameter $k$ can result in variation in the calculated index values: a small $k$ of 3 or 6 month produces the index more sensitive to individual months, while a large $k$ of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives dryness information relative to the long term condition.
-As will be shown in section [SECTION EXAMPLE], this variation may even lead to conflicting conclusions on the dry/wet condition of the area, highlighting the importance to account for index uncertainty when interpreting index values for decision making.
+The choice of time scales parameter $k$ can result in variation in the calculated index values: a small $k$ of 3 or 6 months produces the index more sensitive to individual months, while a large $k$ of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives dryness information relative to the long term condition.
+As will be shown in section [SECTION EXAMPLE], this variation may even lead to conflicting conclusions on the dry/wet condition of the area, highlighting the importance to account for index uncertainty when interpreting index values for decision-making.
 
@@
 
-While many indexes calculate independently on spatial locations, borrowing spatial information can happen at the unobservable locations.
-Also, variables can be collected in different formats: some from satellite imageries as raster data, potentially at different resolutions, some from ground stations as vector data.
+While many indexes are calculated independently on spatial locations, borrowing spatial information can happen at unobservable locations.
+Also, variables can be collected in different formats: some from satellite imageries as raster data, potentially at different resolutions, and some from ground stations as vector data.
 These spatial data need to be merged into a uniform unit to construct indexes.
@@
 These properties include a stable variance, normal distribution, or a certain scale required by some algorithm further down the pipeline.
-Variable tranformation is a general notion of a functional transformation on the variable: $$h_{\tau}(x(\mathbf{s};\mathbf{t})),$$ where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
-Transformation is needed for data that are highly skewed and some common transformations includes log, quadratic, and square root transformation.
+Variable transformation is a general notion of a functional transformation on the variable: $$h_{\tau}(x(\mathbf{s};\mathbf{t})),$$ where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation if any, and $d_{\tau}$ is the number of parameter of $\tau$.
+Transformation is needed for data that are highly skewed and some common transformations include log, quadratic, and square root transformation.
 
@@
 
-While scaling can be seen as a specific type of variable tranformation, it is separated into own step to make the step explicit in the pipeline.
-The key difference between the two steps is that variable transformation typically changes the shape of the data, while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
+While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
+The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
 For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
 @fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step.
-All the scaling operations in green results in the same distribution as the original data.
+All the scaling operations in green result in the same distribution as the original data.
 
@@
 
-When constructing indexes from multivariate information, dimension reduction methods combine those information into a univariate series.
+When constructing indexes from multivariate information, dimension reduction methods combine that information into a univariate series.
 In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate): $$x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),$$ where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$.
@@
 <!-- With a probability model imposed,  -->
-Distribution fit can be seen as model fitting in its simplest term.
+Distribution fit can be seen as the model fitting in its simplest term.
 It can be represented by $$F_{\eta}(x(\mathbf{s}; \mathbf{t})),$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
-A distribution fit typically aims at finding the distribution that best fit the data.
+A distribution fit typically aims at finding the distribution that best fits the data.
 Analysts may start from a pool of candidate distributions with a chosen fitting method and goodness of fit measure. 
-While it is useful to find the ultimate best distribution to fits the data, from a probabilistic perspective, the fitting procedure itself has uncertainty associated with the data fed and parameter chosen.
+While it is useful to find the ultimate best distribution to fits the data, from a probabilistic perspective, the fitting procedure itself has an uncertainty associated with the data fed and the parameter chosen.
 A reasonable alternative is to understand how much the index values can vary given different distributions, fitting methods, and goodness of fit tests, and whether these variations are negligible in a given application.
@@
 
-This step maps the univariate series into a different scale, typically for the ease of comparison across regions.
+This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
 For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
-In drought indexes, i.e.
-SPI or SPEI, the quantiles from the fitted distribution is converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$.
-Normalising is usually used in the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable.
+In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. 
+Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. 
 While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers. 
-Also, the use of normal scale needs to be interpreted with caution.
+Also, the use of a normal scale needs to be interpreted with caution. 
 @fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
@@
 In public communication, the index values are usually accompanied by a categorical grade.
-The categorised grades is an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes.
+The categorised grades are an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes.
 The mapping from continuous index values to the discrete grades is called simplification in the pipeline and it can be written as a piece-wise function:
@@
 where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the thresholds for each category.
-In SPI, drought are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$.
+In SPI, droughts are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$.
 In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.

### 2022-11-30 paper/paper.tex
gramma check

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 pipeline includes modules for operations in the spatial, temporal, and
-multivariate aspect of the data: temporal processing, spatial
-aggregation, variable transformation, scaling, dimension reduction, and
-normalising, as well as modules for comparing and communicating indexes
-(benchmarking and simplification). These steps are not required to be
-arranged in a fixed order and not all the steps have to appear when
-constructing each index. While the pipeline starts with the raw data,
-steps prior to this, i.e.~define a useful concept to construct the
-index, collect high quality data relevant for measuring the concept, are
-also cricial to the sucess of an index. The framework proposed in this
-paper assumes the objective of an index has been defined and relevant
-variables have been collected and quality controlled.
+multivariate aspects of the data as well as modules for comparing and
+communicating indexes (benchmarking and simplification). Analysts are
+free to select the modules they need and arrange them in the order they
+see fit to construct indexes. While the starting point of the pipeline
+is raw data, there are steps prior to this that are crucial to the
+success of an index. For example, the defined index needs to be useful
+for measuring the concept of interest and variables need to be collected
+from reliable sources with proper quality control.
+
+The notation used in this section will be first introduced before
+elaborating on each pipeline module. Let
+\(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with spatial,
+temporal, and multivariate aspects: the spatial dimension
+\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
+space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
+temporal dimension \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
+defined in the 1D space:
+\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). When more than
+one variable is involved, the multivariate data can also be written as:
+\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\).
 
@@
 \caption{\label{fig-pipeline-steps}Diagram of pipeline steps for index
-construction.}
+construction. will need to be updated with better design and the
+distribution fitting step.}
 
 \end{figure}
-
-\textbf{Notation}
-
-Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with
-spatial, temporal, and multivariate aspect: the spatial dimension
-\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
-space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
-temporal dimension \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
-defined in the 1D space:
-\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). When more than
-one variable are involved, the multivariate data can also be written as:
-\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\).
 
@@
 The construction of an index sometimes needs to consider information
-from neighbouring time periods. Temporal processing is a general
+from neighbouring time periods. The temporal processing is a general
 operator on the time dimension of the data in the form of
@@
 associated with the temporal operation and \(d_{\psi}\) is the number of
-parameter of \(\psi\). An typical example of temporal processing is
+parameter of \(\psi\). A typical example of temporal processing is
 aggregation in the drought index SPI to measure the lack of
@@
 where \(j^\prime\) is the new time index after the aggregation. In this
-notation, each spatial location is seaprately aggregated and
+notation, each spatial location is separately aggregated and
 precipitation is ummed from \(k\) month back, \(j^\prime - k + 1\), to
@@
 The choice of time scales parameter \(k\) can result in variation in the
-calculated index values: a small \(k\) of 3 or 6 month produces the
+calculated index values: a small \(k\) of 3 or 6 months produces the
 index more sensitive to individual months, while a large \(k\) of 24 or
@@
 highlighting the importance to account for index uncertainty when
-interpreting index values for decision making.
+interpreting index values for decision-making.
 
@@
 
-While many indexes calculate independently on spatial locations,
-borrowing spatial information can happen at the unobservable locations.
+While many indexes are calculated independently on spatial locations,
+borrowing spatial information can happen at unobservable locations.
 Also, variables can be collected in different formats: some from
 satellite imageries as raster data, potentially at different
-resolutions, some from ground stations as vector data. These spatial
+resolutions, and some from ground stations as vector data. These spatial
 data need to be merged into a uniform unit to construct indexes. In the
@@
 distribution, or a certain scale required by some algorithm further down
-the pipeline. Variable tranformation is a general notion of a functional
-transformation on the variable: \[h_{\tau}(x(\mathbf{s};\mathbf{t})),\]
-where \(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in
-the transformation, if any, and \(d_{\tau}\) is the number of parameter
-of \(\tau\). Transformation is needed for data that are highly skewed
-and some common transformations includes log, quadratic, and square root
+the pipeline. Variable transformation is a general notion of a
+functional transformation on the variable:
+\[h_{\tau}(x(\mathbf{s};\mathbf{t})),\] where
+\(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in the
+transformation if any, and \(d_{\tau}\) is the number of parameter of
+\(\tau\). Transformation is needed for data that are highly skewed and
+some common transformations include log, quadratic, and square root
 transformation.
@@
 
-While scaling can be seen as a specific type of variable tranformation,
-it is separated into own step to make the step explicit in the pipeline.
-The key difference between the two steps is that variable transformation
-typically changes the shape of the data, while scaling only changes the
-data scale and can usually be written in the form of
+While scaling can be seen as a specific type of variable transformation,
+it is separated into its own step to make the step explicit in the
+pipeline. The key difference between the two steps is that variable
+transformation typically changes the shape of the data while scaling
+only changes the data scale and can usually be written in the form of
 \[[x(s_i;t_j) - \alpha]/\gamma.\] For example, a z-score standardisation
@@
 operation is a variable transformation or a scaling step. All the
-scaling operations in green results in the same distribution as the
+scaling operations in green result in the same distribution as the
 original data.
@@
 When constructing indexes from multivariate information, dimension
-reduction methods combine those information into a univariate series. In
+reduction methods combine that information into a univariate series. In
 the pipeline, dimension reduction includes methods that take
@@
 
-Distribution fit can be seen as model fitting in its simplest term. It
-can be represented by \[F_{\eta}(x(\mathbf{s}; \mathbf{t})),\] where
+Distribution fit can be seen as the model fitting in its simplest term.
+It can be represented by \[F_{\eta}(x(\mathbf{s}; \mathbf{t})),\] where
 \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
@@
 distribution fit typically aims at finding the distribution that best
-fit the data. Analysts may start from a pool of candidate distributions
+fits the data. Analysts may start from a pool of candidate distributions
 with a chosen fitting method and goodness of fit measure. While it is
 useful to find the ultimate best distribution to fits the data, from a
-probabilistic perspective, the fitting procedure itself has uncertainty
-associated with the data fed and parameter chosen. A reasonable
-alternative is to understand how much the index values can vary given
-different distributions, fitting methods, and goodness of fit tests, and
-whether these variations are negligible in a given application.
+probabilistic perspective, the fitting procedure itself has an
+uncertainty associated with the data fed and the parameter chosen. A
+reasonable alternative is to understand how much the index values can
+vary given different distributions, fitting methods, and goodness of fit
+tests, and whether these variations are negligible in a given
+application.
 
@@
 This step maps the univariate series into a different scale, typically
-for the ease of comparison across regions. For example, a normal scale,
+for ease of comparison across regions. For example, a normal scale,
 {[}0, 1{]}, or {[}0, 100{]} may be favored for reporting certain
 indexes. In drought indexes, i.e.~SPI or SPEI, the quantiles from the
-fitted distribution is converted into the normal scale via the normal
-reverse CDF function: \(\Phi^{-1}(.)\). Normalising is usually used in
+fitted distribution are converted into the normal scale via the normal
+reverse CDF function: \(\Phi^{-1}(.)\). Normalising is usually used at
 the end of the pipeline and its main difference from the scaling step is
@@
 either unnecessary or inaccurately exaggerate or downplay the outliers.
-Also, the use of normal scale needs to be interpreted with caution.
+Also, the use of a normal scale needs to be interpreted with caution.
 Figure~\ref{fig-normalising} illustrates the normal density not being
@@
 compared across time. Here we denote it with \(u[x(s_i, t_j)]\) where
-\(u\) is a scalar of interest in the index constructed, could be a
-constant or a function of the data, i.e.~mean.
+\(u\) is a scalar of interest in the index constructed. A benchmark
+value could be a constant or a function of the data, i.e.~mean.
 
@@
 In public communication, the index values are usually accompanied by a
-categorical grade. The categorised grades is an ordered set of
+categorical grade. The categorised grades are an ordered set of
 descriptive words or colors to communicate the severity or guide the
@@
 \(c_0, c_1, \cdots, c_z\) are the thresholds for each category. In SPI,
-drought are sorted into four categories: mild drought: \([-0.99, 0]\);
+droughts are sorted into four categories: mild drought: \([-0.99, 0]\);
 moderate drought: \([-1.49, -1]\); severe drought: \([-1.99, -1.5]\),

### 2022-12-01 paper/paper.qmd
easy to implement changes from ursula's comments

@@
 An overview of the pipeline is given in @fig-pipeline-steps to illustrate the construction from raw data to the final indexes.
-The pipeline includes modules for operations in the spatial, temporal, and multivariate aspects of the data as well as modules for comparing and communicating indexes (benchmarking and simplification).
+The pipeline includes modules for operations in the spatial, temporal, and multivariate aspects of the data as well as modules for comparing and communicating indexes.
 Analysts are free to select the modules they need and arrange them in the order they see fit to construct indexes.
@@
 
-**Temporal processing**
+### Temporal processing
 
 The construction of an index sometimes needs to consider information from neighbouring time periods.
-The temporal processing is a general operator on the time dimension of the data in the form of  $$f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),$$ where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
+The temporal processing is a general operator on the time dimension of the data in the form of
+
+\begin{equation}
+f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
+\end{equation}
+
+where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. 
 A typical example of temporal processing is aggregation in the drought index SPI to measure the lack of precipitation for meteorological drought.
@@
 
+*more explicit on k will influence 1) long term vs. short term, 2) uncertainty*
 The choice of time scales parameter $k$ can result in variation in the calculated index values: a small $k$ of 3 or 6 months produces the index more sensitive to individual months, while a large $k$ of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives dryness information relative to the long term condition.
@@
 
-**Spatial aggregation**
+### Spatial processing
 
@@
 These spatial data need to be merged into a uniform unit to construct indexes.
-In the pipeline, spatial aggregation includes operations on the spatial dimension of the data and can be written as $$g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),$$ where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.
-When one of the variables collected is only available in a coarser resolution, $i^\prime$, analysts may choose to reduce other variables with finer resolution, $i$, to match the coarser resolution, as in @fig-spatial-agg.
+In the pipeline, spatial aggregation includes operations on the spatial dimension of the data and can be written as
+
+\begin{equation}
+g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
+\end{equation}
+
+where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$. 
+When one of the variables collected is only available in a coarser resolution, $i^\prime$, analysts may choose to reduce other variables with finer resolution, $i$, to match the coarser resolution.
 This is a spatial aggregation step and aggregating using the mean can be written as: $x(s_{i^\prime};t_{j}) = \bar{x}(s_{i \in i^\prime}; t_j),$, where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid.
-*(Feel like the figure here is more of a visual illustration of the equation without enough further implication =.=)*
 
-```{r fig-spatial-agg}
-#| fig-cap: "Illustration of aggregating raster data with finer resolution (right) into coarser resolution (left)." 
-#| fig-width: 10
-#| fig-height: 3
-dt <- crossing(y = 1:2, x = 1:3) %>% 
-  arrange(-y) %>% 
-  mutate(label = as.factor(1:6))
-okabe <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#CC79A7")
+### Variable transformation
 
-p1 <- dt %>% 
-  ggplot(aes(x = x, y = y)) + 
-  geom_tile(aes(fill = label), color = "black") + 
-  geom_text(aes(label = label)) + 
-  scale_fill_manual(values = okabe) + 
-  theme_void() + 
-  theme(aspect.ratio = 2/3)
+Variable transformation, and scaling in the next section, are both pre-processing steps to create variables that fits assumptions for further computing.
+These include a stable variance, normal distribution, or a certain scale required by some algorithm further down the pipeline.
+Variable transformation is a general notion of a functional transformation on the variable:
 
-dt2 <- crossing(y = 1:6, x = 1:9) %>% 
-  arrange(-y) %>% 
-  mutate(x1 = (x-1) %/%3 + 1, y1 = (y-1) %/%3 + 1) %>% 
-  left_join(dt, by = c("x1" = "x", "y1" = "y")) %>% 
-  mutate(id = row_number())
+\begin{equation}
+h_{\tau}(x(\mathbf{s};\mathbf{t})),
+\end{equation}
 
-p2 <- dt2 %>% 
-  ggplot(aes(x = x, y = y)) + 
-  geom_tile(aes(fill = label), color = "black") + 
-  geom_text(aes(label = id)) + 
-  scale_fill_manual(values = okabe) + 
-  theme_void() + 
-  theme(aspect.ratio = 2/3)
-
-(p1 | p2) & theme(legend.position = "none")
-```
-
-**Variable transformation **
-
-Variable transformation, and scaling in the next section, are both pre-processing steps to create variables with good statistical properties.
-These properties include a stable variance, normal distribution, or a certain scale required by some algorithm further down the pipeline.
-Variable transformation is a general notion of a functional transformation on the variable: $$h_{\tau}(x(\mathbf{s};\mathbf{t})),$$ where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation if any, and $d_{\tau}$ is the number of parameter of $\tau$.
+where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation if any, and $d_{\tau}$ is the number of parameter of $\tau$.
 Transformation is needed for data that are highly skewed and some common transformations include log, quadratic, and square root transformation.
 
-**Scaling**
+### Scaling
 
 While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
-The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
+The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of
+
+\begin{equation}
+[x(s_i;t_j) - \alpha]/\gamma.
+\end{equation}
+
 For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
@@
 
-**Dimension reduction**
+### Dimension reduction
 
 When constructing indexes from multivariate information, dimension reduction methods combine that information into a univariate series.
-In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate): $$x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),$$ where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$.
+In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate):
+
+\begin{equation}
+x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),
+\end{equation}
+
+where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$. 
 The most commonly used dimension reduction technique is Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science.
 It can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the data variance given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
+
+*add another perspective that there are statistical methods, i.e. PCA, and also domain knowledge driven methods, i.e. d = P - PET in SPEI.*
 
@@
 
-**Distribution fit**
+### Distribution fit
 
@@
 <!-- With a probability model imposed,  -->
-Distribution fit can be seen as the model fitting in its simplest term.
-It can be represented by $$F_{\eta}(x(\mathbf{s}; \mathbf{t})),$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
+Distribution fit can be seen as the model fitting in its simplest term. It can be represented by
+
+\begin{equation}
+F_{\eta}(x(\mathbf{s}; \mathbf{t})), 
+\end{equation}
+
+where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
 A distribution fit typically aims at finding the distribution that best fits the data.
@@
 
-**Normalising**
+### Normalising
 
@@
 
-**Benchmarking** [TODO]
+### Benchmarking
 
@@
 
-**Simplification**
+### Simplification
 
@@
 
-$$
+\begin{equation}
 \begin{cases}
@@
 \end{cases}
-$$
+\end{equation}
 

### 2022-12-01 paper/paper.tex
easy to implement changes from ursula's comments

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\textbf{Temporal processing}
+\hypertarget{temporal-processing}{%
+\subsubsection{Temporal processing}\label{temporal-processing}}
 
@@
 operator on the time dimension of the data in the form of
-\[f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),\] where
-\(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the parameters
-associated with the temporal operation and \(d_{\psi}\) is the number of
-parameter of \(\psi\). A typical example of temporal processing is
-aggregation in the drought index SPI to measure the lack of
-precipitation for meteorological drought. In SPI, monthly precipitation
-is aggregated by a time scale parameter \(k\):
+
+\begin{equation}
+f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
+\end{equation}
+
+where \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the
+parameters associated with the temporal operation and \(d_{\psi}\) is
+the number of parameter of \(\psi\). A typical example of temporal
+processing is aggregation in the drought index SPI to measure the lack
+of precipitation for meteorological drought. In SPI, monthly
+precipitation is aggregated by a time scale parameter \(k\):
 \(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),\)
@@
 
-The choice of time scales parameter \(k\) can result in variation in the
-calculated index values: a small \(k\) of 3 or 6 months produces the
-index more sensitive to individual months, while a large \(k\) of 24 or
-36, an equivalent to a 2- or 3-year aggregation, gives dryness
-information relative to the long term condition. As will be shown in
-section {[}SECTION EXAMPLE{]}, this variation may even lead to
+\emph{more explicit on k will influence 1) long term vs.~short term, 2)
+uncertainty} The choice of time scales parameter \(k\) can result in
+variation in the calculated index values: a small \(k\) of 3 or 6 months
+produces the index more sensitive to individual months, while a large
+\(k\) of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives
+dryness information relative to the long term condition. As will be
+shown in section {[}SECTION EXAMPLE{]}, this variation may even lead to
 conflicting conclusions on the dry/wet condition of the area,
@@
 
-\textbf{Spatial aggregation}
+\hypertarget{spatial-processing}{%
+\subsubsection{Spatial processing}\label{spatial-processing}}
 
@@
 dimension of the data and can be written as
-\[g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),\] where
-\(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the spatial
-operation parameters and \(d_{\theta}\) is the number of parameter of
-\(\theta\). When one of the variables collected is only available in a
-coarser resolution, \(i^\prime\), analysts may choose to reduce other
-variables with finer resolution, \(i\), to match the coarser resolution,
-as in Figure~\ref{fig-spatial-agg}. This is a spatial aggregation step
-and aggregating using the mean can be written as:
+
+\begin{equation}
+g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
+\end{equation}
+
+where \(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the
+spatial operation parameters and \(d_{\theta}\) is the number of
+parameter of \(\theta\). When one of the variables collected is only
+available in a coarser resolution, \(i^\prime\), analysts may choose to
+reduce other variables with finer resolution, \(i\), to match the
+coarser resolution. This is a spatial aggregation step and aggregating
+using the mean can be written as:
 \(x(s_{i^\prime};t_{j}) = \bar{x}(s_{i \in i^\prime}; t_j),\), where
 \(i \in i^\prime\) includes all the cells from the finer resolution in
-the coarser grid. \emph{(Feel like the figure here is more of a visual
-illustration of the equation without enough further implication =.=)}
+the coarser grid.
 
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-pdf/fig-spatial-agg-1.pdf}
-
-}
-
-\caption{\label{fig-spatial-agg}Illustration of aggregating raster data
-with finer resolution (right) into coarser resolution (left).}
-
-\end{figure}
-
-\textbf{Variable transformation }
+\hypertarget{variable-transformation}{%
+\subsubsection{Variable transformation}\label{variable-transformation}}
 
 Variable transformation, and scaling in the next section, are both
-pre-processing steps to create variables with good statistical
-properties. These properties include a stable variance, normal
-distribution, or a certain scale required by some algorithm further down
-the pipeline. Variable transformation is a general notion of a
-functional transformation on the variable:
-\[h_{\tau}(x(\mathbf{s};\mathbf{t})),\] where
-\(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in the
-transformation if any, and \(d_{\tau}\) is the number of parameter of
-\(\tau\). Transformation is needed for data that are highly skewed and
-some common transformations include log, quadratic, and square root
+pre-processing steps to create variables that fits assumptions for
+further computing. These include a stable variance, normal distribution,
+or a certain scale required by some algorithm further down the pipeline.
+Variable transformation is a general notion of a functional
+transformation on the variable:
+
+\begin{equation}
+h_{\tau}(x(\mathbf{s};\mathbf{t})),
+\end{equation}
+
+where \(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in
+the transformation if any, and \(d_{\tau}\) is the number of parameter
+of \(\tau\). Transformation is needed for data that are highly skewed
+and some common transformations include log, quadratic, and square root
 transformation.
 
-\textbf{Scaling}
+\hypertarget{scaling}{%
+\subsubsection{Scaling}\label{scaling}}
 
@@
 only changes the data scale and can usually be written in the form of
-\[[x(s_i;t_j) - \alpha]/\gamma.\] For example, a z-score standardisation
-can be written in the above form with \(\alpha = \bar{x}(s; t)\) and
-\(\gamma = \sigma(s; t)\), a min-max standardisation uses
-\(\alpha = \min[x(s_i, t_j)]\) and
+
+\begin{equation}
+[x(s_i;t_j) - \alpha]/\gamma.
+\end{equation}
+
+For example, a z-score standardisation can be written in the above form
+with \(\alpha = \bar{x}(s; t)\) and \(\gamma = \sigma(s; t)\), a min-max
+standardisation uses \(\alpha = \min[x(s_i, t_j)]\) and
 \(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\).
@@
 
-\textbf{Dimension reduction}
+\hypertarget{dimension-reduction}{%
+\subsubsection{Dimension reduction}\label{dimension-reduction}}
 
@@
 univariate):
-\[x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),\]
+
+\begin{equation}
+x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),
+\end{equation}
+
 where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\) reduce the
@@
 the data variance given the constraint \(\sum_{p=1}^P\lambda_p^2 = 1\).
+
+\emph{add another perspective that there are statistical methods,
+i.e.~PCA, and also domain knowledge driven methods, i.e.~d = P - PET in
+SPEI.}
 
@@
 
-\textbf{Distribution fit}
+\hypertarget{distribution-fit}{%
+\subsubsection{Distribution fit}\label{distribution-fit}}
 
@@
 Distribution fit can be seen as the model fitting in its simplest term.
-It can be represented by \[F_{\eta}(x(\mathbf{s}; \mathbf{t})),\] where
-\(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
+It can be represented by
+
+\begin{equation}
+F_{\eta}(x(\mathbf{s}; \mathbf{t})), 
+\end{equation}
+
+where \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
 parameter and \(d_{\eta}\) is the number of parameter of \(\eta\). A
@@
 
-\textbf{Normalising}
+\hypertarget{normalising}{%
+\subsubsection{Normalising}\label{normalising}}
 
@@
 
-\textbf{Benchmarking} {[}TODO{]}
+\hypertarget{benchmarking}{%
+\subsubsection{Benchmarking}\label{benchmarking}}
 
@@
 
-\textbf{Simplification}
+\hypertarget{simplification}{%
+\subsubsection{Simplification}\label{simplification}}
 
@@
 
-\[
+\begin{equation}
 \begin{cases}
@@
 \end{cases}
-\]
+\end{equation}
 

### 2022-12-02 paper/paper.qmd
rearrange spatial processing

@@
 where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. 
-A typical example of temporal processing is aggregation in the drought index SPI to measure the lack of precipitation for meteorological drought.
+A typical example of temporal processing is aggregation, which is used in the drought index SPI to measure the lack of precipitation for meteorological drought.
 In SPI, monthly precipitation is aggregated by a time scale parameter $k$: $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),$ where $j^\prime$ is the new time index after the aggregation.
-In this notation, each spatial location is separately aggregated and precipitation is ummed from $k$ month back, $j^\prime - k + 1$, to the current period, $j^\prime$, to create the aggregated series, indexed by $j^\prime$.
+In this notation, each spatial location is separately aggregated and precipitation is summed from $k$ month back, $j^\prime - k + 1$, to the current period, $j^\prime$, to create the aggregated series, indexed by $j^\prime$.
 
 *more explicit on k will influence 1) long term vs. short term, 2) uncertainty*
+
 The choice of time scales parameter $k$ can result in variation in the calculated index values: a small $k$ of 3 or 6 months produces the index more sensitive to individual months, while a large $k$ of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives dryness information relative to the long term condition.
@@
 
-While many indexes are calculated independently on spatial locations, borrowing spatial information can happen at unobservable locations.
-Also, variables can be collected in different formats: some from satellite imageries as raster data, potentially at different resolutions, and some from ground stations as vector data.
-These spatial data need to be merged into a uniform unit to construct indexes.
-In the pipeline, spatial aggregation includes operations on the spatial dimension of the data and can be written as
+Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing.
+The process can be written as a general operation in the form of
 
 \begin{equation}
-g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
+x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
 \end{equation}
 
-where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$. 
-When one of the variables collected is only available in a coarser resolution, $i^\prime$, analysts may choose to reduce other variables with finer resolution, $i$, to match the coarser resolution.
-This is a spatial aggregation step and aggregating using the mean can be written as: $x(s_{i^\prime};t_{j}) = \bar{x}(s_{i \in i^\prime}; t_j),$, where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid.
+where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$. 
+An example of spatial processing is to align variables collected in different resolutions.
+When variables are collected at different resolutions, analysts may choose to down-sample those in a finer resolution, $i$, to match those in a coarser resolution, $i^\prime$.
+This is a spatial aggregation and if aggregate using the mean, it can be written as
+
+\begin{equation}
+g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}},
+\end{equation}
+
+where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the  coarser grid.
+Other examples of spatial processing include 1) borrowing information from neighbouring spatial locations to interpolate unobserved locations and 2) fusing variables from ground measures with satellite imageries.
 

### 2022-12-02 paper/paper.tex
rearrange spatial processing

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 multivariate aspects of the data as well as modules for comparing and
-communicating indexes (benchmarking and simplification). Analysts are
-free to select the modules they need and arrange them in the order they
-see fit to construct indexes. While the starting point of the pipeline
-is raw data, there are steps prior to this that are crucial to the
-success of an index. For example, the defined index needs to be useful
-for measuring the concept of interest and variables need to be collected
-from reliable sources with proper quality control.
+communicating indexes. Analysts are free to select the modules they need
+and arrange them in the order they see fit to construct indexes. While
+the starting point of the pipeline is raw data, there are steps prior to
+this that are crucial to the success of an index. For example, the
+defined index needs to be useful for measuring the concept of interest
+and variables need to be collected from reliable sources with proper
+quality control.
 
@@
 the number of parameter of \(\psi\). A typical example of temporal
-processing is aggregation in the drought index SPI to measure the lack
-of precipitation for meteorological drought. In SPI, monthly
-precipitation is aggregated by a time scale parameter \(k\):
+processing is aggregation, which is used in the drought index SPI to
+measure the lack of precipitation for meteorological drought. In SPI,
+monthly precipitation is aggregated by a time scale parameter \(k\):
 \(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),\)
@@
 notation, each spatial location is separately aggregated and
-precipitation is ummed from \(k\) month back, \(j^\prime - k + 1\), to
+precipitation is summed from \(k\) month back, \(j^\prime - k + 1\), to
 the current period, \(j^\prime\), to create the aggregated series,
@@
 \emph{more explicit on k will influence 1) long term vs.~short term, 2)
-uncertainty} The choice of time scales parameter \(k\) can result in
-variation in the calculated index values: a small \(k\) of 3 or 6 months
-produces the index more sensitive to individual months, while a large
-\(k\) of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives
-dryness information relative to the long term condition. As will be
-shown in section {[}SECTION EXAMPLE{]}, this variation may even lead to
+uncertainty}
+
+The choice of time scales parameter \(k\) can result in variation in the
+calculated index values: a small \(k\) of 3 or 6 months produces the
+index more sensitive to individual months, while a large \(k\) of 24 or
+36, an equivalent to a 2- or 3-year aggregation, gives dryness
+information relative to the long term condition. As will be shown in
+section {[}SECTION EXAMPLE{]}, this variation may even lead to
 conflicting conclusions on the dry/wet condition of the area,
@@
 
-While many indexes are calculated independently on spatial locations,
-borrowing spatial information can happen at unobservable locations.
-Also, variables can be collected in different formats: some from
-satellite imageries as raster data, potentially at different
-resolutions, and some from ground stations as vector data. These spatial
-data need to be merged into a uniform unit to construct indexes. In the
-pipeline, spatial aggregation includes operations on the spatial
-dimension of the data and can be written as
+Spatial processing may be needed when indexes are not calculated
+independently on each collected location or when variables collected
+from multiple sources need to be fused before further processing. The
+process can be written as a general operation in the form of
 
 \begin{equation}
-g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
+x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
 \end{equation}
@@
 where \(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the
-spatial operation parameters and \(d_{\theta}\) is the number of
-parameter of \(\theta\). When one of the variables collected is only
-available in a coarser resolution, \(i^\prime\), analysts may choose to
-reduce other variables with finer resolution, \(i\), to match the
-coarser resolution. This is a spatial aggregation step and aggregating
-using the mean can be written as:
-\(x(s_{i^\prime};t_{j}) = \bar{x}(s_{i \in i^\prime}; t_j),\), where
-\(i \in i^\prime\) includes all the cells from the finer resolution in
-the coarser grid.
+associated parameters in the process and \(d_{\theta}\) is the number of
+parameter of \(\theta\). An example of spatial processing is to align
+variables collected in different resolutions. When variables are
+collected at different resolutions, analysts may choose to down-sample
+those in a finer resolution, \(i\), to match those in a coarser
+resolution, \(i^\prime\). This is a spatial aggregation and if aggregate
+using the mean, it can be written as
+
+\begin{equation}
+g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}},
+\end{equation}
+
+where \(i \in i^\prime\) includes all the cells from the finer
+resolution in the coarser grid and \(n_{i^\prime}\) is the number of
+observations falls into the coarser grid. Other examples of spatial
+processing include 1) borrowing information from neighbouring spatial
+locations to interpolate unobserved locations and 2) fusing variables
+from ground measures with satellite imageries.
 

### 2022-12-02 paper/paper.qmd
more on dimension reduction

@@
 
+*variables == indicators*
+
 An overview of the pipeline is given in @fig-pipeline-steps to illustrate the construction from raw data to the final indexes.
@@
 
+*Effective drought index*
+
 ### Spatial processing
@@
 
-Variable transformation, and scaling in the next section, are both pre-processing steps to create variables that fits assumptions for further computing.
-These include a stable variance, normal distribution, or a certain scale required by some algorithm further down the pipeline.
+The purpose of variable transformation is to create variables that fits assumptions for further computing.
+These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline.
 Variable transformation is a general notion of a functional transformation on the variable:
@@
 Transformation is needed for data that are highly skewed and some common transformations include log, quadratic, and square root transformation.
+
+<!-- A common assumption is normality in drought index.  -->
 
@@
 @fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step.
-All the scaling operations in green result in the same distribution as the original data.
+While both variable transformation and scaling are pre-processing steps, the scaling operations in green show the same distribution as the original data.
 
@@
 
+\newpage
+
 ### Dimension reduction
 
-When constructing indexes from multivariate information, dimension reduction methods combine that information into a univariate series.
+When indexes are constructed from multivariate information, dimension reduction methods combine that information into a univariate series.
 In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate):
@@
 The most commonly used dimension reduction technique is Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science.
-It can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the data variance given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.
+It can be seen as a special case of weighting, where variables are summed up in a linear combination: $$x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$ with restrictions imposed on the weight coefficient: $\sum_{p=1}^P\lambda_p^2 = 1$.
+In other cases of weighting, the coefficients can be as simple as giving equal weight to each variables.
 
-*add another perspective that there are statistical methods, i.e. PCA, and also domain knowledge driven methods, i.e. d = P - PET in SPEI.*
+Some dimension reduction can also be formulated from domain-specific knowledge.
+This can be theories that describe the physics of the phenomenon being indexed or practical formulations used to extract useful features from the raw variables.
+For example, in the index SPEI, a difference series is calculated between precipitation and potential evapotranspiration (PET) and the validity of this formulation is backed up by climate water balance model [Thornthwaite, 1948], which describes [...].
+*Add another example of remote sensing variables i.e.
+NDVI = (NIR - Red) / (NIR + Red)?*
 
-The PCA is a special case of weighting, where individual variables are summed up in a linear combination with potential restrictions imposed on the weight coefficients.
-These coefficients may be simple mathematical constructions that give equal weight to each variable or can be derived with specialised knowledge by experts in the field.
-While suggested weights can indicate norms adored by practitioners, analysts should be given the flexibility to experiment with different combinations when constructing indexes. 
-This could help understand index behavior from its sensitivity to the variables and suggest alternative weights that better suit specific tasks.
+<!-- Spectral satellite imageries often collected at different bands and it is common to combine values from multiple bands to compose new variables relevant to the index.
+The most common example of this is the Normalized Difference Vegetation Index (NDVI), which uses near-infrared (NIR) ($x_1(s;t)$) and red channel ($x_2(s;t)$): -->
 
-<!-- sometimes called feature extraction in the machine learning community  -->
-<!-- With drought indexes, the extraction of meaningful variables from the original data is usually supported by the water balance model, for example, in SPEI, the step that create $d$ out of precipitation and potential evapotranspiration (PET) has theoretical backup from [see paper.] -->
+<!-- \begin{equation} -->
+<!-- \text{NDVI}(s;t) = \frac{x_1(s;t) - x_2(s;t)}{x_1(s;t) + x_2(s;t)}, -->
+<!-- \end{equation} -->
+
+<!-- which is used to evaluate the vegetation from near-infrared (NIR) and red channel. -->
+
+While suggested weights and formulas can indicate norms adored by practitioners, analysts should be given the flexibility to experiment with different combinations when constructing indexes. 
+This could help understand index behavior from its sensitivity to the variables and suggest alternative weights that better suit the specific tasks.
 

### 2022-12-02 paper/paper.tex
more on dimension reduction

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
+\emph{variables == indicators}
+
 An overview of the pipeline is given in Figure~\ref{fig-pipeline-steps}
@@
 
+\emph{Effective drought index}
+
 \hypertarget{spatial-processing}{%
@@
 
-Variable transformation, and scaling in the next section, are both
-pre-processing steps to create variables that fits assumptions for
-further computing. These include a stable variance, normal distribution,
-or a certain scale required by some algorithm further down the pipeline.
-Variable transformation is a general notion of a functional
-transformation on the variable:
+The purpose of variable transformation is to create variables that fits
+assumptions for further computing. These assumptions include a stable
+variance, normal distribution, or a certain scale required by some
+algorithms down the pipeline. Variable transformation is a general
+notion of a functional transformation on the variable:
 
@@
 pre-processing operations and uses color to differentiate whether the
-operation is a variable transformation or a scaling step. All the
-scaling operations in green result in the same distribution as the
-original data.
+operation is a variable transformation or a scaling step. While both
+variable transformation and scaling are pre-processing steps, the
+scaling operations in green show the same distribution as the original
+data.
 
@@
 
+\newpage
+
 \hypertarget{dimension-reduction}{%
@@
 
-When constructing indexes from multivariate information, dimension
+When indexes are constructed from multivariate information, dimension
 reduction methods combine that information into a univariate series. In
@@
 also called Empirical Orthogonal Function (EOF) in earth science. It can
-be written as
-\(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
-where \(\lambda_p\) is the loading of the PC1, derived from maximising
-the data variance given the constraint \(\sum_{p=1}^P\lambda_p^2 = 1\).
+be seen as a special case of weighting, where variables are summed up in
+a linear combination:
+\[x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]
+with restrictions imposed on the weight coefficient:
+\(\sum_{p=1}^P\lambda_p^2 = 1\). In other cases of weighting, the
+coefficients can be as simple as giving equal weight to each variables.
 
-\emph{add another perspective that there are statistical methods,
-i.e.~PCA, and also domain knowledge driven methods, i.e.~d = P - PET in
-SPEI.}
+Some dimension reduction can also be formulated from domain-specific
+knowledge. This can be theories that describe the physics of the
+phenomenon being indexed or practical formulations used to extract
+useful features from the raw variables. For example, in the index SPEI,
+a difference series is calculated between precipitation and potential
+evapotranspiration (PET) and the validity of this formulation is backed
+up by climate water balance model {[}Thornthwaite, 1948{]}, which
+describes {[}\ldots{]}. \emph{Add another example of remote sensing
+variables i.e.~NDVI = (NIR - Red) / (NIR + Red)?}
 
-The PCA is a special case of weighting, where individual variables are
-summed up in a linear combination with potential restrictions imposed on
-the weight coefficients. These coefficients may be simple mathematical
-constructions that give equal weight to each variable or can be derived
-with specialised knowledge by experts in the field. While suggested
-weights can indicate norms adored by practitioners, analysts should be
-given the flexibility to experiment with different combinations when
-constructing indexes. This could help understand index behavior from its
-sensitivity to the variables and suggest alternative weights that better
-suit specific tasks.
+While suggested weights and formulas can indicate norms adored by
+practitioners, analysts should be given the flexibility to experiment
+with different combinations when constructing indexes. This could help
+understand index behavior from its sensitivity to the variables and
+suggest alternative weights that better suit the specific tasks.
 

### 2022-12-16 paper/paper.qmd
changes after the meeting

@@
 #| out-width: 100%
+# library(indri)
+# library(dplyr)
+# library(lubridate)
+# library(lmomco)
+# library(ggplot2)
+# library(tsibble)
+# library(SPEI)
+# data(wichita)
+# 
+# dt <- wichita %>% 
+#   aggregate(var = prcp, scale = c(6, 12), index = ym, id = id, gran = month) %>%
+#   filter(!is.na(.agg)) %>%
+#   normalise(dist = list(gev(), loglogistic()), var = .agg) %>%
+#   augment(col = .agg)
+# 
+# dt %>% 
+#   ggplot(aes(x = ym, y = .index, color = .dist)) + 
+#   geom_line()
 knitr::include_graphics(here::here("figures/toy-example-spei.png"))
@@
 
+*any index can be broken down into multiple steps and then we can do things with it: swap, change parameter, etc*
+
 *variables == indicators*
@@
 An overview of the pipeline is given in @fig-pipeline-steps to illustrate the construction from raw data to the final indexes.
-The pipeline includes modules for operations in the spatial, temporal, and multivariate aspects of the data as well as modules for comparing and communicating indexes.
+The pipeline includes eight modules for operations in the spatial, temporal, and multivariate aspects of the data as well as modules for comparing and communicating indexes.
 Analysts are free to select the modules they need and arrange them in the order they see fit to construct indexes.
@@
 
-The notation used in this section will be first introduced before elaborating on each pipeline module.
+Before elaborating each of the eight pipeline modules as subsections, the data notation will be first introduced.
 Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspects: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
@@
   scale_x_continuous(breaks = seq(-3, 3, 1)) + 
-  xlab("Density") + 
-  ylab("Quantile") + 
+  xlab("Normalised Score") + 
+  ylab("Probability") + 
   theme_bw() + 
@@
 
-# Incorporating new buliding blocks into the pipeline {#sec-incorporating-new-buliding-blocks-into-the-pipeline}
+# Incorporating alternative methods into the pipeline components {#sec-incorporating-new-buliding-blocks-into-the-pipeline}
 
@@
 
-## Constructing Standardised Precipitation Index (SPI)
+## A drought index example
 
@@
 
-## Calculating SPEI with raster data
+easier to make comparison
+pick a publication, propose a new index that has this benefit but also correlate highly with the big name indexes
+
+## An example on sustainable development indicators
+
+Weighting
+
+## Calculating indexes on raster data
+
+<!-- Raster data is common data format for delivering large satellite spatio-temporal data.  -->
+
+<!-- Issue 1: variables may be collected at different spatial and temporal extent/ scale, different resolutions and granularities.  -->
+
+<!-- cleaning these variables involves spatial and temporal processing.  -->
+
+<!-- Issue 2: what the data the step is operated on?? -->
+
+<!-- When calculating on raster data, it is confusing on what data the operation is on.
+Is the operation operated per pixel or across regions, is the operation operated across all the time point, across year, across month etc.
+-->
+
+<!-- While it may be clear to researchers themselves, but can be confusing for others who read the analysis for the first time.  -->
+
+<!-- The data structure should be made explicit at each operation on the spatio-temporal raster data, in terms of inputs and outputs. -->
+
+<!-- The raster examples  -->
 

### 2022-12-16 paper/paper.tex
changes after the meeting

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
+\emph{any index can be broken down into multiple steps and then we can
+do things with it: swap, change parameter, etc}
+
 \emph{variables == indicators}
@@
 to illustrate the construction from raw data to the final indexes. The
-pipeline includes modules for operations in the spatial, temporal, and
-multivariate aspects of the data as well as modules for comparing and
-communicating indexes. Analysts are free to select the modules they need
-and arrange them in the order they see fit to construct indexes. While
-the starting point of the pipeline is raw data, there are steps prior to
-this that are crucial to the success of an index. For example, the
-defined index needs to be useful for measuring the concept of interest
-and variables need to be collected from reliable sources with proper
-quality control.
+pipeline includes eight modules for operations in the spatial, temporal,
+and multivariate aspects of the data as well as modules for comparing
+and communicating indexes. Analysts are free to select the modules they
+need and arrange them in the order they see fit to construct indexes.
+While the starting point of the pipeline is raw data, there are steps
+prior to this that are crucial to the success of an index. For example,
+the defined index needs to be useful for measuring the concept of
+interest and variables need to be collected from reliable sources with
+proper quality control.
 
-The notation used in this section will be first introduced before
-elaborating on each pipeline module. Let
+Before elaborating each of the eight pipeline modules as subsections,
+the data notation will be first introduced. Let
 \(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with spatial,
@@
 \hypertarget{sec-incorporating-new-buliding-blocks-into-the-pipeline}{%
-\section{Incorporating new buliding blocks into the
-pipeline}\label{sec-incorporating-new-buliding-blocks-into-the-pipeline}}
+\section{Incorporating alternative methods into the pipeline
+components}\label{sec-incorporating-new-buliding-blocks-into-the-pipeline}}
 
@@
 
-\hypertarget{constructing-standardised-precipitation-index-spi}{%
-\subsection{Constructing Standardised Precipitation Index
-(SPI)}\label{constructing-standardised-precipitation-index-spi}}
+\hypertarget{a-drought-index-example}{%
+\subsection{A drought index example}\label{a-drought-index-example}}
 
@@
 
-\hypertarget{calculating-spei-with-raster-data}{%
-\subsection{Calculating SPEI with raster
-data}\label{calculating-spei-with-raster-data}}
+easier to make comparison pick a publication, propose a new index that
+has this benefit but also correlate highly with the big name indexes
+
+\hypertarget{an-example-on-sustainable-development-indicators}{%
+\subsection{An example on sustainable development
+indicators}\label{an-example-on-sustainable-development-indicators}}
+
+Weighting
+
+\hypertarget{calculating-indexes-on-raster-data}{%
+\subsection{Calculating indexes on raster
+data}\label{calculating-indexes-on-raster-data}}
 

### 2023-01-29 paper/paper.qmd
put up materials for HDI example

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+options(width = 500)
 library(tidyverse)
@@
 library(patchwork)
+library(indri)
+library(GGally)
 ```
@@
 
+\newpage
+
 ## An example on sustainable development indicators
 
-Weighting
+Another type of indexes that are commonly used in social science is to rank entities (i.e.
+countries) on a topic through a range of indicators.
+This includes university ranking from QS, Times Higher Education; the Sustainable Development Goals (SDG) Index [@SDSN_Bertelsmann_2021] by the Sustainable Development Solutions Network (SDSN), the Human Development Index (HDI) [@UNDP_2021] by United Nations Development Programme (UNDP), etc.
+These indexes, or rankings, typically have a set of predefined dimensions, or pillars, and indicators are first reduced to these pillars and then reduced to an index.
+In the following example, we will first show how to compute the Human Development Index (HDI) using the pipeline developed, and then demonstrate how the alternative expressions and parameters can be tested with two functions: `swap_exprs()` and `swap_values()`.
+
+Human Development Index (HDI) measures the development of a country through [...], more than just the economics.
+It uses three dimensions and four indicators: health (life expectancy at birth), education (expected years of schooling, mean years of schooling), and standard of living (GNI per capita).
+The technical notes [@UNDP_2021_tech_notes] from United Nation Development Programme have documented the procedures to calculate HDI and we summarise them as below:
+
+  - take log on GNI per capita,
+  - rescale the four indicators using mini-max with the provided minimum and maximum, 
+  - aggregate the two education indicators using the arithmetic mean
+  - aggregate the three dimensions into the index using the geometrical mean
+
+After the initial data import and wrangling, these steps can be calculated in the pipeline,as:
+
+```{r}
+dt <- readxl::read_xlsx(path = here::here("data/hdi.xlsx") , skip = 4)
+raw <- dt %>%
+  janitor::clean_names() %>%
+  dplyr::select(-paste0("x", seq(4, 14, 2))) %>%
+  dplyr::rename(id = x1, countries = x2) %>%
+  dplyr::mutate(across(c(id, human_development_index_hdi:hdi_rank), as.numeric)) %>%
+  dplyr::filter(!is.na(id)) %>%
+  dplyr::mutate(across(4:7, ~round(.x, digits = 3))) %>% 
+  dplyr::select(-c(human_development_index_hdi,
+                   gni_per_capita_rank_minus_hdi_rank, 
+                   hdi_rank)) 
+colnames(raw) <- c("id", "country", "life_exp", "exp_sch", "avg_sch", "gni_pc")
+
+scaling_params <- tibble::tribble(
+   ~Dimension, ~Indicator, ~Var,  ~Minimum, ~Maximum,
+  "Health",              "Life expectancy at birth (years)",   "life_exp",    "20",          "85",
+  "Education",           "Expected years of schooling (years)",  "exp_sch",    "0",          "18",
+  "Education",           "Mean years of schooling (years)",   "avg_sch",      "0",          "15",
+  "Standard of living",  "GNI per capita (2017 PPP$)",     "gni_pc",       "100",      "75000"
+  ) %>%
+  mutate(across(contains("mum"), as.numeric),
+         across(contains("mum"), ~ifelse(Var == "gni_pc", log10(.x), .x)))
+```
+
+```{r echo = TRUE, message=TRUE}
+dt <- raw %>% init(id = country, indicators = life_exp:gni_pc)
+(res <- dt %>%
+  var_trans(gni_pc = log10(gni_pc)) %>% 
+  var_trans(.method = rescale_minmax, .vars = life_exp:gni_pc,
+            min = scaling_params$Minimum, max = scaling_params$Maximum) %>% 
+  dim_red(sch = (exp_sch + avg_sch) / 2) %>%
+  dim_red(index = (life_exp * sch * gni_pc)^(1/3)))
+```
+
+The output is an `indri` object where a summary of the operations performed have been provided.
+The dataset contains the ID, country name,  rescaled variables (`life_exp`, `exp_sch`, `avg_sch`, `gni_pc`), the average of two indicators in the education dimension (`sch`), and the index calculated (`index`).
+
+A set of questions that can interest index analysts are the what-ifs: how would the index/ ranking changes if a particular parameter or expression changes in the pipeline.
+In the context of HDI, the index is calculated as the geometrical mean of the three dimensions, what if an arithmetic mean or different weights is used?
+The maximum and minimum limit is carefully chosen with justification provided in the technical notes:
+
+  > The low minimum value for gross national income (GNI) per capita, $100, is justified by the considerable amount of unmeasured subsistence and nonmarket production in economies close to the minimum, which is not captured in the official data
+
+  What if this lower minimum is raised or decreased?
+
+The function `swap_exprs()` and `swap_values()` can be used to answer these what-ifs.
+The two functions first uses a set of arguments to locate the operations the analysts would like to test, and then accepts a list of expressions or values for testing, and lastly, the initial `indri` (raw data).
+Another five expressions with different weighting emphasis are tested on calculating the final HDI:
+
+```{r echo = TRUE}
+res2 <- res %>%
+  swap_exprs(
+    .var = index,
+    .expr = list(
+      index1 = (life_exp + sch + gni_pc)/3,
+      index2 = 0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc,
+      index3 = 0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc,
+      index4 = 0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc,
+      # PCA recommended weight
+      index5 = 0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc), 
+    .raw_data = dt)
+```
+
+The result contains another 5 variations of indexes calculated from the expression above.
+Country ranks can be added using the base function `rank()`.
+Figure @fig-hdi-expr shows the scatterplot matrix of the country ranking against each other, including the original index (`index0`).
+From the plot, we learn that the geoemtrical mean (`rank0`), the arithmetic mean (`rank1`), and the PCA recommended weights (`rank5`) give almost similar ranks.
+This validates the method chosen by the original index.
+However, when the three dimensions are given different weights (`rank2`, `rank3`, and `rank4`), the country ranking can vary substantially.
+A cluster of points have been highlighted in the panel `rank3` vs.
+`rank4`.
+These countries are ranked highly according to `rank4` but become mediocre when ranked by `index3`.
+Looking into the three dimensions of these countries in Table @tbl-hdi-expr shows that these countries (Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) have high life expectancy and GNI, but relatively low score on education, hence they are ranked lower when education (`sch`) is given a larger weight in `index3` (`0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc`).
+
+```{r fig-hdi-expr}
+#| fig-height: 8
+#| fig-width: 12
+#| fig-cap: "my plot"
+with_rank <- res2$data %>% 
+  mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>% 
+  rename_with(~gsub(".index", "", .x), contains("rank.index"))
+
+lower_cluster <- with_rank %>% 
+  filter(rank4 < 30, rank3 > 50) %>% 
+  select(country, life_exp, sch, gni_pc, index3, index4, rank3, rank4)
+
+with_highlight <- with_rank %>% 
+  mutate(highlight = ifelse(country %in% lower_cluster$country, TRUE, FALSE))
+
+p1 <- ggpairs(
+  with_highlight, columns = 14:19,
+  upper = list(continuous = "points", combo = "facethist", 
+               discrete = "facetbar", na = "na"),
+  lower = NULL, diag = NULL) +
+  theme(aspect.ratio = 1)
+
+# https://stackoverflow.com/questions/65091593/how-to-mark-certain-point-in-ggpairs
+p1$plots[[23]]$mapping <- 
+    `class<-`(c(p1$plots[[23]]$mapping, aes(color = highlight)), "uneval")
+p1 <- p1 + scale_color_manual(values = c("black", "red"))
+p1
+```
+
+```{r echo = FALSE}
+#| tbl-cap:  my tables
+#| label: tbl-hdi-expr
+lower_cluster %>% 
+  knitr::kable(digits = 3)
+```
+
+```{r echo = TRUE}
+res3 <- res %>%
+  swap_values(
+    .module = var_trans, .step = rescale_minmax, .res = gni_pc,
+    .var = min, .values = log10(700),
+    .raw_data = dt)
+```
+
+\newpage
 

### 2023-01-29 paper/paper.tex
put up materials for HDI example

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
+\newpage
+
 \hypertarget{an-example-on-sustainable-development-indicators}{%
@@
 
-Weighting
+Another type of indexes that are commonly used in social science is to
+rank entities (i.e.~countries) on a topic through a range of indicators.
+This includes university ranking from QS, Times Higher Education; the
+Sustainable Development Goals (SDG) Index (Network and Stiftung 2021) by
+the Sustainable Development Solutions Network (SDSN), the Human
+Development Index (HDI) (Programme 2021a) by United Nations Development
+Programme (UNDP), etc. These indexes, or rankings, typically have a set
+of predefined dimensions, or pillars, and indicators are first reduced
+to these pillars and then reduced to an index. In the following example,
+we will first show how to compute the Human Development Index (HDI)
+using the pipeline developed, and then demonstrate how the alternative
+expressions and parameters can be tested with two functions:
+\texttt{swap\_exprs()} and \texttt{swap\_values()}.
+
+Human Development Index (HDI) measures the development of a country
+through {[}\ldots{]}, more than just the economics. It uses three
+dimensions and four indicators: health (life expectancy at birth),
+education (expected years of schooling, mean years of schooling), and
+standard of living (GNI per capita). The technical notes (Programme
+2021b) from United Nation Development Programme have documented the
+procedures to calculate HDI and we summarise them as below:
+
+\begin{itemize}
+\tightlist
+\item
+  take log on GNI per capita,
+\item
+  rescale the four indicators using mini-max with the provided minimum
+  and maximum,
+\item
+  aggregate the two education indicators using the arithmetic mean
+\item
+  aggregate the three dimensions into the index using the geometrical
+  mean
+\end{itemize}
+
+After the initial data import and wrangling, these steps can be
+calculated in the pipeline,as:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
+\NormalTok{(res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log10}\NormalTok{(gni\_pc)) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
+            \AttributeTok{min =}\NormalTok{ scaling\_params}\SpecialCharTok{$}\NormalTok{Minimum, }\AttributeTok{max =}\NormalTok{ scaling\_params}\SpecialCharTok{$}\NormalTok{Maximum) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{)))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+Index pipeline:
+
+Summary: 
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `log10(gni_pc)` -> gni_pc
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = life_exp, min = 20.000000, max = 85.000000)` -> life_exp
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = exp_sch, min = 0.000000, max = 18.000000)` -> exp_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = avg_sch, min = 0.000000, max = 15.000000)` -> avg_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = gni_pc, min = 2.000000, max = 4.875061)` -> gni_pc
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(exp_sch + avg_sch)/2` -> sch
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index
+\end{verbatim}
+
+\begin{verbatim}
+
+Data: 
+# A tibble: 191 x 8
+      id country                life_exp exp_sch avg_sch gni_pc   sch index
+   <dbl> <chr>                     <dbl>   <dbl>   <dbl>  <dbl> <dbl> <dbl>
+ 1     1 Switzerland               0.984   0.917   0.924  0.983 0.920 0.962
+ 2     2 Norway                    0.973   1.01    0.867  0.978 0.939 0.963
+ 3     3 Iceland                   0.964   1.06    0.918  0.955 0.991 0.970
+ 4     4 Hong Kong, China (SAR)    1.01    0.960   0.815  0.973 0.887 0.954
+ 5     5 Australia                 0.993   1.17    0.848  0.936 1.01  0.979
+ 6     6 Denmark                   0.944   1.04    0.864  0.967 0.952 0.954
+ 7     7 Sweden                    0.969   1.08    0.841  0.952 0.960 0.960
+ 8     8 Ireland                   0.954   1.05    0.772  1.00  0.912 0.955
+ 9     9 Germany                   0.933   0.945   0.939  0.952 0.942 0.942
+10    10 Netherlands               0.949   1.04    0.839  0.956 0.939 0.948
+# ... with 181 more rows
+\end{verbatim}
+
+The output is an \texttt{indri} object where a summary of the operations
+performed have been provided. The dataset contains the ID, country name,
+rescaled variables (\texttt{life\_exp}, \texttt{exp\_sch},
+\texttt{avg\_sch}, \texttt{gni\_pc}), the average of two indicators in
+the education dimension (\texttt{sch}), and the index calculated
+(\texttt{index}).
+
+A set of questions that can interest index analysts are the what-ifs:
+how would the index/ ranking changes if a particular parameter or
+expression changes in the pipeline. In the context of HDI, the index is
+calculated as the geometrical mean of the three dimensions, what if an
+arithmetic mean or different weights is used? The maximum and minimum
+limit is carefully chosen with justification provided in the technical
+notes:
+
+\begin{quote}
+The low minimum value for gross national income (GNI) per capita, \$100,
+is justified by the considerable amount of unmeasured subsistence and
+nonmarket production in economies close to the minimum, which is not
+captured in the official data
+\end{quote}
+
+What if this lower minimum is raised or decreased?
+
+The function \texttt{swap\_exprs()} and \texttt{swap\_values()} can be
+used to answer these what-ifs. The two functions first uses a set of
+arguments to locate the operations the analysts would like to test, and
+then accepts a list of expressions or values for testing, and lastly,
+the initial \texttt{indri} (raw data). Another five expressions with
+different weighting emphasis are tested on calculating the final HDI:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{res2 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{swap\_exprs}\NormalTok{(}
+    \AttributeTok{.var =}\NormalTok{ index,}
+    \AttributeTok{.expr =} \FunctionTok{list}\NormalTok{(}
+      \AttributeTok{index1 =}\NormalTok{ (life\_exp }\SpecialCharTok{+}\NormalTok{ sch }\SpecialCharTok{+}\NormalTok{ gni\_pc)}\SpecialCharTok{/}\DecValTok{3}\NormalTok{,}
+      \AttributeTok{index2 =} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
+      \AttributeTok{index3 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
+      \AttributeTok{index4 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
+      \CommentTok{\# PCA recommended weight}
+      \AttributeTok{index5 =} \FloatTok{0.569} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.576} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.586} \SpecialCharTok{*}\NormalTok{ gni\_pc), }
+    \AttributeTok{.raw\_data =}\NormalTok{ dt)}
+\end{Highlighting}
+\end{Shaded}
+
+The result contains another 5 variations of indexes calculated from the
+expression above. Country ranks can be added using the base function
+\texttt{rank()}. Figure Figure~\ref{fig-hdi-expr} shows the scatterplot
+matrix of the country ranking against each other, including the original
+index (\texttt{index0}). From the plot, we learn that the geoemtrical
+mean (\texttt{rank0}), the arithmetic mean (\texttt{rank1}), and the PCA
+recommended weights (\texttt{rank5}) give almost similar ranks. This
+validates the method chosen by the original index. However, when the
+three dimensions are given different weights (\texttt{rank2},
+\texttt{rank3}, and \texttt{rank4}), the country ranking can vary
+substantially. A cluster of points have been highlighted in the panel
+\texttt{rank3} vs.~\texttt{rank4}. These countries are ranked highly
+according to \texttt{rank4} but become mediocre when ranked by
+\texttt{index3}. Looking into the three dimensions of these countries in
+Table Table~\ref{tbl-hdi-expr} shows that these countries (Andorra,
+Qatar, San Marino, Kuwait, and Brunei Darussalam) have high life
+expectancy and GNI, but relatively low score on education, hence they
+are ranked lower when education (\texttt{sch}) is given a larger weight
+in \texttt{index3}
+(\texttt{0.1\ *\ life\_exp\ +\ 0.1\ *\ sch\ +\ 0.8\ *\ gni\_pc}).
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/fig-hdi-expr-1.pdf}
+
+}
+
+\caption{\label{fig-hdi-expr}my plot}
+
+\end{figure}
+
+\hypertarget{tbl-hdi-expr}{}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2727}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1364}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1061}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1061}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1061}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0909}}@{}}
+\caption{\label{tbl-hdi-expr}my tables}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+country
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+life\_exp
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+sch
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+gni\_pc
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+index3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+index4
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+rank3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+rank4
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+country
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+life\_exp
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+sch
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+gni\_pc
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+index3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+index4
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+rank3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+rank4
+\end{minipage} \\
+\midrule()
+\endhead
+Andorra & 0.929 & 0.721 & 0.942 & 0.764 & 0.919 & 69 & 27 \\
+Qatar & 0.912 & 0.684 & 1.023 & 0.741 & 0.978 & 83 & 5 \\
+San Marino & 0.937 & 0.701 & 0.947 & 0.749 & 0.921 & 76 & 25 \\
+Kuwait & 0.903 & 0.670 & 0.947 & 0.721 & 0.915 & 96 & 29 \\
+Brunei Darussalam & 0.841 & 0.694 & 0.977 & 0.737 & 0.935 & 85 & 20 \\
+\bottomrule()
+\end{longtable}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{res3 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{swap\_values}\NormalTok{(}
+    \AttributeTok{.module =}\NormalTok{ var\_trans, }\AttributeTok{.step =}\NormalTok{ rescale\_minmax, }\AttributeTok{.res =}\NormalTok{ gni\_pc,}
+    \AttributeTok{.var =}\NormalTok{ min, }\AttributeTok{.values =} \FunctionTok{log10}\NormalTok{(}\DecValTok{700}\NormalTok{),}
+    \AttributeTok{.raw\_data =}\NormalTok{ dt)}
+\end{Highlighting}
+\end{Shaded}
+
+\newpage
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-SDSN_Bertelsmann_2021}{}}%
+Network, Sustainable Development Solutions, and Bertelsmann Stiftung.
+2021. {``SDG Index and Dashboards.''} \url{https://sdgindex.org/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021}{}}%
+Programme, United Nations Development. 2021a. {``Human Development Index
+(HDI).''}
+\url{https://hdr.undp.org/en/content/human-development-index-hdi}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021_tech_notes}{}}%
+---------. 2021b. {``Human Development Report 2021-2022: The Next
+Frontier: Human Development and the Anthropocene.''}
+\url{https://hdr.undp.org/sites/default/files/2021-22_HDR/hdr2021-22_technical_notes.pdf}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-sutherland_orca_2000}{}}%

### 2023-01-30 paper/paper.qmd
first draft on HDI index

--- 2022-12-16 paper/paper.qmd
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+options(width = 90)
 library(tidyverse)
@@
 
+\newpage
+
 ## An example on sustainable development indicators
 
-Weighting
+<!-- Indexes are commonly used in social science is to rank entities (i.e.
+countries) on a topic through a range of indicators.
+This includes university ranking from QS, Times Higher Education; the Sustainable Development Goals (SDG) Index [@SDSN_Bertelsmann_2021] by the Sustainable Development Solutions Network (SDSN), the Human Development Index (HDI) [@UNDP_2021] by United Nations Development Programme (UNDP), etc.
+These indexes, or rankings, typically have a set of predefined dimensions, or pillars, and indicators are first reduced to these pillars and then reduced to an index.
+-->
+
+In the following example, we will first constructing the Human Development Index (HDI) using the pipeline syntax and then show how the index, and ranking, may changes once an alternative expression or parameter in the computation is changed in two scenarios.
+
+Human Development Index (HDI) measures the development of countries from human living.
+Rather than focusing solely on the contribution of economic growth to the standard of living, HDI identifies another two important dimension in human development: a long living life and being knowledgeable.
+The three dimensions are measured using four indicators: life expectancy at birth (health), expected years of schooling (education), mean years of schooling (education), and Gross National Income (GNI) *per capita* (standard of living).
+The technical notes [@UNDP_2021_tech_notes] have documented the procedures to calculate HDI and they are summarised below:
+
+  1. take log on GNI *per capita*,
+  2. rescale the four indicators into [0, 1] using mini-max,
+  3. aggregate the two education indicators using arithmetic mean
+  4. aggregate the three dimensions into the index using geometric mean
+
+The values used in mini-max rescaling is summarised in @tbl-rescale-params and the justification of these numbers can be found in the technical notes mentioned above.
+
+```{r}
+#| tbl-cap-location: bottom
+#| tbl-cap: Maximum and minimum values used to rescale the four HDI indicators into [0, 1] range. The maximum of GNI per capita is taken as the common log of 75,000, approximatly 4.875.
+#| label: tbl-rescale-params
+scaling_params <- tibble::tribble(
+   ~Dimension, ~Indicator, ~Variable,  ~Minimum, ~Maximum,
+  "Health",              "Life expectancy at birth (years)",   "life_exp",    "20",          "85",
+  "Education",           "Expected years of schooling (years)",  "exp_sch",    "0",          "18",
+  "Education",           "Mean years of schooling (years)",   "avg_sch",      "0",          "15",
+  "Standard of living",  "GNI per capita (2017 PPP$)",     "gni_pc",       "100",      "75000"
+  ) %>%
+  mutate(across(contains("mum"), as.numeric),
+         across(contains("mum"), ~ifelse(Variable == "gni_pc", log10(.x), .x)))
+knitr::kable(scaling_params, digits = 3)
+```
+
+Among the four steps listed in the calculation, the first two are variable transformation.
+The next two steps can be considered as dimension reduction given it intends to combine multivariate data into univariate.
+With the index pipeline proposed in the paper, HDI can be calculated as:
+
+```{r}
+dt <- readxl::read_xlsx(path = here::here("data/hdi.xlsx") , skip = 4)
+raw <- dt %>%
+  janitor::clean_names() %>%
+  dplyr::select(-paste0("x", seq(4, 14, 2))) %>%
+  dplyr::rename(id = x1, countries = x2) %>%
+  dplyr::mutate(across(c(id, human_development_index_hdi:hdi_rank), as.numeric)) %>%
+  dplyr::filter(!is.na(id)) %>%
+  dplyr::mutate(across(4:7, ~round(.x, digits = 3))) %>% 
+  dplyr::select(-c(human_development_index_hdi,
+                   gni_per_capita_rank_minus_hdi_rank, 
+                   hdi_rank)) 
+colnames(raw) <- c("id", "country", "life_exp", "exp_sch", "avg_sch", "gni_pc")
+```
+
+```{r echo = TRUE, message=TRUE}
+dt <- raw %>% init(id = country, indicators = life_exp:gni_pc)
+(res <- dt %>%
+  var_trans(gni_pc = log10(gni_pc)) %>% 
+  var_trans(.method = rescale_minmax, .vars = life_exp:gni_pc,
+            min = c(20, 0, 0, 2), max = c(85, 18, 15, log10(75000))) %>% 
+  dim_red(sch = (exp_sch + avg_sch) / 2) %>%
+  dim_red(index = (life_exp * sch * gni_pc)^(1/3)))
+```
+
+The output is an `indri` object with a summary of operations followed by the data output.
+The data output contains the ID, country name, rescaled variables (`life_exp`, `exp_sch`, `avg_sch`, and `gni_pc`), the aggregated education dimension (`sch`), and the final index calculated (`index`).
+
+Index analysts can be interested in studying the property of the index and the countries through the "what-if" questions. In the context of HDI, this can be:
+
+  - what if arithmetic mean, or different weights, is used to reduce the three dimensions into HDI? 
+  - what if the maximum and minimum values used in the rescaling changes?
+
+We will now demonstrate how these questions can be answered using the function `swap_exprs()` and `swap_values()`.
+Both function first use a set of arguments (`.var`, `.module`, `.step`, and `.res`) to locate the operation to be tested.
+Then, `.expr` and `.values` will accept a list of expression or a vector of values respectively for testing.
+Lastly, the initial `indri` (raw data) needs to be provided.
+The following code tests on five expressions (arithmetic mean, three weighting schemes each emphasizing on one dimension, and principal component analysis recommended weight):
+
+```{r echo = TRUE, message=TRUE}
+(res2 <- res %>%
+  swap_exprs(
+    .var = index,
+    .expr = list(
+      index1 = (life_exp + sch + gni_pc)/3,
+      # PCA recommended weight
+      index2 = 0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc, 
+      index3 = 0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc,
+      index4 = 0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc,
+      index5 = 0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc),
+    .raw_data = dt))
+```
+
+The resulting data includes five more columns corresponding to the five alternative index.
+The country ranking can be further computed on the data and plotted in a scatterplot matrix as in @fig-hdi-expr.
+The plot shows almost similar ranks suggested by the indexes calculated using geometric mean (`rank0`), the arithmetic mean (`rank1`), and the PCA recommended weights (`rank2`).
+The agreement of the three methods supports the use of geometric mean as the dimension reduction method in the original index.
+The scatterplot matrix also reveals the difference in ranking when the three dimensions are given different weights in `rank3`, `rank4`, and `rank5`.
+The panel `rank4` vs `rank5` shows a cluster of highlighted points with relatively high ranks according to `rank5` but low in `rank4`.
+These countries are printed in @tbl-hdi-expr prints along with their three dimensions, calculated indexes and rankings.
+While having a high life expectancy and GNI *per capita*, these countries (Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) score relatively low in education, which is weighted heavily in `index4` (`0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc`).
+
+```{r fig-hdi-expr}
+#| fig-height: 8
+#| fig-width: 12
+#| fig-cap: "Scatterplot matrix of country rankings computed from different dimension reduction expressions in HDI: geometric mean (`rank0`), arithmetic mean (`rank1`), principal component analysis recommended weights (`rank2`), and heavier weight on life expectancy, education, and GNI per capita (`rank3` - `rank5`).
+Geometric mean, arithmetic mean, and PCA produce similar rankings where countries are aligned in a diagnoal line, while country rankings varies when compare among `rank3` to `rank5`.
+Panel `rank5` vs.
+`rank4` highlights a cluster of countries that rank high by `rank5` but low by `rank4`.
+The plot demonstrates the necessity to tests on various alternative expressions to study the robustness of the method and to capture distinctive characteristics of different countries.
+"
+with_rank <- res2$data %>% 
+  mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>% 
+  rename_with(~gsub(".index", "", .x), contains("rank.index"))
+
+lower_cluster <- with_rank %>% 
+  filter(rank5 < 30, rank4 > 50) %>% 
+  select(country, life_exp, sch, gni_pc, index3, index4, rank3, rank4)
+
+with_highlight <- with_rank %>% 
+  mutate(highlight = ifelse(country %in% lower_cluster$country, TRUE, FALSE))
+
+p1 <- ggpairs(
+  with_highlight, columns = 14:19,
+  lower = list(continuous = "points", combo = "facethist", 
+               discrete = "facetbar", na = "na"),
+  upper = NULL, diag = NULL, switch = "both") +
+  theme(aspect.ratio = 1)
+
+# https://stackoverflow.com/questions/65091593/how-to-mark-certain-point-in-ggpairs
+p1$plots[[35]]$mapping <- 
+    `class<-`(c(p1$plots[[35]]$mapping, aes(color = highlight)), "uneval")
+p1 <- p1 + scale_color_manual(values = c("black", "red"))
+p1
+```
+
+```{r echo = FALSE}
+#| tbl-cap: A selected number of countries with low rankings when education is given a heavy weight (`index4`) while ranks high when a heavier weight is given on GNI per capital (`index5`).
+#| label: tbl-hdi-expr
+lower_cluster %>% 
+  rename( `life expectancy` = life_exp, education = sch, `GNI per capita` = gni_pc) %>% 
+  knitr::kable(digits = 3)
+```
+
+The maximum and minimum limit is carefully chosen with justification provided in the technical notes:
+
+  > The low minimum value for gross national income (GNI) per capita, $100, is justified by the considerable amount of unmeasured subsistence and nonmarket production in economies close to the minimum, which is not captured in the official data
+
+  What if this lower minimum is raised or decreased?
+
+```{r echo = TRUE}
+res3 <- res %>%
+  swap_values(
+    .module = var_trans, .step = rescale_minmax, .res = gni_pc,
+    .var = min, .values = log10(700),
+    .raw_data = dt)
+```
+
+\newpage
 

### 2023-01-30 paper/paper.tex
first draft on HDI index

--- 2022-12-16 paper/paper.tex
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-Weighting
+In the following example, we will first constructing the Human
+Development Index (HDI) using the pipeline syntax and then show how the
+index, and ranking, may changes once an alternative expression or
+parameter in the computation is changed in two scenarios.
+
+Human Development Index (HDI) measures the development of countries from
+human living. Rather than focusing solely on the contribution of
+economic growth to the standard of living, HDI identifies another two
+important dimension in human development: a long living life and being
+knowledgeable. The three dimensions are measured using four indicators:
+life expectancy at birth (health), expected years of schooling
+(education), mean years of schooling (education), and Gross National
+Income (GNI) \emph{per capita} (standard of living). The technical notes
+(United Nations Development Programme 2021) have documented the
+procedures to calculate HDI and they are summarised below:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  take log on GNI \emph{per capita},
+\item
+  rescale the four indicators into {[}0, 1{]} using mini-max,
+\item
+  aggregate the two education indicators using arithmetic mean
+\item
+  aggregate the three dimensions into the index using geometric mean
+\end{enumerate}
+
+The values used in mini-max rescaling is summarised in
+Table~\ref{tbl-rescale-params} and the justification of these numbers
+can be found in the technical notes mentioned above.
+
+\hypertarget{tbl-rescale-params}{}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2375}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4500}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1125}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1000}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1000}}@{}}
+\caption{\label{tbl-rescale-params}Maximum and minimum values used to
+rescale the four HDI indicators into {[}0, 1{]} range. The maximum of
+GNI per capita is taken as the common log of 75,000, approximatly
+4.875.}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Dimension
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Indicator
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Variable
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Minimum
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Maximum
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Dimension
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Indicator
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Variable
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Minimum
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Maximum
+\end{minipage} \\
+\midrule()
+\endhead
+Health & Life expectancy at birth (years) & life\_exp & 20 & 85.000 \\
+Education & Expected years of schooling (years) & exp\_sch & 0 &
+18.000 \\
+Education & Mean years of schooling (years) & avg\_sch & 0 & 15.000 \\
+Standard of living & GNI per capita (2017 PPP\$) & gni\_pc & 2 &
+4.875 \\
+\bottomrule()
+\end{longtable}
+
+Among the four steps listed in the calculation, the first two are
+variable transformation. The next two steps can be considered as
+dimension reduction given it intends to combine multivariate data into
+univariate. With the index pipeline proposed in the paper, HDI can be
+calculated as:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
+\NormalTok{(res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log10}\NormalTok{(gni\_pc)) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
+            \AttributeTok{min =} \FunctionTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{), }\AttributeTok{max =} \FunctionTok{c}\NormalTok{(}\DecValTok{85}\NormalTok{, }\DecValTok{18}\NormalTok{, }\DecValTok{15}\NormalTok{, }\FunctionTok{log10}\NormalTok{(}\DecValTok{75000}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{)))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+Index pipeline:
+
+Summary: 
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `log10(gni_pc)` -> gni_pc
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = life_exp, min = 20.000000, max = 85.000000)` -> life_exp
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = exp_sch, min = 0.000000, max = 18.000000)` -> exp_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = avg_sch, min = 0.000000, max = 15.000000)` -> avg_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = gni_pc, min = 2.000000, max = 4.875061)` -> gni_pc
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(exp_sch + avg_sch)/2` -> sch
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index
+\end{verbatim}
+
+\begin{verbatim}
+
+Data: 
+# A tibble: 191 x 8
+      id country                life_exp exp_sch avg_sch gni_pc   sch index
+   <dbl> <chr>                     <dbl>   <dbl>   <dbl>  <dbl> <dbl> <dbl>
+ 1     1 Switzerland               0.984   0.917   0.924  0.983 0.920 0.962
+ 2     2 Norway                    0.973   1.01    0.867  0.978 0.939 0.963
+ 3     3 Iceland                   0.964   1.06    0.918  0.955 0.991 0.970
+ 4     4 Hong Kong, China (SAR)    1.01    0.960   0.815  0.973 0.887 0.954
+ 5     5 Australia                 0.993   1.17    0.848  0.936 1.01  0.979
+ 6     6 Denmark                   0.944   1.04    0.864  0.967 0.952 0.954
+ 7     7 Sweden                    0.969   1.08    0.841  0.952 0.960 0.960
+ 8     8 Ireland                   0.954   1.05    0.772  1.00  0.912 0.955
+ 9     9 Germany                   0.933   0.945   0.939  0.952 0.942 0.942
+10    10 Netherlands               0.949   1.04    0.839  0.956 0.939 0.948
+# ... with 181 more rows
+\end{verbatim}
+
+The output is an \texttt{indri} object with a summary of operations
+followed by the data output. The data output contains the ID, country
+name, rescaled variables (\texttt{life\_exp}, \texttt{exp\_sch},
+\texttt{avg\_sch}, and \texttt{gni\_pc}), the aggregated education
+dimension (\texttt{sch}), and the final index calculated
+(\texttt{index}).
+
+Index analysts can be interested in studying the property of the index
+and the countries through the ``what-if'' questions. In the context of
+HDI, this can be:
+
+\begin{itemize}
+\tightlist
+\item
+  what if arithmetic mean, or different weights, is used to reduce the
+  three dimensions into HDI?
+\item
+  what if the maximum and minimum values used in the rescaling changes?
+\end{itemize}
+
+We will now demonstrate how these questions can be answered using the
+function \texttt{swap\_exprs()} and \texttt{swap\_values()}. Both
+function first use a set of arguments (\texttt{.var}, \texttt{.module},
+\texttt{.step}, and \texttt{.res}) to locate the operation to be tested.
+Then, \texttt{.expr} and \texttt{.values} will accept a list of
+expression or a vector of values respectively for testing. Lastly, the
+initial \texttt{indri} (raw data) needs to be provided. The following
+code tests on five expressions (arithmetic mean, three weighting schemes
+each emphasizing on one dimension, and principal component analysis
+recommended weight):
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{(res2 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{swap\_exprs}\NormalTok{(}
+    \AttributeTok{.var =}\NormalTok{ index,}
+    \AttributeTok{.expr =} \FunctionTok{list}\NormalTok{(}
+      \AttributeTok{index1 =}\NormalTok{ (life\_exp }\SpecialCharTok{+}\NormalTok{ sch }\SpecialCharTok{+}\NormalTok{ gni\_pc)}\SpecialCharTok{/}\DecValTok{3}\NormalTok{,}
+      \CommentTok{\# PCA recommended weight}
+      \AttributeTok{index2 =} \FloatTok{0.569} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.576} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.586} \SpecialCharTok{*}\NormalTok{ gni\_pc, }
+      \AttributeTok{index3 =} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
+      \AttributeTok{index4 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
+      \AttributeTok{index5 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ gni\_pc),}
+    \AttributeTok{.raw\_data =}\NormalTok{ dt))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+Index pipeline:
+
+Summary: 
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `log10(gni_pc)` -> gni_pc
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = life_exp, min = 20, max = 85)` -> life_exp
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = exp_sch, min = 0, max = 18)` -> exp_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = avg_sch, min = 0, max = 15)` -> avg_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = gni_pc, min = 2, max = 4.875061)` -> gni_pc
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(exp_sch + avg_sch)/2` -> sch
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index0
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(life_exp + sch + gni_pc)/3` -> index1
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc` -> index2
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc` -> index3
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc` -> index4
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc` -> index5
+\end{verbatim}
+
+\begin{verbatim}
+
+Data: 
+# A tibble: 191 x 13
+      id country   life_~1 exp_sch avg_sch gni_pc   sch index0 index1 index2 index3 index4
+   <dbl> <chr>       <dbl>   <dbl>   <dbl>  <dbl> <dbl>  <dbl>  <dbl>  <dbl>  <dbl>  <dbl>
+ 1     1 Switzerl~   0.984   0.917   0.924  0.983 0.920  0.962  0.963   1.67  0.978  0.933
+ 2     2 Norway      0.973   1.01    0.867  0.978 0.939  0.963  0.963   1.67  0.970  0.946
+ 3     3 Iceland     0.964   1.06    0.918  0.955 0.991  0.970  0.970   1.68  0.966  0.985
+ 4     4 Hong Kon~   1.01    0.960   0.815  0.973 0.887  0.954  0.956   1.65  0.992  0.908
+ 5     5 Australia   0.993   1.17    0.848  0.936 1.01   0.979  0.979   1.69  0.989  1.00 
+ 6     6 Denmark     0.944   1.04    0.864  0.967 0.952  0.954  0.954   1.65  0.947  0.953
+ 7     7 Sweden      0.969   1.08    0.841  0.952 0.960  0.960  0.960   1.66  0.966  0.960
+ 8     8 Ireland     0.954   1.05    0.772  1.00  0.912  0.955  0.956   1.66  0.955  0.925
+ 9     9 Germany     0.933   0.945   0.939  0.952 0.942  0.942  0.942   1.63  0.936  0.942
+10    10 Netherla~   0.949   1.04    0.839  0.956 0.939  0.948  0.948   1.64  0.949  0.941
+# ... with 181 more rows, 1 more variable: index5 <dbl>, and abbreviated variable name
+#   1: life_exp
+\end{verbatim}
+
+The resulting data includes five more columns corresponding to the five
+alternative index. The country ranking can be further computed on the
+data and plotted in a scatterplot matrix as in
+Figure~\ref{fig-hdi-expr}. The plot shows almost similar ranks suggested
+by the indexes calculated using geometric mean (\texttt{rank0}), the
+arithmetic mean (\texttt{rank1}), and the PCA recommended weights
+(\texttt{rank2}). The agreement of the three methods supports the use of
+geometric mean as the dimension reduction method in the original index.
+The scatterplot matrix also reveals the difference in ranking when the
+three dimensions are given different weights in \texttt{rank3},
+\texttt{rank4}, and \texttt{rank5}. The panel \texttt{rank4} vs
+\texttt{rank5} shows a cluster of highlighted points with relatively
+high ranks according to \texttt{rank5} but low in \texttt{rank4}. These
+countries are printed in Table~\ref{tbl-hdi-expr} prints along with
+their three dimensions, calculated indexes and rankings. While having a
+high life expectancy and GNI \emph{per capita}, these countries
+(Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) score
+relatively low in education, which is weighted heavily in
+\texttt{index4}
+(\texttt{0.1\ *\ life\_exp\ +\ 0.1\ *\ sch\ +\ 0.8\ *\ gni\_pc}).
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/fig-hdi-expr-1.pdf}
+
+}
+
+\caption{\label{fig-hdi-expr}Scatterplot matrix of country rankings
+computed from different dimension reduction expressions in HDI:
+geometric mean (\texttt{rank0}), arithmetic mean (\texttt{rank1}),
+principal component analysis recommended weights (\texttt{rank2}), and
+heavier weight on life expectancy, education, and GNI per capita
+(\texttt{rank3} - \texttt{rank5}). Geometric mean, arithmetic mean, and
+PCA produce similar rankings where countries are aligned in a diagnoal
+line, while country rankings varies when compare among \texttt{rank3} -
+\texttt{rank5}. Panel \texttt{rank5} vs.~\texttt{rank4} highlights a
+cluster of countries that rank high by \texttt{rank5} but low by
+\texttt{rank4}. The plot demonstrates the necessity to tests on various
+alternative expressions to study the robustness of the method and to
+capture distinctive characteristics of different countries.}
+
+\end{figure}
+
+\hypertarget{tbl-hdi-expr}{}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2118}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1882}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1176}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1765}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0824}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0824}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0706}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0706}}@{}}
+\caption{\label{tbl-hdi-expr}A selected number of countries with low
+rankings when education is given a heavy weight (\texttt{index4}) while
+ranks high when a heavier weight is given on GNI per capital
+(\texttt{index5}).}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+country
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+life expectancy
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+education
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+GNI per capita
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+index3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+index4
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+rank3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+rank4
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+country
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+life expectancy
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+education
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+GNI per capita
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+index3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+index4
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+rank3
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+rank4
+\end{minipage} \\
+\midrule()
+\endhead
+Andorra & 0.929 & 0.721 & 0.942 & 0.909 & 0.764 & 32 & 69 \\
+Qatar & 0.912 & 0.684 & 1.023 & 0.900 & 0.741 & 34 & 83 \\
+San Marino & 0.937 & 0.701 & 0.947 & 0.914 & 0.749 & 31 & 76 \\
+Kuwait & 0.903 & 0.670 & 0.947 & 0.884 & 0.721 & 39 & 96 \\
+Brunei Darussalam & 0.841 & 0.694 & 0.977 & 0.840 & 0.737 & 60 & 85 \\
+\bottomrule()
+\end{longtable}
+
+The maximum and minimum limit is carefully chosen with justification
+provided in the technical notes:
+
+\begin{quote}
+The low minimum value for gross national income (GNI) per capita, \$100,
+is justified by the considerable amount of unmeasured subsistence and
+nonmarket production in economies close to the minimum, which is not
+captured in the official data
+\end{quote}
+
+What if this lower minimum is raised or decreased?
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{res3 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{swap\_values}\NormalTok{(}
+    \AttributeTok{.module =}\NormalTok{ var\_trans, }\AttributeTok{.step =}\NormalTok{ rescale\_minmax, }\AttributeTok{.res =}\NormalTok{ gni\_pc,}
+    \AttributeTok{.var =}\NormalTok{ min, }\AttributeTok{.values =} \FunctionTok{log10}\NormalTok{(}\DecValTok{700}\NormalTok{),}
+    \AttributeTok{.raw\_data =}\NormalTok{ dt)}
+\end{Highlighting}
+\end{Shaded}
+
+\newpage
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021_tech_notes}{}}%
+United Nations Development Programme. 2021. {``Human Development Report
+2021-2022: The Next Frontier: Human Development and the Anthropocene.''}
+\url{https://hdr.undp.org/sites/default/files/2021-22_HDR/hdr2021-22_technical_notes.pdf}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
-United Nations Development Programme. 2022. {``Human Development Report
-2021-22.''} New York. \url{http://report.hdr.undp.org}.
+---------. 2022. {``Human Development Report 2021-22.''} New York.
+\url{http://report.hdr.undp.org}.
 

### 2023-01-30 paper/paper.qmd
edit + put up materials for swap_values()

@@
 
-In the following example, we will first constructing the Human Development Index (HDI) using the pipeline syntax and then show how the index, and ranking, may changes once an alternative expression or parameter in the computation is changed in two scenarios.
+In the following example, the Human Development Index (HDI) will first be constructed using the pipeline syntax.
+The section will then demonstrate how expressions or parameters in the pipeline can be swapped to alternatives to study the index property.
 
-Human Development Index (HDI) measures the development of countries from human living.
-Rather than focusing solely on the contribution of economic growth to the standard of living, HDI identifies another two important dimension in human development: a long living life and being knowledgeable.
-The three dimensions are measured using four indicators: life expectancy at birth (health), expected years of schooling (education), mean years of schooling (education), and Gross National Income (GNI) *per capita* (standard of living).
+Human Development Index (HDI) measures the development of countries through more than just economic growth, but also people's life expectancy and opportunity to receive education.
+These three identified dimensions are measured using four indicators: life expectancy at birth (health), expected years of schooling (education), mean years of schooling (education), and Gross National Income (GNI) *per capita* (standard of living).
 The technical notes [@UNDP_2021_tech_notes] have documented the procedures to calculate HDI and they are summarised below:
@@
   2. rescale the four indicators into [0, 1] using mini-max,
-  3. aggregate the two education indicators using arithmetic mean
-  4. aggregate the three dimensions into the index using geometric mean
+  3. aggregate the two education indicators using arithmetic mean, and
+  4. aggregate the three dimensions into the index using geometric mean.
 
-The values used in mini-max rescaling is summarised in @tbl-rescale-params and the justification of these numbers can be found in the technical notes mentioned above.
+The values used in mini-max rescaling are summarised in @tbl-rescale-params and the justification of these numbers can be found in the technical notes mentioned above.
 
@@
 Among the four steps listed in the calculation, the first two are variable transformation.
-The next two steps can be considered as dimension reduction given it intends to combine multivariate data into univariate.
+The next two can be considered as the dimension reduction step in the data pipeline given its intention to combine multivariate data into univariate.
 With the index pipeline proposed in the paper, HDI can be calculated as:
@@
 (res <- dt %>%
-  var_trans(gni_pc = log10(gni_pc)) %>% 
+  var_trans(gni_pc = log(gni_pc, base = 10)) %>% 
   var_trans(.method = rescale_minmax, .vars = life_exp:gni_pc,
@@
   - what if arithmetic mean, or different weights, is used to reduce the three dimensions into HDI? 
-  - what if the maximum and minimum values used in the rescaling changes?
+  - what if an alternative maximum or minimum value is used in the rescaling step?
 
-We will now demonstrate how these questions can be answered using the function `swap_exprs()` and `swap_values()`.
-Both function first use a set of arguments (`.var`, `.module`, `.step`, and `.res`) to locate the operation to be tested.
-Then, `.expr` and `.values` will accept a list of expression or a vector of values respectively for testing.
+These questions can be answered using the function `swap_exprs()` and `swap_values()`.
+Both functions first use a set of arguments (`.var`, `.module`, `.step`, and `.res`) to locate the operation to be tested.
+Then, `.expr` and `.values` will accept a list of expressions or a vector of values for testing.
 Lastly, the initial `indri` (raw data) needs to be provided.
-The following code tests on five expressions (arithmetic mean, three weighting schemes each emphasizing on one dimension, and principal component analysis recommended weight):
+
+*What if an alternative dimension reduction expression is used to reduce the three dimensions into the HDI?*
+
+A common method to reduce the data dimension is through linear combination.
+With different weighting schemes, the combination can be compared to study the variable importance of each dimension or to reveal the underlying structures of countries.
+The following code tests five expressions (arithmetic mean, principal component analysis weight, and three weighting schemes emphasizing life expectancy, education, and GNI per capita respectively):
 
@@
 
-The resulting data includes five more columns corresponding to the five alternative index.
+The resulting data includes five more columns corresponding to the five alternative indexes.
 The country ranking can be further computed on the data and plotted in a scatterplot matrix as in @fig-hdi-expr.
@@
 The panel `rank4` vs `rank5` shows a cluster of highlighted points with relatively high ranks according to `rank5` but low in `rank4`.
-These countries are printed in @tbl-hdi-expr prints along with their three dimensions, calculated indexes and rankings.
+These countries are printed in @tbl-hdi-expr prints along with their three dimensions, calculated indexes, and rankings.
 While having a high life expectancy and GNI *per capita*, these countries (Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) score relatively low in education, which is weighted heavily in `index4` (`0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc`).
@@
 `rank4` highlights a cluster of countries that rank high by `rank5` but low by `rank4`.
-The plot demonstrates the necessity to tests on various alternative expressions to study the robustness of the method and to capture distinctive characteristics of different countries.
+The plot demonstrates the necessity to test on various alternative expressions to verify the robustness of the method and to capture distinctive characteristics of different countries.
 "
@@
 
-The maximum and minimum limit is carefully chosen with justification provided in the technical notes:
+\newpage
+
+*what if the minimum value for GNI per capita is changed?*
+
+The lower limit for GNI per capita is set to be the common log of $100, with justification provided in the technical notes as:
 
@@
 
-  What if this lower minimum is raised or decreased?
+The open source data provided by United Nations Development Programme suggest the minimum GNI *per capita* is `$732` and this opens the questions on how the index/ranking will change if the lower minimum (`$100`) is raised or decreased.
+We can answer this question by testing on a set of different values using `.values` in the function `swap_values()`:
 
-```{r echo = TRUE}
-res3 <- res %>%
+```{r echo = TRUE, message = TRUE}
+(res3 <- res %>%
   swap_values(
@@
     .var = min, .values = log10(700),
-    .raw_data = dt)
+    .raw_data = dt)) 
+```
+
+\newpage
+
+The function computes the GNI *per capita* in both the original value (`gni_pc0`) and the new value provided (`gni_pc1`).
+All the subsequent operations are also computed for both cases and result in `index0` and `index1`.
+
+```{r}
+#| fig-height: 6
+#| fig-width: 6
+#| fig-align: center
+new_rank <- res3$data %>%
+  mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>%
+  rename_with(~gsub(".index", "", .x), contains("rank.index")) %>% 
+  mutate(difference = rank1 - rank0)
+
+try <- new_rank %>%
+  dplyr::select(id, country, contains("life_exp"), contains("index"), contains("rank"), difference) %>% 
+  arrange(-abs(difference))
+
+new_rank %>% 
+  ggplot(aes(x = rank0, y = rank1)) + 
+  geom_point() + 
+  geom_abline(slope = 1, intercept = 0, color = "blue") +
+  # geom_text(data = try %>% filter(difference > 15), color = "red", aes(label = country)) +
+  theme(aspect.ratio = 1) +
+  xlab("log10(100) as the rescale minimum for GNI per capita") +
+  ylab("log10(700) as the rescale minimum for GNI per capita") 
 ```

### 2023-01-30 paper/paper.tex
edit + put up materials for swap_values()

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-In the following example, we will first constructing the Human
-Development Index (HDI) using the pipeline syntax and then show how the
-index, and ranking, may changes once an alternative expression or
-parameter in the computation is changed in two scenarios.
-
-Human Development Index (HDI) measures the development of countries from
-human living. Rather than focusing solely on the contribution of
-economic growth to the standard of living, HDI identifies another two
-important dimension in human development: a long living life and being
-knowledgeable. The three dimensions are measured using four indicators:
-life expectancy at birth (health), expected years of schooling
-(education), mean years of schooling (education), and Gross National
-Income (GNI) \emph{per capita} (standard of living). The technical notes
-(United Nations Development Programme 2021) have documented the
-procedures to calculate HDI and they are summarised below:
+In the following example, the Human Development Index (HDI) will first
+be constructed using the pipeline syntax. The section will then
+demonstrate how expressions or parameters in the pipeline can be swapped
+to alternatives to study the index property.
+
+Human Development Index (HDI) measures the development of countries
+through more than just economic growth, but also people's life
+expectancy and opportunity to receive education. These three identified
+dimensions are measured using four indicators: life expectancy at birth
+(health), expected years of schooling (education), mean years of
+schooling (education), and Gross National Income (GNI) \emph{per capita}
+(standard of living). The technical notes (United Nations Development
+Programme 2021) have documented the procedures to calculate HDI and they
+are summarised below:
 
@@
 \item
-  aggregate the two education indicators using arithmetic mean
-\item
-  aggregate the three dimensions into the index using geometric mean
+  aggregate the two education indicators using arithmetic mean, and
+\item
+  aggregate the three dimensions into the index using geometric mean.
 \end{enumerate}
 
-The values used in mini-max rescaling is summarised in
+The values used in mini-max rescaling are summarised in
 Table~\ref{tbl-rescale-params} and the justification of these numbers
@@
 Among the four steps listed in the calculation, the first two are
-variable transformation. The next two steps can be considered as
-dimension reduction given it intends to combine multivariate data into
-univariate. With the index pipeline proposed in the paper, HDI can be
-calculated as:
+variable transformation. The next two can be considered as the dimension
+reduction step in the data pipeline given its intention to combine
+multivariate data into univariate. With the index pipeline proposed in
+the paper, HDI can be calculated as:
 
@@
 \NormalTok{(res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log10}\NormalTok{(gni\_pc)) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log}\NormalTok{(gni\_pc, }\AttributeTok{base =} \DecValTok{10}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
   \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
@@
 \begin{verbatim}
-var_trans: `log10(gni_pc)` -> gni_pc
+var_trans: `log(gni_pc, base = 10)` -> gni_pc
 \end{verbatim}
@@
 \item
-  what if the maximum and minimum values used in the rescaling changes?
+  what if an alternative maximum or minimum value is used in the
+  rescaling step?
 \end{itemize}
 
-We will now demonstrate how these questions can be answered using the
-function \texttt{swap\_exprs()} and \texttt{swap\_values()}. Both
-function first use a set of arguments (\texttt{.var}, \texttt{.module},
-\texttt{.step}, and \texttt{.res}) to locate the operation to be tested.
-Then, \texttt{.expr} and \texttt{.values} will accept a list of
-expression or a vector of values respectively for testing. Lastly, the
-initial \texttt{indri} (raw data) needs to be provided. The following
-code tests on five expressions (arithmetic mean, three weighting schemes
-each emphasizing on one dimension, and principal component analysis
-recommended weight):
+These questions can be answered using the function
+\texttt{swap\_exprs()} and \texttt{swap\_values()}. Both functions first
+use a set of arguments (\texttt{.var}, \texttt{.module}, \texttt{.step},
+and \texttt{.res}) to locate the operation to be tested. Then,
+\texttt{.expr} and \texttt{.values} will accept a list of expressions or
+a vector of values for testing. Lastly, the initial \texttt{indri} (raw
+data) needs to be provided.
+
+\emph{What if an alternative dimension reduction expression is used to
+reduce the three dimensions into the HDI?}
+
+A common method to reduce the data dimension is through linear
+combination. With different weighting schemes, the combination can be
+compared to study the variable importance of each dimension or to reveal
+the underlying structures of countries. The following code tests five
+expressions (arithmetic mean, principal component analysis weight, and
+three weighting schemes emphasizing life expectancy, education, and GNI
+per capita respectively):
 
@@
 The resulting data includes five more columns corresponding to the five
-alternative index. The country ranking can be further computed on the
+alternative indexes. The country ranking can be further computed on the
 data and plotted in a scatterplot matrix as in
@@
 countries are printed in Table~\ref{tbl-hdi-expr} prints along with
-their three dimensions, calculated indexes and rankings. While having a
+their three dimensions, calculated indexes, and rankings. While having a
 high life expectancy and GNI \emph{per capita}, these countries
@@
 PCA produce similar rankings where countries are aligned in a diagnoal
-line, while country rankings varies when compare among \texttt{rank3} -
+line, while country rankings varies when compare among \texttt{rank3} to
 \texttt{rank5}. Panel \texttt{rank5} vs.~\texttt{rank4} highlights a
 cluster of countries that rank high by \texttt{rank5} but low by
-\texttt{rank4}. The plot demonstrates the necessity to tests on various
-alternative expressions to study the robustness of the method and to
+\texttt{rank4}. The plot demonstrates the necessity to test on various
+alternative expressions to verify the robustness of the method and to
 capture distinctive characteristics of different countries.}
@@
 
-The maximum and minimum limit is carefully chosen with justification
-provided in the technical notes:
+\newpage
+
+\emph{what if the minimum value for GNI per capita is changed?}
+
+The lower limit for GNI per capita is set to be the common log of \$100,
+with justification provided in the technical notes as:
 
@@
 
-What if this lower minimum is raised or decreased?
+The open source data provided by United Nations Development Programme
+suggest the minimum GNI \emph{per capita} is \texttt{\$732} and this
+opens the questions on how the index/ranking will change if the lower
+minimum (\texttt{\$100}) is raised or decreased. We can answer this
+question by testing on a set of different values using \texttt{.values}
+in the function \texttt{swap\_values()}:
 
@@
 \begin{Highlighting}[]
-\NormalTok{res3 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
+\NormalTok{(res3 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
   \FunctionTok{swap\_values}\NormalTok{(}
@@
     \AttributeTok{.var =}\NormalTok{ min, }\AttributeTok{.values =} \FunctionTok{log10}\NormalTok{(}\DecValTok{700}\NormalTok{),}
-    \AttributeTok{.raw\_data =}\NormalTok{ dt)}
+    \AttributeTok{.raw\_data =}\NormalTok{ dt)) }
 \end{Highlighting}
 \end{Shaded}
+
+\begin{verbatim}
+Index pipeline:
+
+Summary: 
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `log(gni_pc, base = 10)` -> gni_pc
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = life_exp, min = 20, max = 85)` -> life_exp
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = exp_sch, min = 0, max = 18)` -> exp_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = avg_sch, min = 0, max = 15)` -> avg_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = gni_pc, min = 2, max = 4.875061)` -> gni_pc0
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(exp_sch + avg_sch)/2` -> sch
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(life_exp * sch * gni_pc0)^(1/3)` -> index0
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = gni_pc, min = 2.84509804001426, max = 4.875061)` ->
+gni_pc1
+\end{verbatim}
+
+\begin{verbatim}
+dim_red: `(life_exp * sch * gni_pc1)^(1/3)` -> index1
+\end{verbatim}
+
+\begin{verbatim}
+
+Data: 
+# A tibble: 191 x 10
+      id country              life_~1 exp_sch avg_sch gni_pc gni_pc0 index0 gni_pc1 index1
+   <dbl> <chr>                  <dbl>   <dbl>   <dbl>  <dbl>   <dbl>  <dbl>   <dbl>  <dbl>
+ 1     1 Switzerland            0.984   0.917   0.924   4.83   0.983  0.962   0.976  0.960
+ 2     2 Norway                 0.973   1.01    0.867   4.81   0.978  0.963   0.968  0.960
+ 3     3 Iceland                0.964   1.06    0.918   4.75   0.955  0.970   0.937  0.964
+ 4     4 Hong Kong, China (S~   1.01    0.960   0.815   4.80   0.973  0.954   0.961  0.951
+ 5     5 Australia              0.993   1.17    0.848   4.69   0.936  0.979   0.910  0.970
+ 6     6 Denmark                0.944   1.04    0.864   4.78   0.967  0.954   0.954  0.950
+ 7     7 Sweden                 0.969   1.08    0.841   4.74   0.952  0.960   0.932  0.953
+ 8     8 Ireland                0.954   1.05    0.772   4.88   1.00   0.955   1.00   0.956
+ 9     9 Germany                0.933   0.945   0.939   4.74   0.952  0.942   0.932  0.936
+10    10 Netherlands            0.949   1.04    0.839   4.75   0.956  0.948   0.937  0.942
+# ... with 181 more rows, and abbreviated variable name 1: life_exp
+\end{verbatim}
+
+\newpage
+
+The function computes the GNI \emph{per capita} in both the original
+value (\texttt{gni\_pc0}) and the new value provided
+(\texttt{gni\_pc1}). All the subsequent operations are also computed for
+both cases and result in \texttt{index0} and \texttt{index1}.
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/unnamed-chunk-8-1.pdf}
+
+}
+
+\end{figure}
 

### 2023-02-09 paper/paper.qmd
some build up on compute_indexes() example

@@
 
-Consider a commonly used drought index: Standardized Precipitation-Evapotranspiration Index (SPEI)  [@spei].
-Its construction involves: 1) calculating potential evapotranspiration (PET) from average temperature, `tavg`, and its difference, `d`, with precipitation, `prcp`, 2) aggregating difference series with a time scale, 3) fitting the aggregated series with a chosen distribution to get density values, and 4) converting the density values to normal quantiles.
-Under the pipeline approach, SPEI will be constructed as:
+Consider a commonly used drought index: Standardized Precipitation-Evapotranspiration Index (SPEI)  [@spei]. Its construction involves:
+
+  1) transform the average temperature (`TMED`) into potential evapotranspiration (`pet`) 
+  2) combine precipitation (`prcp`) and potential evapotranspiration (`pet`) into a single variable `diff`
+  3) aggregate the difference series with a sliding window (`.scale`) 
+  4) fit a distribution to the aggregated series, and 
+  5) derive the index value from the normal density values.
+
+Conventionally approach may combine all these steps into a single function, with some level of modularity. However, these modules may only work for the selected index offered by the package.
+
+Under the pipeline approach, analysts first need to identify which module each step belongs to.
+
+Below shows the pseudocode for constructing SPEI with the pipeline:
 
@@
 DATA %>% 
-  trans_pet(method = thornthwaite, var = tavg) %>% 
-  aggregate(scale = 12, var = d) %>% 
-  normalise(dist = gamma, fit_method = lmom) %>% 
-  augment(var = .agg_prcp)
+  var_trans(.method = thornthwaite, Tave = TMED, ..., .new_name = "pet") %>%
+  dim_red(diff = prcp - pet) %>%
+  aggregate(.var = diff, .scale = 12, .new_name = "agg") %>%
+  dist_fit(.method = "lmoms", .var = agg, .dist = DIST) %>%
+  augment(.var = agg)
 ```
@@
 
-*Change the examples to two scales values, compare how indexes look like*
+The pipeline construct allows for multiple `.scales` and `.dist` to be evaluated in `aggregate()` and `dist_fit()` to compare index under different parameterisations.
+The result can then be passed into the `ggplot2` to crease visualisation.
+@fig-toy-example compares the SPEI calculated with two distributions (log-logistic and pearson III).
 
-Here each command corresponds to one step described above with additional arguments specific to the step. 
-The pipeline construction produces the same index value as command like `spei(...)`, as shown in @fig-toy-example, but with additional benefit:
+Apart from evaluating multiples parameters, the pipeline approach allows the the steps written for one index can be directly extrapolate to another index building within the pipeline.
+The flexibility of the pipeline also integrate well with other existing packages, for examples, fitting distributions using L-moment is commonly used when constructing drought indexes.
+The package `lmomco` provides general L-moment fits to a wide range of distributions and users can easily access to all the distributions within the pipeline.
 
-  - Multiple scales and multiple distributions can be fitted using the `c(...)` syntax to compare index values constructed from different parameterisations; 
-  - Intermediate results can be checked after each step; and
-  - Additional steps and analysis can be wired into the workflow for index diagnostics and customised user need.
+<!-- additional benefit: -->
 
-  *everyone is providing single command function, this is also the same for spi*
-
-  *may not worthwhile if only one index, as long as start changing*
-
-  *just want to calculate index, single function is fine*
+<!--   - Multiple scales and multiple distributions can be fitted using the `c(...)` syntax to compare index values constructed from different parameterisations;  -->
+<!--   - Intermediate results can be checked after each step; and -->
+<!--   - Additional steps and analysis can be wired into the workflow for index diagnostics and customised user need. -->
 
@@
 #| fig-align: center
-#| fig-cap: "Standardised Precipitation Evapotranspiration Index (SPEI) calculated with two distribution fits in Step 3 described above: generalised logistic (`glogist`) and generalised extreme value (`gev`) distribution, using the `wichita` data from the package `SPEI`."
-#| out-height: 30%
-#| out-width: 100%
-# library(indri)
-# library(dplyr)
-# library(lubridate)
-# library(lmomco)
-# library(ggplot2)
-# library(tsibble)
-# library(SPEI)
-# data(wichita)
-# 
-# dt <- wichita %>% 
-#   aggregate(var = prcp, scale = c(6, 12), index = ym, id = id, gran = month) %>%
-#   filter(!is.na(.agg)) %>%
-#   normalise(dist = list(gev(), loglogistic()), var = .agg) %>%
-#   augment(col = .agg)
-# 
-# dt %>% 
-#   ggplot(aes(x = ym, y = .index, color = .dist)) + 
-#   geom_line()
-knitr::include_graphics(here::here("figures/toy-example-spei.png"))
+#| fig-width: 10
+#| fig-height: 3
+#| fig-cap: "Standardised Precipitation Evapotranspiration Index (SPEI) calculated with two distribution fits in Step 3 described above: generalised logistic and pearson III distribution, using the `wichita` data from the package `SPEI`."
+library(indri)
+library(lmomco)
+library(ggplot2)
+library(tsibble)
+library(SPEI)
+library(tidyverse)
+library(lubridate)
+data(wichita)
+
+dt <- wichita %>%
+  mutate(ym = make_yearmonth(YEAR, MONTH), id = 1) %>% 
+  init(id = id, time = ym, indicators = PRCP:ACSH) %>%
+  var_trans(.method = thornthwaite, Tave = TMED, lat = 37.6475, .new_name = "pet") %>%
+  dim_red(diff = PRCP - pet) %>%
+  aggregate(.var = diff, .scale = 12) %>%
+  dist_fit(.dist = list(loglogistic(), pearsonIII()), .method = "lmoms", .var = .agg) %>%
+  augment(.var = .agg)
+
+dt$data %>%
+  mutate(.dist = ifelse(.dist == "parglo", "Log-logistic", "PearsonIII")) %>% 
+  ggplot(aes(x = ym, y = .index, color = .dist)) +
+  geom_line() + 
+   theme_bw() +
+  theme(
+    panel.grid = element_blank()
+  ) +
+  scale_x_yearmonth(name = "Year", date_break = "2 years", date_label = "%Y") +
+  scale_color_brewer(palette = "Dark2", name = "distribution") + 
+  ylab("SPEI")
+
 ```
@@
 
+\newpage
+
 # Examples {#sec-examples}
@@
 ## A drought index example
+
+Drought, as a complex natural hazard event, is difficult to measure.
+
+There is no uniform definition on drought [@ref], a range of drought indexes are usually used simultaneously to quantify the severity of a drought event.
+
+In the package PACKAGE, a collection of drought indexes have been implemented, including the most commonly used SPI and SPEI (`idx_spi()` and `idx_spei()`).
+They can be computed collectively on a dataset under `compute_indexes()`:
+
+```{r eval = FALSE}
+DATA %>% 
+  init(id = ..., time = ...) %>% 
+  compute_indexes(
+    spei = idx_spei(...),
+    spi = idx_spei(...)
+  )
+```
+
+The individual index functions are convenient wrappers around the pipeline building blocks.
+
+For SPI and SPEI, the wrapper functions also allow for multiple inputs of scales and distribution parameters (`.scale` and `.dist`) to allow computation on multiple time scale and distribution fitting.
+The following code compute the two indexes (SPI and SPEI) for different time scales (6, 12, 18, 24, 30, 36) and different distributions (Pearson III, log-logistic, lognormal, and general extreme value distributions) for [NUMBER] weather stations in Eastern Australia shown in @fig-station-map: (TODO: should talk about the data source)
+
+```{r eval = FALSE}
+scale <- c(6, 12, 18, 24, 30, 36)
+dist_spei <- c(gev(), loglogistic(), pearsonIII(), lognormal())
+dist_spi <- c(gamma(), weibull())
+dt %>% 
+  compute_indexes(
+    spei = idx_spei(
+      .pet_method = "thornthwaite", .tavg = tavg, .lat = lat, 
+      .scale = scale, .dist = dist_spei),
+    spi = idx_spi(.scale = scale,  .dist = dist_spi)
+  )
+```
+
+```{r fig-station-map}
+#| fig-cap: "..."
+#| echo: false
+
+map_dt <- ozmaps::abs_ste %>%
+  filter(NAME %in% c("New South Wales", "Victoria", "Queensland")) %>%
+  rmapshaper::ms_simplify(keep = 0.01)
+
+map_dt %>%
+  ggplot() +
+  geom_sf(fill = "transparent") +
+  theme_void() 
+  #geom_point(data = , aes(x = long, y = lat))
+```
 
@@
 #| fig-width: 12
-#| fig-cap: "Scatterplot matrix of country rankings computed from different dimension reduction expressions in HDI: geometric mean (`rank0`), arithmetic mean (`rank1`), principal component analysis recommended weights (`rank2`), and heavier weight on life expectancy, education, and GNI per capita (`rank3` - `rank5`).
+#| fig-cap: "Comparing country ranks from six different index constructions: geometric mean (`rank0`), arithmetic mean (`rank1`), principal component analysis recommended weights (`rank2`), and heavier weight on life expectancy, education, and GNI per capita (`rank3` - `rank5`).
 Geometric mean, arithmetic mean, and PCA produce similar rankings where countries are aligned in a diagnoal line, while country rankings varies when compare among `rank3` to `rank5`.
@@
 with_highlight <- with_rank %>% 
-  mutate(highlight = ifelse(country %in% lower_cluster$country, TRUE, FALSE))
+  mutate(highlight = ifelse(country %in% lower_cluster$country, TRUE, FALSE)) %>% 
+  rename(`geometric mean` = rank0, `arithmetic mean` = rank1,
+         `PCA weights` = rank2, `weight 0.8 on life expec.` = rank3,
+         `weight 0.8 on education`  = rank4, `weight 0.8 on GNI per capita` = rank5)
 
@@
                discrete = "facetbar", na = "na"),
-  upper = NULL, diag = NULL, switch = "both") +
+  upper = NULL, diag = NULL,  switch = "both", 
+  labeller = label_wrap_gen(width = 15, multi_line = TRUE) ) +
   theme(aspect.ratio = 1)
@@
 
-\newpage
+<!-- \newpage -->
 
-*what if the minimum value for GNI per capita is changed?*
+<!-- *what if the minimum value for GNI per capita is changed?* -->
 
-The lower limit for GNI per capita is set to be the common log of $100, with justification provided in the technical notes as:
+<!-- The lower limit for GNI per capita is set to be the common log of $100, with justification provided in the technical notes as: -->
 
-  > The low minimum value for gross national income (GNI) per capita, $100, is justified by the considerable amount of unmeasured subsistence and nonmarket production in economies close to the minimum, which is not captured in the official data
+<!--   > The low minimum value for gross national income (GNI) per capita, $100, is justified by the considerable amount of unmeasured subsistence and nonmarket production in economies close to the minimum, which is not captured in the official data -->
 
-The open source data provided by United Nations Development Programme suggest the minimum GNI *per capita* is `$732` and this opens the questions on how the index/ranking will change if the lower minimum (`$100`) is raised or decreased.
-We can answer this question by testing on a set of different values using `.values` in the function `swap_values()`:
+<!-- The open source data provided by United Nations Development Programme suggest the minimum GNI *per capita* is `$732` and this opens the questions on how the index/ranking will change if the lower minimum (`$100`) is raised or decreased.
+We can answer this question by testing on a set of different values using `.values` in the function `swap_values()`: -->
 
-```{r echo = TRUE, message = TRUE}
-(res3 <- res %>%
-  swap_values(
-    .module = var_trans, .step = rescale_minmax, .res = gni_pc,
-    .var = min, .values = log10(700),
-    .raw_data = dt)) 
-```
+<!-- ```{r echo = TRUE, message = TRUE} -->
+<!-- (res3 <- res %>% -->
+<!--   swap_values( -->
+<!--     .module = var_trans, .step = rescale_minmax, .res = gni_pc, -->
+<!--     .var = max, .values = log10(150000), -->
+<!--     .raw_data = dt) -->
+<!--    )  -->
+<!-- ``` -->
 
-\newpage
+<!-- \newpage -->
 
-The function computes the GNI *per capita* in both the original value (`gni_pc0`) and the new value provided (`gni_pc1`).
+<!-- The function computes the GNI *per capita* in both the original value (`gni_pc0`) and the new value provided (`gni_pc1`).
 All the subsequent operations are also computed for both cases and result in `index0` and `index1`.
+-->
 
-```{r}
-#| fig-height: 6
-#| fig-width: 6
-#| fig-align: center
-new_rank <- res3$data %>%
-  mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>%
-  rename_with(~gsub(".index", "", .x), contains("rank.index")) %>% 
-  mutate(difference = rank1 - rank0)
+<!-- ```{r} -->
+<!-- #| fig-height: 6 -->
+<!-- #| fig-width: 6 -->
+<!-- #| fig-align: center -->
+<!-- new_rank <- res3$data %>% -->
+<!--   mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>% -->
+<!--   rename_with(~gsub(".index", "", .x), contains("rank.index")) %>%  -->
+<!--   mutate(difference = rank1 - rank0) -->
 
-try <- new_rank %>%
-  dplyr::select(id, country, contains("life_exp"), contains("index"), contains("rank"), difference) %>% 
-  arrange(-abs(difference))
+<!-- try <- new_rank %>% -->
+<!--   dplyr::select(id, country, contains("gni_pc"), contains("index"), contains("rank"), difference) %>%  -->
+<!--   arrange(-abs(difference)) -->
 
-new_rank %>% 
-  ggplot(aes(x = rank0, y = rank1)) + 
-  geom_point() + 
-  geom_abline(slope = 1, intercept = 0, color = "blue") +
-  # geom_text(data = try %>% filter(difference > 15), color = "red", aes(label = country)) +
-  theme(aspect.ratio = 1) +
-  xlab("log10(100) as the rescale minimum for GNI per capita") +
-  ylab("log10(700) as the rescale minimum for GNI per capita") 
-```
+<!-- new_rank %>%  -->
+<!--   ggplot(aes(x = rank0, y = rank1)) +  -->
+<!--   geom_point() +  -->
+<!--   geom_abline(slope = 1, intercept = 0, color = "blue") + -->
+<!--   # geom_text(data = try %>% filter(difference > 15), color = "red", aes(label = country)) + -->
+<!--   theme(aspect.ratio = 1) + -->
+<!--   xlab("log10(100) as the rescale minimum for GNI per capita") + -->
+<!--   ylab("log10(700) as the rescale minimum for GNI per capita")  -->
+<!-- ``` -->
 

### 2023-02-09 paper/paper.tex
some build up on compute_indexes() example

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 Precipitation-Evapotranspiration Index (SPEI) (Vicente-Serrano,
-Beguería, and López-Moreno 2010). Its construction involves: 1)
-calculating potential evapotranspiration (PET) from average temperature,
-\texttt{tavg}, and its difference, \texttt{d}, with precipitation,
-\texttt{prcp}, 2) aggregating difference series with a time scale, 3)
-fitting the aggregated series with a chosen distribution to get density
-values, and 4) converting the density values to normal quantiles. Under
-the pipeline approach, SPEI will be constructed as:
+Beguería, and López-Moreno 2010). Its construction involves:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\tightlist
+\item
+  transform the average temperature (\texttt{TMED}) into potential
+  evapotranspiration (\texttt{pet})
+\item
+  combine precipitation (\texttt{prcp}) and potential evapotranspiration
+  (\texttt{pet}) into a single variable \texttt{diff}
+\item
+  aggregate the difference series with a sliding window
+  (\texttt{.scale})
+\item
+  fit a distribution to the aggregated series, and
+\item
+  derive the index value from the normal density values.
+\end{enumerate}
+
+Conventionally approach may combine all these steps into a single
+function, with some level of modularity. However, these modules may only
+work for the selected index offered by the package.
+
+Under the pipeline approach, analysts first need to identify which
+module each step belongs to.
+
+Below shows the pseudocode for constructing SPEI with the pipeline:
 
@@
 \NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{trans\_pet}\NormalTok{(}\AttributeTok{method =}\NormalTok{ thornthwaite, }\AttributeTok{var =}\NormalTok{ tavg) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{var =}\NormalTok{ d) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{normalise}\NormalTok{(}\AttributeTok{dist =}\NormalTok{ gamma, }\AttributeTok{fit\_method =}\NormalTok{ lmom) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{augment}\NormalTok{(}\AttributeTok{var =}\NormalTok{ .agg\_prcp)}
+  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ thornthwaite, }\AttributeTok{Tave =}\NormalTok{ TMED, ..., }\AttributeTok{.new\_name =} \StringTok{"pet"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{diff =}\NormalTok{ prcp }\SpecialCharTok{{-}}\NormalTok{ pet) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{.var =}\NormalTok{ diff, }\AttributeTok{.scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{.new\_name =} \StringTok{"agg"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{dist\_fit}\NormalTok{(}\AttributeTok{.method =} \StringTok{"lmoms"}\NormalTok{, }\AttributeTok{.var =}\NormalTok{ agg, }\AttributeTok{.dist =}\NormalTok{ DIST) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{augment}\NormalTok{(}\AttributeTok{.var =}\NormalTok{ agg)}
 \end{Highlighting}
@@
 
-\emph{Change the examples to two scales values, compare how indexes look
-like}
-
-Here each command corresponds to one step described above with
-additional arguments specific to the step. The pipeline construction
-produces the same index value as command like \texttt{spei(...)}, as
-shown in Figure~\ref{fig-toy-example}, but with additional benefit:
-
-\begin{itemize}
-\tightlist
-\item
-  Multiple scales and multiple distributions can be fitted using the
-  \texttt{c(...)} syntax to compare index values constructed from
-  different parameterisations;
-\item
-  Intermediate results can be checked after each step; and
-\item
-  Additional steps and analysis can be wired into the workflow for index
-  diagnostics and customised user need.
-\end{itemize}
-
-\emph{everyone is providing single command function, this is also the
-same for spi}
-
-\emph{may not worthwhile if only one index, as long as start changing}
-
-\emph{just want to calculate index, single function is fine}
+The pipeline construct allows for multiple \texttt{.scales} and
+\texttt{.dist} to be evaluated in \texttt{aggregate()} and
+\texttt{dist\_fit()} to compare index under different parameterisations.
+The result can then be passed into the \texttt{ggplot2} to crease
+visualisation. Figure~\ref{fig-toy-example} compares the SPEI calculated
+with two distributions (log-logistic and pearson III).
+
+Apart from evaluating multiples parameters, the pipeline approach allows
+the the steps written for one index can be directly extrapolate to
+another index building within the pipeline. The flexibility of the
+pipeline also integrate well with other existing packages, for examples,
+fitting distributions using L-moment is commonly used when constructing
+drought indexes. The package \texttt{lmomco} provides general L-moment
+fits to a wide range of distributions and users can easily access to all
+the distributions within the pipeline.
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=0.3\textheight]{../figures/toy-example-spei.png}
+{\centering \includegraphics{paper_files/figure-pdf/fig-toy-example-1.pdf}
 
@@
 Evapotranspiration Index (SPEI) calculated with two distribution fits in
-Step 3 described above: generalised logistic (\texttt{glogist}) and
-generalised extreme value (\texttt{gev}) distribution, using the
-\texttt{wichita} data from the package \texttt{SPEI}.}
+Step 3 described above: generalised logistic and pearson III
+distribution, using the \texttt{wichita} data from the package
+\texttt{SPEI}.}
 
@@
 
+\newpage
+
 \hypertarget{sec-examples}{%
@@
 \subsection{A drought index example}\label{a-drought-index-example}}
+
+Drought, as a complex natural hazard event, is difficult to measure.
+
+There is no uniform definition on drought (\textbf{ref?}), a range of
+drought indexes are usually used simultaneously to quantify the severity
+of a drought event.
+
+In the package PACKAGE, a collection of drought indexes have been
+implemented, including the most commonly used SPI and SPEI
+(\texttt{idx\_spi()} and \texttt{idx\_spei()}). They can be computed
+collectively on a dataset under \texttt{compute\_indexes()}:
+
+The individual index functions are convenient wrappers around the
+pipeline building blocks.
+
+For SPI and SPEI, the wrapper functions also allow for multiple inputs
+of scales and distribution parameters (\texttt{.scale} and
+\texttt{.dist}) to allow computation on multiple time scale and
+distribution fitting. The following code compute the two indexes (SPI
+and SPEI) for different time scales (6, 12, 18, 24, 30, 36) and
+different distributions (Pearson III, log-logistic, lognormal, and
+general extreme value distributions) for {[}NUMBER{]} weather stations
+in Eastern Australia shown in Figure~\ref{fig-station-map}: (TODO:
+should talk about the data source)
+
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/fig-station-map-1.pdf}
+
+}
+
+\caption{\label{fig-station-map}\ldots{}}
+
+\end{figure}
 
@@
  1     1 Switzerland               0.984   0.917   0.924  0.983 0.920 0.962
- 2     2 Norway                    0.973   1.01    0.867  0.978 0.939 0.963
- 3     3 Iceland                   0.964   1.06    0.918  0.955 0.991 0.970
- 4     4 Hong Kong, China (SAR)    1.01    0.960   0.815  0.973 0.887 0.954
- 5     5 Australia                 0.993   1.17    0.848  0.936 1.01  0.979
- 6     6 Denmark                   0.944   1.04    0.864  0.967 0.952 0.954
- 7     7 Sweden                    0.969   1.08    0.841  0.952 0.960 0.960
- 8     8 Ireland                   0.954   1.05    0.772  1.00  0.912 0.955
+ 2     2 Norway                    0.973   1       0.867  0.978 0.933 0.961
+ 3     3 Iceland                   0.964   1       0.918  0.955 0.959 0.959
+ 4     4 Hong Kong, China (SAR)    1       0.960   0.815  0.973 0.887 0.952
+ 5     5 Australia                 0.993   1       0.848  0.936 0.924 0.951
+ 6     6 Denmark                   0.944   1       0.864  0.967 0.932 0.948
+ 7     7 Sweden                    0.969   1       0.841  0.952 0.920 0.947
+ 8     8 Ireland                   0.954   1       0.772  1     0.886 0.945
  9     9 Germany                   0.933   0.945   0.939  0.952 0.942 0.942
-10    10 Netherlands               0.949   1.04    0.839  0.956 0.939 0.948
+10    10 Netherlands               0.949   1       0.839  0.956 0.919 0.941
 # ... with 181 more rows
@@
  1     1 Switzerl~   0.984   0.917   0.924  0.983 0.920  0.962  0.963   1.67  0.978  0.933
- 2     2 Norway      0.973   1.01    0.867  0.978 0.939  0.963  0.963   1.67  0.970  0.946
- 3     3 Iceland     0.964   1.06    0.918  0.955 0.991  0.970  0.970   1.68  0.966  0.985
- 4     4 Hong Kon~   1.01    0.960   0.815  0.973 0.887  0.954  0.956   1.65  0.992  0.908
- 5     5 Australia   0.993   1.17    0.848  0.936 1.01   0.979  0.979   1.69  0.989  1.00 
- 6     6 Denmark     0.944   1.04    0.864  0.967 0.952  0.954  0.954   1.65  0.947  0.953
- 7     7 Sweden      0.969   1.08    0.841  0.952 0.960  0.960  0.960   1.66  0.966  0.960
- 8     8 Ireland     0.954   1.05    0.772  1.00  0.912  0.955  0.956   1.66  0.955  0.925
+ 2     2 Norway      0.973   1       0.867  0.978 0.933  0.961  0.961   1.66  0.969  0.942
+ 3     3 Iceland     0.964   1       0.918  0.955 0.959  0.959  0.959   1.66  0.963  0.959
+ 4     4 Hong Kon~   1       0.960   0.815  0.973 0.887  0.952  0.953   1.65  0.986  0.907
+ 5     5 Australia   0.993   1       0.848  0.936 0.924  0.951  0.951   1.65  0.980  0.932
+ 6     6 Denmark     0.944   1       0.864  0.967 0.932  0.948  0.948   1.64  0.945  0.937
+ 7     7 Sweden      0.969   1       0.841  0.952 0.920  0.947  0.947   1.64  0.962  0.928
+ 8     8 Ireland     0.954   1       0.772  1     0.886  0.945  0.947   1.64  0.952  0.904
  9     9 Germany     0.933   0.945   0.939  0.952 0.942  0.942  0.942   1.63  0.936  0.942
-10    10 Netherla~   0.949   1.04    0.839  0.956 0.939  0.948  0.948   1.64  0.949  0.941
+10    10 Netherla~   0.949   1       0.839  0.956 0.919  0.941  0.941   1.63  0.947  0.926
 # ... with 181 more rows, 1 more variable: index5 <dbl>, and abbreviated variable name
@@
 
-\caption{\label{fig-hdi-expr}Scatterplot matrix of country rankings
-computed from different dimension reduction expressions in HDI:
-geometric mean (\texttt{rank0}), arithmetic mean (\texttt{rank1}),
-principal component analysis recommended weights (\texttt{rank2}), and
-heavier weight on life expectancy, education, and GNI per capita
-(\texttt{rank3} - \texttt{rank5}). Geometric mean, arithmetic mean, and
-PCA produce similar rankings where countries are aligned in a diagnoal
-line, while country rankings varies when compare among \texttt{rank3} to
-\texttt{rank5}. Panel \texttt{rank5} vs.~\texttt{rank4} highlights a
-cluster of countries that rank high by \texttt{rank5} but low by
-\texttt{rank4}. The plot demonstrates the necessity to test on various
-alternative expressions to verify the robustness of the method and to
-capture distinctive characteristics of different countries.}
+\caption{\label{fig-hdi-expr}Comparing country ranks from six different
+index constructions: geometric mean (\texttt{rank0}), arithmetic mean
+(\texttt{rank1}), principal component analysis recommended weights
+(\texttt{rank2}), and heavier weight on life expectancy, education, and
+GNI per capita (\texttt{rank3} - \texttt{rank5}). Geometric mean,
+arithmetic mean, and PCA produce similar rankings where countries are
+aligned in a diagnoal line, while country rankings varies when compare
+among \texttt{rank3} to \texttt{rank5}. Panel \texttt{rank5}
+vs.~\texttt{rank4} highlights a cluster of countries that rank high by
+\texttt{rank5} but low by \texttt{rank4}. The plot demonstrates the
+necessity to test on various alternative expressions to verify the
+robustness of the method and to capture distinctive characteristics of
+different countries.}
 
@@
 Andorra & 0.929 & 0.721 & 0.942 & 0.909 & 0.764 & 32 & 69 \\
-Qatar & 0.912 & 0.684 & 1.023 & 0.900 & 0.741 & 34 & 83 \\
-San Marino & 0.937 & 0.701 & 0.947 & 0.914 & 0.749 & 31 & 76 \\
+Qatar & 0.912 & 0.684 & 1.000 & 0.898 & 0.739 & 34 & 84 \\
+San Marino & 0.937 & 0.701 & 0.947 & 0.914 & 0.749 & 30 & 76 \\
 Kuwait & 0.903 & 0.670 & 0.947 & 0.884 & 0.721 & 39 & 96 \\
@@
 \end{longtable}
-
-\newpage
-
-\emph{what if the minimum value for GNI per capita is changed?}
-
-The lower limit for GNI per capita is set to be the common log of \$100,
-with justification provided in the technical notes as:
-
-\begin{quote}
-The low minimum value for gross national income (GNI) per capita, \$100,
-is justified by the considerable amount of unmeasured subsistence and
-nonmarket production in economies close to the minimum, which is not
-captured in the official data
-\end{quote}
-
-The open source data provided by United Nations Development Programme
-suggest the minimum GNI \emph{per capita} is \texttt{\$732} and this
-opens the questions on how the index/ranking will change if the lower
-minimum (\texttt{\$100}) is raised or decreased. We can answer this
-question by testing on a set of different values using \texttt{.values}
-in the function \texttt{swap\_values()}:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{(res3 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{swap\_values}\NormalTok{(}
-    \AttributeTok{.module =}\NormalTok{ var\_trans, }\AttributeTok{.step =}\NormalTok{ rescale\_minmax, }\AttributeTok{.res =}\NormalTok{ gni\_pc,}
-    \AttributeTok{.var =}\NormalTok{ min, }\AttributeTok{.values =} \FunctionTok{log10}\NormalTok{(}\DecValTok{700}\NormalTok{),}
-    \AttributeTok{.raw\_data =}\NormalTok{ dt)) }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-Index pipeline:
-
-Summary: 
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `log(gni_pc, base = 10)` -> gni_pc
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = life_exp, min = 20, max = 85)` -> life_exp
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = exp_sch, min = 0, max = 18)` -> exp_sch
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = avg_sch, min = 0, max = 15)` -> avg_sch
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = gni_pc, min = 2, max = 4.875061)` -> gni_pc0
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `(exp_sch + avg_sch)/2` -> sch
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `(life_exp * sch * gni_pc0)^(1/3)` -> index0
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = gni_pc, min = 2.84509804001426, max = 4.875061)` ->
-gni_pc1
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `(life_exp * sch * gni_pc1)^(1/3)` -> index1
-\end{verbatim}
-
-\begin{verbatim}
-
-Data: 
-# A tibble: 191 x 10
-      id country              life_~1 exp_sch avg_sch gni_pc gni_pc0 index0 gni_pc1 index1
-   <dbl> <chr>                  <dbl>   <dbl>   <dbl>  <dbl>   <dbl>  <dbl>   <dbl>  <dbl>
- 1     1 Switzerland            0.984   0.917   0.924   4.83   0.983  0.962   0.976  0.960
- 2     2 Norway                 0.973   1.01    0.867   4.81   0.978  0.963   0.968  0.960
- 3     3 Iceland                0.964   1.06    0.918   4.75   0.955  0.970   0.937  0.964
- 4     4 Hong Kong, China (S~   1.01    0.960   0.815   4.80   0.973  0.954   0.961  0.951
- 5     5 Australia              0.993   1.17    0.848   4.69   0.936  0.979   0.910  0.970
- 6     6 Denmark                0.944   1.04    0.864   4.78   0.967  0.954   0.954  0.950
- 7     7 Sweden                 0.969   1.08    0.841   4.74   0.952  0.960   0.932  0.953
- 8     8 Ireland                0.954   1.05    0.772   4.88   1.00   0.955   1.00   0.956
- 9     9 Germany                0.933   0.945   0.939   4.74   0.952  0.942   0.932  0.936
-10    10 Netherlands            0.949   1.04    0.839   4.75   0.956  0.948   0.937  0.942
-# ... with 181 more rows, and abbreviated variable name 1: life_exp
-\end{verbatim}
-
-\newpage
-
-The function computes the GNI \emph{per capita} in both the original
-value (\texttt{gni\_pc0}) and the new value provided
-(\texttt{gni\_pc1}). All the subsequent operations are also computed for
-both cases and result in \texttt{index0} and \texttt{index1}.
-
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-pdf/unnamed-chunk-8-1.pdf}
-
-}
-
-\end{figure}
 

### 2023-02-11 paper/paper.qmd
add example on calculating indexes + two captions

@@
 library(indri)
+library(lmomco)
+library(lubridate)
+library(SPEI)
+library(tsibble)
 library(GGally)
+library(patchwork)
 ```
@@
 
-```{r eval = FALSE}
+```{r eval = FALSE, echo = TRUE}
 DATA %>% 
@@
 For SPI and SPEI, the wrapper functions also allow for multiple inputs of scales and distribution parameters (`.scale` and `.dist`) to allow computation on multiple time scale and distribution fitting.
-The following code compute the two indexes (SPI and SPEI) for different time scales (6, 12, 18, 24, 30, 36) and different distributions (Pearson III, log-logistic, lognormal, and general extreme value distributions) for [NUMBER] weather stations in Eastern Australia shown in @fig-station-map: (TODO: should talk about the data source)
 
-```{r eval = FALSE}
-scale <- c(6, 12, 18, 24, 30, 36)
-dist_spei <- c(gev(), loglogistic(), pearsonIII(), lognormal())
-dist_spi <- c(gamma(), weibull())
-dt %>% 
+The climate data is sourced from Global Historical Climatology Network Daily (GHCND).
+
+stations in Queensland, Australia. 
+With precipitation, maximum and minimum temperature available since January 1990.
+
+This gives `r length(unique(queensland$id))` stations.
+
+The following code compute the two indexes (SPI and SPEI) at different time scales (6, 12, 24, 36) and different distributions (log-logistic and general extreme value distributions for SPEI and gamma for SPI).
+
+```{r cache=TRUE, echo = TRUE}
+.scale <- c(6, 12, 24, 36)
+(idx <- queensland %>%
+  init(id = id, time = ym) %>%
   compute_indexes(
@@
       .pet_method = "thornthwaite", .tavg = tavg, .lat = lat,
-      .scale = scale, .dist = dist_spei),
-    spi = idx_spi(.scale = scale,  .dist = dist_spi)
-  )
+      .scale = .scale, .dist = c(gev(), loglogistic())),
+    spi = idx_spi(.scale = .scale)
+  ))
 ```
 
-```{r fig-station-map}
-#| fig-cap: "..."
-#| echo: false
+The output from `compute_indexes()` contains index values and associated parameter in a long tibble.
+Analysis and visualisation can then be conducted to understand the effect of scale and distribution on the index value or explore the spatial and temporal pattern of wet/dry condition as in @fig-compute-spatial and @fig-compute-temporal.
 
-map_dt <- ozmaps::abs_ste %>%
-  filter(NAME %in% c("New South Wales", "Victoria", "Queensland")) %>%
-  rmapshaper::ms_simplify(keep = 0.01)
+```{r}
+library(sf)
+qld_lga <- absmapsdata::lga2018 %>%
+  filter(state_name_2016 == "Queensland") %>%
+  rmapshaper::ms_simplify(keep = 0.3) %>%
+  mutate(lga_name_2018 = str_replace(lga_name_2018, " \\([S|R|C|]\\)", ""),
+         lga_name_2018 = str_replace(lga_name_2018, " \\(Qld\\)", ""))
+```
 
-map_dt %>%
+<!-- maybe combine the two figures: Exploratory data analysis on the computed indexes: ??? -->
+
+```{r fig-compute-spatial}
+#| fig-cap: "Spatial distribution of Standardized Precipitation Index (SPI-12) in Queensland, Australia during two major flood and drought events: 2010/11 and 2019/20.
+The map shows a continous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 Janurary, likely due to the increased rainfall in February from the meteorological record."
+#| fig-width: 12
+#| fig-height: 5
+queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
+  rmapshaper::ms_simplify(keep = 0.02)
+queensland_map %>%
+  ggplot() +
+  geom_sf(fill = "transparent", color = "grey50", linewidth = 0.4) +
+  geom_point(
+    data = idx %>%
+      filter(.idx == "spi", .scale == 12) %>%
+      filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
+               ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )), 
+    aes(x = long, y = lat, color = .index)) +
+  scale_color_distiller(palette = "BrBG", direction = 1, name = "Index", ) +
+  theme_void() +
+  facet_wrap(vars(ym), ncol = 6)
+```
+
+```{r fig-compute-temporal}
+#| fig-cap: 'Time series plot of Standardized Precipitation-Evapotranspiration Index (SPEI) at the Texas post office station (highlighted by a diamond shape in panel a).
+The SPEI is calculated at four time scales (6, 12, 24, and 36 months) and fitted with two distributions (Log Logistic and GEV).
+The dashed line at -2 represents the class "extreme drought" by the SPEI.
+A larger time scale gives a smoother index series, while also takes longer to recover from an extreme situation as seen in the  2019/20 drought period.
+The SPEI values from two distribution fits mostly agree, while GEV can results in more extreme values, i.e.
+in 1998 and 2020.'
+#| fig-width: 12
+#| fig-height: 6
+texas <- idx %>%
+    filter(name == "TEXAS POST OFFICE", .idx == "spei") %>% distinct(long, lat)
+p1 <- queensland_map %>%
   ggplot() +
   geom_sf(fill = "transparent") +
+  geom_point(data = queensland %>%
+               distinct(id, long, lat, name),
+             aes(x = long, y = lat)) +
+  geom_point(data = texas, aes(x = long, y = lat),
+             color = "orange", shape = 18, fill = "orange", size = 4) +  
   theme_void()
-  #geom_point(data = , aes(x = long, y = lat))
+
+p2 <- idx %>%
+  filter(name == "TEXAS POST OFFICE", .idx == "spei") %>%
+  mutate(.dist = ifelse(.dist == "pargev", "GEV", "Log Logistic")) %>%
+  rename(scale = .scale) %>% 
+  ggplot(aes(x = ym, y = .index, color = .dist, group = .dist)) +
+  geom_line() +
+  theme_benchmark() +
+  facet_wrap(vars(scale), labeller = label_both, ncol = 1) +
+  scale_x_yearmonth(breaks = "2 year", date_labels = "%Y") +
+  scale_color_brewer(palette = "Dark2", name = "Distribution") + 
+  xlab("Year") + 
+  ylab("Index")
+
+(p1 | p2)  + 
+  patchwork::plot_annotation(tag_levels = "a") + 
+  patchwork::plot_layout(guides = "collect")  &
+  theme(legend.position = "bottom")
+
 ```
-
--   a basic workflow and congruence with results in the `SPEI` pkg
--   allow multiple distribution fit
--   allow bootstrap uncertainty
-
-easier to make comparison
-pick a publication, propose a new index that has this benefit but also correlate highly with the big name indexes
 

### 2023-02-11 paper/paper.tex
add example on calculating indexes + two captions

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners, frame hidden, enhanced, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ ..., }\AttributeTok{time =}\NormalTok{ ...) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{compute\_indexes}\NormalTok{(}
+    \AttributeTok{spei =} \FunctionTok{idx\_spei}\NormalTok{(...),}
+    \AttributeTok{spi =} \FunctionTok{idx\_spei}\NormalTok{(...)}
+\NormalTok{  )}
+\end{Highlighting}
+\end{Shaded}
+
 The individual index functions are convenient wrappers around the
@@
 \texttt{.dist}) to allow computation on multiple time scale and
-distribution fitting. The following code compute the two indexes (SPI
-and SPEI) for different time scales (6, 12, 18, 24, 30, 36) and
-different distributions (Pearson III, log-logistic, lognormal, and
-general extreme value distributions) for {[}NUMBER{]} weather stations
-in Eastern Australia shown in Figure~\ref{fig-station-map}: (TODO:
-should talk about the data source)
+distribution fitting.
+
+The climate data is sourced from Global Historical Climatology Network
+Daily (GHCND).
+
+stations in Queensland, Australia. With precipitation, maximum and
+minimum temperature available since January 1990.
+
+This gives 29 stations.
+
+The following code compute the two indexes (SPI and SPEI) at different
+time scales (6, 12, 24, 36) and different distributions (log-logistic
+and general extreme value distributions for SPEI and gamma for SPI).
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{.scale }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{6}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{24}\NormalTok{, }\DecValTok{36}\NormalTok{)}
+\NormalTok{(idx }\OtherTok{\textless{}{-}}\NormalTok{ queensland }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ id, }\AttributeTok{time =}\NormalTok{ ym) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{compute\_indexes}\NormalTok{(}
+    \AttributeTok{spei =} \FunctionTok{idx\_spei}\NormalTok{(}
+      \AttributeTok{.pet\_method =} \StringTok{"thornthwaite"}\NormalTok{, }\AttributeTok{.tavg =}\NormalTok{ tavg, }\AttributeTok{.lat =}\NormalTok{ lat,}
+      \AttributeTok{.scale =}\NormalTok{ .scale, }\AttributeTok{.dist =} \FunctionTok{c}\NormalTok{(}\FunctionTok{gev}\NormalTok{(), }\FunctionTok{loglogistic}\NormalTok{())),}
+    \AttributeTok{spi =} \FunctionTok{idx\_spi}\NormalTok{(}\AttributeTok{.scale =}\NormalTok{ .scale)}
+\NormalTok{  ))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+# A tibble: 128,586 x 19
+   .idx  .period id              ym  prcp  tmax  tmin  tavg  long   lat name    pet   diff
+   <chr>   <dbl> <chr>        <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <chr> <dbl>  <dbl>
+ 1 spei        6 ASN00029~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5 KOWA~  67.8  102. 
+ 2 spei        6 ASN00029~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5 KOWA~  67.8  102. 
+ 3 spei        6 ASN00029~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7 MOUN~  30.8  -30.8
+ 4 spei        6 ASN00029~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7 MOUN~  30.8  -30.8
+ 5 spei        6 ASN00031~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9 CAIR~  66.3  728. 
+ 6 spei        6 ASN00031~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9 CAIR~  66.3  728. 
+ 7 spei        6 ASN00031~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1 WALK~  48.4  456. 
+ 8 spei        6 ASN00031~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1 WALK~  48.4  456. 
+ 9 spei        6 ASN00032~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6 SOUT~  54.5 1915. 
+10 spei        6 ASN00032~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6 SOUT~  54.5 1915. 
+# ... with 128,576 more rows, and 6 more variables: .scale <dbl>, .agg <dbl>,
+#   .method <chr>, .fitted <dbl>, .dist <chr>, .index <dbl>
+\end{verbatim}
+
+The output from \texttt{compute\_indexes()} contains index values and
+associated parameter in a long tibble. Analysis and visualisation can
+then be conducted to understand the effect of scale and distribution on
+the index value or explore the spatial and temporal pattern of wet/dry
+condition as in Figure~\ref{fig-compute-spatial} and
+Figure~\ref{fig-compute-temporal}.
 
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-station-map-1.pdf}
+{\centering \includegraphics{paper_files/figure-pdf/fig-compute-spatial-1.pdf}
 
@@
 
-\caption{\label{fig-station-map}\ldots{}}
+\caption{\label{fig-compute-spatial}Spatial distribution of Standardized
+Precipitation Index (SPI-12) in Queensland, Australia during two major
+flood and drought events: 2010/11 and 2019/20. The map shows a continous
+wet period during the 2010/11 flood period and a mitigated drought
+situation, after its worst in 2019 December and 2020 Janurary, likely
+due to the increased rainfall in February from the meteorological
+record.}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  a basic workflow and congruence with results in the \texttt{SPEI} pkg
-\item
-  allow multiple distribution fit
-\item
-  allow bootstrap uncertainty
-\end{itemize}
-
-easier to make comparison pick a publication, propose a new index that
-has this benefit but also correlate highly with the big name indexes
+\begin{figure}
+
+{\centering \includegraphics{paper_files/figure-pdf/fig-compute-temporal-1.pdf}
+
+}
+
+\caption{\label{fig-compute-temporal}Time series plot of Standardized
+Precipitation-Evapotranspiration Index (SPEI) at the Texas post office
+station (highlighted by a diamond shape in panel a). The SPEI is
+calculated at four time scales (6, 12, 24, and 36 months) and fitted
+with two distributions (Log Logistic and GEV). The dashed line at -2
+represents the class ``extreme drought'' by the SPEI. A larger time
+scale gives a smoother index series, while also takes longer to recover
+from an extreme situation as seen in the 2019/20 drought period. The
+SPEI values from two distribution fits mostly agree, while GEV can
+results in more extreme values, i.e.~in 1998 and 2020.}
+
+\end{figure}
 

### 2023-02-12 paper/paper.qmd
write up example compute_indexes()

@@
 
-## How does the pipeline constructin of an index look like?
+## How does the pipeline constructin of an index look like? {#sec-toy-example}
 
@@
 
-Drought, as a complex natural hazard event, is difficult to measure.
+A common task for drought researchers is to compute indexes at different parameter combinations.
+This can be used to identify the spatial and temporal extent of drought events, recommend the best parameter choice, or compare the effectiveness of indexes for monitoring drought.
+The example below computes two indexes: SPI and SPEI, at various time scales and fitted distributions, for stations in the state of Queensland in Australia.
+The purpose of the example is to demonstrate the interfaces the PACKAGE package built to allow easy computing at different parameter combinations.
 
-There is no uniform definition on drought [@ref], a range of drought indexes are usually used simultaneously to quantify the severity of a drought event.
+The state of Queensland in Australia is frequently affected by natural disaster events such as flood and drought, which can have significant impacts on its agricultural industry.
+This study uses daily data from Global Historical Climatology Network Daily (GHCND), accessed via the package `rnoaa` to examine drought/flood condition in Queensland.
+Daily data is average into monthly and stations are excluded if monthly data contains missings, which is required for calculating both SPI and SPEI.
+This gives `r length(unique(queensland$id))` stations with complete records from 1990 January to 2022 April.
 
-In the package PACKAGE, a collection of drought indexes have been implemented, including the most commonly used SPI and SPEI (`idx_spi()` and `idx_spei()`).
-They can be computed collectively on a dataset under `compute_indexes()`:
-
-```{r eval = FALSE, echo = TRUE}
-DATA %>% 
-  init(id = ..., time = ...) %>% 
-  compute_indexes(
-    spei = idx_spei(...),
-    spi = idx_spei(...)
-  )
-```
-
-The individual index functions are convenient wrappers around the pipeline building blocks.
-
-For SPI and SPEI, the wrapper functions also allow for multiple inputs of scales and distribution parameters (`.scale` and `.dist`) to allow computation on multiple time scale and distribution fitting.
-
-The climate data is sourced from Global Historical Climatology Network Daily (GHCND).
-
-stations in Queensland, Australia. 
-With precipitation, maximum and minimum temperature available since January 1990.
-
-This gives `r length(unique(queensland$id))` stations.
-
-The following code compute the two indexes (SPI and SPEI) at different time scales (6, 12, 24, 36) and different distributions (log-logistic and general extreme value distributions for SPEI and gamma for SPI).
+The function `compute_indexes()` can be used to collectively compute multiple indexes.
+The PACKAGE offers wrapper functions, with the prefix idx_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function.
+For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise).
+Each `idx_xxx()` function specifies the relevant parameters relevant to the index: the thornthwaite method is used to calculate PET in SPEI, with the average temperature (`tavg`) and latitude (`lat`) used as inputs.
+The SPEI is computed at four time scales (6, 12, 24, and 36 months) and fitted with two distributions (Log-logistic and General Extreme Value (GEV)).
+The SPI is also computed at the same four time scales and uses the default gamma distribution to fit the aggregated series.
 
@@
 The output from `compute_indexes()` contains index values and associated parameter in a long tibble.
-Analysis and visualisation can then be conducted to understand the effect of scale and distribution on the index value or explore the spatial and temporal pattern of wet/dry condition as in @fig-compute-spatial and @fig-compute-temporal.
+It includes the original variables (`id`, `ym`, `prcp`, `tmax`, `tmin`, `tavg`, `long`, `lat`, and `name`), index parameters (`.idx`, `.scale`, `.method`, and `.dist`), intermediate variables (`pet`, `.agg`, and `.fitted`), and the final index (`.index`).
+This data can be visualised across space or time, or simultaneously, to explore the wet/dry condition in Queensland.
+@fig-compute-spatial visualises the spatial distribution of SPI at two periods (2010 October - 2011 March and 2019 October - 2020 March) with significant natural disaster events: 2010/11 Queensland flood and 2019 Australia drought, which contributes to the notorious 2019/20 bushfire.
+@fig-compute-temporal displays the sensitivity of the SPEI series for one particular station, Texas post office, at different time scales and fitted distributions.
+These two plots demonstrate some possibilities to explore the indexes after they are computed from `compute_indexes()`.
 
@@
 ```
-
-<!-- maybe combine the two figures: Exploratory data analysis on the computed indexes: ??? -->
 

### 2023-02-12 paper/paper.tex
write up example compute_indexes()

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners, frame hidden, enhanced, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{how-does-the-pipeline-constructin-of-an-index-look-like}{%
+\hypertarget{sec-toy-example}{%
 \subsection{How does the pipeline constructin of an index look
-like?}\label{how-does-the-pipeline-constructin-of-an-index-look-like}}
+like?}\label{sec-toy-example}}
 
@@
 
-Drought, as a complex natural hazard event, is difficult to measure.
-
-There is no uniform definition on drought (\textbf{ref?}), a range of
-drought indexes are usually used simultaneously to quantify the severity
-of a drought event.
-
-In the package PACKAGE, a collection of drought indexes have been
-implemented, including the most commonly used SPI and SPEI
-(\texttt{idx\_spi()} and \texttt{idx\_spei()}). They can be computed
-collectively on a dataset under \texttt{compute\_indexes()}:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ ..., }\AttributeTok{time =}\NormalTok{ ...) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{compute\_indexes}\NormalTok{(}
-    \AttributeTok{spei =} \FunctionTok{idx\_spei}\NormalTok{(...),}
-    \AttributeTok{spi =} \FunctionTok{idx\_spei}\NormalTok{(...)}
-\NormalTok{  )}
-\end{Highlighting}
-\end{Shaded}
-
-The individual index functions are convenient wrappers around the
-pipeline building blocks.
-
-For SPI and SPEI, the wrapper functions also allow for multiple inputs
-of scales and distribution parameters (\texttt{.scale} and
-\texttt{.dist}) to allow computation on multiple time scale and
-distribution fitting.
-
-The climate data is sourced from Global Historical Climatology Network
-Daily (GHCND).
-
-stations in Queensland, Australia. With precipitation, maximum and
-minimum temperature available since January 1990.
-
-This gives 29 stations.
-
-The following code compute the two indexes (SPI and SPEI) at different
-time scales (6, 12, 24, 36) and different distributions (log-logistic
-and general extreme value distributions for SPEI and gamma for SPI).
+A common task for drought researchers is to compute indexes at different
+parameter combinations. This can be used to identify the spatial and
+temporal extent of drought events, recommend the best parameter choice,
+or compare the effectiveness of indexes for monitoring drought. The
+example below computes two indexes: SPI and SPEI, at various time scales
+and fitted distributions, for stations in the state of Queensland in
+Australia. The purpose of the example is to demonstrate the interfaces
+the PACKAGE package built to allow easy computing at different parameter
+combinations.
+
+The state of Queensland in Australia is frequently affected by natural
+disaster events such as flood and drought, which can have significant
+impacts on its agricultural industry. This study uses daily data from
+Global Historical Climatology Network Daily (GHCND), accessed via the
+package \texttt{rnoaa} to examine drought/flood condition in Queensland.
+Daily data is average into monthly and stations are excluded if monthly
+data contains missings, which is required for calculating both SPI and
+SPEI. This gives 29 stations with complete records from 1990 January to
+2022 April.
+
+The function \texttt{compute\_indexes()} can be used to collectively
+compute multiple indexes. The PACKAGE offers wrapper functions, with the
+prefix idx\_, that simplify the calculation of commonly used indexes by
+combining a set of pipeline steps into a single function. For example,
+the function \texttt{idx\_spei()} includes the five steps previously
+described in Section~\ref{sec-toy-example} (variable transformation,
+dimension reduction, temporal aggregation, distribution fit, and
+normalise). Each \texttt{idx\_xxx()} function specifies the relevant
+parameters relevant to the index: the thornthwaite method is used to
+calculate PET in SPEI, with the average temperature (\texttt{tavg}) and
+latitude (\texttt{lat}) used as inputs. The SPEI is computed at four
+time scales (6, 12, 24, and 36 months) and fitted with two distributions
+(Log-logistic and General Extreme Value (GEV)). The SPI is also computed
+at the same four time scales and uses the default gamma distribution to
+fit the aggregated series.
 
@@
 The output from \texttt{compute\_indexes()} contains index values and
-associated parameter in a long tibble. Analysis and visualisation can
-then be conducted to understand the effect of scale and distribution on
-the index value or explore the spatial and temporal pattern of wet/dry
-condition as in Figure~\ref{fig-compute-spatial} and
-Figure~\ref{fig-compute-temporal}.
+associated parameter in a long tibble. It includes the original
+variables (\texttt{id}, \texttt{ym}, \texttt{prcp}, \texttt{tmax},
+\texttt{tmin}, \texttt{tavg}, \texttt{long}, \texttt{lat}, and
+\texttt{name}), index parameters (\texttt{.idx}, \texttt{.scale},
+\texttt{.method}, and \texttt{.dist}), intermediate variables
+(\texttt{pet}, \texttt{.agg}, and \texttt{.fitted}), and the final index
+(\texttt{.index}). This data can be visualised across space or time, or
+simultaneously, to explore the wet/dry condition in Queensland.
+Figure~\ref{fig-compute-spatial} visualises the spatial distribution of
+SPI at two periods (2010 October - 2011 March and 2019 October - 2020
+March) with significant natural disaster events: 2010/11 Queensland
+flood and 2019 Australia drought, which contributes to the notorious
+2019/20 bushfire. Figure~\ref{fig-compute-temporal} displays the
+sensitivity of the SPEI series for one particular station, Texas post
+office, at different time scales and fitted distributions. These two
+plots demonstrate some possibilities to explore the indexes after they
+are computed from \texttt{compute\_indexes()}.
 

### 2023-02-13 paper/paper.qmd
conclusion

@@
 
-\newpage
-
 # Examples {#sec-examples}
@@
 
-\newpage
+# Conclusion
 
-## Calculating indexes on raster data
+The paper presents a data pipeline with nine modules for constructing and analysing indexes.
+The pipeline increases transparency in the practice for index analysts to experiment with different index design and parameter choices to better design and apply their indexes.
+The significance of this work is its ability to provide a universal framework for index construction, which can be applied across different domains.
 
-<!-- Raster data is common data format for delivering large satellite spatio-temporal data.  -->
-
-<!-- Issue 1: variables may be collected at different spatial and temporal extent/ scale, different resolutions and granularities.  -->
-
-<!-- cleaning these variables involves spatial and temporal processing.  -->
-
-<!-- Issue 2: what the data the step is operated on?? -->
-
-<!-- When calculating on raster data, it is confusing on what data the operation is on.
-Is the operation operated per pixel or across regions, is the operation operated across all the time point, across year, across month etc.
--->
-
-<!-- While it may be clear to researchers themselves, but can be confusing for others who read the analysis for the first time.  -->
-
-<!-- The data structure should be made explicit at each operation on the spatio-temporal raster data, in terms of inputs and outputs. -->
-
-<!-- The raster examples  -->
+Examples have been given in the drought indexes and human development index to demonstrate computing of indexes with different parameters combinations and how alternative index design can provide insights to understand distinctive country characteristics that could sometimes be overlooked.
+The accompanied package, PACKAGE, is not meant to provide comprehensive implementation for all indexes across all domains.
+Instead, it demonstrates implementing individual pipeline steps that are versatile to multiple indexes and composing new indexes from existing steps.
+Domain experts are welcomed to adopt the pipeline approach to develop specialised packages for specific-domains indexes.
 

### 2023-02-13 paper/paper.tex
conclusion

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, frame hidden, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\newpage
-
 \hypertarget{sec-examples}{%
@@
 
-\newpage
-
-\hypertarget{calculating-indexes-on-raster-data}{%
-\subsection{Calculating indexes on raster
-data}\label{calculating-indexes-on-raster-data}}
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+The paper presents a data pipeline with nine modules for constructing
+and analysing indexes. The pipeline increases transparency in the
+practice for index analysts to experiment with different index design
+and parameter choices to better design and apply their indexes. The
+significance of this work is its ability to provide a universal
+framework for index construction, which can be applied across different
+domains.
+
+Examples have been given in the drought indexes and human development
+index to demonstrate computing of indexes with different parameters
+combinations and how alternative index design can provide insights to
+understand distinctive country characteristics that could sometimes be
+overlooked. The accompanied package, PACKAGE, is not meant to provide
+comprehensive implementation for all indexes across all domains.
+Instead, it demonstrates implementing individual pipeline steps that are
+versatile to multiple indexes and composing new indexes from existing
+steps. Domain experts are welcomed to adopt the pipeline approach to
+develop specialised packages for specific-domains indexes.
 

### 2023-02-27 paper/paper.qmd
change color palette - yellow in the middle for diverge; more edit in HDI example

@@
 
+<!-- # 1. unpack or the argumebt s  -->
+<!-- # 2. construct the syntax to evalute: ~rescale_minmax(life_exp, min = xxx, max = xxxx) -->
+<!-- # 3. evalute  -->
+<!-- # 4. update the data object  -->
+
 <!-- # [Extending the pipeline] a different name here: alternate a module with a different (method) -->
@@
 # Examples {#sec-examples}
+
+\newpage
 
@@
 The example below computes two indexes: SPI and SPEI, at various time scales and fitted distributions, for stations in the state of Queensland in Australia.
-The purpose of the example is to demonstrate the interfaces the PACKAGE package built to allow easy computing at different parameter combinations.
+The purpose of the example is to demonstrate the interfaces the tidyindex package built to allow easy computing at different parameter combinations.
 
@@
 The function `compute_indexes()` can be used to collectively compute multiple indexes.
-The PACKAGE offers wrapper functions, with the prefix idx_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function.
+The tidyindex offers wrapper functions, with the prefix idx_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function.
 For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise).
@@
     aes(x = long, y = lat, color = .index)) +
-  scale_color_distiller(palette = "BrBG", direction = 1, name = "Index", ) +
+  colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
   theme_void() +
@@
 ```
-
-\newpage
 
@@
 dt <- raw %>% init(id = country, indicators = life_exp:gni_pc)
-(res <- dt %>%
+res <- dt %>%
   var_trans(gni_pc = log(gni_pc, base = 10)) %>% 
@@
   dim_red(sch = (exp_sch + avg_sch) / 2) %>%
-  dim_red(index = (life_exp * sch * gni_pc)^(1/3)))
+  dim_red(index = (life_exp * sch * gni_pc)^(1/3))
 ```
 
-The output is an `indri` object with a summary of operations followed by the data output.
+<!-- The output is an `indri` object with a summary of operations followed by the data output.
 The data output contains the ID, country name, rescaled variables (`life_exp`, `exp_sch`, `avg_sch`, and `gni_pc`), the aggregated education dimension (`sch`), and the final index calculated (`index`).
+-->
 
-Index analysts can be interested in studying the property of the index and the countries through the "what-if" questions. In the context of HDI, this can be:
-
-  - what if arithmetic mean, or different weights, is used to reduce the three dimensions into HDI? 
-  - what if an alternative maximum or minimum value is used in the rescaling step?
-
-These questions can be answered using the function `swap_exprs()` and `swap_values()`.
-Both functions first use a set of arguments (`.var`, `.module`, `.step`, and `.res`) to locate the operation to be tested.
-Then, `.expr` and `.values` will accept a list of expressions or a vector of values for testing.
-Lastly, the initial `indri` (raw data) needs to be provided.
-
-*What if an alternative dimension reduction expression is used to reduce the three dimensions into the HDI?*
-
-A common method to reduce the data dimension is through linear combination.
+Index analysts can be interested in studying the property of the index and the countries through the "what-if" questions.
+In the context of HDI, this can be what if an alternative dimension reduction expression is used to reduce the three dimensions into the HDI?
+Apart from the geometrical mean, linear combination is a common method to reduce the data dimension.
 With different weighting schemes, the combination can be compared to study the variable importance of each dimension or to reveal the underlying structures of countries.
-The following code tests five expressions (arithmetic mean, principal component analysis weight, and three weighting schemes emphasizing life expectancy, education, and GNI per capita respectively):
+The following code tests five expressions (arithmetic mean, principal component analysis weight, and three weighting schemes emphasizing life expectancy, education, and GNI per capita respectively) using the `swap_exprs()` function.
+The function first locates the operation to be tested (`.var = index`), then accepts a list of expressions for testing `.expr`, before requesting the raw data:
 
@@
 
+\newpage
+
 # Conclusion
@@
 Examples have been given in the drought indexes and human development index to demonstrate computing of indexes with different parameters combinations and how alternative index design can provide insights to understand distinctive country characteristics that could sometimes be overlooked.
-The accompanied package, PACKAGE, is not meant to provide comprehensive implementation for all indexes across all domains.
+The accompanied package, tidyindex, is not meant to provide comprehensive implementation for all indexes across all domains.
 Instead, it demonstrates implementing individual pipeline steps that are versatile to multiple indexes and composing new indexes from existing steps.

### 2023-02-27 paper/paper.tex
change color palette - yellow in the middle for diverge; more edit in HDI example

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, frame hidden, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
+\newpage
+
 \hypertarget{a-drought-index-example}{%
@@
 Australia. The purpose of the example is to demonstrate the interfaces
-the PACKAGE package built to allow easy computing at different parameter
-combinations.
+the tidyindex package built to allow easy computing at different
+parameter combinations.
 
@@
 The function \texttt{compute\_indexes()} can be used to collectively
-compute multiple indexes. The PACKAGE offers wrapper functions, with the
-prefix idx\_, that simplify the calculation of commonly used indexes by
-combining a set of pipeline steps into a single function. For example,
-the function \texttt{idx\_spei()} includes the five steps previously
-described in Section~\ref{sec-toy-example} (variable transformation,
-dimension reduction, temporal aggregation, distribution fit, and
-normalise). Each \texttt{idx\_xxx()} function specifies the relevant
-parameters relevant to the index: the thornthwaite method is used to
-calculate PET in SPEI, with the average temperature (\texttt{tavg}) and
-latitude (\texttt{lat}) used as inputs. The SPEI is computed at four
-time scales (6, 12, 24, and 36 months) and fitted with two distributions
-(Log-logistic and General Extreme Value (GEV)). The SPI is also computed
-at the same four time scales and uses the default gamma distribution to
-fit the aggregated series.
+compute multiple indexes. The tidyindex offers wrapper functions, with
+the prefix idx\_, that simplify the calculation of commonly used indexes
+by combining a set of pipeline steps into a single function. For
+example, the function \texttt{idx\_spei()} includes the five steps
+previously described in Section~\ref{sec-toy-example} (variable
+transformation, dimension reduction, temporal aggregation, distribution
+fit, and normalise). Each \texttt{idx\_xxx()} function specifies the
+relevant parameters relevant to the index: the thornthwaite method is
+used to calculate PET in SPEI, with the average temperature
+(\texttt{tavg}) and latitude (\texttt{lat}) used as inputs. The SPEI is
+computed at four time scales (6, 12, 24, and 36 months) and fitted with
+two distributions (Log-logistic and General Extreme Value (GEV)). The
+SPI is also computed at the same four time scales and uses the default
+gamma distribution to fit the aggregated series.
 
@@
 
-\newpage
-
 \hypertarget{an-example-on-sustainable-development-indicators}{%
@@
 \NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
-\NormalTok{(res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
+\NormalTok{res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
   \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log}\NormalTok{(gni\_pc, }\AttributeTok{base =} \DecValTok{10}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
@@
   \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{)))}
+  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{))}
 \end{Highlighting}
@@
 
-\begin{verbatim}
-Index pipeline:
-
-Summary: 
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `log(gni_pc, base = 10)` -> gni_pc
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = life_exp, min = 20.000000, max = 85.000000)` -> life_exp
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = exp_sch, min = 0.000000, max = 18.000000)` -> exp_sch
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = avg_sch, min = 0.000000, max = 15.000000)` -> avg_sch
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = gni_pc, min = 2.000000, max = 4.875061)` -> gni_pc
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `(exp_sch + avg_sch)/2` -> sch
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index
-\end{verbatim}
-
-\begin{verbatim}
-
-Data: 
-# A tibble: 191 x 8
-      id country                life_exp exp_sch avg_sch gni_pc   sch index
-   <dbl> <chr>                     <dbl>   <dbl>   <dbl>  <dbl> <dbl> <dbl>
- 1     1 Switzerland               0.984   0.917   0.924  0.983 0.920 0.962
- 2     2 Norway                    0.973   1       0.867  0.978 0.933 0.961
- 3     3 Iceland                   0.964   1       0.918  0.955 0.959 0.959
- 4     4 Hong Kong, China (SAR)    1       0.960   0.815  0.973 0.887 0.952
- 5     5 Australia                 0.993   1       0.848  0.936 0.924 0.951
- 6     6 Denmark                   0.944   1       0.864  0.967 0.932 0.948
- 7     7 Sweden                    0.969   1       0.841  0.952 0.920 0.947
- 8     8 Ireland                   0.954   1       0.772  1     0.886 0.945
- 9     9 Germany                   0.933   0.945   0.939  0.952 0.942 0.942
-10    10 Netherlands               0.949   1       0.839  0.956 0.919 0.941
-# ... with 181 more rows
-\end{verbatim}
-
-The output is an \texttt{indri} object with a summary of operations
-followed by the data output. The data output contains the ID, country
-name, rescaled variables (\texttt{life\_exp}, \texttt{exp\_sch},
-\texttt{avg\_sch}, and \texttt{gni\_pc}), the aggregated education
-dimension (\texttt{sch}), and the final index calculated
-(\texttt{index}).
-
 Index analysts can be interested in studying the property of the index
 and the countries through the ``what-if'' questions. In the context of
-HDI, this can be:
-
-\begin{itemize}
-\tightlist
-\item
-  what if arithmetic mean, or different weights, is used to reduce the
-  three dimensions into HDI?
-\item
-  what if an alternative maximum or minimum value is used in the
-  rescaling step?
-\end{itemize}
-
-These questions can be answered using the function
-\texttt{swap\_exprs()} and \texttt{swap\_values()}. Both functions first
-use a set of arguments (\texttt{.var}, \texttt{.module}, \texttt{.step},
-and \texttt{.res}) to locate the operation to be tested. Then,
-\texttt{.expr} and \texttt{.values} will accept a list of expressions or
-a vector of values for testing. Lastly, the initial \texttt{indri} (raw
-data) needs to be provided.
-
-\emph{What if an alternative dimension reduction expression is used to
-reduce the three dimensions into the HDI?}
-
-A common method to reduce the data dimension is through linear
-combination. With different weighting schemes, the combination can be
+HDI, this can be what if an alternative dimension reduction expression
+is used to reduce the three dimensions into the HDI? Apart from the
+geometrical mean, linear combination is a common method to reduce the
+data dimension. With different weighting schemes, the combination can be
 compared to study the variable importance of each dimension or to reveal
@@
 three weighting schemes emphasizing life expectancy, education, and GNI
-per capita respectively):
+per capita respectively) using the \texttt{swap\_exprs()} function. The
+function first locates the operation to be tested
+(\texttt{.var\ =\ index}), then accepts a list of expressions for
+testing \texttt{.expr}, before requesting the raw data:
 
@@
 
+\newpage
+
 \hypertarget{conclusion}{%
@@
 understand distinctive country characteristics that could sometimes be
-overlooked. The accompanied package, PACKAGE, is not meant to provide
+overlooked. The accompanied package, tidyindex, is not meant to provide
 comprehensive implementation for all indexes across all domains.

### 2023-07-04 tidyindex.qmd
as tf article and restructure rpepo

--- 2023-02-27 paper/paper.qmd
@@
 format:
-  arxiv-pdf:
+  tandf-pdf:
     keep-tex: true  
@@
 library(patchwork)
-library(indri)
+library(tidyindex)
 library(lmomco)
@@
 
-```{r eval = FALSE, echo = TRUE}
+````
 DATA %>% 
@@
   augment(.var = agg)
-```
+````
 
@@
 
-```{r fig-toy-example, eval = TRUE, echo = FALSE}
+```{r}
+knitr::include_graphics(here::here("figures/toy-example-spei.png"))
+```
+
+```{r fig-toy-example, eval = FALSE, echo = FALSE}
 #| fig-align: center
@@
 #| fig-cap: "Standardised Precipitation Evapotranspiration Index (SPEI) calculated with two distribution fits in Step 3 described above: generalised logistic and pearson III distribution, using the `wichita` data from the package `SPEI`."
-library(indri)
+library(tidyindex)
 library(lmomco)
@@
   mutate(ym = make_yearmonth(YEAR, MONTH), id = 1) %>% 
-  init(id = id, time = ym, indicators = PRCP:ACSH) %>%
-  var_trans(.method = thornthwaite, Tave = TMED, lat = 37.6475, .new_name = "pet") %>%
+  init(id = id, time = ym) %>%
+  var_trans(.pet_method = "thornthwaite", .vars = TMED, .lat = 37.6475) %>%
   dim_red(diff = PRCP - pet) %>%
@@
 # Examples {#sec-examples}
-
-\newpage
 

### 2023-07-04 tidyindex.tex
as tf article and restructure rpepo

--- 2023-02-27 paper/paper.tex
@@
 \documentclass[
-]{article}
+]{interact}
 
@@
 
-\usepackage{arxiv}
 \usepackage{orcidlink}
-\usepackage{amsmath}
-\usepackage[T1]{fontenc}
 \makeatletter
@@
 \title{Demo arXiv template}
-\author{
-\textbf{H. Sherry Zhang}~\orcidlink{0000-0002-7122-1463}\\Department of
-Econometrics and Business Statistics\\Monash University\\Melbourne,
-VIC\\\href{mailto:huize.zhang@monash.edu}{huize.zhang@monash.edu}\\\\\\
-\textbf{Collaborators}\\Department of Econometrics and Business
-Statistics\\Monash University\\Melbourne, VIC\\}
-\date{}
+\author{H. Sherry
+Zhang$\textsuperscript{1}$~\orcidlink{0000-0002-7122-1463}, Collaborators$\textsuperscript{1}$}
+
+\thanks{CONTACT: H. Sherry
+Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
 \begin{document}
+\captionsetup{labelsep=space}
 \maketitle
+\textsuperscript{1} Department of Econometrics and Business
+Statistics, Monash University, Melbourne, VIC, Australia
 \begin{abstract}
@@
 \end{abstract}
-{\bfseries \emph Keywords}
-\def\sep{\textbullet\ }
-indexes \sep data pipeline \sep 
+\begin{keywords}
+\def\sep{;\ }
+indexes\sep data pipeline\sep 
 software design
-
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\end{keywords}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ thornthwaite, }\AttributeTok{Tave =}\NormalTok{ TMED, ..., }\AttributeTok{.new\_name =} \StringTok{"pet"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{diff =}\NormalTok{ prcp }\SpecialCharTok{{-}}\NormalTok{ pet) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{.var =}\NormalTok{ diff, }\AttributeTok{.scale =} \DecValTok{12}\NormalTok{, }\AttributeTok{.new\_name =} \StringTok{"agg"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{dist\_fit}\NormalTok{(}\AttributeTok{.method =} \StringTok{"lmoms"}\NormalTok{, }\AttributeTok{.var =}\NormalTok{ agg, }\AttributeTok{.dist =}\NormalTok{ DIST) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{augment}\NormalTok{(}\AttributeTok{.var =}\NormalTok{ agg)}
-\end{Highlighting}
-\end{Shaded}
+\begin{verbatim}
+DATA %>% 
+  var_trans(.method = thornthwaite, Tave = TMED, ..., .new_name = "pet") %>%
+  dim_red(diff = prcp - pet) %>%
+  aggregate(.var = diff, .scale = 12, .new_name = "agg") %>%
+  dist_fit(.method = "lmoms", .var = agg, .dist = DIST) %>%
+  augment(.var = agg)
+\end{verbatim}
 
@@
 The result can then be passed into the \texttt{ggplot2} to crease
-visualisation. Figure~\ref{fig-toy-example} compares the SPEI calculated
+visualisation. \textbf{?@fig-toy-example} compares the SPEI calculated
 with two distributions (log-logistic and pearson III).
@@
 
-\begin{figure}
-
-{\centering \includegraphics{paper_files/figure-pdf/fig-toy-example-1.pdf}
-
-}
-
-\caption{\label{fig-toy-example}Standardised Precipitation
-Evapotranspiration Index (SPEI) calculated with two distribution fits in
-Step 3 described above: generalised logistic and pearson III
-distribution, using the \texttt{wichita} data from the package
-\texttt{SPEI}.}
-
-\end{figure}
-
-\newpage
+\includegraphics[width=10in,height=\textheight]{figures/toy-example-spei.png}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{../figures/pipeline-steps.png}
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-steps.png}
 
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
 
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-normalising-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-normalising-1.pdf}
 
@@
 \section{Examples}\label{sec-examples}}
-
-\newpage
 
@@
 \begin{verbatim}
+[1] "Checking for missing values (`NA`): all the data must be complete. Input type is vector. Assuming the data are monthly time series starting in January, all regular (non-leap) years."
+\end{verbatim}
+
+\begin{verbatim}
 # A tibble: 128,586 x 19
@@
    <chr>   <dbl> <chr>        <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <chr> <dbl>  <dbl>
- 1 spei        6 ASN00029~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5 KOWA~  67.8  102. 
- 2 spei        6 ASN00029~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5 KOWA~  67.8  102. 
- 3 spei        6 ASN00029~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7 MOUN~  30.8  -30.8
- 4 spei        6 ASN00029~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7 MOUN~  30.8  -30.8
- 5 spei        6 ASN00031~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9 CAIR~  66.3  728. 
- 6 spei        6 ASN00031~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9 CAIR~  66.3  728. 
- 7 spei        6 ASN00031~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1 WALK~  48.4  456. 
- 8 spei        6 ASN00031~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1 WALK~  48.4  456. 
- 9 spei        6 ASN00032~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6 SOUT~  54.5 1915. 
-10 spei        6 ASN00032~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6 SOUT~  54.5 1915. 
-# ... with 128,576 more rows, and 6 more variables: .scale <dbl>, .agg <dbl>,
-#   .method <chr>, .fitted <dbl>, .dist <chr>, .index <dbl>
+ 1 spei        6 ASN00029~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5 KOWA~  85.9   84.1
+ 2 spei        6 ASN00029~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5 KOWA~  85.9   84.1
+ 3 spei        6 ASN00029~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7 MOUN~  47.6  -47.6
+ 4 spei        6 ASN00029~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7 MOUN~  47.6  -47.6
+ 5 spei        6 ASN00031~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9 CAIR~  80.3  714. 
+ 6 spei        6 ASN00031~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9 CAIR~  80.3  714. 
+ 7 spei        6 ASN00031~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1 WALK~  48.0  456. 
+ 8 spei        6 ASN00031~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1 WALK~  48.0  456. 
+ 9 spei        6 ASN00032~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6 SOUT~  70.2 1900. 
+10 spei        6 ASN00032~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6 SOUT~  70.2 1900. 
+# i 128,576 more rows
+# i 6 more variables: .scale <dbl>, .agg <dbl>, .method <chr>, .fitted <dbl>,
+#   .dist <chr>, .index <dbl>
 \end{verbatim}
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-compute-spatial-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-compute-spatial-1.pdf}
 
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-compute-temporal-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-compute-temporal-1.pdf}
 
@@
 # A tibble: 191 x 13
-      id country   life_~1 exp_sch avg_sch gni_pc   sch index0 index1 index2 index3 index4
+      id country  life_exp exp_sch avg_sch gni_pc   sch index0 index1 index2 index3 index4
    <dbl> <chr>       <dbl>   <dbl>   <dbl>  <dbl> <dbl>  <dbl>  <dbl>  <dbl>  <dbl>  <dbl>
- 1     1 Switzerl~   0.984   0.917   0.924  0.983 0.920  0.962  0.963   1.67  0.978  0.933
+ 1     1 Switzer~    0.984   0.917   0.924  0.983 0.920  0.962  0.963   1.67  0.978  0.933
  2     2 Norway      0.973   1       0.867  0.978 0.933  0.961  0.961   1.66  0.969  0.942
  3     3 Iceland     0.964   1       0.918  0.955 0.959  0.959  0.959   1.66  0.963  0.959
- 4     4 Hong Kon~   1       0.960   0.815  0.973 0.887  0.952  0.953   1.65  0.986  0.907
- 5     5 Australia   0.993   1       0.848  0.936 0.924  0.951  0.951   1.65  0.980  0.932
+ 4     4 Hong Ko~    1       0.960   0.815  0.973 0.887  0.952  0.953   1.65  0.986  0.907
+ 5     5 Austral~    0.993   1       0.848  0.936 0.924  0.951  0.951   1.65  0.980  0.932
  6     6 Denmark     0.944   1       0.864  0.967 0.932  0.948  0.948   1.64  0.945  0.937
@@
  9     9 Germany     0.933   0.945   0.939  0.952 0.942  0.942  0.942   1.63  0.936  0.942
-10    10 Netherla~   0.949   1       0.839  0.956 0.919  0.941  0.941   1.63  0.947  0.926
-# ... with 181 more rows, 1 more variable: index5 <dbl>, and abbreviated variable name
-#   1: life_exp
+10    10 Netherl~    0.949   1       0.839  0.956 0.919  0.941  0.941   1.63  0.947  0.926
+# i 181 more rows
+# i 1 more variable: index5 <dbl>
 \end{verbatim}
@@
 
-{\centering \includegraphics{paper_files/figure-pdf/fig-hdi-expr-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-hdi-expr-1.pdf}
 

### 2023-07-04 tidyindex.qmd
outline the challenges with evaluating indexes and our proposal

@@
 knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
-options(width = 90)
+options(width = 70)
 library(tidyverse)
@@
 
-<!-- FYI:  -->
-
-<!--   - A measure is a variable, but is not an index.  -->
-<!--   - Not all multivariate quantities are indexes (i.e. those remote sensor measures). It is possible that an index is constructed from a single variable (i.e. SPI). -->
-<!--   - indexes is not a single/ set of measures -->
-
 **What is the challenges with current index construction**
 
-*see if there is any paper describing this type of pains*
-
-*useful to reference tidy data and tidy model that makes the workflow on modelling tidy somewhere in introduction*
-
-Currently, index construction lacks a standardised workflow. 
-It is often up to researchers or research institutions to decide whether to provide open source code on the new indexes, what would be the best user interface for other researchers to use the new indexes, and how easily the new indexes can be compared with other existing indexes.
-This makes the computation lack transparency and indexes cumbersome to experiment with:
-
-  - Researchers who wish to validate the indexes calculated from large institutes need to reinvent the wheels themselves since the source code used for computing is often not available for public consumption;
-  - Open-source code provided by research groups has a narrow margin for exploring other options outside the provided;
-  - Similar steps used by different indexes are difficult to spot since the design of the user interface for indexes often includes all the steps under a single function call; and
-  - It is generally hard to inspect intermediate results during the index construction if users wish to check the output of a certain step.
-
-<!-- Institutes like National Oceanic and Atmospheric Administration (NOAA) and United States Drought Monitor (USDM) moderate indexes in their domains. -->
-<!-- They provide public information to describe the methodology used in the calculation, publish the index values for public consumption while the source code used for computing is often not available for public consumption.
--->
-
-<!-- For example, many indexes involve a step to fit variables into a particular distribution. -->
-<!-- Index researchers may provide options for different fits, but it is generally hard to experiment with options outside the provided distributions. -->
-
-<!-- For example, many indexes needs standardisation in its workflow, it can be used to rescale the input variables or the index series. -->
-<!-- But the current construction does not make it clear through its user interface but sweep it, along with all the other steps in the index construction, under a single function call.  -->
+  - Difficult to reach consensus on composite/multivariate indexes. 
+  - Given the complexity of multivariate indexes as compared to univariate indexes, it raises more doubt to persuade people to use it. 
+  - Typical ways to evaluate new indexes is to compare it with the standard indexes (i.e. SPI in drought index) or a response variable (i.e. crop yield for monitoring agricultural drought). 
+  - However, more doubt can still raises: Whether the variable used to construct the index is a good set of variables. 
+  - Does the index series capture the variables measures. Whether other ways of combining the variables results in difference in the index. 
+  - Being able to answer these questions help to prove the robustness, extendibility and dimensionality of the index, as proposed by xxx as criteria for (selecting?) indexes.
 
@@
 
-This paper proposes a data pipeline for index construction. 
-By recognising the common steps shared by many indexes, we develop a pipeline that breaks down index construction into multiple modules and allow operations in various modules to be combined like building blocks to construct indexes.
-The pipeline approach is general while adaptable to most index construction. 
-It allows indexes to be created, studied, and compared in a structured tidy form and enables statistical analysis of indexes to be performed easily: 
-More specifically, it enables researchers to 1) validate the indexes calculated from external organisations, 2) unify various indexes under the same framework for computing, 3) swap or adjust individual steps in the index construction to study their contribution, 4) calculate uncertainty on indexes through bootstrap or others, 5) enhance existing indexes through comparing and studying their statistical properties, and finally, 6) propose new indexes from combining different steps in existing indexes.
+  - In order to evaluate indexes, indexes need to be broken down into pieces that it constitutes and then we can analyse how each component will affect the index. 
+  - The OECD handbook [@oecd_handbook_2008] provides technical guidelines to construct composite indexes for comparing country performance for policy-making. 
+  This paper identifies a more general pipeline framework where statistical indexes (of all kinds) can be built from. 
+  Based on the pipeline, an R package is built to validate the indexes, [more details on the functionality here] from understanding multivariate relationships to uncertainty analysis.
+
+<!-- The pipeline approach is general while adaptable to most index construction.  -->
+<!-- It allows indexes to be created, studied, and compared in a structured tidy form and enables statistical analysis of indexes to be performed easily:  -->
 
@@
 
-This work is of interest to researchers actively developing new indexes since it encourages new indexes to be delivered in an easy-to-reproduce design. 
-It would also provide analysts who wish to compute a range of indexes in their analysis a uniform interface to build relevant indexes from raw data. 
-For statisticians and software developing engineers, this work frames the process of index construction in a more user-oriented workflow and could motivate similar research for other process in scientific computing.
+  - This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes. 
+  - Helping index analysts to identify weakness in existing indexes for methodology improvement.
 
-The rest of the paper is structured as follows: @sec-data-pipeline-in-r reviews the concept of data pipeline in R. 
-The pipeline framework for index construction is presented in @sec-a-pipeline-for-building-statistical-indexes.
-@sec-incorporating-new-buliding-blocks-into-the-pipeline explains how to include a new building block in each pipeline module.
+The rest of the paper is structured as follows: @sec-pipeline reviews the concept of data pipeline in R.
+The pipeline framework for index construction is presented in @sec-index-pipeline.
+@sec-dev explains how to include a new building block in each pipeline module.
 Examples are given in @sec-examples to demonstrate the index construction with the pipeline built.
 
-# Data pipeline {#sec-data-pipeline-in-r}
+# Data pipeline {#sec-pipeline}
 
@@
 
-# A pipeline for building statistical indexes {#sec-a-pipeline-for-building-statistical-indexes}
+# A pipeline for building statistical indexes {#sec-index-pipeline}
 
@@
 
-```{r}
+```{r width = "100%"}
 knitr::include_graphics(here::here("figures/toy-example-spei.png"))
@@
 
-# Incorporating alternative methods into the pipeline components {#sec-incorporating-new-buliding-blocks-into-the-pipeline}
+# Incorporating alternative methods into the pipeline components {#sec-dev}
 

### 2023-07-04 tidyindex.tex
outline the challenges with evaluating indexes and our proposal

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-\emph{see if there is any paper describing this type of pains}
-
-\emph{useful to reference tidy data and tidy model that makes the
-workflow on modelling tidy somewhere in introduction}
-
-Currently, index construction lacks a standardised workflow. It is often
-up to researchers or research institutions to decide whether to provide
-open source code on the new indexes, what would be the best user
-interface for other researchers to use the new indexes, and how easily
-the new indexes can be compared with other existing indexes. This makes
-the computation lack transparency and indexes cumbersome to experiment
-with:
-
 \begin{itemize}
@@
 \item
-  Researchers who wish to validate the indexes calculated from large
-  institutes need to reinvent the wheels themselves since the source
-  code used for computing is often not available for public consumption;
-\item
-  Open-source code provided by research groups has a narrow margin for
-  exploring other options outside the provided;
-\item
-  Similar steps used by different indexes are difficult to spot since
-  the design of the user interface for indexes often includes all the
-  steps under a single function call; and
-\item
-  It is generally hard to inspect intermediate results during the index
-  construction if users wish to check the output of a certain step.
+  Difficult to reach consensus on composite/multivariate indexes.
+\item
+  Given the complexity of multivariate indexes as compared to univariate
+  indexes, it raises more doubt to persuade people to use it.
+\item
+  Typical ways to evaluate new indexes is to compare it with the
+  standard indexes (i.e.~SPI in drought index) or a response variable
+  (i.e.~crop yield for monitoring agricultural drought).
+\item
+  However, more doubt can still raises: Whether the variable used to
+  construct the index is a good set of variables.
+\item
+  Does the index series capture the variables measures. Whether other
+  ways of combining the variables results in difference in the index.
+\item
+  Being able to answer these questions help to prove the robustness,
+  extendibility and dimensionality of the index, as proposed by xxx as
+  criteria for (selecting?) indexes.
 \end{itemize}
@@
 
-This paper proposes a data pipeline for index construction. By
-recognising the common steps shared by many indexes, we develop a
-pipeline that breaks down index construction into multiple modules and
-allow operations in various modules to be combined like building blocks
-to construct indexes. The pipeline approach is general while adaptable
-to most index construction. It allows indexes to be created, studied,
-and compared in a structured tidy form and enables statistical analysis
-of indexes to be performed easily: More specifically, it enables
-researchers to 1) validate the indexes calculated from external
-organisations, 2) unify various indexes under the same framework for
-computing, 3) swap or adjust individual steps in the index construction
-to study their contribution, 4) calculate uncertainty on indexes through
-bootstrap or others, 5) enhance existing indexes through comparing and
-studying their statistical properties, and finally, 6) propose new
-indexes from combining different steps in existing indexes.
+\begin{itemize}
+\tightlist
+\item
+  In order to evaluate indexes, indexes need to be broken down into
+  pieces that it constitutes and then we can analyse how each component
+  will affect the index.
+\item
+  The OECD handbook (OECD, European Union, and Joint Research Centre -
+  European Commission 2008) provides technical guidelines to construct
+  composite indexes for comparing country performance for policy-making.
+  This paper identifies a more general pipeline framework where
+  statistical indexes (of all kinds) can be built from. Based on the
+  pipeline, an R package is built to validate the indexes, {[}more
+  details on the functionality here{]} from understanding multivariate
+  relationships to uncertainty analysis.
+\end{itemize}
 
@@
 
-This work is of interest to researchers actively developing new indexes
-since it encourages new indexes to be delivered in an easy-to-reproduce
-design. It would also provide analysts who wish to compute a range of
-indexes in their analysis a uniform interface to build relevant indexes
-from raw data. For statisticians and software developing engineers, this
-work frames the process of index construction in a more user-oriented
-workflow and could motivate similar research for other process in
-scientific computing.
+\begin{itemize}
+\tightlist
+\item
+  This work provides researchers actively developing new indexes with
+  tools to evaluate the proposed indexes.
+\item
+  Helping index analysts to identify weakness in existing indexes for
+  methodology improvement.
+\end{itemize}
 
 The rest of the paper is structured as follows:
-Section~\ref{sec-data-pipeline-in-r} reviews the concept of data
-pipeline in R. The pipeline framework for index construction is
-presented in
-Section~\ref{sec-a-pipeline-for-building-statistical-indexes}.
-Section~\ref{sec-incorporating-new-buliding-blocks-into-the-pipeline}
-explains how to include a new building block in each pipeline module.
-Examples are given in Section~\ref{sec-examples} to demonstrate the
-index construction with the pipeline built.
-
-\hypertarget{sec-data-pipeline-in-r}{%
-\section{Data pipeline}\label{sec-data-pipeline-in-r}}
+Section~\ref{sec-pipeline} reviews the concept of data pipeline in R.
+The pipeline framework for index construction is presented in
+Section~\ref{sec-index-pipeline}. Section~\ref{sec-dev} explains how to
+include a new building block in each pipeline module. Examples are given
+in Section~\ref{sec-examples} to demonstrate the index construction with
+the pipeline built.
+
+\hypertarget{sec-pipeline}{%
+\section{Data pipeline}\label{sec-pipeline}}
 
@@
 
-\hypertarget{sec-a-pipeline-for-building-statistical-indexes}{%
+\hypertarget{sec-index-pipeline}{%
 \section{A pipeline for building statistical
-indexes}\label{sec-a-pipeline-for-building-statistical-indexes}}
+indexes}\label{sec-index-pipeline}}
 
@@
 
-\hypertarget{sec-incorporating-new-buliding-blocks-into-the-pipeline}{%
+\hypertarget{sec-dev}{%
 \section{Incorporating alternative methods into the pipeline
-components}\label{sec-incorporating-new-buliding-blocks-into-the-pipeline}}
+components}\label{sec-dev}}
 
@@
 \begin{verbatim}
-var_trans: `rescale_minmax(var = life_exp, min = 20, max = 85)` -> life_exp
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = exp_sch, min = 0, max = 18)` -> exp_sch
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = avg_sch, min = 0, max = 15)` -> avg_sch
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = gni_pc, min = 2, max = 4.875061)` -> gni_pc
+var_trans: `rescale_minmax(var = life_exp, min = 20, max = 85)` ->
+life_exp
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = exp_sch, min = 0, max = 18)` ->
+exp_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = avg_sch, min = 0, max = 15)` ->
+avg_sch
+\end{verbatim}
+
+\begin{verbatim}
+var_trans: `rescale_minmax(var = gni_pc, min = 2, max = 4.875061)` ->
+gni_pc
 \end{verbatim}
@@
 # A tibble: 191 x 13
-      id country  life_exp exp_sch avg_sch gni_pc   sch index0 index1 index2 index3 index4
-   <dbl> <chr>       <dbl>   <dbl>   <dbl>  <dbl> <dbl>  <dbl>  <dbl>  <dbl>  <dbl>  <dbl>
- 1     1 Switzer~    0.984   0.917   0.924  0.983 0.920  0.962  0.963   1.67  0.978  0.933
- 2     2 Norway      0.973   1       0.867  0.978 0.933  0.961  0.961   1.66  0.969  0.942
- 3     3 Iceland     0.964   1       0.918  0.955 0.959  0.959  0.959   1.66  0.963  0.959
- 4     4 Hong Ko~    1       0.960   0.815  0.973 0.887  0.952  0.953   1.65  0.986  0.907
- 5     5 Austral~    0.993   1       0.848  0.936 0.924  0.951  0.951   1.65  0.980  0.932
- 6     6 Denmark     0.944   1       0.864  0.967 0.932  0.948  0.948   1.64  0.945  0.937
- 7     7 Sweden      0.969   1       0.841  0.952 0.920  0.947  0.947   1.64  0.962  0.928
- 8     8 Ireland     0.954   1       0.772  1     0.886  0.945  0.947   1.64  0.952  0.904
- 9     9 Germany     0.933   0.945   0.939  0.952 0.942  0.942  0.942   1.63  0.936  0.942
-10    10 Netherl~    0.949   1       0.839  0.956 0.919  0.941  0.941   1.63  0.947  0.926
+      id country   life_exp exp_sch avg_sch gni_pc   sch index0 index1
+   <dbl> <chr>        <dbl>   <dbl>   <dbl>  <dbl> <dbl>  <dbl>  <dbl>
+ 1     1 Switzerl~    0.984   0.917   0.924  0.983 0.920  0.962  0.963
+ 2     2 Norway       0.973   1       0.867  0.978 0.933  0.961  0.961
+ 3     3 Iceland      0.964   1       0.918  0.955 0.959  0.959  0.959
+ 4     4 Hong Kon~    1       0.960   0.815  0.973 0.887  0.952  0.953
+ 5     5 Australia    0.993   1       0.848  0.936 0.924  0.951  0.951
+ 6     6 Denmark      0.944   1       0.864  0.967 0.932  0.948  0.948
+ 7     7 Sweden       0.969   1       0.841  0.952 0.920  0.947  0.947
+ 8     8 Ireland      0.954   1       0.772  1     0.886  0.945  0.947
+ 9     9 Germany      0.933   0.945   0.939  0.952 0.942  0.942  0.942
+10    10 Netherla~    0.949   1       0.839  0.956 0.919  0.941  0.941
 # i 181 more rows
-# i 1 more variable: index5 <dbl>
+# i 4 more variables: index2 <dbl>, index3 <dbl>, index4 <dbl>,
+#   index5 <dbl>
 \end{verbatim}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-oecd_handbook_2008}{}}%
+OECD, European Union, and Joint Research Centre - European Commission.
+2008. \emph{Handbook on {Constructing} {Composite} {Indicators}:
+{Methodology} and {User} {Guide}}. OECD.
+\url{https://doi.org/10.1787/9789264043466-en}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-sutherland_orca_2000}{}}%

### 2023-07-04 tidyindex.qmd
outline for introducing done

@@
 
-<!-- Each individual index follows its own data pipeline and it can be difficult to evaluate how an index can be affected by tweaking parameters in a certain step, rearranging the order of steps, using to different method.
--->
-
-<!-- However, even if most of the indexes have available existing implementation available to users, there would be hundreds of them and it will be burdensome to remember, classify, and compare between.
--->
-
 **Why index is useful, why people care about indexes**
@@
 Many concepts relevant to decision making cannot be directly measured, however, they are crucial for resource allocation, early prevention, and other operational purpose.
-For example, fire authorities would be interested to quantify fire risk since bushfires can have a huge impact on monetary loss, health, and the local ecosystem.
-Climatologists would be interested in monitoring the change in global climate since variability in atmospheric and oceanic conditions has a direct impact on global weather and climate.
-Usually this concept of interest is associated with more than one variables and these variables need to be integrated to make decisions on the subject matter.
-<!-- From a modelling perspective, these unobserved quantities are called latent variables and existing methods can incorporate these latent variables in the model structure.  -->
-A common approach to quantify concepts like these is to construct an index using these relevant variables. 
-This allows researchers to compare the quantity of interest across entities (i.e. countries, regions) and also cross time.
+
+  - USDM and European Drought Observatory publishes drought indicator to [...], 
+  - NOAA and BOM producing forecast on El Nino alert system to [...]
+  - OECD and UN producing indexes measuring socio-economic aspects of country for policy making.
+
+There decision making often needs to consider multiple aspects and indexes is a statistical tool to pack these information into a single number, making it easier for policy makers and public communication.
 
@@
 
-In this article, an index is defined as a tool to quantify a concept of interest that does not have a direct measure.
-The concept of interest doesn't have a direct measure can because it is impractical to measure at the population level. 
-For example, it would be nearly impossible to include all the available stocks in the market to characterise stock market behavior, so indexes like Dow Jones Industrial Average, S&P 500, and Nasdaq Composite select a representative set of stocks to measure the overall market behavior.
-Also belonging to this category are the economic indexes like the Consumer Price Index, where price changes of a basket of items are weighted to measure inflation. 
-The lack of direct measure could also because the concept itself is an unobservable human construction, rather than a physical quantity that can be measured.
-Many natural hazard and social concepts falls into this category. 
-This includes drought indexes constructed from meteorological, agricultural, hydrological, and social-economic variables, e.g.
-Standardised Precipitation Index (SPI) [@mckee1993relationship] and Aggregated Drought Index (ADI) [@keyantash_aggregate_2004] among others.
-Social development indexes like Human Development Index [@hdi] and Global Liveability Index [@gli] measure various aspects of the quality of human capital and urban life.
-
-*still need to tweak the tone a bit: "they are called index, they are not the index we will talk about"*
-
-Despite many quantity having the term *index* in their name, they cannot be technically classified as indexes according to the definition given above. 
-The reason for these quantities to lose their index memberships is that they are variables can be accurately measured given the instrument precision. 
-This includes quantities like precipitation of the driest month or percentage of days when maximum temperature is below 10th percentile. 
-They are measures of precipitation and percentage of days under specific conditions (dries month, maximum temperature below 10th percentile).
-They are variables, or indicators, that can be used to construct indexes but are not indexes themselves. 
-Similarly, a set of remote sensing indexes are not indexes, since they are measures of electromagnetic wave reflectance.
-This includes Normalized Difference Vegetation Index (NDVI) [@tucker_red_1979], derived from the ratio of difference over sum on two segments in the spectrum, also called band: near-infrared (NIR) and red.
-So are the "indexes" derived from NDVI, e.g. Vegetation Condition Index [@kogan_application_1995]. 
-Notice that this does not exclude all the construction derived from remotes sensor variables to be valid indexes.
-For example, Vegetation Drought Response Index [@brown_vegetation_2008] is a valid index since it integrates climate, satellite, and biophysical variables to quantify vegetation stress.
+  - Indexes can be composed from single variables or multiple variables (composite index)
+  - to make the definition even more complex is that indexes themselves are used with other variables to compose indexes (many drought indexes using an index called SPI). 
+  - Some indexes may be more simple than you think:  
+    - a single transformation (NDVI [@tucker_red_1979] takes two spectrum: red and near-infrared and calculates as $(\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})$)
+    - indexes sometimes are similar to features extracted from time series data: ETCCDI climate change indices [@climate-indexes]: SU - number of summer days: annual count of days when TX > 25 degree
+  - The index pipeline and software implementation we propose are applicable to all indexes mentioned above, while the main focus of this paper is on indexes that combine multiple variables/indexes since more skills are required to diagnose those indexes to understand how they behave.
 

### 2023-07-04 tidyindex.tex
outline for introducing done

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 however, they are crucial for resource allocation, early prevention, and
-other operational purpose. For example, fire authorities would be
-interested to quantify fire risk since bushfires can have a huge impact
-on monetary loss, health, and the local ecosystem. Climatologists would
-be interested in monitoring the change in global climate since
-variability in atmospheric and oceanic conditions has a direct impact on
-global weather and climate. Usually this concept of interest is
-associated with more than one variables and these variables need to be
-integrated to make decisions on the subject matter. A common approach to
-quantify concepts like these is to construct an index using these
-relevant variables. This allows researchers to compare the quantity of
-interest across entities (i.e.~countries, regions) and also cross time.
+other operational purpose.
+
+\begin{itemize}
+\tightlist
+\item
+  USDM and European Drought Observatory publishes drought indicator to
+  {[}\ldots{]},
+\item
+  NOAA and BOM producing forecast on El Nino alert system to
+  {[}\ldots{]}
+\item
+  OECD and UN producing indexes measuring socio-economic aspects of
+  country for policy making.
+\end{itemize}
+
+There decision making often needs to consider multiple aspects and
+indexes is a statistical tool to pack these information into a single
+number, making it easier for policy makers and public communication.
 
@@
 
-In this article, an index is defined as a tool to quantify a concept of
-interest that does not have a direct measure. The concept of interest
-doesn't have a direct measure can because it is impractical to measure
-at the population level. For example, it would be nearly impossible to
-include all the available stocks in the market to characterise stock
-market behavior, so indexes like Dow Jones Industrial Average, S\&P 500,
-and Nasdaq Composite select a representative set of stocks to measure
-the overall market behavior. Also belonging to this category are the
-economic indexes like the Consumer Price Index, where price changes of a
-basket of items are weighted to measure inflation. The lack of direct
-measure could also because the concept itself is an unobservable human
-construction, rather than a physical quantity that can be measured. Many
-natural hazard and social concepts falls into this category. This
-includes drought indexes constructed from meteorological, agricultural,
-hydrological, and social-economic variables, e.g.~Standardised
-Precipitation Index (SPI) (McKee et al. 1993) and Aggregated Drought
-Index (ADI) (Keyantash and Dracup 2004) among others. Social development
-indexes like Human Development Index (United Nations Development
-Programme 2022) and Global Liveability Index (Economist Intelligence
-Unit 2019) measure various aspects of the quality of human capital and
-urban life.
-
-\emph{still need to tweak the tone a bit: ``they are called index, they
-are not the index we will talk about''}
-
-Despite many quantity having the term \emph{index} in their name, they
-cannot be technically classified as indexes according to the definition
-given above. The reason for these quantities to lose their index
-memberships is that they are variables can be accurately measured given
-the instrument precision. This includes quantities like precipitation of
-the driest month or percentage of days when maximum temperature is below
-10th percentile. They are measures of precipitation and percentage of
-days under specific conditions (dries month, maximum temperature below
-10th percentile). They are variables, or indicators, that can be used to
-construct indexes but are not indexes themselves. Similarly, a set of
-remote sensing indexes are not indexes, since they are measures of
-electromagnetic wave reflectance. This includes Normalized Difference
-Vegetation Index (NDVI) (Tucker 1979), derived from the ratio of
-difference over sum on two segments in the spectrum, also called band:
-near-infrared (NIR) and red. So are the ``indexes'' derived from NDVI,
-e.g.~Vegetation Condition Index (Kogan 1995). Notice that this does not
-exclude all the construction derived from remotes sensor variables to be
-valid indexes. For example, Vegetation Drought Response Index (Brown et
-al. 2008) is a valid index since it integrates climate, satellite, and
-biophysical variables to quantify vegetation stress.
+\begin{itemize}
+\tightlist
+\item
+  Indexes can be composed from single variables or multiple variables
+  (composite index)
+\item
+  to make the definition even more complex is that indexes themselves
+  are used with other variables to compose indexes (many drought indexes
+  using an index called SPI).
+\item
+  Some indexes may be more simple than you think:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    a single transformation (NDVI (Tucker 1979) takes two spectrum: red
+    and near-infrared and calculates as
+    \((\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})\))
+  \item
+    indexes sometimes are similar to features extracted from time series
+    data: ETCCDI climate change indices (Zhang 2020): SU - number of
+    summer days: annual count of days when TX \textgreater{} 25 degree
+  \end{itemize}
+\item
+  The index pipeline and software implementation we propose are
+  applicable to all indexes mentioned above, while the main focus of
+  this paper is on indexes that combine multiple variables/indexes since
+  more skills are required to diagnose those indexes to understand how
+  they behave.
+\end{itemize}
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-brown_vegetation_2008}{}}%
-Brown, Jesslyn F., Brian D. Wardlow, Tsegaye Tadesse, Michael J. Hayes,
-and Bradley C. Reed. 2008. {``The {Vegetation} {Drought} {Response}
-{Index} ({VegDRI}): {A} {New} {Integrated} {Approach} for {Monitoring}
-{Drought} {Stress} in {Vegetation}.''} \emph{GIScience \& Remote
-Sensing} 45 (1): 16--46.
-\url{https://doi.org/10.2747/1548-1603.45.1.16}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-buja_elements_1988}{}}%
@@
 Statistics}, 277--308. Wadsworth, Belmont.
-
-\leavevmode\vadjust pre{\hypertarget{ref-gli}{}}%
-Economist Intelligence Unit. 2019. {``The Global Liveability Index
-2019.''} The Economist.
-\url{https://www.cbeinternational.ca/pdf/Liveability-Free-report-2019.pdf}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-keyantash_aggregate_2004}{}}%
-Keyantash, John A., and John A. Dracup. 2004. {``An Aggregate Drought
-Index: {Assessing} Drought Severity Based on Fluctuations in the
-Hydrologic Cycle and Surface Water Storage.''} \emph{Water Resources
-Research} 40 (9). \url{https://doi.org/10.1029/2003WR002610}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kogan_application_1995}{}}%
-Kogan, F. N. 1995. {``Application of Vegetation Index and Brightness
-Temperature for Drought Detection.''} \emph{Advances in Space Research},
-Natural {Hazards}: {Monitoring} and {Assessment} {Using} {Remote}
-{Sensing} {Technique}, 15 (11): 91--100.
-\url{https://doi.org/10.1016/0273-1177(95)00079-T}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mckee1993relationship}{}}%
-McKee, Thomas B, Nolan J Doesken, John Kleist, et al. 1993. {``The
-Relationship of Drought Frequency and Duration to Time Scales.''} In
-\emph{Proceedings of the 8th Conference on Applied Climatology},
-17:179--83. 22. Boston, MA, USA.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
----------. 2022. {``Human Development Report 2021-22.''} New York.
-\url{http://report.hdr.undp.org}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-climate-indexes}{}}%
+Zhang, Xuebin. 2020. {``ETCCDI Climate Change Indices.''}
+\url{http://etccdi.pacificclimate.org/index.shtml}.
+
 \end{CSLReferences}

### 2023-07-09 tidyindex.qmd
re-write the index definition section and shift of research direction in intro

@@
 ---
-title: "Demo arXiv template"
+title: "A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data"
 format:
@@
         address: 29 Ancora Imparo Way, Clayton
-        city: Melbourne, VIC
+        city: Melbourne, Victoria
         country: Australia
@@
     email: huize.zhang@monash.edu
-    url: https://huizezhangsh.netlify.app/
-  - name: Collaborators
-    affiliations:
+  - name: Dianne Cook
+    affiliation: 
       - name: Monash University
@@
         address: 29 Ancora Imparo Way, Clayton
-        city: Melbourne, VIC
+        city: Melbourne, Victoria
         country: Australia
+    orcid: 0000-0002-3813-7155
+  - name: Ursula Laa
+    affiliation: 
+      - name: University of Natural Resources and Life Sciences
+        department: Institute of Statistics
+        address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
+        city: Vienna
+        country: Austria
+    orcid: 0000-0002-0249-6439
+  - name: Nicolas Langrené
+    affiliation: 
+      - name: BNU-HKBU United International College
+        department: Department of Mathematical Sciences
+        address: 2000 Jintong Road, Tangjiawan, Zhuhai
+        city: Zhuhai, Guangdong
+        country: China
+    orcid: 0000-0001-7601-4618
+  - name: Patricia Menéndez
+    affiliation:
+      - name: Monash University
+        department: Department of Econometrics and Business Statistics
+        address: 29 Ancora Imparo Way, Clayton
+        city: Melbourne, Victoria
+        country: Australia
+    orcid: 0000-0003-0701-6315
 abstract: |
@@
 
+While the [decision to make] is often multi-faceted, simple indexes tend to be favoured when it comes to communicating to stakeholders and public.
+
 **Define what is an index, what is not**
 
-  - Indexes can be composed from single variables or multiple variables (composite index)
-  - to make the definition even more complex is that indexes themselves are used with other variables to compose indexes (many drought indexes using an index called SPI). 
-  - Some indexes may be more simple than you think:  
-    - a single transformation (NDVI [@tucker_red_1979] takes two spectrum: red and near-infrared and calculates as $(\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})$)
-    - indexes sometimes are similar to features extracted from time series data: ETCCDI climate change indices [@climate-indexes]: SU - number of summer days: annual count of days when TX > 25 degree
-  - The index pipeline and software implementation we propose are applicable to all indexes mentioned above, while the main focus of this paper is on indexes that combine multiple variables/indexes since more skills are required to diagnose those indexes to understand how they behave.
+While indexes are commonly used in operation and research, their compositions and structures can vary substantially.
+An index can be composed from a singe variable, such as the Standardized Precipitation Index (SPI) [@spi], or multiple variables, also known as composite indexes or composite indicators, like the Human Development Index [@hdi].
+Some indexes can be expressed as a single equation, like Normalized Difference Vegetation Index (NDVI) [@ndvi], calculated as $(\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})$.
+In other cases, indexes themselves can be used as variables to construct indexes.
+For example, SPI is used to in many complex indexes, such as the U.S.
+Drought monitor and European Combined Drought Index (CDI) [@cdi].
+While the index framework proposed in this paper applies to all the aforementioned indexes, our focus will be on studying complex indexes that combine multiple variables using statistical methods.
+
+The initial plan aims to investigate how variable combinations can affect the indexes and whether we can use statistical methods to recommend better combination for decision making.
+However, when analysing specific index examples, we find the composition of indexes is much more complicated than originally expected.
+In addition to combining multiple variables, indexes can have a spatial and/or temporal component to capture the neighbouring effect in space and the prolonged effect in time.
+[Constructing indexes from combining variables with spatial and temporal component is further complicated by different treatments to process data from different sources such as satellite imageries, in-situ stations, and large physical models.]
 
@@
 
-  - Difficult to reach consensus on composite/multivariate indexes. 
-  - Given the complexity of multivariate indexes as compared to univariate indexes, it raises more doubt to persuade people to use it. 
-  - Typical ways to evaluate new indexes is to compare it with the standard indexes (i.e. SPI in drought index) or a response variable (i.e. crop yield for monitoring agricultural drought). 
+While numerous indexes are proposed by researchers, most well-known indexes only use one or two variables.
+More complex indexes tend to have difficulties to reach consensus given the uncertainty of their behaviors in different scenarios.
+Typical ways to evaluate new indexes is to compare it with the standard/ well-known indexes, such as SPI or SPEI for drought indexes, or a response variable relevant for the monitoring purpose, i.e.
+crop yield for monitoring agricultural drought.
+
   - However, more doubt can still raises: Whether the variable used to construct the index is a good set of variables. 
@@
 
-```{r width = "100%"}
+```{r out.width = "100%"}
 knitr::include_graphics(here::here("figures/toy-example-spei.png"))
@@
 
-## Pipeline steps for constructing indies
+## Pipeline steps for constructing indexes
 
@@
 For example, the defined index needs to be useful for measuring the concept of interest and variables need to be collected from reliable sources with proper quality control.
+
+  * align the spatial resolution and temporal frequency of data collected from different satellite products
+  * obtain and clean variables from countries, potentially aggregate from regional data into country level
+  * merge ground-based and satellite-based data
 

### 2023-07-09 tidyindex.tex
re-write the index definition section and shift of research direction in intro

@@
 \hypersetup{
-  pdftitle={Demo arXiv template},
-  pdfauthor={H. Sherry Zhang; Collaborators},
+  pdftitle={A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data},
+  pdfauthor={H. Sherry Zhang; Dianne Cook; Ursula Laa; Nicolas Langrené; Patricia Menéndez},
   pdfkeywords={indexes, data pipeline, software design},
@@
 
-\title{Demo arXiv template}
+\title{A Tidy Framework and Infrastructure to Systematically Assemble
+Spatio-temporal Indexes from Multivariate Data}
 \author{H. Sherry
-Zhang$\textsuperscript{1}$~\orcidlink{0000-0002-7122-1463}, Collaborators$\textsuperscript{1}$}
+Zhang$\textsuperscript{1}$~\orcidlink{0000-0002-7122-1463}, Dianne
+Cook$\textsuperscript{1}$~\orcidlink{0000-0002-3813-7155}, Ursula
+Laa$\textsuperscript{2}$~\orcidlink{0000-0002-0249-6439}, Nicolas
+Langrené$\textsuperscript{3}$~\orcidlink{0000-0001-7601-4618}, Patricia
+Menéndez$\textsuperscript{1}$~\orcidlink{0000-0003-0701-6315}}
 
@@
 \textsuperscript{1} Department of Econometrics and Business
-Statistics, Monash University, Melbourne, VIC, Australia
+Statistics, Monash University, Melbourne,
+Victoria, Australia\\ \textsuperscript{2} Institute of
+Statistics, University of Natural Resources and Life
+Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
+Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
+Guangdong, China
 \begin{abstract}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, boxrule=0pt, enhanced, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
+While the {[}decision to make{]} is often multi-faceted, simple indexes
+tend to be favoured when it comes to communicating to stakeholders and
+public.
+
 \textbf{Define what is an index, what is not}
+
+While indexes are commonly used in operation and research, their
+compositions and structures can vary substantially. An index can be
+composed from a singe variable, such as the Standardized Precipitation
+Index (SPI) (McKee, Doesken, and Kleist 1993), or multiple variables,
+also known as composite indexes or composite indicators, like the Human
+Development Index (United Nations Development Programme 2022). Some
+indexes can be expressed as a single equation, like Normalized
+Difference Vegetation Index (NDVI) (Tucker 1979), calculated as
+\((\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})\). In other cases,
+indexes themselves can be used as variables to construct indexes. For
+example, SPI is used to in many complex indexes, such as the U.S.
+Drought monitor and European Combined Drought Index (CDI)
+(Sepulcre-Canto et al. 2012). While the index framework proposed in this
+paper applies to all the aforementioned indexes, our focus will be on
+studying complex indexes that combine multiple variables using
+statistical methods.
+
+The initial plan aims to investigate how variable combinations can
+affect the indexes and whether we can use statistical methods to
+recommend better combination for decision making. However, when
+analysing specific index examples, we find the composition of indexes is
+much more complicated than originally expected. In addition to combining
+multiple variables, indexes can have a spatial and/or temporal component
+to capture the neighbouring effect in space and the prolonged effect in
+time. {[}Constructing indexes from combining variables with spatial and
+temporal component is further complicated by different treatments to
+process data from different sources such as satellite imageries, in-situ
+stations, and large physical models.{]}
+
+\textbf{What is the challenges with current index construction}
+
+While numerous indexes are proposed by researchers, most well-known
+indexes only use one or two variables. More complex indexes tend to have
+difficulties to reach consensus given the uncertainty of their behaviors
+in different scenarios. Typical ways to evaluate new indexes is to
+compare it with the standard/ well-known indexes, such as SPI or SPEI
+for drought indexes, or a response variable relevant for the monitoring
+purpose, i.e.~crop yield for monitoring agricultural drought.
 
@@
 \tightlist
-\item
-  Indexes can be composed from single variables or multiple variables
-  (composite index)
-\item
-  to make the definition even more complex is that indexes themselves
-  are used with other variables to compose indexes (many drought indexes
-  using an index called SPI).
-\item
-  Some indexes may be more simple than you think:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    a single transformation (NDVI (Tucker 1979) takes two spectrum: red
-    and near-infrared and calculates as
-    \((\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})\))
-  \item
-    indexes sometimes are similar to features extracted from time series
-    data: ETCCDI climate change indices (Zhang 2020): SU - number of
-    summer days: annual count of days when TX \textgreater{} 25 degree
-  \end{itemize}
-\item
-  The index pipeline and software implementation we propose are
-  applicable to all indexes mentioned above, while the main focus of
-  this paper is on indexes that combine multiple variables/indexes since
-  more skills are required to diagnose those indexes to understand how
-  they behave.
-\end{itemize}
-
-\textbf{What is the challenges with current index construction}
-
-\begin{itemize}
-\tightlist
-\item
-  Difficult to reach consensus on composite/multivariate indexes.
-\item
-  Given the complexity of multivariate indexes as compared to univariate
-  indexes, it raises more doubt to persuade people to use it.
-\item
-  Typical ways to evaluate new indexes is to compare it with the
-  standard indexes (i.e.~SPI in drought index) or a response variable
-  (i.e.~crop yield for monitoring agricultural drought).
 \item
@@
 
-\includegraphics[width=10in,height=\textheight]{figures/toy-example-spei.png}
-
-\hypertarget{pipeline-steps-for-constructing-indies}{%
+\includegraphics[width=1\textwidth,height=\textheight]{figures/toy-example-spei.png}
+
+\hypertarget{pipeline-steps-for-constructing-indexes}{%
 \subsection{Pipeline steps for constructing
-indies}\label{pipeline-steps-for-constructing-indies}}
+indexes}\label{pipeline-steps-for-constructing-indexes}}
 
@@
 proper quality control.
+
+\begin{itemize}
+\tightlist
+\item
+  align the spatial resolution and temporal frequency of data collected
+  from different satellite products
+\item
+  obtain and clean variables from countries, potentially aggregate from
+  regional data into country level
+\item
+  merge ground-based and satellite-based data
+\end{itemize}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-spi}{}}%
+McKee, Thomas B, Nolan J Doesken, and John Kleist. 1993. {``The
+Relationship of Drought Frequency and Duration to Time Scales.''} In
+\emph{Proceedings of the 8th Conference of Applied Climatology},
+179--84. Anaheim, CA: American Meteorological Society.
+
 \leavevmode\vadjust pre{\hypertarget{ref-oecd_handbook_2008}{}}%
@@
 \url{https://doi.org/10.1787/9789264043466-en}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-cdi}{}}%
+Sepulcre-Canto, G., S. Horion, A. Singleton, H. Carrao, and J. Vogt.
+2012. {``Development of a {Combined} {Drought} {Indicator} to Detect
+Agricultural Drought in {Europe}.''} \emph{Natural Hazards and Earth
+System Sciences} 12 (11): 3519--31.
+\url{https://doi.org/10.5194/nhess-12-3519-2012}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tucker_red_1979}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-ndvi}{}}%
 Tucker, Compton J. 1979. {``Red and Photographic Infrared Linear
@@
 \url{https://hdr.undp.org/sites/default/files/2021-22_HDR/hdr2021-22_technical_notes.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
+---------. 2022. {``Human Development Report 2021-22.''} New York.
+\url{http://report.hdr.undp.org}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-climate-indexes}{}}%
-Zhang, Xuebin. 2020. {``ETCCDI Climate Change Indices.''}
-\url{http://etccdi.pacificclimate.org/index.shtml}.
-
 \end{CSLReferences}

### 2023-07-10 tidyindex.qmd
more writing in the intro

@@
 The initial plan aims to investigate how variable combinations can affect the indexes and whether we can use statistical methods to recommend better combination for decision making.
-However, when analysing specific index examples, we find the composition of indexes is much more complicated than originally expected.
-In addition to combining multiple variables, indexes can have a spatial and/or temporal component to capture the neighbouring effect in space and the prolonged effect in time.
-[Constructing indexes from combining variables with spatial and temporal component is further complicated by different treatments to process data from different sources such as satellite imageries, in-situ stations, and large physical models.]
+However, when analysing specific index examples, we find indexes are often not just a simple weighted linear aggregation of variables.
+When indexes have a spatial and/or temporal component, different choices on aggregating space and time will introduce uncertainty.
+[There is always new variables found to be related to make the decision and should be introduced into the index.
+The decision made in pre-processing data coming from different sources such as satellite imageries, in-situ stations, and large physical models.] While numerous indexes are proposed in the literature, the well-known ones tend to prefer a simple procedure: they may only consider one or two variables, i.e.
+SPI or use a decision-tree based process to incorporate multiple variables, i.e.
+(How El Nino is monitored by major climate institutes).
+More complex indexes often have difficulties to reach consensus, due to their uncertainty when applied in different scenarios.
 
-**What is the challenges with current index construction**
+**What is the challenges with current index construction**/ **what can be done if people adopt this pipeline/ why it is beneficial?**/**who would benefit from this paper**
 
-While numerous indexes are proposed by researchers, most well-known indexes only use one or two variables.
-More complex indexes tend to have difficulties to reach consensus given the uncertainty of their behaviors in different scenarios.
-Typical ways to evaluate new indexes is to compare it with the standard/ well-known indexes, such as SPI or SPEI for drought indexes, or a response variable relevant for the monitoring purpose, i.e.
+Typical ways to evaluate indexes is to compare it with the standard/ well-known indexes, such as SPI or SPEI for drought indexes, or a response variable relevant for the monitoring purpose, i.e.
 crop yield for monitoring agricultural drought.
-
-  - However, more doubt can still raises: Whether the variable used to construct the index is a good set of variables. 
-  - Does the index series capture the variables measures. Whether other ways of combining the variables results in difference in the index. 
-  - Being able to answer these questions help to prove the robustness, extendibility and dimensionality of the index, as proposed by xxx as criteria for (selecting?) indexes.
-
-**what can be done if people adopt this pipeline/ why it is beneficial?**
-
-  - In order to evaluate indexes, indexes need to be broken down into pieces that it constitutes and then we can analyse how each component will affect the index. 
-  - The OECD handbook [@oecd_handbook_2008] provides technical guidelines to construct composite indexes for comparing country performance for policy-making. 
-  This paper identifies a more general pipeline framework where statistical indexes (of all kinds) can be built from. 
-  Based on the pipeline, an R package is built to validate the indexes, [more details on the functionality here] from understanding multivariate relationships to uncertainty analysis.
-
-<!-- The pipeline approach is general while adaptable to most index construction.  -->
-<!-- It allows indexes to be created, studied, and compared in a structured tidy form and enables statistical analysis of indexes to be performed easily:  -->
-
-**who would benefit from this paper**
-
-  - This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes. 
-  - Helping index analysts to identify weakness in existing indexes for methodology improvement.
+While these evaluations show the accuracy of the index to capture, they doesn't address the robustness of the index,[ given variation of the input parameters, or a change of different study areas or across time.] To evaluate the robustness of indexes, indexes need to be broken down into their fundamental building blocks, and then analyse the effect of each component.
+The OECD handbook [@oecd_handbook_2008] has recommended a set of guidelines for construct composite indexes for measuring country's socio-economic performance.
+This paper identifies a more general pipeline framework from which indexes can be built from.
+Based on the pipeline, an R package, tidyindex, is developed to [validate the indexes].
+This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes.
+It also helps index analysts to identify weakness in existing indexes for methodology improvement.
 

### 2023-07-10 tidyindex.tex
more writing in the intro

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, boxrule=0pt, enhanced, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 recommend better combination for decision making. However, when
-analysing specific index examples, we find the composition of indexes is
-much more complicated than originally expected. In addition to combining
-multiple variables, indexes can have a spatial and/or temporal component
-to capture the neighbouring effect in space and the prolonged effect in
-time. {[}Constructing indexes from combining variables with spatial and
-temporal component is further complicated by different treatments to
-process data from different sources such as satellite imageries, in-situ
-stations, and large physical models.{]}
-
-\textbf{What is the challenges with current index construction}
-
-While numerous indexes are proposed by researchers, most well-known
-indexes only use one or two variables. More complex indexes tend to have
-difficulties to reach consensus given the uncertainty of their behaviors
-in different scenarios. Typical ways to evaluate new indexes is to
-compare it with the standard/ well-known indexes, such as SPI or SPEI
-for drought indexes, or a response variable relevant for the monitoring
-purpose, i.e.~crop yield for monitoring agricultural drought.
-
-\begin{itemize}
-\tightlist
-\item
-  However, more doubt can still raises: Whether the variable used to
-  construct the index is a good set of variables.
-\item
-  Does the index series capture the variables measures. Whether other
-  ways of combining the variables results in difference in the index.
-\item
-  Being able to answer these questions help to prove the robustness,
-  extendibility and dimensionality of the index, as proposed by xxx as
-  criteria for (selecting?) indexes.
-\end{itemize}
-
+analysing specific index examples, we find indexes are often not just a
+simple weighted linear aggregation of variables. When indexes have a
+spatial and/or temporal component, different choices on aggregating
+space and time will introduce uncertainty. {[}There is always new
+variables found to be related to make the decision and should be
+introduced into the index. The decision made in pre-processing data
+coming from different sources such as satellite imageries, in-situ
+stations, and large physical models.{]} While numerous indexes are
+proposed in the literature, the well-known ones tend to prefer a simple
+procedure: they may only consider one or two variables, i.e.~SPI or use
+a decision-tree based process to incorporate multiple variables,
+i.e.~(How El Nino is monitored by major climate institutes). More
+complex indexes often have difficulties to reach consensus, due to their
+uncertainty when applied in different scenarios.
+
+\textbf{What is the challenges with current index construction}/
 \textbf{what can be done if people adopt this pipeline/ why it is
-beneficial?}
-
-\begin{itemize}
-\tightlist
-\item
-  In order to evaluate indexes, indexes need to be broken down into
-  pieces that it constitutes and then we can analyse how each component
-  will affect the index.
-\item
-  The OECD handbook (OECD, European Union, and Joint Research Centre -
-  European Commission 2008) provides technical guidelines to construct
-  composite indexes for comparing country performance for policy-making.
-  This paper identifies a more general pipeline framework where
-  statistical indexes (of all kinds) can be built from. Based on the
-  pipeline, an R package is built to validate the indexes, {[}more
-  details on the functionality here{]} from understanding multivariate
-  relationships to uncertainty analysis.
-\end{itemize}
-
-\textbf{who would benefit from this paper}
-
-\begin{itemize}
-\tightlist
-\item
-  This work provides researchers actively developing new indexes with
-  tools to evaluate the proposed indexes.
-\item
-  Helping index analysts to identify weakness in existing indexes for
+beneficial?}/\textbf{who would benefit from this paper}
+
+Typical ways to evaluate indexes is to compare it with the standard/
+well-known indexes, such as SPI or SPEI for drought indexes, or a
+response variable relevant for the monitoring purpose, i.e.~crop yield
+for monitoring agricultural drought. While these evaluations show the
+accuracy of the index to capture, they doesn't address the robustness of
+the index,{[} given variation of the input parameters, or a change of
+different study areas or across time.{]} To evaluate the robustness of
+indexes, indexes need to be broken down into their fundamental building
+blocks, and then analyse the effect of each component. The OECD handbook
+(OECD, European Union, and Joint Research Centre - European Commission
+2008) has recommended a set of guidelines for construct composite
+indexes for measuring country's socio-economic performance. This paper
+identifies a more general pipeline framework from which indexes can be
+built from. Based on the pipeline, an R package, tidyindex, is developed
+to {[}validate the indexes{]}. This work provides researchers actively
+developing new indexes with tools to evaluate the proposed indexes. It
+also helps index analysts to identify weakness in existing indexes for
 methodology improvement.
-\end{itemize}
 

### 2023-07-11 tidyindex.qmd
a new draft version of the introduction

@@
 
-**Why index is useful, why people care about indexes**
+  <!-- - Why index is useful, why people care about indexes -->
+  <!-- - Define what is an index, what is not -->
+  <!-- - What is the challenges with current index construction -->
+  <!-- - what can be done if people adopt this pipeline/ why it is beneficial? -->
+  <!-- - who would benefit from this paper -->
 
-*incorporate the following in why using index: multiple pieces of information (variables) that need to be taken into account*
+Indexes are commonly-used tools to monitor, communicate and make decisions across various domains in society.
+Their applications can be found in the environment (e.g.
+Air Quality Index, El Niño-Southern Oscillation Index), in the economy (Consumer Price Index, stock market indexes), and in the social sciences (e.g.
+QS University Rankings, Human Development Index).
 
-Many concepts relevant to decision making cannot be directly measured, however, they are crucial for resource allocation, early prevention, and other operational purpose.
+To construct an index, experts typically first define an underlying concept of interest to measure, then collect relevant variables for measuring the concept, and lastly summarise these variables into a single numerical or categorical value.
+An index can be as simple as counting the number of days when the maximum temperature is greater than 25 degree.
+This is "the number of summer days" (SU), a climate index listed in ETCCDI Climate Change Indices [@climate-indexes].
+In some cases, indexes are weighted linear combinations of variables.
+For example, Human Development Index (HDI) [@hdi] combines the three aspects (life expectancy, years of schooling, and GNI per capita) to measure country's development.
+In more complex scenarios, indexes are constructed by combining multiple variables, often accompanied by spatial and temporal considerations arising from diverse data formats, including in-situ stations, satellite imageries, and model outputs.
+An example of this is the Combined Drought Index (CDI) [@cdi], curated by the European Drought Observatory, which uses a decision tree to combine Standardized Precipitation Index (SPI) [@spi] with the satellite measure, Fraction of Absorbed Photosynthetically Active Radiation (FAPAR), and modelled soil moisture.
 
-  - USDM and European Drought Observatory publishes drought indicator to [...], 
-  - NOAA and BOM producing forecast on El Nino alert system to [...]
-  - OECD and UN producing indexes measuring socio-economic aspects of country for policy making.
+Our initial interest is to understand how different variable combinations contribute to the uncertainty of an index.
+However, as we start to understand how indexes are composed, we find many other steps in the construction also introduce uncertainty.
+In order to evaluate the uncertainty, an index needs to be broken down into its fundamental building blocks to analyse the effect of each component.
+This shifts the focus to investigate the decomposition of indexes into their constituent parts for analysing uncertainty.
+While the OECD handbook [@oecd_handbook_2008] has provided a set of steps for constructing composite socio-economic indexes, there is a need to extend these guidelines to accommodate more complex construction steps required for indexes in general.
 
-There decision making often needs to consider multiple aspects and indexes is a statistical tool to pack these information into a single number, making it easier for policy makers and public communication.
-
-While the [decision to make] is often multi-faceted, simple indexes tend to be favoured when it comes to communicating to stakeholders and public.
-
-**Define what is an index, what is not**
-
-While indexes are commonly used in operation and research, their compositions and structures can vary substantially.
-An index can be composed from a singe variable, such as the Standardized Precipitation Index (SPI) [@spi], or multiple variables, also known as composite indexes or composite indicators, like the Human Development Index [@hdi].
-Some indexes can be expressed as a single equation, like Normalized Difference Vegetation Index (NDVI) [@ndvi], calculated as $(\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})$.
-In other cases, indexes themselves can be used as variables to construct indexes.
-For example, SPI is used to in many complex indexes, such as the U.S.
-Drought monitor and European Combined Drought Index (CDI) [@cdi].
-While the index framework proposed in this paper applies to all the aforementioned indexes, our focus will be on studying complex indexes that combine multiple variables using statistical methods.
-
-The initial plan aims to investigate how variable combinations can affect the indexes and whether we can use statistical methods to recommend better combination for decision making.
-However, when analysing specific index examples, we find indexes are often not just a simple weighted linear aggregation of variables.
-When indexes have a spatial and/or temporal component, different choices on aggregating space and time will introduce uncertainty.
-[There is always new variables found to be related to make the decision and should be introduced into the index.
-The decision made in pre-processing data coming from different sources such as satellite imageries, in-situ stations, and large physical models.] While numerous indexes are proposed in the literature, the well-known ones tend to prefer a simple procedure: they may only consider one or two variables, i.e.
-SPI or use a decision-tree based process to incorporate multiple variables, i.e.
-(How El Nino is monitored by major climate institutes).
-More complex indexes often have difficulties to reach consensus, due to their uncertainty when applied in different scenarios.
-
-**What is the challenges with current index construction**/ **what can be done if people adopt this pipeline/ why it is beneficial?**/**who would benefit from this paper**
-
-Typical ways to evaluate indexes is to compare it with the standard/ well-known indexes, such as SPI or SPEI for drought indexes, or a response variable relevant for the monitoring purpose, i.e.
-crop yield for monitoring agricultural drought.
-While these evaluations show the accuracy of the index to capture, they doesn't address the robustness of the index,[ given variation of the input parameters, or a change of different study areas or across time.] To evaluate the robustness of indexes, indexes need to be broken down into their fundamental building blocks, and then analyse the effect of each component.
-The OECD handbook [@oecd_handbook_2008] has recommended a set of guidelines for construct composite indexes for measuring country's socio-economic performance.
-This paper identifies a more general pipeline framework from which indexes can be built from.
-Based on the pipeline, an R package, tidyindex, is developed to [validate the indexes].
-This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes.
-It also helps index analysts to identify weakness in existing indexes for methodology improvement.
+In this work, we develop a data pipeline framework where indexes from different domains can be built from.
+Based on the pipeline, an R package, tidyindex, is developed for practitioners to construct indexes and understand how an index responds to changes in different components in the pipeline.
+This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes on their robustness for wider adoption.
+It also helps index analysts to diagnose an index to identify its weakness for methodology improvement.
 
@@
 Examples are given in @sec-examples to demonstrate the index construction with the pipeline built.
+
+<!-- Evaluating indexes is challenging because the underlying concept an index measures is often a latent variable and lacks direct benchmarks for comparison.
+In literature, researchers typically demonstrate the accuracy of an index through comparing it to some well-known indexes, such as SPI for drought indexes, or to a relevant response variable, such as crop yield for monitoring agricultural drought.
+For social indexes, @uncertainty proposes a Monte Carlo algorithm to compute index uncertainty through sampling the linear weights.
+-->
+
+<!-- These sometimes make indexes difficult to reach consensus for wider use and large institutes often prefer simple indexes for more stable performance and easier interpretation. -->
+
+<!-- When indexes have a spatial and/or temporal component, different choices on how space and time is aggregated will introduce uncertainty.
+A slight change in the variable value can sometimes introduce a big difference for the index.
+-->
 
@@
 
+Future work:
+  - integrate more complex dimension reduction methods to calculate weights 
+  - strengthen the spatial processing module
+
 # Reference

### 2023-07-11 tidyindex.tex
a new draft version of the introduction

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, interior hidden, boxrule=0pt, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\textbf{Why index is useful, why people care about indexes}
-
-\emph{incorporate the following in why using index: multiple pieces of
-information (variables) that need to be taken into account}
-
-Many concepts relevant to decision making cannot be directly measured,
-however, they are crucial for resource allocation, early prevention, and
-other operational purpose.
-
-\begin{itemize}
-\tightlist
-\item
-  USDM and European Drought Observatory publishes drought indicator to
-  {[}\ldots{]},
-\item
-  NOAA and BOM producing forecast on El Nino alert system to
-  {[}\ldots{]}
-\item
-  OECD and UN producing indexes measuring socio-economic aspects of
-  country for policy making.
-\end{itemize}
-
-There decision making often needs to consider multiple aspects and
-indexes is a statistical tool to pack these information into a single
-number, making it easier for policy makers and public communication.
-
-While the {[}decision to make{]} is often multi-faceted, simple indexes
-tend to be favoured when it comes to communicating to stakeholders and
-public.
-
-\textbf{Define what is an index, what is not}
-
-While indexes are commonly used in operation and research, their
-compositions and structures can vary substantially. An index can be
-composed from a singe variable, such as the Standardized Precipitation
-Index (SPI) (McKee, Doesken, and Kleist 1993), or multiple variables,
-also known as composite indexes or composite indicators, like the Human
-Development Index (United Nations Development Programme 2022). Some
-indexes can be expressed as a single equation, like Normalized
-Difference Vegetation Index (NDVI) (Tucker 1979), calculated as
-\((\text{NIR} + \text{Red})/(\text{NIR} + \text{Red})\). In other cases,
-indexes themselves can be used as variables to construct indexes. For
-example, SPI is used to in many complex indexes, such as the U.S.
-Drought monitor and European Combined Drought Index (CDI)
-(Sepulcre-Canto et al. 2012). While the index framework proposed in this
-paper applies to all the aforementioned indexes, our focus will be on
-studying complex indexes that combine multiple variables using
-statistical methods.
-
-The initial plan aims to investigate how variable combinations can
-affect the indexes and whether we can use statistical methods to
-recommend better combination for decision making. However, when
-analysing specific index examples, we find indexes are often not just a
-simple weighted linear aggregation of variables. When indexes have a
-spatial and/or temporal component, different choices on aggregating
-space and time will introduce uncertainty. {[}There is always new
-variables found to be related to make the decision and should be
-introduced into the index. The decision made in pre-processing data
-coming from different sources such as satellite imageries, in-situ
-stations, and large physical models.{]} While numerous indexes are
-proposed in the literature, the well-known ones tend to prefer a simple
-procedure: they may only consider one or two variables, i.e.~SPI or use
-a decision-tree based process to incorporate multiple variables,
-i.e.~(How El Nino is monitored by major climate institutes). More
-complex indexes often have difficulties to reach consensus, due to their
-uncertainty when applied in different scenarios.
-
-\textbf{What is the challenges with current index construction}/
-\textbf{what can be done if people adopt this pipeline/ why it is
-beneficial?}/\textbf{who would benefit from this paper}
-
-Typical ways to evaluate indexes is to compare it with the standard/
-well-known indexes, such as SPI or SPEI for drought indexes, or a
-response variable relevant for the monitoring purpose, i.e.~crop yield
-for monitoring agricultural drought. While these evaluations show the
-accuracy of the index to capture, they doesn't address the robustness of
-the index,{[} given variation of the input parameters, or a change of
-different study areas or across time.{]} To evaluate the robustness of
-indexes, indexes need to be broken down into their fundamental building
-blocks, and then analyse the effect of each component. The OECD handbook
-(OECD, European Union, and Joint Research Centre - European Commission
-2008) has recommended a set of guidelines for construct composite
-indexes for measuring country's socio-economic performance. This paper
-identifies a more general pipeline framework from which indexes can be
-built from. Based on the pipeline, an R package, tidyindex, is developed
-to {[}validate the indexes{]}. This work provides researchers actively
-developing new indexes with tools to evaluate the proposed indexes. It
-also helps index analysts to identify weakness in existing indexes for
-methodology improvement.
+Indexes are commonly-used tools to monitor, communicate and make
+decisions across various domains in society. Their applications can be
+found in the environment (e.g.~Air Quality Index, El Niño-Southern
+Oscillation Index), in the economy (Consumer Price Index, stock market
+indexes), and in the social sciences (e.g.~QS University Rankings, Human
+Development Index).
+
+To construct an index, experts typically first define an underlying
+concept of interest to measure, then collect relevant variables for
+measuring the concept, and lastly summarise these variables into a
+single numerical or categorical value. An index can be as simple as
+counting the number of days when the maximum temperature is greater than
+25 degree. This is ``the number of summer days'' (SU), a climate index
+listed in ETCCDI Climate Change Indices (Zhang 2020). In some cases,
+indexes are weighted linear combinations of variables. For example,
+Human Development Index (HDI) (United Nations Development Programme
+2022) combines the three aspects (life expectancy, years of schooling,
+and GNI per capita) to measure country's development. In more complex
+scenarios, indexes are constructed by combining multiple variables,
+often accompanied by spatial and temporal considerations arising from
+diverse data formats, including in-situ stations, satellite imageries,
+and model outputs. An example of this is the Combined Drought Index
+(CDI) (Sepulcre-Canto et al. 2012), curated by the European Drought
+Observatory, which uses a decision tree to combine Standardized
+Precipitation Index (SPI) (McKee, Doesken, and Kleist 1993) with the
+satellite measure, Fraction of Absorbed Photosynthetically Active
+Radiation (FAPAR), and modelled soil moisture.
+
+Our initial interest is to understand how different variable
+combinations contribute to the uncertainty of an index. However, as we
+start to understand how indexes are composed, we find many other steps
+in the construction also introduce uncertainty. In order to evaluate the
+uncertainty, an index needs to be broken down into its fundamental
+building blocks to analyse the effect of each component. This shifts the
+focus to investigate the decomposition of indexes into their constituent
+parts for analysing uncertainty. While the OECD handbook (OECD, European
+Union, and Joint Research Centre - European Commission 2008) has
+provided a set of steps for constructing composite socio-economic
+indexes, there is a need to extend these guidelines to accommodate more
+complex construction steps required for indexes in general.
+
+In this work, we develop a data pipeline framework where indexes from
+different domains can be built from. Based on the pipeline, an R
+package, tidyindex, is developed for practitioners to construct indexes
+and understand how an index responds to changes in different components
+in the pipeline. This work provides researchers actively developing new
+indexes with tools to evaluate the proposed indexes on their robustness
+for wider adoption. It also helps index analysts to diagnose an index to
+identify its weakness for methodology improvement.
 
@@
 
+Future work: - integrate more complex dimension reduction methods to
+calculate weights - strengthen the spatial processing module
+
 \hypertarget{reference}{%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ndvi}{}}%
-Tucker, Compton J. 1979. {``Red and Photographic Infrared Linear
-Combinations for Monitoring Vegetation.''} \emph{Remote Sensing of
-Environment} 8 (2): 127--50.
-\url{https://doi.org/10.1016/0034-4257(79)90013-0}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021_tech_notes}{}}%

### 2023-07-19 tidyindex.qmd
plots for GGGI

@@
 knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
-options(width = 70)
+#options(width = 70)
 library(tidyverse)
@@
 library(patchwork)
+library(tourr)
 ```
@@
 
-## An example on sustainable development indicators
+## Global Gender Gap Index
 
-<!-- Indexes are commonly used in social science is to rank entities (i.e.
-countries) on a topic through a range of indicators.
-This includes university ranking from QS, Times Higher Education; the Sustainable Development Goals (SDG) Index [@SDSN_Bertelsmann_2021] by the Sustainable Development Solutions Network (SDSN), the Human Development Index (HDI) [@UNDP_2021] by United Nations Development Programme (UNDP), etc.
-These indexes, or rankings, typically have a set of predefined dimensions, or pillars, and indicators are first reduced to these pillars and then reduced to an index.
--->
-
-In the following example, the Human Development Index (HDI) will first be constructed using the pipeline syntax.
-The section will then demonstrate how expressions or parameters in the pipeline can be swapped to alternatives to study the index property.
-
-Human Development Index (HDI) measures the development of countries through more than just economic growth, but also people's life expectancy and opportunity to receive education.
-These three identified dimensions are measured using four indicators: life expectancy at birth (health), expected years of schooling (education), mean years of schooling (education), and Gross National Income (GNI) *per capita* (standard of living).
-The technical notes [@UNDP_2021_tech_notes] have documented the procedures to calculate HDI and they are summarised below:
-
-  1. take log on GNI *per capita*,
-  2. rescale the four indicators into [0, 1] using mini-max,
-  3. aggregate the two education indicators using arithmetic mean, and
-  4. aggregate the three dimensions into the index using geometric mean.
-
-The values used in mini-max rescaling are summarised in @tbl-rescale-params and the justification of these numbers can be found in the technical notes mentioned above.
-
-```{r}
-#| tbl-cap-location: bottom
-#| tbl-cap: Maximum and minimum values used to rescale the four HDI indicators into [0, 1] range. The maximum of GNI per capita is taken as the common log of 75,000, approximatly 4.875.
-#| label: tbl-rescale-params
-scaling_params <- tibble::tribble(
-   ~Dimension, ~Indicator, ~Variable,  ~Minimum, ~Maximum,
-  "Health",              "Life expectancy at birth (years)",   "life_exp",    "20",          "85",
-  "Education",           "Expected years of schooling (years)",  "exp_sch",    "0",          "18",
-  "Education",           "Mean years of schooling (years)",   "avg_sch",      "0",          "15",
-  "Standard of living",  "GNI per capita (2017 PPP$)",     "gni_pc",       "100",      "75000"
-  ) %>%
-  mutate(across(contains("mum"), as.numeric),
-         across(contains("mum"), ~ifelse(Variable == "gni_pc", log10(.x), .x)))
-knitr::kable(scaling_params, digits = 3)
+```{r eval = FALSE}
+a <- dim_weights %>% select(-variable) %>% t() %>% as_tibble(rownames = "id")
+colnames(a) <- c("id", dim_weights$variable)
+a %>% 
+  pivot_longer(-id, names_to = "vars", values_to = "value") %>% 
+  ggplot(aes(x = value, group = vars)) + 
+  geom_histogram() + 
+  facet_wrap(vars(vars))
 ```
 
-Among the four steps listed in the calculation, the first two are variable transformation.
-The next two can be considered as the dimension reduction step in the data pipeline given its intention to combine multivariate data into univariate.
-With the index pipeline proposed in the paper, HDI can be calculated as:
+```{r gggi-verify}
+set.seed(123)
+n <- 1000
+w_idx <- map(1:n, function(x){
+  w_idx <- stats::runif(4, min = 0.1, max = 1)
+  w_idx <- w_idx/sum(w_idx)
+  w_idx %>% as.matrix(nrow = 4, ncol = 1)
+  })
+rand_w <- reduce(w_idx, cbind) %>% as_tibble()
+colnames(rand_w) <- paste0("V", 1:n)
 
-```{r}
-dt <- readxl::read_xlsx(path = here::here("data/hdi.xlsx") , skip = 4)
-raw <- dt %>%
-  janitor::clean_names() %>%
-  dplyr::select(-paste0("x", seq(4, 14, 2))) %>%
-  dplyr::rename(id = x1, countries = x2) %>%
-  dplyr::mutate(across(c(id, human_development_index_hdi:hdi_rank), as.numeric)) %>%
-  dplyr::filter(!is.na(id)) %>%
-  dplyr::mutate(across(4:7, ~round(.x, digits = 3))) %>% 
-  dplyr::select(-c(human_development_index_hdi,
-                   gni_per_capita_rank_minus_hdi_rank, 
-                   hdi_rank)) 
-colnames(raw) <- c("id", "country", "life_exp", "exp_sch", "avg_sch", "gni_pc")
+dim_weights <- gggi_weights %>%
+  filter(!is.na(pillar_weight)) %>%
+  select(variable, pillar_weight) %>% 
+  bind_cols(rand_w)
+
+dt <- gggi %>% 
+  select(country:political_empowerment) %>% 
+  init(id = country) %>%
+  add_meta(dim_weights, var_col = variable)
+
+dt2 <- dt %>%
+  dimension_reduction(
+    index_new = aggregate_linear(
+      ~economic_participation_and_opportunity:political_empowerment,
+      weight = pillar_weight))
+
+dt2$data %>%
+  ggplot(aes(x = index, y = index_new)) +
+  geom_abline(slope = 1, intercept = 0, color = "grey") + 
+  geom_point() +
+  theme_bw() + 
+  theme(aspect.ratio = 1, panel.grid.minor = element_blank()) 
 ```
 
-```{r echo = TRUE, message=TRUE}
-dt <- raw %>% init(id = country, indicators = life_exp:gni_pc)
-res <- dt %>%
-  var_trans(gni_pc = log(gni_pc, base = 10)) %>% 
-  var_trans(.method = rescale_minmax, .vars = life_exp:gni_pc,
-            min = c(20, 0, 0, 2), max = c(85, 18, 15, log10(75000))) %>% 
-  dim_red(sch = (exp_sch + avg_sch) / 2) %>%
-  dim_red(index = (life_exp * sch * gni_pc)^(1/3))
+```{r gggi-weights, fig.height=20, fig.align='center', fig.width=10}
+dt3 <- dt2 %>%
+  swap_values(.id = 1, .param = weight,
+              .value = list(pillar_weight, paste0("V", 2:100)),
+              .raw_data = dt)
+
+idx_wide <- dt3$data %>% select(country, dplyr::contains("index_new"))
+idx_wide2 <- idx_wide %>% mutate(index_avg = rowMeans(idx_wide[,2:(ncol(idx_wide))]))
+idx_long2 <- idx_wide2 %>% pivot_longer(-country, names_to = "index", values_to = "value") %>% 
+  left_join(gggi %>% select(country, rank), by = "country")
+
+idx_long2 %>%
+  ggplot(aes(x = value, y = fct_reorder(country, -rank), group = country)) +
+  geom_boxplot() + 
+  #geom_point(color = "grey80", size = 0.5) +
+  geom_point(data = idx_long2 %>% filter(index == "index_new1"), color = "red") + 
+  theme_bw() + 
+  ylab("Country") + 
+  xlab("Index")
 ```
 
-<!-- The output is an `indri` object with a summary of operations followed by the data output.
-The data output contains the ID, country name, rescaled variables (`life_exp`, `exp_sch`, `avg_sch`, and `gni_pc`), the aggregated education dimension (`sch`), and the final index calculated (`index`).
--->
+```{r eval = FALSE}
+dt <- tidyindex::gggi %>% 
+  dplyr::select(country: political_empowerment) %>%
+  filter(region %in% c("South Asia", "Sub-Saharan Africa")) %>%
+  arrange(index) 
+colnames(dt) <- c(colnames(dt)[1:4], "economy", "education", "health", "politics")
 
-Index analysts can be interested in studying the property of the index and the countries through the "what-if" questions.
-In the context of HDI, this can be what if an alternative dimension reduction expression is used to reduce the three dimensions into the HDI?
-Apart from the geometrical mean, linear combination is a common method to reduce the data dimension.
-With different weighting schemes, the combination can be compared to study the variable importance of each dimension or to reveal the underlying structures of countries.
-The following code tests five expressions (arithmetic mean, principal component analysis weight, and three weighting schemes emphasizing life expectancy, education, and GNI per capita respectively) using the `swap_exprs()` function.
-The function first locates the operation to be tested (`.var = index`), then accepts a list of expressions for testing `.expr`, before requesting the raw data:
+w_proj2idx <- function(w){w/sum(w)}
+w_idx2proj <- function(w){w/(sqrt(sum(w^2))) %>% matrix(nrow = length(w))}
+find_next <- function(matrix){
+  start <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1)
+  tourr:::step_angle(tourr:::geodesic_info(start, matrix), 0.4)
+}
+b <- array(dim = c(4, 1, 9))
+b[,,c(1,3,5,7,9)] <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1) %>% w_idx2proj()
+b[,,2] <- matrix(c(0.5, 0.2, 0.2, 0.1), ncol = 1) %>% find_next()
+b[,,4] <- matrix(c(0.2, 0.5, 0.2, 0.1), ncol= 1) %>% find_next()
+b[,,6] <- matrix(c(0.1, 0.2, 0.5, 0.2), ncol= 1) %>% find_next()
+b[,,8] <- matrix(c(0.15, 0.2, 0.15, 0.5), ncol = 1) %>% find_next()
 
-```{r echo = TRUE, message=TRUE}
-(res2 <- res %>%
-  swap_exprs(
-    .var = index,
-    .expr = list(
-      index1 = (life_exp + sch + gni_pc)/3,
-      # PCA recommended weight
-      index2 = 0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc, 
-      index3 = 0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc,
-      index4 = 0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc,
-      index5 = 0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc),
-    .raw_data = dt))
+render(
+  dt[,5:8], planned2_tour(b), dev = "png",
+  display_idx(cex = 2, label_cex = 2,
+              label = dt$country, abb_vars = FALSE),
+  width = 800, height = 1300, apf = 1/20, frames = 100,
+  here::here("figures/idx-tour/idx-tour-%03d.png")
+)
+
+gifski::gifski(
+  list.files(here::here("figures/idx-tour"), full.names = TRUE),
+  gif_file = here::here("figures/idx-tour.gif"),
+  delay = 0.15, width = 800, height = 1300)
+
 ```
 
-The resulting data includes five more columns corresponding to the five alternative indexes.
-The country ranking can be further computed on the data and plotted in a scatterplot matrix as in @fig-hdi-expr.
-The plot shows almost similar ranks suggested by the indexes calculated using geometric mean (`rank0`), the arithmetic mean (`rank1`), and the PCA recommended weights (`rank2`).
-The agreement of the three methods supports the use of geometric mean as the dimension reduction method in the original index.
-The scatterplot matrix also reveals the difference in ranking when the three dimensions are given different weights in `rank3`, `rank4`, and `rank5`.
-The panel `rank4` vs `rank5` shows a cluster of highlighted points with relatively high ranks according to `rank5` but low in `rank4`.
-These countries are printed in @tbl-hdi-expr prints along with their three dimensions, calculated indexes, and rankings.
-While having a high life expectancy and GNI *per capita*, these countries (Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) score relatively low in education, which is weighted heavily in `index4` (`0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc`).
-
-```{r fig-hdi-expr}
-#| fig-height: 8
-#| fig-width: 12
-#| fig-cap: "Comparing country ranks from six different index constructions: geometric mean (`rank0`), arithmetic mean (`rank1`), principal component analysis recommended weights (`rank2`), and heavier weight on life expectancy, education, and GNI per capita (`rank3` - `rank5`).
-Geometric mean, arithmetic mean, and PCA produce similar rankings where countries are aligned in a diagnoal line, while country rankings varies when compare among `rank3` to `rank5`.
-Panel `rank5` vs.
-`rank4` highlights a cluster of countries that rank high by `rank5` but low by `rank4`.
-The plot demonstrates the necessity to test on various alternative expressions to verify the robustness of the method and to capture distinctive characteristics of different countries.
-"
-with_rank <- res2$data %>% 
-  mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>% 
-  rename_with(~gsub(".index", "", .x), contains("rank.index"))
-
-lower_cluster <- with_rank %>% 
-  filter(rank5 < 30, rank4 > 50) %>% 
-  select(country, life_exp, sch, gni_pc, index3, index4, rank3, rank4)
-
-with_highlight <- with_rank %>% 
-  mutate(highlight = ifelse(country %in% lower_cluster$country, TRUE, FALSE)) %>% 
-  rename(`geometric mean` = rank0, `arithmetic mean` = rank1,
-         `PCA weights` = rank2, `weight 0.8 on life expec.` = rank3,
-         `weight 0.8 on education`  = rank4, `weight 0.8 on GNI per capita` = rank5)
-
-p1 <- ggpairs(
-  with_highlight, columns = 14:19,
-  lower = list(continuous = "points", combo = "facethist", 
-               discrete = "facetbar", na = "na"),
-  upper = NULL, diag = NULL,  switch = "both", 
-  labeller = label_wrap_gen(width = 15, multi_line = TRUE) ) +
-  theme(aspect.ratio = 1)
-
-# https://stackoverflow.com/questions/65091593/how-to-mark-certain-point-in-ggpairs
-p1$plots[[35]]$mapping <- 
-    `class<-`(c(p1$plots[[35]]$mapping, aes(color = highlight)), "uneval")
-p1 <- p1 + scale_color_manual(values = c("black", "red"))
-p1
+```{r idx-tour, fig.align='center', fig.height= 12, fig.width=12}
+#| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index.
+The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse.
+Full animation is available at [vimeo link]."
+frames <- c("001", "013", "037", "061", "085")
+ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
+rl <- lapply(ani, png::readPNG)
+gl <-  lapply(rl, grid::rasterGrob)
+wrap_plots(gl, ncol = 3) 
 ```
-
-```{r echo = FALSE}
-#| tbl-cap: A selected number of countries with low rankings when education is given a heavy weight (`index4`) while ranks high when a heavier weight is given on GNI per capital (`index5`).
-#| label: tbl-hdi-expr
-lower_cluster %>% 
-  rename( `life expectancy` = life_exp, education = sch, `GNI per capita` = gni_pc) %>% 
-  knitr::kable(digits = 3)
-```
-
-<!-- \newpage -->
-
-<!-- *what if the minimum value for GNI per capita is changed?* -->
-
-<!-- The lower limit for GNI per capita is set to be the common log of $100, with justification provided in the technical notes as: -->
-
-<!--   > The low minimum value for gross national income (GNI) per capita, $100, is justified by the considerable amount of unmeasured subsistence and nonmarket production in economies close to the minimum, which is not captured in the official data -->
-
-<!-- The open source data provided by United Nations Development Programme suggest the minimum GNI *per capita* is `$732` and this opens the questions on how the index/ranking will change if the lower minimum (`$100`) is raised or decreased.
-We can answer this question by testing on a set of different values using `.values` in the function `swap_values()`: -->
-
-<!-- ```{r echo = TRUE, message = TRUE} -->
-<!-- (res3 <- res %>% -->
-<!--   swap_values( -->
-<!--     .module = var_trans, .step = rescale_minmax, .res = gni_pc, -->
-<!--     .var = max, .values = log10(150000), -->
-<!--     .raw_data = dt) -->
-<!--    )  -->
-<!-- ``` -->
-
-<!-- \newpage -->
-
-<!-- The function computes the GNI *per capita* in both the original value (`gni_pc0`) and the new value provided (`gni_pc1`).
-All the subsequent operations are also computed for both cases and result in `index0` and `index1`.
--->
-
-<!-- ```{r} -->
-<!-- #| fig-height: 6 -->
-<!-- #| fig-width: 6 -->
-<!-- #| fig-align: center -->
-<!-- new_rank <- res3$data %>% -->
-<!--   mutate(across(contains("index"), ~rank(-.x), .names = "rank.{.col}")) %>% -->
-<!--   rename_with(~gsub(".index", "", .x), contains("rank.index")) %>%  -->
-<!--   mutate(difference = rank1 - rank0) -->
-
-<!-- try <- new_rank %>% -->
-<!--   dplyr::select(id, country, contains("gni_pc"), contains("index"), contains("rank"), difference) %>%  -->
-<!--   arrange(-abs(difference)) -->
-
-<!-- new_rank %>%  -->
-<!--   ggplot(aes(x = rank0, y = rank1)) +  -->
-<!--   geom_point() +  -->
-<!--   geom_abline(slope = 1, intercept = 0, color = "blue") + -->
-<!--   # geom_text(data = try %>% filter(difference > 15), color = "red", aes(label = country)) + -->
-<!--   theme(aspect.ratio = 1) + -->
-<!--   xlab("log10(100) as the rescale minimum for GNI per capita") + -->
-<!--   ylab("log10(700) as the rescale minimum for GNI per capita")  -->
-<!-- ``` -->
-
-\newpage
 

### 2023-07-19 tidyindex.tex
plots for GGGI

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, sharp corners, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{an-example-on-sustainable-development-indicators}{%
-\subsection{An example on sustainable development
-indicators}\label{an-example-on-sustainable-development-indicators}}
-
-In the following example, the Human Development Index (HDI) will first
-be constructed using the pipeline syntax. The section will then
-demonstrate how expressions or parameters in the pipeline can be swapped
-to alternatives to study the index property.
-
-Human Development Index (HDI) measures the development of countries
-through more than just economic growth, but also people's life
-expectancy and opportunity to receive education. These three identified
-dimensions are measured using four indicators: life expectancy at birth
-(health), expected years of schooling (education), mean years of
-schooling (education), and Gross National Income (GNI) \emph{per capita}
-(standard of living). The technical notes (United Nations Development
-Programme 2021) have documented the procedures to calculate HDI and they
-are summarised below:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  take log on GNI \emph{per capita},
-\item
-  rescale the four indicators into {[}0, 1{]} using mini-max,
-\item
-  aggregate the two education indicators using arithmetic mean, and
-\item
-  aggregate the three dimensions into the index using geometric mean.
-\end{enumerate}
-
-The values used in mini-max rescaling are summarised in
-Table~\ref{tbl-rescale-params} and the justification of these numbers
-can be found in the technical notes mentioned above.
-
-\hypertarget{tbl-rescale-params}{}
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2375}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.4500}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1125}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1000}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1000}}@{}}
-\caption{\label{tbl-rescale-params}Maximum and minimum values used to
-rescale the four HDI indicators into {[}0, 1{]} range. The maximum of
-GNI per capita is taken as the common log of 75,000, approximatly
-4.875.}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Dimension
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Variable
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Minimum
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Maximum
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-Dimension
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Variable
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Minimum
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Maximum
-\end{minipage} \\
-\midrule()
-\endhead
-Health & Life expectancy at birth (years) & life\_exp & 20 & 85.000 \\
-Education & Expected years of schooling (years) & exp\_sch & 0 &
-18.000 \\
-Education & Mean years of schooling (years) & avg\_sch & 0 & 15.000 \\
-Standard of living & GNI per capita (2017 PPP\$) & gni\_pc & 2 &
-4.875 \\
-\bottomrule()
-\end{longtable}
-
-Among the four steps listed in the calculation, the first two are
-variable transformation. The next two can be considered as the dimension
-reduction step in the data pipeline given its intention to combine
-multivariate data into univariate. With the index pipeline proposed in
-the paper, HDI can be calculated as:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{dt }\OtherTok{\textless{}{-}}\NormalTok{ raw }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country, }\AttributeTok{indicators =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc)}
-\NormalTok{res }\OtherTok{\textless{}{-}}\NormalTok{ dt }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{gni\_pc =} \FunctionTok{log}\NormalTok{(gni\_pc, }\AttributeTok{base =} \DecValTok{10}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{var\_trans}\NormalTok{(}\AttributeTok{.method =}\NormalTok{ rescale\_minmax, }\AttributeTok{.vars =}\NormalTok{ life\_exp}\SpecialCharTok{:}\NormalTok{gni\_pc,}
-            \AttributeTok{min =} \FunctionTok{c}\NormalTok{(}\DecValTok{20}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{2}\NormalTok{), }\AttributeTok{max =} \FunctionTok{c}\NormalTok{(}\DecValTok{85}\NormalTok{, }\DecValTok{18}\NormalTok{, }\DecValTok{15}\NormalTok{, }\FunctionTok{log10}\NormalTok{(}\DecValTok{75000}\NormalTok{))) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{sch =}\NormalTok{ (exp\_sch }\SpecialCharTok{+}\NormalTok{ avg\_sch) }\SpecialCharTok{/} \DecValTok{2}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{dim\_red}\NormalTok{(}\AttributeTok{index =}\NormalTok{ (life\_exp }\SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{*}\NormalTok{ gni\_pc)}\SpecialCharTok{\^{}}\NormalTok{(}\DecValTok{1}\SpecialCharTok{/}\DecValTok{3}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-Index analysts can be interested in studying the property of the index
-and the countries through the ``what-if'' questions. In the context of
-HDI, this can be what if an alternative dimension reduction expression
-is used to reduce the three dimensions into the HDI? Apart from the
-geometrical mean, linear combination is a common method to reduce the
-data dimension. With different weighting schemes, the combination can be
-compared to study the variable importance of each dimension or to reveal
-the underlying structures of countries. The following code tests five
-expressions (arithmetic mean, principal component analysis weight, and
-three weighting schemes emphasizing life expectancy, education, and GNI
-per capita respectively) using the \texttt{swap\_exprs()} function. The
-function first locates the operation to be tested
-(\texttt{.var\ =\ index}), then accepts a list of expressions for
-testing \texttt{.expr}, before requesting the raw data:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{(res2 }\OtherTok{\textless{}{-}}\NormalTok{ res }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{swap\_exprs}\NormalTok{(}
-    \AttributeTok{.var =}\NormalTok{ index,}
-    \AttributeTok{.expr =} \FunctionTok{list}\NormalTok{(}
-      \AttributeTok{index1 =}\NormalTok{ (life\_exp }\SpecialCharTok{+}\NormalTok{ sch }\SpecialCharTok{+}\NormalTok{ gni\_pc)}\SpecialCharTok{/}\DecValTok{3}\NormalTok{,}
-      \CommentTok{\# PCA recommended weight}
-      \AttributeTok{index2 =} \FloatTok{0.569} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.576} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.586} \SpecialCharTok{*}\NormalTok{ gni\_pc, }
-      \AttributeTok{index3 =} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
-      \AttributeTok{index4 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ gni\_pc,}
-      \AttributeTok{index5 =} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ life\_exp }\SpecialCharTok{+} \FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ sch }\SpecialCharTok{+} \FloatTok{0.8} \SpecialCharTok{*}\NormalTok{ gni\_pc),}
-    \AttributeTok{.raw\_data =}\NormalTok{ dt))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-Index pipeline:
-
-Summary: 
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `log(gni_pc, base = 10)` -> gni_pc
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = life_exp, min = 20, max = 85)` ->
-life_exp
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = exp_sch, min = 0, max = 18)` ->
-exp_sch
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = avg_sch, min = 0, max = 15)` ->
-avg_sch
-\end{verbatim}
-
-\begin{verbatim}
-var_trans: `rescale_minmax(var = gni_pc, min = 2, max = 4.875061)` ->
-gni_pc
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `(exp_sch + avg_sch)/2` -> sch
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `(life_exp * sch * gni_pc)^(1/3)` -> index0
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `(life_exp + sch + gni_pc)/3` -> index1
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `0.569 * life_exp + 0.576 * sch + 0.586 * gni_pc` -> index2
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `0.8 * life_exp + 0.1 * sch + 0.1 * gni_pc` -> index3
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `0.1 * life_exp + 0.8 * sch + 0.1 * gni_pc` -> index4
-\end{verbatim}
-
-\begin{verbatim}
-dim_red: `0.1 * life_exp + 0.1 * sch + 0.8 * gni_pc` -> index5
-\end{verbatim}
-
-\begin{verbatim}
-
-Data: 
-# A tibble: 191 x 13
-      id country   life_exp exp_sch avg_sch gni_pc   sch index0 index1
-   <dbl> <chr>        <dbl>   <dbl>   <dbl>  <dbl> <dbl>  <dbl>  <dbl>
- 1     1 Switzerl~    0.984   0.917   0.924  0.983 0.920  0.962  0.963
- 2     2 Norway       0.973   1       0.867  0.978 0.933  0.961  0.961
- 3     3 Iceland      0.964   1       0.918  0.955 0.959  0.959  0.959
- 4     4 Hong Kon~    1       0.960   0.815  0.973 0.887  0.952  0.953
- 5     5 Australia    0.993   1       0.848  0.936 0.924  0.951  0.951
- 6     6 Denmark      0.944   1       0.864  0.967 0.932  0.948  0.948
- 7     7 Sweden       0.969   1       0.841  0.952 0.920  0.947  0.947
- 8     8 Ireland      0.954   1       0.772  1     0.886  0.945  0.947
- 9     9 Germany      0.933   0.945   0.939  0.952 0.942  0.942  0.942
-10    10 Netherla~    0.949   1       0.839  0.956 0.919  0.941  0.941
-# i 181 more rows
-# i 4 more variables: index2 <dbl>, index3 <dbl>, index4 <dbl>,
-#   index5 <dbl>
-\end{verbatim}
-
-The resulting data includes five more columns corresponding to the five
-alternative indexes. The country ranking can be further computed on the
-data and plotted in a scatterplot matrix as in
-Figure~\ref{fig-hdi-expr}. The plot shows almost similar ranks suggested
-by the indexes calculated using geometric mean (\texttt{rank0}), the
-arithmetic mean (\texttt{rank1}), and the PCA recommended weights
-(\texttt{rank2}). The agreement of the three methods supports the use of
-geometric mean as the dimension reduction method in the original index.
-The scatterplot matrix also reveals the difference in ranking when the
-three dimensions are given different weights in \texttt{rank3},
-\texttt{rank4}, and \texttt{rank5}. The panel \texttt{rank4} vs
-\texttt{rank5} shows a cluster of highlighted points with relatively
-high ranks according to \texttt{rank5} but low in \texttt{rank4}. These
-countries are printed in Table~\ref{tbl-hdi-expr} prints along with
-their three dimensions, calculated indexes, and rankings. While having a
-high life expectancy and GNI \emph{per capita}, these countries
-(Andorra, Qatar, San Marino, Kuwait, and Brunei Darussalam) score
-relatively low in education, which is weighted heavily in
-\texttt{index4}
-(\texttt{0.1\ *\ life\_exp\ +\ 0.1\ *\ sch\ +\ 0.8\ *\ gni\_pc}).
+\hypertarget{global-gender-gap-index}{%
+\subsection{Global Gender Gap Index}\label{global-gender-gap-index}}
+
+\includegraphics{tidyindex_files/figure-pdf/gggi-verify-1.pdf}
 
@@
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/fig-hdi-expr-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/gggi-weights-1.pdf}
 
@@
 
-\caption{\label{fig-hdi-expr}Comparing country ranks from six different
-index constructions: geometric mean (\texttt{rank0}), arithmetic mean
-(\texttt{rank1}), principal component analysis recommended weights
-(\texttt{rank2}), and heavier weight on life expectancy, education, and
-GNI per capita (\texttt{rank3} - \texttt{rank5}). Geometric mean,
-arithmetic mean, and PCA produce similar rankings where countries are
-aligned in a diagnoal line, while country rankings varies when compare
-among \texttt{rank3} to \texttt{rank5}. Panel \texttt{rank5}
-vs.~\texttt{rank4} highlights a cluster of countries that rank high by
-\texttt{rank5} but low by \texttt{rank4}. The plot demonstrates the
-necessity to test on various alternative expressions to verify the
-robustness of the method and to capture distinctive characteristics of
-different countries.}
-
 \end{figure}
 
-\hypertarget{tbl-hdi-expr}{}
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2118}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1882}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1176}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1765}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0824}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0824}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0706}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0706}}@{}}
-\caption{\label{tbl-hdi-expr}A selected number of countries with low
-rankings when education is given a heavy weight (\texttt{index4}) while
-ranks high when a heavier weight is given on GNI per capital
-(\texttt{index5}).}\tabularnewline
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-country
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-life expectancy
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-education
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-GNI per capita
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-index3
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-index4
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-rank3
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-rank4
-\end{minipage} \\
-\midrule()
-\endfirsthead
-\toprule()
-\begin{minipage}[b]{\linewidth}\raggedright
-country
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-life expectancy
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-education
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-GNI per capita
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-index3
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-index4
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-rank3
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-rank4
-\end{minipage} \\
-\midrule()
-\endhead
-Andorra & 0.929 & 0.721 & 0.942 & 0.909 & 0.764 & 32 & 69 \\
-Qatar & 0.912 & 0.684 & 1.000 & 0.898 & 0.739 & 34 & 84 \\
-San Marino & 0.937 & 0.701 & 0.947 & 0.914 & 0.749 & 30 & 76 \\
-Kuwait & 0.903 & 0.670 & 0.947 & 0.884 & 0.721 & 39 & 96 \\
-Brunei Darussalam & 0.841 & 0.694 & 0.977 & 0.840 & 0.737 & 60 & 85 \\
-\bottomrule()
-\end{longtable}
-
-\newpage
+\begin{figure}
+
+{\centering \includegraphics{tidyindex_files/figure-pdf/idx-tour-1.pdf}
+
+}
+
+\caption{Five frames selected from varying the linear weights of four
+dimensions in Global Gender Gap Index. The weights vary slightly from
+the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe
+how the index and ranking reponse. Full animation is available at
+{[}vimeo link{]}.}
+
+\end{figure}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-UNDP_2021_tech_notes}{}}%
-United Nations Development Programme. 2021. {``Human Development Report
-2021-2022: The Next Frontier: Human Development and the Anthropocene.''}
-\url{https://hdr.undp.org/sites/default/files/2021-22_HDR/hdr2021-22_technical_notes.pdf}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
----------. 2022. {``Human Development Report 2021-22.''} New York.
-\url{http://report.hdr.undp.org}.
+United Nations Development Programme. 2022. {``Human Development Report
+2021-22.''} New York. \url{http://report.hdr.undp.org}.
 

### 2023-07-24 tidyindex.qmd
write up for the gggi index section

@@
 
-Indexes are commonly-used tools to monitor, communicate and make decisions across various domains in society.
+Indexes are commonly used to combine multivariate information into a single number for monitoring, communicating, and decision-making across various domains in society.
 Their applications can be found in the environment (e.g.
@@
 To construct an index, experts typically first define an underlying concept of interest to measure, then collect relevant variables for measuring the concept, and lastly summarise these variables into a single numerical or categorical value.
+Many decisions are factored into this process and
+
 An index can be as simple as counting the number of days when the maximum temperature is greater than 25 degree.
@@
 
+Global Gender Gap Index (GGGI) [@WEF2023] is published by the World Economic Forum to measure the gender parity of 146 countries across four dimensions: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment.
+The index is expressed as a percentage and can be interpreted as the percentage of the gender gap that has been closed.
+Fourteen variables from the four dimensions are collected for both genders and calculated as females-to-males ratios.
+Within each dimension, variables are linear aggregated with weights shown in @tbl-gggi-weights.
+The weights is determined so that a one percentage point change in the standard deviation of each variable contributes equally to the index and is calculated as taking the inverse of the standard deviation of each variable and scaling them to sum to 1 within each dimension.
+The four dimensions are then aggregated with equal weights (0.25) into the final index.
+The 2023 GGGI data is available in the country's economy profile from the Global Gender Gap Report 2023 [@WEF2023] and can be accessed through the `tidyindex` package as `gggi`, along with the corresponding weights `gggi_weights`.
+
+```{r tbl-gggi-weights}
+#| tbl-cap: Weights of the fourteen variables in Global Gender Gap Index
+#| tbl-cap-location: bottom
+gggi_weights %>%  
+   mutate(variable = str_replace_all(variable, "_", " ") %>%
+            str_to_sentence()) %>% 
+  select(c(1,2, 5:7)) %>% 
+  mutate(dimension = c(rep("Economy", 5), rep("Education" ,4), 
+                       rep("Health", 2), rep("Politics", 3))) %>% 
+  set_names(str_to_sentence) %>% 
+  knitr::kable(digits = 3)
+```
+
+A natural thing to do when provided with the index data is to reproduce the index using the described methodology to verify the index calculation.
+In the case of GGGI, the construction can be simplified as a single linear aggregation with variable weight (rescaled inverse of variable standard deviation) multiplied by the dimension weight (0.25 for all dimensions):
+
+````
+gggi %>% 
+  init(id = country) %>%
+  add_meta(gggi_weights, var_col = variable) %>% 
+  dimension_reduction(
+    index_new = aggregate_linear(
+      ~labour_force_participation:years_with_female_head_of_state,
+      weight = weight)) 
+````
+
+The result can be compared with the GGGI values available from the report as shown in @fig-compare-gggi, validating the reproducibility of the index from the country data.
+
 ```{r eval = FALSE}
@@
 
-```{r gggi-verify}
+```{r fig-compare-gggi}
+#| fig-cap: "Verifying the calculation of the Global Gender Gap Index (GGGI).
+The index can be reproduced from the methodology described in the Global Gender Gap Report 2023 after removing the countries with missing variables, the treatment of which is unclear."
 set.seed(123)
@@
 
-dim_weights <- gggi_weights %>%
-  filter(!is.na(pillar_weight)) %>%
-  select(variable, pillar_weight) %>% 
-  bind_cols(rand_w)
-
-dt <- gggi %>% 
-  select(country:political_empowerment) %>% 
+dt2 <- gggi %>% 
   init(id = country) %>%
-  add_meta(dim_weights, var_col = variable)
-
-dt2 <- dt %>%
+  add_meta(gggi_weights, var_col = variable) %>% 
   dimension_reduction(
     index_new = aggregate_linear(
-      ~economic_participation_and_opportunity:political_empowerment,
-      weight = pillar_weight))
+      ~labour_force_participation:years_with_female_head_of_state,
+      weight = weight))
 
 dt2$data %>%
+  filter(!is.na(index_new)) %>% 
   ggplot(aes(x = index, y = index_new)) +
@@
   theme_bw() + 
-  theme(aspect.ratio = 1, panel.grid.minor = element_blank()) 
+  theme(aspect.ratio = 1, panel.grid.minor = element_blank()) + 
+  xlab("Benchmark") + 
+  ylab("Index calculated")
 ```
 
-```{r gggi-weights, fig.height=20, fig.align='center', fig.width=10}
+```{r eval = FALSE, gggi-weights, fig.height=20, fig.align='center', fig.width=10}
 dt3 <- dt2 %>%
@@
 ```
+
+Index analysts could be interested in understanding how the index values and country ranking change when the weights assigned the each dimension is slightly changed.
+This can help assess the sensitivity of the country ranking, test the robustness of the indexes to different variables, and identify the strengths and weaknesses of countries in each aspect.
+We run a local tour focusing on the four dimensions of countries in the South Asia and Sub-Saharan Africa region.
+The local tour generates an animation starting with an equal weight of 0.25 for all four dimensions and gradually increases and reduces the weight of one variable at a time.
+In @fig-idx-tour, we select five frames from the tour animation: the first frame represents equal weights, and the subsequent four frames have a higher weight for each of the four dimensions.
+When health receives a higher weight, the impact on the index value is minimal.
+Index values and ranking change slightly when the dimension economy and education are given a greater weight, highlighting countries with a relatively better (moving right) or worse (moving left) in these single dimensions.
+For example, when economy is given a higher weight, the index moves to the left for Ethiopia and Senegal, revealing it as a weakness for these two countries to improve compared to their peers with similar rankings.
+These information will not be observed when using a single equal weight scheme since the advantageous and disadvantageous dimension of a country will be averaged out.
+
+Notably, increasing the weight of the politics dimension leads to a pronounced decrease of index value for almost all the countries.
+This can further impact the interpretation on how much of the gender gap has been closed.
+Analysts need to further consider the robustness of the index and potentially investigate the construction of the politics dimension, including the variable selection process.
 
@@
 
-```{r idx-tour, fig.align='center', fig.height= 12, fig.width=12}
+```{r fig-idx-tour, fig.align='center', fig.height= 12, fig.width=12}
 #| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index.

### 2023-07-24 tidyindex.tex
write up for the gggi index section

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-Indexes are commonly-used tools to monitor, communicate and make
-decisions across various domains in society. Their applications can be
-found in the environment (e.g.~Air Quality Index, El Niño-Southern
-Oscillation Index), in the economy (Consumer Price Index, stock market
-indexes), and in the social sciences (e.g.~QS University Rankings, Human
-Development Index).
+Indexes are commonly used to combine multivariate information into a
+single number for monitoring, communicating, and decision-making across
+various domains in society. Their applications can be found in the
+environment (e.g.~Air Quality Index, El Niño-Southern Oscillation
+Index), in the economy (Consumer Price Index, stock market indexes), and
+in the social sciences (e.g.~QS University Rankings, Human Development
+Index).
 
@@
 measuring the concept, and lastly summarise these variables into a
-single numerical or categorical value. An index can be as simple as
-counting the number of days when the maximum temperature is greater than
-25 degree. This is ``the number of summer days'' (SU), a climate index
-listed in ETCCDI Climate Change Indices (Zhang 2020). In some cases,
-indexes are weighted linear combinations of variables. For example,
-Human Development Index (HDI) (United Nations Development Programme
-2022) combines the three aspects (life expectancy, years of schooling,
-and GNI per capita) to measure country's development. In more complex
-scenarios, indexes are constructed by combining multiple variables,
-often accompanied by spatial and temporal considerations arising from
-diverse data formats, including in-situ stations, satellite imageries,
-and model outputs. An example of this is the Combined Drought Index
-(CDI) (Sepulcre-Canto et al. 2012), curated by the European Drought
-Observatory, which uses a decision tree to combine Standardized
-Precipitation Index (SPI) (McKee, Doesken, and Kleist 1993) with the
-satellite measure, Fraction of Absorbed Photosynthetically Active
-Radiation (FAPAR), and modelled soil moisture.
+single numerical or categorical value. Many decisions are factored into
+this process and
+
+An index can be as simple as counting the number of days when the
+maximum temperature is greater than 25 degree. This is ``the number of
+summer days'' (SU), a climate index listed in ETCCDI Climate Change
+Indices (Zhang 2020). In some cases, indexes are weighted linear
+combinations of variables. For example, Human Development Index (HDI)
+(United Nations Development Programme 2022) combines the three aspects
+(life expectancy, years of schooling, and GNI per capita) to measure
+country's development. In more complex scenarios, indexes are
+constructed by combining multiple variables, often accompanied by
+spatial and temporal considerations arising from diverse data formats,
+including in-situ stations, satellite imageries, and model outputs. An
+example of this is the Combined Drought Index (CDI) (Sepulcre-Canto et
+al. 2012), curated by the European Drought Observatory, which uses a
+decision tree to combine Standardized Precipitation Index (SPI) (McKee,
+Doesken, and Kleist 1993) with the satellite measure, Fraction of
+Absorbed Photosynthetically Active Radiation (FAPAR), and modelled soil
+moisture.
 
@@
 
-\includegraphics{tidyindex_files/figure-pdf/gggi-verify-1.pdf}
+Global Gender Gap Index (GGGI) (World Economic Forum 2023) is published
+by the World Economic Forum to measure the gender parity of 146
+countries across four dimensions: Economic Participation and
+Opportunity, Educational Attainment, Health and Survival, and Political
+Empowerment. The index is expressed as a percentage and can be
+interpreted as the percentage of the gender gap that has been closed.
+Fourteen variables from the four dimensions are collected for both
+genders and calculated as females-to-males ratios. Within each
+dimension, variables are linear aggregated with weights shown in
+Table~\ref{tbl-gggi-weights}. The weights is determined so that a one
+percentage point change in the standard deviation of each variable
+contributes equally to the index and is calculated as taking the inverse
+of the standard deviation of each variable and scaling them to sum to 1
+within each dimension. The four dimensions are then aggregated with
+equal weights (0.25) into the final index. The 2023 GGGI data is
+available in the country's economy profile from the Global Gender Gap
+Report 2023 (World Economic Forum 2023) and can be accessed through the
+\texttt{tidyindex} package as \texttt{gggi}, along with the
+corresponding weights \texttt{gggi\_weights}.
+
+\hypertarget{tbl-gggi-weights}{}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.5185}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1235}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1358}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1358}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0864}}@{}}
+\caption{\label{tbl-gggi-weights}Weights of the fourteen variables in
+Global Gender Gap Index}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Variable
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Dimension
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Var\_weight
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Dim\_weight
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Weight
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Variable
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Dimension
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Var\_weight
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Dim\_weight
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Weight
+\end{minipage} \\
+\midrule()
+\endhead
+Labour force participation & Economy & 0.199 & 0.25 & 0.050 \\
+Wage equality for similar work & Economy & 0.310 & 0.25 & 0.078 \\
+Estimated earned income & Economy & 0.221 & 0.25 & 0.055 \\
+Legislators senior officials and managers & Economy & 0.149 & 0.25 &
+0.037 \\
+Professional and technical workers & Economy & 0.121 & 0.25 & 0.030 \\
+Literacy rate & Education & 0.191 & 0.25 & 0.048 \\
+Enrolment in primary education & Education & 0.459 & 0.25 & 0.115 \\
+Enrolment in secondary education & Education & 0.230 & 0.25 & 0.058 \\
+Enrolment in tertiary education & Education & 0.121 & 0.25 & 0.030 \\
+Sex ratio at birth & Health & 0.693 & 0.25 & 0.173 \\
+Healthy life expectancy & Health & 0.307 & 0.25 & 0.077 \\
+Women in parliament & Politics & 0.310 & 0.25 & 0.078 \\
+Women in ministerial positions & Politics & 0.247 & 0.25 & 0.062 \\
+Years with female head of state & Politics & 0.443 & 0.25 & 0.111 \\
+\bottomrule()
+\end{longtable}
+
+A natural thing to do when provided with the index data is to reproduce
+the index using the described methodology to verify the index
+calculation. In the case of GGGI, the construction can be simplified as
+a single linear aggregation with variable weight (rescaled inverse of
+variable standard deviation) multiplied by the dimension weight (0.25
+for all dimensions):
+
+\begin{verbatim}
+gggi %>% 
+  init(id = country) %>%
+  add_meta(gggi_weights, var_col = variable) %>% 
+  dimension_reduction(
+    index_new = aggregate_linear(
+      ~labour_force_participation:years_with_female_head_of_state,
+      weight = weight)) 
+\end{verbatim}
+
+The result can be compared with the GGGI values available from the
+report as shown in Figure~\ref{fig-compare-gggi}, validating the
+reproducibility of the index from the country data.
 
@@
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/gggi-weights-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-compare-gggi-1.pdf}
 
@@
 
+\caption{\label{fig-compare-gggi}Verifying the calculation of the Global
+Gender Gap Index (GGGI). The index can be reproduced from the
+methodology described in the Global Gender Gap Report 2023 after
+removing the countries with missing variables, the treatment of which is
+unclear.}
+
 \end{figure}
 
+Index analysts could be interested in understanding how the index values
+and country ranking change when the weights assigned the each dimension
+is slightly changed. This can help assess the sensitivity of the country
+ranking, test the robustness of the indexes to different variables, and
+identify the strengths and weaknesses of countries in each aspect. We
+run a local tour focusing on the four dimensions of countries in the
+South Asia and Sub-Saharan Africa region. The local tour generates an
+animation starting with an equal weight of 0.25 for all four dimensions
+and gradually increases and reduces the weight of one variable at a
+time. In Figure~\ref{fig-idx-tour}, we select five frames from the tour
+animation. The first frame represents equal weights, and the subsequent
+four frames have a higher weight for each of the four dimensions. When
+health receives a higher weight, the impact on the index value is
+minimal. Index values and ranking change slightly when the dimension
+economy and education are given a greater weight, highlighting countries
+with a relatively better (moving right) or worse (moving left) in these
+single dimensions. For example, when economy is given a higher weight,
+the index moves to the left for Ethiopia and Senegal, revealing it as a
+weakness for these two countries to improve compared to their peers with
+similar rankings. These information will not be observed when using a
+single equal weight scheme since the advantageous and disadvantageous
+dimension of a country will be averaged out.
+
+Notably, increasing the weight of the politics dimension leads to a
+pronounced decrease of index value for almost all the countries. This
+can further impact the interpretation on how much of the gender gap has
+been closed. Analysts need to further consider the robustness of the
+index and potentially investigate the construction of the politics
+dimension, including the variable selection process.
+
 \begin{figure}
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/idx-tour-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-idx-tour-1.pdf}
 
@@
 
-\caption{Five frames selected from varying the linear weights of four
-dimensions in Global Gender Gap Index. The weights vary slightly from
-the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe
-how the index and ranking reponse. Full animation is available at
-{[}vimeo link{]}.}
+\caption{\label{fig-idx-tour}Five frames selected from varying the
+linear weights of four dimensions in Global Gender Gap Index. The
+weights vary slightly from the official simple average weights (0.25,
+0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full
+animation is available at {[}vimeo link{]}.}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-WEF2023}{}}%
+World Economic Forum. 2023. {``{The Global Gender Gap Report 2023}.''}
+\url{https://www3.weforum.org/docs/WEF_GGGR_2023.pdf}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-xie_reactive_2014}{}}%

### 2023-07-24 tidyindex.qmd
vimeo link to idx tour video

@@
 The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse.
-Full animation is available at [vimeo link]."
+Full animation is available at https://vimeo.com/847874016?share=copy."
 frames <- c("001", "013", "037", "061", "085")

### 2023-07-24 tidyindex.tex
vimeo link to idx tour video

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 time. In Figure~\ref{fig-idx-tour}, we select five frames from the tour
-animation. The first frame represents equal weights, and the subsequent
+animation: the first frame represents equal weights, and the subsequent
 four frames have a higher weight for each of the four dimensions. When
@@
 0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full
-animation is available at {[}vimeo link{]}.}
+animation is available at https://vimeo.com/847874016?share=copy.}
 

### 2023-07-25 tidyindex.qmd
proofread the GGGI example section

--- 2023-07-19 tidyindex.qmd
@@
 
-To construct an index, experts typically first define an underlying concept of interest to measure, then collect relevant variables for measuring the concept, and lastly summarise these variables into a single numerical or categorical value.
-An index can be as simple as counting the number of days when the maximum temperature is greater than 25 degree.
-This is "the number of summer days" (SU), a climate index listed in ETCCDI Climate Change Indices [@climate-indexes].
-In some cases, indexes are weighted linear combinations of variables.
-For example, Human Development Index (HDI) [@hdi] combines the three aspects (life expectancy, years of schooling, and GNI per capita) to measure country's development.
-In more complex scenarios, indexes are constructed by combining multiple variables, often accompanied by spatial and temporal considerations arising from diverse data formats, including in-situ stations, satellite imageries, and model outputs.
-An example of this is the Combined Drought Index (CDI) [@cdi], curated by the European Drought Observatory, which uses a decision tree to combine Standardized Precipitation Index (SPI) [@spi] with the satellite measure, Fraction of Absorbed Photosynthetically Active Radiation (FAPAR), and modelled soil moisture.
+To construct an index, experts typically first define an underlying concept of interest to measure.
+This concept is latent and often without direct measure but of social and public importance.
+Relevant and available variables are then defined and collected for measuring the concept and combined as an index using statistical methods.
+Many decisions are factored into the construction process of an index.
+For example, decision on which variables to include may depend on the data availability and the choice in other similar indexes.
+Indexes using weighted linear combination of variables, for example, Human Development Index (HDI) [@hdi] combines the three aspects (life expectancy, years of schooling, and GNI per capita), needs to decide on the weights of each variable.
+In the case of a drought index, analysts may need to decide among similar indexes, on the time scale to aggregate precipitation, and the distribution used to fit the data.
+All these decisions can introduce uncertainty to the index and lead to different interpretation and actions in operation.
 
-Our initial interest is to understand how different variable combinations contribute to the uncertainty of an index.
-However, as we start to understand how indexes are composed, we find many other steps in the construction also introduce uncertainty.
-In order to evaluate the uncertainty, an index needs to be broken down into its fundamental building blocks to analyse the effect of each component.
-This shifts the focus to investigate the decomposition of indexes into their constituent parts for analysing uncertainty.
-While the OECD handbook [@oecd_handbook_2008] has provided a set of steps for constructing composite socio-economic indexes, there is a need to extend these guidelines to accommodate more complex construction steps required for indexes in general.
+To understand how different factors contribute to the uncertainty of an index, it needs to be broken down into its fundamental building blocks to analyse the effect of each component.
+Indexes from various disciplines, i.e.
+[...] seems very different, can be boiled down into a [...].
+Such decomposition of indexes standardise the index construction pipeline and offers benefit for comparing among indexes.
+In social indexes, the OECD handbook [@oecd_handbook_2008] has provided a set of steps for constructing composite socio-economic indexes, but there is a need to extend these guidelines to accommodate more complex construction steps required for indexes in general.
 
@@
 
+The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023].
+The index is composed of 14 variables, expressed as female-to-male ratios, which are first aggregated in a linear combination into the four dimensions using the weight from the `V-weight` column in @tbl-gggi-weights.
+The weight is calculated as the inverse of the standard deviation of each variable and scaling to sum to 1 within each dimension to allow a one percentage point change in the standard deviation of each variable to contribute equally to the index.
+The four dimensions are then aggregated in another linear combination with equal weight tp obtain the index.
+The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi`, along with the corresponding weights `gggi_weights`.
+
+```{r tbl-gggi-weights}
+#| tbl-cap: Weights of the fourteen variables in Global Gender Gap Index
+#| tbl-cap-location: bottom
+gggi_weights %>%  
+   mutate(variable = str_replace_all(variable, "_", " ") %>%
+            str_to_sentence()) %>% 
+  select(c(1,2, 5:7)) %>% 
+  mutate(dimension = c(rep("Economy", 5), rep("Education" ,4), 
+                       rep("Health", 2), rep("Politics", 3))) %>% 
+  knitr::kable(
+    digits = 3, 
+    col.names = c("Variable", "Dimension", "V-weight",
+        "D-weight", "Weight")
+    ) 
+```
+
+A natural thing to do when provided with the index data is to reproduce the index.
+This helps index analysts to verify the index calculation and become familiar with the methodology.
+For GGGI, the construction can be simplified as a single linear aggregation step in the dimension reduction module, with the `Weight` column in @tbl-gggi-weights, which is the product of the variable weight (`V-weight`) and dimension weight (`D-weight`).
+
+````
+gggi %>% 
+  init(id = country) %>%
+  add_meta(gggi_weights, var_col = variable) %>% 
+  dimension_reduction(
+    index_new = aggregate_linear(
+      ~labour_force_participation:years_with_female_head_of_state,
+      weight = weight)) 
+````
+
+The result can be compared with the GGGI values available in the report as shown in @fig-compare-gggi, validating the reproducibility of the index for country with no missing variables.
+
 ```{r eval = FALSE}
@@
 ```
+
+To understand the uncertainty of this dimension reduction step while avoiding the missingness issue on the variable level, we can run a local tour to slightly vary the weight of each dimension to see how index value and country ranking changes.
+We select countries in the South Asia and Sub-Saharan Africa region and gradually increase the weight of one variable and reduce back to equal weight all four dimensions, one at a time, to produce an animation of how GGGI changes in each country.
+Five frames (equal weights and one for each dimension with a relatively higher weight) selected from the tour animation are shown in @fig-idx-tour and you can find the link to the full animation in figure caption.
+
+Many insights can be derived from the animation and these selected frames.
+These can be helpful in understanding the sensitivity of the index value and country ranking to each variable and identifying the strengths and weaknesses of each country relative to others.
+When economy and education are given a greater weight, the variation in index values and ranking highlights countries that performs relatively better (moving right) or worse (moving left) in these two single dimensions.
+For example, the index moves to the left for Ethiopia and Senegal when economy is given a higher weight.
+This reveals the economy as a weakness for these two countries compared to their similarly-ranked peers.
+When health receives a higher weight, the impact on the index value is minimal.
+Notably, increasing the weight in politics leads to a pronounced decrease of index value for almost all the countries.
+Given the index value has a direct interpretation on how much of the gender gap has been closed, such a weight variation can cast a negative light on the progression to gender parity.
+Analysts need to further investigate the construction of the politics dimension and consider whether this is desirable for the index.
 
@@
 
-```{r idx-tour, fig.align='center', fig.height= 12, fig.width=12}
+```{r fig-idx-tour, fig.align='center', fig.height= 12, fig.width=12}
 #| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index.
 The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse.
-Full animation is available at [vimeo link]."
+Full animation is available at https://vimeo.com/847874016?share=copy."
 frames <- c("001", "013", "037", "061", "085")

### 2023-07-25 tidyindex.tex
proofread the GGGI example section

--- 2023-07-19 tidyindex.tex
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-Indexes are commonly-used tools to monitor, communicate and make
-decisions across various domains in society. Their applications can be
-found in the environment (e.g.~Air Quality Index, El Niño-Southern
-Oscillation Index), in the economy (Consumer Price Index, stock market
-indexes), and in the social sciences (e.g.~QS University Rankings, Human
-Development Index).
+Indexes are commonly used to combine multivariate information into a
+single number for monitoring, communicating, and decision-making across
+various domains in society. Their applications can be found in the
+environment (e.g.~Air Quality Index, El Niño-Southern Oscillation
+Index), in the economy (Consumer Price Index, stock market indexes), and
+in the social sciences (e.g.~QS University Rankings, Human Development
+Index).
 
 To construct an index, experts typically first define an underlying
-concept of interest to measure, then collect relevant variables for
-measuring the concept, and lastly summarise these variables into a
-single numerical or categorical value. An index can be as simple as
-counting the number of days when the maximum temperature is greater than
-25 degree. This is ``the number of summer days'' (SU), a climate index
-listed in ETCCDI Climate Change Indices (Zhang 2020). In some cases,
-indexes are weighted linear combinations of variables. For example,
-Human Development Index (HDI) (United Nations Development Programme
-2022) combines the three aspects (life expectancy, years of schooling,
-and GNI per capita) to measure country's development. In more complex
-scenarios, indexes are constructed by combining multiple variables,
-often accompanied by spatial and temporal considerations arising from
-diverse data formats, including in-situ stations, satellite imageries,
-and model outputs. An example of this is the Combined Drought Index
-(CDI) (Sepulcre-Canto et al. 2012), curated by the European Drought
-Observatory, which uses a decision tree to combine Standardized
-Precipitation Index (SPI) (McKee, Doesken, and Kleist 1993) with the
-satellite measure, Fraction of Absorbed Photosynthetically Active
-Radiation (FAPAR), and modelled soil moisture.
-
-Our initial interest is to understand how different variable
-combinations contribute to the uncertainty of an index. However, as we
-start to understand how indexes are composed, we find many other steps
-in the construction also introduce uncertainty. In order to evaluate the
-uncertainty, an index needs to be broken down into its fundamental
-building blocks to analyse the effect of each component. This shifts the
-focus to investigate the decomposition of indexes into their constituent
-parts for analysing uncertainty. While the OECD handbook (OECD, European
-Union, and Joint Research Centre - European Commission 2008) has
-provided a set of steps for constructing composite socio-economic
-indexes, there is a need to extend these guidelines to accommodate more
-complex construction steps required for indexes in general.
+concept of interest to measure. This concept is latent and often without
+direct measure but of social and public importance. Relevant and
+available variables are then defined and collected for measuring the
+concept and combined as an index using statistical methods. Many
+decisions are factored into the construction process of an index. For
+example, decision on which variables to include may depend on the data
+availability and the choice in other similar indexes. Indexes using
+weighted linear combination of variables, for example, Human Development
+Index (HDI) (United Nations Development Programme 2022) combines the
+three aspects (life expectancy, years of schooling, and GNI per capita),
+needs to decide on the weights of each variable. In the case of a
+drought index, analysts may need to decide among similar indexes, on the
+time scale to aggregate precipitation, and the distribution used to fit
+the data. All these decisions can introduce uncertainty to the index and
+lead to different interpretation and actions in operation.
+
+To understand how different factors contribute to the uncertainty of an
+index, it needs to be broken down into its fundamental building blocks
+to analyse the effect of each component. Indexes from various
+disciplines, i.e.~{[}\ldots{]} seems very different, can be boiled down
+into a {[}\ldots{]}. Such decomposition of indexes standardise the index
+construction pipeline and offers benefit for comparing among indexes. In
+social indexes, the OECD handbook (OECD, European Union, and Joint
+Research Centre - European Commission 2008) has provided a set of steps
+for constructing composite socio-economic indexes, but there is a need
+to extend these guidelines to accommodate more complex construction
+steps required for indexes in general.
 
@@
 
-\includegraphics{tidyindex_files/figure-pdf/gggi-verify-1.pdf}
+The Global Gender Gap Index (GGGI), published annually by the World
+Economic Forum, measures gender parity by assessing relative gaps
+between men and women in four key areas: Economic Participation and
+Opportunity, Educational Attainment, Health and Survival, and Political
+Empowerment (World Economic Forum 2023). The index is composed of 14
+variables, expressed as female-to-male ratios, which are first
+aggregated in a linear combination into the four dimensions using the
+weight from the \texttt{V-weight} column in
+Table~\ref{tbl-gggi-weights}. The weight is calculated as the inverse of
+the standard deviation of each variable and scaling to sum to 1 within
+each dimension to allow a one percentage point change in the standard
+deviation of each variable to contribute equally to the index. The four
+dimensions are then aggregated in another linear combination with equal
+weight tp obtain the index. The 2023 GGGI data is available from the
+Global Gender Gap Report 2023 in the country's economy profile and can
+be accessed in R via the \texttt{tidyindex} package as \texttt{gggi},
+along with the corresponding weights \texttt{gggi\_weights}.
+
+\hypertarget{tbl-gggi-weights}{}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.5455}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1299}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0909}}@{}}
+\caption{\label{tbl-gggi-weights}Weights of the fourteen variables in
+Global Gender Gap Index}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Variable
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Dimension
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+V-weight
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+D-weight
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Weight
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+Variable
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Dimension
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+V-weight
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+D-weight
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+Weight
+\end{minipage} \\
+\midrule()
+\endhead
+Labour force participation & Economy & 0.199 & 0.25 & 0.050 \\
+Wage equality for similar work & Economy & 0.310 & 0.25 & 0.078 \\
+Estimated earned income & Economy & 0.221 & 0.25 & 0.055 \\
+Legislators senior officials and managers & Economy & 0.149 & 0.25 &
+0.037 \\
+Professional and technical workers & Economy & 0.121 & 0.25 & 0.030 \\
+Literacy rate & Education & 0.191 & 0.25 & 0.048 \\
+Enrolment in primary education & Education & 0.459 & 0.25 & 0.115 \\
+Enrolment in secondary education & Education & 0.230 & 0.25 & 0.058 \\
+Enrolment in tertiary education & Education & 0.121 & 0.25 & 0.030 \\
+Sex ratio at birth & Health & 0.693 & 0.25 & 0.173 \\
+Healthy life expectancy & Health & 0.307 & 0.25 & 0.077 \\
+Women in parliament & Politics & 0.310 & 0.25 & 0.078 \\
+Women in ministerial positions & Politics & 0.247 & 0.25 & 0.062 \\
+Years with female head of state & Politics & 0.443 & 0.25 & 0.111 \\
+\bottomrule()
+\end{longtable}
+
+A natural thing to do when provided with the index data is to reproduce
+the index. This helps index analysts to verify the index calculation and
+become familiar with the methodology. For GGGI, the construction can be
+simplified as a single linear aggregation step in the dimension
+reduction module, with the \texttt{Weight} column in
+Table~\ref{tbl-gggi-weights}, which is the product of the variable
+weight (\texttt{V-weight}) and dimension weight (\texttt{D-weight}).
+
+\begin{verbatim}
+gggi %>% 
+  init(id = country) %>%
+  add_meta(gggi_weights, var_col = variable) %>% 
+  dimension_reduction(
+    index_new = aggregate_linear(
+      ~labour_force_participation:years_with_female_head_of_state,
+      weight = weight)) 
+\end{verbatim}
+
+The result can be compared with the GGGI values available in the report
+as shown in Figure~\ref{fig-compare-gggi}, validating the
+reproducibility of the index for country with no missing variables.
 
@@
 
+\caption{\label{fig-compare-gggi}Verifying the calculation of the Global
+Gender Gap Index (GGGI). The index can be reproduced from the
+methodology described in the Global Gender Gap Report 2023 after
+removing the countries with missing variables, the treatment of which is
+unclear.}
+
 \end{figure}
 
+To understand the uncertainty of this dimension reduction step while
+avoiding the missingness issue on the variable level, we can run a local
+tour to slightly vary the weight of each dimension to see how index
+value and country ranking changes. We select countries in the South Asia
+and Sub-Saharan Africa region and gradually increase the weight of one
+variable and reduce back to equal weight all four dimensions, one at a
+time, to produce an animation of how GGGI changes in each country. Five
+frames (equal weights and one for each dimension with a relatively
+higher weight) selected from the tour animation are shown in
+Figure~\ref{fig-idx-tour} and you can find the link to the full
+animation in figure caption.
+
+Many insights can be derived from the animation and these selected
+frames. These can be helpful in understanding the sensitivity of the
+index value and country ranking to each variable and identifying the
+strengths and weaknesses of each country relative to others. When
+economy and education are given a greater weight, the variation in index
+values and ranking highlights countries that performs relatively better
+(moving right) or worse (moving left) in these two single dimensions.
+For example, the index moves to the left for Ethiopia and Senegal when
+economy is given a higher weight. This reveals the economy as a weakness
+for these two countries compared to their similarly-ranked peers. When
+health receives a higher weight, the impact on the index value is
+minimal. Notably, increasing the weight in politics leads to a
+pronounced decrease of index value for almost all the countries. Given
+the index value has a direct interpretation on how much of the gender
+gap has been closed, such a weight variation can cast a negative light
+on the progression to gender parity. Analysts need to further
+investigate the construction of the politics dimension and consider
+whether this is desirable for the index.
+
 \begin{figure}
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/idx-tour-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-idx-tour-1.pdf}
 
@@
 
-\caption{Five frames selected from varying the linear weights of four
-dimensions in Global Gender Gap Index. The weights vary slightly from
-the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe
-how the index and ranking reponse. Full animation is available at
-{[}vimeo link{]}.}
+\caption{\label{fig-idx-tour}Five frames selected from varying the
+linear weights of four dimensions in Global Gender Gap Index. The
+weights vary slightly from the official simple average weights (0.25,
+0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full
+animation is available at https://vimeo.com/847874016?share=copy.}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-spi}{}}%
-McKee, Thomas B, Nolan J Doesken, and John Kleist. 1993. {``The
-Relationship of Drought Frequency and Duration to Time Scales.''} In
-\emph{Proceedings of the 8th Conference of Applied Climatology},
-179--84. Anaheim, CA: American Meteorological Society.
-
 \leavevmode\vadjust pre{\hypertarget{ref-oecd_handbook_2008}{}}%
@@
 \url{https://doi.org/10.1787/9789264043466-en}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cdi}{}}%
-Sepulcre-Canto, G., S. Horion, A. Singleton, H. Carrao, and J. Vogt.
-2012. {``Development of a {Combined} {Drought} {Indicator} to Detect
-Agricultural Drought in {Europe}.''} \emph{Natural Hazards and Earth
-System Sciences} 12 (11): 3519--31.
-\url{https://doi.org/10.5194/nhess-12-3519-2012}.
 

### 2023-07-26 tidyindex.qmd
an overview paragraph for the examples and example titles

@@
 For example, decision on which variables to include may depend on the data availability and the choice in other similar indexes.
-Indexes using weighted linear combination of variables, for example, Human Development Index (HDI) [@hdi] combines the three aspects (life expectancy, years of schooling, and GNI per capita), needs to decide on the weights of each variable.
+In indexes constructed from linear combination of variables, decisions need to be made on the weight assigned to each variable.
 In the case of a drought index, analysts may need to decide among similar indexes, on the time scale to aggregate precipitation, and the distribution used to fit the data.
@@
 
-## A drought index example
+This section uses the example of drought and social indexes to show the analysis made possible with the index pipeline.
+The drought index example computes two indexes with various time scales and distributions simultaneously using the pipeline framework to understand the flood and drought events in Queensland.
+The social index example focuses on the dimension reduction in Global Gender Gap Index to explore the impact of weight changes in linear combination on index value and country ranking.
+
+## Every distribution, every scale, every index all at once
 
@@
 
-## Global Gender Gap Index
+## Does a minor change in variable weights cause a tornado?
 
@@
 The weight is calculated as the inverse of the standard deviation of each variable and scaling to sum to 1 within each dimension to allow a one percentage point change in the standard deviation of each variable to contribute equally to the index.
-The four dimensions are then aggregated in another linear combination with equal weight tp obtain the index.
+The four dimensions are then aggregated in another linear combination with equal weight to obtain the index.
 The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi`, along with the corresponding weights `gggi_weights`.
@@
   dplyr::select(country: political_empowerment) %>%
-  filter(region %in% c("South Asia", "Sub-Saharan Africa")) %>%
+  filter(region %in% c("Sub-Saharan Africa")) %>%
   arrange(index) 

### 2023-07-26 tidyindex.tex
an overview paragraph for the examples and example titles

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 example, decision on which variables to include may depend on the data
-availability and the choice in other similar indexes. Indexes using
-weighted linear combination of variables, for example, Human Development
-Index (HDI) (United Nations Development Programme 2022) combines the
-three aspects (life expectancy, years of schooling, and GNI per capita),
-needs to decide on the weights of each variable. In the case of a
-drought index, analysts may need to decide among similar indexes, on the
-time scale to aggregate precipitation, and the distribution used to fit
-the data. All these decisions can introduce uncertainty to the index and
+availability and the choice in other similar indexes. In indexes
+constructed from linear combination of variables, decisions need to be
+made on the weight assigned to each variable. In the case of a drought
+index, analysts may need to decide among similar indexes, on the time
+scale to aggregate precipitation, and the distribution used to fit the
+data. All these decisions can introduce uncertainty to the index and
 lead to different interpretation and actions in operation.
@@
 
-\hypertarget{a-drought-index-example}{%
-\subsection{A drought index example}\label{a-drought-index-example}}
+This section uses the example of drought and social indexes to show the
+analysis made possible with the index pipeline. The drought index
+example computes two indexes with various time scales and distributions
+simultaneously using the pipeline framework to understand the flood and
+drought events in Queensland. The social index example focuses on the
+dimension reduction in Global Gender Gap Index to explore the impact of
+weight changes in linear combination on index value and country ranking.
+
+\hypertarget{every-distribution-every-scale-every-index-all-at-once}{%
+\subsection{Every distribution, every scale, every index all at
+once}\label{every-distribution-every-scale-every-index-all-at-once}}
 
@@
 
-\hypertarget{global-gender-gap-index}{%
-\subsection{Global Gender Gap Index}\label{global-gender-gap-index}}
+\hypertarget{does-a-minor-change-in-variable-weights-cause-a-tornado}{%
+\subsection{Does a minor change in variable weights cause a
+tornado?}\label{does-a-minor-change-in-variable-weights-cause-a-tornado}}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
-United Nations Development Programme. 2022. {``Human Development Report
-2021-22.''} New York. \url{http://report.hdr.undp.org}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%

### 2023-07-29 tidyindex.qmd
various changes

@@
 This concept is latent and often without direct measure but of social and public importance.
-Relevant and available variables are then defined and collected for measuring the concept and combined as an index using statistical methods.
-Many decisions are factored into the construction process of an index.
+Relevant and available variables are then defined, collected, and combined as an index using statistical methods to approximate the concept.
+This construction process is often not straight forward and many decisions need to be made, whether explicitly or not.
 For example, decision on which variables to include may depend on the data availability and the choice in other similar indexes.
-In indexes constructed from linear combination of variables, decisions need to be made on the weight assigned to each variable.
-In the case of a drought index, analysts may need to decide among similar indexes, on the time scale to aggregate precipitation, and the distribution used to fit the data.
-All these decisions can introduce uncertainty to the index and lead to different interpretation and actions in operation.
+Indexes constructed from linear combination of variables need to decide on the weight assigned to each variable.
+Some indexes have a spatial and/or temporal component and variables can be aggregated to different spatial resolutions and temporal scales for different monitoring purpose.
+These decisions can result in different index values and have different practical implication.
 
-To understand how different factors contribute to the uncertainty of an index, it needs to be broken down into its fundamental building blocks to analyse the effect of each component.
-Indexes from various disciplines, i.e.
-[...] seems very different, can be boiled down into a [...].
-Such decomposition of indexes standardise the index construction pipeline and offers benefit for comparing among indexes.
+To be able to test different decision choices systematically for an index, the index needs to be broken down into its fundamental building blocks to analyse the effect of each component.
+Such decomposition of indexes standardise the index construction pipeline and offers benefit for comparing among indexes and calculate index uncertainty.
 In social indexes, the OECD handbook [@oecd_handbook_2008] has provided a set of steps for constructing composite socio-economic indexes, but there is a need to extend these guidelines to accommodate more complex construction steps required for indexes in general.
@@
 The pipeline includes eight modules for operations in the spatial, temporal, and multivariate aspects of the data as well as modules for comparing and communicating indexes.
-Analysts are free to select the modules they need and arrange them in the order they see fit to construct indexes.
-While the starting point of the pipeline is raw data, there are steps prior to this that are crucial to the success of an index. 
-For example, the defined index needs to be useful for measuring the concept of interest and variables need to be collected from reliable sources with proper quality control.
+Analysts can choose a subset of modules and reorder them as needed to construct an index.
 
-  * align the spatial resolution and temporal frequency of data collected from different satellite products
-  * obtain and clean variables from countries, potentially aggregate from regional data into country level
-  * merge ground-based and satellite-based data
+Data pre-processing may happen before the start of the index pipeline to prepare multivariate data from different locations/ countries ready for constructing an index.
+For remote sensing data, it includes aligning the spatial resolution and temporal frequency of data collected from different satellite products, or merging in-situ stations with the satellite data.
 
-Before elaborating each of the eight pipeline modules as subsections, the data notation will be first introduced.
+The notation used for introducing the pipeline modules is as follows.
 Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspects: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
@@
 where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$. 
-The most commonly used dimension reduction technique is Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science.
+The most commonly used dimension reduction technique is Principal Component Analysis (PCA).
 It can be seen as a special case of weighting, where variables are summed up in a linear combination: $$x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$ with restrictions imposed on the weight coefficient: $\sum_{p=1}^P\lambda_p^2 = 1$.
@@
   dplyr::select(country: political_empowerment) %>%
-  filter(region %in% c("Sub-Saharan Africa")) %>%
+  filter(country %in% c("Namibia", "South Africa", "Bangladesh", "Botswana",
+                        "Ethiopia", "Kenya", "Senegal", "Afghanistan")) %>%
   arrange(index) 
@@
 b[,,c(1,3,5,7,9)] <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1) %>% w_idx2proj()
-b[,,2] <- matrix(c(0.5, 0.2, 0.2, 0.1), ncol = 1) %>% find_next()
-b[,,4] <- matrix(c(0.2, 0.5, 0.2, 0.1), ncol= 1) %>% find_next()
-b[,,6] <- matrix(c(0.1, 0.2, 0.5, 0.2), ncol= 1) %>% find_next()
-b[,,8] <- matrix(c(0.15, 0.2, 0.15, 0.5), ncol = 1) %>% find_next()
+b[,,2] <- matrix(c(0.7, 0.1, 0.1, 0.1), ncol = 1) %>% find_next()
+b[,,4] <- matrix(c(0.1, 0.7, 0.1, 0.1), ncol= 1) %>% find_next()
+b[,,6] <- matrix(c(0.1, 0.1, 0.5, 0.1), ncol= 1) %>% find_next()
+b[,,8] <- matrix(c(0.1, 0.1, 0.1, 0.5), ncol = 1) %>% find_next()
 
@@
   dt[,5:8], planned2_tour(b), dev = "png",
-  display_idx(cex = 2, label_cex = 2,
+  display_idx(cex = 4, label_cex = 3.5, panel_height_ratio = c(3,1), 
+              axis_cex = 2.5, weight_cex = 3, label_col = "grey70", 
               label = dt$country, abb_vars = FALSE),
-  width = 800, height = 1300, apf = 1/20, frames = 100,
+  width = 900, height = 900, apf = 1/20, frames = 100,
   here::here("figures/idx-tour/idx-tour-%03d.png")
@@
   gif_file = here::here("figures/idx-tour.gif"),
-  delay = 0.15, width = 800, height = 1300)
+  delay = 0.15, width = 900, height = 900)
 
@@
 
-```{r fig-idx-tour, fig.align='center', fig.height= 12, fig.width=12}
+```{r fig-idx-tour, fig.align='center', fig.height=8} 
 #| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index.
@@
 gl <-  lapply(rl, grid::rasterGrob)
-wrap_plots(gl, ncol = 3) 
+wrap_plots(gl, ncol = 2)  
 ```

### 2023-07-29 tidyindex.tex
various changes

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 direct measure but of social and public importance. Relevant and
-available variables are then defined and collected for measuring the
-concept and combined as an index using statistical methods. Many
-decisions are factored into the construction process of an index. For
-example, decision on which variables to include may depend on the data
-availability and the choice in other similar indexes. In indexes
-constructed from linear combination of variables, decisions need to be
-made on the weight assigned to each variable. In the case of a drought
-index, analysts may need to decide among similar indexes, on the time
-scale to aggregate precipitation, and the distribution used to fit the
-data. All these decisions can introduce uncertainty to the index and
-lead to different interpretation and actions in operation.
-
-To understand how different factors contribute to the uncertainty of an
-index, it needs to be broken down into its fundamental building blocks
-to analyse the effect of each component. Indexes from various
-disciplines, i.e.~{[}\ldots{]} seems very different, can be boiled down
-into a {[}\ldots{]}. Such decomposition of indexes standardise the index
-construction pipeline and offers benefit for comparing among indexes. In
-social indexes, the OECD handbook (OECD, European Union, and Joint
-Research Centre - European Commission 2008) has provided a set of steps
-for constructing composite socio-economic indexes, but there is a need
-to extend these guidelines to accommodate more complex construction
-steps required for indexes in general.
+available variables are then defined, collected, and combined as an
+index using statistical methods to approximate the concept. This
+construction process is often not straight forward and many decisions
+need to be made, whether explicitly or not. For example, decision on
+which variables to include may depend on the data availability and the
+choice in other similar indexes. Indexes constructed from linear
+combination of variables need to decide on the weight assigned to each
+variable. Some indexes have a spatial and/or temporal component and
+variables can be aggregated to different spatial resolutions and
+temporal scales for different monitoring purpose. These decisions can
+result in different index values and have different practical
+implication.
+
+To be able to test different decision choices systematically for an
+index, the index needs to be broken down into its fundamental building
+blocks to analyse the effect of each component. Such decomposition of
+indexes standardise the index construction pipeline and offers benefit
+for comparing among indexes and calculate index uncertainty. In social
+indexes, the OECD handbook (OECD, European Union, and Joint Research
+Centre - European Commission 2008) has provided a set of steps for
+constructing composite socio-economic indexes, but there is a need to
+extend these guidelines to accommodate more complex construction steps
+required for indexes in general.
 
@@
 and multivariate aspects of the data as well as modules for comparing
-and communicating indexes. Analysts are free to select the modules they
-need and arrange them in the order they see fit to construct indexes.
-While the starting point of the pipeline is raw data, there are steps
-prior to this that are crucial to the success of an index. For example,
-the defined index needs to be useful for measuring the concept of
-interest and variables need to be collected from reliable sources with
-proper quality control.
-
-\begin{itemize}
-\tightlist
-\item
-  align the spatial resolution and temporal frequency of data collected
-  from different satellite products
-\item
-  obtain and clean variables from countries, potentially aggregate from
-  regional data into country level
-\item
-  merge ground-based and satellite-based data
-\end{itemize}
-
-Before elaborating each of the eight pipeline modules as subsections,
-the data notation will be first introduced. Let
-\(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with spatial,
-temporal, and multivariate aspects: the spatial dimension
+and communicating indexes. Analysts can choose a subset of modules and
+reorder them as needed to construct an index.
+
+Data pre-processing may happen before the start of the index pipeline to
+prepare multivariate data from different locations/ countries ready for
+constructing an index. For remote sensing data, it includes aligning the
+spatial resolution and temporal frequency of data collected from
+different satellite products, or merging in-situ stations with the
+satellite data.
+
+The notation used for introducing the pipeline modules is as follows.
+Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with
+spatial, temporal, and multivariate aspects: the spatial dimension
 \(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
@@
 variable dimension from \(P\) to \(P^*\). The most commonly used
-dimension reduction technique is Principal Component Analysis (PCA),
-also called Empirical Orthogonal Function (EOF) in earth science. It can
-be seen as a special case of weighting, where variables are summed up in
-a linear combination:
+dimension reduction technique is Principal Component Analysis (PCA). It
+can be seen as a special case of weighting, where variables are summed
+up in a linear combination:
 \[x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]
@@
 dimensions are then aggregated in another linear combination with equal
-weight tp obtain the index. The 2023 GGGI data is available from the
+weight to obtain the index. The 2023 GGGI data is available from the
 Global Gender Gap Report 2023 in the country's economy profile and can

### 2023-07-29 tidyindex.qmd
work on dimension reduction

@@
 
-When indexes are constructed from multivariate information, dimension reduction methods combine that information into a univariate series.
-In the pipeline, dimension reduction includes methods that take multivariate inputs and output the data in a lower dimension (often univariate):
+Dimension reduction summarises the multivariate information into univariate, which can be denoted as:
 
 \begin{equation}
-x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),
+x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
 \end{equation}
 
-where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$ reduce the variable dimension from $P$ to $P^*$. 
-The most commonly used dimension reduction technique is Principal Component Analysis (PCA).
-It can be seen as a special case of weighting, where variables are summed up in a linear combination: $$x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$ with restrictions imposed on the weight coefficient: $\sum_{p=1}^P\lambda_p^2 = 1$.
-In other cases of weighting, the coefficients can be as simple as giving equal weight to each variables.
+where $p = 1, 2, \cdots, P$.
+The combination can be based on domain-specific knowledge, originated from theories describing the underlying physical process.
+For example, the SPEI uses a water balance model ($D = P - \text{PET}$) to calculate the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$).
 
-Some dimension reduction can also be formulated from domain-specific knowledge.
-This can be theories that describe the physics of the phenomenon being indexed or practical formulations used to extract useful features from the raw variables.
-For example, in the index SPEI, a difference series is calculated between precipitation and potential evapotranspiration (PET) and the validity of this formulation is backed up by climate water balance model [Thornthwaite, 1948], which describes [...].
-*Add another example of remote sensing variables i.e.
-NDVI = (NIR - Red) / (NIR + Red)?*
+Another widely used approach is linear combination, which aggregates a collection of variables in a linear additive structure, expressed as: $$x(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$
+
+where $\lambda_p$ denotes the weight assigned to variable $x_p$.
+Most indexes uses an equal weight that sum to 1, which allows each variable to contribute equally to the index.
+A linear combination can also be viewed as a linear projection of multivariate information by the weight vector.
+Projecting data from higher to lower dimension inevitably leads to information loss.
+For example, two countries can receive similar index, but their components could significantly differ -- one with average scores across all components and another with extreme scores the opposite ends.
+The selected set of weights need to be examined to understand its effect on the index and its implications for decision-making.
+To do this, analysts can vary the coefficients in the linear projection to observe how the index value and countries' ranking change.
+These changes can also be visualized using a method called tour by generating an animation of the projections among different sets of weights.
+
+<!-- Other weight constraint used includes $\sum_{p=1}^P\lambda_p^2 = 1$, is used by principal component analysis.  -->
 
@@
 <!-- which is used to evaluate the vegetation from near-infrared (NIR) and red channel. -->
-
-While suggested weights and formulas can indicate norms adored by practitioners, analysts should be given the flexibility to experiment with different combinations when constructing indexes. 
-This could help understand index behavior from its sensitivity to the variables and suggest alternative weights that better suit the specific tasks.
 

### 2023-07-29 tidyindex.tex
work on dimension reduction

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, enhanced, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-When indexes are constructed from multivariate information, dimension
-reduction methods combine that information into a univariate series. In
-the pipeline, dimension reduction includes methods that take
-multivariate inputs and output the data in a lower dimension (often
-univariate):
+Dimension reduction summarises the multivariate information into
+univariate, which can be denoted as:
 
 \begin{equation}
-x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),
+x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
 \end{equation}
 
-where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\) reduce the
-variable dimension from \(P\) to \(P^*\). The most commonly used
-dimension reduction technique is Principal Component Analysis (PCA). It
-can be seen as a special case of weighting, where variables are summed
-up in a linear combination:
-\[x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]
-with restrictions imposed on the weight coefficient:
-\(\sum_{p=1}^P\lambda_p^2 = 1\). In other cases of weighting, the
-coefficients can be as simple as giving equal weight to each variables.
-
-Some dimension reduction can also be formulated from domain-specific
-knowledge. This can be theories that describe the physics of the
-phenomenon being indexed or practical formulations used to extract
-useful features from the raw variables. For example, in the index SPEI,
-a difference series is calculated between precipitation and potential
-evapotranspiration (PET) and the validity of this formulation is backed
-up by climate water balance model {[}Thornthwaite, 1948{]}, which
-describes {[}\ldots{]}. \emph{Add another example of remote sensing
-variables i.e.~NDVI = (NIR - Red) / (NIR + Red)?}
-
-While suggested weights and formulas can indicate norms adored by
-practitioners, analysts should be given the flexibility to experiment
-with different combinations when constructing indexes. This could help
-understand index behavior from its sensitivity to the variables and
-suggest alternative weights that better suit the specific tasks.
+where \(p = 1, 2, \cdots, P\). The combination can be based on
+domain-specific knowledge, originated from theories describing the
+underlying physical process. For example, the SPEI uses a water balance
+model (\(D = P - \text{PET}\)) to calculate the difference series
+(\(D\)) from precipitation (\(P\)) and potential evapotranspiration
+(\(\text{PET}\)).
+
+Another widely used approach is linear combination, which aggregates a
+collection of variables in a linear additive structure, expressed as:
+\[x(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]
+
+where \(\lambda_p\) denotes the weight assigned to variable \(x_p\).
+Most indexes uses an equal weight that sum to 1, which allows each
+variable to contribute equally to the index. A linear combination can
+also be viewed as a linear projection of multivariate information by the
+weight vector. Projecting data from higher to lower dimension inevitably
+leads to information loss. For example, two countries can receive
+similar index, but their components could significantly differ -- one
+with average scores across all components and another with extreme
+scores the opposite ends. The selected set of weights need to be
+examined to understand its effect on the index and its implications for
+decision-making. To do this, analysts can vary the coefficients in the
+linear projection to observe how the index value and countries' ranking
+change. These changes can also be visualized using a method called tour
+by generating an animation of the projections among different sets of
+weights.
 

### 2023-07-31 tidyindex.qmd
Update Intro paragraph 1

@@
 
-Indexes are commonly used to combine multivariate information into a single number for monitoring, communicating, and decision-making across various domains in society.
-Their applications can be found in the environment (e.g.
-Air Quality Index, El Niño-Southern Oscillation Index), in the economy (Consumer Price Index, stock market indexes), and in the social sciences (e.g.
-QS University Rankings, Human Development Index).
+Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making.
+Examples of those include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings or Human Development Index among many others.
 

### 2023-07-31 tidyindex.qmd
Update Intro paragraph 2

@@
 
-To construct an index, experts typically first define an underlying concept of interest to measure.
-This concept is latent and often without direct measure but of social and public importance.
-Relevant and available variables are then defined, collected, and combined as an index using statistical methods to approximate the concept.
-This construction process is often not straight forward and many decisions need to be made, whether explicitly or not.
-For example, decision on which variables to include may depend on the data availability and the choice in other similar indexes.
-Indexes constructed from linear combination of variables need to decide on the weight assigned to each variable.
-Some indexes have a spatial and/or temporal component and variables can be aggregated to different spatial resolutions and temporal scales for different monitoring purpose.
-These decisions can result in different index values and have different practical implication.
+To construct an index, experts typically start by defining a concept of interest that requires measurement.
+This concept often lacks a direct measurable attribute or can only be measure as a composite of various processes, yet it holds social and public significance.
+To create an index, once the underlying processes involved are identified, relevant and available variables are then defined, collected, and combined using statistical methods into an index that aims at measure the process of interest.
+The construction process is often not straightforward, and decisions need to be made, such as the selection of variables to be included, which might depend on data availability and the statistical definition of the index to be used, among others.
+For instance, the indexes constructed from linear combination of variables need to decide on the weight assigned to each variable.
+Some indexes have a spatial and/or temporal components, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes.
+Hence, all these decisions can result in different index values and have different practical implications.
 

### 2023-07-31 tidyindex.qmd
Update Intro paragraph 3

@@
 
-To be able to test different decision choices systematically for an index, the index needs to be broken down into its fundamental building blocks to analyse the effect of each component.
-Such decomposition of indexes standardise the index construction pipeline and offers benefit for comparing among indexes and calculate index uncertainty.
-In social indexes, the OECD handbook [@oecd_handbook_2008] has provided a set of steps for constructing composite socio-economic indexes, but there is a need to extend these guidelines to accommodate more complex construction steps required for indexes in general.
+To be able to test different decision choices systematically for an index, the index needs to be broken down into its fundamental building blocks to analyse the contribution and effect of each component.
+Such decomposition of index components provides the means to standardise index construction via a pipeline and offers benefits for comparing among indexes and calculating index uncertainty.
+In social indexes for instance, the OECD handbook [@oecd_handbook_2008] has provided a set of steps and recommendations for constructing composite socio-economic indexes.
+However, there is still a need to extend these guidelines and methods to accommodate the inclusion of more methodological complex steps required for indexes in general.
 

### 2023-07-31 tidyindex.qmd
Update Intro paragraph 4

@@
 
-In this work, we develop a data pipeline framework where indexes from different domains can be built from.
-Based on the pipeline, an R package, tidyindex, is developed for practitioners to construct indexes and understand how an index responds to changes in different components in the pipeline.
-This work provides researchers actively developing new indexes with tools to evaluate the proposed indexes on their robustness for wider adoption.
-It also helps index analysts to diagnose an index to identify its weakness for methodology improvement.
+In this work, we aim at providing statistical and computational methods to develop a data pipeline framework for constructing and customize indexes using data.
+As a companion the R package, tidyindex, is developed  to construct and explore different.
+Furthermore, the tidyindex package can be used to explore the effects of changes on index definition, index construction steps, and construction methods.
+This work provides researchers actively developing new indexes or aiming at improving indexes with the tools to easily modify and evaluate indexes.
+It also helps index analysts diagnose indexes, carry out sensitivity analysis to assess small perturbations in the index, and identify potential weaknesses for methodology improvement.
 

### 2023-07-31 tidyindex.qmd
Update Intro paragraph 5

@@
 To be able to test different decision choices systematically for an index, the index needs to be broken down into its fundamental building blocks to analyse the contribution and effect of each component.
+We call this process of breaking the index construction into different steps the index pipeline.
 Such decomposition of index components provides the means to standardise index construction via a pipeline and offers benefits for comparing among indexes and calculating index uncertainty.
@@
 The pipeline framework for index construction is presented in @sec-index-pipeline.
-@sec-dev explains how to include a new building block in each pipeline module.
+@sec-dev explains how to add new steps into the index methodology in the form of new building blocks inside the index construction pipeline.
 Examples are given in @sec-examples to demonstrate the index construction with the pipeline built.

### 2023-07-31 tidyindex.tex
knit

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, frame hidden, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-Indexes are commonly used to combine multivariate information into a
-single number for monitoring, communicating, and decision-making across
-various domains in society. Their applications can be found in the
-environment (e.g.~Air Quality Index, El Niño-Southern Oscillation
-Index), in the economy (Consumer Price Index, stock market indexes), and
-in the social sciences (e.g.~QS University Rankings, Human Development
-Index).
-
-To construct an index, experts typically first define an underlying
-concept of interest to measure. This concept is latent and often without
-direct measure but of social and public importance. Relevant and
-available variables are then defined, collected, and combined as an
-index using statistical methods to approximate the concept. This
-construction process is often not straight forward and many decisions
-need to be made, whether explicitly or not. For example, decision on
-which variables to include may depend on the data availability and the
-choice in other similar indexes. Indexes constructed from linear
-combination of variables need to decide on the weight assigned to each
-variable. Some indexes have a spatial and/or temporal component and
-variables can be aggregated to different spatial resolutions and
-temporal scales for different monitoring purpose. These decisions can
-result in different index values and have different practical
-implication.
+Indexes are commonly used to combine and summarize different sources of
+information into a single number for monitoring, communicating, and
+decision-making. Examples of those include the Air Quality Index, El
+Niño-Southern Oscillation Index, Consumer Price Index, QS University
+Rankings or Human Development Index among many others.
+
+To construct an index, experts typically start by defining a concept of
+interest that requires measurement. This concept often lacks a direct
+measurable attribute or can only be measure as a composite of various
+processes, yet it holds social and public significance. To create an
+index, once the underlying processes involved are identified, relevant
+and available variables are then defined, collected, and combined using
+statistical methods into an index that aims at measure the process of
+interest. The construction process is often not straightforward, and
+decisions need to be made, such as the selection of variables to be
+included, which might depend on data availability and the statistical
+definition of the index to be used, among others. For instance, the
+indexes constructed from linear combination of variables need to decide
+on the weight assigned to each variable. Some indexes have a spatial
+and/or temporal components, and variables can be aggregated to different
+spatial resolutions and temporal scales, leading to various indexes for
+different monitoring purposes. Hence, all these decisions can result in
+different index values and have different practical implications.
 
@@
 index, the index needs to be broken down into its fundamental building
-blocks to analyse the effect of each component. Such decomposition of
-indexes standardise the index construction pipeline and offers benefit
-for comparing among indexes and calculate index uncertainty. In social
-indexes, the OECD handbook (OECD, European Union, and Joint Research
-Centre - European Commission 2008) has provided a set of steps for
-constructing composite socio-economic indexes, but there is a need to
-extend these guidelines to accommodate more complex construction steps
-required for indexes in general.
-
-In this work, we develop a data pipeline framework where indexes from
-different domains can be built from. Based on the pipeline, an R
-package, tidyindex, is developed for practitioners to construct indexes
-and understand how an index responds to changes in different components
-in the pipeline. This work provides researchers actively developing new
-indexes with tools to evaluate the proposed indexes on their robustness
-for wider adoption. It also helps index analysts to diagnose an index to
-identify its weakness for methodology improvement.
+blocks to analyse the contribution and effect of each component. We call
+this process of breaking the index construction into different steps the
+index pipeline. Such decomposition of index components provides the
+means to standardise index construction via a pipeline and offers
+benefits for comparing among indexes and calculating index uncertainty.
+In social indexes for instance, the OECD handbook (OECD, European Union,
+and Joint Research Centre - European Commission 2008) has provided a set
+of steps and recommendations for constructing composite socio-economic
+indexes. However, there is still a need to extend these guidelines and
+methods to accommodate the inclusion of more methodological complex
+steps required for indexes in general.
+
+In this work, we aim at providing statistical and computational methods
+to develop a data pipeline framework for constructing and customize
+indexes using data. As a companion the R package, tidyindex, is
+developed to construct and explore different. Furthermore, the tidyindex
+package can be used to explore the effects of changes on index
+definition, index construction steps, and construction methods. This
+work provides researchers actively developing new indexes or aiming at
+improving indexes with the tools to easily modify and evaluate indexes.
+It also helps index analysts diagnose indexes, carry out sensitivity
+analysis to assess small perturbations in the index, and identify
+potential weaknesses for methodology improvement.
 
@@
 Section~\ref{sec-index-pipeline}. Section~\ref{sec-dev} explains how to
-include a new building block in each pipeline module. Examples are given
-in Section~\ref{sec-examples} to demonstrate the index construction with
+add new steps into the index methodology in the form of new building
+blocks inside the index construction pipeline. Examples are given in
+Section~\ref{sec-examples} to demonstrate the index construction with
 the pipeline built.

### 2023-08-01 tidyindex.qmd
add pipeline diagrams

@@
 #| out-width: 100%
-knitr::include_graphics(here::here("figures/pipeline-steps.png"))
+knitr::include_graphics(here::here("figures/pipeline-overall.png"))
 ```
@@
 
+```{r fig-pipeline-spei}
+#| fig-align: center
+#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
+#| out-height: 90%
+#| out-width: 100%
+knitr::include_graphics(here::here("figures/pipeline-spei.png"))
+```
+
 The function `compute_indexes()` can be used to collectively compute multiple indexes.
@@
 
-```{r cache=TRUE, echo = TRUE}
+```{r cache=TRUE, echo = TRUE, message=FALSE}
 .scale <- c(6, 12, 24, 36)
-(idx <- queensland %>%
+idx <- queensland %>%
   init(id = id, time = ym) %>%
@@
     spi = idx_spi(.scale = .scale)
-  ))
+  )
 ```
@@
 For GGGI, the construction can be simplified as a single linear aggregation step in the dimension reduction module, with the `Weight` column in @tbl-gggi-weights, which is the product of the variable weight (`V-weight`) and dimension weight (`D-weight`).
+
+```{r fig-pipeline-steps-gggi}
+#| fig-align: center
+#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
+#| out-height: 90%
+#| out-width: 100%
+knitr::include_graphics(here::here("figures/pipeline-gggi.png"))
+```
 

### 2023-08-01 tidyindex.tex
add pipeline diagrams

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, frame hidden, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-steps.png}
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-overall.png}
 
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-spei.png}
+
+}
+
+\caption{\label{fig-pipeline-spei}Diagram of pipeline steps for index
+construction. will need to be updated with better design and the
+distribution fitting step.}
+
+\end{figure}
+
 The function \texttt{compute\_indexes()} can be used to collectively
@@
 \NormalTok{.scale }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{6}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{24}\NormalTok{, }\DecValTok{36}\NormalTok{)}
-\NormalTok{(idx }\OtherTok{\textless{}{-}}\NormalTok{ queensland }\SpecialCharTok{\%\textgreater{}\%}
+\NormalTok{idx }\OtherTok{\textless{}{-}}\NormalTok{ queensland }\SpecialCharTok{\%\textgreater{}\%}
   \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ id, }\AttributeTok{time =}\NormalTok{ ym) }\SpecialCharTok{\%\textgreater{}\%}
@@
     \AttributeTok{spi =} \FunctionTok{idx\_spi}\NormalTok{(}\AttributeTok{.scale =}\NormalTok{ .scale)}
-\NormalTok{  ))}
+\NormalTok{  )}
 \end{Highlighting}
@@
 [1] "Checking for missing values (`NA`): all the data must be complete. Input type is vector. Assuming the data are monthly time series starting in January, all regular (non-leap) years."
-\end{verbatim}
-
-\begin{verbatim}
-# A tibble: 128,586 x 19
-   .idx  .period id              ym  prcp  tmax  tmin  tavg  long   lat name    pet   diff
-   <chr>   <dbl> <chr>        <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <chr> <dbl>  <dbl>
- 1 spei        6 ASN00029~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5 KOWA~  85.9   84.1
- 2 spei        6 ASN00029~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5 KOWA~  85.9   84.1
- 3 spei        6 ASN00029~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7 MOUN~  47.6  -47.6
- 4 spei        6 ASN00029~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7 MOUN~  47.6  -47.6
- 5 spei        6 ASN00031~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9 CAIR~  80.3  714. 
- 6 spei        6 ASN00031~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9 CAIR~  80.3  714. 
- 7 spei        6 ASN00031~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1 WALK~  48.0  456. 
- 8 spei        6 ASN00031~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1 WALK~  48.0  456. 
- 9 spei        6 ASN00032~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6 SOUT~  70.2 1900. 
-10 spei        6 ASN00032~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6 SOUT~  70.2 1900. 
-# i 128,576 more rows
-# i 6 more variables: .scale <dbl>, .agg <dbl>, .method <chr>, .fitted <dbl>,
-#   .dist <chr>, .index <dbl>
 \end{verbatim}
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-gggi.png}
+
+}
+
+\caption{\label{fig-pipeline-steps-gggi}Diagram of pipeline steps for
+index construction. will need to be updated with better design and the
+distribution fitting step.}
+
+\end{figure}
+
 \begin{verbatim}

### 2023-08-01 tidyindex.qmd
rearrange and organise after meeting

@@
 
-## How does the pipeline constructin of an index look like? {#sec-toy-example}
-
-Consider a commonly used drought index: Standardized Precipitation-Evapotranspiration Index (SPEI)  [@spei]. Its construction involves:
-
-  1) transform the average temperature (`TMED`) into potential evapotranspiration (`pet`) 
-  2) combine precipitation (`prcp`) and potential evapotranspiration (`pet`) into a single variable `diff`
-  3) aggregate the difference series with a sliding window (`.scale`) 
-  4) fit a distribution to the aggregated series, and 
-  5) derive the index value from the normal density values.
-
-Conventionally approach may combine all these steps into a single function, with some level of modularity. However, these modules may only work for the selected index offered by the package.
-
-Under the pipeline approach, analysts first need to identify which module each step belongs to.
-
-Below shows the pseudocode for constructing SPEI with the pipeline:
-
-````
-DATA %>% 
-  var_trans(.method = thornthwaite, Tave = TMED, ..., .new_name = "pet") %>%
-  dim_red(diff = prcp - pet) %>%
-  aggregate(.var = diff, .scale = 12, .new_name = "agg") %>%
-  dist_fit(.method = "lmoms", .var = agg, .dist = DIST) %>%
-  augment(.var = agg)
-````
-
-<!-- *FWI:*  -->
-
-<!--   - *verbs as name for each step: `aggregate`, `normalise`* -->
-<!--   - *use NSE for variable name: `prcp`, `c(tavg, prcp)`* -->
-<!--   - *created variables are prefixed with dot: `.agg_prcp`* -->
-<!--   - *avoid string and use function to avoid "gamma" vs. "gam": `gamma()`, `lmom()`* -->
-
-The pipeline construct allows for multiple `.scales` and `.dist` to be evaluated in `aggregate()` and `dist_fit()` to compare index under different parameterisations.
-The result can then be passed into the `ggplot2` to crease visualisation.
-@fig-toy-example compares the SPEI calculated with two distributions (log-logistic and pearson III).
-
-Apart from evaluating multiples parameters, the pipeline approach allows the the steps written for one index can be directly extrapolate to another index building within the pipeline.
-The flexibility of the pipeline also integrate well with other existing packages, for examples, fitting distributions using L-moment is commonly used when constructing drought indexes.
-The package `lmomco` provides general L-moment fits to a wide range of distributions and users can easily access to all the distributions within the pipeline.
-
-<!-- additional benefit: -->
-
-<!--   - Multiple scales and multiple distributions can be fitted using the `c(...)` syntax to compare index values constructed from different parameterisations;  -->
-<!--   - Intermediate results can be checked after each step; and -->
-<!--   - Additional steps and analysis can be wired into the workflow for index diagnostics and customised user need. -->
-
-```{r out.width = "100%"}
-knitr::include_graphics(here::here("figures/toy-example-spei.png"))
-```
-
-```{r fig-toy-example, eval = FALSE, echo = FALSE}
-#| fig-align: center
-#| fig-width: 10
-#| fig-height: 3
-#| fig-cap: "Standardised Precipitation Evapotranspiration Index (SPEI) calculated with two distribution fits in Step 3 described above: generalised logistic and pearson III distribution, using the `wichita` data from the package `SPEI`."
-library(tidyindex)
-library(lmomco)
-library(ggplot2)
-library(tsibble)
-library(SPEI)
-library(tidyverse)
-library(lubridate)
-data(wichita)
-
-dt <- wichita %>%
-  mutate(ym = make_yearmonth(YEAR, MONTH), id = 1) %>% 
-  init(id = id, time = ym) %>%
-  var_trans(.pet_method = "thornthwaite", .vars = TMED, .lat = 37.6475) %>%
-  dim_red(diff = PRCP - pet) %>%
-  aggregate(.var = diff, .scale = 12) %>%
-  dist_fit(.dist = list(loglogistic(), pearsonIII()), .method = "lmoms", .var = .agg) %>%
-  augment(.var = .agg)
-
-dt$data %>%
-  mutate(.dist = ifelse(.dist == "parglo", "Log-logistic", "PearsonIII")) %>% 
-  ggplot(aes(x = ym, y = .index, color = .dist)) +
-  geom_line() + 
-   theme_bw() +
-  theme(
-    panel.grid = element_blank()
-  ) +
-  scale_x_yearmonth(name = "Year", date_break = "2 years", date_label = "%Y") +
-  scale_color_brewer(palette = "Dark2", name = "distribution") + 
-  ylab("SPEI")
-
-```
-
 ## Pipeline steps for constructing indexes
-
-<!-- *constructing time series index should also be encapsulated in the framework* -->
-
-*any index can be broken down into multiple steps and then we can do things with it: swap, change parameter, etc*
-
-*variables == indicators*
 
@@
 
-The construction of an index sometimes needs to consider information from neighbouring time periods.
+**Input: **
+
+**Output: **
+
+**Computation:** The construction of an index sometimes needs to consider information from neighbouring time periods.
 The temporal processing is a general operator on the time dimension of the data in the form of
@@
 
-Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing.
+**Input: **
+
+**Output: **
+
+**Computation:** Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing.
 The process can be written as a general operation in the form of
@@
 
-The purpose of variable transformation is to create variables that fits assumptions for further computing.
+**Input: **
+
+**Output: **
+
+**Computation:** The purpose of variable transformation is to create variables that fits assumptions for further computing.
 These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline.
@@
 
-While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
+**Input: **
+
+**Output: **
+
+**Computation:** While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
 The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of
@@
 
-\newpage
-
 ### Dimension reduction
 
-Dimension reduction summarises the multivariate information into univariate, which can be denoted as:
+**Input: **
+
+**Output: **
+
+**Computation:** Dimension reduction summarises the multivariate information into univariate, which can be denoted as:
 
@@
 
-<!-- With a probability model imposed,  -->
-Distribution fit can be seen as the model fitting in its simplest term. It can be represented by
+**Input: **
+
+**Output: **
+
+**Computation:** Distribution fit can be seen as the model fitting in its simplest term. It can be represented by
 
@@
 
-This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
+**Input: **
+
+**Output: **
+
+**Computation:** This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
 For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
@@
 
-Benchmarking sets a constant value to allow the constructed index to be compared across time.
+**Input: **
+
+**Output: **
+
+**Computation:** Benchmarking sets a constant value to allow the constructed index to be compared across time.
 Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed. A benchmark value could be a constant or a function of the data, i.e. mean.
@@
 
-In public communication, the index values are usually accompanied by a categorical grade.
+**Input: **
+
+**Output: **
+
+**Computation:** In public communication, the index values are usually accompanied by a categorical grade.
 The categorised grades are an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes.
@@
 
-# Incorporating alternative methods into the pipeline components {#sec-dev}
+# Software design {#sec-dev}
 

### 2023-08-01 tidyindex.tex
rearrange and organise after meeting

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{sec-toy-example}{%
-\subsection{How does the pipeline constructin of an index look
-like?}\label{sec-toy-example}}
-
-Consider a commonly used drought index: Standardized
-Precipitation-Evapotranspiration Index (SPEI) (Vicente-Serrano,
-Beguería, and López-Moreno 2010). Its construction involves:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
-\tightlist
-\item
-  transform the average temperature (\texttt{TMED}) into potential
-  evapotranspiration (\texttt{pet})
-\item
-  combine precipitation (\texttt{prcp}) and potential evapotranspiration
-  (\texttt{pet}) into a single variable \texttt{diff}
-\item
-  aggregate the difference series with a sliding window
-  (\texttt{.scale})
-\item
-  fit a distribution to the aggregated series, and
-\item
-  derive the index value from the normal density values.
-\end{enumerate}
-
-Conventionally approach may combine all these steps into a single
-function, with some level of modularity. However, these modules may only
-work for the selected index offered by the package.
-
-Under the pipeline approach, analysts first need to identify which
-module each step belongs to.
-
-Below shows the pseudocode for constructing SPEI with the pipeline:
-
-\begin{verbatim}
-DATA %>% 
-  var_trans(.method = thornthwaite, Tave = TMED, ..., .new_name = "pet") %>%
-  dim_red(diff = prcp - pet) %>%
-  aggregate(.var = diff, .scale = 12, .new_name = "agg") %>%
-  dist_fit(.method = "lmoms", .var = agg, .dist = DIST) %>%
-  augment(.var = agg)
-\end{verbatim}
-
-The pipeline construct allows for multiple \texttt{.scales} and
-\texttt{.dist} to be evaluated in \texttt{aggregate()} and
-\texttt{dist\_fit()} to compare index under different parameterisations.
-The result can then be passed into the \texttt{ggplot2} to crease
-visualisation. \textbf{?@fig-toy-example} compares the SPEI calculated
-with two distributions (log-logistic and pearson III).
-
-Apart from evaluating multiples parameters, the pipeline approach allows
-the the steps written for one index can be directly extrapolate to
-another index building within the pipeline. The flexibility of the
-pipeline also integrate well with other existing packages, for examples,
-fitting distributions using L-moment is commonly used when constructing
-drought indexes. The package \texttt{lmomco} provides general L-moment
-fits to a wide range of distributions and users can easily access to all
-the distributions within the pipeline.
-
-\includegraphics[width=1\textwidth,height=\textheight]{figures/toy-example-spei.png}
-
 \hypertarget{pipeline-steps-for-constructing-indexes}{%
@@
 indexes}\label{pipeline-steps-for-constructing-indexes}}
-
-\emph{any index can be broken down into multiple steps and then we can
-do things with it: swap, change parameter, etc}
-
-\emph{variables == indicators}
 
@@
 
-The construction of an index sometimes needs to consider information
-from neighbouring time periods. The temporal processing is a general
-operator on the time dimension of the data in the form of
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} The construction of an index sometimes needs to
+consider information from neighbouring time periods. The temporal
+processing is a general operator on the time dimension of the data in
+the form of
 
@@
 
-Spatial processing may be needed when indexes are not calculated
-independently on each collected location or when variables collected
-from multiple sources need to be fused before further processing. The
-process can be written as a general operation in the form of
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} Spatial processing may be needed when indexes are
+not calculated independently on each collected location or when
+variables collected from multiple sources need to be fused before
+further processing. The process can be written as a general operation in
+the form of
 
@@
 
-The purpose of variable transformation is to create variables that fits
-assumptions for further computing. These assumptions include a stable
-variance, normal distribution, or a certain scale required by some
-algorithms down the pipeline. Variable transformation is a general
-notion of a functional transformation on the variable:
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} The purpose of variable transformation is to
+create variables that fits assumptions for further computing. These
+assumptions include a stable variance, normal distribution, or a certain
+scale required by some algorithms down the pipeline. Variable
+transformation is a general notion of a functional transformation on the
+variable:
 
@@
 
-While scaling can be seen as a specific type of variable transformation,
-it is separated into its own step to make the step explicit in the
-pipeline. The key difference between the two steps is that variable
-transformation typically changes the shape of the data while scaling
-only changes the data scale and can usually be written in the form of
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} While scaling can be seen as a specific type of
+variable transformation, it is separated into its own step to make the
+step explicit in the pipeline. The key difference between the two steps
+is that variable transformation typically changes the shape of the data
+while scaling only changes the data scale and can usually be written in
+the form of
 
@@
 
-\newpage
-
 \hypertarget{dimension-reduction}{%
@@
 
-Dimension reduction summarises the multivariate information into
-univariate, which can be denoted as:
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} Dimension reduction summarises the multivariate
+information into univariate, which can be denoted as:
 
@@
 
-Distribution fit can be seen as the model fitting in its simplest term.
-It can be represented by
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} Distribution fit can be seen as the model fitting
+in its simplest term. It can be represented by
 
@@
 
-This step maps the univariate series into a different scale, typically
-for ease of comparison across regions. For example, a normal scale,
-{[}0, 1{]}, or {[}0, 100{]} may be favored for reporting certain
-indexes. In drought indexes, i.e.~SPI or SPEI, the quantiles from the
-fitted distribution are converted into the normal scale via the normal
-reverse CDF function: \(\Phi^{-1}(.)\). Normalising is usually used at
-the end of the pipeline and its main difference from the scaling step is
-that here the change of scale also changes the distribution of the
-variable. While being commonly used, this step can get criticism from
-analysts for forcing the data into the decided scale, which can be
-either unnecessary or inaccurately exaggerate or downplay the outliers.
-Also, the use of a normal scale needs to be interpreted with caution.
-Figure~\ref{fig-normalising} illustrates the normal density not being
-directly proportional to its probability of occurrence. This is
-concerning, especially at the extreme values, since a small difference
-in the tail density can have magnitudes of difference in its probability
-of occurrence.
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} This step maps the univariate series into a
+different scale, typically for ease of comparison across regions. For
+example, a normal scale, {[}0, 1{]}, or {[}0, 100{]} may be favored for
+reporting certain indexes. In drought indexes, i.e.~SPI or SPEI, the
+quantiles from the fitted distribution are converted into the normal
+scale via the normal reverse CDF function: \(\Phi^{-1}(.)\). Normalising
+is usually used at the end of the pipeline and its main difference from
+the scaling step is that here the change of scale also changes the
+distribution of the variable. While being commonly used, this step can
+get criticism from analysts for forcing the data into the decided scale,
+which can be either unnecessary or inaccurately exaggerate or downplay
+the outliers. Also, the use of a normal scale needs to be interpreted
+with caution. Figure~\ref{fig-normalising} illustrates the normal
+density not being directly proportional to its probability of
+occurrence. This is concerning, especially at the extreme values, since
+a small difference in the tail density can have magnitudes of difference
+in its probability of occurrence.
 
@@
 
-Benchmarking sets a constant value to allow the constructed index to be
-compared across time. Here we denote it with \(u[x(s_i, t_j)]\) where
-\(u\) is a scalar of interest in the index constructed. A benchmark
-value could be a constant or a function of the data, i.e.~mean.
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} Benchmarking sets a constant value to allow the
+constructed index to be compared across time. Here we denote it with
+\(u[x(s_i, t_j)]\) where \(u\) is a scalar of interest in the index
+constructed. A benchmark value could be a constant or a function of the
+data, i.e.~mean.
 
@@
 
-In public communication, the index values are usually accompanied by a
-categorical grade. The categorised grades are an ordered set of
-descriptive words or colors to communicate the severity or guide the
-comprehension of the indexes. The mapping from continuous index values
-to the discrete grades is called simplification in the pipeline and it
-can be written as a piece-wise function:
+\textbf{Input: }
+
+\textbf{Output: }
+
+\textbf{Computation:} In public communication, the index values are
+usually accompanied by a categorical grade. The categorised grades are
+an ordered set of descriptive words or colors to communicate the
+severity or guide the comprehension of the indexes. The mapping from
+continuous index values to the discrete grades is called simplification
+in the pipeline and it can be written as a piece-wise function:
 
@@
 \hypertarget{sec-dev}{%
-\section{Incorporating alternative methods into the pipeline
-components}\label{sec-dev}}
+\section{Software design}\label{sec-dev}}
 
@@
 example, the function \texttt{idx\_spei()} includes the five steps
-previously described in Section~\ref{sec-toy-example} (variable
+previously described in \textbf{?@sec-toy-example} (variable
 transformation, dimension reduction, temporal aggregation, distribution
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%
-Vicente-Serrano, Sergio M., Santiago Beguería, and Juan I. López-Moreno.
-2010. {``A {Multiscalar} {Drought} {Index} {Sensitive} to {Global}
-{Warming}: {The} {Standardized} {Precipitation} {Evapotranspiration}
-{Index}.''} \emph{Journal of Climate} 23 (7): 1696--1718.
-\url{https://journals.ametsoc.org/view/journals/clim/23/7/2009jcli2909.1.xml}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%

### 2023-08-03 tidyindex.qmd
first two paragraphs in Sec 2

@@
 
-*Think about if there is another word for data pipeline*
+<!-- **Why you should care about pipeline** -->
 
-**Why you should care about pipeline**
+Data pipeline refers to data transforming through a series of functions, resembling the concept of function composition in mathematics.
+To accomplish specific tasks, analysts initially write lines of code and later generalise them into functions.
+This generalization transforms the code from a specific use-case to something reuseable and easy to control with arguments.
+A pipeline represents a further generalization of functions into a workflow.
+It chains individual functions together, rather than creating a single monolithic pieces, to solve a data analysis problem.
+The resulting pipeline can be reused for other similar problems, and ideally, allows for easy switching of functions within the pipeline, similar to controlling function parameters through arguments.
 
-Data pipeline is not a new concept to computing. It refers to a set of data processing elements connected in series, where the output of one element is the input of the next one.
-@wickham_plumbing_2009 argues that whether made explicit or not, the pipeline has to be presented in every graphics program.
-The paper also argues that breaking down graphic rendering into steps is beneficial for understanding the implementation and comparing between different graphic systems. 
-The discussion on pipeline construction is well documented in early interactive graphics software: @buja_elements_1988, @sutherland_orca_2000, and @xie_reactive_2014 and their pipeline steps include non-linear transformation, variable standardization, randomization and dimension reduction.
+<!-- One of the most commonly known pipeline examples is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program, i.e.
+````command 1 | command 2 | command 3 | ...````.
+To solve a complex problem, the Unix system builds simple programs that do one thing well and work well together.
+-->
 
-**What is pipeline, its underlying software design philosophy, and how these are reflected in R**
+<!-- **What is pipeline, its underlying software design philosophy, and how these are reflected in R** -->
 
-One of the most commonly known pipeline examples is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program, i.e.
-
-````
-command 1 | command 2 | command 3 | ...
-````
-
-To solve a complex problem, the Unix system builds simple programs that do one thing well and work well together.
-This design is also reflected in the tidyverse ecosystem in R.
-To solve a complicated data problem using tidyverse, analysts typically build the solution using a collection of tools from the tidyverse toolbox. 
-The data object can flow smoothly from one command to the next, safeguarded by the tidy data format [@wickham_tidy_2014], which prescribes three rules on how to lay out tabular data.
-The tidyverse tools also embrace a strong human-centered design where function names are intuitive and easy to reference through autocomplete.
-With the tidyverse design principle in mind, the tidymodel suite enables analysts to build machine learning models through the data pipeline. 
-It includes typical tasks required in machine learning like data resampling, feature engineering, model fitting, model tuning, and model evaluation.
-An advantage of tidymodel pipeline over separate software for individual models is that analysts no longer need to write model-specific syntax to work with each model, but pipeline-specific syntax that is applicable to all the models implemented in tidymodel.
-This allows users to easily experiment with a collection of machine learning models.
+Early discussion on data pipeline in R can be found in the interactive graphics literature [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014]. 
+Specifically, @wickham_plumbing_2009 differentiates two different approaches used for structuring for (graphic) software: pipes and pipelines.
+The pipe approach offers functions, or components, and analysts concatenate functions together to form a pipeline to solve the specific problem at hands.
+This approach is used by the tidyverse suite, specifically in the dplyr package, where analysts apply functions in a sequence to address a wide array of data wrangling tasks.
+On the other hand, the pipeline approach addresses the data analysis tasks in a more structured manner, with a few centred steps used in almost all the workflow.
+For example, a supervised machine learning workflow typically involves steps like data splitting, model training and model evaluation, among others, regardless of the specific model used.
+A pipeline approach, like the tidymodel suite, modularizes these components and provides an unified syntax to specify different models.
 

### 2023-08-03 tidyindex.tex
first two paragraphs in Sec 2

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-\emph{Think about if there is another word for data pipeline}
-
-\textbf{Why you should care about pipeline}
-
-Data pipeline is not a new concept to computing. It refers to a set of
-data processing elements connected in series, where the output of one
-element is the input of the next one. Wickham et al. (2009) argues that
-whether made explicit or not, the pipeline has to be presented in every
-graphics program. The paper also argues that breaking down graphic
-rendering into steps is beneficial for understanding the implementation
-and comparing between different graphic systems. The discussion on
-pipeline construction is well documented in early interactive graphics
-software: Buja et al. (1988), Sutherland et al. (2000), and Xie,
-Hofmann, and Cheng (2014) and their pipeline steps include non-linear
-transformation, variable standardization, randomization and dimension
-reduction.
-
-\textbf{What is pipeline, its underlying software design philosophy, and
-how these are reflected in R}
-
-One of the most commonly known pipeline examples is perhaps the Unix
-pipeline where programs can be concatenated with \texttt{\textbar{}} to
-flow the output from the last program into the next program, i.e.~
-
-\begin{verbatim}
-command 1 | command 2 | command 3 | ...
-\end{verbatim}
-
-To solve a complex problem, the Unix system builds simple programs that
-do one thing well and work well together. This design is also reflected
-in the tidyverse ecosystem in R. To solve a complicated data problem
-using tidyverse, analysts typically build the solution using a
-collection of tools from the tidyverse toolbox. The data object can flow
-smoothly from one command to the next, safeguarded by the tidy data
-format (Wickham 2014), which prescribes three rules on how to lay out
-tabular data. The tidyverse tools also embrace a strong human-centered
-design where function names are intuitive and easy to reference through
-autocomplete. With the tidyverse design principle in mind, the tidymodel
-suite enables analysts to build machine learning models through the data
-pipeline. It includes typical tasks required in machine learning like
-data resampling, feature engineering, model fitting, model tuning, and
-model evaluation. An advantage of tidymodel pipeline over separate
-software for individual models is that analysts no longer need to write
-model-specific syntax to work with each model, but pipeline-specific
-syntax that is applicable to all the models implemented in tidymodel.
-This allows users to easily experiment with a collection of machine
-learning models.
+Data pipeline refers to data transforming through a series of functions,
+resembling the concept of function composition in mathematics. To
+accomplish specific tasks, analysts initially write lines of code and
+later generalise them into functions. This generalization transforms the
+code from a specific use-case to something reuseable and easy to control
+with arguments. A pipeline represents a further generalization of
+functions into a workflow. It chains individual functions together,
+rather than creating a single monolithic pieces, to solve a data
+analysis problem. The resulting pipeline can be reused for other similar
+problems, and ideally, allows for easy switching of functions within the
+pipeline, similar to controlling function parameters through arguments.
+
+Early discussion on data pipeline in R can be found in the interactive
+graphics literature (Buja et al. 1988; Sutherland et al. 2000; Xie,
+Hofmann, and Cheng 2014). Specifically, Wickham et al. (2009)
+differentiates two different approaches used for structuring for
+(graphic) software: pipes and pipelines. The pipe approach offers
+functions, or components, and analysts concatenate functions together to
+form a pipeline to solve the specific problem at hands. This approach is
+used by the tidyverse suite, specifically in the dplyr package, where
+analysts apply functions in a sequence to address a wide array of data
+wrangling tasks. On the other hand, the pipeline approach addresses the
+data analysis tasks in a more structured manner, with a few centred
+steps used in almost all the workflow. For example, a supervised machine
+learning workflow typically involves steps like data splitting, model
+training and model evaluation, among others, regardless of the specific
+model used. A pipeline approach, like the tidymodel suite, modularizes
+these components and provides an unified syntax to specify different
+models.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%
-Wickham, Hadley. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
-Software} 59 (September): 1--23.
-\url{https://doi.org/10.18637/jss.v059.i10}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%

### 2023-08-05 tidyindex.qmd
data pipeline section ready

@@
 
-Data pipeline refers to data transforming through a series of functions, resembling the concept of function composition in mathematics.
-To accomplish specific tasks, analysts initially write lines of code and later generalise them into functions.
-This generalization transforms the code from a specific use-case to something reuseable and easy to control with arguments.
+Data pipeline refers to data transforming through a series of steps, similar to the concept composition of function in mathematics.
+To accomplish specific tasks, analysts initially write code snippets and later generalise them into functions.
+This generalization transforms the code from a specific use-case to an abstraction that is reuseable and easy to control with arguments.
 A pipeline represents a further generalization of functions into a workflow.
@@
 
-**Constructing indexes would also benefit from pipeline and embracing the aforementioned design philosophy.**
+<!-- **Constructing indexes would also benefit from pipeline and embracing the aforementioned design philosophy.** -->
 
-In index construction, data pipeline is often presented in a workflow diagram in the research paper to illustrate how the raw data is transformed into the final indexes. 
-This agrees with Wickham's argument on the presence of the data pipeline, however, more often than not, the pipeline is not made explicit in the software.
-Often the time, all the steps are lumped into a single wrapper function, rather than being split into smaller, modulated functions. 
-This increases the cost of maintaining and understanding the code base, gives analysts little freedom to customise the indexes for specific needs, and hinders reusing existing code for building new indexes.
-A pipeline approach unites a range of indexes under a single data pipeline and analysts can compose indexes from pipeline steps like building Legos from individual bricks.
-In this workflow, analysts are not limited by indexes that have been already proposed and can easily combine pipeline steps to compose novel indexes.
-Analysis of the indexes (i.e.
-calculation of uncertainty) is also feasible by adding external code into the pipeline.
+<!-- The workflow to construct indexes are more similar to the one for machine learning models with several components commonly used across most of the indexes.
+Indexes typically take inputs from multivariate data and combine them into a univariate series -- this is a dimension reduction step.
+Indexes are often series that traced over time, hence a temporal component.
+-->
+
+For constructing indexes, many indexes have provided available publicly scripts for computation, i.e.
+Human Development Index (HDI) [@hdi].
+Some have made the first level of generalisation from scripts to functions and creates R packages specialised for calculating particular indexes, i.e.
+the `SPEI` package [@SPEI] for calculating drought indexes SPI and SPEI.
+Our pipeline workflow takes this generalization a step further, enabling easier and reproducible computation of an array of indexes in different fields.
+Instead of providing individual functions for each index with common modules shared within the code base, our approach provides explicit, user-facing modules for constructing pipeline workflow.
+Analysts can create all the indexes they wish by concatenating these modules in the same syntax, regardless of their origins.
+
+<!-- This approach allows for quick experimentation with alternatives and facilitates building new indexes.
+Consequently, maintenance is streamlined in the long term, as all the indexes are developed in-house using a unified syntax.
+-->
 

### 2023-08-05 tidyindex.tex
data pipeline section ready

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, breakable, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-Data pipeline refers to data transforming through a series of functions,
-resembling the concept of function composition in mathematics. To
-accomplish specific tasks, analysts initially write lines of code and
+Data pipeline refers to data transforming through a series of steps,
+similar to the concept composition of function in mathematics. To
+accomplish specific tasks, analysts initially write code snippets and
 later generalise them into functions. This generalization transforms the
-code from a specific use-case to something reuseable and easy to control
-with arguments. A pipeline represents a further generalization of
-functions into a workflow. It chains individual functions together,
-rather than creating a single monolithic pieces, to solve a data
-analysis problem. The resulting pipeline can be reused for other similar
-problems, and ideally, allows for easy switching of functions within the
-pipeline, similar to controlling function parameters through arguments.
+code from a specific use-case to an abstraction that is reuseable and
+easy to control with arguments. A pipeline represents a further
+generalization of functions into a workflow. It chains individual
+functions together, rather than creating a single monolithic pieces, to
+solve a data analysis problem. The resulting pipeline can be reused for
+other similar problems, and ideally, allows for easy switching of
+functions within the pipeline, similar to controlling function
+parameters through arguments.
 
@@
 
-\textbf{Constructing indexes would also benefit from pipeline and
-embracing the aforementioned design philosophy.}
-
-In index construction, data pipeline is often presented in a workflow
-diagram in the research paper to illustrate how the raw data is
-transformed into the final indexes. This agrees with Wickham's argument
-on the presence of the data pipeline, however, more often than not, the
-pipeline is not made explicit in the software. Often the time, all the
-steps are lumped into a single wrapper function, rather than being split
-into smaller, modulated functions. This increases the cost of
-maintaining and understanding the code base, gives analysts little
-freedom to customise the indexes for specific needs, and hinders reusing
-existing code for building new indexes. A pipeline approach unites a
-range of indexes under a single data pipeline and analysts can compose
-indexes from pipeline steps like building Legos from individual bricks.
-In this workflow, analysts are not limited by indexes that have been
-already proposed and can easily combine pipeline steps to compose novel
-indexes. Analysis of the indexes (i.e.~calculation of uncertainty) is
-also feasible by adding external code into the pipeline.
+For constructing indexes, many indexes have provided available publicly
+scripts for computation, i.e.~Human Development Index (HDI) (United
+Nations Development Programme 2022). Some have made the first level of
+generalisation from scripts to functions and creates R packages
+specialised for calculating particular indexes, i.e.~the \texttt{SPEI}
+package (Beguería and Vicente-Serrano 2017) for calculating drought
+indexes SPI and SPEI. Our pipeline workflow takes this generalization a
+step further, enabling easier and reproducible computation of an array
+of indexes in different fields. Instead of providing individual
+functions for each index with common modules shared within the code
+base, our approach provides explicit, user-facing modules for
+constructing pipeline workflow. Analysts can create all the indexes they
+wish by concatenating these modules in the same syntax, regardless of
+their origins.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-SPEI}{}}%
+Beguería, Santiago, and Sergio M. Vicente-Serrano. 2017. \emph{SPEI:
+Calculation of the Standardised Precipitation-Evapotranspiration Index}.
+\url{https://CRAN.R-project.org/package=SPEI}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-buja_elements_1988}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
+United Nations Development Programme. 2022. {``Human Development Report
+2021-22.''} New York.
+\url{https://hdr.undp.org/data-center/human-development-index\#/indicies/HDI}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%

### 2023-08-05 tidyindex.qmd
a paragraph on why you can decompose the procedure into modules

@@
 
-<!-- The workflow to construct indexes are more similar to the one for machine learning models with several components commonly used across most of the indexes.
-Indexes typically take inputs from multivariate data and combine them into a univariate series -- this is a dimension reduction step.
-Indexes are often series that traced over time, hence a temporal component.
--->
-
 For constructing indexes, many indexes have provided available publicly scripts for computation, i.e.
@@
 # A pipeline for building statistical indexes {#sec-index-pipeline}
+
+In constructing various indexes, the primary aim is to transform the data, often multivariate, into a univariate index.
+Spatial and temporal considerations are also factored into the process when observational units and time periods are not independent.
+However, despite the variations in contextual information for indexes in different fields, the underlying statistical methodology remains consistent across diverse domains.
+Each index can be represented as a series of modular statistical operations on the data.
+This allows us to decompose the index construction process into a unified pipeline workflow with a standardized set of data processing steps to be applied across different indexes.
 

### 2023-08-05 tidyindex.tex
a paragraph on why you can decompose the procedure into modules

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, breakable, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 indexes}\label{sec-index-pipeline}}
+
+In constructing various indexes, the primary aim is to transform the
+data, often multivariate, into a univariate index. Spatial and temporal
+considerations are also factored into the process when observational
+units and time periods are not independent. However, despite the
+variations in contextual information for indexes in different fields,
+the underlying statistical methodology remains consistent across diverse
+domains. Each index can be represented as a series of modular
+statistical operations on the data. This allows us to decompose the
+index construction process into a unified pipeline workflow with a
+standardized set of data processing steps to be applied across different
+indexes.
 

### 2023-08-06 tidyindex.qmd
a software design paragraph and some input/output for temporal processing

@@
 
-## Pipeline steps for constructing indexes
+## Pipeline steps for constructing indexes {#sec-pipeline-steps}
 
@@
 
-**Input: **
+**Input: ** multivariate data with temporal coordinates
 
-**Output: **
+**Output: ** multivariate data with aggregated temporal axis
 
@@
 \begin{equation}
-f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
+x(\mathbf{s};\mathbf{t}^\prime)  f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
 \end{equation}
@@
 where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. 
-A typical example of temporal processing is aggregation, which is used in the drought index SPI to measure the lack of precipitation for meteorological drought.
-In SPI, monthly precipitation is aggregated by a time scale parameter $k$: $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),$ where $j^\prime$ is the new time index after the aggregation.
-In this notation, each spatial location is separately aggregated and precipitation is summed from $k$ month back, $j^\prime - k + 1$, to the current period, $j^\prime$, to create the aggregated series, indexed by $j^\prime$.
+A typical example of temporal processing is aggregation and in the drought community, these indexes are often called multi-scalar indexes.
+For example,  Standardized Precipitation Index (SPI) aggregates monthly precipitation using a time scale parameter $k$ in the form of
 
-*more explicit on k will influence 1) long term vs. short term, 2) uncertainty*
+$$x(s_i;t_{j^\prime}) = \sum_{j = j^\prime}^{j^\prime + k - 1}x(s_i; t_j),$$
 
-The choice of time scales parameter $k$ can result in variation in the calculated index values: a small $k$ of 3 or 6 months produces the index more sensitive to individual months, while a large $k$ of 24 or 36, an equivalent to a 2- or 3-year aggregation, gives dryness information relative to the long term condition.
-As will be shown in section [SECTION EXAMPLE], this variation may even lead to conflicting conclusions on the dry/wet condition of the area, highlighting the importance to account for index uncertainty when interpreting index values for decision-making.
-
-*Effective drought index*
+where $j^\prime$ is the new time index after the aggregation.
+Multi-scalar drought indexes are widely used to study different types of drought events.
+A smaller value of the time scales parameter $k$ is used for short term drought events, while a large $k$ of 18 or 24, an equivalent to a 1.5- or 2-year aggregation, is often used for long term assessment.
 
@@
 
-**Input: **
+**Input: ** Multivariate data with coordinates
 
-**Output: **
+**Output: ** Multivariate data with a new set of coordinates
 
@@
 # Software design {#sec-dev}
+
+Tidyindex, as suggested by its name, follows the style and design principles of the tidyverse suite.
+The package is designed to simplify index construction for human analysts, ensure consistency across functions, and enable composability across modules to construct indexes.
+With the package, analysts can pass a data object through a set of pipeline modules introduced in @sec-pipeline-steps, offering different operations as standalone functions.
+For example, the dimension reduction module includes linear weighting, geometrical weighting, and manual expressions as possible methods to combine multivariate data.
+These methods are initially evaluated as a recipe of the operation without the data, and then, at the pipeline level, evaluating with the data using tidy evaluation, with the package `rlang` heavily used throughout the package.
 

### 2023-08-06 tidyindex.tex
a software design paragraph and some input/output for temporal processing

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, breakable, boxrule=0pt, frame hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{pipeline-steps-for-constructing-indexes}{%
+\hypertarget{sec-pipeline-steps}{%
 \subsection{Pipeline steps for constructing
-indexes}\label{pipeline-steps-for-constructing-indexes}}
+indexes}\label{sec-pipeline-steps}}
 
@@
 
-\textbf{Input: }
-
-\textbf{Output: }
+\textbf{Input: } multivariate data with temporal coordinates
+
+\textbf{Output: } multivariate data with aggregated temporal axis
 
@@
 the number of parameter of \(\psi\). A typical example of temporal
-processing is aggregation, which is used in the drought index SPI to
-measure the lack of precipitation for meteorological drought. In SPI,
-monthly precipitation is aggregated by a time scale parameter \(k\):
-\(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j),\)
-where \(j^\prime\) is the new time index after the aggregation. In this
-notation, each spatial location is separately aggregated and
-precipitation is summed from \(k\) month back, \(j^\prime - k + 1\), to
-the current period, \(j^\prime\), to create the aggregated series,
-indexed by \(j^\prime\).
-
-\emph{more explicit on k will influence 1) long term vs.~short term, 2)
-uncertainty}
-
-The choice of time scales parameter \(k\) can result in variation in the
-calculated index values: a small \(k\) of 3 or 6 months produces the
-index more sensitive to individual months, while a large \(k\) of 24 or
-36, an equivalent to a 2- or 3-year aggregation, gives dryness
-information relative to the long term condition. As will be shown in
-section {[}SECTION EXAMPLE{]}, this variation may even lead to
-conflicting conclusions on the dry/wet condition of the area,
-highlighting the importance to account for index uncertainty when
-interpreting index values for decision-making.
-
-\emph{Effective drought index}
+processing is aggregation and in the drought community, these indexes
+are often called multi-scalar indexes. For example, Standardized
+Precipitation Index (SPI) aggregates monthly precipitation using a time
+scale parameter \(k\) in the form of
+
+\[x(s_i;t_{j^\prime}) = \sum_{j = j^\prime}^{j^\prime + k - 1}x(s_i; t_j),\]
+
+where \(j^\prime\) is the new time index after the aggregation.
+Multi-scalar drought indexes are widely used to study different types of
+drought events. A smaller value of the time scales parameter \(k\) is
+used for short term drought events, while a large \(k\) of 18 or 24, an
+equivalent to a 1.5- or 2-year aggregation, is often used for long term
+assessment.
 
@@
 
-\textbf{Input: }
-
-\textbf{Output: }
+\textbf{Input: } Multivariate data with coordinates
+
+\textbf{Output: } Multivariate data with a new set of coordinates
 
@@
 \section{Software design}\label{sec-dev}}
+
+Tidyindex, as suggested by its name, follows the style and design
+principles of the tidyverse suite. The package is designed to simplify
+index construction for human analysts, ensure consistency across
+functions, and enable composability across modules to construct indexes.
+With the package, analysts can pass a data object through a set of
+pipeline modules introduced in Section~\ref{sec-pipeline-steps},
+offering different operations as standalone functions. For example, the
+dimension reduction module includes linear weighting, geometrical
+weighting, and manual expressions as possible methods to combine
+multivariate data. These methods are initially evaluated as a recipe of
+the operation without the data, and then, at the pipeline level,
+evaluating with the data using tidy evaluation, with the package
+\texttt{rlang} heavily used throughout the package.
 

### 2023-08-08 tidyindex.qmd
refine Section 2 to Tidy framework and bridge to pipeline

@@
 
-# Data pipeline {#sec-pipeline}
+# Tidy framework {#sec-pipeline}
 
@@
 
-Data pipeline refers to data transforming through a series of steps, similar to the concept composition of function in mathematics.
-To accomplish specific tasks, analysts initially write code snippets and later generalise them into functions.
-This generalization transforms the code from a specific use-case to an abstraction that is reuseable and easy to control with arguments.
-A pipeline represents a further generalization of functions into a workflow.
-It chains individual functions together, rather than creating a single monolithic pieces, to solve a data analysis problem.
-The resulting pipeline can be reused for other similar problems, and ideally, allows for easy switching of functions within the pipeline, similar to controlling function parameters through arguments.
-
-<!-- One of the most commonly known pipeline examples is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program, i.e.
-````command 1 | command 2 | command 3 | ...````.
-To solve a complex problem, the Unix system builds simple programs that do one thing well and work well together.
--->
+The tidy framework consists of two key components: tidy data and tidy tools.
+The concept of tidy data [@wickham_tidy_2014] prescribes specific rules for organising data in an analysis, with observations as rows, variables as columns, and types of observational units as tables.
+Tidy tools, on the other hand, are concatenated to establish a sequential process through which the tidy data object flows, creating a pipeline for data processing.
+In this workflow, all the functions take inputs and outputs as tidy data, enabling direct concatenation of functions and saving analysts from bridging the outputs to subsequent operations.
+Additionally, the pipeline workflow breaks down complex problems into smaller, more manageable pieces.
+In an interactive data analysis, this flexibility to link various operations together as needed makes the pipeline workflow preferable over a monolithic design, where all the steps are predetermined and tightly integrated into a single piece.
 
@@
 
-Early discussion on data pipeline in R can be found in the interactive graphics literature [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014]. 
-Specifically, @wickham_plumbing_2009 differentiates two different approaches used for structuring for (graphic) software: pipes and pipelines.
-The pipe approach offers functions, or components, and analysts concatenate functions together to form a pipeline to solve the specific problem at hands.
-This approach is used by the tidyverse suite, specifically in the dplyr package, where analysts apply functions in a sequence to address a wide array of data wrangling tasks.
-On the other hand, the pipeline approach addresses the data analysis tasks in a more structured manner, with a few centred steps used in almost all the workflow.
-For example, a supervised machine learning workflow typically involves steps like data splitting, model training and model evaluation, among others, regardless of the specific model used.
-A pipeline approach, like the tidymodel suite, modularizes these components and provides an unified syntax to specify different models.
+Dating back to earlier times, the pipeline approach to structure data analysis draws inspiration from the interactive graphics literature [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009].
+More recently, the tidy framework has provided a conceptual guide for the tidyverse suite [@wickham_welcome_2019] for general-purpose data wrangling, a concept that has gained popularity within the R community.
+Subsequently, more packages have since been developed to analyze data using the tidy framework for specific domains, for example, `tidytext` for text data [@silge_tidytext_2016], and `naniar` for missing data [@tierney_expanding_2023], and `tidymodels` for building machine learning models [@tidymodels].
+To create a tidy workflow tailored to a specific domain, developers first need to identify the fundamental building blocks in the typical workflow.
+These components are then implemented as modules, which can be combined to form the pipeline.
+For example, in supervised machine learning models, steps such as data splitting, model training and model evaluation are commonly used in most workflow.
+In the `tidymodels`, these steps are correspondingly implemented as package `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen.
 
@@
 
-# A pipeline for building statistical indexes {#sec-index-pipeline}
+# Details of the index pipline {#sec-index-pipeline}
 
@@
 This allows us to decompose the index construction process into a unified pipeline workflow with a standardized set of data processing steps to be applied across different indexes.
-
-## Pipeline steps for constructing indexes {#sec-pipeline-steps}
 
@@
 # Software design {#sec-dev}
-
-Tidyindex, as suggested by its name, follows the style and design principles of the tidyverse suite.
-The package is designed to simplify index construction for human analysts, ensure consistency across functions, and enable composability across modules to construct indexes.
-With the package, analysts can pass a data object through a set of pipeline modules introduced in @sec-pipeline-steps, offering different operations as standalone functions.
-For example, the dimension reduction module includes linear weighting, geometrical weighting, and manual expressions as possible methods to combine multivariate data.
-These methods are initially evaluated as a recipe of the operation without the data, and then, at the pipeline level, evaluating with the data using tidy evaluation, with the package `rlang` heavily used throughout the package.
 

### 2023-08-08 tidyindex.tex
refine Section 2 to Tidy framework and bridge to pipeline

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, breakable, boxrule=0pt, frame hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 \hypertarget{sec-pipeline}{%
-\section{Data pipeline}\label{sec-pipeline}}
-
-Data pipeline refers to data transforming through a series of steps,
-similar to the concept composition of function in mathematics. To
-accomplish specific tasks, analysts initially write code snippets and
-later generalise them into functions. This generalization transforms the
-code from a specific use-case to an abstraction that is reuseable and
-easy to control with arguments. A pipeline represents a further
-generalization of functions into a workflow. It chains individual
-functions together, rather than creating a single monolithic pieces, to
-solve a data analysis problem. The resulting pipeline can be reused for
-other similar problems, and ideally, allows for easy switching of
-functions within the pipeline, similar to controlling function
-parameters through arguments.
-
-Early discussion on data pipeline in R can be found in the interactive
-graphics literature (Buja et al. 1988; Sutherland et al. 2000; Xie,
-Hofmann, and Cheng 2014). Specifically, Wickham et al. (2009)
-differentiates two different approaches used for structuring for
-(graphic) software: pipes and pipelines. The pipe approach offers
-functions, or components, and analysts concatenate functions together to
-form a pipeline to solve the specific problem at hands. This approach is
-used by the tidyverse suite, specifically in the dplyr package, where
-analysts apply functions in a sequence to address a wide array of data
-wrangling tasks. On the other hand, the pipeline approach addresses the
-data analysis tasks in a more structured manner, with a few centred
-steps used in almost all the workflow. For example, a supervised machine
-learning workflow typically involves steps like data splitting, model
-training and model evaluation, among others, regardless of the specific
-model used. A pipeline approach, like the tidymodel suite, modularizes
-these components and provides an unified syntax to specify different
-models.
+\section{Tidy framework}\label{sec-pipeline}}
+
+The tidy framework consists of two key components: tidy data and tidy
+tools. The concept of tidy data (Wickham 2014) prescribes specific rules
+for organising data in an analysis, with observations as rows, variables
+as columns, and types of observational units as tables. Tidy tools, on
+the other hand, are concatenated to establish a sequential process
+through which the tidy data object flows, creating a pipeline for data
+processing. In this workflow, all the functions take inputs and outputs
+as tidy data, enabling direct concatenation of functions and saving
+analysts from bridging the outputs to subsequent operations.
+Additionally, the pipeline workflow breaks down complex problems into
+smaller, more manageable pieces. In an interactive data analysis, this
+flexibility to link various operations together as needed makes the
+pipeline workflow preferable over a monolithic design, where all the
+steps are predetermined and tightly integrated into a single piece.
+
+Dating back to earlier times, the pipeline approach to structure data
+analysis draws inspiration from the interactive graphics literature
+(Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng 2014;
+Wickham et al. 2009). More recently, the tidy framework has provided a
+conceptual guide for the tidyverse suite (Wickham et al. 2019) for
+general-purpose data wrangling, a concept that has gained popularity
+within the R community. Subsequently, more packages have since been
+developed to analyze data using the tidy framework for specific domains,
+for example, \texttt{tidytext} for text data (Silge and Robinson 2016),
+and \texttt{naniar} for missing data (Tierney and Cook 2023), and
+\texttt{tidymodels} for building machine learning models (Kuhn and
+Wickham 2020). To create a tidy workflow tailored to a specific domain,
+developers first need to identify the fundamental building blocks in the
+typical workflow. These components are then implemented as modules,
+which can be combined to form the pipeline. For example, in supervised
+machine learning models, steps such as data splitting, model training
+and model evaluation are commonly used in most workflow. In the
+\texttt{tidymodels}, these steps are correspondingly implemented as
+package \texttt{rsample}, \texttt{parsnip}, and \texttt{yardstick},
+agnostic to the specific model chosen.
 
@@
 \hypertarget{sec-index-pipeline}{%
-\section{A pipeline for building statistical
-indexes}\label{sec-index-pipeline}}
+\section{Details of the index pipline}\label{sec-index-pipeline}}
 
@@
 indexes.
-
-\hypertarget{sec-pipeline-steps}{%
-\subsection{Pipeline steps for constructing
-indexes}\label{sec-pipeline-steps}}
 
@@
 \section{Software design}\label{sec-dev}}
-
-Tidyindex, as suggested by its name, follows the style and design
-principles of the tidyverse suite. The package is designed to simplify
-index construction for human analysts, ensure consistency across
-functions, and enable composability across modules to construct indexes.
-With the package, analysts can pass a data object through a set of
-pipeline modules introduced in Section~\ref{sec-pipeline-steps},
-offering different operations as standalone functions. For example, the
-dimension reduction module includes linear weighting, geometrical
-weighting, and manual expressions as possible methods to combine
-multivariate data. These methods are initially evaluated as a recipe of
-the operation without the data, and then, at the pipeline level,
-evaluating with the data using tidy evaluation, with the package
-\texttt{rlang} heavily used throughout the package.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-tidymodels}{}}%
+Kuhn, Max, and Hadley Wickham. 2020. \emph{Tidymodels: A Collection of
+Packages for Modeling and Machine Learning Using Tidyverse Principles.}
+\url{https://www.tidymodels.org}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-oecd_handbook_2008}{}}%
@@
 \url{https://doi.org/10.1787/9789264043466-en}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-silge_tidytext_2016}{}}%
+Silge, Julia, and David Robinson. 2016. {``Tidytext: {Text} {Mining} and
+{Analysis} {Using} {Tidy} {Data} {Principles} in {R}.''} \emph{The
+Journal of Open Source Software} 1 (3): 37.
+\url{https://doi.org/10.21105/joss.00037}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-tierney_expanding_2023}{}}%
+Tierney, Nicholas, and Dianne Cook. 2023. {``Expanding {Tidy} {Data}
+{Principles} to {Facilitate} {Missing} {Data} {Exploration},
+{Visualization} and {Assessment} of {Imputations}.''} \emph{Journal of
+Statistical Software} 105 (February): 1--31.
+\url{https://doi.org/10.18637/jss.v105.i07}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
@@
 \url{https://hdr.undp.org/data-center/human-development-index\#/indicies/HDI}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%
+Wickham, Hadley. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
+Software} 59 (September): 1--23.
+\url{https://doi.org/10.18637/jss.v059.i10}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_welcome_2019}{}}%
+Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
+D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
+{``Welcome to the {Tidyverse}.''} \emph{Journal of Open Source Software}
+4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
 

### 2023-08-08 tidyindex.tex
small tweaks on pipeline step details

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, enhanced, breakable, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \hypertarget{temporal-processing}{%
-\subsubsection{Temporal processing}\label{temporal-processing}}
-
-\textbf{Input: } multivariate data with temporal coordinates
-
-\textbf{Output: } multivariate data with aggregated temporal axis
+\subsection{Temporal processing}\label{temporal-processing}}
+
+\textbf{Input: } \(x(\mathbf{s};\mathbf{t})\): multivariate data with
+temporal coordinates
+
+\textbf{Output: } \(x(\mathbf{s};\mathbf{t}^\prime)\): multivariate data
+with aggregated temporal axis
 
@@
 \begin{equation}
-x(\mathbf{s};\mathbf{t}^\prime)  f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
+x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
 \end{equation}
@@
 \hypertarget{spatial-processing}{%
-\subsubsection{Spatial processing}\label{spatial-processing}}
+\subsection{Spatial processing}\label{spatial-processing}}
 
@@
 \hypertarget{variable-transformation}{%
-\subsubsection{Variable transformation}\label{variable-transformation}}
+\subsection{Variable transformation}\label{variable-transformation}}
 
@@
 \hypertarget{scaling}{%
-\subsubsection{Scaling}\label{scaling}}
+\subsection{Scaling}\label{scaling}}
 
@@
 \hypertarget{dimension-reduction}{%
-\subsubsection{Dimension reduction}\label{dimension-reduction}}
+\subsection{Dimension reduction}\label{dimension-reduction}}
 
@@
 \hypertarget{distribution-fit}{%
-\subsubsection{Distribution fit}\label{distribution-fit}}
+\subsection{Distribution fit}\label{distribution-fit}}
 
@@
 \hypertarget{normalising}{%
-\subsubsection{Normalising}\label{normalising}}
+\subsection{Normalising}\label{normalising}}
 
@@
 \hypertarget{benchmarking}{%
-\subsubsection{Benchmarking}\label{benchmarking}}
+\subsection{Benchmarking}\label{benchmarking}}
 
@@
 \hypertarget{simplification}{%
-\subsubsection{Simplification}\label{simplification}}
+\subsection{Simplification}\label{simplification}}
 

### 2023-08-08 tidyindex.qmd
small tweaks on pipeline step details
revise section 2 to include flexibility and readability

@@
 The concept of tidy data [@wickham_tidy_2014] prescribes specific rules for organising data in an analysis, with observations as rows, variables as columns, and types of observational units as tables.
-Tidy tools, on the other hand, are concatenated to establish a sequential process through which the tidy data object flows, creating a pipeline for data processing.
-In this workflow, all the functions take inputs and outputs as tidy data, enabling direct concatenation of functions and saving analysts from bridging the outputs to subsequent operations.
-Additionally, the pipeline workflow breaks down complex problems into smaller, more manageable pieces.
-In an interactive data analysis, this flexibility to link various operations together as needed makes the pipeline workflow preferable over a monolithic design, where all the steps are predetermined and tightly integrated into a single piece.
+Tidy tools, on the other hand, are concatenated in a sequence through which the tidy object flows, creating a pipeline for data processing.
+These pipelines are data centric, meaning all the functions take a tidy data object as inputs and return a processed tidy data object, directly ready for the next operations without further reshaping needed.
+Also, the pipeline approach corresponds to the modular programming practice, which break down complex problems into smaller and more manageable pieces, as oppose to a monolithic design, where all the steps are predetermined and integrated into a single piece.
 
@@
 
-Dating back to earlier times, the pipeline approach to structure data analysis draws inspiration from the interactive graphics literature [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009].
-More recently, the tidy framework has provided a conceptual guide for the tidyverse suite [@wickham_welcome_2019] for general-purpose data wrangling, a concept that has gained popularity within the R community.
-Subsequently, more packages have since been developed to analyze data using the tidy framework for specific domains, for example, `tidytext` for text data [@silge_tidytext_2016], and `naniar` for missing data [@tierney_expanding_2023], and `tidymodels` for building machine learning models [@tidymodels].
+Examples of using a pipeline approach to structure data analysis in R can be traced back to the interactive graphics literature in the earlier time [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009].
+@wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems.
+While this comment is made in the context of graphics program, it is also applicable to other general or domain specific data analysis workflow.
+More recently, the tidyverse suite [@wickham_welcome_2019] takes the pipeline approach for general-purpose data wrangling and has gained popularity within the R community.
+The pipeline-style code can be directly read as a series of operations applied successively on the tidy data object, offering a method to document the data wrangling process.
+Unlike textual documentation, which may omit seemingly obvious or less significant information, the pipeline approach captures all the computational details while retaining the conciseness of a coding language, making it easy for human analysts to comprehend.
+
+Since the success of tidyverse, more packages have been developed to analyze data using the tidy framework for domains specific applications, a noticeable example of which is `tidymodels` for building machine learning models [@tidymodels].
 To create a tidy workflow tailored to a specific domain, developers first need to identify the fundamental building blocks in the typical workflow.
@@
 
-For constructing indexes, many indexes have provided available publicly scripts for computation, i.e.
+<!-- For constructing indexes, many indexes have provided available publicly scripts for computation, i.e.
 Human Development Index (HDI) [@hdi].
@@
 the `SPEI` package [@SPEI] for calculating drought indexes SPI and SPEI.
-Our pipeline workflow takes this generalization a step further, enabling easier and reproducible computation of an array of indexes in different fields.
-Instead of providing individual functions for each index with common modules shared within the code base, our approach provides explicit, user-facing modules for constructing pipeline workflow.
-Analysts can create all the indexes they wish by concatenating these modules in the same syntax, regardless of their origins.
+-->
+
+For constructing indexes, rather than implementing each index separately, as has been done for some indexes, the pipeline approach introduces explicit, user-facing modules that analysts can compose into indexes.
+This approach provides several advantages to analysts and researchers working on indexes.
+The index construction process can be complicated to describe and readable pipeline code complements conventional diagrams and textual description for understand the computational details.
+The uniform interface allows analysts to assemble a diverse array of indexes with a shared set of components, reducing the learning curve for adopting new implementation for different indexes.
+Importantly, each pipeline module and its corresponding steps are implemented as standalone functions, providing great flexibility for analysis the index.
+Analysts can rearrange the order in the pipeline or insert new step to experiment with possible new definition for the index or to understand index behaviour.
 
@@
 
-### Temporal processing
+## Temporal processing
 
-**Input: ** multivariate data with temporal coordinates
+**Input: ** $x(\mathbf{s};\mathbf{t})$: multivariate data with temporal coordinates
 
-**Output: ** multivariate data with aggregated temporal axis
+**Output: ** $x(\mathbf{s};\mathbf{t}^\prime)$: multivariate data with aggregated temporal axis
 
@@
 
-### Spatial processing
+## Spatial processing
 
@@
 
-### Variable transformation
+## Variable transformation
 
@@
 
-### Scaling
+## Scaling
 
@@
 
-### Dimension reduction
+## Dimension reduction
 
@@
 
-### Distribution fit
+## Distribution fit
 
@@
 
-### Normalising
+## Normalising
 
@@
 
-### Benchmarking
+## Benchmarking
 
@@
 
-### Simplification
+## Simplification
 

### 2023-08-08 tidyindex.tex
revise section 2 to include flexibility and readability

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, enhanced, breakable, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 as columns, and types of observational units as tables. Tidy tools, on
-the other hand, are concatenated to establish a sequential process
-through which the tidy data object flows, creating a pipeline for data
-processing. In this workflow, all the functions take inputs and outputs
-as tidy data, enabling direct concatenation of functions and saving
-analysts from bridging the outputs to subsequent operations.
-Additionally, the pipeline workflow breaks down complex problems into
-smaller, more manageable pieces. In an interactive data analysis, this
-flexibility to link various operations together as needed makes the
-pipeline workflow preferable over a monolithic design, where all the
-steps are predetermined and tightly integrated into a single piece.
-
-Dating back to earlier times, the pipeline approach to structure data
-analysis draws inspiration from the interactive graphics literature
-(Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng 2014;
-Wickham et al. 2009). More recently, the tidy framework has provided a
-conceptual guide for the tidyverse suite (Wickham et al. 2019) for
-general-purpose data wrangling, a concept that has gained popularity
-within the R community. Subsequently, more packages have since been
-developed to analyze data using the tidy framework for specific domains,
-for example, \texttt{tidytext} for text data (Silge and Robinson 2016),
-and \texttt{naniar} for missing data (Tierney and Cook 2023), and
-\texttt{tidymodels} for building machine learning models (Kuhn and
-Wickham 2020). To create a tidy workflow tailored to a specific domain,
-developers first need to identify the fundamental building blocks in the
-typical workflow. These components are then implemented as modules,
-which can be combined to form the pipeline. For example, in supervised
-machine learning models, steps such as data splitting, model training
-and model evaluation are commonly used in most workflow. In the
-\texttt{tidymodels}, these steps are correspondingly implemented as
-package \texttt{rsample}, \texttt{parsnip}, and \texttt{yardstick},
-agnostic to the specific model chosen.
-
-For constructing indexes, many indexes have provided available publicly
-scripts for computation, i.e.~Human Development Index (HDI) (United
-Nations Development Programme 2022). Some have made the first level of
-generalisation from scripts to functions and creates R packages
-specialised for calculating particular indexes, i.e.~the \texttt{SPEI}
-package (Beguería and Vicente-Serrano 2017) for calculating drought
-indexes SPI and SPEI. Our pipeline workflow takes this generalization a
-step further, enabling easier and reproducible computation of an array
-of indexes in different fields. Instead of providing individual
-functions for each index with common modules shared within the code
-base, our approach provides explicit, user-facing modules for
-constructing pipeline workflow. Analysts can create all the indexes they
-wish by concatenating these modules in the same syntax, regardless of
-their origins.
+the other hand, are concatenated in a sequence through which the tidy
+object flows, creating a pipeline for data processing. These pipelines
+are data centric, meaning all the functions take a tidy data object as
+inputs and return a processed tidy data object, directly ready for the
+next operations without further reshaping needed. Also, the pipeline
+approach corresponds to the modular programming practice, which break
+down complex problems into smaller and more manageable pieces, as oppose
+to a monolithic design, where all the steps are predetermined and
+integrated into a single piece.
+
+Examples of using a pipeline approach to structure data analysis in R
+can be traced back to the interactive graphics literature in the earlier
+time (Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng
+2014; Wickham et al. 2009). Wickham et al. (2009) argues that whether
+made explicit or not, pipeline has to be presented in every graphics
+program and making them explicit is beneficial for understanding the
+implementation and comparing between different graphic systems. While
+this comment is made in the context of graphics program, it is also
+applicable to other general or domain specific data analysis workflow.
+More recently, the tidyverse suite (Wickham et al. 2019) takes the
+pipeline approach for general-purpose data wrangling and has gained
+popularity within the R community. The pipeline-style code can be
+directly read as a series of operations applied successively on the tidy
+data object, offering a method to document the data wrangling process.
+Unlike textual documentation, which may omit seemingly obvious or less
+significant information, the pipeline approach captures all the
+computational details while retaining the conciseness of a coding
+language, making it easy for human analysts to comprehend.
+
+Since the success of tidyverse, more packages have been developed to
+analyze data using the tidy framework for domains specific applications,
+a noticeable example of which is \texttt{tidymodels} for building
+machine learning models (Kuhn and Wickham 2020). To create a tidy
+workflow tailored to a specific domain, developers first need to
+identify the fundamental building blocks in the typical workflow. These
+components are then implemented as modules, which can be combined to
+form the pipeline. For example, in supervised machine learning models,
+steps such as data splitting, model training and model evaluation are
+commonly used in most workflow. In the \texttt{tidymodels}, these steps
+are correspondingly implemented as package \texttt{rsample},
+\texttt{parsnip}, and \texttt{yardstick}, agnostic to the specific model
+chosen.
+
+For constructing indexes, rather than implementing each index
+separately, as has been done for some indexes, the pipeline approach
+introduces explicit, user-facing modules that analysts can compose into
+indexes. This approach provides several advantages to analysts and
+researchers working on indexes. The index construction process can be
+complicated to describe and readable pipeline code complements
+conventional diagrams and textual description for understand the
+computational details. The uniform interface allows analysts to assemble
+a diverse array of indexes with a shared set of components, reducing the
+learning curve for adopting new implementation for different indexes.
+Importantly, each pipeline module and its corresponding steps are
+implemented as standalone functions, providing great flexibility for
+anlaysis the index. Analysts can rearrange the order in the pipeline or
+insert new step to experiment with possible new definition for the index
+or to understand index behaviour.
 
@@
 \url{https://doi.org/10.1787/9789264043466-en}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-silge_tidytext_2016}{}}%
-Silge, Julia, and David Robinson. 2016. {``Tidytext: {Text} {Mining} and
-{Analysis} {Using} {Tidy} {Data} {Principles} in {R}.''} \emph{The
-Journal of Open Source Software} 1 (3): 37.
-\url{https://doi.org/10.21105/joss.00037}.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tierney_expanding_2023}{}}%
-Tierney, Nicholas, and Dianne Cook. 2023. {``Expanding {Tidy} {Data}
-{Principles} to {Facilitate} {Missing} {Data} {Exploration},
-{Visualization} and {Assessment} of {Imputations}.''} \emph{Journal of
-Statistical Software} 105 (February): 1--31.
-\url{https://doi.org/10.18637/jss.v105.i07}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hdi}{}}%
-United Nations Development Programme. 2022. {``Human Development Report
-2021-22.''} New York.
-\url{https://hdr.undp.org/data-center/human-development-index\#/indicies/HDI}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%

### 2023-08-09 tidyindex.qmd
changes after morning re-read

@@
 Tidy tools, on the other hand, are concatenated in a sequence through which the tidy object flows, creating a pipeline for data processing.
-These pipelines are data centric, meaning all the functions take a tidy data object as inputs and return a processed tidy data object, directly ready for the next operations without further reshaping needed.
-Also, the pipeline approach corresponds to the modular programming practice, which break down complex problems into smaller and more manageable pieces, as oppose to a monolithic design, where all the steps are predetermined and integrated into a single piece.
+These pipelines are data centric, meaning all the functions take a tidy data object as inputs and return a processed tidy data object, directly ready for the next operations to be applied.
+Also, the pipeline approach corresponds to the modular programming practice, which breaks down complex problems into smaller and more manageable pieces, as oppose to a monolithic design, where all the steps are predetermined and integrated into a single piece.
+The flexibility provided by the modularity makes it easier to modify certain steps in the pipeline and to maintain and extend the code base.
 
@@
 @wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems.
-While this comment is made in the context of graphics program, it is also applicable to other general or domain specific data analysis workflow.
+While this comment is made in the context of graphics program, it is also applicable generally to data analysis workflow.
 More recently, the tidyverse suite [@wickham_welcome_2019] takes the pipeline approach for general-purpose data wrangling and has gained popularity within the R community.
-The pipeline-style code can be directly read as a series of operations applied successively on the tidy data object, offering a method to document the data wrangling process.
-Unlike textual documentation, which may omit seemingly obvious or less significant information, the pipeline approach captures all the computational details while retaining the conciseness of a coding language, making it easy for human analysts to comprehend.
+The pipeline-style code can be directly read as a series of operations applied successively on the tidy data object, offering a method to document the data wrangling process for analysts to comprehend.
+Unlike textual documentation, which may omit seemingly obvious or less significant information, the pipeline approach captures all the computational details while retaining conciseness.
 
@@
 In the `tidymodels`, these steps are correspondingly implemented as package `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen.
+The uniform interface in tidymodels frees analysts from recalling model-specific syntax for performing the same operation across different models, increasing the efficiency to work with different models simultaneously.
 
@@
 
-For constructing indexes, rather than implementing each index separately, as has been done for some indexes, the pipeline approach introduces explicit, user-facing modules that analysts can compose into indexes.
-This approach provides several advantages to analysts and researchers working on indexes.
-The index construction process can be complicated to describe and readable pipeline code complements conventional diagrams and textual description for understand the computational details.
-The uniform interface allows analysts to assemble a diverse array of indexes with a shared set of components, reducing the learning curve for adopting new implementation for different indexes.
-Importantly, each pipeline module and its corresponding steps are implemented as standalone functions, providing great flexibility for analysis the index.
-Analysts can rearrange the order in the pipeline or insert new step to experiment with possible new definition for the index or to understand index behaviour.
+For constructing indexes, rather than implementing each index separately, as has been done for some indexes, the pipeline approach introduces explicit, user-facing, and standalone modules for analysts to assemble indexes.
+Analysts can rearrange the order in the pipeline or insert new step to experiment with possible new index definition or to understand index behaviour.
+Also, the readability of the pipeline code can complement conventional diagrams and textual information to understand the how indexes are constructed.
+The uniform interface of pipeline allows analysts to assemble a diverse array of indexes with a shared set of components, reducing the learning curve for adopting new implementation for different indexes.
 

### 2023-08-09 tidyindex.tex
changes after morning re-read

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 inputs and return a processed tidy data object, directly ready for the
-next operations without further reshaping needed. Also, the pipeline
-approach corresponds to the modular programming practice, which break
-down complex problems into smaller and more manageable pieces, as oppose
-to a monolithic design, where all the steps are predetermined and
-integrated into a single piece.
+next operations to be applied. Also, the pipeline approach corresponds
+to the modular programming practice, which breaks down complex problems
+into smaller and more manageable pieces, as oppose to a monolithic
+design, where all the steps are predetermined and integrated into a
+single piece. The flexibility provided by the modularity makes it easier
+to modify certain steps in the pipeline and to maintain and extend the
+code base.
 
@@
 this comment is made in the context of graphics program, it is also
-applicable to other general or domain specific data analysis workflow.
-More recently, the tidyverse suite (Wickham et al. 2019) takes the
-pipeline approach for general-purpose data wrangling and has gained
-popularity within the R community. The pipeline-style code can be
-directly read as a series of operations applied successively on the tidy
-data object, offering a method to document the data wrangling process.
-Unlike textual documentation, which may omit seemingly obvious or less
-significant information, the pipeline approach captures all the
-computational details while retaining the conciseness of a coding
-language, making it easy for human analysts to comprehend.
+applicable generally to data analysis workflow. More recently, the
+tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
+general-purpose data wrangling and has gained popularity within the R
+community. The pipeline-style code can be directly read as a series of
+operations applied successively on the tidy data object, offering a
+method to document the data wrangling process for analysts to
+comprehend. Unlike textual documentation, which may omit seemingly
+obvious or less significant information, the pipeline approach captures
+all the computational details while retaining conciseness.
 
@@
 \texttt{parsnip}, and \texttt{yardstick}, agnostic to the specific model
-chosen.
+chosen. The uniform interface in tidymodels frees analysts from
+recalling model-specific syntax for performing the same operation across
+different models, increasing the efficiency to work with different
+models simultaneously.
 
@@
 separately, as has been done for some indexes, the pipeline approach
-introduces explicit, user-facing modules that analysts can compose into
-indexes. This approach provides several advantages to analysts and
-researchers working on indexes. The index construction process can be
-complicated to describe and readable pipeline code complements
-conventional diagrams and textual description for understand the
-computational details. The uniform interface allows analysts to assemble
-a diverse array of indexes with a shared set of components, reducing the
-learning curve for adopting new implementation for different indexes.
-Importantly, each pipeline module and its corresponding steps are
-implemented as standalone functions, providing great flexibility for
-anlaysis the index. Analysts can rearrange the order in the pipeline or
-insert new step to experiment with possible new definition for the index
-or to understand index behaviour.
+introduces explicit, user-facing, and standalone modules for analysts to
+assemble indexes. Analysts can rearrange the order in the pipeline or
+insert new step to experiment with possible new index definition or to
+understand index behaviour. Also, the readability of the pipeline code
+can complement conventional diagrams and textual information to
+understand the how indexes are constructed. The uniform interface of
+pipeline allows analysts to assemble a diverse array of indexes with a
+shared set of components, reducing the learning curve for adopting new
+implementation for different indexes.
 

### 2023-08-09 tidyindex.tex
Update section 2

@@
 the other hand, are concatenated in a sequence through which the tidy
-object flows, creating a pipeline for data processing. These pipelines
-are data centric, meaning all the functions take a tidy data object as
-inputs and return a processed tidy data object, directly ready for the
+data flows, creating a pipeline for data processing and modelling. These pipelines
+are data centric, meaning all the tidy tools or functions take a tidy data object as
+input and return a processed tidy data object, directly ready for the
 next operations to be applied. Also, the pipeline approach corresponds
 to the modular programming practice, which breaks down complex problems
-into smaller and more manageable pieces, as oppose to a monolithic
-design, where all the steps are predetermined and integrated into a
-single piece. The flexibility provided by the modularity makes it easier
-to modify certain steps in the pipeline and to maintain and extend the
+into smaller and more manageable pieces.
+% , as oppose to a monolithic
+% design, where all the steps are predetermined and integrated into a
+% single piece. 
+The flexibility provided by the modularity makes it easier
+to modify any step in the pipeline and to maintain and extend the
 code base.
 
-Examples of using a pipeline approach to structure data analysis in R
-can be traced back to the interactive graphics literature in the earlier
-time (Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng
+Examples of using a pipeline approach for data analysis 
+can be traced back to the interactive graphics literature including 
+Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng
 2014; Wickham et al. 2009). Wickham et al. (2009) argues that whether
-made explicit or not, pipeline has to be presented in every graphics
+made explicit or not, a pipeline has to be presented in every graphics
 program and making them explicit is beneficial for understanding the
 implementation and comparing between different graphic systems. While
-this comment is made in the context of graphics program, it is also
-applicable generally to data analysis workflow. More recently, the
+this comment is made in the context of graphics program (WHAT DO YOU MEAN by graphics program?), it is also
+applicable generally to any data analysis workflow. More recently, the
 tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
@@
 community. The pipeline-style code can be directly read as a series of
-operations applied successively on the tidy data object, offering a
-method to document the data wrangling process for analysts to
-comprehend. Unlike textual documentation, which may omit seemingly
-obvious or less significant information, the pipeline approach captures
-all the computational details while retaining conciseness.
+operations applied successively on a tidy data object, offering a
+method to create a reproducible data wrangling process that is also easy to document.
+% Unlike textual documentation, which may omit seemingly
+% obvious or less significant information, the pipeline approach captures
+% all the computational details while retaining conciseness.
 
@@
 workflow tailored to a specific domain, developers first need to
-identify the fundamental building blocks in the typical workflow. These
+identify the fundamental building blocks to create a workflow. These
 components are then implemented as modules, which can be combined to
@@
 For constructing indexes, rather than implementing each index
-separately, as has been done for some indexes, the pipeline approach
-introduces explicit, user-facing, and standalone modules for analysts to
-assemble indexes. Analysts can rearrange the order in the pipeline or
-insert new step to experiment with possible new index definition or to
-understand index behaviour. Also, the readability of the pipeline code
-can complement conventional diagrams and textual information to
-understand the how indexes are constructed. The uniform interface of
+separately, as it has been done for some indexes, the pipeline approach
+introduces explicit, user-facing (??), and standalone modules for analysts to
+assemble indexes. Analysts can then rearrange the pipeline or
+insert new steps to create indexes and/or to experiment with possible index definition while also testing
+the index behaviour. 
+% 
+% In addition, the readability of the pipeline code
+% can complement conventional diagrams and textual information to
+% understand the how indexes are constructed.
+
+(I don't think this is needed -->)The uniform interface of a
 pipeline allows analysts to assemble a diverse array of indexes with a

### 2023-08-09 tidyindex.qmd
update the changes to qmd and knit

@@
 The concept of tidy data [@wickham_tidy_2014] prescribes specific rules for organising data in an analysis, with observations as rows, variables as columns, and types of observational units as tables.
-Tidy tools, on the other hand, are concatenated in a sequence through which the tidy object flows, creating a pipeline for data processing.
-These pipelines are data centric, meaning all the functions take a tidy data object as inputs and return a processed tidy data object, directly ready for the next operations to be applied.
+Tidy tools, on the other hand, are concatenated in a sequence through which the tidy data flows, creating a pipeline for data processing and modelling.
+These pipelines are data centric, meaning all the tidy tools or functions take a tidy data object as inputs and return a processed tidy data object, directly ready for the next operations to be applied.
 Also, the pipeline approach corresponds to the modular programming practice, which breaks down complex problems into smaller and more manageable pieces, as oppose to a monolithic design, where all the steps are predetermined and integrated into a single piece.
@@
 
-Examples of using a pipeline approach to structure data analysis in R can be traced back to the interactive graphics literature in the earlier time [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009].
-@wickham_plumbing_2009 argues that whether made explicit or not, pipeline has to be presented in every graphics program and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems.
-While this comment is made in the context of graphics program, it is also applicable generally to data analysis workflow.
+Examples of using a pipeline approach for data analysis can be traced back to the interactive graphics literature, including @buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009].
+@wickham_plumbing_2009 argues that whether made explicit or not, a pipeline has to be presented in every graphics program and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems.
+While this comment is made in the context of interactive graphics program, it is also applicable generally to any data analysis workflow.
 More recently, the tidyverse suite [@wickham_welcome_2019] takes the pipeline approach for general-purpose data wrangling and has gained popularity within the R community.
-The pipeline-style code can be directly read as a series of operations applied successively on the tidy data object, offering a method to document the data wrangling process for analysts to comprehend.
-Unlike textual documentation, which may omit seemingly obvious or less significant information, the pipeline approach captures all the computational details while retaining conciseness.
+The pipeline-style code can be directly read as a series of operations applied successively on tidy data object, offering a method to document the data wrangling process with all the computational details for reproducibility.
 
 Since the success of tidyverse, more packages have been developed to analyze data using the tidy framework for domains specific applications, a noticeable example of which is `tidymodels` for building machine learning models [@tidymodels].
-To create a tidy workflow tailored to a specific domain, developers first need to identify the fundamental building blocks in the typical workflow.
+To create a tidy workflow tailored to a specific domain, developers first need to identify the fundamental building blocks to create a workflow.
 These components are then implemented as modules, which can be combined to form the pipeline.
@@
 
-For constructing indexes, rather than implementing each index separately, as has been done for some indexes, the pipeline approach introduces explicit, user-facing, and standalone modules for analysts to assemble indexes.
-Analysts can rearrange the order in the pipeline or insert new step to experiment with possible new index definition or to understand index behaviour.
+For constructing indexes, rather than implementing each index separately, as it has been done for some indexes, the pipeline approach introduces explicit and standalone modules for analysts to assemble indexes.
+Analysts can then rearrange the order of modules in the pipeline or insert new modules to experiment with possible new index definition or to understand index behaviour.
 Also, the readability of the pipeline code can complement conventional diagrams and textual information to understand the how indexes are constructed.

### 2023-08-09 tidyindex.tex
update the changes to qmd and knit

--- 2023-08-09 tidyindex.tex
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, breakable, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 the other hand, are concatenated in a sequence through which the tidy
-object flows, creating a pipeline for data processing. These pipelines
-are data centric, meaning all the functions take a tidy data object as
-inputs and return a processed tidy data object, directly ready for the
-next operations to be applied. Also, the pipeline approach corresponds
-to the modular programming practice, which breaks down complex problems
-into smaller and more manageable pieces, as oppose to a monolithic
-design, where all the steps are predetermined and integrated into a
-single piece. The flexibility provided by the modularity makes it easier
-to modify certain steps in the pipeline and to maintain and extend the
-code base.
-
-Examples of using a pipeline approach to structure data analysis in R
-can be traced back to the interactive graphics literature in the earlier
-time (Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng
-2014; Wickham et al. 2009). Wickham et al. (2009) argues that whether
-made explicit or not, pipeline has to be presented in every graphics
+data flows, creating a pipeline for data processing and modelling. These
+pipelines are data centric, meaning all the tidy tools or functions take
+a tidy data object as inputs and return a processed tidy data object,
+directly ready for the next operations to be applied. Also, the pipeline
+approach corresponds to the modular programming practice, which breaks
+down complex problems into smaller and more manageable pieces, as oppose
+to a monolithic design, where all the steps are predetermined and
+integrated into a single piece. The flexibility provided by the
+modularity makes it easier to modify certain steps in the pipeline and
+to maintain and extend the code base.
+
+Examples of using a pipeline approach for data analysis can be traced
+back to the interactive graphics literature, including Buja et al.
+(1988); Sutherland et al. (2000); Xie, Hofmann, and Cheng (2014);
+Wickham et al. (2009){]}. Wickham et al. (2009) argues that whether made
+explicit or not, a pipeline has to be presented in every graphics
 program and making them explicit is beneficial for understanding the
 implementation and comparing between different graphic systems. While
-this comment is made in the context of graphics program, it is also
-applicable generally to data analysis workflow. More recently, the
-tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
-general-purpose data wrangling and has gained popularity within the R
-community. The pipeline-style code can be directly read as a series of
-operations applied successively on the tidy data object, offering a
-method to document the data wrangling process for analysts to
-comprehend. Unlike textual documentation, which may omit seemingly
-obvious or less significant information, the pipeline approach captures
-all the computational details while retaining conciseness.
+this comment is made in the context of interactive graphics program, it
+is also applicable generally to any data analysis workflow. More
+recently, the tidyverse suite (Wickham et al. 2019) takes the pipeline
+approach for general-purpose data wrangling and has gained popularity
+within the R community. The pipeline-style code can be directly read as
+a series of operations applied successively on tidy data object,
+offering a method to document the data wrangling process with all the
+computational details for reproducibility.
 
@@
 For constructing indexes, rather than implementing each index
-separately, as has been done for some indexes, the pipeline approach
-introduces explicit, user-facing, and standalone modules for analysts to
-assemble indexes. Analysts can rearrange the order in the pipeline or
-insert new step to experiment with possible new index definition or to
-understand index behaviour. Also, the readability of the pipeline code
-can complement conventional diagrams and textual information to
-understand the how indexes are constructed. The uniform interface of
-pipeline allows analysts to assemble a diverse array of indexes with a
-shared set of components, reducing the learning curve for adopting new
-implementation for different indexes.
+separately, as it has been done for some indexes, the pipeline approach
+introduces explicit and standalone modules for analysts to assemble
+indexes. Analysts can then rearrange the order of modules in the
+pipeline or insert new modules to experiment with possible new index
+definition or to understand index behaviour. Also, the readability of
+the pipeline code can complement conventional diagrams and textual
+information to understand the how indexes are constructed. The uniform
+interface of pipeline allows analysts to assemble a diverse array of
+indexes with a shared set of components, reducing the learning curve for
+adopting new implementation for different indexes.
 

### 2023-08-09 tidyindex.qmd
pipeline paragraph of section 2

@@
 
-For constructing indexes, rather than implementing each index separately, as it has been done for some indexes, the pipeline approach introduces explicit and standalone modules for analysts to assemble indexes.
-Analysts can then rearrange the order of modules in the pipeline or insert new modules to experiment with possible new index definition or to understand index behaviour.
+For constructing indexes, the pipeline approach adopts explicit and standalone modules that can be assembled in different ways.
+Index developers can choose the appropriate modules and arrange them accordingly to generate the data pipeline that is needed for their purpose.
+The pipeline approach provides many advantages:
+
+- makes the computation more transparent, and thus more easily debugged.
+- allows for rapidly processing new data to check how different features, like outliers, might affect the index value.
+- provides the capacity to measure uncertainty by computing confidence intervals from multiple samples as generated by bootstrapping to original data.
+- enables systematic comparison of surrogate indexes designed to measure the same phenomenon.
+- it may even be possible to automate diagramatic explanations and documentation of the index.
+
+Adoption of this pipeline approach would provide uniformity to the field of index development, research and application.
+
+<!-- They might even  insert new modules to experiment with possible new index definitions or to understand an index's behaviour.
 Also, the readability of the pipeline code can complement conventional diagrams and textual information to understand the how indexes are constructed.
-The uniform interface of pipeline allows analysts to assemble a diverse array of indexes with a shared set of components, reducing the learning curve for adopting new implementation for different indexes.
+The uniform interface of pipeline allows analysts to assemble a diverse array of indexes with a shared set of components, reducing the learning curve for adopting new implementation for different indexes.-->
 
@@
 
-# Details of the index pipline {#sec-index-pipeline}
+# Details of the index pipeline {#sec-index-pipeline}
 
@@
 library(sf)
+# options(timeout = 1000)
+# remotes::install_github("wfmackey/absmapsdata")
 qld_lga <- absmapsdata::lga2018 %>%

### 2023-08-09 tidyindex.tex
pipeline paragraph of section 2

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
 \fi
@@
 \makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
 \makeatother
@@
 \@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
 \makeatother
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, breakable, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-For constructing indexes, rather than implementing each index
-separately, as it has been done for some indexes, the pipeline approach
-introduces explicit and standalone modules for analysts to assemble
-indexes. Analysts can then rearrange the order of modules in the
-pipeline or insert new modules to experiment with possible new index
-definition or to understand index behaviour. Also, the readability of
-the pipeline code can complement conventional diagrams and textual
-information to understand the how indexes are constructed. The uniform
-interface of pipeline allows analysts to assemble a diverse array of
-indexes with a shared set of components, reducing the learning curve for
-adopting new implementation for different indexes.
+For constructing indexes, the pipeline approach adopts explicit and
+standalone modules that can be assembled in different ways. Index
+developers can choose the appropriate modules and arrange them
+accordingly to generate the data pipeline that is needed for their
+purpose. The pipeline approach provides many advantages:
+
+\begin{itemize}
+\tightlist
+\item
+  makes the computation more transparent, and thus more easily debugged.
+\item
+  allows for rapidly processing new data to check how different
+  features, like outliers, might affect the index value.
+\item
+  provides the capacity to measure uncertainty by computing confidence
+  intervals from multiple samples as generated by bootstrapping to
+  original data.
+\item
+  enables systematic comparison of surrogate indexes designed to measure
+  the same phenomenon.
+\item
+  it may even be possible to automate diagramatic explanations and
+  documentation of the index.
+\end{itemize}
+
+Adoption of this pipeline approach would provide uniformity to the field
+of index development, research and application.
 
 \hypertarget{sec-index-pipeline}{%
-\section{Details of the index pipline}\label{sec-index-pipeline}}
+\section{Details of the index pipeline}\label{sec-index-pipeline}}
 
@@
 Global Gender Gap Index}\tabularnewline
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endfirsthead
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 Labour force participation & Economy & 0.199 & 0.25 & 0.050 \\
@@
 Years with female head of state & Politics & 0.443 & 0.25 & 0.111 \\
-\bottomrule()
 \end{longtable}

### 2023-08-09 tidyindex.qmd
remove redundant qld_lga code

@@
 
-Examples of using a pipeline approach for data analysis can be traced back to the interactive graphics literature, including @buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009].
+Examples of using a pipeline approach for data analysis can be traced back to the interactive graphics literature, including @buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009.
 @wickham_plumbing_2009 argues that whether made explicit or not, a pipeline has to be presented in every graphics program and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems.
@@
 
-```{r}
-library(sf)
-# options(timeout = 1000)
-# remotes::install_github("wfmackey/absmapsdata")
-qld_lga <- absmapsdata::lga2018 %>%
-  filter(state_name_2016 == "Queensland") %>%
-  rmapshaper::ms_simplify(keep = 0.3) %>%
-  mutate(lga_name_2018 = str_replace(lga_name_2018, " \\([S|R|C|]\\)", ""),
-         lga_name_2018 = str_replace(lga_name_2018, " \\(Qld\\)", ""))
-```
-
 ```{r fig-compute-spatial}

### 2023-08-09 tidyindex.tex
remove redundant qld_lga code

--- 2023-08-09 tidyindex.tex
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, breakable, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 (1988); Sutherland et al. (2000); Xie, Hofmann, and Cheng (2014);
-Wickham et al. (2009){]}. Wickham et al. (2009) argues that whether made
+Wickham et al. (2009). Wickham et al. (2009) argues that whether made
 explicit or not, a pipeline has to be presented in every graphics

### 2023-08-11 tidyindex.qmd
a draft on introducing the general notation

@@
 
-The notation used for introducing the pipeline modules is as follows.
-Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspects: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$.
-When more than one variable is involved, the multivariate data can also be written as: $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$.
+Now, we introduce the notation used for describing pipeline modules. Consider multivariate spatio-temporal data,
+
+\begin{equation}
+\begin{aligned}
+x^l(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^L(s;t)] \\
+& l = 1, 2, \cdots L, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
+\end{aligned}
+\end{equation}
+
+where $x^l$ is the $l$th variable, $s$ is the spatial location, and $t$ is the temporal order.
+<!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.--> The spatial space, $D_s \subseteq \mathbb{R}^d$, is generic to include different spatial representations, including a 1D set of $n$ locations $\{D_s: s_1, s_2, \cdots, s_n\}$, a 2D set of longitude and latitude pairs, and $d$-D coordinate set.
+The higher dimension in the temporal space, $D_t \subseteq \mathbb{R}^m$, can be used to represent for a $m$-D deconstruction of time.
+For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$.
+In the case when the spatial dimension is fixed, multivariate temporal data can be simplified as $x^l_s(t)$, and similarly,  multivariate spatial data can be simplified as $x^l_t(s)$.
 
@@
 
-**Input: ** $x(\mathbf{s};\mathbf{t})$: multivariate data with temporal coordinates
+`Input`: $x^l_s(t)$,
 
-**Output: ** $x(\mathbf{s};\mathbf{t}^\prime)$: multivariate data with aggregated temporal axis
+`Output`:  $x^l_s(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.
 
-**Computation:** The construction of an index sometimes needs to consider information from neighbouring time periods.
+`Computation`: The construction of an index sometimes needs to consider information from neighbouring time periods.
 The temporal processing is a general operator on the time dimension of the data in the form of
@@
 
-**Input: ** Multivariate data with coordinates
+`Input`:  $x^l_t(s)$
 
-**Output: ** Multivariate data with a new set of coordinates
+`Output`: $x^l_t(s^\prime)$ where $s^\prime \in D_{s^{\prime}}$
 
-**Computation:** Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing.
+`Computation`: Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing.
 The process can be written as a general operation in the form of
@@
 
-**Input: **
+`Input`: $x^l(s; t)$
 
-**Output: **
+`Output`: $y^l(s; t)$
 
-**Computation:** The purpose of variable transformation is to create variables that fits assumptions for further computing.
+`Computation`: The purpose of variable transformation is to create variables that fits assumptions for further computing.
 These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline.
@@
 
-**Input: **
+`Input`: $x^l(s; t)$
 
-**Output: **
+`Output`: $x^l(s; t)$
 
-**Computation:** While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
+`Computation`: While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
 The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of
@@
 
-**Input: **
+`Input`: $x^l(s; t)$
 
-**Output: **
+`Output`: $x^{l^\prime}(s; t)$, where $l^{\prime} = 1, 2, \cdots, L^{\prime}$
 
-**Computation:** Dimension reduction summarises the multivariate information into univariate, which can be denoted as:
+`Computation`: Dimension reduction summarises the multivariate information into univariate, which can be denoted as:
 
@@
 
-*model fit? *
+<!-- *model fit? * -->
 
-**Input: **
+`Input`:  $x^l(s; t)$
 
-**Output: **
+`Output`:
 
-**Computation:** Distribution fit can be seen as the model fitting in its simplest term. It can be represented by
+`Computation`: Distribution fit can be seen as the model fitting in its simplest term. It can be represented by
 
@@
 
-**Input: **
+`Input`:  $x^l(s; t)$
 
-**Output: **
+`Output`:
 
-**Computation:** This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
+`Computation`: This step maps the univariate series into a different scale, typically for ease of comparison across regions. 
 For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
@@
 
-**Input: **
+`Input`:  $x^l(s; t)$
 
-**Output: **
+`Output`:
 
-**Computation:** Benchmarking sets a constant value to allow the constructed index to be compared across time.
+$D(x^l(s,t), a)$ where $D(., .)$ is a distance function between
+
+`Computation`: Benchmarking sets a constant value to allow the constructed index to be compared across time.
 Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed. A benchmark value could be a constant or a function of the data, i.e. mean.
@@
 
-**Input: **
+`Input`:  $x^l(s; t)$
 
-**Output: **
+`Output`: $C^l(s;t)$ where $\{C: c_0, c_1, \cdots, c_z\}$ is a discrete set
 
-**Computation:** In public communication, the index values are usually accompanied by a categorical grade.
+`Computation`: In public communication, the index values are usually accompanied by a categorical grade.
 The categorised grades are an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes.
@@
 \begin{cases}
-C_0 & c_1 \leq (s_i; t_j) < c_0 \\
-C_1 & c_2 \leq x(s_i; t_j) < c_1 \\
-C_2 & c_3 \leq x(s_i; t_j) < c_2 \\
+c_0 & C_1 \leq x^l(s; t) < C_0 \\
+c_1 & C_2 \leq x^l(s; t) < C_1 \\
+c_2 & C_3 \leq x^l(s; t) < C_2 \\
 \cdots \\
-C_z & c_z \leq x(s_i; t_j)
+c_z & C_z \leq x^l(s; t)
 \end{cases}

### 2023-08-11 tidyindex.tex
a draft on introducing the general notation

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-The notation used for introducing the pipeline modules is as follows.
-Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with
-spatial, temporal, and multivariate aspects: the spatial dimension
-\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
-space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
-temporal dimension \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
-defined in the 1D space:
-\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). When more than
-one variable is involved, the multivariate data can also be written as:
-\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\).
+Now, we introduce the notation used for describing pipeline modules.
+Consider multivariate spatio-temporal data,
+
+\begin{equation}
+\begin{aligned}
+x^l(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^L(s;t)] \\
+& l = 1, 2, \cdots L, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
+\end{aligned}
+\end{equation}
+
+where \(x^l\) is the \(l\)th variable, \(s\) is the spatial location,
+and \(t\) is the temporal order. The spatial space,
+\(D_s \subseteq \mathbb{R}^d\), is generic to include different spatial
+representations, including a 1D set of \(n\) locations
+\(\{D_s: s_1, s_2, \cdots, s_n\}\), a 2D set of longitude and latitude
+pairs, and \(d\)-D coordinate set. The higher dimension in the temporal
+space, \(D_t \subseteq \mathbb{R}^m\), can be used to represent for a
+\(m\)-D deconstruction of time. For example, a year-month combination,
+\((Y, M)^\prime\), can be used to record monthly precipitation series
+from January 1990 to December 2020, where \(Y = 1990, \cdots, 2020\) and
+\(M = 1, \cdots, 12\). In the case when the spatial dimension is fixed,
+multivariate temporal data can be simplified as \(x^l_s(t)\), and
+similarly, multivariate spatial data can be simplified as \(x^l_t(s)\).
 
@@
 
-\textbf{Input: } \(x(\mathbf{s};\mathbf{t})\): multivariate data with
-temporal coordinates
-
-\textbf{Output: } \(x(\mathbf{s};\mathbf{t}^\prime)\): multivariate data
-with aggregated temporal axis
-
-\textbf{Computation:} The construction of an index sometimes needs to
+\texttt{Input}: \(x^l_s(t)\),
+
+\texttt{Output}: \(x^l_s(t^{\prime})\), where
+\(t^\prime \in D_{t^{\prime}}\).
+
+\texttt{Computation}: The construction of an index sometimes needs to
 consider information from neighbouring time periods. The temporal
@@
 
-\textbf{Input: } Multivariate data with coordinates
-
-\textbf{Output: } Multivariate data with a new set of coordinates
-
-\textbf{Computation:} Spatial processing may be needed when indexes are
+\texttt{Input}: \(x^l_t(s)\)
+
+\texttt{Output}: \(x^l_t(s^\prime)\) where
+\(s^\prime \in D_{s^{\prime}}\)
+
+\texttt{Computation}: Spatial processing may be needed when indexes are
 not calculated independently on each collected location or when
@@
 
-\textbf{Input: }
-
-\textbf{Output: }
-
-\textbf{Computation:} The purpose of variable transformation is to
+\texttt{Input}: \(x^l(s; t)\)
+
+\texttt{Output}: \(y^l(s; t)\)
+
+\texttt{Computation}: The purpose of variable transformation is to
 create variables that fits assumptions for further computing. These
@@
 
-\textbf{Input: }
-
-\textbf{Output: }
-
-\textbf{Computation:} While scaling can be seen as a specific type of
+\texttt{Input}: \(x^l(s; t)\)
+
+\texttt{Output}: \(x^l(s; t)\)
+
+\texttt{Computation}: While scaling can be seen as a specific type of
 variable transformation, it is separated into its own step to make the
@@
 
-\textbf{Input: }
-
-\textbf{Output: }
-
-\textbf{Computation:} Dimension reduction summarises the multivariate
+\texttt{Input}: \(x^l(s; t)\)
+
+\texttt{Output}: \(x^{l^\prime}(s; t)\), where
+\(l^{\prime} = 1, 2, \cdots, L^{\prime}\)
+
+\texttt{Computation}: Dimension reduction summarises the multivariate
 information into univariate, which can be denoted as:
@@
 
-\emph{model fit? }
-
-\textbf{Input: }
-
-\textbf{Output: }
-
-\textbf{Computation:} Distribution fit can be seen as the model fitting
+\texttt{Input}: \(x^l(s; t)\)
+
+\texttt{Output}:
+
+\texttt{Computation}: Distribution fit can be seen as the model fitting
 in its simplest term. It can be represented by
@@
 
-\textbf{Input: }
-
-\textbf{Output: }
-
-\textbf{Computation:} This step maps the univariate series into a
+\texttt{Input}: \(x^l(s; t)\)
+
+\texttt{Output}:
+
+\texttt{Computation}: This step maps the univariate series into a
 different scale, typically for ease of comparison across regions. For
@@
 
-\textbf{Input: }
-
-\textbf{Output: }
-
-\textbf{Computation:} Benchmarking sets a constant value to allow the
+\texttt{Input}: \(x^l(s; t)\)
+
+\texttt{Output}:
+
+\(D(x^l(s,t), a)\) where \(D(., .)\) is a distance function between
+
+\texttt{Computation}: Benchmarking sets a constant value to allow the
 constructed index to be compared across time. Here we denote it with
@@
 
-\textbf{Input: }
-
-\textbf{Output: }
-
-\textbf{Computation:} In public communication, the index values are
+\texttt{Input}: \(x^l(s; t)\)
+
+\texttt{Output}: \(C^l(s;t)\) where \(\{C: c_0, c_1, \cdots, c_z\}\) is
+a discrete set
+
+\texttt{Computation}: In public communication, the index values are
 usually accompanied by a categorical grade. The categorised grades are

### 2023-08-14 tidyindex.qmd
update the gggi plot

@@
 
-To understand the uncertainty of this dimension reduction step while avoiding the missingness issue on the variable level, we can run a local tour to slightly vary the weight of each dimension to see how index value and country ranking changes.
-We select countries in the South Asia and Sub-Saharan Africa region and gradually increase the weight of one variable and reduce back to equal weight all four dimensions, one at a time, to produce an animation of how GGGI changes in each country.
-Five frames (equal weights and one for each dimension with a relatively higher weight) selected from the tour animation are shown in @fig-idx-tour and you can find the link to the full animation in figure caption.
+To understand the uncertainty of this dimension reduction step from combining four dimensions (Economy, Education, Health, and Politics), we run a local tour [TODO: reference to local tour] that varies the weight of each dimension slightly to observe how index value and country ranking changes.
+The local tour produces an animation that gradually increases the weight of one variable and reduces it back to equal weight, one at a time.
+A set of countries are selected for this exercise, which includes 1) the top four with GGGI > 0.85, 2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the bottom five countries with GGGI < 0.6.
+Five frames (equal weights and one for each dimension with a relatively higher weight) are selected from the local tour animation to show in @fig-idx-tour and the full animation is available at  https://vimeo.com/847874016?share=copy.
 
-Many insights can be derived from the animation and these selected frames.
-These can be helpful in understanding the sensitivity of the index value and country ranking to each variable and identifying the strengths and weaknesses of each country relative to others.
+@fig-idx-tour helps to understand the how different dimension reduction alternatives affect the whole index distribution and individual countries.
+
 When economy and education are given a greater weight, the variation in index values and ranking highlights countries that performs relatively better (moving right) or worse (moving left) in these two single dimensions.
@@
 This reveals the economy as a weakness for these two countries compared to their similarly-ranked peers.
+
 When health receives a higher weight, the impact on the index value is minimal.
+
 Notably, increasing the weight in politics leads to a pronounced decrease of index value for almost all the countries.
@@
   dplyr::select(country: political_empowerment) %>%
-  filter(country %in% c("Namibia", "South Africa", "Bangladesh", "Botswana",
-                        "Ethiopia", "Kenya", "Senegal", "Afghanistan")) %>%
+  filter(between(index, 0.72, 0.729)  | index < 0.6 | index >= 0.85) %>%
   arrange(index) 
@@
 w_idx2proj <- function(w){w/(sqrt(sum(w^2))) %>% matrix(nrow = length(w))}
-find_next <- function(matrix){
+find_next <- function(matrix, angle = 0.4){
   start <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1)
-  tourr:::step_angle(tourr:::geodesic_info(start, matrix), 0.4)
+  tourr:::step_angle(tourr:::geodesic_info(start, matrix), angle = angle)
 }
-b <- array(dim = c(4, 1, 9))
-b[,,c(1,3,5,7,9)] <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1) %>% w_idx2proj()
-b[,,2] <- matrix(c(0.7, 0.1, 0.1, 0.1), ncol = 1) %>% find_next()
-b[,,4] <- matrix(c(0.1, 0.7, 0.1, 0.1), ncol= 1) %>% find_next()
-b[,,6] <- matrix(c(0.1, 0.1, 0.5, 0.1), ncol= 1) %>% find_next()
-b[,,8] <- matrix(c(0.1, 0.1, 0.1, 0.5), ncol = 1) %>% find_next()
+b <- array(dim = c(4, 1, 3))
+b[,,1] <- matrix(c(0.3, 0.3, 0.3, 0.1), ncol = 1) %>% find_next()
+b[,,2] <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1) %>% w_idx2proj()
+b[,,3] <- matrix(c(0.1, 0.1, 0.1, 0.7), ncol= 1) %>% find_next(angle = 0.6)
 
@@
   dt[,5:8], planned2_tour(b), dev = "png",
-  display_idx(cex = 4, label_cex = 3.5, panel_height_ratio = c(3,1), 
-              axis_cex = 2.5, weight_cex = 3, label_col = "grey70", 
+  display_idx(half_range = 2,cex = 4, label_cex = 3.5, col = "black",
+              panel_height_ratio = c(4,1),
+              axis_bar_lwd = c(rep(3, 3), 7),
+              axis_bar_col = c(rep("#000000", 3), "red"),
+              axis_bar_label_cex = c(rep(2.5, 3), 3),
+              axis_bar_label_col = c(rep("#000000", 3), "red"), 
+              axis_label_cex = 3, frame_x_pos = 0.16, 
+              axis_var_cex = 2.5,
               label = dt$country, abb_vars = FALSE),
-  width = 900, height = 900, apf = 1/20, frames = 100,
+  width = 900, height = 1400, apf = 1/20, frames = 120,
   here::here("figures/idx-tour/idx-tour-%03d.png")
@@
   gif_file = here::here("figures/idx-tour.gif"),
-  delay = 0.15, width = 900, height = 900)
+  delay = 0.15, width = 900, height = 1400)
 
@@
 
-```{r fig-idx-tour, fig.align='center', fig.height=8} 
+* Bangladesh uses politics to compensate economy
+* increase economy -> 
+* increase education -> 
+* increase health -> shrink the range of index (bring top and bottom country closer)
+* increase politics -> all the country has index value smaller
+
+```{r fig-idx-tour, fig.align='center', fig.height=5} 
 #| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index.
@@
 Full animation is available at https://vimeo.com/847874016?share=copy."
-frames <- c("001", "013", "037", "061", "085")
+#frames <- c("001", "013", "037", "061", "085", "109")
+frames <- c("001", "006", "012", "018", "024", "029")
 ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
@@
 gl <-  lapply(rl, grid::rasterGrob)
-wrap_plots(gl, ncol = 2)  
+wrap_plots(gl, ncol = 3)  
 ```
+
+<!-- These can be helpful in understanding the sensitivity of the index value and country ranking to each variable and identifying the strengths and weaknesses of each country relative to others.  -->
 

### 2023-08-14 tidyindex.tex
update the gggi plot

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-To understand the uncertainty of this dimension reduction step while
-avoiding the missingness issue on the variable level, we can run a local
-tour to slightly vary the weight of each dimension to see how index
-value and country ranking changes. We select countries in the South Asia
-and Sub-Saharan Africa region and gradually increase the weight of one
-variable and reduce back to equal weight all four dimensions, one at a
-time, to produce an animation of how GGGI changes in each country. Five
-frames (equal weights and one for each dimension with a relatively
-higher weight) selected from the tour animation are shown in
-Figure~\ref{fig-idx-tour} and you can find the link to the full
-animation in figure caption.
-
-Many insights can be derived from the animation and these selected
-frames. These can be helpful in understanding the sensitivity of the
-index value and country ranking to each variable and identifying the
-strengths and weaknesses of each country relative to others. When
-economy and education are given a greater weight, the variation in index
-values and ranking highlights countries that performs relatively better
-(moving right) or worse (moving left) in these two single dimensions.
-For example, the index moves to the left for Ethiopia and Senegal when
-economy is given a higher weight. This reveals the economy as a weakness
-for these two countries compared to their similarly-ranked peers. When
-health receives a higher weight, the impact on the index value is
-minimal. Notably, increasing the weight in politics leads to a
-pronounced decrease of index value for almost all the countries. Given
-the index value has a direct interpretation on how much of the gender
-gap has been closed, such a weight variation can cast a negative light
-on the progression to gender parity. Analysts need to further
-investigate the construction of the politics dimension and consider
-whether this is desirable for the index.
+To understand the uncertainty of this dimension reduction step from
+combining four dimensions (Economy, Education, Health, and Politics), we
+run a local tour {[}TODO: reference to local tour{]} that varies the
+weight of each dimension slightly to observe how index value and country
+ranking changes. The local tour produces an animation that gradually
+increases the weight of one variable and reduces it back to equal
+weight, one at a time. A set of countries are selected for this
+exercise, which includes 1) the top four with GGGI \textgreater{} 0.85,
+2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama,
+Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the
+bottom five countries with GGGI \textless{} 0.6. Five frames (equal
+weights and one for each dimension with a relatively higher weight) are
+selected from the local tour animation to show in
+Figure~\ref{fig-idx-tour} and the full animation is available at
+https://vimeo.com/847874016?share=copy.
+
+Figure~\ref{fig-idx-tour} helps to understand the how different
+dimension reduction alternatives affect the whole index distribution and
+individual countries.
+
+When economy and education are given a greater weight, the variation in
+index values and ranking highlights countries that performs relatively
+better (moving right) or worse (moving left) in these two single
+dimensions. For example, the index moves to the left for Ethiopia and
+Senegal when economy is given a higher weight. This reveals the economy
+as a weakness for these two countries compared to their similarly-ranked
+peers.
+
+When health receives a higher weight, the impact on the index value is
+minimal.
+
+Notably, increasing the weight in politics leads to a pronounced
+decrease of index value for almost all the countries. Given the index
+value has a direct interpretation on how much of the gender gap has been
+closed, such a weight variation can cast a negative light on the
+progression to gender parity. Analysts need to further investigate the
+construction of the politics dimension and consider whether this is
+desirable for the index.
+
+\begin{itemize}
+\tightlist
+\item
+  Bangladesh uses politics to compensate economy
+\item
+  increase economy -\textgreater{}
+\item
+  increase education -\textgreater{}
+\item
+  increase health -\textgreater{} shrink the range of index (bring top
+  and bottom country closer)
+\item
+  increase politics -\textgreater{} all the country has index value
+  smaller
+\end{itemize}
 

### 2023-08-14 tidyindex.qmd
outline insights from the gggi plot

@@
 
-When economy and education are given a greater weight, the variation in index values and ranking highlights countries that performs relatively better (moving right) or worse (moving left) in these two single dimensions.
-For example, the index moves to the left for Ethiopia and Senegal when economy is given a higher weight.
-This reveals the economy as a weakness for these two countries compared to their similarly-ranked peers.
+* Bangladesh 
+* when decreasing the weight on politics, difference between countries narrower. 
+* increase politics -> all the country has lower index value
 
-When health receives a higher weight, the impact on the index value is minimal.
-
-Notably, increasing the weight in politics leads to a pronounced decrease of index value for almost all the countries.
+<!-- Notably, increasing the weight in politics leads to a pronounced decrease of index value for almost all the countries.
 Given the index value has a direct interpretation on how much of the gender gap has been closed, such a weight variation can cast a negative light on the progression to gender parity.
 Analysts need to further investigate the construction of the politics dimension and consider whether this is desirable for the index.
+-->
 
@@
 
-* Bangladesh uses politics to compensate economy
-* increase economy -> 
-* increase education -> 
-* increase health -> shrink the range of index (bring top and bottom country closer)
-* increase politics -> all the country has index value smaller
-
 ```{r fig-idx-tour, fig.align='center', fig.height=5} 
@@
 ```
+
+<!-- When economy and education are given a greater weight, the variation in index values and ranking highlights countries that performs relatively better (moving right) or worse (moving left) in these two single dimensions.
+For example, the index moves to the left for Ethiopia and Senegal when economy is given a higher weight.
+This reveals the economy as a weakness for these two countries compared to their similarly-ranked peers.
+-->
 

### 2023-08-14 tidyindex.tex
outline insights from the gggi plot

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-When economy and education are given a greater weight, the variation in
-index values and ranking highlights countries that performs relatively
-better (moving right) or worse (moving left) in these two single
-dimensions. For example, the index moves to the left for Ethiopia and
-Senegal when economy is given a higher weight. This reveals the economy
-as a weakness for these two countries compared to their similarly-ranked
-peers.
-
-When health receives a higher weight, the impact on the index value is
-minimal.
-
-Notably, increasing the weight in politics leads to a pronounced
-decrease of index value for almost all the countries. Given the index
-value has a direct interpretation on how much of the gender gap has been
-closed, such a weight variation can cast a negative light on the
-progression to gender parity. Analysts need to further investigate the
-construction of the politics dimension and consider whether this is
-desirable for the index.
-
 \begin{itemize}
@@
 \item
-  Bangladesh uses politics to compensate economy
-\item
-  increase economy -\textgreater{}
-\item
-  increase education -\textgreater{}
-\item
-  increase health -\textgreater{} shrink the range of index (bring top
-  and bottom country closer)
-\item
-  increase politics -\textgreater{} all the country has index value
-  smaller
+  Bangladesh
+\item
+  when decreasing the weight on politics, difference between countries
+  narrower.
+\item
+  increase politics -\textgreater{} all the country has lower index
+  value
 \end{itemize}

### 2023-08-14 tidyindex.qmd
various small changes

@@
 \begin{aligned}
-x^l(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^L(s;t)] \\
-& l = 1, 2, \cdots L, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
+\mathbf{x}(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^P(s;t)] \\
+& p = 1, 2, \cdots P, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
 \end{aligned}
@@
 
-where $x^l$ is the $l$th variable, $s$ is the spatial location, and $t$ is the temporal order.
+where $x^l$ is the $p$th variable, $s$ is the spatial location, and $t$ is the temporal order.
 <!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.--> The spatial space, $D_s \subseteq \mathbb{R}^d$, is generic to include different spatial representations, including a 1D set of $n$ locations $\{D_s: s_1, s_2, \cdots, s_n\}$, a 2D set of longitude and latitude pairs, and $d$-D coordinate set.
@@
 For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$.
-In the case when the spatial dimension is fixed, multivariate temporal data can be simplified as $x^l_s(t)$, and similarly,  multivariate spatial data can be simplified as $x^l_t(s)$.
+To simplify the notation, a subscript letter $s$ or $t$ is used to denote the data have a fixed spatial locations or temporal order.
+When the spatial locations are fixed, the data become multivariate time series, denoted as $x^l_s(t)$.
+Similarly,  $x^l_t(s)$ represents multivariate spatial data when the temporal dimension is fixed.
 
@@
 
-`Input`: $x^l_s(t)$,
+`Input`: $x_s(t)$,
 
@@
 
-```{r cache=TRUE, echo = TRUE, message=FALSE}
+```{r cache=TRUE, echo = TRUE, message=FALSE, eval=FALSE}
 .scale <- c(6, 12, 24, 36)
@@
   )
+```
+
+```{r eval = FALSE}
+save(idx, file = here::here("data/idx.rda"))
 ```
@@
 The map shows a continous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 Janurary, likely due to the increased rainfall in February from the meteorological record."
-#| fig-width: 12
-#| fig-height: 5
+#| fig-width: 6
+load(here::here("data/idx.rda"))
 queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
@@
 in 1998 and 2020.'
-#| fig-width: 12
-#| fig-height: 6
+#| fig-width: 8
+#| fig-height: 5
 texas <- idx %>%
@@
   patchwork::plot_annotation(tag_levels = "a") + 
-  patchwork::plot_layout(guides = "collect")  &
+  patchwork::plot_layout(guides = "collect", widths = c(1, 4))  &
   theme(legend.position = "bottom")
@@
 
-The result can be compared with the GGGI values available in the report as shown in @fig-compare-gggi, validating the reproducibility of the index for country with no missing variables.
+The result can be compared with the GGGI values available in the report, validating the reproducibility of the index for country with no missing variables.
 
@@
       weight = weight)) 
-
-dt2$data %>%
-  filter(!is.na(index_new)) %>% 
-  ggplot(aes(x = index, y = index_new)) +
-  geom_abline(slope = 1, intercept = 0, color = "grey") + 
-  geom_point() +
-  theme_bw() + 
-  theme(aspect.ratio = 1, panel.grid.minor = element_blank()) + 
-  xlab("Benchmark") + 
-  ylab("Index calculated")
 ```
@@
 A set of countries are selected for this exercise, which includes 1) the top four with GGGI > 0.85, 2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the bottom five countries with GGGI < 0.6.
-Five frames (equal weights and one for each dimension with a relatively higher weight) are selected from the local tour animation to show in @fig-idx-tour and the full animation is available at  https://vimeo.com/847874016?share=copy.
+Five frames (equal weights and one for each dimension with a relatively higher weight) are selected from the local tour animation to show in @fig-idx-tour and the full animation is available at  https://vimeo.com/847874016.
 
@@
 The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse.
-Full animation is available at https://vimeo.com/847874016?share=copy."
+Full animation is available at https://vimeo.com/847874016."
 #frames <- c("001", "013", "037", "061", "085", "109")

### 2023-08-14 tidyindex.tex
various small changes

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-where \(x^l\) is the \(l\)th variable, \(s\) is the spatial location,
+where \(x^l\) is the \(p\)th variable, \(s\) is the spatial location,
 and \(t\) is the temporal order. The spatial space,
@@
 from January 1990 to December 2020, where \(Y = 1990, \cdots, 2020\) and
-\(M = 1, \cdots, 12\). In the case when the spatial dimension is fixed,
-multivariate temporal data can be simplified as \(x^l_s(t)\), and
-similarly, multivariate spatial data can be simplified as \(x^l_t(s)\).
+\(M = 1, \cdots, 12\). To simplify the notation, a subscript letter
+\(s\) or \(t\) is used to denote the data have a fixed spatial locations
+or temporal order. When the spatial locations are fixed, the data become
+multivariate time series, denoted as \(x^l_s(t)\). Similarly,
+\(x^l_t(s)\) represents multivariate spatial data when the temporal
+dimension is fixed.
 
@@
 
-\texttt{Input}: \(x^l_s(t)\),
+\texttt{Input}: \(x_s(t)\),
 
@@
     \AttributeTok{spei =} \FunctionTok{idx\_spei}\NormalTok{(}
-      \AttributeTok{.pet\_method =} \StringTok{"thornthwaite"}\NormalTok{, }\AttributeTok{.tavg =}\NormalTok{ tavg, }\AttributeTok{.lat =}\NormalTok{ lat,}
+      \AttributeTok{.pet\_method =} \StringTok{"thornthwaite"}\NormalTok{, }\AttributeTok{.tavg =}\NormalTok{ tavg, }\AttributeTok{.lat =}\NormalTok{ lat, }
       \AttributeTok{.scale =}\NormalTok{ .scale, }\AttributeTok{.dist =} \FunctionTok{c}\NormalTok{(}\FunctionTok{gev}\NormalTok{(), }\FunctionTok{loglogistic}\NormalTok{())),}
@@
 \end{Shaded}
-
-\begin{verbatim}
-[1] "Checking for missing values (`NA`): all the data must be complete. Input type is vector. Assuming the data are monthly time series starting in January, all regular (non-leap) years."
-\end{verbatim}
 
@@
 
-The result can be compared with the GGGI values available in the report
-as shown in Figure~\ref{fig-compare-gggi}, validating the
-reproducibility of the index for country with no missing variables.
-
-\begin{figure}
-
-{\centering \includegraphics{tidyindex_files/figure-pdf/fig-compare-gggi-1.pdf}
-
-}
-
-\caption{\label{fig-compare-gggi}Verifying the calculation of the Global
-Gender Gap Index (GGGI). The index can be reproduced from the
-methodology described in the Global Gender Gap Report 2023 after
-removing the countries with missing variables, the treatment of which is
-unclear.}
-
-\end{figure}
+The result can be compared with the GGGI values available in the report,
+validating the reproducibility of the index for country with no missing
+variables.
 
@@
 Figure~\ref{fig-idx-tour} and the full animation is available at
-https://vimeo.com/847874016?share=copy.
+https://vimeo.com/847874016.
 
@@
 0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full
-animation is available at https://vimeo.com/847874016?share=copy.}
+animation is available at https://vimeo.com/847874016.}
 

### 2023-08-14 tidyindex.qmd
a table listing all the notations + visual mode reformat

@@
 <!-- - Why index is useful, why people care about indexes -->
+
 <!-- - Define what is an index, what is not -->
+
 <!-- - What is the challenges with current index construction -->
+
 <!-- - what can be done if people adopt this pipeline/ why it is beneficial? -->
+
 <!-- - who would benefit from this paper -->
@@
 
-Now, we introduce the notation used for describing pipeline modules. Consider multivariate spatio-temporal data,
+Now, we introduce the notation used for describing pipeline modules. Consider multivariate spatio-temporal process,
 
+```{=tex}
 \begin{equation}
-\begin{aligned}
-\mathbf{x}(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^P(s;t)] \\
-& p = 1, 2, \cdots P, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
-\end{aligned}
+\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
 \end{equation}
+```
+where:
 
-where $x^l$ is the $p$th variable, $s$ is the spatial location, and $t$ is the temporal order.
-<!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.--> The spatial space, $D_s \subseteq \mathbb{R}^d$, is generic to include different spatial representations, including a 1D set of $n$ locations $\{D_s: s_1, s_2, \cdots, s_n\}$, a 2D set of longitude and latitude pairs, and $d$-D coordinate set.
-The higher dimension in the temporal space, $D_t \subseteq \mathbb{R}^m$, can be used to represent for a $m$-D deconstruction of time.
-For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$.
-To simplify the notation, a subscript letter $s$ or $t$ is used to denote the data have a fixed spatial locations or temporal order.
+-   $x$ is the variable in $\mathbb{R}^p$,
+
+-   $s$ represents the geographic locations in the space $D_s \subseteq \mathbb{R}^m$.
+This includes a collection of countries of interest and paired longitude and latitude coordinates of locations, and
+
+-   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. Sometimes, it can be time decomposed into several components, such as year, month, quarter, and season.
+
+<!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.-->
+
+<!--For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$.
+.-->
+
+<!-- To simplify the notation, a subscript letter $s$ or $t$ is used to denote the data have a fixed spatial locations or temporal order.
 When the spatial locations are fixed, the data become multivariate time series, denoted as $x^l_s(t)$.
 Similarly,  $x^l_t(s)$ represents multivariate spatial data when the temporal dimension is fixed.
+-->
+
+| No.
+| Module                  | Input             | Operation                                                                                                                                          | Output                                                              |
+|-----|-----------------|----------|--------------------|---------------------|
+| 1   | Temporal processing     | $x(s;t)$          | aaaaa                                                                                                                                              | $x(s; t^\prime)$                                                    |
+| 2   | Spatial processing      | $x(s;t)$          | aaaaa                                                                                                                                              |  $x(s^\prime; t)$                                                   |
+| 3   | Variable transformation | $x(s; t)$         | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????
+|
+| 4   | Scaling                 | $x(s; t)$         | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $v(s;t)$ where bbbbb                                                |
+| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $y \in \mathbb{R}^d$ and $d < p$            |
+| 6   | Distribution fit        | $x(s; t)$         | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1 |
+| 7   | Normalising             | $x(s; t)$         | $\Phi^{-1}[x(s, t)]$                                                                                                                               | $z(s; t)$ where $z \sim N(.,.)$                                     |
+| 8   | Benchmarking            | $x(s; t)$         |                                                                                                                                                    | $b(s;t)$                                                            |
+| 9   | Simplification          | $x(s; t)$         | ccccc                                                                                                                                              | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                              |
+
+: An notation overview of the input, operation, and output of each pipeline module.
+
+Some notes:
+
+-   still $x(s,t)$ notation because for all the other modules, $s$ and $t$ are all fixed, then $x(s, t)$ becomes $x$????
+-   aaaaa: just a general function over the input????
+-   bbbbb: input and output has the same distributional shape
+-   ccccc: latex issue
 
@@
 
-`Input`: $x_s(t)$,
+`Input`: $x(t)$, time series of a single variable with fixed locations
 
-`Output`:  $x^l_s(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.
+`Output`: $x(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.
 
@@
 
+```{=tex}
 \begin{equation}
@@
 \end{equation}
-
+```
 where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
@@
 \end{equation}
-
+```
 where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$.
@@
 \end{equation}
-
+```
 where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the coarser grid.
@@
 \end{equation}
-
+```
 where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation if any, and $d_{\tau}$ is the number of parameter of $\tau$.
@@
 \end{equation}
-
+```
 For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
@@
 <!-- Take one or multiple variables to create a new variable,  -->
+
 <!-- Restrict it to single variable, square root, log etc could be linearly, also non-linear -->
@@
 \end{equation}
-
+```
 where $p = 1, 2, \cdots, P$.
@@
 <!-- \begin{equation} -->
+
 <!-- \text{NDVI}(s;t) = \frac{x_1(s;t) - x_2(s;t)}{x_1(s;t) + x_2(s;t)}, -->
+
 <!-- \end{equation} -->
@@
 \end{equation}
-
+```
 where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
@@
 `Computation`: This step maps the univariate series into a different scale, typically for ease of comparison across regions.
-For example, a normal scale, [0, 1], or [0, 100] may be favored for reporting certain indexes. 
-In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. 
+For example, a normal scale, \[0, 1\], or \[0, 100\] may be favored for reporting certain indexes.
+In drought indexes, i.e.
+SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$.
 Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable.
@@
 `Computation`: Benchmarking sets a constant value to allow the constructed index to be compared across time.
-Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed. A benchmark value could be a constant or a function of the data, i.e. mean.
+Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed.
+A benchmark value could be a constant or a function of the data, i.e.
+mean.
 
@@
 
+```{=tex}
 \begin{equation}
 \begin{cases}
-c_0 & C_1 \leq x^l(s; t) < C_0 \\
-c_1 & C_2 \leq x^l(s; t) < C_1 \\
-c_2 & C_3 \leq x^l(s; t) < C_2 \\
+a_0 & C_1 \leq x(s; t) < C_0 \\
+a_1 & C_2 \leq x(s; t) < C_1 \\
+a_2 & C_3 \leq x(s; t) < C_2 \\
 \cdots \\
-c_z & C_z \leq x^l(s; t)
+a_z & C_z \leq x(s; t)
 \end{cases}
 \end{equation}
-
+```
 where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the thresholds for each category.
@@
 <!-- # 1. unpack or the argumebt s  -->
+
 <!-- # 2. construct the syntax to evalute: ~rescale_minmax(life_exp, min = xxx, max = xxxx) -->
+
 <!-- # 3. evalute  -->
+
 <!-- # 4. update the data object  -->
@@
 The function `compute_indexes()` can be used to collectively compute multiple indexes.
-The tidyindex offers wrapper functions, with the prefix idx_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function.
+The tidyindex offers wrapper functions, with the prefix idx\_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function.
 For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise).
@@
 
-````
     gggi %>% 
@@
           weight = weight))
-````
 
@@
 
-To understand the uncertainty of this dimension reduction step from combining four dimensions (Economy, Education, Health, and Politics), we run a local tour [TODO: reference to local tour] that varies the weight of each dimension slightly to observe how index value and country ranking changes.
+To understand the uncertainty of this dimension reduction step from combining four dimensions (Economy, Education, Health, and Politics), we run a local tour \[TODO: reference to local tour\] that varies the weight of each dimension slightly to observe how index value and country ranking changes.
 The local tour produces an animation that gradually increases the weight of one variable and reduces it back to equal weight, one at a time.
-A set of countries are selected for this exercise, which includes 1) the top four with GGGI > 0.85, 2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the bottom five countries with GGGI < 0.6.
+A set of countries are selected for this exercise, which includes 1) the top four with GGGI \> 0.85, 2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the bottom five countries with GGGI \< 0.6.
 Five frames (equal weights and one for each dimension with a relatively higher weight) are selected from the local tour animation to show in @fig-idx-tour and the full animation is available at https://vimeo.com/847874016.
@@
 
-* Bangladesh 
-* when decreasing the weight on politics, difference between countries narrower. 
-* increase politics -> all the country has lower index value
+-   Bangladesh
+-   when decreasing the weight on politics, difference between countries narrower.
+-   increase politics -\> all the country has lower index value
 
@@
 
-Future work:
-  - integrate more complex dimension reduction methods to calculate weights 
-  - strengthen the spatial processing module
+Future work: - integrate more complex dimension reduction methods to calculate weights - strengthen the spatial processing module
 

### 2023-08-14 tidyindex.tex
a table listing all the notations + visual mode reformat

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 Now, we introduce the notation used for describing pipeline modules.
-Consider multivariate spatio-temporal data,
+Consider multivariate spatio-temporal process,
 
 \begin{equation}
-\begin{aligned}
-\mathbf{x}(s;t) = [x^1(s;t), x^2(s;t), & \cdots, x^P(s;t)] \\
-& p = 1, 2, \cdots P, s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
-\end{aligned}
+\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
 \end{equation}
 
-where \(x^l\) is the \(p\)th variable, \(s\) is the spatial location,
-and \(t\) is the temporal order. The spatial space,
-\(D_s \subseteq \mathbb{R}^d\), is generic to include different spatial
-representations, including a 1D set of \(n\) locations
-\(\{D_s: s_1, s_2, \cdots, s_n\}\), a 2D set of longitude and latitude
-pairs, and \(d\)-D coordinate set. The higher dimension in the temporal
-space, \(D_t \subseteq \mathbb{R}^m\), can be used to represent for a
-\(m\)-D deconstruction of time. For example, a year-month combination,
-\((Y, M)^\prime\), can be used to record monthly precipitation series
-from January 1990 to December 2020, where \(Y = 1990, \cdots, 2020\) and
-\(M = 1, \cdots, 12\). To simplify the notation, a subscript letter
-\(s\) or \(t\) is used to denote the data have a fixed spatial locations
-or temporal order. When the spatial locations are fixed, the data become
-multivariate time series, denoted as \(x^l_s(t)\). Similarly,
-\(x^l_t(s)\) represents multivariate spatial data when the temporal
-dimension is fixed.
+where:
+
+\begin{itemize}
+\item
+  \(x\) is the variable in \(\mathbb{R}^p\),
+\item
+  \(s\) represents the geographic locations in the space
+  \(D_s \subseteq \mathbb{R}^d\). This includes a collection of
+  countries of interest and paired longitude and latitude coordinates of
+  locations, and
+\item
+  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^m\).
+  Sometimes, it can be time decomposed into several components, such as
+  year, month, quarter, and season.
+\end{itemize}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0685}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1370}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2740}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2877}}@{}}
+\caption{An notation overview of the input, operation, and output of
+each pipeline module.}\tabularnewline
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+No.
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Module
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Input
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Operation
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Output
+\end{minipage} \\
+\midrule()
+\endfirsthead
+\toprule()
+\begin{minipage}[b]{\linewidth}\raggedright
+No.
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Module
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Input
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Operation
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Output
+\end{minipage} \\
+\midrule()
+\endhead
+1 & Temporal processing & \(x(s;t)\) & aaaaa & \(x(s; t^\prime)\) \\
+2 & Spatial processing & \(x(s;t)\) & aaaaa & \(x(s^\prime; t)\) \\
+3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
+???? \\
+4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(v(s;t)\)
+where bbbbb \\
+5 & Dimension reduction & \(\mathbf{x}(s;t)\) & aaaaa &
+\(\mathbf{y}(s;t)\) where \(y \in \mathbb{R}^d\) and \(d < p\) \\
+6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
+\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
+distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
+\(p(.)\) denotes the probability value between 0 and 1 \\
+7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
+where \(z \sim N(.,.)\) \\
+8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
+9 & Simplification & \(x(s; t)\) & ccccc &
+\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
+\bottomrule()
+\end{longtable}
+
+Some notes:
+
+\begin{itemize}
+\tightlist
+\item
+  still \(x(s,t)\) notation because for all the other modules, \(s\) and
+  \(t\) are all fixed, then \(x(s, t)\) becomes \(x\)????
+\item
+  aaaaa: just a general function over the input????
+\item
+  bbbbb: input and output has the same distributional shape
+\item
+  ccccc: latex issue
+\end{itemize}
 
@@
 
-\texttt{Input}: \(x_s(t)\),
-
-\texttt{Output}: \(x^l_s(t^{\prime})\), where
+\texttt{Input}: \(x(t)\), time series of a single variable with fixed
+locations
+
+\texttt{Output}: \(x(t^{\prime})\), where
 \(t^\prime \in D_{t^{\prime}}\).
@@
 \begin{cases}
-c_0 & C_1 \leq x^l(s; t) < C_0 \\
-c_1 & C_2 \leq x^l(s; t) < C_1 \\
-c_2 & C_3 \leq x^l(s; t) < C_2 \\
+a_0 & C_1 \leq x(s; t) < C_0 \\
+a_1 & C_2 \leq x(s; t) < C_1 \\
+a_2 & C_3 \leq x(s; t) < C_2 \\
 \cdots \\
-c_z & C_z \leq x^l(s; t)
+a_z & C_z \leq x(s; t)
 \end{cases}

### 2023-08-15 tidyindex.qmd
rework on temporal processing

@@
 
--   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. Sometimes, it can be time decomposed into several components, such as year, month, quarter, and season.
+-   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. Sometimes, it can be time breaks into other temporal components, such as year, month, quarter, and season.
 
@@
 | Module                  | Input             | Operation                                                                                                                                          | Output                                                              |
-|-----|-----------------|----------|--------------------|---------------------|
-| 1   | Temporal processing     | $x(s;t)$          | aaaaa                                                                                                                                              | $x(s; t^\prime)$                                                    |
-| 2   | Spatial processing      | $x(s;t)$          | aaaaa                                                                                                                                              |  $x(s^\prime; t)$                                                   |
+|-------------|-------------|-------------|-----------------|-----------------|
+| 1   | Temporal processing     | $x(t)$            | $f[x(.)]$                                                                                                                                          | $x^*(t^\prime)$ where $t\prime \in D_{t^\prime}$                    |
+| 2   | Spatial processing      | $x(s)$            | aaaaa                                                                                                                                              | $x^\#(s^\prime)$                                                    |
 | 3   | Variable transformation | $x(s; t)$         | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????
@@
 
-`Output`: $x(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.
+`Output`: $x^*(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.
 
-`Computation`: The construction of an index sometimes needs to consider information from neighbouring time periods.
-The temporal processing is a general operator on the time dimension of the data in the form of
+`Computation`: The temporal processing module takes a single variable in a set of fixed locations, $x(t)$, as the input.
+This step *transforms* the original series into a new series to incorporate the temporal neighbouring effect: $x^*(.) = f[x(.)]$.
+In temporal processing, the temporal dimension $t$ may also transition to $t^\prime \subseteq D_{t^\prime}$, allowing for the decomposition of time into its constituent temporal components.
+For example, this can be used to break down daily data into monthly.
+This gives the output of the temporal processing module as $x^*(t^\prime)$.
 
-```{=tex}
-\begin{equation}
-x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
-\end{equation}
-```
-where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.
-A typical example of temporal processing is aggregation and in the drought community, these indexes are often called multi-scalar indexes.
-For example, Standardized Precipitation Index (SPI) aggregates monthly precipitation using a time scale parameter $k$ in the form of
+An example of temporal processing is in Standardized Precipitation Index (SPI), where monthly precipitation is summed over a rolling window of size $k$, also known as the time scale.
+Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be written as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.
+For SPI, the choice of the time scale $k$, is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).
 
-$$x(s_i;t_{j^\prime}) = \sum_{j = j^\prime}^{j^\prime + k - 1}x(s_i; t_j),$$
+<!-- The construction of an index sometimes needs to consider information from neighbouring time periods.
+The temporal processing is a general operator on the time dimension of the data in the form of -->
 
-where $j^\prime$ is the new time index after the aggregation.
-Multi-scalar drought indexes are widely used to study different types of drought events.
-A smaller value of the time scales parameter $k$ is used for short term drought events, while a large $k$ of 18 or 24, an equivalent to a 1.5- or 2-year aggregation, is often used for long term assessment.
+<!-- ```{=tex} -->
+
+<!-- \begin{equation} -->
+
+<!-- x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})), -->
+
+<!-- \end{equation} -->
+
+<!-- ``` -->
+
+<!-- where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. -->
 

### 2023-08-15 tidyindex.tex
rework on temporal processing

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \begin{equation}
-\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
+\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
 \end{equation}
@@
   \(s\) represents the geographic locations in the space
-  \(D_s \subseteq \mathbb{R}^d\). This includes a collection of
+  \(D_s \subseteq \mathbb{R}^m\). This includes a collection of
   countries of interest and paired longitude and latitude coordinates of
@@
 \item
-  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^m\).
-  Sometimes, it can be time decomposed into several components, such as
-  year, month, quarter, and season.
+  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
+  Sometimes, it can be time breaks into other temporal components, such
+  as year, month, quarter, and season.
 \end{itemize}
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0685}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
   >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1370}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2740}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2877}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}@{}}
 \caption{An notation overview of the input, operation, and output of
@@
 \endhead
-1 & Temporal processing & \(x(s;t)\) & aaaaa & \(x(s; t^\prime)\) \\
-2 & Spatial processing & \(x(s;t)\) & aaaaa & \(x(s^\prime; t)\) \\
+1 & Temporal processing & \(x(t)\) & \(f[x(.)]\) & \(x^*(t^\prime)\)
+where \(t\prime \in D_{t^\prime}\) \\
+2 & Spatial processing & \(x(s)\) & aaaaa & \(x^\#(s^\prime)\) \\
 3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
@@
 
-\texttt{Output}: \(x(t^{\prime})\), where
+\texttt{Output}: \(x^*(t^{\prime})\), where
 \(t^\prime \in D_{t^{\prime}}\).
 
-\texttt{Computation}: The construction of an index sometimes needs to
-consider information from neighbouring time periods. The temporal
-processing is a general operator on the time dimension of the data in
-the form of
-
-\begin{equation}
-x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
-\end{equation}
-
-where \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the
-parameters associated with the temporal operation and \(d_{\psi}\) is
-the number of parameter of \(\psi\). A typical example of temporal
-processing is aggregation and in the drought community, these indexes
-are often called multi-scalar indexes. For example, Standardized
-Precipitation Index (SPI) aggregates monthly precipitation using a time
-scale parameter \(k\) in the form of
-
-\[x(s_i;t_{j^\prime}) = \sum_{j = j^\prime}^{j^\prime + k - 1}x(s_i; t_j),\]
-
-where \(j^\prime\) is the new time index after the aggregation.
-Multi-scalar drought indexes are widely used to study different types of
-drought events. A smaller value of the time scales parameter \(k\) is
-used for short term drought events, while a large \(k\) of 18 or 24, an
-equivalent to a 1.5- or 2-year aggregation, is often used for long term
-assessment.
+\texttt{Computation}: The temporal processing module takes a single
+variable in a set of fixed locations, \(x(t)\), as the input. This step
+\emph{transforms} the original series into a new series to incorporate
+the temporal neighbouring effect: \(x^*(.) = f[x(.)]\). In temporal
+processing, the temporal dimension \(t\) may also transition to
+\(t^\prime \subseteq D_{t^\prime}\), allowing for the decomposition of
+time into its constituent temporal components. For example, this can be
+used to break down daily data into monthly. This gives the output of the
+temporal processing module as \(x^*(t^\prime)\).
+
+An example of temporal processing is in Standardized Precipitation Index
+(SPI), where monthly precipitation is summed over a rolling window of
+size \(k\), also known as the time scale. Assuming an original time
+series indexed by \(t= 1, 2, \cdots, T\), the aggregated series, indexed
+by \(t^\prime = 1, 2, \cdots, T-k+1\), can be written as
+\(x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)\). For SPI,
+the choice of the time scale \(k\), is used to control the accumulation
+period for the water deficit, enabling the assessment of drought
+severity across various types (meteorological, agricultural, and
+hydrological).
 

### 2023-08-15 tidyindex.qmd
rework on spatial processing and dimension reduction + proofread

@@
 | Module                  | Input                                                       | Operation                                                                                                                                          | Output                                                                  |
-|-------------|-------------|-------------|-----------------|-----------------|
-| 1   | Temporal processing     | $x(t)$            | $f[x(.)]$                                                                                                                                          | $x^*(t^\prime)$ where $t\prime \in D_{t^\prime}$                    |
-| 2   | Spatial processing      | $x(s)$            | aaaaa                                                                                                                                              | $x^\#(s^\prime)$                                                    |
+|---------------|---------------|---------------|---------------|---------------|
+| 1   | Temporal processing     | $x(t)$                                                      | $f_{\Psi}[x(.)]$                                                                                                                                   | $x^*(t^\prime)$ where $t\prime \in D_{t^\prime}$                        |
+| 2   | Spatial processing      | $x(s)$                                                      | $g_{\Theta}[x(.)]$                                                                                                                                 | $x^\#(s^\prime)$ where $s^\prime \in D_{s^\prime}$                      |
 | 3   | Variable transformation | $x(s; t)$                                                   | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????
@@
 | 4   | Scaling                 | $x(s; t)$                                                   | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $v(s;t)$ where bbbbb                                                    |
-| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $y \in \mathbb{R}^d$ and $d < p$            |
+| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^p$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$ |
 | 6   | Distribution fit        | $x(s; t)$                                                   | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1     |
@@
 
-`Computation`: The temporal processing module takes a single variable in a set of fixed locations, $x(t)$, as the input.
-This step *transforms* the original series into a new series to incorporate the temporal neighbouring effect: $x^*(.) = f[x(.)]$.
-In temporal processing, the temporal dimension $t$ may also transition to $t^\prime \subseteq D_{t^\prime}$, allowing for the decomposition of time into its constituent temporal components.
-For example, this can be used to break down daily data into monthly.
-This gives the output of the temporal processing module as $x^*(t^\prime)$.
+`Computation`: The temporal processing module takes a single variable in a set of fixed locations, $x(t)$, as input.
+This step transforms the original series into a new series to incorporate the temporal neighbouring effect: $x^*(.) = f_{\Psi}[x(.)]$, where $\Psi$ is a vector of hyper-parameters associated with the transformation.
+In temporal processing, the transition of temporal dimension from $t$ to $t^\prime \subseteq D_{t^\prime}$ allows for the decomposition of time into its constituent temporal components.
+For example, to break down daily data into monthly.
 
-An example of temporal processing is in Standardized Precipitation Index (SPI), where monthly precipitation is summed over a rolling window of size $k$, also known as the time scale.
-Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be written as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.
-For SPI, the choice of the time scale $k$, is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).
+In the Standardized Precipitation Index (SPI), temporal processing is used to sum the monthly precipitation over a rolling window of size $k$, also known as the time scale, prior to the distribution fitting and normalising module.
+Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.
+For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).
 
@@
 
-`Input`: $x^l_t(s)$
+`Input`: $x(s)$
 
-`Output`: $x^l_t(s^\prime)$ where $s^\prime \in D_{s^{\prime}}$
+`Output`: $x^\#(s^\prime)$ where $s^\prime \in D_{s^{\prime}}$
 
-`Computation`: Spatial processing may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing.
-The process can be written as a general operation in the form of
+`Computation`: The spatial processing module takes a single variable with a fixed temporal dimension, $x(s)$, as input.
+This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(.) = g_{\Theta}(x(.))$, where $\Theta$ is a vector of hyper-parameters associated with the transformation.
+The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions.
+This also includes the aggregation of variables into different levels, such as city, state, and countries scales.
 
-```{=tex}
-\begin{equation}
-x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
-\end{equation}
-```
-where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$.
+<!-- may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing.
+The process can be written as a general operation in the form of -->
+
+<!-- ```{=tex} -->
+
+<!-- \begin{equation} -->
+
+<!-- x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})), -->
+
+<!-- \end{equation} -->
+
+<!-- ``` -->
+
+<!-- where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$.
 An example of spatial processing is to align variables collected in different resolutions.
 When variables are collected at different resolutions, analysts may choose to down-sample those in a finer resolution, $i$, to match those in a coarser resolution, $i^\prime$.
-This is a spatial aggregation and if aggregate using the mean, it can be written as
+This is a spatial aggregation and if aggregate using the mean, it can be written as -->
 
-```{=tex}
-\begin{equation}
-g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}},
-\end{equation}
-```
-where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the coarser grid.
+<!-- ```{=tex} -->
+
+<!-- \begin{equation} -->
+
+<!-- g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}}, -->
+
+<!-- \end{equation} -->
+
+<!-- ``` -->
+
+<!-- where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the coarser grid.
 Other examples of spatial processing include 1) borrowing information from neighbouring spatial locations to interpolate unobserved locations and 2) fusing variables from ground measures with satellite imageries.
+-->
 
@@
 
-`Input`: $x^l(s; t)$
+`Input`: $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^p$
 
-`Output`: $x^{l^\prime}(s; t)$, where $l^{\prime} = 1, 2, \cdots, L^{\prime}$
+`Output`: $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$
 
-`Computation`: Dimension reduction summarises the multivariate information into univariate, which can be denoted as:
+`Computation`: Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$ , as the input.
+It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output.
 
-```{=tex}
-\begin{equation}
-x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
-\end{equation}
-```
-where $p = 1, 2, \cdots, P$.
-The combination can be based on domain-specific knowledge, originated from theories describing the underlying physical process.
-For example, the SPEI uses a water balance model ($D = P - \text{PET}$) to calculate the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$).
+The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods.
+The Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
+This is the only step that differentiates it from the Standardized Precipitation Index (SPI).
 
-Another widely used approach is linear combination, which aggregates a collection of variables in a linear additive structure, expressed as: $$x(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),$$
-
-where $\lambda_p$ denotes the weight assigned to variable $x_p$.
-Most indexes uses an equal weight that sum to 1, which allows each variable to contribute equally to the index.
-A linear combination can also be viewed as a linear projection of multivariate information by the weight vector.
-Projecting data from higher to lower dimension inevitably leads to information loss.
-For example, two countries can receive similar index, but their components could significantly differ -- one with average scores across all components and another with extreme scores the opposite ends.
-The selected set of weights need to be examined to understand its effect on the index and its implications for decision-making.
-To do this, analysts can vary the coefficients in the linear projection to observe how the index value and countries' ranking change.
-These changes can also be visualized using a method called tour by generating an animation of the projections among different sets of weights.
+When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques.
+Many social indexes use a linear combination in an additive structure, $y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),$ where $\lambda_i$ is the weight assigned to each individual variable $x_i$.
+A linear combination can also be preceived as a linear projection of multivariate information through the weight vector.
+Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making.
+One can visualise the data projection across different sets of weights through animation using a method called tour.
 

### 2023-08-15 tidyindex.tex
rework on spatial processing and dimension reduction + proofread

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, frame hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, sharp corners, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
 \caption{An notation overview of the input, operation, and output of
@@
 \endhead
-1 & Temporal processing & \(x(t)\) & \(f[x(.)]\) & \(x^*(t^\prime)\)
-where \(t\prime \in D_{t^\prime}\) \\
-2 & Spatial processing & \(x(s)\) & aaaaa & \(x^\#(s^\prime)\) \\
+1 & Temporal processing & \(x(t)\) & \(f_{\Psi}[x(.)]\) &
+\(x^*(t^\prime)\) where \(t\prime \in D_{t^\prime}\) \\
+2 & Spatial processing & \(x(s)\) & \(g_{\Theta}[x(.)]\) &
+\(x^\#(s^\prime)\) where \(s^\prime \in D_{s^\prime}\) \\
 3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
@@
 where bbbbb \\
-5 & Dimension reduction & \(\mathbf{x}(s;t)\) & aaaaa &
-\(\mathbf{y}(s;t)\) where \(y \in \mathbb{R}^d\) and \(d < p\) \\
+5 & Dimension reduction & \(\mathbf{x}(s;t)\) where
+\(\mathbf{x} \subseteq \mathbb{R}^p\) & aaaaa & \(\mathbf{y}(s;t)\)
+where \(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\) \\
 6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
@@
 \texttt{Computation}: The temporal processing module takes a single
-variable in a set of fixed locations, \(x(t)\), as the input. This step
-\emph{transforms} the original series into a new series to incorporate
-the temporal neighbouring effect: \(x^*(.) = f[x(.)]\). In temporal
-processing, the temporal dimension \(t\) may also transition to
-\(t^\prime \subseteq D_{t^\prime}\), allowing for the decomposition of
-time into its constituent temporal components. For example, this can be
-used to break down daily data into monthly. This gives the output of the
-temporal processing module as \(x^*(t^\prime)\).
-
-An example of temporal processing is in Standardized Precipitation Index
-(SPI), where monthly precipitation is summed over a rolling window of
-size \(k\), also known as the time scale. Assuming an original time
-series indexed by \(t= 1, 2, \cdots, T\), the aggregated series, indexed
-by \(t^\prime = 1, 2, \cdots, T-k+1\), can be written as
+variable in a set of fixed locations, \(x(t)\), as input. This step
+transforms the original series into a new series to incorporate the
+temporal neighbouring effect: \(x^*(.) = f_{\Psi}[x(.)]\), where
+\(\Psi\) is a vector of hyper-parameters associated with the
+transformation. In temporal processing, the transition of temporal
+dimension from \(t\) to \(t^\prime \subseteq D_{t^\prime}\) allows for
+the decomposition of time into its constituent temporal components. For
+example, to break down daily data into monthly.
+
+In the Standardized Precipitation Index (SPI), temporal processing is
+used to sum the monthly precipitation over a rolling window of size
+\(k\), also known as the time scale, prior to the distribution fitting
+and normalising module. Assuming an original time series indexed by
+\(t= 1, 2, \cdots, T\), the aggregated precipitation series, indexed by
+\(t^\prime = 1, 2, \cdots, T-k+1\), can be expressed as
 \(x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)\). For SPI,
-the choice of the time scale \(k\), is used to control the accumulation
+the choice of the time scale \(k\) is used to control the accumulation
 period for the water deficit, enabling the assessment of drought
@@
 
-\texttt{Input}: \(x^l_t(s)\)
-
-\texttt{Output}: \(x^l_t(s^\prime)\) where
+\texttt{Input}: \(x(s)\)
+
+\texttt{Output}: \(x^\#(s^\prime)\) where
 \(s^\prime \in D_{s^{\prime}}\)
 
-\texttt{Computation}: Spatial processing may be needed when indexes are
-not calculated independently on each collected location or when
-variables collected from multiple sources need to be fused before
-further processing. The process can be written as a general operation in
-the form of
-
-\begin{equation}
-x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),
-\end{equation}
-
-where \(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the
-associated parameters in the process and \(d_{\theta}\) is the number of
-parameter of \(\theta\). An example of spatial processing is to align
-variables collected in different resolutions. When variables are
-collected at different resolutions, analysts may choose to down-sample
-those in a finer resolution, \(i\), to match those in a coarser
-resolution, \(i^\prime\). This is a spatial aggregation and if aggregate
-using the mean, it can be written as
-
-\begin{equation}
-g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}},
-\end{equation}
-
-where \(i \in i^\prime\) includes all the cells from the finer
-resolution in the coarser grid and \(n_{i^\prime}\) is the number of
-observations falls into the coarser grid. Other examples of spatial
-processing include 1) borrowing information from neighbouring spatial
-locations to interpolate unobserved locations and 2) fusing variables
-from ground measures with satellite imageries.
+\texttt{Computation}: The spatial processing module takes a single
+variable with a fixed temporal dimension, \(x(s)\), as input. This step
+transforms the variable from the original spatial dimension \(s\) into
+the new dimension \(s^\prime \in D_{s^\prime}\) through
+\(x^\#(.) = g_{\Theta}(x(.))\), where \(\Theta\) is a vector of
+hyper-parameters associated with the transformation. The change of
+spatial dimension allows for the alignment of variables collected from
+different measurements, such as in-situ stations and satellite imagery,
+or originating from different resolutions. This also includes the
+aggregation of variables into different levels, such as city, state, and
+countries scales.
 
@@
 
-\texttt{Input}: \(x^l(s; t)\)
-
-\texttt{Output}: \(x^{l^\prime}(s; t)\), where
-\(l^{\prime} = 1, 2, \cdots, L^{\prime}\)
-
-\texttt{Computation}: Dimension reduction summarises the multivariate
-information into univariate, which can be denoted as:
-
-\begin{equation}
-x_p(\mathbf{s}; \mathbf{t}) \rightarrow x(\mathbf{s}; \mathbf{t}),
-\end{equation}
-
-where \(p = 1, 2, \cdots, P\). The combination can be based on
-domain-specific knowledge, originated from theories describing the
-underlying physical process. For example, the SPEI uses a water balance
-model (\(D = P - \text{PET}\)) to calculate the difference series
-(\(D\)) from precipitation (\(P\)) and potential evapotranspiration
-(\(\text{PET}\)).
-
-Another widely used approach is linear combination, which aggregates a
-collection of variables in a linear additive structure, expressed as:
-\[x(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t}),\]
-
-where \(\lambda_p\) denotes the weight assigned to variable \(x_p\).
-Most indexes uses an equal weight that sum to 1, which allows each
-variable to contribute equally to the index. A linear combination can
-also be viewed as a linear projection of multivariate information by the
-weight vector. Projecting data from higher to lower dimension inevitably
-leads to information loss. For example, two countries can receive
-similar index, but their components could significantly differ -- one
-with average scores across all components and another with extreme
-scores the opposite ends. The selected set of weights need to be
-examined to understand its effect on the index and its implications for
-decision-making. To do this, analysts can vary the coefficients in the
-linear projection to observe how the index value and countries' ranking
-change. These changes can also be visualized using a method called tour
-by generating an animation of the projections among different sets of
-weights.
+\texttt{Input}: \(\mathbf{x}(s;t)\) where
+\(\mathbf{x} \subseteq \mathbb{R}^p\)
+
+\texttt{Output}: \(\mathbf{y}(s;t)\) where
+\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\)
+
+\texttt{Computation}: Dimension reduction takes the multivariate
+information \(\mathbf{x}(s;t)\), where
+\(\mathbf{x} \subseteq \mathbb{R}^p\) , as the input. It summarises the
+high-dimensional information into a lower dimension representation
+\(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\) and \(d < p\),
+as the output.
+
+The transformation can be based on domain-specific knowledge,
+originating from theories describing the underlying physical processes,
+or guided in statistical methods. The Standardized
+Precipitation-Evapotranspiration Index (SPEI) calculates the difference
+series (\(D\)) from precipitation (\(P\)) and potential
+evapotranspiration (\(\text{PET}\)), using a water balance model
+(\(D = P - \text{PET}\)). This is the only step that differentiates it
+from the Standardized Precipitation Index (SPI).
+
+When a physical process is absent to describe the phenomenon,
+multivariate data are often combined using statistical methods, which
+can be further classified into linear and non-linear techniques. Many
+social indexes use a linear combination in an additive structure,
+\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
+is the weight assigned to each individual variable \(x_i\). A linear
+combination can also be preceived as a linear projection of multivariate
+information through the weight vector. Projecting data from higher to
+lower dimension unavoidably leads to information loss and the used
+weights need careful examination to understand their effects on the
+index and implications for interpretation and decision-making. One can
+visualise the data projection across different sets of weights through
+animation using a method called tour.
 

### 2023-08-15 tidyindex.tex
Update notation sec 3

@@
 \item
-  \(x\) is the variable in \(\mathbb{R}^p\),
-\item
-  \(s\) represents the geographic locations in the space
-  \(D_s \subseteq \mathbb{R}^m\). This includes a collection of
-  countries of interest and paired longitude and latitude coordinates of
-  locations, and
-\item
-  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
-  Sometimes, it can be time breaks into other temporal components, such
-  as year, month, quarter, and season.
+  \(x^j(s, j)\) represents a  variable of interest for example precipitaton, $j = 1, \ldots , p$ and 
+\item
+  \(s\) represents the geographic location in the space
+  \(D_s \subseteq \mathbb{R}^m\). Examples of geographic locations include a collection of
+  countries, longitude and latitude coordinates or regions of interest and, 
+\item
+  \(t\) denotes the temporal measurements in \(D_t \subseteq \mathbb{R}^n\).
+  For instance, time measurements could be recorded hourly, yearly, monthly, quaterly or by season.
 \end{itemize}

### 2023-08-15 tidyindex.qmd
migrate changes from .text to qmd

@@
 
--   $x$ is the variable in $\mathbb{R}^p$,
+-   $x^i(s, t)$ represents a variable of interest for example precipitation,$j = 1, \cdots, p$ and
 
 -   $s$ represents the geographic locations in the space $D_s \subseteq \mathbb{R}^m$.
-This includes a collection of countries of interest and paired longitude and latitude coordinates of locations, and
+Examples of geographic locations include a collection of countries, longitude and latitude coordinates or regions of interest and,
 
--   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. Sometimes, it can be time breaks into other temporal components, such as year, month, quarter, and season.
+-   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. For instance, time measurements could be recorded hourly, yearly, monthly, quaterly or by season..
 
@@
 | Module                  | Input                                                       | Operation                                                                                                                                          | Output                                                                  |
-|---------------|---------------|---------------|---------------|---------------|
+|------------|------------|------------|--------------------------|------------|
 | 1   | Temporal processing     | $x(t)$                                                      | $f_{\Psi}[x(.)]$                                                                                                                                   | $x^*(t^\prime)$ where $t\prime \in D_{t^\prime}$                        |

### 2023-08-15 tidyindex.tex
migrate changes from .text to qmd

--- 2023-08-15 tidyindex.tex
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, sharp corners, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, breakable, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 \item
-  \(x\) is the variable in \(\mathbb{R}^p\),
+  \(x^i(s, t)\) represents a variable of interest for example
+  precipitation,\(j = 1, \cdots, p\) and
 \item
   \(s\) represents the geographic locations in the space
-  \(D_s \subseteq \mathbb{R}^m\). This includes a collection of
-  countries of interest and paired longitude and latitude coordinates of
-  locations, and
+  \(D_s \subseteq \mathbb{R}^m\). Examples of geographic locations
+  include a collection of countries, longitude and latitude coordinates
+  or regions of interest and,
 \item
   \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
-  Sometimes, it can be time breaks into other temporal components, such
-  as year, month, quarter, and season.
+  For instance, time measurements could be recorded hourly, yearly,
+  monthly, quaterly or by season..
 \end{itemize}
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3514}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}@{}}
 \caption{An notation overview of the input, operation, and output of

### 2023-08-15 tidyindex.qmd
Update table and section 3

@@
 
--   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. For instance, time measurements could be recorded hourly, yearly, monthly, quaterly or by season..
+-   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. For instance, time measurements could be recorded hourly, yearly, monthly, quaterly or by season.
+
+In what follows when geographical or temporal components of the $x(s,t)$ process are fixed we will be using suffix notation.
+For example, $x_{s}(t)$ represents the data for a fixed location $s$ as a function of time $t$.
+While $x_{t}(s)$ denotes the spatial varying procces for a fixed $t$.
 
@@
 | Module                  | Input                                                                  | Operation                                                                                                                                          | Output                                                                 |
-|------------|------------|------------|--------------------------|------------|
-| 1   | Temporal processing     | $x(t)$                                                      | $f_{\Psi}[x(.)]$                                                                                                                                   | $x^*(t^\prime)$ where $t\prime \in D_{t^\prime}$                        |
-| 2   | Spatial processing      | $x(s)$                                                      | $g_{\Theta}[x(.)]$                                                                                                                                 | $x^\#(s^\prime)$ where $s^\prime \in D_{s^\prime}$                      |
-| 3   | Variable transformation | $x(s; t)$                                                   | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????
+|-------------|-------------|-------------|--------------------|-------------|
+| 1   | Temporal processing     | $x_{s}(t)$                                                             | $f_{\Psi}[x(.)]$                                                                                                                                   | $\tilde{x}^i(t^\prime)$ where $t\prime \in D_{t^\prime}$                       |
+| 2   | Spatial processing      | $x_t(s)$                                                               | $g[x(.)]$                                                                                                                                 | $x^i(s^\prime)$ where $s^\prime \in D_{s^\prime}$                     |
+| 3   | Variable transformation | $x(s; t)$                                                              | $T(x(s;t))$                                                                                                                                        | $Tx^i_t(s)$ ????
 |
-| 4   | Scaling                 | $x(s; t)$                                                   | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $v(s;t)$ where bbbbb                                                    |
-| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^p$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$ |
+| 4   | Scaling                 | $x(s; t)$                                                              | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $Sx^i(s;t)$ where bbbbb                                                   |
+| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and$d < q$ |
 | 6   | Distribution fit        | $x(s; t)$                                                              | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1    |
@@
 
-`Computation`: The temporal processing module takes a single variable in a set of fixed locations, $x(t)$, as input.
-This step transforms the original series into a new series to incorporate the temporal neighbouring effect: $x^*(.) = f_{\Psi}[x(.)]$, where $\Psi$ is a vector of hyper-parameters associated with the transformation.
+`Computation`: The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$  as a function of time.
+In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$,  $\tilde{x}_s^i(t^') = f[[x_s(t)]$ where $t^'$ refers to the new temporal resolution after aggregation.
+An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$.
+That is also known as the time scale.
+<!-- Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.
+-->
+For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).
+
+<!-- to incorporate the temporal neighbouring effect: $x^*(.) = f_{\Psi}[x(.)]$, where $\Psi$ is a vector of hyper-parameters associated with the transformation.
 In temporal processing, the transition of temporal dimension from $t$ to $t^\prime \subseteq D_{t^\prime}$ allows for the decomposition of time into its constituent temporal components.
-For example, to break down daily data into monthly.
-
-In the Standardized Precipitation Index (SPI), temporal processing is used to sum the monthly precipitation over a rolling window of size $k$, also known as the time scale, prior to the distribution fitting and normalising module.
-Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.
-For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).
+For example,  -->
 
@@
 The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods.
-The Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference series ($D$) from precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
-This is the only step that differentiates it from the Standardized Precipitation Index (SPI).
+For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
+This is the only step that differs from the Standardized Precipitation Index (SPI).
 
+(Move this to the example and remove notation)
 When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques.
@@
 
+```         
 gggi %>% 
@@
       weight = weight)) 
+```
 

### 2023-08-15 tidyindex.qmd
knit

@@
 `Computation`: The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$  as a function of time.
-In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$,  $\tilde{x}_s^i(t^') = f[[x_s(t)]$ where $t^'$ refers to the new temporal resolution after aggregation.
+In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$,  $\tilde{x}_s^i(t^\prime) = f[[x_s(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation.
 An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$.

### 2023-08-15 tidyindex.tex
knit

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced, frame hidden]}{\end{tcolorbox}}\fi
 
@@
   For instance, time measurements could be recorded hourly, yearly,
-  monthly, quaterly or by season..
+  monthly, quaterly or by season.
 \end{itemize}
 
+In what follows when geographical or temporal components of the
+\(x(s,t)\) process are fixed we will be using suffix notation. For
+example, \(x_{s}(t)\) represents the data for a fixed location \(s\) as
+a function of time \(t\). While \(x_{t}(s)\) denotes the spatial varying
+procces for a fixed \(t\).
+
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3514}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1622}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}@{}}
 \caption{An notation overview of the input, operation, and output of
@@
 \endhead
-1 & Temporal processing & \(x(t)\) & \(f_{\Psi}[x(.)]\) &
-\(x^*(t^\prime)\) where \(t\prime \in D_{t^\prime}\) \\
-2 & Spatial processing & \(x(s)\) & \(g_{\Theta}[x(.)]\) &
-\(x^\#(s^\prime)\) where \(s^\prime \in D_{s^\prime}\) \\
-3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
-???? \\
-4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(v(s;t)\)
+1 & Temporal processing & \(x_{s}(t)\) & \(f_{\Psi}[x(.)]\) &
+\(\tilde{x}^i(t^\prime)\) where \(t\prime \in D_{t^\prime}\) \\
+2 & Spatial processing & \(x_t(s)\) & \(g[x(.)]\) & \(x^i(s^\prime)\)
+where \(s^\prime \in D_{s^\prime}\) \\
+3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) &
+\(Tx^i_t(s)\) ???? \\
+4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(Sx^i(s;t)\)
 where bbbbb \\
 5 & Dimension reduction & \(\mathbf{x}(s;t)\) where
-\(\mathbf{x} \subseteq \mathbb{R}^p\) & aaaaa & \(\mathbf{y}(s;t)\)
-where \(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\) \\
+\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) & aaaaa &
+\(\mathbf{y}(s;t)\) where \(\mathbf{y} \subseteq \mathbb{R}^d\)
+and\(d < q\) \\
 6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
@@
 
-\texttt{Computation}: The temporal processing module takes a single
-variable in a set of fixed locations, \(x(t)\), as input. This step
-transforms the original series into a new series to incorporate the
-temporal neighbouring effect: \(x^*(.) = f_{\Psi}[x(.)]\), where
-\(\Psi\) is a vector of hyper-parameters associated with the
-transformation. In temporal processing, the transition of temporal
-dimension from \(t\) to \(t^\prime \subseteq D_{t^\prime}\) allows for
-the decomposition of time into its constituent temporal components. For
-example, to break down daily data into monthly.
-
-In the Standardized Precipitation Index (SPI), temporal processing is
-used to sum the monthly precipitation over a rolling window of size
-\(k\), also known as the time scale, prior to the distribution fitting
-and normalising module. Assuming an original time series indexed by
-\(t= 1, 2, \cdots, T\), the aggregated precipitation series, indexed by
-\(t^\prime = 1, 2, \cdots, T-k+1\), can be expressed as
-\(x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)\). For SPI,
-the choice of the time scale \(k\) is used to control the accumulation
-period for the water deficit, enabling the assessment of drought
-severity across various types (meteorological, agricultural, and
+\texttt{Computation}: The temporal processing module takes as input
+argument a single variable \(x_s^i(t)\) at location \(s\) as a function
+of time. In this step the original time series can be transformed or
+summarized into a new one via time aggregation, the transformation is
+represented by the function \(f\),
+\(\tilde{x}_s^i(t^\prime) = f[[x_s(t)]\) where \(t^\prime\) refers to
+the new temporal resolution after aggregation. An example of a temporal
+processing done in the computation of the Standardized Precipitation
+Index (SPI), consists of summing the monthly precipitation series over a
+rolling time window of size \(k\). That is also known as the time scale.
+For SPI, the choice of the time scale \(k\) is used to control the
+accumulation period for the water deficit, enabling the assessment of
+drought severity across various types (meteorological, agricultural, and
 hydrological).
@@
 originating from theories describing the underlying physical processes,
-or guided in statistical methods. The Standardized
+or guided in statistical methods. For example, the Standardized
 Precipitation-Evapotranspiration Index (SPEI) calculates the difference
-series (\(D\)) from precipitation (\(P\)) and potential
-evapotranspiration (\(\text{PET}\)), using a water balance model
-(\(D = P - \text{PET}\)). This is the only step that differentiates it
-from the Standardized Precipitation Index (SPI).
-
-When a physical process is absent to describe the phenomenon,
-multivariate data are often combined using statistical methods, which
-can be further classified into linear and non-linear techniques. Many
-social indexes use a linear combination in an additive structure,
+\(D\) between precipitation (\(P\)) and potential evapotranspiration
+(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
+This is the only step that differs from the Standardized Precipitation
+Index (SPI).
+
+(Move this to the example and remove notation) When a physical process
+is absent to describe the phenomenon, multivariate data are often
+combined using statistical methods, which can be further classified into
+linear and non-linear techniques. Many social indexes use a linear
+combination in an additive structure,
 \(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)

### 2023-08-15 tidyindex.qmd
clean up after working with Patricia

--- 2023-08-15 tidyindex.qmd
@@
 
-In what follows when geographical or temporal components of the $x(s,t)$ process are fixed we will be using suffix notation.
+In what follows when geographic or temporal components of the $x(s,t)$ process are fixed we will be using suffix notation.
 For example, $x_{s}(t)$ represents the data for a fixed location $s$ as a function of time $t$.
-While $x_{t}(s)$ denotes the spatial varying procces for a fixed $t$.
-
-<!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.-->
-
-<!--For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$.
-.-->
-
-<!-- To simplify the notation, a subscript letter $s$ or $t$ is used to denote the data have a fixed spatial locations or temporal order.
-When the spatial locations are fixed, the data become multivariate time series, denoted as $x^l_s(t)$.
-Similarly,  $x^l_t(s)$ represents multivariate spatial data when the temporal dimension is fixed.
--->
+While $x_{t}(s)$ denotes the spatial varying proccess for a fixed $t$.
+An overview of the notation for pipeline input, operation, and output is present in @tbl-notation.
 
@@
 | Module                  | Input                                                                  | Operation                                                                         | Output                                                                 |
-|-------------|-------------|-------------|--------------------|-------------|
-| 1   | Temporal processing     | $x_{s}(t)$                                                             | $f_{\Psi}[x(.)]$                                                                                                                                   | $\tilde{x}^i(t^\prime)$ where $t\prime \in D_{t^\prime}$                       |
-| 2   | Spatial processing      | $x_t(s)$                                                               | $g[x(.)]$                                                                                                                                 | $x^i(s^\prime)$ where $s^\prime \in D_{s^\prime}$                     |
-| 3   | Variable transformation | $x(s; t)$                                                              | $T(x(s;t))$                                                                                                                                        | $Tx^i_t(s)$ ????
+|-|-------------------|--------------|----------------------|----------------------|
+| 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |
+| 2   | Spatial processing      | $x^i_t(s)$                                                             | $g[x^i_t(s)]$                                                                         | $x^i_t(s^\prime)$, where $s^\prime \in D_{s^\prime}$ ????
 |
-| 4   | Scaling                 | $x(s; t)$                                                              | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $Sx^i(s;t)$ where bbbbb                                                   |
-| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and$d < q$ |
-| 6   | Distribution fit        | $x(s; t)$                                                              | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1    |
-| 7   | Normalising             | $x(s; t)$                                                              | $\Phi^{-1}[x(s, t)]$                                                                                                                               | $z(s; t)$ where $z \sim N(.,.)$                                        |
-| 8   | Benchmarking            | $x(s; t)$                                                              |                                                                                                                                                    | $b(s;t)$                                                               |
-| 9   | Simplification          | $x(s; t)$                                                              | ccccc                                                                                                                                              | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |
+| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x(s;t)]$                                                                       | $Tx^i_t(s)$ ????
+|
+| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????
+|
+| 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
+| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x(s;t)]$, where $F_{\mathbf{\theta}}(.)$ is the CDF function | $p(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
+| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x(s; t)]$                                                              | $z(s; t)$, where $z$ has a normal distribution                          |
+| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x(s;t)]$                                                                       | $b(s;t)$                                                               |
+| 9   | Simplification          | $x^i(s; t)$                                                            | ccccc                                                                             | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |
 
-: An notation overview of the input, operation, and output of each pipeline module.
-
-Some notes:
-
--   still $x(s,t)$ notation because for all the other modules, $s$ and $t$ are all fixed, then $x(s, t)$ becomes $x$????
--   aaaaa: just a general function over the input????
--   bbbbb: input and output has the same distributional shape
--   ccccc: latex issue
+: An notation overview of the input, operation, and output of each pipeline module. {#tbl-notation}
 
@@
 #| fig-align: center
-#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
+#| fig-cap: "Diagram of pipeline modules for index construction. The highlighted path illustrates one possible construction using the dimension reduction and simplification module."
 #| out-height: 90%
@@
 
-`Input`: $x(t)$, time series of a single variable with fixed locations
-
-`Output`: $x^*(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.
-
-`Computation`: The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$  as a function of time.
-In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$,  $\tilde{x}_s^i(t^') = f[[x_s(t)]$ where $t^'$ refers to the new temporal resolution after aggregation.
+The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$ as a function of time.
+In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $\tilde{x}_s^i(t^\prime) = f[x^i_s(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation.
 An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$.
 That is also known as the time scale.
+For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).
+
 <!-- Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.
 -->
-For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).
-
-<!-- to incorporate the temporal neighbouring effect: $x^*(.) = f_{\Psi}[x(.)]$, where $\Psi$ is a vector of hyper-parameters associated with the transformation.
-In temporal processing, the transition of temporal dimension from $t$ to $t^\prime \subseteq D_{t^\prime}$ allows for the decomposition of time into its constituent temporal components.
-For example,  -->
-
-<!-- The construction of an index sometimes needs to consider information from neighbouring time periods.
-The temporal processing is a general operator on the time dimension of the data in the form of -->
-
-<!-- ```{=tex} -->
-
-<!-- \begin{equation} -->
-
-<!-- x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})), -->
-
-<!-- \end{equation} -->
-
-<!-- ``` -->
-
-<!-- where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. -->
 
@@
 
-`Input`: $x(s)$
-
-`Output`: $x^\#(s^\prime)$ where $s^\prime \in D_{s^{\prime}}$
-
-`Computation`: The spatial processing module takes a single variable with a fixed temporal dimension, $x(s)$, as input.
+The spatial processing module takes a single variable with a fixed temporal dimension, $x(s)$, as input.
 This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(.) = g_{\Theta}(x(.))$, where $\Theta$ is a vector of hyper-parameters associated with the transformation.
@@
 This also includes the aggregation of variables into different levels, such as city, state, and countries scales.
-
-<!-- may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing.
-The process can be written as a general operation in the form of -->
-
-<!-- ```{=tex} -->
-
-<!-- \begin{equation} -->
-
-<!-- x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})), -->
-
-<!-- \end{equation} -->
-
-<!-- ``` -->
-
-<!-- where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$.
-An example of spatial processing is to align variables collected in different resolutions.
-When variables are collected at different resolutions, analysts may choose to down-sample those in a finer resolution, $i$, to match those in a coarser resolution, $i^\prime$.
-This is a spatial aggregation and if aggregate using the mean, it can be written as -->
-
-<!-- ```{=tex} -->
-
-<!-- \begin{equation} -->
-
-<!-- g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}}, -->
-
-<!-- \end{equation} -->
-
-<!-- ``` -->
-
-<!-- where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the coarser grid.
-Other examples of spatial processing include 1) borrowing information from neighbouring spatial locations to interpolate unobserved locations and 2) fusing variables from ground measures with satellite imageries.
--->
 
@@
 
-`Input`: $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^p$
-
-`Output`: $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$
-
-`Computation`: Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$ , as the input.
+Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$ , as the input.
 It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output.
-
 The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods.
@@
 This is the only step that differs from the Standardized Precipitation Index (SPI).
-
-(Move this to the example and remove notation)
-When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques.
-Many social indexes use a linear combination in an additive structure, $y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),$ where $\lambda_i$ is the weight assigned to each individual variable $x_i$.
-A linear combination can also be preceived as a linear projection of multivariate information through the weight vector.
-Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making.
-One can visualise the data projection across different sets of weights through animation using a method called tour.
 
@@
 
+(Moved from the pipeline module to here) When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques.
+Many social indexes use a linear combination in an additive structure, $y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),$ where $\lambda_i$ is the weight assigned to each individual variable $x_i$.
+A linear combination can also be preceived as a linear projection of multivariate information through the weight vector.
+Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making.
+One can visualise the data projection across different sets of weights through animation using a method called tour.
+
 A natural thing to do when provided with the index data is to reproduce the index.
@@
 
-```         
     gggi %>% 
@@
           weight = weight))
-```
 

### 2023-08-15 tidyindex.tex
clean up after working with Patricia

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-In what follows when geographical or temporal components of the
-\(x(s,t)\) process are fixed we will be using suffix notation. For
-example, \(x_{s}(t)\) represents the data for a fixed location \(s\) as
-a function of time \(t\). While \(x_{t}(s)\) denotes the spatial varying
-procces for a fixed \(t\).
-
+In what follows when geographic or temporal components of the \(x(s,t)\)
+process are fixed we will be using suffix notation. For example,
+\(x_{s}(t)\) represents the data for a fixed location \(s\) as a
+function of time \(t\). While \(x_{t}(s)\) denotes the spatial varying
+proccess for a fixed \(t\). An overview of the notation for pipeline
+input, operation, and output is present in Table~\ref{tbl-notation}.
+
+\hypertarget{tbl-notation}{}
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}@{}}
-\caption{An notation overview of the input, operation, and output of
-each pipeline module.}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0128}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1795}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}
+\caption{\label{tbl-notation}An notation overview of the input,
+operation, and output of each pipeline module.}\tabularnewline
 \toprule()
@@
 \endhead
-1 & Temporal processing & \(x_{s}(t)\) & \(f_{\Psi}[x(.)]\) &
-\(\tilde{x}^i(t^\prime)\) where \(t\prime \in D_{t^\prime}\) \\
-2 & Spatial processing & \(x_t(s)\) & \(g[x(.)]\) & \(x^i(s^\prime)\)
-where \(s^\prime \in D_{s^\prime}\) \\
-3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) &
+1 & Temporal processing & \(x^i_{s}(t)\) & \(f[x^i_s(t)]\) &
+\(\tilde{x}^i_s(t^\prime)\), where \(t^\prime \in D_{t^\prime}\) \\
+2 & Spatial processing & \(x^i_t(s)\) & \(g[x^i_t(s)]\) &
+\(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
+3 & Variable transformation & \(x^i(s; t)\) & \(T[x(s;t)]\) &
 \(Tx^i_t(s)\) ???? \\
-4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(Sx^i(s;t)\)
-where bbbbb \\
-5 & Dimension reduction & \(\mathbf{x}(s;t)\) where
-\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) & aaaaa &
-\(\mathbf{y}(s;t)\) where \(\mathbf{y} \subseteq \mathbb{R}^d\)
-and\(d < q\) \\
-6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
-\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
-distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
-\(p(.)\) denotes the probability value between 0 and 1 \\
-7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
-where \(z \sim N(.,.)\) \\
-8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
-9 & Simplification & \(x(s; t)\) & ccccc &
+4 & Scaling & \(x^i(s; t)\) & \([x(s;t) - \alpha]/\gamma\) &
+\(Sx^i(s;t)\) ???? \\
+5 & Dimension reduction & \(\mathbf{x}(s;t)\), where
+\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
+\(h[\mathbf{x}(s;t)]\) & \(\mathbf{y}(s;t)\), where
+\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
+6 & Distribution fit & \(x^i(s; t)\) & \(F[x(s;t)]\), where
+\(F_{\mathbf{\theta}}(.)\) is the CDF function & \(p(s;t)\), where
+\(p(.) \in [0, 1]\) is a probability value \\
+7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x(s; t)]\) & \(z(s; t)\),
+where \(z\) has a normal distribution \\
+8 & Benchmarking & \(x^i(s; t)\) & \(u[x(s;t)]\) & \(b(s;t)\) \\
+9 & Simplification & \(x^i(s; t)\) & ccccc &
 \(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
@@
 
-Some notes:
-
-\begin{itemize}
-\tightlist
-\item
-  still \(x(s,t)\) notation because for all the other modules, \(s\) and
-  \(t\) are all fixed, then \(x(s, t)\) becomes \(x\)????
-\item
-  aaaaa: just a general function over the input????
-\item
-  bbbbb: input and output has the same distributional shape
-\item
-  ccccc: latex issue
-\end{itemize}
-
 \begin{figure}
@@
 
-\caption{\label{fig-pipeline-steps}Diagram of pipeline steps for index
-construction. will need to be updated with better design and the
-distribution fitting step.}
+\caption{\label{fig-pipeline-steps}Diagram of pipeline modules for index
+construction. The highlighted path illustrates one possible construction
+using the dimension reduction and simplification module.}
 
@@
 
-\texttt{Input}: \(x(t)\), time series of a single variable with fixed
-locations
-
-\texttt{Output}: \(x^*(t^{\prime})\), where
-\(t^\prime \in D_{t^{\prime}}\).
-
-\texttt{Computation}: The temporal processing module takes as input
-argument a single variable \(x_s^i(t)\) at location \(s\) as a function
-of time. In this step the original time series can be transformed or
-summarized into a new one via time aggregation, the transformation is
-represented by the function \(f\),
-\(\tilde{x}_s^i(t^\prime) = f[[x_s(t)]\) where \(t^\prime\) refers to
-the new temporal resolution after aggregation. An example of a temporal
-processing done in the computation of the Standardized Precipitation
-Index (SPI), consists of summing the monthly precipitation series over a
-rolling time window of size \(k\). That is also known as the time scale.
-For SPI, the choice of the time scale \(k\) is used to control the
-accumulation period for the water deficit, enabling the assessment of
-drought severity across various types (meteorological, agricultural, and
-hydrological).
+The temporal processing module takes as input argument a single variable
+\(x_s^i(t)\) at location \(s\) as a function of time. In this step the
+original time series can be transformed or summarized into a new one via
+time aggregation, the transformation is represented by the function
+\(f\), \(\tilde{x}_s^i(t^\prime) = f[x^i_s(t)]\) where \(t^\prime\)
+refers to the new temporal resolution after aggregation. An example of a
+temporal processing done in the computation of the Standardized
+Precipitation Index (SPI), consists of summing the monthly precipitation
+series over a rolling time window of size \(k\). That is also known as
+the time scale. For SPI, the choice of the time scale \(k\) is used to
+control the accumulation period for the water deficit, enabling the
+assessment of drought severity across various types (meteorological,
+agricultural, and hydrological).
 
@@
 
-\texttt{Input}: \(x(s)\)
-
-\texttt{Output}: \(x^\#(s^\prime)\) where
-\(s^\prime \in D_{s^{\prime}}\)
-
-\texttt{Computation}: The spatial processing module takes a single
-variable with a fixed temporal dimension, \(x(s)\), as input. This step
-transforms the variable from the original spatial dimension \(s\) into
-the new dimension \(s^\prime \in D_{s^\prime}\) through
+The spatial processing module takes a single variable with a fixed
+temporal dimension, \(x(s)\), as input. This step transforms the
+variable from the original spatial dimension \(s\) into the new
+dimension \(s^\prime \in D_{s^\prime}\) through
 \(x^\#(.) = g_{\Theta}(x(.))\), where \(\Theta\) is a vector of
@@
 
-\texttt{Input}: \(\mathbf{x}(s;t)\) where
-\(\mathbf{x} \subseteq \mathbb{R}^p\)
-
-\texttt{Output}: \(\mathbf{y}(s;t)\) where
-\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\)
-
-\texttt{Computation}: Dimension reduction takes the multivariate
-information \(\mathbf{x}(s;t)\), where
-\(\mathbf{x} \subseteq \mathbb{R}^p\) , as the input. It summarises the
-high-dimensional information into a lower dimension representation
-\(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\) and \(d < p\),
-as the output.
-
-The transformation can be based on domain-specific knowledge,
-originating from theories describing the underlying physical processes,
-or guided in statistical methods. For example, the Standardized
+Dimension reduction takes the multivariate information
+\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\) , as
+the input. It summarises the high-dimensional information into a lower
+dimension representation \(\mathbf{y}(s;t)\), where
+\(y \subseteq \mathbb{R}^d\) and \(d < p\), as the output. The
+transformation can be based on domain-specific knowledge, originating
+from theories describing the underlying physical processes, or guided in
+statistical methods. For example, the Standardized
 Precipitation-Evapotranspiration Index (SPEI) calculates the difference
@@
 Index (SPI).
-
-(Move this to the example and remove notation) When a physical process
-is absent to describe the phenomenon, multivariate data are often
-combined using statistical methods, which can be further classified into
-linear and non-linear techniques. Many social indexes use a linear
-combination in an additive structure,
-\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
-is the weight assigned to each individual variable \(x_i\). A linear
-combination can also be preceived as a linear projection of multivariate
-information through the weight vector. Projecting data from higher to
-lower dimension unavoidably leads to information loss and the used
-weights need careful examination to understand their effects on the
-index and implications for interpretation and decision-making. One can
-visualise the data projection across different sets of weights through
-animation using a method called tour.
 
@@
 
+(Moved from the pipeline module to here) When a physical process is
+absent to describe the phenomenon, multivariate data are often combined
+using statistical methods, which can be further classified into linear
+and non-linear techniques. Many social indexes use a linear combination
+in an additive structure,
+\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
+is the weight assigned to each individual variable \(x_i\). A linear
+combination can also be preceived as a linear projection of multivariate
+information through the weight vector. Projecting data from higher to
+lower dimension unavoidably leads to information loss and the used
+weights need careful examination to understand their effects on the
+index and implications for interpretation and decision-making. One can
+visualise the data projection across different sets of weights through
+animation using a method called tour.
+
 A natural thing to do when provided with the index data is to reproduce

### 2023-08-15 tidyindex.qmd
more small tweaks

@@
 
--   $x^i(s, t)$ represents a variable of interest for example precipitation,$j = 1, \cdots, p$ and
+-   $x^i(s, t)$ represents a variable of interest for example precipitation, $i = 1, \cdots, p$ and
 
@@
 |
-| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x(s;t)]$                                                                       | $Tx^i_t(s)$ ????
+| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x^i(s;t)]$                                                                       | $Tx^i(s;t)$ ????
 |
-| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????
+| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x^i(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????
 |
 | 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
-| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x(s;t)]$, where $F_{\mathbf{\theta}}(.)$ is the CDF function | $p(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
-| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x(s; t)]$                                                              | $z(s; t)$, where $z$ has a normal distribution                          |
-| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x(s;t)]$                                                                       | $b(s;t)$                                                               |
-| 9   | Simplification          | $x^i(s; t)$                                                            | ccccc                                                                             | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |
+| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F_{\mathbf{\theta}}(.)$ is the CDF function | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
+| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x^i(s; t)]$                                                              | $z^i(s; t)$, where $z$ has a normal distribution                          |
+| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x^i(s;t)]$                                                                       | $b^i(s;t)$                                                               |
+| 9   | Simplification          | $x^i(s; t)$                                                            | $v[x^i(s;t)]$                                                                            | $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |
 
@@
 
-Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$ , as the input.
-It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output.
+Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$, as the input.
+It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < q$, as the output.
 The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods.
@@
 \begin{equation}
+v[x_i(s;t)] = 
 \begin{cases}

### 2023-08-15 tidyindex.tex
more small tweaks

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
 
@@
   \(x^i(s, t)\) represents a variable of interest for example
-  precipitation,\(j = 1, \cdots, p\) and
+  precipitation, \(i = 1, \cdots, p\) and
 \item
@@
 \(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
-3 & Variable transformation & \(x^i(s; t)\) & \(T[x(s;t)]\) &
-\(Tx^i_t(s)\) ???? \\
-4 & Scaling & \(x^i(s; t)\) & \([x(s;t) - \alpha]/\gamma\) &
+3 & Variable transformation & \(x^i(s; t)\) & \(T[x^i(s;t)]\) &
+\(Tx^i(s;t)\) ???? \\
+4 & Scaling & \(x^i(s; t)\) & \([x^i(s;t) - \alpha]/\gamma\) &
 \(Sx^i(s;t)\) ???? \\
@@
 \(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
-6 & Distribution fit & \(x^i(s; t)\) & \(F[x(s;t)]\), where
-\(F_{\mathbf{\theta}}(.)\) is the CDF function & \(p(s;t)\), where
+6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where
+\(F_{\mathbf{\theta}}(.)\) is the CDF function & \(p^i(s;t)\), where
 \(p(.) \in [0, 1]\) is a probability value \\
-7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x(s; t)]\) & \(z(s; t)\),
-where \(z\) has a normal distribution \\
-8 & Benchmarking & \(x^i(s; t)\) & \(u[x(s;t)]\) & \(b(s;t)\) \\
-9 & Simplification & \(x^i(s; t)\) & ccccc &
-\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
+7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x^i(s; t)]\) &
+\(z^i(s; t)\), where \(z\) has a normal distribution \\
+8 & Benchmarking & \(x^i(s; t)\) & \(u[x^i(s;t)]\) & \(b^i(s;t)\) \\
+9 & Simplification & \(x^i(s; t)\) & \(v[x^i(s;t)]\) &
+\(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
 \bottomrule()
@@
 Dimension reduction takes the multivariate information
-\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\) , as
-the input. It summarises the high-dimensional information into a lower
-dimension representation \(\mathbf{y}(s;t)\), where
-\(y \subseteq \mathbb{R}^d\) and \(d < p\), as the output. The
-transformation can be based on domain-specific knowledge, originating
-from theories describing the underlying physical processes, or guided in
-statistical methods. For example, the Standardized
-Precipitation-Evapotranspiration Index (SPEI) calculates the difference
-\(D\) between precipitation (\(P\)) and potential evapotranspiration
-(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
-This is the only step that differs from the Standardized Precipitation
-Index (SPI).
+\(\mathbf{x}(s;t)\), where
+\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\), as the input. It
+summarises the high-dimensional information into a lower dimension
+representation \(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\)
+and \(d < q\), as the output. The transformation can be based on
+domain-specific knowledge, originating from theories describing the
+underlying physical processes, or guided in statistical methods. For
+example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
+calculates the difference \(D\) between precipitation (\(P\)) and
+potential evapotranspiration (\(\text{PET}\)), using a water balance
+model (\(D = P - \text{PET}\)). This is the only step that differs from
+the Standardized Precipitation Index (SPI).
 

### 2023-08-15 tidyindex.qmd
rework on variable transformation and scaling

@@
 | 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
-| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F_{\mathbf{\theta}}(.)$ is the CDF function | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
+| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F$ is the CDF function | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
 | 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x^i(s; t)]$                                                              | $z^i(s; t)$, where $z$ has a normal distribution                          |
@@
 
-The spatial processing module takes a single variable with a fixed temporal dimension, $x(s)$, as input.
-This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(.) = g_{\Theta}(x(.))$, where $\Theta$ is a vector of hyper-parameters associated with the transformation.
+The spatial processing module takes a single variable with a fixed temporal dimension, $x_t(s)$, as input.
+This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(s;t) = g[x(s;t)]$.
 The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions.
@@
 
-`Input`: $x^l(s; t)$
-
-`Output`: $y^l(s; t)$
-
-`Computation`: The purpose of variable transformation is to create variables that fits assumptions for further computing.
-These assumptions include a stable variance, normal distribution, or a certain scale required by some algorithms down the pipeline.
-Variable transformation is a general notion of a functional transformation on the variable:
-
-```{=tex}
-\begin{equation}
-h_{\tau}(x(\mathbf{s};\mathbf{t})),
-\end{equation}
-```
-where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation if any, and $d_{\tau}$ is the number of parameter of $\tau$.
-Transformation is needed for data that are highly skewed and some common transformations include log, quadratic, and square root transformation.
-
-<!-- A common assumption is normality in drought index.  -->
+Variable transformation takes input of a single variable $x^i(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^i(s;t)$.
+When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality.
+For example, in Human Development Index (HDI), a logarithmic transformation is applied to the variable Gross National Income per capita (GNI), to reduce its impact on HDI, particularly for countries with high GNI values.
 
@@
 
-`Input`: $x^l(s; t)$
-
-`Output`: $x^l(s; t)$
-
-`Computation`: While scaling can be seen as a specific type of variable transformation, it is separated into its own step to make the step explicit in the pipeline.
-The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of
-
-```{=tex}
-\begin{equation}
-[x(s_i;t_j) - \alpha]/\gamma.
-\end{equation}
-```
-For example, a z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
-@fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step.
-While both variable transformation and scaling are pre-processing steps, the scaling operations in green show the same distribution as the original data.
-
-<!-- Take one or multiple variables to create a new variable,  -->
-
-<!-- Restrict it to single variable, square root, log etc could be linearly, also non-linear -->
-
-<!-- change the shape of the variable -->
-
-<!-- <!-- A conversion of non-normal variables into (approximately) normal. wrong! -->
-
-<!-- GAM, can you do additive model pairwise/ three-way -->
-
-<!-- While being a common technique in data pre-processing, scaling has an impact on the outliers of the data and analysts need to be cautious on checking its effect. -->
+Unlike variable transformation, scaling maintains the distributional shape of the variable.
+It includes techniques such as centering, z-score standardisation, and min-max standardisation and can be expressed as $[x^i(s;t) - \alpha]/\gamma$.
+In Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardisation.
 
 ```{r fig-scale-var-trans-compare}
-#| fig-cap: "Comparison of operations in scaling (green) and variable transformation (orange) steps in free scale.
-Variables after the scaling operations have the same distribution as the origin, while the distribution changes after variable transformation.
+#| fig-cap: "Comparison of the module scaling (green) and variable transformation (orange).
+While both modules change the variable range, scaling maintains the same distributional shape, which is not the case with variable transformation.
 "
@@
            r"($x^{1/3}$)",
-           r"($\frac{x - min(x)}{max(x) - min(x)}$)",
+           r"($\frac{x - x_{min}}{x_{max} - x_{min}}$)",
            r"($\frac{x^{0.5} - 1}{0.5}$)",
@@
   facet_wrap(vars(var), scales = "free") + 
-  scale_color_brewer(palette = "Dark2") + 
+  scale_color_brewer(palette = "Dark2", name = "module") + 
   theme_bw() + 

### 2023-08-15 tidyindex.tex
rework on variable transformation and scaling

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 \(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
-6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where
-\(F_{\mathbf{\theta}}(.)\) is the CDF function & \(p^i(s;t)\), where
-\(p(.) \in [0, 1]\) is a probability value \\
+6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where \(F\) is
+the CDF function & \(p^i(s;t)\), where \(p(.) \in [0, 1]\) is a
+probability value \\
 7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x^i(s; t)]\) &
@@
 The spatial processing module takes a single variable with a fixed
-temporal dimension, \(x(s)\), as input. This step transforms the
+temporal dimension, \(x_t(s)\), as input. This step transforms the
 variable from the original spatial dimension \(s\) into the new
 dimension \(s^\prime \in D_{s^\prime}\) through
-\(x^\#(.) = g_{\Theta}(x(.))\), where \(\Theta\) is a vector of
-hyper-parameters associated with the transformation. The change of
-spatial dimension allows for the alignment of variables collected from
-different measurements, such as in-situ stations and satellite imagery,
-or originating from different resolutions. This also includes the
-aggregation of variables into different levels, such as city, state, and
-countries scales.
+\(x^\#(s;t) = g[x(s;t)]\). The change of spatial dimension allows for
+the alignment of variables collected from different measurements, such
+as in-situ stations and satellite imagery, or originating from different
+resolutions. This also includes the aggregation of variables into
+different levels, such as city, state, and countries scales.
 
@@
 
-\texttt{Input}: \(x^l(s; t)\)
-
-\texttt{Output}: \(y^l(s; t)\)
-
-\texttt{Computation}: The purpose of variable transformation is to
-create variables that fits assumptions for further computing. These
-assumptions include a stable variance, normal distribution, or a certain
-scale required by some algorithms down the pipeline. Variable
-transformation is a general notion of a functional transformation on the
-variable:
-
-\begin{equation}
-h_{\tau}(x(\mathbf{s};\mathbf{t})),
-\end{equation}
-
-where \(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in
-the transformation if any, and \(d_{\tau}\) is the number of parameter
-of \(\tau\). Transformation is needed for data that are highly skewed
-and some common transformations include log, quadratic, and square root
-transformation.
+Variable transformation takes input of a single variable \(x^i(s;t)\)
+and reshapes its distribution using the function \(T[.]\) to produce
+\(x^i(s;t)\). When a variable has a skewed distribution, transformations
+such as log, square root, or cubic root can adjust the distribution
+towards normality. For example, in Human Development Index (HDI), a
+logarithmic transformation is applied to the variable Gross National
+Income per capita (GNI), to reduce its impact on HDI, particularly for
+countries with high GNI values.
 
@@
 
-\texttt{Input}: \(x^l(s; t)\)
-
-\texttt{Output}: \(x^l(s; t)\)
-
-\texttt{Computation}: While scaling can be seen as a specific type of
-variable transformation, it is separated into its own step to make the
-step explicit in the pipeline. The key difference between the two steps
-is that variable transformation typically changes the shape of the data
-while scaling only changes the data scale and can usually be written in
-the form of
-
-\begin{equation}
-[x(s_i;t_j) - \alpha]/\gamma.
-\end{equation}
-
-For example, a z-score standardisation can be written in the above form
-with \(\alpha = \bar{x}(s; t)\) and \(\gamma = \sigma(s; t)\), a min-max
-standardisation uses \(\alpha = \min[x(s_i, t_j)]\) and
-\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\).
-Figure~\ref{fig-scale-var-trans-compare} shows a collection of variable
-pre-processing operations and uses color to differentiate whether the
-operation is a variable transformation or a scaling step. While both
-variable transformation and scaling are pre-processing steps, the
-scaling operations in green show the same distribution as the original
-data.
+Unlike variable transformation, scaling maintains the distributional
+shape of the variable. It includes techniques such as centering, z-score
+standardisation, and min-max standardisation and can be expressed as
+\([x^i(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
+three dimensions (health, education, and economy) are converted into the
+same scale (0-1) using min-max standardisation.
 
@@
 
-\caption{\label{fig-scale-var-trans-compare}Comparison of operations in
-scaling (green) and variable transformation (orange) steps in free
-scale. Variables after the scaling operations have the same distribution
-as the origin, while the distribution changes after variable
-transformation.}
+\caption{\label{fig-scale-var-trans-compare}Comparison of the module
+scaling (green) and variable transformation (orange). While both modules
+change the variable range, scaling maintains the same distributional
+shape, which is not the case with variable transformation.}
 

### 2023-08-15 tidyindex.qmd
rework distribution fit/ benchmark/ simplification

@@
 | 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
-| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F$ is the CDF function | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
+| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F$ is the CDF | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
 | 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x^i(s; t)]$                                                              | $z^i(s; t)$, where $z$ has a normal distribution                          |
@@
 
-<!-- Other weight constraint used includes $\sum_{p=1}^P\lambda_p^2 = 1$, is used by principal component analysis.  -->
-
-<!-- Spectral satellite imageries often collected at different bands and it is common to combine values from multiple bands to compose new variables relevant to the index.
-The most common example of this is the Normalized Difference Vegetation Index (NDVI), which uses near-infrared (NIR) ($x_1(s;t)$) and red channel ($x_2(s;t)$): -->
-
-<!-- \begin{equation} -->
-
-<!-- \text{NDVI}(s;t) = \frac{x_1(s;t) - x_2(s;t)}{x_1(s;t) + x_2(s;t)}, -->
-
-<!-- \end{equation} -->
-
-<!-- which is used to evaluate the vegetation from near-infrared (NIR) and red channel. -->
-
 ## Distribution fit
 
-<!-- *model fit? * -->
-
-`Input`: $x^l(s; t)$
-
-`Output`:
-
-`Computation`: Distribution fit can be seen as the model fitting in its simplest term. It can be represented by
-
-```{=tex}
-\begin{equation}
-F_{\eta}(x(\mathbf{s}; \mathbf{t})), 
-\end{equation}
-```
-where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$.
-A distribution fit typically aims at finding the distribution that best fits the data.
-Analysts may start from a pool of candidate distributions with a chosen fitting method and goodness of fit measure.
-While it is useful to find the ultimate best distribution to fits the data, from a probabilistic perspective, the fitting procedure itself has an uncertainty associated with the data fed and the parameter chosen.
-A reasonable alternative is to understand how much the index values can vary given different distributions, fitting methods, and goodness of fit tests, and whether these variations are negligible in a given application.
+Distribution fit applies the cumulative distribution function $F$ of a distribution on the variable $x^i(s; t)$ to obtain the probability values $p^i(s;t) \in [0, 1]$.
+In SPEI, many distributions, including log-logistic, pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series.
+Different fitting methods, and different goodness of fit tests may be used to compare the distribution choice on the index value.
 
@@
 
-`Input`: $x^l(s; t)$
-
-`Output`:
-
-`Computation`: This step maps the univariate series into a different scale, typically for ease of comparison across regions.
+<!-- This step maps the univariate series into a different scale, typically for ease of comparison across regions.
 For example, a normal scale, \[0, 1\], or \[0, 100\] may be favored for reporting certain indexes.
@@
 This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.
-
-```{r fig-normalising}
-#| fig-cap: "Scatterplot of normal quantiles against their density values.
-THree tail density values are highlighted with its probability of occurence labelled.
-Probability is calculated assuming monthly data: with a density of -2, the probability of occurrence is 1/pnorm(-2)/12 = 4 years.
-The non-linear relationship between the two quantities suggests normalised indexes need to be interpreted with caution since a slight change in the tail distribution can result in magnitudes of difference in its probability of occurrence."
-#| fig-width: 9
-library(tidyverse)
-all <- tibble(x = seq(-3, 3, 0.01), y = pnorm(x)) 
-dt <- tibble(x = c(-3, -2.5, -2), y = pnorm(x)) 
-label <- tibble(x = c(-3, -2.5, -2), y = pnorm(x), 
-                label = c("once in 62 years", "once in 13 years", "once in 4 years"))
-
-dt %>% 
-  ggplot() + 
-  geom_point(aes(x= x, y = y)) + 
-  geom_line(data = all, aes(x = x, y = y)) + 
-  ggrepel::geom_label_repel(
-    data = label, aes(x = x, y = y, label =label),
-    nudge_y = 0.2, nudge_x = 0.1,
-    arrow = arrow(length = unit(0.08, "inches")), min.segment.length = 0) +
-  scale_x_continuous(breaks = seq(-3, 3, 1)) + 
-  xlab("Normalised Score") + 
-  ylab("Probability") + 
-  theme_bw() + 
-  theme(panel.grid.minor = element_blank())
-```
+-->
 
@@
 
-`Input`: $x^l(s; t)$
-
-`Output`:
-
-$D(x^l(s,t), a)$ where $D(., .)$ is a distance function between
-
-`Computation`: Benchmarking sets a constant value to allow the constructed index to be compared across time.
-Here we denote it with $u[x(s_i, t_j)]$ where $u$ is a scalar of interest in the index constructed.
-A benchmark value could be a constant or a function of the data, i.e.
-mean.
+Benchmark sets a value $b^i(s,t)$ for comparing against the original variable $x^i(s;t)$.
+This benchmark can be a fixed value consistently across space and time or determined by the data through the function $u[x^i(s;t)]$.
+Once a benchmark being set, observations can be highlighted for adjustments in other modules or can serve targets for monitoring and planning.
 
@@
 
-`Input`: $x^l(s; t)$
-
-`Output`: $C^l(s;t)$ where $\{C: c_0, c_1, \cdots, c_z\}$ is a discrete set
-
-`Computation`: In public communication, the index values are usually accompanied by a categorical grade.
-The categorised grades are an ordered set of descriptive words or colors to communicate the severity or guide the comprehension of the indexes.
-The mapping from continuous index values to the discrete grades is called simplification in the pipeline and it can be written as a piece-wise function:
+Simplification takes a continuous variable $x^i(s;t)$ and categorises it into a discrete set $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$ through a piecewise function,
 
@@
 \begin{cases}
-a_0 & C_1 \leq x(s; t) < C_0 \\
-a_1 & C_2 \leq x(s; t) < C_1 \\
-a_2 & C_3 \leq x(s; t) < C_2 \\
+a_0 & C_1 \leq x^i(s; t) < C_0 \\
+a_1 & C_2 \leq x^i(s; t) < C_1 \\
+a_2 & C_3 \leq x^i(s; t) < C_2 \\
 \cdots \\
-a_z & C_z \leq x(s; t)
+a_z & C_z \leq x^i(s; t)
 \end{cases}
@@
 ```
-where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the thresholds for each category.
-In SPI, droughts are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$.
-In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.
+
+This is typically used at the end of the index pipeline to simplify the index to communicate to the public the severity of the concept of interest measured by the index.
+An example of simplification is to map the calculated SPI to four categories: mild, moderate, severe, and extreme drought.
 

### 2023-08-15 tidyindex.tex
rework distribution fit/ benchmark/ simplification

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where \(F\) is
-the CDF function & \(p^i(s;t)\), where \(p(.) \in [0, 1]\) is a
-probability value \\
+the CDF & \(p^i(s;t)\), where \(p(.) \in [0, 1]\) is a probability
+value \\
 7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x^i(s; t)]\) &
@@
 
-\texttt{Input}: \(x^l(s; t)\)
-
-\texttt{Output}:
-
-\texttt{Computation}: Distribution fit can be seen as the model fitting
-in its simplest term. It can be represented by
-
-\begin{equation}
-F_{\eta}(x(\mathbf{s}; \mathbf{t})), 
-\end{equation}
-
-where \(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
-parameter and \(d_{\eta}\) is the number of parameter of \(\eta\). A
-distribution fit typically aims at finding the distribution that best
-fits the data. Analysts may start from a pool of candidate distributions
-with a chosen fitting method and goodness of fit measure. While it is
-useful to find the ultimate best distribution to fits the data, from a
-probabilistic perspective, the fitting procedure itself has an
-uncertainty associated with the data fed and the parameter chosen. A
-reasonable alternative is to understand how much the index values can
-vary given different distributions, fitting methods, and goodness of fit
-tests, and whether these variations are negligible in a given
-application.
+Distribution fit applies the cumulative distribution function \(F\) of a
+distribution on the variable \(x^i(s; t)\) to obtain the probability
+values \(p^i(s;t) \in [0, 1]\). In SPEI, many distributions, including
+log-logistic, pearson III, lognormal, and general extreme distribution,
+are candidates for the aggregated series. Different fitting methods, and
+different goodness of fit tests may be used to compare the distribution
+choice on the index value.
 
@@
 
-\texttt{Input}: \(x^l(s; t)\)
-
-\texttt{Output}:
-
-\texttt{Computation}: This step maps the univariate series into a
-different scale, typically for ease of comparison across regions. For
-example, a normal scale, {[}0, 1{]}, or {[}0, 100{]} may be favored for
-reporting certain indexes. In drought indexes, i.e.~SPI or SPEI, the
-quantiles from the fitted distribution are converted into the normal
-scale via the normal reverse CDF function: \(\Phi^{-1}(.)\). Normalising
-is usually used at the end of the pipeline and its main difference from
-the scaling step is that here the change of scale also changes the
-distribution of the variable. While being commonly used, this step can
-get criticism from analysts for forcing the data into the decided scale,
-which can be either unnecessary or inaccurately exaggerate or downplay
-the outliers. Also, the use of a normal scale needs to be interpreted
-with caution. Figure~\ref{fig-normalising} illustrates the normal
-density not being directly proportional to its probability of
-occurrence. This is concerning, especially at the extreme values, since
-a small difference in the tail density can have magnitudes of difference
-in its probability of occurrence.
-
-\begin{figure}
-
-{\centering \includegraphics{tidyindex_files/figure-pdf/fig-normalising-1.pdf}
-
-}
-
-\caption{\label{fig-normalising}Scatterplot of normal quantiles against
-their density values. THree tail density values are highlighted with its
-probability of occurence labelled. Probability is calculated assuming
-monthly data: with a density of -2, the probability of occurrence is
-1/pnorm(-2)/12 = 4 years. The non-linear relationship between the two
-quantities suggests normalised indexes need to be interpreted with
-caution since a slight change in the tail distribution can result in
-magnitudes of difference in its probability of occurrence.}
-
-\end{figure}
-
 \hypertarget{benchmarking}{%
@@
 
-\texttt{Input}: \(x^l(s; t)\)
-
-\texttt{Output}:
-
-\(D(x^l(s,t), a)\) where \(D(., .)\) is a distance function between
-
-\texttt{Computation}: Benchmarking sets a constant value to allow the
-constructed index to be compared across time. Here we denote it with
-\(u[x(s_i, t_j)]\) where \(u\) is a scalar of interest in the index
-constructed. A benchmark value could be a constant or a function of the
-data, i.e.~mean.
+Benchmark sets a value \(b^i(s,t)\) for comparing against the original
+variable \(x^i(s;t)\). This benchmark can be a fixed value consistently
+across space and time or determined by the data through the function
+\(u[x^i(s;t)]\). Once a benchmark being set, observations can be
+highlighted for adjustments in other modules or can serve targets for
+monitoring and planning.
 
@@
 
-\texttt{Input}: \(x^l(s; t)\)
-
-\texttt{Output}: \(C^l(s;t)\) where \(\{C: c_0, c_1, \cdots, c_z\}\) is
-a discrete set
-
-\texttt{Computation}: In public communication, the index values are
-usually accompanied by a categorical grade. The categorised grades are
-an ordered set of descriptive words or colors to communicate the
-severity or guide the comprehension of the indexes. The mapping from
-continuous index values to the discrete grades is called simplification
-in the pipeline and it can be written as a piece-wise function:
+Simplification takes a continuous variable \(x^i(s;t)\) and categorises
+it into a discrete set \(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\)
+through a piecewise function,
 
@@
 
-where \(C_0, C_1,\cdots ,C_z\) are the categories and
-\(c_0, c_1, \cdots, c_z\) are the thresholds for each category. In SPI,
-droughts are sorted into four categories: mild drought: \([-0.99, 0]\);
-moderate drought: \([-1.49, -1]\); severe drought: \([-1.99, -1.5]\),
-and extreme drought: \([-\infty, -2]\). In this case,
-\(C_0, C_1, C_2, C_3\) are the drought categories: mild, moderate,
-severe, and extreme drought (\(z = 3\)) and
-\(c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2\) are the cutoff value for each
-class.
+This is typically used at the end of the index pipeline to simplify the
+index to communicate to the public the severity of the concept of
+interest measured by the index. An example of simplification is to map
+the calculated SPI to four categories: mild, moderate, severe, and
+extreme drought.
 

### 2023-08-15 tidyindex.qmd
proofread before subsections in Section 3

@@
 
-An overview of the pipeline is given in @fig-pipeline-steps to illustrate the construction from raw data to the final indexes.
-The pipeline includes eight modules for operations in the spatial, temporal, and multivariate aspects of the data as well as modules for comparing and communicating indexes.
-Analysts can choose a subset of modules and reorder them as needed to construct an index.
-
-Data pre-processing may happen before the start of the index pipeline to prepare multivariate data from different locations/ countries ready for constructing an index.
-For remote sensing data, it includes aligning the spatial resolution and temporal frequency of data collected from different satellite products, or merging in-situ stations with the satellite data.
+An overview of the pipeline is presented in @fig-pipeline-steps, illustrating the nine available modules designed to obtain the index from the data.
+These modules include operations for temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
+Analysts have the flexibility to construct indexes by connecting modules according to their preferences.
 
@@
 | Module                  | Input                                                                  | Operation                                                                         | Output                                                                 |
-|-|-------------------|--------------|----------------------|----------------------|
+|-|-------------------|-----------------|-------------------|----------------------|
 | 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |

### 2023-08-15 tidyindex.tex
proofread before subsections in Section 3

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-An overview of the pipeline is given in Figure~\ref{fig-pipeline-steps}
-to illustrate the construction from raw data to the final indexes. The
-pipeline includes eight modules for operations in the spatial, temporal,
-and multivariate aspects of the data as well as modules for comparing
-and communicating indexes. Analysts can choose a subset of modules and
-reorder them as needed to construct an index.
-
-Data pre-processing may happen before the start of the index pipeline to
-prepare multivariate data from different locations/ countries ready for
-constructing an index. For remote sensing data, it includes aligning the
-spatial resolution and temporal frequency of data collected from
-different satellite products, or merging in-situ stations with the
-satellite data.
+An overview of the pipeline is presented in
+Figure~\ref{fig-pipeline-steps}, illustrating the nine available modules
+designed to obtain the index from the data. These modules include
+operations for temporal and spatial aggregation, variable transformation
+and combination, distribution fitting, benchmark setting, and index
+communication. Analysts have the flexibility to construct indexes by
+connecting modules according to their preferences.
 
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1795}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2179}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
   >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}

### 2023-08-16 tidyindex.qmd
work on the GGGI example

@@
 The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023].
-The index is composed of 14 variables, expressed as female-to-male ratios, which are first aggregated in a linear combination into the four dimensions using the weight from the `V-weight` column in @tbl-gggi-weights.
-The weight is calculated as the inverse of the standard deviation of each variable and scaling to sum to 1 within each dimension to allow a one percentage point change in the standard deviation of each variable to contribute equally to the index.
-The four dimensions are then aggregated in another linear combination with equal weight to obtain the index.
-The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi`, along with the corresponding weights `gggi_weights`.
+The index, compiled from 14 variables expressed as female-to-male ratios, first aggregates these variables into four dimensions through linear combinations.
+These dimensions are then combined though another linear combination.
+@fig-pp-gggi illustrates this pipeline construction by applying the dimension reduction module twice on the data to generate the index.
+@tbl-gggi-weights presents the The variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction.
+In the table, the variable weights are computed as the inverse of the standard deviation of each variable, scaled to sum to 1.
+This weights ensures that a one percentage point change in the standard deviation of each variable to contribute equally to the index.
+Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weight.
+The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi` and @tbl-gggi-weights as `gggi_weights`.
+
+```{r fig-pp-gggi}
+#| fig-align: center
+#| fig-cap: "Index pipeline for the Global Gender Gap Index (GGGI). The index is constructed as applying the module dimension reduction twice on the data."
+#| out-height: 90%
+#| out-width: 100%
+knitr::include_graphics(here::here("figures/pipeline-gggi.png"))
+```
 
 ```{r tbl-gggi-weights}
-#| tbl-cap: Weights of the fourteen variables in Global Gender Gap Index
+#| tbl-cap: Weights used to compute the Global Gender Gap Index
 #| tbl-cap-location: bottom
@@
 
-(Moved from the pipeline module to here) When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques.
-Many social indexes use a linear combination in an additive structure, $y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),$ where $\lambda_i$ is the weight assigned to each individual variable $x_i$.
-A linear combination can also be preceived as a linear projection of multivariate information through the weight vector.
-Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making.
-One can visualise the data projection across different sets of weights through animation using a method called tour.
+The index can be reproduce with tidyindex as:
 
-A natural thing to do when provided with the index data is to reproduce the index.
-This helps index analysts to verify the index calculation and become familiar with the methodology.
-For GGGI, the construction can be simplified as a single linear aggregation step in the dimension reduction module, with the `Weight` column in @tbl-gggi-weights, which is the product of the variable weight (`V-weight`) and dimension weight (`D-weight`).
-
-```{r fig-pipeline-steps-gggi}
-#| fig-align: center
-#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
-#| out-height: 90%
-#| out-width: 100%
-knitr::include_graphics(here::here("figures/pipeline-gggi.png"))
-```
-
+```{r eval = FALSE, echo = TRUE}
 gggi %>% 
-      init(id = country) %>%
-      add_meta(gggi_weights, var_col = variable) %>% 
-      dimension_reduction(
-        index_new = aggregate_linear(
-          ~labour_force_participation:years_with_female_head_of_state,
-          weight = weight))
-
-The result can be compared with the GGGI values available in the report, validating the reproducibility of the index for country with no missing variables.
-
-```{r eval = FALSE}
-a <- dim_weights %>% select(-variable) %>% t() %>% as_tibble(rownames = "id")
-colnames(a) <- c("id", dim_weights$variable)
-a %>% 
-  pivot_longer(-id, names_to = "vars", values_to = "value") %>% 
-  ggplot(aes(x = value, group = vars)) + 
-  geom_histogram() + 
-  facet_wrap(vars(vars))
-```
-
-```{r fig-compare-gggi}
-#| fig-cap: "Verifying the calculation of the Global Gender Gap Index (GGGI).
-The index can be reproduced from the methodology described in the Global Gender Gap Report 2023 after removing the countries with missing variables, the treatment of which is unclear."
-set.seed(123)
-n <- 1000
-w_idx <- map(1:n, function(x){
-  w_idx <- stats::runif(4, min = 0.1, max = 1)
-  w_idx <- w_idx/sum(w_idx)
-  w_idx %>% as.matrix(nrow = 4, ncol = 1)
-  })
-rand_w <- reduce(w_idx, cbind) %>% as_tibble()
-colnames(rand_w) <- paste0("V", 1:n)
-
-dt2 <- gggi %>% 
   init(id = country) %>%
@@
 
-```{r eval = FALSE, gggi-weights, fig.height=20, fig.align='center', fig.width=10}
-dt3 <- dt2 %>%
-  swap_values(.id = 1, .param = weight,
-              .value = list(pillar_weight, paste0("V", 2:100)),
-              .raw_data = dt)
+After initialising the `gggi` object and attaching the `gggi_weights` as metadata, a single linear aggregation step within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata.
+While computing the index from the original 14 variables, it remains unclear how the missing values are handled within the index, which impacts 68 out of the total 146 countries.
+However, after aggregating variables into the four dimensions, where no missing values exist, the index is reproducible for all the countries.
 
-idx_wide <- dt3$data %>% select(country, dplyr::contains("index_new"))
-idx_wide2 <- idx_wide %>% mutate(index_avg = rowMeans(idx_wide[,2:(ncol(idx_wide))]))
-idx_long2 <- idx_wide2 %>% pivot_longer(-country, names_to = "index", values_to = "value") %>% 
-  left_join(gggi %>% select(country, rank), by = "country")
-
-idx_long2 %>%
-  ggplot(aes(x = value, y = fct_reorder(country, -rank), group = country)) +
-  geom_boxplot() + 
-  #geom_point(color = "grey80", size = 0.5) +
-  geom_point(data = idx_long2 %>% filter(index == "index_new1"), color = "red") + 
-  theme_bw() + 
-  ylab("Country") + 
-  xlab("Index")
-```
+A linear combination can also be perceived as a linear projection of multivariate information through the weight vector.
+Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making.
+One can visualise the data projection across different sets of weights through animation using a method called tour.
 

### 2023-08-16 tidyindex.tex
work on the GGGI example

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 Opportunity, Educational Attainment, Health and Survival, and Political
-Empowerment (World Economic Forum 2023). The index is composed of 14
-variables, expressed as female-to-male ratios, which are first
-aggregated in a linear combination into the four dimensions using the
-weight from the \texttt{V-weight} column in
-Table~\ref{tbl-gggi-weights}. The weight is calculated as the inverse of
-the standard deviation of each variable and scaling to sum to 1 within
-each dimension to allow a one percentage point change in the standard
-deviation of each variable to contribute equally to the index. The four
-dimensions are then aggregated in another linear combination with equal
-weight to obtain the index. The 2023 GGGI data is available from the
-Global Gender Gap Report 2023 in the country's economy profile and can
-be accessed in R via the \texttt{tidyindex} package as \texttt{gggi},
-along with the corresponding weights \texttt{gggi\_weights}.
+Empowerment (World Economic Forum 2023). The index, compiled from 14
+variables expressed as female-to-male ratios, first aggregates these
+variables into four dimensions through linear combinations. These
+dimensions are then combined though another linear combination.
+Figure~\ref{fig-pp-gggi} illustrates this pipeline construction by
+applying the dimension reduction module twice on the data to generate
+the index. Table~\ref{tbl-gggi-weights} presents the The variable
+weights (\texttt{V-weight}) and dimension weights (\texttt{D-weight})
+used in the two dimension reduction. In the table, the variable weights
+are computed as the inverse of the standard deviation of each variable,
+scaled to sum to 1. This weights ensures that a one percentage point
+change in the standard deviation of each variable to contribute equally
+to the index. Dimension weights are equal across the four dimensions and
+the last column, \texttt{weight}, multiples the variable and dimension
+weights to produce a single set of weight. The 2023 GGGI data is
+available from the Global Gender Gap Report 2023 in the country's
+economy profile and can be accessed in R via the \texttt{tidyindex}
+package as \texttt{gggi} and Table~\ref{tbl-gggi-weights} as
+\texttt{gggi\_weights}.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-gggi.png}
+
+}
+
+\caption{\label{fig-pp-gggi}Index pipeline for the Global Gender Gap
+Index (GGGI). The index is constructed as applying the module dimension
+reduction twice on the data.}
+
+\end{figure}
 
@@
   >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0909}}@{}}
-\caption{\label{tbl-gggi-weights}Weights of the fourteen variables in
-Global Gender Gap Index}\tabularnewline
+\caption{\label{tbl-gggi-weights}Weights used to compute the Global
+Gender Gap Index}\tabularnewline
 \toprule()
@@
 
-(Moved from the pipeline module to here) When a physical process is
-absent to describe the phenomenon, multivariate data are often combined
-using statistical methods, which can be further classified into linear
-and non-linear techniques. Many social indexes use a linear combination
-in an additive structure,
-\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
-is the weight assigned to each individual variable \(x_i\). A linear
-combination can also be preceived as a linear projection of multivariate
-information through the weight vector. Projecting data from higher to
-lower dimension unavoidably leads to information loss and the used
-weights need careful examination to understand their effects on the
+The index can be reproduce with tidyindex as:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{gggi }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{add\_meta}\NormalTok{(gggi\_weights, }\AttributeTok{var\_col =}\NormalTok{ variable) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{dimension\_reduction}\NormalTok{(}
+    \AttributeTok{index\_new =} \FunctionTok{aggregate\_linear}\NormalTok{(}
+      \SpecialCharTok{\textasciitilde{}}\NormalTok{labour\_force\_participation}\SpecialCharTok{:}\NormalTok{years\_with\_female\_head\_of\_state,}
+      \AttributeTok{weight =}\NormalTok{ weight)) }
+\end{Highlighting}
+\end{Shaded}
+
+After initialising the \texttt{gggi} object and attaching the
+\texttt{gggi\_weights} as metadata, a single linear aggregation step
+within the dimension reduction module is applied to the 14 variables
+(from column \texttt{labour\_force\_participation} to
+\texttt{years\_with\_female\_head\_of\_state}), using the weight
+specified in the \texttt{weight} column of the attached metadata. While
+computing the index from the original 14 variables, it remains unclear
+how the missing values are handled within the index, which impacts 68
+out of the total 146 countries. However, after aggregating variables
+into the four dimensions, when no missing values exist, the index is
+reproducible for all the countries.
+
+A linear combination can also be perceived as a linear projection of
+multivariate information through the weight vector. Projecting data from
+higher to lower dimension unavoidably leads to information loss and the
+used weights need careful examination to understand their effects on the
 index and implications for interpretation and decision-making. One can
@@
 animation using a method called tour.
-
-A natural thing to do when provided with the index data is to reproduce
-the index. This helps index analysts to verify the index calculation and
-become familiar with the methodology. For GGGI, the construction can be
-simplified as a single linear aggregation step in the dimension
-reduction module, with the \texttt{Weight} column in
-Table~\ref{tbl-gggi-weights}, which is the product of the variable
-weight (\texttt{V-weight}) and dimension weight (\texttt{D-weight}).
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-gggi.png}
-
-}
-
-\caption{\label{fig-pipeline-steps-gggi}Diagram of pipeline steps for
-index construction. will need to be updated with better design and the
-distribution fitting step.}
-
-\end{figure}
-
-\begin{verbatim}
-gggi %>% 
-  init(id = country) %>%
-  add_meta(gggi_weights, var_col = variable) %>% 
-  dimension_reduction(
-    index_new = aggregate_linear(
-      ~labour_force_participation:years_with_female_head_of_state,
-      weight = weight)) 
-\end{verbatim}
-
-The result can be compared with the GGGI values available in the report,
-validating the reproducibility of the index for country with no missing
-variables.
 

### 2023-08-16 tidyindex.qmd
GGGI example revised

@@
 
-After initialising the `gggi` object and attaching the `gggi_weights` as metadata, a single linear aggregation step within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata.
+After initialising the `gggi` object and attaching the `gggi_weights` as metadata, a single linear combination step within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata.
 While computing the index from the original 14 variables, it remains unclear how the missing values are handled within the index, which impacts 68 out of the total 146 countries.
@@
 
-A linear combination can also be perceived as a linear projection of multivariate information through the weight vector.
-Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making.
-One can visualise the data projection across different sets of weights through animation using a method called tour.
+A linear combination can also be interpreted as a linear projection of multivariate information with a weight vector.
+Projecting data from higher to lower dimension unavoidably leads to information loss and the weights used require careful examination to understand their effects on the index and implications for interpretation and decision-making.
+We use a visualisation method called "tour" to animate the data projection across different sets of weights, aiming to understand how changes on the politics weight will affect the index value.
+We select a set countries for this exercise, which includes:
 
-To understand the uncertainty of this dimension reduction step from combining four dimensions (Economy, Education, Health, and Politics), we run a local tour \[TODO: reference to local tour\] that varies the weight of each dimension slightly to observe how index value and country ranking changes.
-The local tour produces an animation that gradually increases the weight of one variable and reduces it back to equal weight, one at a time.
-A set of countries are selected for this exercise, which includes 1) the top four with GGGI \> 0.85, 2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the bottom five countries with GGGI \< 0.6.
-Five frames (equal weights and one for each dimension with a relatively higher weight) are selected from the local tour animation to show in @fig-idx-tour and the full animation is available at https://vimeo.com/847874016.
+  1) Top-ranked countries with GGGI > 0.85,
+  2) Countries ranked between 57 and 62, with GGGI values from 0.72 to 0.73, and
+  3) Low-ranked countries with GGGI < 0.6.
 
-@fig-idx-tour helps to understand the how different dimension reduction alternatives affect the whole index distribution and individual countries.
+We run a local tour \[TODO: reference to local tour\] to slightly vary the weight of the politics dimension from the original 0.25 while keeping the weights constant for the other three dimensions.
 
--   Bangladesh
--   when decreasing the weight on politics, difference between countries narrower.
--   increase politics -\> all the country has lower index value
+In @fig-idx-tour, six frames have been chosen from the animation available at https://vimeo.com/847874016.
+When the weight of politics is reduced (Frame 1 and 6 vs.
+Frame 12), the difference in GGGI values between the top Nordic countries and mid- or low-ranked countries narrows, suggesting a smaller variation among countries in achieving gender parity.
+Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs.
+Frame 12), nearly all the countries in the three categories experience a decrease in GGGI values.
+A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight.
+This leads to its index value being almost similar to those of the top-ranked Nordic countries when the politics dimension is assigned a weight of 0.52 in Frame 29.
 
-<!-- Notably, increasing the weight in politics leads to a pronounced decrease of index value for almost all the countries.
-Given the index value has a direct interpretation on how much of the gender gap has been closed, such a weight variation can cast a negative light on the progression to gender parity.
-Analysts need to further investigate the construction of the politics dimension and consider whether this is desirable for the index.
--->
+In GGGI, the index value has a direct interpretation as the percentage of gender gap that has been closed.
+When countries exhibit closer index values, or consistent decreasing index values, it can be directly interpreted as country's progress towards gender parity.
+Analysts must carefully consider whether such interpretation is reasonable and review the index construction process.
 
@@
 ```{r fig-idx-tour, fig.align='center', fig.height=5}
-#| fig-cap: "Five frames selected from varying the linear weights of four dimensions in Global Gender Gap Index.
-The weights vary slightly from the official simple average weights (0.25, 0.25, 0.25, 0.25) to observe how the index and ranking reponse.
+#| fig-cap: "Six frames selected from varying the weights of the politics dimension in Global Gender Gap Index (GGGI).
+The top panel shows the GGGI value against the country, ranked by its original index value in Frame 12.
+The bottom panel displays the weight used to produce the index value in the top panel, with each frame corresponds to a set of weights.
+Countries selected includes 1) top-ranked countries with GGGI > 0.85, 2) countries ranked between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked countries with GGGI < 0.6.
+Compared to Frame 12 where equal weigts are used for the four dimension, a reduced weight in politics (Frame 1 and 6) shows narrower gaps between top and mid- or lower- ranked countries, while an increase in the politics weight (Frame 18, 24, and 29) leads to a systematic decrease of GGGI values across all the countries, except for Bangladesh.
 Full animation is available at https://vimeo.com/847874016."

### 2023-08-16 tidyindex.tex
GGGI example revised

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, sharp corners, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 
@@
 After initialising the \texttt{gggi} object and attaching the
-\texttt{gggi\_weights} as metadata, a single linear aggregation step
+\texttt{gggi\_weights} as metadata, a single linear combination step
 within the dimension reduction module is applied to the 14 variables
@@
 out of the total 146 countries. However, after aggregating variables
-into the four dimensions, when no missing values exist, the index is
+into the four dimensions, where no missing values exist, the index is
 reproducible for all the countries.
 
-A linear combination can also be perceived as a linear projection of
-multivariate information through the weight vector. Projecting data from
+A linear combination can also be interpreted as a linear projection of
+multivariate information with a weight vector. Projecting data from
 higher to lower dimension unavoidably leads to information loss and the
-used weights need careful examination to understand their effects on the
-index and implications for interpretation and decision-making. One can
-visualise the data projection across different sets of weights through
-animation using a method called tour.
-
-To understand the uncertainty of this dimension reduction step from
-combining four dimensions (Economy, Education, Health, and Politics), we
-run a local tour {[}TODO: reference to local tour{]} that varies the
-weight of each dimension slightly to observe how index value and country
-ranking changes. The local tour produces an animation that gradually
-increases the weight of one variable and reduces it back to equal
-weight, one at a time. A set of countries are selected for this
-exercise, which includes 1) the top four with GGGI \textgreater{} 0.85,
-2) a set of countries with GGGI between 0.72 and 0.73 (Brazil, Panama,
-Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and 3) the
-bottom five countries with GGGI \textless{} 0.6. Five frames (equal
-weights and one for each dimension with a relatively higher weight) are
-selected from the local tour animation to show in
-Figure~\ref{fig-idx-tour} and the full animation is available at
-https://vimeo.com/847874016.
-
-Figure~\ref{fig-idx-tour} helps to understand the how different
-dimension reduction alternatives affect the whole index distribution and
-individual countries.
-
-\begin{itemize}
+weights used require careful examination to understand their effects on
+the index and implications for interpretation and decision-making. We
+use a visualisation method called ``tour'' to animate the data
+projection across different sets of weights, aiming to understand how
+changes on the politics weight will affect the index value. We select a
+set countries for this exercise, which includes:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
 \tightlist
 \item
-  Bangladesh
-\item
-  when decreasing the weight on politics, difference between countries
-  narrower.
-\item
-  increase politics -\textgreater{} all the country has lower index
-  value
-\end{itemize}
+  Top-ranked countries with GGGI \textgreater{} 0.85,
+\item
+  Countries ranked between 57 and 62, with GGGI values from 0.72 to
+  0.73, and
+\item
+  Low-ranked countries with GGGI \textless{} 0.6.
+\end{enumerate}
+
+We run a local tour {[}TODO: reference to local tour{]} to slightly vary
+the weight of the politics dimension from the original 0.25 while
+keeping the weights constant for the other three dimensions.
+
+In Figure~\ref{fig-idx-tour}, six frames have been chosen from the
+animation available at https://vimeo.com/847874016. When the weight of
+politics is reduced (Frame 1 and 6 vs.~Frame 12), the difference in GGGI
+values between the top Nordic countries and mid- or low-ranked countries
+narrows, suggesting a smaller variation among countries in achieving
+gender parity. Conversely, when the weight of politics increases (Frame
+18, 24, and 29 vs.~Frame 12), nearly all the countries in the three
+categories experience a decrease in GGGI values. A noticeable exception
+to this trend is Bangladesh, where its index value moves in the same
+direction as the politics weight. This leads to its index value being
+almost similar to those of the top-ranked Nordic countries when the
+politics dimension is assigned a weight of 0.52 in Frame 29.
+
+In GGGI, the index value has a direct interpretation as the percentage
+of gender gap that has been closed. When countries exhibit closer index
+values, or consistent decreasing index values, it can be directly
+interpreted as country's progress towards gender parity. Analysts must
+carefully consider whether such interpretation is reasonable and review
+the index construction process.
 
@@
 
-\caption{\label{fig-idx-tour}Five frames selected from varying the
-linear weights of four dimensions in Global Gender Gap Index. The
-weights vary slightly from the official simple average weights (0.25,
-0.25, 0.25, 0.25) to observe how the index and ranking reponse. Full
-animation is available at https://vimeo.com/847874016.}
+\caption{\label{fig-idx-tour}Six frames selected from varying the
+weights of the politics dimension in Global Gender Gap Index (GGGI). The
+top panel shows the GGGI value against the country, ranked by its
+original index value in Frame 12. The bottom panel displays the weight
+used to produce the index value in the top panel, with each frame
+corresponds to a set of weights. Countries selected includes 1)
+top-ranked countries with GGGI \textgreater{} 0.85, 2) countries ranked
+between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked
+countries with GGGI \textless{} 0.6. Compared to Frame 12 where equal
+weigts are used for the four dimension, a reduced weight in politics
+(Frame 1 and 6) shows narrower gaps between top and mid- or lower-
+ranked countries, while an increase in the politics weight (Frame 18,
+24, and 29) leads to a systematic decrease of GGGI values across all the
+countries, except for Bangladesh. Full animation is available at
+https://vimeo.com/847874016.}
 

### 2023-08-16 tidyindex.qmd
minor edit and changes in GGGI examples

@@
 The index, compiled from 14 variables expressed as female-to-male ratios, first aggregates these variables into four dimensions through linear combinations.
-These dimensions are then combined though another linear combination.
+These dimensions are then combined though another linear combination to form the index.
 @fig-pp-gggi illustrates this pipeline construction by applying the dimension reduction module twice on the data to generate the index.
-@tbl-gggi-weights presents the The variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction.
+@tbl-gggi-weights presents the variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction.
 In the table, the variable weights are computed as the inverse of the standard deviation of each variable, scaled to sum to 1.
-This weights ensures that a one percentage point change in the standard deviation of each variable to contribute equally to the index.
+These weights ensure that a one percentage point change in the standard deviation of each variable to contribute equally to the index.
 Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weight.
-The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi` and @tbl-gggi-weights as `gggi_weights`.
 
@@
 
-The index can be reproduce with tidyindex as:
+The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi` and @tbl-gggi-weights as `gggi_weights`.
+The index can be reproduced with the pacakge as:
 
@@
 Projecting data from higher to lower dimension unavoidably leads to information loss and the weights used require careful examination to understand their effects on the index and implications for interpretation and decision-making.
-We use a visualisation method called "tour" to animate the data projection across different sets of weights, aiming to understand how changes on the politics weight will affect the index value.
-We select a set countries for this exercise, which includes:
+By making slight adjustments to the weight vector, we can observe how index values and country rankings change.
+For illustration, we select a set countries including:
 
@@
 
-We run a local tour \[TODO: reference to local tour\] to slightly vary the weight of the politics dimension from the original 0.25 while keeping the weights constant for the other three dimensions.
+We slightly vary the weight of the politics dimension from the original 0.25 while keeping the weights constant for the other three dimensions.
+This process creates an animation showing the movement of index values in response to changing weights.
+This visualisation technique, which presents a sequence of data projections, is referred to as a "tour".
+The specific kind of tour that moves between the original and nearby projections is known as a "local tour".
 
@@
 In GGGI, the index value has a direct interpretation as the percentage of gender gap that has been closed.
-When countries exhibit closer index values, or consistent decreasing index values, it can be directly interpreted as country's progress towards gender parity.
-Analysts must carefully consider whether such interpretation is reasonable and review the index construction process.
+When countries exhibit closer index values or consistent, it can be directly interpreted as a reflection of a country's progress toward gender parity.
+Ideally, an index should be robust against minor changes in its construction components.
+This example provides researchers with means to observe changes in the index resulting from variations in the weights used in linear combination.
+This approach can be applied broadly to different sets of weights of interest, extending beyond the change in a single dimension illustrated by the example.
 
@@
 ```{r fig-idx-tour, fig.align='center', fig.height=5}
-#| fig-cap: "Six frames selected from varying the weights of the politics dimension in Global Gender Gap Index (GGGI).
+#| fig-cap: "Six frames selected to explore how varying the weights of the politics dimension changes the index values and country ranking in Global Gender Gap Index (GGGI).
 The top panel shows the GGGI value against the country, ranked by its original index value in Frame 12.

### 2023-08-16 tidyindex.tex
minor edit and changes in GGGI examples

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, sharp corners, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 variables into four dimensions through linear combinations. These
-dimensions are then combined though another linear combination.
-Figure~\ref{fig-pp-gggi} illustrates this pipeline construction by
-applying the dimension reduction module twice on the data to generate
-the index. Table~\ref{tbl-gggi-weights} presents the The variable
-weights (\texttt{V-weight}) and dimension weights (\texttt{D-weight})
-used in the two dimension reduction. In the table, the variable weights
-are computed as the inverse of the standard deviation of each variable,
-scaled to sum to 1. This weights ensures that a one percentage point
-change in the standard deviation of each variable to contribute equally
-to the index. Dimension weights are equal across the four dimensions and
-the last column, \texttt{weight}, multiples the variable and dimension
-weights to produce a single set of weight. The 2023 GGGI data is
-available from the Global Gender Gap Report 2023 in the country's
-economy profile and can be accessed in R via the \texttt{tidyindex}
-package as \texttt{gggi} and Table~\ref{tbl-gggi-weights} as
-\texttt{gggi\_weights}.
+dimensions are then combined though another linear combination to form
+the index. Figure~\ref{fig-pp-gggi} illustrates this pipeline
+construction by applying the dimension reduction module twice on the
+data to generate the index. Table~\ref{tbl-gggi-weights} presents the
+variable weights (\texttt{V-weight}) and dimension weights
+(\texttt{D-weight}) used in the two dimension reduction. In the table,
+the variable weights are computed as the inverse of the standard
+deviation of each variable, scaled to sum to 1. These weights ensure
+that a one percentage point change in the standard deviation of each
+variable to contribute equally to the index. Dimension weights are equal
+across the four dimensions and the last column, \texttt{weight},
+multiples the variable and dimension weights to produce a single set of
+weight.
 
@@
 
-The index can be reproduce with tidyindex as:
+The 2023 GGGI data is available from the Global Gender Gap Report 2023
+in the country's economy profile and can be accessed in R via the
+\texttt{tidyindex} package as \texttt{gggi} and
+Table~\ref{tbl-gggi-weights} as \texttt{gggi\_weights}. The index can be
+reproduced with the pacakge as:
 
@@
 weights used require careful examination to understand their effects on
-the index and implications for interpretation and decision-making. We
-use a visualisation method called ``tour'' to animate the data
-projection across different sets of weights, aiming to understand how
-changes on the politics weight will affect the index value. We select a
-set countries for this exercise, which includes:
+the index and implications for interpretation and decision-making. By
+making slight adjustments to the weight vector, we can observe how index
+values and country rankings change. For illustration, we select a set
+countries including:
 
@@
 
-We run a local tour {[}TODO: reference to local tour{]} to slightly vary
-the weight of the politics dimension from the original 0.25 while
-keeping the weights constant for the other three dimensions.
+We slightly vary the weight of the politics dimension from the original
+0.25 while keeping the weights constant for the other three dimensions.
+This process creates an animation showing the movement of index values
+in response to changing weights. This visualisation technique, which
+presents a sequence of data projections, is referred to as a ``tour''.
+The specific kind of tour that moves between the original and nearby
+projections is known as a ``local tour''.
 
@@
 of gender gap that has been closed. When countries exhibit closer index
-values, or consistent decreasing index values, it can be directly
-interpreted as country's progress towards gender parity. Analysts must
-carefully consider whether such interpretation is reasonable and review
-the index construction process.
+values or consistent, it can be directly interpreted as a reflection of
+a country's progress toward gender parity. Ideally, an index should be
+robust against minor changes in its construction components. This
+example provides researchers with means to observe changes in the index
+resulting from variations in the weights used in linear combination.
+This approach can be applied broadly to different sets of weights of
+interest, extending beyond the change in a single dimension illustrated
+by the example.
 
@@
 
-\caption{\label{fig-idx-tour}Six frames selected from varying the
-weights of the politics dimension in Global Gender Gap Index (GGGI). The
-top panel shows the GGGI value against the country, ranked by its
-original index value in Frame 12. The bottom panel displays the weight
-used to produce the index value in the top panel, with each frame
-corresponds to a set of weights. Countries selected includes 1)
-top-ranked countries with GGGI \textgreater{} 0.85, 2) countries ranked
-between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked
-countries with GGGI \textless{} 0.6. Compared to Frame 12 where equal
-weigts are used for the four dimension, a reduced weight in politics
-(Frame 1 and 6) shows narrower gaps between top and mid- or lower-
-ranked countries, while an increase in the politics weight (Frame 18,
-24, and 29) leads to a systematic decrease of GGGI values across all the
-countries, except for Bangladesh. Full animation is available at
-https://vimeo.com/847874016.}
+\caption{\label{fig-idx-tour}Six frames selected to explore how varying
+the weights of the politics dimension changes the index values and
+country ranking in Global Gender Gap Index (GGGI). The top panel shows
+the GGGI value against the country, ranked by its original index value
+in Frame 12. The bottom panel displays the weight used to produce the
+index value in the top panel, with each frame corresponds to a set of
+weights. Countries selected includes 1) top-ranked countries with GGGI
+\textgreater{} 0.85, 2) countries ranked between 57 and 62 with GGGI
+from 0.72 to 0.73, and 3) low-ranked countries with GGGI \textless{}
+0.6. Compared to Frame 12 where equal weigts are used for the four
+dimension, a reduced weight in politics (Frame 1 and 6) shows narrower
+gaps between top and mid- or lower- ranked countries, while an increase
+in the politics weight (Frame 18, 24, and 29) leads to a systematic
+decrease of GGGI values across all the countries, except for Bangladesh.
+Full animation is available at https://vimeo.com/847874016.}
 

### 2023-08-17 tidyindex.qmd
use j for variable in the subscript instead in notation

@@
 \begin{equation}
-\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
+\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
 \end{equation}
@@
 
--   $x^i(s, t)$ represents a variable of interest for example precipitation, $i = 1, \cdots, p$ and
+-   $x_j(s, t)$ represents a variable of interest for example precipitation, $j = 1, \cdots, p$ and
 
@@
 
-In what follows when geographic or temporal components of the $x(s,t)$ process are fixed we will be using suffix notation.
-For example, $x_{s}(t)$ represents the data for a fixed location $s$ as a function of time $t$.
-While $x_{t}(s)$ denotes the spatial varying proccess for a fixed $t$.
+In what follows when geographic or temporal components of the $x_j(s,t)$ process are fixed we will be using suffix notation.
+For example, $x_{sj}(t)$ represents the data for a fixed location $s$ as a function of time $t$.
+While $x_{tj}(s)$ denotes the spatial varying process for a fixed $t$.
 An overview of the notation for pipeline input, operation, and output is present in @tbl-notation.
@@
 | Module                  | Input                                                                   | Operation                           | Output                                                                        |
-|-|-------------------|-----------------|-------------------|----------------------|
-| 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |
-| 2   | Spatial processing      | $x^i_t(s)$                                                             | $g[x^i_t(s)]$                                                                         | $x^i_t(s^\prime)$, where $s^\prime \in D_{s^\prime}$ ????
-|
-| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x^i(s;t)]$                                                                       | $Tx^i(s;t)$ ????
-|
-| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x^i(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????
-|
-| 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
-| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F$ is the CDF | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
-| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x^i(s; t)]$                                                              | $z^i(s; t)$, where $z$ has a normal distribution                          |
-| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x^i(s;t)]$                                                                       | $b^i(s;t)$                                                               |
-| 9   | Simplification          | $x^i(s; t)$                                                            | $v[x^i(s;t)]$                                                                            | $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |
+|--|---------------|----------|---------------|--------------------------|
+| 1   | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
+| 2   | Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
+| 3   | Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
+| 4   | Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
+| 5   | Dimension reduction     | $\mathbf{x}(s;t) \quad \mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q$      |
+| 6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $p_j(s;t) \quad p(.) \in [0, 1]$                   |
+| 7   | Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
+| 8   | Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
+| 9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                      |
 
@@
 
-The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$ as a function of time.
-In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $\tilde{x}_s^i(t^\prime) = f[x^i_s(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation.
+The temporal processing module takes as input argument a single variable $x_{sj}(t)$ at location $s$ as a function of time.
+In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation.
 An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$.
@@
 
-<!-- Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.
--->
-
 ## Spatial processing
 
-The spatial processing module takes a single variable with a fixed temporal dimension, $x_t(s)$, as input.
-This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(s;t) = g[x(s;t)]$.
+The spatial processing module takes a single variable with a fixed temporal dimension, $x_{tj}(s)$, as input.
+This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]$.
 The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions.
@@
 
-Variable transformation takes input of a single variable $x^i(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^i(s;t)$.
+Variable transformation takes input of a single variable $x_j(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^{\text{Trans}}_{j}(s;t)$.
 When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality.
@@
 Unlike variable transformation, scaling maintains the distributional shape of the variable.
-It includes techniques such as centering, z-score standardisation, and min-max standardisation and can be expressed as $[x^i(s;t) - \alpha]/\gamma$.
+It includes techniques such as centering, z-score standardisation, and min-max standardisation and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$.
 In Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardisation.
@@
 
-Distribution fit applies the cumulative distribution function $F$ of a distribution on the variable $x^i(s; t)$ to obtain the probability values $p^i(s;t) \in [0, 1]$.
+Distribution fit applies the Cumulative Distribution Function (CDF) $F$ of a distribution on the variable $x_j(s; t)$ to obtain the probability values $p_j(s;t) \in [0, 1]$.
 In SPEI, many distributions, including log-logistic, pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series.
@@
 
-<!-- This step maps the univariate series into a different scale, typically for ease of comparison across regions.
-For example, a normal scale, \[0, 1\], or \[0, 100\] may be favored for reporting certain indexes.
-In drought indexes, i.e.
-SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$.
-Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable.
-While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers.
-Also, the use of a normal scale needs to be interpreted with caution.
-@fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence.
-This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence.
--->
+Normalising applies the inverse normal CDF $\Phi^{-1}$ on the input data to obtain the normal density $z_{j}(s;t)$.
+Normalising can sometimes be confused with the scaling or variable transformation module, which do not involve using a normal distribution to transform the variable.
+It is arguably whether normalising and distribution fit should be combined or separated into two modules.
+A decision has been made to separate them into two modules given the different types of output each module present (probability values for distribution fit and normal density values for normalising).
 
@@
 
-Benchmark sets a value $b^i(s,t)$ for comparing against the original variable $x^i(s;t)$.
-This benchmark can be a fixed value consistently across space and time or determined by the data through the function $u[x^i(s;t)]$.
+Benchmark sets a value $b_j(s,t)$ for comparing against the original variable $x_j(s;t)$.
+This benchmark can be a fixed value consistently across space and time or determined by the data through the function $u[x_j(s;t)]$.
 Once a benchmark being set, observations can be highlighted for adjustments in other modules or can serve targets for monitoring and planning.
@@
 
-Simplification takes a continuous variable $x^i(s;t)$ and categorises it into a discrete set $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$ through a piecewise function,
+Simplification takes a continuous variable $x_j(s;t)$ and categorises it into a discrete set $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$ through a piecewise function,
 
@@
 ```
-
 This is typically used at the end of the index pipeline to simplify the index to communicate to the public the severity of the concept of interest measured by the index.
@@
 
-  1) Top-ranked countries with GGGI > 0.85,
+1)  Top-ranked countries with GGGI \> 0.85,
 2)  Countries ranked between 57 and 62, with GGGI values from 0.72 to 0.73, and
-  3) Low-ranked countries with GGGI < 0.6.
+3)  Low-ranked countries with GGGI \< 0.6.
 

### 2023-08-17 tidyindex.tex
use j for variable in the subscript instead in notation

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 \item
-  \(x^i(s, t)\) represents a variable of interest for example
-  precipitation, \(i = 1, \cdots, p\) and
+  \(x_j(s, t)\) represents a variable of interest for example
+  precipitation, \(j = 1, \cdots, p\) and
 \item
@@
 
-In what follows when geographic or temporal components of the \(x(s,t)\)
-process are fixed we will be using suffix notation. For example,
-\(x_{s}(t)\) represents the data for a fixed location \(s\) as a
-function of time \(t\). While \(x_{t}(s)\) denotes the spatial varying
-proccess for a fixed \(t\). An overview of the notation for pipeline
-input, operation, and output is present in Table~\ref{tbl-notation}.
+In what follows when geographic or temporal components of the
+\(x_j(s,t)\) process are fixed we will be using suffix notation. For
+example, \(x_{sj}(t)\) represents the data for a fixed location \(s\) as
+a function of time \(t\). While \(x_{tj}(s)\) denotes the spatial
+varying process for a fixed \(t\). An overview of the notation for
+pipeline input, operation, and output is present in
+Table~\ref{tbl-notation}.
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0128}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2179}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0294}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2206}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1471}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2206}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3824}}@{}}
 \caption{\label{tbl-notation}An notation overview of the input,
@@
 \endhead
-1 & Temporal processing & \(x^i_{s}(t)\) & \(f[x^i_s(t)]\) &
-\(\tilde{x}^i_s(t^\prime)\), where \(t^\prime \in D_{t^\prime}\) \\
-2 & Spatial processing & \(x^i_t(s)\) & \(g[x^i_t(s)]\) &
-\(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
-3 & Variable transformation & \(x^i(s; t)\) & \(T[x^i(s;t)]\) &
-\(Tx^i(s;t)\) ???? \\
-4 & Scaling & \(x^i(s; t)\) & \([x^i(s;t) - \alpha]/\gamma\) &
-\(Sx^i(s;t)\) ???? \\
-5 & Dimension reduction & \(\mathbf{x}(s;t)\), where
-\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
-\(h[\mathbf{x}(s;t)]\) & \(\mathbf{y}(s;t)\), where
-\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
-6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where \(F\) is
-the CDF & \(p^i(s;t)\), where \(p(.) \in [0, 1]\) is a probability
-value \\
-7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x^i(s; t)]\) &
-\(z^i(s; t)\), where \(z\) has a normal distribution \\
-8 & Benchmarking & \(x^i(s; t)\) & \(u[x^i(s;t)]\) & \(b^i(s;t)\) \\
-9 & Simplification & \(x^i(s; t)\) & \(v[x^i(s;t)]\) &
-\(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
+1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
+\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
+2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
+\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
+3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
+\(x^{\text{Trans}}_j(s;t)\) \\
+4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
+\(x^{\text{Scale}}_j(s;t)\) \\
+5 & Dimension reduction &
+\(\mathbf{x}(s;t) \quad \mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
+\(h[\mathbf{x}(s;t)]\) &
+\(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q\) \\
+6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
+\(p_j(s;t) \quad p(.) \in [0, 1]\) \\
+7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
+\(z_j(s; t)\) \\
+8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
+9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
+\(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
 \bottomrule()
@@
 The temporal processing module takes as input argument a single variable
-\(x_s^i(t)\) at location \(s\) as a function of time. In this step the
+\(x_{sj}(t)\) at location \(s\) as a function of time. In this step the
 original time series can be transformed or summarized into a new one via
 time aggregation, the transformation is represented by the function
-\(f\), \(\tilde{x}_s^i(t^\prime) = f[x^i_s(t)]\) where \(t^\prime\)
-refers to the new temporal resolution after aggregation. An example of a
-temporal processing done in the computation of the Standardized
-Precipitation Index (SPI), consists of summing the monthly precipitation
-series over a rolling time window of size \(k\). That is also known as
-the time scale. For SPI, the choice of the time scale \(k\) is used to
-control the accumulation period for the water deficit, enabling the
-assessment of drought severity across various types (meteorological,
-agricultural, and hydrological).
+\(f\), \(x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]\) where
+\(t^\prime\) refers to the new temporal resolution after aggregation. An
+example of a temporal processing done in the computation of the
+Standardized Precipitation Index (SPI), consists of summing the monthly
+precipitation series over a rolling time window of size \(k\). That is
+also known as the time scale. For SPI, the choice of the time scale
+\(k\) is used to control the accumulation period for the water deficit,
+enabling the assessment of drought severity across various types
+(meteorological, agricultural, and hydrological).
 
@@
 The spatial processing module takes a single variable with a fixed
-temporal dimension, \(x_t(s)\), as input. This step transforms the
+temporal dimension, \(x_{tj}(s)\), as input. This step transforms the
 variable from the original spatial dimension \(s\) into the new
 dimension \(s^\prime \in D_{s^\prime}\) through
-\(x^\#(s;t) = g[x(s;t)]\). The change of spatial dimension allows for
-the alignment of variables collected from different measurements, such
-as in-situ stations and satellite imagery, or originating from different
-resolutions. This also includes the aggregation of variables into
-different levels, such as city, state, and countries scales.
+\(x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]\). The change of spatial
+dimension allows for the alignment of variables collected from different
+measurements, such as in-situ stations and satellite imagery, or
+originating from different resolutions. This also includes the
+aggregation of variables into different levels, such as city, state, and
+countries scales.
 
@@
 
-Variable transformation takes input of a single variable \(x^i(s;t)\)
+Variable transformation takes input of a single variable \(x_j(s;t)\)
 and reshapes its distribution using the function \(T[.]\) to produce
-\(x^i(s;t)\). When a variable has a skewed distribution, transformations
-such as log, square root, or cubic root can adjust the distribution
-towards normality. For example, in Human Development Index (HDI), a
-logarithmic transformation is applied to the variable Gross National
-Income per capita (GNI), to reduce its impact on HDI, particularly for
-countries with high GNI values.
+\(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
+distribution, transformations such as log, square root, or cubic root
+can adjust the distribution towards normality. For example, in Human
+Development Index (HDI), a logarithmic transformation is applied to the
+variable Gross National Income per capita (GNI), to reduce its impact on
+HDI, particularly for countries with high GNI values.
 
@@
 standardisation, and min-max standardisation and can be expressed as
-\([x^i(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
+\([x_{j}(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
 three dimensions (health, education, and economy) are converted into the
@@
 
-Distribution fit applies the cumulative distribution function \(F\) of a
-distribution on the variable \(x^i(s; t)\) to obtain the probability
-values \(p^i(s;t) \in [0, 1]\). In SPEI, many distributions, including
-log-logistic, pearson III, lognormal, and general extreme distribution,
-are candidates for the aggregated series. Different fitting methods, and
-different goodness of fit tests may be used to compare the distribution
-choice on the index value.
+Distribution fit applies the Cumulative Distribution Function (CDF)
+\(F\) of a distribution on the variable \(x_j(s; t)\) to obtain the
+probability values \(p_j(s;t) \in [0, 1]\). In SPEI, many distributions,
+including log-logistic, pearson III, lognormal, and general extreme
+distribution, are candidates for the aggregated series. Different
+fitting methods, and different goodness of fit tests may be used to
+compare the distribution choice on the index value.
 
@@
 
+Normalising applies the inverse normal CDF \(\Phi^{-1}\) on the input
+data to obtain the normal density \(z_{j}(s;t)\). Normalising can
+sometimes be confused with the scaling or variable transformation
+module, which do not involve using a normal distribution to transform
+the variable. It is arguably whether normalising and distribution fit
+should be combined or separated into two modules. A decision has been
+made to separate them into two modules given the different types of
+output each module present (probability values for distribution fit and
+normal density values for normalising).
+
 \hypertarget{benchmarking}{%
@@
 
-Benchmark sets a value \(b^i(s,t)\) for comparing against the original
-variable \(x^i(s;t)\). This benchmark can be a fixed value consistently
+Benchmark sets a value \(b_j(s,t)\) for comparing against the original
+variable \(x_j(s;t)\). This benchmark can be a fixed value consistently
 across space and time or determined by the data through the function
-\(u[x^i(s;t)]\). Once a benchmark being set, observations can be
+\(u[x_j(s;t)]\). Once a benchmark being set, observations can be
 highlighted for adjustments in other modules or can serve targets for
@@
 
-Simplification takes a continuous variable \(x^i(s;t)\) and categorises
-it into a discrete set \(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\)
+Simplification takes a continuous variable \(x_j(s;t)\) and categorises
+it into a discrete set \(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\)
 through a piecewise function,

### 2023-08-17 tidyindex.qmd
software design

@@
 
-# Details of the index pipeline {#sec-index-pipeline}
+# Details of the index pipeline {#sec-pipeline}
 
@@
 
-<!-- # 1. unpack or the argumebt s  -->
+The R package `tidyindex` implements the index pipeline modules described in @sec-pipeline. These modules compute an index in a sequential manner, as shown below:
 
-<!-- # 2. construct the syntax to evalute: ~rescale_minmax(life_exp, min = xxx, max = xxxx) -->
+```{r eval = FALSE, echo = TRUE}
+DATA |> 
+  module1(...) |>
+  module2(...) |>
+  module3(...) |>
+  ...
+```
 
-<!-- # 3. evalute  -->
+Each module offers a variety of alternatives, each represented by a distinct function.
+For example, within the `dimension_reduction()` module, three methods are available: `aggregate_linear()`, `aggregate_geometrical()`, and `manual_input()` and they can be used as:
 
-<!-- # 4. update the data object  -->
+```{r eval = FALSE, , echo = TRUE}
+dimension_reduction(V1 = aggregate_linear(...))
+dimension_reduction(V2 = aggregate_geometrical(...))
+dimension_reduction(V3 = manual_input(...))
+```
 
-<!-- # [Extending the pipeline] a different name here: alternate a module with a different (method) -->
+Each method can be independently evaluated as a recipe, for example,
+
+```{r eval = FALSE, , echo = TRUE}
+manual_input(~x1 + x2)
+```
+
+takes a formula to combine the variable `x1` and `x2` and will return:
+
+````
+[1] "manual_input"
+attr(,"formula")
+[1] "x1 + x2"
+attr(,"class")
+[1] "dim_red"
+````
+
+This recipe will then be evaluated in the pipeline module with data to obtain the numerical results.
+The package also offers wrapper functions that combine multiple steps for specific indexes.
+For instance, the `idx_spi()` function bundles three steps (temporal aggregation, distribution fit, and normalising) into a single command, simplifying the syntax for computation.
+Analysts are also encouraged to create their customised indexes from existing modules.
+
+```{r eval = FALSE, echo = TRUE}
+idx_spi <- function(...){
+  DATA |> 
+    aggregate(...) |>
+    dist_fit(...) |> 
+    augment(...)
+}
+```
 
@@
 This section uses the example of drought and social indexes to show the analysis made possible with the index pipeline.
-The drought index example computes two indexes with various time scales and distributions simultaneously using the pipeline framework to understand the flood and drought events in Queensland.
-The social index example focuses on the dimension reduction in Global Gender Gap Index to explore the impact of weight changes in linear combination on index value and country ranking.
+The drought index example computes two indexes (SPI and SPEI) with various time scales and distributions simultaneously using the pipeline framework to understand the flood and drought events in Queensland.
+The second example focuses on the dimension reduction step in the Global Gender Gap Index to explore how the changes of linear combination weights affect the index values and country rankings.
 
@@
 ```{r fig-idx-tour, fig.align='center', fig.height=5}
-#| fig-cap: "Six frames selected to explore how varying the weights of the politics dimension changes the index values and country ranking in Global Gender Gap Index (GGGI).
+#| fig-cap: "Six frames selected to explore how varying the weights of the politics dimension changes the index values and country rankings in Global Gender Gap Index (GGGI).
 The top panel shows the GGGI value against the country, ranked by its original index value in Frame 12.
-The bottom panel displays the weight used to produce the index value in the top panel, with each frame corresponds to a set of weights.
+The bottom panel displays the weight used to produce the index values in the top panel, with each frame corresponds to a set of weights.
 Countries selected includes 1) top-ranked countries with GGGI > 0.85, 2) countries ranked between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked countries with GGGI < 0.6.

### 2023-08-17 tidyindex.tex
software design

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 The pipeline framework for index construction is presented in
-Section~\ref{sec-index-pipeline}. Section~\ref{sec-dev} explains how to
-add new steps into the index methodology in the form of new building
-blocks inside the index construction pipeline. Examples are given in
+\textbf{?@sec-index-pipeline}. Section~\ref{sec-dev} explains how to add
+new steps into the index methodology in the form of new building blocks
+inside the index construction pipeline. Examples are given in
 Section~\ref{sec-examples} to demonstrate the index construction with
@@
 
-\hypertarget{sec-index-pipeline}{%
-\section{Details of the index pipeline}\label{sec-index-pipeline}}
+\hypertarget{sec-pipeline}{%
+\section{Details of the index pipeline}\label{sec-pipeline}}
 
@@
 
+The R package \texttt{tidyindex} implements the index pipeline modules
+described in Section~\ref{sec-pipeline}. These modules compute an index
+in a sequential manner, as shown below:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{DATA }\SpecialCharTok{|\textgreater{}} 
+  \FunctionTok{module1}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{module2}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{module3}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
+\NormalTok{  ...}
+\end{Highlighting}
+\end{Shaded}
+
+Each module offers a variety of alternatives, each represented by a
+distinct function. For example, within the
+\texttt{dimension\_reduction()} module, three methods are available:
+\texttt{aggregate\_linear()}, \texttt{aggregate\_geometrical()}, and
+\texttt{manual\_input()} and they can be used as:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{dimension\_reduction}\NormalTok{(}\AttributeTok{V1 =} \FunctionTok{aggregate\_linear}\NormalTok{(...))}
+\FunctionTok{dimension\_reduction}\NormalTok{(}\AttributeTok{V2 =} \FunctionTok{aggregate\_geometrical}\NormalTok{(...))}
+\FunctionTok{dimension\_reduction}\NormalTok{(}\AttributeTok{V3 =} \FunctionTok{manual\_input}\NormalTok{(...))}
+\end{Highlighting}
+\end{Shaded}
+
+Each method can be independently evaluated as a recipe, for example,
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{manual\_input}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x1 }\SpecialCharTok{+}\NormalTok{ x2)}
+\end{Highlighting}
+\end{Shaded}
+
+takes a formula to combine the variable \texttt{x1} and \texttt{x2} and
+will return:
+
+\begin{verbatim}
+[1] "manual_input"
+attr(,"formula")
+[1] "x1 + x2"
+attr(,"class")
+[1] "dim_red"
+\end{verbatim}
+
+This recipe will then be evaluated in the pipeline module with data to
+obtain the numerical results. The package also offers wrapper functions
+that combine multiple steps for specific indexes. For instance, the
+\texttt{idx\_spi()} function bundles three steps (temporal aggregation,
+distribution fit, and normalising) into a single command, simplifying
+the syntax for computation. Analysts are also encouraged to create their
+customised indexes from existing modules.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{idx\_spi }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(...)\{}
+\NormalTok{  DATA }\SpecialCharTok{|\textgreater{}} 
+    \FunctionTok{aggregate}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
+    \FunctionTok{dist\_fit}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} 
+    \FunctionTok{augment}\NormalTok{(...)}
+\NormalTok{\}}
+\end{Highlighting}
+\end{Shaded}
+
 \hypertarget{sec-examples}{%
@@
 analysis made possible with the index pipeline. The drought index
-example computes two indexes with various time scales and distributions
-simultaneously using the pipeline framework to understand the flood and
-drought events in Queensland. The social index example focuses on the
-dimension reduction in Global Gender Gap Index to explore the impact of
-weight changes in linear combination on index value and country ranking.
+example computes two indexes (SPI and SPEI) with various time scales and
+distributions simultaneously using the pipeline framework to understand
+the flood and drought events in Queensland. The second example focuses
+on the dimension reduction step in the Global Gender Gap Index to
+explore how the changes of linear combination weights affect the index
+values and country rankings.
 
@@
 the weights of the politics dimension changes the index values and
-country ranking in Global Gender Gap Index (GGGI). The top panel shows
+country rankings in Global Gender Gap Index (GGGI). The top panel shows
 the GGGI value against the country, ranked by its original index value
 in Frame 12. The bottom panel displays the weight used to produce the
-index value in the top panel, with each frame corresponds to a set of
+index values in the top panel, with each frame corresponds to a set of
 weights. Countries selected includes 1) top-ranked countries with GGGI

### 2023-08-17 tidyindex.qmd
add abstract

@@
 abstract: |
-  - indexes, useful, quantify severity, early monitoring, 
-  - A huge number of indexes have been proposed by domain experts, however, a large majority of them are not being adopted, reused, and compared in research or in practice. 
-  - One of the reasons for this is the plenty of indexes are quite complex and there is no obvious easy-to-use implementation to apply them to user's data. 
-  - The paper describes a general pipeline framework to construct indexes from spatio-temporal data, 
-  - This allows all the indexes to be constructed through a uniform data pipeline and different indexes to vary on the details of each step in the data pipeline and their orders. 
-  - The pipeline proposed aim to smooth the workflow of index construction through breaking down the complicated steps proposed by various indexes into small building blocks shared by most of the indexes.
-  - The framework will be demonstrated with drought indexes as examples, but appliable in general to environmental indexes constructed from multivariate spatio-temporal data
+  Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making.
+While many work has focused on defining new indexes for the application of interest, less attention has been directed towards understanding how the a given index behave under different scenarios.
+This research introduces a data pipeline comparising 9 modules to assemble indexes from a series of modular building blocks.
+This modular data pipeline is universally applicable to index computation and allows analysts to incorporate the analysis on index into the construction pipeline.
+Analysts can compute with different index parameters, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess index robustness .
+The paper presents two examples to illustrate the usage of the pipeline.
+The first example claculates two drought indexes with different parameters to investigate the spatio-temporal distribution of drought in Queensland, Australia.
+The second example on the Global Gender Gap Index (GGGI) explores the effects of adjusting linear combination weights on a country's index values and rankings.
 keywords:

### 2023-08-17 tidyindex.tex
add abstract

@@
 \begin{abstract}
-\begin{itemize}
-\tightlist
-\item
-  indexes, useful, quantify severity, early monitoring,
-\item
-  A huge number of indexes have been proposed by domain experts,
-  however, a large majority of them are not being adopted, reused, and
-  compared in research or in practice.
-\item
-  One of the reasons for this is the plenty of indexes are quite complex
-  and there is no obvious easy-to-use implementation to apply them to
-  user's data.
-\item
-  The paper describes a general pipeline framework to construct indexes
-  from spatio-temporal data,
-\item
-  This allows all the indexes to be constructed through a uniform data
-  pipeline and different indexes to vary on the details of each step in
-  the data pipeline and their orders.
-\item
-  The pipeline proposed aim to smooth the workflow of index construction
-  through breaking down the complicated steps proposed by various
-  indexes into small building blocks shared by most of the indexes.
-\item
-  The framework will be demonstrated with drought indexes as examples,
-  but appliable in general to environmental indexes constructed from
-  multivariate spatio-temporal data
-\end{itemize}
+Indexes are useful for summarizing multivariate information into single
+metrics for monitoring, communicating, and decision-making. While many
+work has focused on defining new indexes for the application of
+interest, less attention has been directed towards understanding how the
+a given index behave under different scenarios. This research introduces
+a data pipeline comparising 9 modules to assemble indexes from a series
+of modular building blocks. This modular data pipeline is universally
+applicable to index computation and allows analysts to incorporate the
+analysis on index into the construction pipeline. Analysts can compute
+with different index parameters, adjust steps in the index definition by
+adding, removing, and swapping them to experiment with various index
+designs, calculate uncertainty measures, and assess index robustness .
+The paper presents two examples to illustrate the usage of the pipeline.
+The first example claculates two drought indexes with different
+parameters to investigate the spatio-temporal distribution of drought in
+Queensland, Australia. The second example on the Global Gender Gap Index
+(GGGI) explores the effects of adjusting linear combination weights on a
+country's index values and rankings.
 \end{abstract}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, boxrule=0pt, frame hidden, breakable, interior hidden]}{\end{tcolorbox}}\fi
 

### 2023-08-17 tidyindex.qmd
changes to the second last para in intro

@@
 For instance, the indexes constructed from linear combination of variables need to decide on the weight assigned to each variable.
-Some indexes have a spatial and/or temporal components, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes.
+Some indexes have a spatial and/or temporal component, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes.
 Hence, all these decisions can result in different index values and have different practical implications.
@@
 
-In this work, we aim at providing statistical and computational methods to develop a data pipeline framework for constructing and customize indexes using data.
-As a companion the R package, tidyindex, is developed to construct and explore different.
-Furthermore, the tidyindex package can be used to explore the effects of changes on index definition, index construction steps, and construction methods.
-This work provides researchers actively developing new indexes or aiming at improving indexes with the tools to easily modify and evaluate indexes.
-It also helps index analysts diagnose indexes, carry out sensitivity analysis to assess small perturbations in the index, and identify potential weaknesses for methodology improvement.
+In this work, we provide statistical and computational methods for developing a data pipeline framework to construct and customize indexes using data.
+The proposed pipeline comprises various modules, including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
+Given the decisions analysts need to make when combining multivariate data into indexes, the proposed pipeline enables the evaluation of how the specific choice can affect the index, as well as how the index may appear under alternative options.
+Furthermore, uncertainty calculation can also flow through the pipeline, providing the index with confidence measures.
 
-The rest of the paper is structured as follows: @sec-pipeline reviews the concept of data pipeline in R.
-The pipeline framework for index construction is presented in @sec-index-pipeline.
-@sec-dev explains how to add new steps into the index methodology in the form of new building blocks inside the index construction pipeline.
-Examples are given in @sec-examples to demonstrate the index construction with the pipeline built.
+The rest of the paper is structured as follows: @sec-tidy reviews the tidy framework in R and how index construction can benefit from such a framework.
+The details of the pipeline modules are presented in @sec-pipeline.
+@sec-dev explains the design of the `tidyindex` pacakge that implements the modules.
+Examples are given in @sec-examples to illustrate the use cases of the pipeline.
 
@@
 
-# Tidy framework {#sec-pipeline}
+# Tidy framework {#sec-tidy}
 

### 2023-08-17 tidyindex.tex
changes to the second last para in intro

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, boxrule=0pt, frame hidden, breakable, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 on the weight assigned to each variable. Some indexes have a spatial
-and/or temporal components, and variables can be aggregated to different
+and/or temporal component, and variables can be aggregated to different
 spatial resolutions and temporal scales, leading to various indexes for
@@
 
-In this work, we aim at providing statistical and computational methods
-to develop a data pipeline framework for constructing and customize
-indexes using data. As a companion the R package, tidyindex, is
-developed to construct and explore different. Furthermore, the tidyindex
-package can be used to explore the effects of changes on index
-definition, index construction steps, and construction methods. This
-work provides researchers actively developing new indexes or aiming at
-improving indexes with the tools to easily modify and evaluate indexes.
-It also helps index analysts diagnose indexes, carry out sensitivity
-analysis to assess small perturbations in the index, and identify
-potential weaknesses for methodology improvement.
-
-The rest of the paper is structured as follows:
-Section~\ref{sec-pipeline} reviews the concept of data pipeline in R.
-The pipeline framework for index construction is presented in
-\textbf{?@sec-index-pipeline}. Section~\ref{sec-dev} explains how to add
-new steps into the index methodology in the form of new building blocks
-inside the index construction pipeline. Examples are given in
-Section~\ref{sec-examples} to demonstrate the index construction with
-the pipeline built.
-
-\hypertarget{sec-pipeline}{%
-\section{Tidy framework}\label{sec-pipeline}}
+In this work, we provide statistical and computational methods for
+developing a data pipeline framework to construct and customize indexes
+using data. The proposed pipeline comprises various modules, including
+temporal and spatial aggregation, variable transformation and
+combination, distribution fitting, benchmark setting, and index
+communication. Given the decisions analysts need to make when combining
+multivariate data into indexes, the proposed pipeline enables the
+evaluation of how the specific choice can affect the index, as well as
+how the index may appear under alternative options. Furthermore,
+uncertainty calculation can also flow through the pipeline, providing
+the index with confidence measures.
+
+The rest of the paper is structured as follows: Section~\ref{sec-tidy}
+reviews the tidy framework in R and how index construction can benefit
+from such a framework. The details of the pipeline modules are presented
+in Section~\ref{sec-pipeline}. Section~\ref{sec-dev} explains the design
+of the \texttt{tidyindex} pacakge that implements the modules. Examples
+are given in Section~\ref{sec-examples} to illustrate the use cases of
+the pipeline.
+
+\hypertarget{sec-tidy}{%
+\section{Tidy framework}\label{sec-tidy}}
 

### 2023-08-17 tidyindex.qmd
go through the drought example

@@
 knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
-#options(width = 70)
+options(width = 70)
 library(tidyverse)
@@
 The details of the pipeline modules are presented in @sec-pipeline.
-@sec-dev explains the design of the `tidyindex` pacakge that implements the modules.
+@sec-software explains the design of the `tidyindex` pacakge that implements the modules.
 Examples are given in @sec-examples to illustrate the use cases of the pipeline.
@@
 
-# Software design {#sec-dev}
+# Software design {#sec-software}
 
@@
 
-A common task for drought researchers is to compute indexes at different parameter combinations.
-This can be used to identify the spatial and temporal extent of drought events, recommend the best parameter choice, or compare the effectiveness of indexes for monitoring drought.
-The example below computes two indexes: SPI and SPEI, at various time scales and fitted distributions, for stations in the state of Queensland in Australia.
-The purpose of the example is to demonstrate the interfaces the tidyindex package built to allow easy computing at different parameter combinations.
+The state of Queensland in Australia frequently experience natural disaster events such as flood and drought, which can significantly impact its agricultural industry.
+This example uses daily data from Global Historical Climatology Network Daily (GHCND), aggregated into monthly precipitation, to compute two drought indexes -- SPI and SPEI -- at various time scales and fitted distributions, for  `r length(unique(queensland$id))` stations in the state of Queensland in Australia, spanning from January 1990 to April 2022.
+This example showcases the basic calculation of indexes with different parameter specifications within the pipeline framework.
+The dataset used in this example is available in the `tidyindex` package as `queensland` and blow prints the first few rows of the data:
 
-The state of Queensland in Australia is frequently affected by natural disaster events such as flood and drought, which can have significant impacts on its agricultural industry.
-This study uses daily data from Global Historical Climatology Network Daily (GHCND), accessed via the package `rnoaa` to examine drought/flood condition in Queensland.
-Daily data is average into monthly and stations are excluded if monthly data contains missings, which is required for calculating both SPI and SPEI.
-This gives `r length(unique(queensland$id))` stations with complete records from 1990 January to 2022 April.
+```{r}
+queensland %>% head(5)
+```
 
-```{r fig-pipeline-spei}
+```{r fig-spei}
 #| fig-align: center
-#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
+#| fig-cap: "Index pipeline for two drought indexes: the Standardized Precipitation Index (SPI) and the Standardized Precipitation-Evapotranspiration Index (SPEI).
+Both indexes share similar construction steps with SPEI having two steps additional steps (variable transformation and dimension reduction) to convert temperature into evapotranspiration and combine it with the precipitation series.
+"
 #| out-height: 90%
 #| out-width: 100%
-knitr::include_graphics(here::here("figures/pipeline-spei.png"))
+knitr::include_graphics(path = here::here("figures/pipeline-spei.png"))
 ```
 
-The function `compute_indexes()` can be used to collectively compute multiple indexes.
-The tidyindex offers wrapper functions, with the prefix idx\_, that simplify the calculation of commonly used indexes by combining a set of pipeline steps into a single function.
-For example, the function `idx_spei()` includes the five steps previously described in @sec-toy-example (variable transformation, dimension reduction, temporal aggregation, distribution fit, and normalise).
-Each `idx_xxx()` function specifies the relevant parameters relevant to the index: the thornthwaite method is used to calculate PET in SPEI, with the average temperature (`tavg`) and latitude (`lat`) used as inputs.
-The SPEI is computed at four time scales (6, 12, 24, and 36 months) and fitted with two distributions (Log-logistic and General Extreme Value (GEV)).
-The SPI is also computed at the same four time scales and uses the default gamma distribution to fit the aggregated series.
+@fig-spei illustrates the pipeline steps of the two indexes.
+The two indexes are similar with the distinct that SPEI involves two additional steps -- variable transformation and dimension reduction -- prior to temporal processing.
+As introduced in @sec-software, wrapper functions are available for both indexes as `idx_spi()` and `idx_spei()`, which allows for the specification of different time scales and distributions for fitting the aggregated series.
+In `tidyindex`, multiple indexes can be calculated collectively using the function `compute_indexes()`.
+Both SPI and SPEI are calculated across four time scales (6, 12, 24, and 36 months).
+The SPEI is fitted with two distributions (log-logistic and general extreme value distribution) and the gamma distribution is used for SPI:
 
@@
 
-The output from `compute_indexes()` contains index values and associated parameter in a long tibble.
-It includes the original variables (`id`, `ym`, `prcp`, `tmax`, `tmin`, `tavg`, `long`, `lat`, and `name`), index parameters (`.idx`, `.scale`, `.method`, and `.dist`), intermediate variables (`pet`, `.agg`, and `.fitted`), and the final index (`.index`).
-This data can be visualised across space or time, or simultaneously, to explore the wet/dry condition in Queensland.
-@fig-compute-spatial visualises the spatial distribution of SPI at two periods (2010 October - 2011 March and 2019 October - 2020 March) with significant natural disaster events: 2010/11 Queensland flood and 2019 Australia drought, which contributes to the notorious 2019/20 bushfire.
-@fig-compute-temporal displays the sensitivity of the SPEI series for one particular station, Texas post office, at different time scales and fitted distributions.
-These two plots demonstrate some possibilities to explore the indexes after they are computed from `compute_indexes()`.
+```{r}
+load(here::here("data/idx.rda"))
+```
+
+```{r}
+idx
+```
+
+The output contains the original data, index values (`.index`), parameter used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`pet`, `.agg`, and `.fitted`).
+This data can be visualized to investigate the spatio-temporal distribution of the drought/flood events, as well as the response of index values to different time scales and distribution parameters at specific single locations.
+@fig-compute-spatial and @fig-compute-temporal exemplify two possibilities.
+@fig-compute-spatial presents the spatial distribution of SPI during two periods: October 2010 to March 2011 for the 2010/11 Queensland flood and October 2019 to March 2020 for the 2019 Australia drought, which contributes to the notorious 2019/20 bushfire.
+@fig-compute-temporal displays the sensitivity of the SPEI series at the Texas post office to different time scales and fitted distributions.
+Larger time scales produce a smoother index across time, however, all time scales indicate an extreme drought (corresponding to -2 in SPEI) in 2020, confirming the severity of the drought across different time horizons.
+Moreover, the chosen distribution has less influence on the index, with general extreme value distribution tending to produce more extreme outcomes than log logistic distribution for the extreme events (index > 2 or <-2).
 
@@
 #| fig-width: 6
-load(here::here("data/idx.rda"))
 queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 

### 2023-08-17 tidyindex.tex
go through the drought example

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, sharp corners, boxrule=0pt, enhanced, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 from such a framework. The details of the pipeline modules are presented
-in Section~\ref{sec-pipeline}. Section~\ref{sec-dev} explains the design
-of the \texttt{tidyindex} pacakge that implements the modules. Examples
-are given in Section~\ref{sec-examples} to illustrate the use cases of
-the pipeline.
+in Section~\ref{sec-pipeline}. Section~\ref{sec-software} explains the
+design of the \texttt{tidyindex} pacakge that implements the modules.
+Examples are given in Section~\ref{sec-examples} to illustrate the use
+cases of the pipeline.
 
@@
 
-\hypertarget{sec-dev}{%
-\section{Software design}\label{sec-dev}}
+\hypertarget{sec-software}{%
+\section{Software design}\label{sec-software}}
 
@@
 
-A common task for drought researchers is to compute indexes at different
-parameter combinations. This can be used to identify the spatial and
-temporal extent of drought events, recommend the best parameter choice,
-or compare the effectiveness of indexes for monitoring drought. The
-example below computes two indexes: SPI and SPEI, at various time scales
-and fitted distributions, for stations in the state of Queensland in
-Australia. The purpose of the example is to demonstrate the interfaces
-the tidyindex package built to allow easy computing at different
-parameter combinations.
-
-The state of Queensland in Australia is frequently affected by natural
-disaster events such as flood and drought, which can have significant
-impacts on its agricultural industry. This study uses daily data from
-Global Historical Climatology Network Daily (GHCND), accessed via the
-package \texttt{rnoaa} to examine drought/flood condition in Queensland.
-Daily data is average into monthly and stations are excluded if monthly
-data contains missings, which is required for calculating both SPI and
-SPEI. This gives 29 stations with complete records from 1990 January to
-2022 April.
+The state of Queensland in Australia frequently experience natural
+disaster events such as flood and drought, which can significantly
+impact its agricultural industry. This example uses daily data from
+Global Historical Climatology Network Daily (GHCND), aggregated into
+monthly precipitation, to compute two drought indexes -- SPI and SPEI --
+at various time scales and fitted distributions, for 29 stations in the
+state of Queensland in Australia, spanning from January 1990 to April
+2022. This example showcases the basic calculation of indexes with
+different parameter specifications within the pipeline framework. The
+dataset used in this example is available in the \texttt{tidyindex}
+package as \texttt{queensland} and blow prints the first few rows of the
+data:
+
+\begin{verbatim}
+# A tibble: 5 x 9
+  id                ym  prcp  tmax  tmin  tavg  long   lat name       
+  <chr>          <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl> <chr>      
+1 ASN00029038 1990 Jan  1682  34.3  24.7  29.5  142. -15.5 KOWANYAMA ~
+2 ASN00029038 1990 Feb   416  35.2  23.2  29.2  142. -15.5 KOWANYAMA ~
+3 ASN00029038 1990 Mar  2026  32.5  23.6  28.0  142. -15.5 KOWANYAMA ~
+4 ASN00029038 1990 Apr   597  32.9  17.7  25.3  142. -15.5 KOWANYAMA ~
+5 ASN00029038 1990 May   244  31.8  20.1  25.9  142. -15.5 KOWANYAMA ~
+\end{verbatim}
 
@@
 
-\caption{\label{fig-pipeline-spei}Diagram of pipeline steps for index
-construction. will need to be updated with better design and the
-distribution fitting step.}
+\caption{\label{fig-spei}Index pipeline for two drought indexes: the
+Standardized Precipitation Index (SPI) and the Standardized
+Precipitation-Evapotranspiration Index (SPEI). Both indexes share
+similar construction steps with SPEI having two steps additional steps
+(variable transformation and dimension reduction) to convert temperature
+into evapotranspiration and combine it with the precipitation series.}
 
@@
 
-The function \texttt{compute\_indexes()} can be used to collectively
-compute multiple indexes. The tidyindex offers wrapper functions, with
-the prefix idx\_, that simplify the calculation of commonly used indexes
-by combining a set of pipeline steps into a single function. For
-example, the function \texttt{idx\_spei()} includes the five steps
-previously described in \textbf{?@sec-toy-example} (variable
-transformation, dimension reduction, temporal aggregation, distribution
-fit, and normalise). Each \texttt{idx\_xxx()} function specifies the
-relevant parameters relevant to the index: the thornthwaite method is
-used to calculate PET in SPEI, with the average temperature
-(\texttt{tavg}) and latitude (\texttt{lat}) used as inputs. The SPEI is
-computed at four time scales (6, 12, 24, and 36 months) and fitted with
-two distributions (Log-logistic and General Extreme Value (GEV)). The
-SPI is also computed at the same four time scales and uses the default
-gamma distribution to fit the aggregated series.
+Figure~\ref{fig-spei} illustrates the pipeline steps of the two indexes.
+The two indexes are similar with the distinct that SPEI involves two
+additional steps -- variable transformation and dimension reduction --
+prior to temporal processing. As introduced in
+Section~\ref{sec-software}, wrapper functions are available for both
+indexes as \texttt{idx\_spi()} and \texttt{idx\_spei()}, which allows
+for the specification of different time scales and distributions for
+fitting the aggregated series. In \texttt{tidyindex}, multiple indexes
+can be calculated collectively using the function
+\texttt{compute\_indexes()}. Both SPI and SPEI are calculated across
+four time scales (6, 12, 24, and 36 months). The SPEI is fitted with two
+distributions (log-logistic and general extreme value distribution) and
+the gamma distribution is used for SPI:
 
@@
 
-The output from \texttt{compute\_indexes()} contains index values and
-associated parameter in a long tibble. It includes the original
-variables (\texttt{id}, \texttt{ym}, \texttt{prcp}, \texttt{tmax},
-\texttt{tmin}, \texttt{tavg}, \texttt{long}, \texttt{lat}, and
-\texttt{name}), index parameters (\texttt{.idx}, \texttt{.scale},
-\texttt{.method}, and \texttt{.dist}), intermediate variables
-(\texttt{pet}, \texttt{.agg}, and \texttt{.fitted}), and the final index
-(\texttt{.index}). This data can be visualised across space or time, or
-simultaneously, to explore the wet/dry condition in Queensland.
-Figure~\ref{fig-compute-spatial} visualises the spatial distribution of
-SPI at two periods (2010 October - 2011 March and 2019 October - 2020
-March) with significant natural disaster events: 2010/11 Queensland
-flood and 2019 Australia drought, which contributes to the notorious
+\begin{verbatim}
+# A tibble: 128,586 x 19
+   .idx  .period id             ym  prcp  tmax  tmin  tavg  long   lat
+   <chr>   <dbl> <chr>       <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
+ 1 spei        6 ASN0002~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5
+ 2 spei        6 ASN0002~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5
+ 3 spei        6 ASN0002~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7
+ 4 spei        6 ASN0002~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7
+ 5 spei        6 ASN0003~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9
+ 6 spei        6 ASN0003~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9
+ 7 spei        6 ASN0003~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1
+ 8 spei        6 ASN0003~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1
+ 9 spei        6 ASN0003~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6
+10 spei        6 ASN0003~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6
+# i 128,576 more rows
+# i 9 more variables: name <chr>, pet <dbl>, diff <dbl>,
+#   .scale <dbl>, .agg <dbl>, .method <chr>, .fitted <dbl>,
+#   .dist <chr>, .index <dbl>
+\end{verbatim}
+
+The output contains the original data, index values (\texttt{.index}),
+parameter used (\texttt{.scale}, \texttt{.method}, and \texttt{.dist}),
+and all the intermediate variables (\texttt{pet}, \texttt{.agg}, and
+\texttt{.fitted}). This data can be visualized to investigate the
+spatio-temporal distribution of the drought/flood events, as well as the
+response of index values to different time scales and distribution
+parameters at specific single locations.
+Figure~\ref{fig-compute-spatial} and Figure~\ref{fig-compute-temporal}
+exemplify two possibilities. Figure~\ref{fig-compute-spatial} presents
+the spatial distribution of SPI during two periods: October 2010 to
+March 2011 for the 2010/11 Queensland flood and October 2019 to March
+2020 for the 2019 Australia drought, which contributes to the notorious
 2019/20 bushfire. Figure~\ref{fig-compute-temporal} displays the
-sensitivity of the SPEI series for one particular station, Texas post
-office, at different time scales and fitted distributions. These two
-plots demonstrate some possibilities to explore the indexes after they
-are computed from \texttt{compute\_indexes()}.
+sensitivity of the SPEI series at the Texas post office to different
+time scales and fitted distributions. Larger time scales produce a
+smoother index across time, however, all time scales indicate an extreme
+drought (corresponding to -2 in SPEI) in 2020, confirming the severity
+of the drought across different time horizons. Moreover, the chosen
+distribution has less influence on the index, with general extreme value
+distribution tending to produce more extreme outcomes than log logistic
+distribution for the extreme events (index \textgreater{} 2 or
+\textless-2).
 

### 2023-08-17 tidyindex.qmd
go through grammarly

@@
 Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making.
-Examples of those include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings or Human Development Index among many others.
+Examples of those include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, or Human Development Index among many others.
 
 To construct an index, experts typically start by defining a concept of interest that requires measurement.
-This concept often lacks a direct measurable attribute or can only be measure as a composite of various processes, yet it holds social and public significance.
-To create an index, once the underlying processes involved are identified, relevant and available variables are then defined, collected, and combined using statistical methods into an index that aims at measure the process of interest.
+This concept often lacks a direct measurable attribute or can only be measured as a composite of various processes, yet it holds social and public significance.
+To create an index, once the underlying processes involved are identified, relevant and available variables are then defined, collected, and combined using statistical methods into an index that aims to measure the process of interest.
 The construction process is often not straightforward, and decisions need to be made, such as the selection of variables to be included, which might depend on data availability and the statistical definition of the index to be used, among others.
-For instance, the indexes constructed from linear combination of variables need to decide on the weight assigned to each variable.
+For instance, the indexes constructed from a linear combination of variables need to decide on the weight assigned to each variable.
 Some indexes have a spatial and/or temporal component, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes.
@@
 
-To be able to test different decision choices systematically for an index, the index needs to be broken down into its fundamental building blocks to analyse the contribution and effect of each component.
+To be able to test different decision choices systematically for an index, the index needs to be broken down into its fundamental building blocks to analyze the contribution and effect of each component.
 We call this process of breaking the index construction into different steps the index pipeline.
-Such decomposition of index components provides the means to standardise index construction via a pipeline and offers benefits for comparing among indexes and calculating index uncertainty.
-In social indexes for instance, the OECD handbook [@oecd_handbook_2008] has provided a set of steps and recommendations for constructing composite socio-economic indexes.
+Such decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes and calculating index uncertainty.
+In social indexes, for instance, the OECD handbook [@oecd_handbook_2008] has provided a set of steps and recommendations for constructing composite socio-economic indexes.
 However, there is still a need to extend these guidelines and methods to accommodate the inclusion of more methodological complex steps required for indexes in general.
@@
 The details of the pipeline modules are presented in @sec-pipeline.
-@sec-software explains the design of the `tidyindex` pacakge that implements the modules.
+@sec-software explains the design of the `tidyindex` package that implements the modules.
 Examples are given in @sec-examples to illustrate the use cases of the pipeline.
@@
 The tidy framework consists of two key components: tidy data and tidy tools.
-The concept of tidy data [@wickham_tidy_2014] prescribes specific rules for organising data in an analysis, with observations as rows, variables as columns, and types of observational units as tables.
-Tidy tools, on the other hand, are concatenated in a sequence through which the tidy data flows, creating a pipeline for data processing and modelling.
-These pipelines are data centric, meaning all the tidy tools or functions take a tidy data object as inputs and return a processed tidy data object, directly ready for the next operations to be applied.
-Also, the pipeline approach corresponds to the modular programming practice, which breaks down complex problems into smaller and more manageable pieces, as oppose to a monolithic design, where all the steps are predetermined and integrated into a single piece.
+The concept of tidy data [@wickham_tidy_2014] prescribes specific rules for organizing data in an analysis, with observations as rows, variables as columns, and types of observational units as tables.
+Tidy tools, on the other hand, are concatenated in a sequence through which the tidy data flows, creating a pipeline for data processing and modeling.
+These pipelines are data-centric, meaning all the tidy tools or functions take a tidy data object as input and return a processed tidy data object, directly ready for the next operations to be applied.
+Also, the pipeline approach corresponds to the modular programming practice, which breaks down complex problems into smaller and more manageable pieces, as opposed to a monolithic design, where all the steps are predetermined and integrated into a single piece.
 The flexibility provided by the modularity makes it easier to modify certain steps in the pipeline and to maintain and extend the code base.
@@
 Examples of using a pipeline approach for data analysis can be traced back to the interactive graphics literature, including @buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009.
-@wickham_plumbing_2009 argues that whether made explicit or not, a pipeline has to be presented in every graphics program and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems.
-While this comment is made in the context of interactive graphics program, it is also applicable generally to any data analysis workflow.
+@wickham_plumbing_2009 argue that whether made explicit or not, a pipeline has to be presented in every graphics program, and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems.
+While this comment is made in the context of interactive graphics programs, it is also applicable generally to any data analysis workflow.
 More recently, the tidyverse suite [@wickham_welcome_2019] takes the pipeline approach for general-purpose data wrangling and has gained popularity within the R community.
-The pipeline-style code can be directly read as a series of operations applied successively on tidy data object, offering a method to document the data wrangling process with all the computational details for reproducibility.
+The pipeline-style code can be directly read as a series of operations applied successively on tidy data objects, offering a method to document the data wrangling process with all the computational details for reproducibility.
 
@@
 These components are then implemented as modules, which can be combined to form the pipeline.
-For example, in supervised machine learning models, steps such as data splitting, model training and model evaluation are commonly used in most workflow.
+For example, in supervised machine learning models, steps such as data splitting, model training, and model evaluation are commonly used in most workflow.
 In the `tidymodels`, these steps are correspondingly implemented as package `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen.
@@
 -   enables systematic comparison of surrogate indexes designed to measure the same phenomenon.
--   it may even be possible to automate diagramatic explanations and documentation of the index.
+-   it may even be possible to automate diagrammatic explanations and documentation of the index.
 
-Adoption of this pipeline approach would provide uniformity to the field of index development, research and application.
+Adoption of this pipeline approach would provide uniformity to the field of index development, research, and application.
 
@@
 
-Now, we introduce the notation used for describing pipeline modules. Consider multivariate spatio-temporal process,
+Now, we introduce the notation used for describing pipeline modules. Consider a multivariate spatio-temporal process,
 
@@
 
--   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. For instance, time measurements could be recorded hourly, yearly, monthly, quaterly or by season.
+-   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. For instance, time measurements could be recorded hourly, yearly, monthly, quarterly, or by season.
 
@@
 In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation.
-An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$.
+An example of temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$.
 That is also known as the time scale.
@@
 The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions.
-This also includes the aggregation of variables into different levels, such as city, state, and countries scales.
+This also includes the aggregation of variables into different levels, such as city, state, and country scales.
 
@@
 
-Variable transformation takes input of a single variable $x_j(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^{\text{Trans}}_{j}(s;t)$.
+Variable transformation takes the input of a single variable $x_j(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^{\text{Trans}}_{j}(s;t)$.
 When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality.
@@
 Unlike variable transformation, scaling maintains the distributional shape of the variable.
-It includes techniques such as centering, z-score standardisation, and min-max standardisation and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$.
-In Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardisation.
+It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$.
+In Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.
 
@@
 Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$, as the input.
-It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < q$, as the output.
-The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods.
+It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < q$, as the output.
+The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods.
 For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
@@
 Distribution fit applies the Cumulative Distribution Function (CDF) $F$ of a distribution on the variable $x_j(s; t)$ to obtain the probability values $p_j(s;t) \in [0, 1]$.
-In SPEI, many distributions, including log-logistic, pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series.
-Different fitting methods, and different goodness of fit tests may be used to compare the distribution choice on the index value.
+In SPEI, many distributions, including log-logistic, Pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series.
+Different fitting methods and different goodness of fit tests may be used to compare the distribution choice on the index value.
 
@@
 
-Normalising applies the inverse normal CDF $\Phi^{-1}$ on the input data to obtain the normal density $z_{j}(s;t)$.
-Normalising can sometimes be confused with the scaling or variable transformation module, which do not involve using a normal distribution to transform the variable.
-It is arguably whether normalising and distribution fit should be combined or separated into two modules.
-A decision has been made to separate them into two modules given the different types of output each module present (probability values for distribution fit and normal density values for normalising).
+Normalizing applies the inverse normal CDF $\Phi^{-1}$ on the input data to obtain the normal density $z_{j}(s;t)$.
+Normalizing can sometimes be confused with the scaling or variable transformation module, which does not involve using a normal distribution to transform the variable.
+It is arguably whether normalizing and distribution fit should be combined or separated into two modules.
+A decision has been made to separate them into two modules given the different types of output each module presents (probability values for distribution fit and normal density values for normalizing).
 
@@
 This benchmark can be a fixed value consistently across space and time or determined by the data through the function $u[x_j(s;t)]$.
-Once a benchmark being set, observations can be highlighted for adjustments in other modules or can serve targets for monitoring and planning.
+Once a benchmark is set, observations can be highlighted for adjustments in other modules or can serve as targets for monitoring and planning.
 
@@
 
-takes a formula to combine the variable `x1` and `x2` and will return:
+takes a formula to combine the variables `x1` and `x2` and return:
 
@@
 
-This recipe will then be evaluated in the pipeline module with data to obtain the numerical results.
+This recipe will then be evaluated in the pipeline module with data to obtain numerical results.
 The package also offers wrapper functions that combine multiple steps for specific indexes.
-For instance, the `idx_spi()` function bundles three steps (temporal aggregation, distribution fit, and normalising) into a single command, simplifying the syntax for computation.
-Analysts are also encouraged to create their customised indexes from existing modules.
+For instance, the `idx_spi()` function bundles three steps (temporal aggregation, distribution fit, and normalizing) into a single command, simplifying the syntax for computation.
+Analysts are also encouraged to create customized indexes from existing modules.
 
@@
 The drought index example computes two indexes (SPI and SPEI) with various time scales and distributions simultaneously using the pipeline framework to understand the flood and drought events in Queensland.
-The second example focuses on the dimension reduction step in the Global Gender Gap Index to explore how the changes of linear combination weights affect the index values and country rankings.
+The second example focuses on the dimension reduction step in the Global Gender Gap Index to explore how the changes in linear combination weights affect the index values and country rankings.
 
@@
 
-The state of Queensland in Australia frequently experience natural disaster events such as flood and drought, which can significantly impact its agricultural industry.
+The state of Queensland in Australia frequently experiences natural disaster events such as flood and drought, which can significantly impact its agricultural industry.
 This example uses daily data from Global Historical Climatology Network Daily (GHCND), aggregated into monthly precipitation, to compute two drought indexes -- SPI and SPEI -- at various time scales and fitted distributions, for  `r length(unique(queensland$id))` stations in the state of Queensland in Australia, spanning from January 1990 to April 2022.
@@
 
-The output contains the original data, index values (`.index`), parameter used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`pet`, `.agg`, and `.fitted`).
+The output contains the original data, index values (`.index`), parameters used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`pet`, `.agg`, and `.fitted`).
 This data can be visualized to investigate the spatio-temporal distribution of the drought/flood events, as well as the response of index values to different time scales and distribution parameters at specific single locations.
@@
 Larger time scales produce a smoother index across time, however, all time scales indicate an extreme drought (corresponding to -2 in SPEI) in 2020, confirming the severity of the drought across different time horizons.
-Moreover, the chosen distribution has less influence on the index, with general extreme value distribution tending to produce more extreme outcomes than log logistic distribution for the extreme events (index > 2 or <-2).
+Moreover, the chosen distribution has less influence on the index, with general extreme value distribution tending to produce more extreme outcomes than log-logistic distribution for the extreme events (index > 2 or <-2).
 
@@
 #| fig-cap: "Spatial distribution of Standardized Precipitation Index (SPI-12) in Queensland, Australia during two major flood and drought events: 2010/11 and 2019/20.
-The map shows a continous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 Janurary, likely due to the increased rainfall in February from the meteorological record."
+The map shows a continuous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 January, likely due to the increased rainfall in February from the meteorological record."
 #| fig-width: 6
@@
 The dashed line at -2 represents the class "extreme drought" by the SPEI.
-A larger time scale gives a smoother index series, while also takes longer to recover from an extreme situation as seen in the  2019/20 drought period.
-The SPEI values from two distribution fits mostly agree, while GEV can results in more extreme values, i.e.
+A larger time scale gives a smoother index series, while also taking longer to recover from an extreme situation as seen in the  2019/20 drought period.
+The SPEI values from the two distribution fit mostly agree, while GEV can result in more extreme values, i.e.
 in 1998 and 2020.'
@@
 The index, compiled from 14 variables expressed as female-to-male ratios, first aggregates these variables into four dimensions through linear combinations.
-These dimensions are then combined though another linear combination to form the index.
+These dimensions are then combined through another linear combination to form the index.
 @fig-pp-gggi illustrates this pipeline construction by applying the dimension reduction module twice on the data to generate the index.
-@tbl-gggi-weights presents the variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction.
+@tbl-gggi-weights presents the variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction steps.
 In the table, the variable weights are computed as the inverse of the standard deviation of each variable, scaled to sum to 1.
-These weights ensure that a one percentage point change in the standard deviation of each variable to contribute equally to the index.
-Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weight.
+These weights ensure that a one percentage point change in the standard deviation of each variable contributes equally to the index.
+Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weights.
 
@@
 The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi` and @tbl-gggi-weights as `gggi_weights`.
-The index can be reproduced with the pacakge as:
+The index can be reproduced with the package as:
 
@@
 
-After initialising the `gggi` object and attaching the `gggi_weights` as metadata, a single linear combination step within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata.
+After initialzing the `gggi` object and attaching the `gggi_weights` as meta-data, a single linear combination step within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata.
 While computing the index from the original 14 variables, it remains unclear how the missing values are handled within the index, which impacts 68 out of the total 146 countries.
@@
 A linear combination can also be interpreted as a linear projection of multivariate information with a weight vector.
-Projecting data from higher to lower dimension unavoidably leads to information loss and the weights used require careful examination to understand their effects on the index and implications for interpretation and decision-making.
+Projecting data from higher to lower dimensions unavoidably leads to information loss and the weights used require careful examination to understand their effects on the index and implications for interpretation and decision-making.
 By making slight adjustments to the weight vector, we can observe how index values and country rankings change.
-For illustration, we select a set countries including:
+For illustration, we select a set of countries including:
 
@@
 This process creates an animation showing the movement of index values in response to changing weights.
-This visualisation technique, which presents a sequence of data projections, is referred to as a "tour".
+This visualization technique, which presents a sequence of data projections, is referred to as a "tour".
 The specific kind of tour that moves between the original and nearby projections is known as a "local tour".
@@
 
-In GGGI, the index value has a direct interpretation as the percentage of gender gap that has been closed.
-When countries exhibit closer index values or consistent, it can be directly interpreted as a reflection of a country's progress toward gender parity.
+In GGGI, the index value has a direct interpretation as the percentage of the gender gap that has been closed.
+When countries exhibit closer or consistent decreasing index values, it can be directly interpreted as a reflection of a country's progress toward gender parity.
 Ideally, an index should be robust against minor changes in its construction components.
-This example provides researchers with means to observe changes in the index resulting from variations in the weights used in linear combination.
+This example provides researchers with means to observe changes in the index resulting from variations in the weights used in linear combinations.
 This approach can be applied broadly to different sets of weights of interest, extending beyond the change in a single dimension illustrated by the example.

### 2023-08-17 tidyindex.tex
go through grammarly

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 Niño-Southern Oscillation Index, Consumer Price Index, QS University
-Rankings or Human Development Index among many others.
+Rankings, or Human Development Index among many others.
 
@@
 interest that requires measurement. This concept often lacks a direct
-measurable attribute or can only be measure as a composite of various
+measurable attribute or can only be measured as a composite of various
 processes, yet it holds social and public significance. To create an
@@
 and available variables are then defined, collected, and combined using
-statistical methods into an index that aims at measure the process of
+statistical methods into an index that aims to measure the process of
 interest. The construction process is often not straightforward, and
@@
 definition of the index to be used, among others. For instance, the
-indexes constructed from linear combination of variables need to decide
-on the weight assigned to each variable. Some indexes have a spatial
-and/or temporal component, and variables can be aggregated to different
-spatial resolutions and temporal scales, leading to various indexes for
-different monitoring purposes. Hence, all these decisions can result in
-different index values and have different practical implications.
+indexes constructed from a linear combination of variables need to
+decide on the weight assigned to each variable. Some indexes have a
+spatial and/or temporal component, and variables can be aggregated to
+different spatial resolutions and temporal scales, leading to various
+indexes for different monitoring purposes. Hence, all these decisions
+can result in different index values and have different practical
+implications.
 
@@
 index, the index needs to be broken down into its fundamental building
-blocks to analyse the contribution and effect of each component. We call
+blocks to analyze the contribution and effect of each component. We call
 this process of breaking the index construction into different steps the
 index pipeline. Such decomposition of index components provides the
-means to standardise index construction via a pipeline and offers
+means to standardize index construction via a pipeline and offers
 benefits for comparing among indexes and calculating index uncertainty.
-In social indexes for instance, the OECD handbook (OECD, European Union,
-and Joint Research Centre - European Commission 2008) has provided a set
-of steps and recommendations for constructing composite socio-economic
-indexes. However, there is still a need to extend these guidelines and
-methods to accommodate the inclusion of more methodological complex
-steps required for indexes in general.
+In social indexes, for instance, the OECD handbook (OECD, European
+Union, and Joint Research Centre - European Commission 2008) has
+provided a set of steps and recommendations for constructing composite
+socio-economic indexes. However, there is still a need to extend these
+guidelines and methods to accommodate the inclusion of more
+methodological complex steps required for indexes in general.
 
@@
 in Section~\ref{sec-pipeline}. Section~\ref{sec-software} explains the
-design of the \texttt{tidyindex} pacakge that implements the modules.
+design of the \texttt{tidyindex} package that implements the modules.
 Examples are given in Section~\ref{sec-examples} to illustrate the use
@@
 tools. The concept of tidy data (Wickham 2014) prescribes specific rules
-for organising data in an analysis, with observations as rows, variables
+for organizing data in an analysis, with observations as rows, variables
 as columns, and types of observational units as tables. Tidy tools, on
 the other hand, are concatenated in a sequence through which the tidy
-data flows, creating a pipeline for data processing and modelling. These
-pipelines are data centric, meaning all the tidy tools or functions take
-a tidy data object as inputs and return a processed tidy data object,
+data flows, creating a pipeline for data processing and modeling. These
+pipelines are data-centric, meaning all the tidy tools or functions take
+a tidy data object as input and return a processed tidy data object,
 directly ready for the next operations to be applied. Also, the pipeline
 approach corresponds to the modular programming practice, which breaks
-down complex problems into smaller and more manageable pieces, as oppose
-to a monolithic design, where all the steps are predetermined and
-integrated into a single piece. The flexibility provided by the
+down complex problems into smaller and more manageable pieces, as
+opposed to a monolithic design, where all the steps are predetermined
+and integrated into a single piece. The flexibility provided by the
 modularity makes it easier to modify certain steps in the pipeline and
@@
 (1988); Sutherland et al. (2000); Xie, Hofmann, and Cheng (2014);
-Wickham et al. (2009). Wickham et al. (2009) argues that whether made
+Wickham et al. (2009). Wickham et al. (2009) argue that whether made
 explicit or not, a pipeline has to be presented in every graphics
-program and making them explicit is beneficial for understanding the
+program, and making them explicit is beneficial for understanding the
 implementation and comparing between different graphic systems. While
-this comment is made in the context of interactive graphics program, it
+this comment is made in the context of interactive graphics programs, it
 is also applicable generally to any data analysis workflow. More
@@
 within the R community. The pipeline-style code can be directly read as
-a series of operations applied successively on tidy data object,
+a series of operations applied successively on tidy data objects,
 offering a method to document the data wrangling process with all the
@@
 form the pipeline. For example, in supervised machine learning models,
-steps such as data splitting, model training and model evaluation are
+steps such as data splitting, model training, and model evaluation are
 commonly used in most workflow. In the \texttt{tidymodels}, these steps
@@
 \item
-  it may even be possible to automate diagramatic explanations and
+  it may even be possible to automate diagrammatic explanations and
   documentation of the index.
@@
 Adoption of this pipeline approach would provide uniformity to the field
-of index development, research and application.
+of index development, research, and application.
 
@@
 Now, we introduce the notation used for describing pipeline modules.
-Consider multivariate spatio-temporal process,
+Consider a multivariate spatio-temporal process,
 
@@
   For instance, time measurements could be recorded hourly, yearly,
-  monthly, quaterly or by season.
+  monthly, quarterly, or by season.
 \end{itemize}
@@
 \(t^\prime\) refers to the new temporal resolution after aggregation. An
-example of a temporal processing done in the computation of the
+example of temporal processing done in the computation of the
 Standardized Precipitation Index (SPI), consists of summing the monthly
@@
 aggregation of variables into different levels, such as city, state, and
-countries scales.
+country scales.
 
@@
 
-Variable transformation takes input of a single variable \(x_j(s;t)\)
-and reshapes its distribution using the function \(T[.]\) to produce
-\(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
+Variable transformation takes the input of a single variable
+\(x_j(s;t)\) and reshapes its distribution using the function \(T[.]\)
+to produce \(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
 distribution, transformations such as log, square root, or cubic root
@@
 shape of the variable. It includes techniques such as centering, z-score
-standardisation, and min-max standardisation and can be expressed as
+standardization, and min-max standardization and can be expressed as
 \([x_{j}(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
 three dimensions (health, education, and economy) are converted into the
-same scale (0-1) using min-max standardisation.
+same scale (0-1) using min-max standardization.
 
@@
 \(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\), as the input. It
-summarises the high-dimensional information into a lower dimension
+summarises the high-dimensional information into a lower-dimension
 representation \(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\)
@@
 domain-specific knowledge, originating from theories describing the
-underlying physical processes, or guided in statistical methods. For
+underlying physical processes, or guided by statistical methods. For
 example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
@@
 probability values \(p_j(s;t) \in [0, 1]\). In SPEI, many distributions,
-including log-logistic, pearson III, lognormal, and general extreme
+including log-logistic, Pearson III, lognormal, and general extreme
 distribution, are candidates for the aggregated series. Different
-fitting methods, and different goodness of fit tests may be used to
+fitting methods and different goodness of fit tests may be used to
 compare the distribution choice on the index value.
@@
 
-Normalising applies the inverse normal CDF \(\Phi^{-1}\) on the input
-data to obtain the normal density \(z_{j}(s;t)\). Normalising can
+Normalizing applies the inverse normal CDF \(\Phi^{-1}\) on the input
+data to obtain the normal density \(z_{j}(s;t)\). Normalizing can
 sometimes be confused with the scaling or variable transformation
-module, which do not involve using a normal distribution to transform
-the variable. It is arguably whether normalising and distribution fit
+module, which does not involve using a normal distribution to transform
+the variable. It is arguably whether normalizing and distribution fit
 should be combined or separated into two modules. A decision has been
 made to separate them into two modules given the different types of
-output each module present (probability values for distribution fit and
-normal density values for normalising).
+output each module presents (probability values for distribution fit and
+normal density values for normalizing).
 
@@
 across space and time or determined by the data through the function
-\(u[x_j(s;t)]\). Once a benchmark being set, observations can be
-highlighted for adjustments in other modules or can serve targets for
+\(u[x_j(s;t)]\). Once a benchmark is set, observations can be
+highlighted for adjustments in other modules or can serve as targets for
 monitoring and planning.
@@
 
-takes a formula to combine the variable \texttt{x1} and \texttt{x2} and
-will return:
+takes a formula to combine the variables \texttt{x1} and \texttt{x2} and
+return:
 
@@
 This recipe will then be evaluated in the pipeline module with data to
-obtain the numerical results. The package also offers wrapper functions
-that combine multiple steps for specific indexes. For instance, the
+obtain numerical results. The package also offers wrapper functions that
+combine multiple steps for specific indexes. For instance, the
 \texttt{idx\_spi()} function bundles three steps (temporal aggregation,
-distribution fit, and normalising) into a single command, simplifying
-the syntax for computation. Analysts are also encouraged to create their
-customised indexes from existing modules.
+distribution fit, and normalizing) into a single command, simplifying
+the syntax for computation. Analysts are also encouraged to create
+customized indexes from existing modules.
 
@@
 on the dimension reduction step in the Global Gender Gap Index to
-explore how the changes of linear combination weights affect the index
+explore how the changes in linear combination weights affect the index
 values and country rankings.
@@
 
-The state of Queensland in Australia frequently experience natural
+The state of Queensland in Australia frequently experiences natural
 disaster events such as flood and drought, which can significantly
@@
 The output contains the original data, index values (\texttt{.index}),
-parameter used (\texttt{.scale}, \texttt{.method}, and \texttt{.dist}),
+parameters used (\texttt{.scale}, \texttt{.method}, and \texttt{.dist}),
 and all the intermediate variables (\texttt{pet}, \texttt{.agg}, and
@@
 distribution has less influence on the index, with general extreme value
-distribution tending to produce more extreme outcomes than log logistic
+distribution tending to produce more extreme outcomes than log-logistic
 distribution for the extreme events (index \textgreater{} 2 or
@@
 Precipitation Index (SPI-12) in Queensland, Australia during two major
-flood and drought events: 2010/11 and 2019/20. The map shows a continous
-wet period during the 2010/11 flood period and a mitigated drought
-situation, after its worst in 2019 December and 2020 Janurary, likely
-due to the increased rainfall in February from the meteorological
+flood and drought events: 2010/11 and 2019/20. The map shows a
+continuous wet period during the 2010/11 flood period and a mitigated
+drought situation, after its worst in 2019 December and 2020 January,
+likely due to the increased rainfall in February from the meteorological
 record.}
@@
 represents the class ``extreme drought'' by the SPEI. A larger time
-scale gives a smoother index series, while also takes longer to recover
+scale gives a smoother index series, while also taking longer to recover
 from an extreme situation as seen in the 2019/20 drought period. The
-SPEI values from two distribution fits mostly agree, while GEV can
-results in more extreme values, i.e.~in 1998 and 2020.}
+SPEI values from the two distribution fit mostly agree, while GEV can
+result in more extreme values, i.e.~in 1998 and 2020.}
 
@@
 variables into four dimensions through linear combinations. These
-dimensions are then combined though another linear combination to form
+dimensions are then combined through another linear combination to form
 the index. Figure~\ref{fig-pp-gggi} illustrates this pipeline
@@
 variable weights (\texttt{V-weight}) and dimension weights
-(\texttt{D-weight}) used in the two dimension reduction. In the table,
-the variable weights are computed as the inverse of the standard
+(\texttt{D-weight}) used in the two dimension reduction steps. In the
+table, the variable weights are computed as the inverse of the standard
 deviation of each variable, scaled to sum to 1. These weights ensure
 that a one percentage point change in the standard deviation of each
-variable to contribute equally to the index. Dimension weights are equal
+variable contributes equally to the index. Dimension weights are equal
 across the four dimensions and the last column, \texttt{weight},
 multiples the variable and dimension weights to produce a single set of
-weight.
+weights.
 
@@
 Table~\ref{tbl-gggi-weights} as \texttt{gggi\_weights}. The index can be
-reproduced with the pacakge as:
+reproduced with the package as:
 
@@
 
-After initialising the \texttt{gggi} object and attaching the
-\texttt{gggi\_weights} as metadata, a single linear combination step
+After initialzing the \texttt{gggi} object and attaching the
+\texttt{gggi\_weights} as meta-data, a single linear combination step
 within the dimension reduction module is applied to the 14 variables
@@
 multivariate information with a weight vector. Projecting data from
-higher to lower dimension unavoidably leads to information loss and the
+higher to lower dimensions unavoidably leads to information loss and the
 weights used require careful examination to understand their effects on
@@
 making slight adjustments to the weight vector, we can observe how index
-values and country rankings change. For illustration, we select a set
+values and country rankings change. For illustration, we select a set of
 countries including:
@@
 This process creates an animation showing the movement of index values
-in response to changing weights. This visualisation technique, which
+in response to changing weights. This visualization technique, which
 presents a sequence of data projections, is referred to as a ``tour''.
@@
 In GGGI, the index value has a direct interpretation as the percentage
-of gender gap that has been closed. When countries exhibit closer index
-values or consistent, it can be directly interpreted as a reflection of
-a country's progress toward gender parity. Ideally, an index should be
-robust against minor changes in its construction components. This
-example provides researchers with means to observe changes in the index
-resulting from variations in the weights used in linear combination.
-This approach can be applied broadly to different sets of weights of
-interest, extending beyond the change in a single dimension illustrated
-by the example.
+of the gender gap that has been closed. When countries exhibit closer or
+consistent decreasing index values, it can be directly interpreted as a
+reflection of a country's progress toward gender parity. Ideally, an
+index should be robust against minor changes in its construction
+components. This example provides researchers with means to observe
+changes in the index resulting from variations in the weights used in
+linear combinations. This approach can be applied broadly to different
+sets of weights of interest, extending beyond the change in a single
+dimension illustrated by the example.
 

### 2023-08-17 tidyindex.qmd
add conclusion

@@
 
-The paper presents a data pipeline with nine modules for constructing and analysing indexes.
-The pipeline increases transparency in the practice for index analysts to experiment with different index design and parameter choices to better design and apply their indexes.
-The significance of this work is its ability to provide a universal framework for index construction, which can be applied across different domains.
-
-Examples have been given in the drought indexes and human development index to demonstrate computing of indexes with different parameters combinations and how alternative index design can provide insights to understand distinctive country characteristics that could sometimes be overlooked.
-The accompanied package, tidyindex, is not meant to provide comprehensive implementation for all indexes across all domains.
-Instead, it demonstrates implementing individual pipeline steps that are versatile to multiple indexes and composing new indexes from existing steps.
-Domain experts are welcomed to adopt the pipeline approach to develop specialised packages for specific-domains indexes.
-
-Future work: - integrate more complex dimension reduction methods to calculate weights - strengthen the spatial processing module
+The paper introduces a data pipeline comprising nine modules designed for the construction and analysis of indexes within the tidy framework.
+The pipeline offers a modular workflow to allow compute index with different parameterizations, to test minor changes to the original index definition, and to quantify uncertainties.
+The framework proposed in the paper is universal to index across diverse domains.
+Examples are provided, including the drought indexes (SPI and SPEI) and Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions and to illustrate how slight adjustment of linear combination weights impact the index.
+The accompanied package, tidyindex, is not intended to offer an exhaustive implementation for all indexes across every domains.
+Instead, it provides a realization of the pipeline framework proposed in the paper.
+When adopting the pipeline approach to construct indexes, analysts may consider developing software that can be readily deployed in the cloud for production purposes.
 

### 2023-08-17 tidyindex.tex
add conclusion

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, interior hidden, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-The paper presents a data pipeline with nine modules for constructing
-and analysing indexes. The pipeline increases transparency in the
-practice for index analysts to experiment with different index design
-and parameter choices to better design and apply their indexes. The
-significance of this work is its ability to provide a universal
-framework for index construction, which can be applied across different
-domains.
-
-Examples have been given in the drought indexes and human development
-index to demonstrate computing of indexes with different parameters
-combinations and how alternative index design can provide insights to
-understand distinctive country characteristics that could sometimes be
-overlooked. The accompanied package, tidyindex, is not meant to provide
-comprehensive implementation for all indexes across all domains.
-Instead, it demonstrates implementing individual pipeline steps that are
-versatile to multiple indexes and composing new indexes from existing
-steps. Domain experts are welcomed to adopt the pipeline approach to
-develop specialised packages for specific-domains indexes.
-
-Future work: - integrate more complex dimension reduction methods to
-calculate weights - strengthen the spatial processing module
+The paper introduces a data pipeline comprising nine modules designed
+for the construction and analysis of indexes within the tidy framework.
+The pipeline offers a modular workflow to allow compute index with
+different parameterizations, to test minor changes to the original index
+definition, and to quantify uncertainties. The framework proposed in the
+paper is universal to index across diverse domains. Examples are
+provided, including the drought indexes (SPI and SPEI) and Global Gender
+Gap Index (GGGI), to demonstrate the index calculation with different
+time scales and distributions and to illustrate how slight adjustment of
+linear combination weights impact the index. The accompanied package,
+tidyindex, is not intended to offer an exhaustive implementation for all
+indexes across every domains. Instead, it provides a realization of the
+pipeline framework proposed in the paper. When adopting the pipeline
+approach to construct indexes, analysts may consider developing software
+that can be readily deployed in the cloud for production purposes.
 

### 2023-08-17 tidyindex.qmd
add acknowledgement

@@
 <!-- **Constructing indexes would also benefit from pipeline and embracing the aforementioned design philosophy.** -->
-
-<!-- For constructing indexes, many indexes have provided available publicly scripts for computation, i.e.
-Human Development Index (HDI) [@hdi].
-Some have made the first level of generalisation from scripts to functions and creates R packages specialised for calculating particular indexes, i.e.
-the `SPEI` package [@SPEI] for calculating drought indexes SPI and SPEI.
--->
 
@@
 
-<!-- When economy and education are given a greater weight, the variation in index values and ranking highlights countries that performs relatively better (moving right) or worse (moving left) in these two single dimensions.
-For example, the index moves to the left for Ethiopia and Senegal when economy is given a higher weight.
-This reveals the economy as a weakness for these two countries compared to their similarly-ranked peers.
--->
-
-<!-- These can be helpful in understanding the sensitivity of the index value and country ranking to each variable and identifying the strengths and weaknesses of each country relative to others.  -->
-
 # Conclusion
@@
 
+# Acknowledgement
+
+This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship.
+The article is created using Quarto in R.
+The source code for reproducing this paper can be found at: https://github.com/huizezhang-sherry/paper-tidyindex.
+
 # Reference

### 2023-08-17 tidyindex.tex
add acknowledgement

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
+\hypertarget{acknowledgement}{%
+\section{Acknowledgement}\label{acknowledgement}}
+
+This work is funded by a Commonwealth Scientific and Industrial Research
+Organisation (CSIRO) Data61 Scholarship. The article is created using
+Quarto in R. The source code for reproducing this paper can be found at:
+https://github.com/huizezhang-sherry/paper-tidyindex.
+
 \hypertarget{reference}{%

### 2023-08-18 tidyindex.qmd
a new section on background to index usage and development + some changes on notation

@@
   - software design
+  - uncertainty
+  - decision-making
 bibliography: bibliography.bib  
@@
 Such decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes and calculating index uncertainty.
-In social indexes, for instance, the OECD handbook [@oecd_handbook_2008] has provided a set of steps and recommendations for constructing composite socio-economic indexes.
-However, there is still a need to extend these guidelines and methods to accommodate the inclusion of more methodological complex steps required for indexes in general.
 
@@
 -->
+# Background to index usage and development
+
+<!-- * point to major review papers describing index dev  -->
+<!-- * point to major uses of indexes  -->
+
+Indexes serve as critical tools to guide understanding and decisions across natural and social sciences.
+In environmental science, climate indexes are produces by major monitoring centres, such as the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early monitoring on natural disaster.
+In economics, indexes provide insight into market trends through combining a collection of representative items.
+In social sciences, indexes are useful to measure social concepts such as human development, gender equity, and university quality, which leads to country ranking and enables cross-country comparison.
+
+While the concept of combining variables into indexes finds widespread application across various domains, indexes tend to develop separately within each field.
+The OECD handbook [@oecd_handbook_2008] provides a framework of steps and recommendations.
+for composing socio-economic composite indexes.
+The drought index handbook [@svoboda2016handbook], instead, provides a practical guide outlining the advantage and disadvantage of various drought indexes, as recommended by World Meteorology Organization.
+Review papers in the literature summarise different methodologies for constructing indexes from different variables and statistical methods, for example, @zargar2011review, @hao_drought_2015 @alahacoon_comprehensive_2022 for drought indexes.
+In specific domains, attention are given to analyze the index construction process, including the implications of the choices made during the index construction [@jones_vulnerability_2007] and uncertainty or sensitivity analysis [@uncertainty; @tate_social_2012].
+
+In software development, many tools encapsulate all the construction steps into a single function to allow computing indexes with provided data.
+This includes the `SPEI` package [@spei] for two drought indexes, the `gpindex` [@gpindex] package for price indexes, and the `fundiversity` package [@fundiversity] for functional diversity indexes in ecology.
+The package `COINr` [@COINr] deviates from this one-size-fit-all approach by following the guidelines proposed in the OECD handbook to construct, analyze, and visualize composite indexes.
 
@@
 
-| No.
-| Module                  | Input                                                                   | Operation                           | Output                                                                        |
-|--|---------------|----------|---------------|--------------------------|
-| 1   | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
-| 2   | Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
-| 3   | Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
-| 4   | Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
-| 5   | Dimension reduction     | $\mathbf{x}(s;t) \quad \mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q$      |
-| 6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $p_j(s;t) \quad p(.) \in [0, 1]$                   |
-| 7   | Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
-| 8   | Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
-| 9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                      |
+| Section | Module                  | Input                                                                   | Operation                           | Output                                                                        |
+|-------|---------------|---------|---------------|--------------------------|
+| 3.1   | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
+| 3.2   | Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
+| 3.3   | Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
+| 3.4   | Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
+| 3.5   | Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q$      |
+| 3.6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $p_j(s;t) \quad p(.) \in [0, 1]$                   |
+| 3.7   | Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
+| 3.8   | Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
+| 3.9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                      |
 
@@
 In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation.
-An example of temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$.
+An example of temporal processing done in the computation of the Standardized Precipitation Index (SPI) [@mckee1993relationship], consists of summing the monthly precipitation series over a rolling time window of size $k$.
 That is also known as the time scale.
@@
 The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods.
-For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
+For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) [@SPEI] calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
 This is the only step that differs from the Standardized Precipitation Index (SPI).
@@
 
-The R package `tidyindex` implements the index pipeline modules described in @sec-pipeline. These modules compute an index in a sequential manner, as shown below:
+The R package `tidyindex` implements a proof-of-concept of the index pipeline modules described in @sec-pipeline. These modules compute an index in a sequential manner, as shown below:
 
@@
 ```
+
+(more changes) The accompanied package, tidyindex, is not intended to offer an exhaustive implementation for all indexes across every domains.
+Instead, it provides a realization of the pipeline framework proposed in the paper.
+When adopting the pipeline approach to construct indexes, analysts may consider developing software that can be readily deployed in the cloud for production purposes.
 
@@
 
+TODO: increase font for variable names
+
+## Confidence interval
+
+also the ABS one - could display some uncertainties
+
 # Conclusion
@@
 Examples are provided, including the drought indexes (SPI and SPEI) and Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions and to illustrate how slight adjustment of linear combination weights impact the index.
-The accompanied package, tidyindex, is not intended to offer an exhaustive implementation for all indexes across every domains.
-Instead, it provides a realization of the pipeline framework proposed in the paper.
-When adopting the pipeline approach to construct indexes, analysts may consider developing software that can be readily deployed in the cloud for production purposes.
 
@@
 This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship.
-The article is created using Quarto in R.
+The article is created using Quarto [@] in R [@].
 The source code for reproducing this paper can be found at: https://github.com/huizezhang-sherry/paper-tidyindex.

### 2023-08-18 tidyindex.tex
a new section on background to index usage and development + some changes on notation

@@
   pdfauthor={H. Sherry Zhang; Dianne Cook; Ursula Laa; Nicolas Langrené; Patricia Menéndez},
-  pdfkeywords={indexes, data pipeline, software design},
+  pdfkeywords={indexes, data pipeline, software
+design, uncertainty, decision-making},
   colorlinks=true,
@@
 \def\sep{;\ }
-indexes\sep data pipeline\sep 
-software design
+indexes\sep data pipeline\sep software design\sep uncertainty\sep 
+decision-making
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 benefits for comparing among indexes and calculating index uncertainty.
-In social indexes, for instance, the OECD handbook (OECD, European
-Union, and Joint Research Centre - European Commission 2008) has
-provided a set of steps and recommendations for constructing composite
-socio-economic indexes. However, there is still a need to extend these
-guidelines and methods to accommodate the inclusion of more
-methodological complex steps required for indexes in general.
 
@@
 cases of the pipeline.
+
+\hypertarget{background-to-index-usage-and-development}{%
+\section{Background to index usage and
+development}\label{background-to-index-usage-and-development}}
+
+Indexes serve as critical tools to guide understanding and decisions
+across natural and social sciences. In environmental science, climate
+indexes are produces by major monitoring centres, such as the United
+States Drought Monitor and National Oceanic and Atmospheric
+Administration, to facilitate agricultural planning and early monitoring
+on natural disaster. In economics, indexes provide insight into market
+trends through combining a collection of representative items. In social
+sciences, indexes are useful to measure social concepts such as human
+development, gender equity, and university quality, which leads to
+country ranking and enables cross-country comparison.
+
+While the concept of combining variables into indexes finds widespread
+application across various domains, indexes tend to develop separately
+within each field. The OECD handbook (OECD, European Union, and Joint
+Research Centre - European Commission 2008) provides a framework of
+steps and recommendations. for composing socio-economic composite
+indexes. The drought index handbook (Svoboda, Fuchs, et al. 2016),
+instead, provides a practical guide outlining the advantage and
+disadvantage of various drought indexes, as recommended by World
+Meteorology Organization. Review papers in the literature summarise
+different methodologies for constructing indexes from different
+variables and statistical methods, for example, Zargar et al. (2011),
+Hao and Singh (2015) Alahacoon and Edirisinghe (2022) for drought
+indexes. In specific domains, attention are given to analyze the index
+construction process, including the implications of the choices made
+during the index construction (Jones and Andrey 2007) and uncertainty or
+sensitivity analysis (Saisana, Saltelli, and Tarantola 2005; Tate 2012).
+
+In software development, many tools encapsulate all the construction
+steps into a single function to allow computing indexes with provided
+data. This includes the \texttt{SPEI} package (Vicente-Serrano,
+Beguería, and López-Moreno 2010) for two drought indexes, the
+\texttt{gpindex} (Martin 2023) package for price indexes, and the
+\texttt{fundiversity} package (Grenié and Gruson 2023) for functional
+diversity indexes in ecology. The package \texttt{COINr} (Becker et al.
+2022) deviates from this one-size-fit-all approach by following the
+guidelines proposed in the OECD handbook to construct, analyze, and
+visualize composite indexes.
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0294}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2206}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1471}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2206}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3824}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
 \caption{\label{tbl-notation}An notation overview of the input,
@@
 \begin{minipage}[b]{\linewidth}\raggedright
-No.
+Section
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
@@
 \endhead
-1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
+3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
 \(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
-2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
+3.2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
 \(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
-3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
+3.3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
 \(x^{\text{Trans}}_j(s;t)\) \\
-4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
+3.4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
 \(x^{\text{Scale}}_j(s;t)\) \\
-5 & Dimension reduction &
-\(\mathbf{x}(s;t) \quad \mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
-\(h[\mathbf{x}(s;t)]\) &
-\(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q\) \\
-6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
+3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
+& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q\) \\
+3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
 \(p_j(s;t) \quad p(.) \in [0, 1]\) \\
-7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
+3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
 \(z_j(s; t)\) \\
-8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
-9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
+3.8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
+3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
 \(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
@@
 example of temporal processing done in the computation of the
-Standardized Precipitation Index (SPI), consists of summing the monthly
-precipitation series over a rolling time window of size \(k\). That is
-also known as the time scale. For SPI, the choice of the time scale
-\(k\) is used to control the accumulation period for the water deficit,
-enabling the assessment of drought severity across various types
-(meteorological, agricultural, and hydrological).
+Standardized Precipitation Index (SPI) (McKee et al. 1993), consists of
+summing the monthly precipitation series over a rolling time window of
+size \(k\). That is also known as the time scale. For SPI, the choice of
+the time scale \(k\) is used to control the accumulation period for the
+water deficit, enabling the assessment of drought severity across
+various types (meteorological, agricultural, and hydrological).
 
@@
 example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
-calculates the difference \(D\) between precipitation (\(P\)) and
-potential evapotranspiration (\(\text{PET}\)), using a water balance
-model (\(D = P - \text{PET}\)). This is the only step that differs from
-the Standardized Precipitation Index (SPI).
+(Beguería and Vicente-Serrano 2017) calculates the difference \(D\)
+between precipitation (\(P\)) and potential evapotranspiration
+(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
+This is the only step that differs from the Standardized Precipitation
+Index (SPI).
 
@@
 
-The R package \texttt{tidyindex} implements the index pipeline modules
-described in Section~\ref{sec-pipeline}. These modules compute an index
-in a sequential manner, as shown below:
+The R package \texttt{tidyindex} implements a proof-of-concept of the
+index pipeline modules described in Section~\ref{sec-pipeline}. These
+modules compute an index in a sequential manner, as shown below:
 
@@
 \end{Shaded}
+
+(more changes) The accompanied package, tidyindex, is not intended to
+offer an exhaustive implementation for all indexes across every domains.
+Instead, it provides a realization of the pipeline framework proposed in
+the paper. When adopting the pipeline approach to construct indexes,
+analysts may consider developing software that can be readily deployed
+in the cloud for production purposes.
 
@@
 
+TODO: increase font for variable names
+
+\hypertarget{confidence-interval}{%
+\subsection{Confidence interval}\label{confidence-interval}}
+
+also the ABS one - could display some uncertainties
+
 \hypertarget{conclusion}{%
@@
 time scales and distributions and to illustrate how slight adjustment of
-linear combination weights impact the index. The accompanied package,
-tidyindex, is not intended to offer an exhaustive implementation for all
-indexes across every domains. Instead, it provides a realization of the
-pipeline framework proposed in the paper. When adopting the pipeline
-approach to construct indexes, analysts may consider developing software
-that can be readily deployed in the cloud for production purposes.
+linear combination weights impact the index.
 
@@
 Organisation (CSIRO) Data61 Scholarship. The article is created using
-Quarto in R. The source code for reproducing this paper can be found at:
-https://github.com/huizezhang-sherry/paper-tidyindex.
+Quarto {[}@{]} in R {[}@{]}. The source code for reproducing this paper
+can be found at: https://github.com/huizezhang-sherry/paper-tidyindex.
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-alahacoon_comprehensive_2022}{}}%
+Alahacoon, Niranga, and Mahesh Edirisinghe. 2022. {``A Comprehensive
+Assessment of Remote Sensing and Traditional Based Drought Monitoring
+Indices at Global and Regional Scale.''} \emph{Geomatics, Natural
+Hazards and Risk} 13 (December): 762--99.
+\url{https://doi.org/10.1080/19475705.2022.2044394}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-COINr}{}}%
+Becker, William, Giulio Caperna, Maria Del Sorbo, Hedvig Norlen, Eleni
+Papadimitriou, and Michaela Saisana. 2022. {``COINr: An r Package for
+Developing Composite Indicators.''} \emph{Journal of Open Source
+Software} 7 (78): 4567. \url{https://doi.org/10.21105/joss.04567}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-SPEI}{}}%
+Beguería, Santiago, and Sergio M. Vicente-Serrano. 2017. \emph{SPEI:
+Calculation of the Standardised Precipitation-Evapotranspiration Index}.
+\url{https://CRAN.R-project.org/package=SPEI}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-buja_elements_1988}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-fundiversity}{}}%
+Grenié, Matthias, and Hugo Gruson. 2023. \emph{{fundiversity}: Easy
+Computation of Functional Diversity Indices}.
+\url{https://doi.org/10.5281/zenodo.4761754}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hao_drought_2015}{}}%
+Hao, Zengchao, and Vijay P. Singh. 2015. {``Drought Characterization
+from a Multivariate Perspective: {A} Review.''} \emph{Journal of
+Hydrology} 527 (August): 668--78.
+\url{https://doi.org/10.1016/j.jhydrol.2015.05.031}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jones_vulnerability_2007}{}}%
+Jones, Brenda, and Jean Andrey. 2007. {``Vulnerability Index
+Construction: Methodological Choices and Their Influence on Identifying
+Vulnerable Neighbourhoods.''} \emph{International Journal of Emergency
+Management} 4 (2): 269--95.
+\url{https://doi.org/10.1504/IJEM.2007.013994}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-tidymodels}{}}%
@@
 \url{https://www.tidymodels.org}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gpindex}{}}%
+Martin, Steve. 2023. \emph{Gpindex: Generalized Price and Quantity
+Indexes}. \url{https://CRAN.R-project.org/package=gpindex}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mckee1993relationship}{}}%
+McKee, Thomas B, Nolan J Doesken, John Kleist, et al. 1993. {``The
+Relationship of Drought Frequency and Duration to Time Scales.''} In
+\emph{Proceedings of the 8th Conference on Applied Climatology},
+17:179--83. 22. Boston, MA, USA.
 
@@
 \url{https://doi.org/10.1787/9789264043466-en}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-uncertainty}{}}%
+Saisana, M., A. Saltelli, and S. Tarantola. 2005. {``{Uncertainty and
+Sensitivity Analysis Techniques as Tools for the Quality Assessment of
+Composite Indicators}.''} \emph{Journal of the Royal Statistical Society
+Series A: Statistics in Society} 168 (2): 307--23.
+\url{https://doi.org/10.1111/j.1467-985X.2005.00350.x}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-svoboda2016handbook}{}}%
+Svoboda, Mark, Brian Fuchs, et al. 2016. {``Handbook of Drought
+Indicators and Indices.''} \emph{Drought and Water Crises: Integrating
+Science, Management, and Policy}, 155--208.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tate_social_2012}{}}%
+Tate, Eric. 2012. {``Social Vulnerability Indices: A Comparative
+Assessment Using Uncertainty and Sensitivity Analysis.''} \emph{Natural
+Hazards} 63 (2): 325--47.
+\url{https://doi.org/10.1007/s11069-012-0152-2}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%
+Vicente-Serrano, Sergio M., Santiago Beguería, and Juan I. López-Moreno.
+2010. {``A {Multiscalar} {Drought} {Index} {Sensitive} to {Global}
+{Warming}: {The} {Standardized} {Precipitation} {Evapotranspiration}
+{Index}.''} \emph{Journal of Climate} 23 (7): 1696--1718.
+\url{https://journals.ametsoc.org/view/journals/clim/23/7/2009jcli2909.1.xml}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-zargar2011review}{}}%
+Zargar, Amin, Rehan Sadiq, Bahman Naser, and Faisal I Khan. 2011. {``A
+Review of Drought Indices.''} \emph{Environmental Reviews} 19 (NA):
+333--49.
+
 \end{CSLReferences}

### 2023-08-18 tidyindex.tex
citation on R and Quarto

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 Organisation (CSIRO) Data61 Scholarship. The article is created using
-Quarto {[}@{]} in R {[}@{]}. The source code for reproducing this paper
-can be found at: https://github.com/huizezhang-sherry/paper-tidyindex.
+Quarto (Allaire et al. 2022) in R (R Core Team 2021). The source code
+for reproducing this paper can be found at:
+https://github.com/huizezhang-sherry/paper-tidyindex.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Allaire_Quarto_2022}{}}%
+Allaire, J. J., Charles Teague, Carlos Scheidegger, Yihui Xie, and
+Christophe Dervieux. 2022. \emph{{Quarto}} (version 1.2).
+\url{https://doi.org/10.5281/zenodo.5960048}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-COINr}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
+R Core Team. 2021. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.R-project.org/}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-uncertainty}{}}%
@@
 Review of Drought Indices.''} \emph{Environmental Reviews} 19 (NA):
-333--49.
+333--49. \url{https://www.jstor.org/stable/envirevi.19.333}.
 

### 2023-08-19 tidyindex.qmd
citation on R and Quarto
revised the abstract

@@
   Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making.
-While many work has focused on defining new indexes for the application of interest, less attention has been directed towards understanding how the a given index behave under different scenarios.
-This research introduces a data pipeline comparising 9 modules to assemble indexes from a series of modular building blocks.
-This modular data pipeline is universally applicable to index computation and allows analysts to incorporate the analysis on index into the construction pipeline.
-Analysts can compute with different index parameters, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess index robustness .
-The paper presents two examples to illustrate the usage of the pipeline.
-The first example claculates two drought indexes with different parameters to investigate the spatio-temporal distribution of drought in Queensland, Australia.
-The second example on the Global Gender Gap Index (GGGI) explores the effects of adjusting linear combination weights on a country's index values and rankings.
+While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their value and variation in values.
+Here we discuss a modular data pipeline recommendation to assemble indexes.
+It is universally applicable to index computation and allows investigation of index behavior as part of the development procedure.
+One can compute with different the index with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess an index's robustness.
+The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on a country's ranking; and how to calculate bootstrap confidence intervals for XXX index.
+The methods are supported by a new R package, called `tidyindex`.
 keywords:
@@
 This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship.
-The article is created using Quarto [@] in R [@].
+The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
 The source code for reproducing this paper can be found at: https://github.com/huizezhang-sherry/paper-tidyindex.

### 2023-08-19 tidyindex.tex
revised the abstract

@@
 Indexes are useful for summarizing multivariate information into single
-metrics for monitoring, communicating, and decision-making. While many
-work has focused on defining new indexes for the application of
-interest, less attention has been directed towards understanding how the
-a given index behave under different scenarios. This research introduces
-a data pipeline comparising 9 modules to assemble indexes from a series
-of modular building blocks. This modular data pipeline is universally
-applicable to index computation and allows analysts to incorporate the
-analysis on index into the construction pipeline. Analysts can compute
-with different index parameters, adjust steps in the index definition by
-adding, removing, and swapping them to experiment with various index
-designs, calculate uncertainty measures, and assess index robustness .
-The paper presents two examples to illustrate the usage of the pipeline.
-The first example claculates two drought indexes with different
-parameters to investigate the spatio-temporal distribution of drought in
-Queensland, Australia. The second example on the Global Gender Gap Index
-(GGGI) explores the effects of adjusting linear combination weights on a
-country's index values and rankings.
+metrics for monitoring, communicating, and decision-making. While most
+work has focused on defining new indexes for specific purposes, more
+attention needs to be directed towards making it possible to understand
+index behavior in different data conditions, and to determine how their
+structure affects their value and variation in values. Here we discuss a
+modular data pipeline recommendation to assemble indexes. It is
+universally applicable to index computation and allows investigation of
+index behavior as part of the development procedure. One can compute
+with different the index with different parameter choices, adjust steps
+in the index definition by adding, removing, and swapping them to
+experiment with various index designs, calculate uncertainty measures,
+and assess an index's robustness. The paper presents three examples to
+illustrate the pipeline framework usage: comparison of two different
+indexes designed to monitor the spatio-temporal distribution of drought
+in Queensland, Australia; the effect of dimension reduction choices on
+the Global Gender Gap Index (GGGI) on a country's ranking; and how to
+calculate bootstrap confidence intervals for XXX index. The methods are
+supported by a new R package, called \texttt{tidyindex}.
 \end{abstract}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 operation, and output of each pipeline module.}\tabularnewline
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
@@
 \(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
-\bottomrule()
 \end{longtable}
@@
 Gender Gap Index}\tabularnewline
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright

### 2023-08-19 tidyindex.qmd
first paragraph of section 2 parts moved into first paragraph of the intro

@@
 Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making.
-Examples of those include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, or Human Development Index among many others.
+They serve as critical tools across the natural and social sciences.
+Examples include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, and the Human Development Index.
+In environmental science climate indexes are produced by major monitoring centers, like the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early detection of natural disasters.
+In economics, indexes provide insight into market trends through combining prices of a basket of goods and services.
+In social sciences, indexes are used to monitor human development, gender equity, or university quality.
+The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared.
 
@@
 
-Indexes serve as critical tools to guide understanding and decisions across natural and social sciences.
-In environmental science, climate indexes are produces by major monitoring centres, such as the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early monitoring on natural disaster.
-In economics, indexes provide insight into market trends through combining a collection of representative items.
-In social sciences, indexes are useful to measure social concepts such as human development, gender equity, and university quality, which leads to country ranking and enables cross-country comparison.
-
-While the concept of combining variables into indexes finds widespread application across various domains, indexes tend to develop separately within each field.
 The OECD handbook [@oecd_handbook_2008] provides a framework of steps and recommendations.

### 2023-08-19 tidyindex.tex
first paragraph of section 2 parts moved into first paragraph of the intro

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 information into a single number for monitoring, communicating, and
-decision-making. Examples of those include the Air Quality Index, El
+decision-making. They serve as critical tools across the natural and
+social sciences. Examples include the Air Quality Index, El
 Niño-Southern Oscillation Index, Consumer Price Index, QS University
-Rankings, or Human Development Index among many others.
+Rankings, and the Human Development Index. In environmental science
+climate indexes are produced by major monitoring centers, like the
+United States Drought Monitor and National Oceanic and Atmospheric
+Administration, to facilitate agricultural planning and early detection
+of natural disasters. In economics, indexes provide insight into market
+trends through combining prices of a basket of goods and services. In
+social sciences, indexes are used to monitor human development, gender
+equity, or university quality. The problem is that every index is
+developed in its own unique way, by different researchers or
+organizations, and often indexes designed for the same purpose cannot
+easily be compared.
 
@@
 
-Indexes serve as critical tools to guide understanding and decisions
-across natural and social sciences. In environmental science, climate
-indexes are produces by major monitoring centres, such as the United
-States Drought Monitor and National Oceanic and Atmospheric
-Administration, to facilitate agricultural planning and early monitoring
-on natural disaster. In economics, indexes provide insight into market
-trends through combining a collection of representative items. In social
-sciences, indexes are useful to measure social concepts such as human
-development, gender equity, and university quality, which leads to
-country ranking and enables cross-country comparison.
-
-While the concept of combining variables into indexes finds widespread
-application across various domains, indexes tend to develop separately
-within each field. The OECD handbook (OECD, European Union, and Joint
-Research Centre - European Commission 2008) provides a framework of
-steps and recommendations. for composing socio-economic composite
-indexes. The drought index handbook (Svoboda, Fuchs, et al. 2016),
-instead, provides a practical guide outlining the advantage and
-disadvantage of various drought indexes, as recommended by World
-Meteorology Organization. Review papers in the literature summarise
-different methodologies for constructing indexes from different
-variables and statistical methods, for example, Zargar et al. (2011),
-Hao and Singh (2015) Alahacoon and Edirisinghe (2022) for drought
-indexes. In specific domains, attention are given to analyze the index
-construction process, including the implications of the choices made
-during the index construction (Jones and Andrey 2007) and uncertainty or
-sensitivity analysis (Saisana, Saltelli, and Tarantola 2005; Tate 2012).
+The OECD handbook (OECD, European Union, and Joint Research Centre -
+European Commission 2008) provides a framework of steps and
+recommendations. for composing socio-economic composite indexes. The
+drought index handbook (Svoboda, Fuchs, et al. 2016), instead, provides
+a practical guide outlining the advantage and disadvantage of various
+drought indexes, as recommended by World Meteorology Organization.
+Review papers in the literature summarise different methodologies for
+constructing indexes from different variables and statistical methods,
+for example, Zargar et al. (2011), Hao and Singh (2015) Alahacoon and
+Edirisinghe (2022) for drought indexes. In specific domains, attention
+are given to analyze the index construction process, including the
+implications of the choices made during the index construction (Jones
+and Andrey 2007) and uncertainty or sensitivity analysis (Saisana,
+Saltelli, and Tarantola 2005; Tate 2012).
 

### 2023-08-19 tidyindex.qmd
tidied up section 2

@@
 -->
-# Background to index usage and development
+
+# Background to index development
 
@@
 
-The OECD handbook [@oecd_handbook_2008] provides a framework of steps and recommendations.
-for composing socio-economic composite indexes.
-The drought index handbook [@svoboda2016handbook], instead, provides a practical guide outlining the advantage and disadvantage of various drought indexes, as recommended by World Meteorology Organization.
-Review papers in the literature summarise different methodologies for constructing indexes from different variables and statistical methods, for example, @zargar2011review, @hao_drought_2015 @alahacoon_comprehensive_2022 for drought indexes.
-In specific domains, attention are given to analyze the index construction process, including the implications of the choices made during the index construction [@jones_vulnerability_2007] and uncertainty or sensitivity analysis [@uncertainty; @tate_social_2012].
+There are many documents providing advice on how to construct indexes for different fields, and review articles describing the range of available indexes for specific purposes.
+The OECD handbook [@oecd_handbook_2008] provides a comprehensive guide for computing socio-economic composite indexes, with detailed steps and recommendations.
+The drought index handbook [@svoboda2016handbook] provides details of various drought indexes and recommendations from the World Meteorology Organization.
+@zargar2011review, @hao_drought_2015 and @alahacoon_comprehensive_2022 are review papers describing the range of possible drought indexes.
 
-In software development, many tools encapsulate all the construction steps into a single function to allow computing indexes with provided data.
-This includes the `SPEI` package [@spei] for two drought indexes, the `gpindex` [@gpindex] package for price indexes, and the `fundiversity` package [@fundiversity] for functional diversity indexes in ecology.
-The package `COINr` [@COINr] deviates from this one-size-fit-all approach by following the guidelines proposed in the OECD handbook to construct, analyze, and visualize composite indexes.
+There is also some attention being given to the diagnosis of indexes, and incorporation of uncertainty.
+@jones_vulnerability_2007 investigates the methodological choices made in the development of indexes for assessing vulnerable neighborhoods.
+@uncertainty describes incorporating uncertainty estimates and conducting sensitivity analysis on composite indexes.
+@tate_social_2012, similarly, makes a comparative assessment of social vulnerability indexes based on uncertainty estimation and sensitivity analysis.
+(XXX Something about Ursula's colleagues paper here too.)
+
+There are also R packages supporting index calculation.
+The  `SPEI` package  [@spei] computes two drought indexes.
+The `gpindex` package [@gpindex] computes price indexes, and the `fundiversity` package [@fundiversity] computes functional diversity indexes for ecological study.
+The package `COINr` [@COINr] is more ambitious, making a start on following the broader guidelines in the OECD handbook to construct, analyze, and visualize composite indexes.
+
+From reviewing this literature, and in the process of developing methods for making it easier to work with multivariate spatiotemporal data, it seems possible to think about indexes in a more organised, cohesive and standard manner.
+Actually, it seems that the area could benefit from a *tidy* approach.
 

### 2023-08-19 tidyindex.tex
tidied up section 2

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{background-to-index-usage-and-development}{%
-\section{Background to index usage and
-development}\label{background-to-index-usage-and-development}}
-
-The OECD handbook (OECD, European Union, and Joint Research Centre -
-European Commission 2008) provides a framework of steps and
-recommendations. for composing socio-economic composite indexes. The
-drought index handbook (Svoboda, Fuchs, et al. 2016), instead, provides
-a practical guide outlining the advantage and disadvantage of various
-drought indexes, as recommended by World Meteorology Organization.
-Review papers in the literature summarise different methodologies for
-constructing indexes from different variables and statistical methods,
-for example, Zargar et al. (2011), Hao and Singh (2015) Alahacoon and
-Edirisinghe (2022) for drought indexes. In specific domains, attention
-are given to analyze the index construction process, including the
-implications of the choices made during the index construction (Jones
-and Andrey 2007) and uncertainty or sensitivity analysis (Saisana,
-Saltelli, and Tarantola 2005; Tate 2012).
-
-In software development, many tools encapsulate all the construction
-steps into a single function to allow computing indexes with provided
-data. This includes the \texttt{SPEI} package (Vicente-Serrano,
-Beguería, and López-Moreno 2010) for two drought indexes, the
-\texttt{gpindex} (Martin 2023) package for price indexes, and the
-\texttt{fundiversity} package (Grenié and Gruson 2023) for functional
-diversity indexes in ecology. The package \texttt{COINr} (Becker et al.
-2022) deviates from this one-size-fit-all approach by following the
-guidelines proposed in the OECD handbook to construct, analyze, and
-visualize composite indexes.
+\hypertarget{background-to-index-development}{%
+\section{Background to index
+development}\label{background-to-index-development}}
+
+There are many documents providing advice on how to construct indexes
+for different fields, and review articles describing the range of
+available indexes for specific purposes. The OECD handbook (OECD,
+European Union, and Joint Research Centre - European Commission 2008)
+provides a comprehensive guide for computing socio-economic composite
+indexes, with detailed steps and recommendations. The drought index
+handbook (Svoboda, Fuchs, et al. 2016) provides details of various
+drought indexes and recommendations from the World Meteorology
+Organization. Zargar et al. (2011), Hao and Singh (2015) and Alahacoon
+and Edirisinghe (2022) are review papers describing the range of
+possible drought indexes.
+
+There is also some attention being given to the diagnosis of indexes,
+and incorporation of uncertainty. Jones and Andrey (2007) investigates
+the methodological choices made in the development of indexes for
+assessing vulnerable neighborhoods. Saisana, Saltelli, and Tarantola
+(2005) describes incorporating uncertainty estimates and conducting
+sensitivity analysis on composite indexes. Tate (2012), similarly, makes
+a comparative assessment of social vulnerability indexes based on
+uncertainty estimation and sensitivity analysis. (XXX Something about
+Ursula's colleagues paper here too.)
+
+There are also R packages supporting index calculation. The
+\texttt{SPEI} package (Vicente-Serrano, Beguería, and López-Moreno 2010)
+computes two drought indexes. The \texttt{gpindex} package (Martin 2023)
+computes price indexes, and the \texttt{fundiversity} package (Grenié
+and Gruson 2023) computes functional diversity indexes for ecological
+study. The package \texttt{COINr} (Becker et al. 2022) is more
+ambitious, making a start on following the broader guidelines in the
+OECD handbook to construct, analyze, and visualize composite indexes.
+
+From reviewing this literature, and in the process of developing methods
+for making it easier to work with multivariate spatiotemporal data, it
+seems possible to think about indexes in a more organised, cohesive and
+standard manner. Actually, it seems that the area could benefit from a
+\emph{tidy} approach.
 

### 2023-08-20 tidyindex.qmd
add the figure for confidence interval

--- 2023-08-19 tidyindex.qmd
@@
   Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making.
-While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their value and variation in values.
-Here we discuss a modular data pipeline recommendation to assemble indexes.
-It is universally applicable to index computation and allows investigation of index behavior as part of the development procedure.
-One can compute with different the index with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess an index's robustness.
-The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on a country's ranking; and how to calculate bootstrap confidence intervals for XXX index.
-The methods are supported by a new R package, called `tidyindex`.
+While many work has focused on defining new indexes for the application of interest, less attention has been directed towards understanding how the a given index behave under different scenarios.
+This research introduces a data pipeline comparising 9 modules to assemble indexes from a series of modular building blocks.
+This modular data pipeline is universally applicable to index computation and allows analysts to incorporate the analysis on index into the construction pipeline.
+Analysts can compute with different index parameters, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess index robustness .
+The paper presents two examples to illustrate the usage of the pipeline.
+The first example claculates two drought indexes with different parameters to investigate the spatio-temporal distribution of drought in Queensland, Australia.
+The second example on the Global Gender Gap Index (GGGI) explores the effects of adjusting linear combination weights on a country's index values and rankings.
 keywords:
@@
   facet_wrap(vars(scale), labeller = label_both, ncol = 1) +
-  scale_x_yearmonth(breaks = "2 year", date_labels = "%Y") +
+  scale_x_yearmonth(breaks = "4 year", date_labels = "%Y") +
   scale_color_brewer(palette = "Dark2", name = "Distribution") + 
@@
 
+```{r echo = TRUE, eval = FALSE}
+DATA %>% 
+  aggregate(.var = prcp, .scale = 24) %>% 
+  dist_fit(.dist = gamma(), .var = .agg, .n_boot = 100) %>% 
+  augment(.var = .agg)
+
+```
+
+@fig-conf-interval presents the 80% and 95% confidence interval for the Texas post office station, in Queensland, Australia.
+Since the start of 2019, both confidence intervals indicate an extreme drought condition (below -2).
+early warning for the bushfire?
+
 also the ABS one - could display some uncertainties
+
+```{r fig-conf-interval}
+#| fig-width: 8
+#| fig-height: 2
+#| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI) for the Texas post office station, in Queensland, Australia.
+A bootstrap sample of 100 is taken from the aggregated pricipitation series to estimate gamma parameters and calculate the index.
+Since the start of 2019, both confidence intervals indicate an extreme drought condition (below -2)."
+res <- queensland %>% 
+  filter(name == "TEXAS POST OFFICE") %>% 
+  init(id = id, time = ym) %>% 
+  aggregate(.var = prcp, .scale = 24) %>% 
+  dist_fit(.dist = gamma(), .var = .agg, .n_boot = 100) %>% 
+  augment(.var = .agg)
+
+res_band <- res$data %>% 
+  group_by(ym, id, name) %>% 
+  summarise(
+    q025 = quantile(.index, 0.025),
+    q05 = quantile(.index, 0.05),
+    q10 = quantile(.index, 0.1),
+    q50 = quantile(.index, 0.5),
+    q90 = quantile(.index, 0.9),
+    q95 = quantile(.index, 0.95),
+    q975 = quantile(.index, 0.975)
+    ) %>% 
+  ungroup()
+
+all <- idx %>% 
+  filter(name == "TEXAS POST OFFICE", .idx == "spi", .scale == 24) %>% 
+  select(ym, .index) %>% 
+  left_join(res_band, by = "ym")
+
+loc <- make_yearmonth(2022, 12)
+
+# the sampling uncertainty is quite large here
+# and this is mainly due to we only have 30 year worth of data
+# each month fit has only 30 points
+all %>% 
+  ggplot(aes(x = ym)) + 
+  geom_hline(yintercept = -2, linetype = "dashed", color = "grey50") + 
+  #geom_text(aes(x = loc, y = -2.7, label = "SPI = -2"), color = "grey50") + 
+  geom_ribbon(aes(ymin = q025, ymax = q975), fill = "steelblue1", alpha = 0.5) + 
+  geom_ribbon(aes(ymin = q10, ymax = q90), fill = "steelblue4", alpha = 0.5) + 
+  geom_line(aes(y = .index)) + 
+  scale_x_yearmonth(name = "Year", date_break = "2 years", date_label = "%Y") +
+  theme_bw() + 
+  facet_wrap(vars(name), ncol = 1) + 
+  theme(panel.grid = element_blank(), 
+        legend.position = "bottom") + 
+  ylab("SPI")
+
+```
 

### 2023-08-20 tidyindex.tex
add the figure for confidence interval

--- 2023-08-19 tidyindex.tex
@@
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{iftex}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
 \fi
@@
 \makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
 \makeatother
@@
 \@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
 \makeatother
@@
 Indexes are useful for summarizing multivariate information into single
-metrics for monitoring, communicating, and decision-making. While most
-work has focused on defining new indexes for specific purposes, more
-attention needs to be directed towards making it possible to understand
-index behavior in different data conditions, and to determine how their
-structure affects their value and variation in values. Here we discuss a
-modular data pipeline recommendation to assemble indexes. It is
-universally applicable to index computation and allows investigation of
-index behavior as part of the development procedure. One can compute
-with different the index with different parameter choices, adjust steps
-in the index definition by adding, removing, and swapping them to
-experiment with various index designs, calculate uncertainty measures,
-and assess an index's robustness. The paper presents three examples to
-illustrate the pipeline framework usage: comparison of two different
-indexes designed to monitor the spatio-temporal distribution of drought
-in Queensland, Australia; the effect of dimension reduction choices on
-the Global Gender Gap Index (GGGI) on a country's ranking; and how to
-calculate bootstrap confidence intervals for XXX index. The methods are
-supported by a new R package, called \texttt{tidyindex}.
+metrics for monitoring, communicating, and decision-making. While many
+work has focused on defining new indexes for the application of
+interest, less attention has been directed towards understanding how the
+a given index behave under different scenarios. This research introduces
+a data pipeline comparising 9 modules to assemble indexes from a series
+of modular building blocks. This modular data pipeline is universally
+applicable to index computation and allows analysts to incorporate the
+analysis on index into the construction pipeline. Analysts can compute
+with different index parameters, adjust steps in the index definition by
+adding, removing, and swapping them to experiment with various index
+designs, calculate uncertainty measures, and assess index robustness .
+The paper presents two examples to illustrate the usage of the pipeline.
+The first example claculates two drought indexes with different
+parameters to investigate the spatio-temporal distribution of drought in
+Queensland, Australia. The second example on the Global Gender Gap Index
+(GGGI) explores the effects of adjusting linear combination weights on a
+country's index values and rankings.
 \end{abstract}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, interior hidden, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 operation, and output of each pipeline module.}\tabularnewline
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endfirsthead
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
@@
 \(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
+\bottomrule()
 \end{longtable}
@@
 Gender Gap Index}\tabularnewline
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Labour force participation & Economy & 0.199 & 0.25 & 0.050 \\
@@
 Years with female head of state & Politics & 0.443 & 0.25 & 0.111 \\
+\bottomrule()
 \end{longtable}
@@
 
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{.var =}\NormalTok{ prcp, }\AttributeTok{.scale =} \DecValTok{24}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{dist\_fit}\NormalTok{(}\AttributeTok{.dist =} \FunctionTok{gamma}\NormalTok{(), }\AttributeTok{.var =}\NormalTok{ .agg, }\AttributeTok{.n\_boot =} \DecValTok{100}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{augment}\NormalTok{(}\AttributeTok{.var =}\NormalTok{ .agg)}
+\end{Highlighting}
+\end{Shaded}
+
+Figure~\ref{fig-conf-interval} presents the 80\% and 95\% confidence
+interval for the Texas post office station, in Queensland, Australia.
+Since the start of 2019, both confidence intervals indicate an extreme
+drought condition (below -2). early warning for the bushfire?
+
 also the ABS one - could display some uncertainties
+
+\begin{figure}
+
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-conf-interval-1.pdf}
+
+}
+
+\caption{\label{fig-conf-interval}80\% and 95\% confidence interval of
+the Standardized Precipitation Index (SPI) for the Texas post office
+station, in Queensland, Australia. A bootstrap sample of 100 is taken
+from the aggregated pricipitation series to estimate gamma parameters
+and calculate the index. Since the start of 2019, both confidence
+intervals indicate an extreme drought condition (below -2).}
+
+\end{figure}
 

### 2023-08-20 tidyindex.tex

--- 2023-08-19 tidyindex.tex
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
 

### 2023-08-20 tidyindex.qmd
writing up the confidence interval example

@@
 
-## Every distribution, every scale, every index all at once
+## Every distribution, every scale, every index all at once {#sec-example1}
 
@@
 
-## Confidence interval
+## Confidence interval [TO BE CHANGED]
+
+Index uncertainties may arise from various stages of the pipeline, including from different parameterisation choices, as illustrated from @sec-example1, or from the statistical procedures in the pipeline.
+Given the lack of direct instruments for measurement, true values for indexes can never be obtained to compare the index against.
+Nevertheless, an index can be provided with confidence interval to report it accuracy.
+In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate the calculation of confidence interval for the Standardized Precipitation Index (SPI).
+Bootstrapping is used to account for the sampling uncertainty in the distribution fit step of the index pipeline and to assess its impact on the SPI series.
+
+In SPI, the distribution fit step fits the gamma distribution to the aggregated precipitation series separately for each month.
+This results in  31 or 32 points, from 1990 Jan to 2022 Apr, for estimating each set of distribution parameters.
+To account for this sampling uncertainty with these samples, bootstrapping is used to generate replicates of the aggregated series.
+In the `tidyindex` package, this bootstrap sampling is activated when the argument `.n_boot` is set to a value other than the default of 1.
+In the following code, the Standardized Precipitation Index (SPI) is calculated using a time scale of 24.
+The bootstrap procedure samples the aggregated precipitation (`.agg`) for 100 iterations ( `.n_boot = 100`) and then fits the gamma distribution.
+The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `agument()`.
 
@@
 
+The confidence interval can then be calculated using the quantile method from the bootstrap samples.
 @fig-conf-interval presents the 80% and 95% confidence interval for the Texas post office station, in Queensland, Australia.
-Since the start of 2019, both confidence intervals indicate an extreme drought condition (below -2).
-early warning for the bushfire?
-
-also the ABS one - could display some uncertainties
+From the start of 2019 to 2020,  the majority of the confidence intervals lie below the extreme drought line (SPI = -2), suggesting a high level of certainty that the Texas post office is suffering from a drastic drought.
+The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantifying the drought/ flood severity in extreme events.
 
@@
 #| fig-height: 2
-#| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI) for the Texas post office station, in Queensland, Australia.
-A bootstrap sample of 100 is taken from the aggregated pricipitation series to estimate gamma parameters and calculate the index.
-Since the start of 2019, both confidence intervals indicate an extreme drought condition (below -2)."
+#| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI-24) for the Texas post office station, in Queensland, Australia.
+A bootstrap sample of 100 is taken from the aggregated pricipitation series to estimate gamma parameters and to calculate the index.
+The dashed line at SPI = -2 represents an extreme drought as defined by the SPI.
+Most parts of the confidence intervals from 2019 to 2020 sit below the extreme drought line (-2) and are relatively wide compared to other time periods.
+This suggests that while it is certain that the Texas post office is suffering from a drastic drought, there is considerable uncertainty in quantifying its severity, given the extremity of the event."
 res <- queensland %>% 
@@
     q025 = quantile(.index, 0.025),
-    q05 = quantile(.index, 0.05),
     q10 = quantile(.index, 0.1),
-    q50 = quantile(.index, 0.5),
     q90 = quantile(.index, 0.9),
-    q95 = quantile(.index, 0.95),
     q975 = quantile(.index, 0.975)
@@
 The framework proposed in the paper is universal to index across diverse domains.
-Examples are provided, including the drought indexes (SPI and SPEI) and Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions and to illustrate how slight adjustment of linear combination weights impact the index.
+Examples are provided, including the drought indexes (SPI and SPEI) and Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions, to illustrate how slight adjustment of linear combination weights impact the index, and to calculate confidence intervals on the index.
+
+<!-- When adopting the pipeline approach to construct indexes, analysts may consider developing software that can be readily deployed in the cloud for production purposes. -->
 

### 2023-08-20 tidyindex.tex
writing up the confidence interval example

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, interior hidden, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{every-distribution-every-scale-every-index-all-at-once}{%
+\hypertarget{sec-example1}{%
 \subsection{Every distribution, every scale, every index all at
-once}\label{every-distribution-every-scale-every-index-all-at-once}}
+once}\label{sec-example1}}
 
@@
 
-\hypertarget{confidence-interval}{%
-\subsection{Confidence interval}\label{confidence-interval}}
+\hypertarget{confidence-interval-to-be-changed}{%
+\subsection{Confidence interval {[}TO BE
+CHANGED{]}}\label{confidence-interval-to-be-changed}}
+
+Index uncertainties may arise from various stages of the pipeline,
+including from different parameterisation choices, as illustrated from
+Section~\ref{sec-example1}, or from the statistical procedures in the
+pipeline. Given the lack of direct instruments for measurement, true
+values for indexes can never be obtained to compare the index against.
+Nevertheless, an index can be provided with confidence interval to
+report it accuracy. In this example, the Texas post office station
+highlighted in Figure~\ref{fig-compute-temporal} is used to illustrate
+the calculation of confidence interval for the Standardized
+Precipitation Index (SPI). Bootstrapping is used to account for the
+sampling uncertainty in the distribution fit step of the index pipeline
+and to assess its impact on the SPI series.
+
+In SPI, the distribution fit step fits the gamma distribution to the
+aggregated precipitation series separately for each month. This results
+in 31 or 32 points, from 1990 Jan to 2022 Apr, for estimating each set
+of distribution parameters. To account for this sampling uncertainty
+with these samples, bootstrapping is used to generate replicates of the
+aggregated series. In the \texttt{tidyindex} package, this bootstrap
+sampling is activated when the argument \texttt{.n\_boot} is set to a
+value other than the default of 1. In the following code, the
+Standardized Precipitation Index (SPI) is calculated using a time scale
+of 24. The bootstrap procedure samples the aggregated precipitation
+(\texttt{.agg}) for 100 iterations ( \texttt{.n\_boot\ =\ 100}) and then
+fits the gamma distribution. The resulting gamma probabilities are then
+transformed into normal densities in the normalizing step with
+\texttt{agument()}.
 
@@
 
-Figure~\ref{fig-conf-interval} presents the 80\% and 95\% confidence
-interval for the Texas post office station, in Queensland, Australia.
-Since the start of 2019, both confidence intervals indicate an extreme
-drought condition (below -2). early warning for the bushfire?
-
-also the ABS one - could display some uncertainties
+The confidence interval can then be calculated using the quantile method
+from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
+80\% and 95\% confidence interval for the Texas post office station, in
+Queensland, Australia. From the start of 2019 to 2020, the majority of
+the confidence intervals lie below the extreme drought line (SPI = -2),
+suggesting a high level of certainty that the Texas post office is
+suffering from a drastic drought. The relatively wide confidence
+interval, as well as during the excessive precipitation events in
+1996-1998 and 1999-2000, suggests a high variation of the gamma
+parameters estimated from the bootstrap samples and its difficulty to
+accurately quantifying the drought/ flood severity in extreme events.
 
@@
 \caption{\label{fig-conf-interval}80\% and 95\% confidence interval of
-the Standardized Precipitation Index (SPI) for the Texas post office
+the Standardized Precipitation Index (SPI-24) for the Texas post office
 station, in Queensland, Australia. A bootstrap sample of 100 is taken
 from the aggregated pricipitation series to estimate gamma parameters
-and calculate the index. Since the start of 2019, both confidence
-intervals indicate an extreme drought condition (below -2).}
+and to calculate the index. The dashed line at SPI = -2 represents an
+extreme drought as defined by the SPI. Most parts of the confidence
+intervals from 2019 to 2020 sit below the extreme drought line (-2) and
+are relatively wide compared to other time periods. This suggests that
+while it is certain that the Texas post office is suffering from a
+drastic drought, there is considerable uncertainty in quantifying its
+severity, given the extremity of the event.}
 
@@
 Gap Index (GGGI), to demonstrate the index calculation with different
-time scales and distributions and to illustrate how slight adjustment of
-linear combination weights impact the index.
+time scales and distributions, to illustrate how slight adjustment of
+linear combination weights impact the index, and to calculate confidence
+intervals on the index.
 

### 2023-08-20 tidyindex.qmd
various small changes

@@
   Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making.
-While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their value and variation in values.
+While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their values and variation in values.
 Here we discuss a modular data pipeline recommendation to assemble indexes.
 It is universally applicable to index computation and allows investigation of index behavior as part of the development procedure.
-One can compute with different the index with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess an index's robustness.
-The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on a country's ranking; and how to calculate bootstrap confidence intervals for XXX index.
+One can compute indexes with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess indexes' robustness.
+The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on countries' ranking; and how to calculate bootstrap confidence intervals for the Standardized Precipitation Index (SPI).
 The methods are supported by a new R package, called `tidyindex`.
@@
 Examples include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, and the Human Development Index.
-In environmental science climate indexes are produced by major monitoring centers, like the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early detection of natural disasters.
+In environmental science, climate indexes are produced by major monitoring centers, like the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early detection of natural disasters.
 In economics, indexes provide insight into market trends through combining prices of a basket of goods and services.
@@
 @uncertainty describes incorporating uncertainty estimates and conducting sensitivity analysis on composite indexes.
-@tate_social_2012, similarly, makes a comparative assessment of social vulnerability indexes based on uncertainty estimation and sensitivity analysis.
-(XXX Something about Ursula's colleagues paper here too.)
+@tate_social_2012 and @tate_uncertainty_2013, similarly, make a comparative assessment of social vulnerability indexes based on uncertainty estimation and sensitivity analysis.
+@laimighofer_how_2022 studies five uncertainty sources (record length, observation period, distribution choice, parameter estimation method, and GOF-test) of drought indexes.
 
@@
 
-From reviewing this literature, and in the process of developing methods for making it easier to work with multivariate spatiotemporal data, it seems possible to think about indexes in a more organised, cohesive and standard manner.
+From reviewing this literature, and in the process of developing methods for making it easier to work with multivariate spatio- temporal data, it seems possible to think about indexes in a more organised, cohesive and standard manner.
 Actually, it seems that the area could benefit from a *tidy* approach.
@@
 "
-#| fig-height: 8
-#| fig-width: 12
+#| fig-height: 6
+#| fig-width: 8
 set.seed(123)
@@
     var %in% c("origin", "z-score", "minmax", "centering"),
-    "Scaling", "Var. Trans.")
+    "Scaling", "Variable Transformation")
     )
@@
 latex_df <- tibble(
-  var = c("origin", "z-score", "qudratic", "log", "square root", "cubic root",
-          "minmax", "boxcox", "centering"),
+  var = c("origin", "z-score", "qudratic", "log", "square root",
+      "cubic root","minmax", "boxcox", "centering"),
   latex = c(
@@
              parse = TRUE) + 
-  facet_wrap(vars(var), scales = "free") + 
+  facet_wrap(vars(factor(var, levels = c(
+      "origin", "centering", "minmax","z-score", "boxcox", "cubic root",
+      "log", "qudratic", "square root"))), scales = "free") + 
   scale_color_brewer(palette = "Dark2", name = "module") + 
@@
 #| fig-width: 6
+#| fig-height: 3
 queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
@@
               axis_bar_label_col = c(rep("#000000", 3), "red"), 
-              axis_label_cex = 3, frame_x_pos = 0.16, 
-              axis_var_cex = 2.5,
+              axis_label_cex = 3, frame_x_pos = 0.16,  frame_cex = 4,
+              axis_var_cex = 3.6, 
               label = dt$country, abb_vars = FALSE),
@@
 
-TODO: increase font for variable names
-
-## Confidence interval [TO BE CHANGED]
+## Decoding uncertainty through the wisdom of the crowd
 
@@
 In SPI, the distribution fit step fits the gamma distribution to the aggregated precipitation series separately for each month.
-This results in  31 or 32 points, from 1990 Jan to 2022 Apr, for estimating each set of distribution parameters.
+This results in  31 or 32 points, from January 1990 to April 2022, for estimating each set of distribution parameters.
 To account for this sampling uncertainty with these samples, bootstrapping is used to generate replicates of the aggregated series.
@@
 In the following code, the Standardized Precipitation Index (SPI) is calculated using a time scale of 24.
-The bootstrap procedure samples the aggregated precipitation (`.agg`) for 100 iterations ( `.n_boot = 100`) and then fits the gamma distribution.
+The bootstrap procedure samples the aggregated precipitation (`.agg`) for 100 iterations (`.n_boot = 100`) and then fits the gamma distribution.
 The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `agument()`.
@@
 From the start of 2019 to 2020,  the majority of the confidence intervals lie below the extreme drought line (SPI = -2), suggesting a high level of certainty that the Texas post office is suffering from a drastic drought.
-The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantifying the drought/ flood severity in extreme events.
+The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantifying the drought and flood severity in extreme events.
 
@@
 The dashed line at SPI = -2 represents an extreme drought as defined by the SPI.
-Most parts of the confidence intervals from 2019 to 2020 sit below the extreme drought line (-2) and are relatively wide compared to other time periods.
+Most parts of the confidence intervals from 2019 to 2020 sit below the extreme drought line and are relatively wide compared to other time periods.
 This suggests that while it is certain that the Texas post office is suffering from a drastic drought, there is considerable uncertainty in quantifying its severity, given the extremity of the event."

### 2023-08-20 tidyindex.tex
various small changes

@@
 index behavior in different data conditions, and to determine how their
-structure affects their value and variation in values. Here we discuss a
-modular data pipeline recommendation to assemble indexes. It is
+structure affects their values and variation in values. Here we discuss
+a modular data pipeline recommendation to assemble indexes. It is
 universally applicable to index computation and allows investigation of
 index behavior as part of the development procedure. One can compute
-with different the index with different parameter choices, adjust steps
-in the index definition by adding, removing, and swapping them to
-experiment with various index designs, calculate uncertainty measures,
-and assess an index's robustness. The paper presents three examples to
-illustrate the pipeline framework usage: comparison of two different
-indexes designed to monitor the spatio-temporal distribution of drought
-in Queensland, Australia; the effect of dimension reduction choices on
-the Global Gender Gap Index (GGGI) on a country's ranking; and how to
-calculate bootstrap confidence intervals for XXX index. The methods are
-supported by a new R package, called \texttt{tidyindex}.
+indexes with different parameter choices, adjust steps in the index
+definition by adding, removing, and swapping them to experiment with
+various index designs, calculate uncertainty measures, and assess
+indexes' robustness. The paper presents three examples to illustrate the
+pipeline framework usage: comparison of two different indexes designed
+to monitor the spatio-temporal distribution of drought in Queensland,
+Australia; the effect of dimension reduction choices on the Global
+Gender Gap Index (GGGI) on countries' ranking; and how to calculate
+bootstrap confidence intervals for the Standardized Precipitation Index
+(SPI). The methods are supported by a new R package, called
+\texttt{tidyindex}.
 \end{abstract}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, interior hidden, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 Niño-Southern Oscillation Index, Consumer Price Index, QS University
-Rankings, and the Human Development Index. In environmental science
+Rankings, and the Human Development Index. In environmental science,
 climate indexes are produced by major monitoring centers, like the
@@
 (2005) describes incorporating uncertainty estimates and conducting
-sensitivity analysis on composite indexes. Tate (2012), similarly, makes
-a comparative assessment of social vulnerability indexes based on
-uncertainty estimation and sensitivity analysis. (XXX Something about
-Ursula's colleagues paper here too.)
+sensitivity analysis on composite indexes. Tate (2012) and Tate (2013),
+similarly, make a comparative assessment of social vulnerability indexes
+based on uncertainty estimation and sensitivity analysis. Laimighofer
+and Laaha (2022) studies five uncertainty sources (record length,
+observation period, distribution choice, parameter estimation method,
+and GOF-test) of drought indexes.
 
@@
 From reviewing this literature, and in the process of developing methods
-for making it easier to work with multivariate spatiotemporal data, it
+for making it easier to work with multivariate spatio- temporal data, it
 seems possible to think about indexes in a more organised, cohesive and
@@
 
-TODO: increase font for variable names
-
-\hypertarget{confidence-interval-to-be-changed}{%
-\subsection{Confidence interval {[}TO BE
-CHANGED{]}}\label{confidence-interval-to-be-changed}}
+\hypertarget{decoding-uncertainty-through-the-wisdom-of-the-crowd}{%
+\subsection{Decoding uncertainty through the wisdom of the
+crowd}\label{decoding-uncertainty-through-the-wisdom-of-the-crowd}}
 
@@
 aggregated precipitation series separately for each month. This results
-in 31 or 32 points, from 1990 Jan to 2022 Apr, for estimating each set
-of distribution parameters. To account for this sampling uncertainty
+in 31 or 32 points, from January 1990 to April 2022, for estimating each
+set of distribution parameters. To account for this sampling uncertainty
 with these samples, bootstrapping is used to generate replicates of the
@@
 of 24. The bootstrap procedure samples the aggregated precipitation
-(\texttt{.agg}) for 100 iterations ( \texttt{.n\_boot\ =\ 100}) and then
+(\texttt{.agg}) for 100 iterations (\texttt{.n\_boot\ =\ 100}) and then
 fits the gamma distribution. The resulting gamma probabilities are then
@@
 parameters estimated from the bootstrap samples and its difficulty to
-accurately quantifying the drought/ flood severity in extreme events.
+accurately quantifying the drought and flood severity in extreme events.
 
@@
 extreme drought as defined by the SPI. Most parts of the confidence
-intervals from 2019 to 2020 sit below the extreme drought line (-2) and
-are relatively wide compared to other time periods. This suggests that
-while it is certain that the Texas post office is suffering from a
-drastic drought, there is considerable uncertainty in quantifying its
-severity, given the extremity of the event.}
+intervals from 2019 to 2020 sit below the extreme drought line and are
+relatively wide compared to other time periods. This suggests that while
+it is certain that the Texas post office is suffering from a drastic
+drought, there is considerable uncertainty in quantifying its severity,
+given the extremity of the event.}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-laimighofer_how_2022}{}}%
+Laimighofer, Johannes, and Gregor Laaha. 2022. {``How Standard Are
+Standardized Drought Indices? {Uncertainty} Components for the {SPI} \&
+{SPEI} Case.''} \emph{Journal of Hydrology} 613 (October): 128385.
+\url{https://doi.org/10.1016/j.jhydrol.2022.128385}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-gpindex}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-tate_uncertainty_2013}{}}%
+---------. 2013. {``Uncertainty {Analysis} for a {Social}
+{Vulnerability} {Index}.''} \emph{Annals of the Association of American
+Geographers} 103 (3): 526--43.
+\url{https://doi.org/10.1080/00045608.2012.700616}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%

### 2023-08-20 tidyindex.qmd
a paragraph on linear combination, pca, and linear projections in dimension reduction

@@
 | 3.5   | Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q$      |
-| 3.6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $p_j(s;t) \quad p(.) \in [0, 1]$                   |
+| 3.6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $P_j(s;t) \quad P(.) \in [0, 1]$                   |
 | 3.7   | Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
@@
 
+Common statistical methods for reducing the variable dimension include linear combinations, Principal Component Analysis (PCA), and linear projections.
+While all the three methods can lead in the same result due to their shared linear approach in combining variables in an additive structure: $\sum_{j = 1}^{p}\lambda_{j}x_j(s;t)$ where $\lambda$ is the weight or principal component, they offer different interpretation of the weight.
+Linear combination typically use equal contribution or use expert suggested weights for simplicity.
+Both PCA and linear projections determine their weights (principal components) through optimizing a function of the data.
+In PCA, this function is the variance while linear projections allow any function of the data to serve as the objective function in the optimization.
+
 ## Distribution fit
 
-Distribution fit applies the Cumulative Distribution Function (CDF) $F$ of a distribution on the variable $x_j(s; t)$ to obtain the probability values $p_j(s;t) \in [0, 1]$.
+Distribution fit applies the Cumulative Distribution Function (CDF) $F$ of a distribution on the variable $x_j(s; t)$ to obtain the probability values $P_j(s;t) \in [0, 1]$.
 In SPEI, many distributions, including log-logistic, Pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series.

### 2023-08-20 tidyindex.tex
a paragraph on linear combination, pca, and linear projections in dimension reduction

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, sharp corners, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
-\(p_j(s;t) \quad p(.) \in [0, 1]\) \\
+\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
 3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
@@
 
+Common statistical methods for reducing the variable dimension include
+linear combinations, Principal Component Analysis (PCA), and linear
+projections. While all the three methods can lead in the same result due
+to their shared linear approach in combining variables in an additive
+structure: \(\sum_{j = 1}^{p}\lambda_{j}x_j(s;t)\) where \(\lambda\) is
+the weight or principal component, they offer different interpretation
+of the weight. Linear combination typically use equal contribution or
+use expert suggested weights for simplicity. Both PCA and linear
+projections determine their weights (principal components) through
+optimizing a function of the data. In PCA, this function is the variance
+while linear projections allow any function of the data to serve as the
+objective function in the optimization.
+
 \hypertarget{distribution-fit}{%
@@
 \(F\) of a distribution on the variable \(x_j(s; t)\) to obtain the
-probability values \(p_j(s;t) \in [0, 1]\). In SPEI, many distributions,
+probability values \(P_j(s;t) \in [0, 1]\). In SPEI, many distributions,
 including log-logistic, Pearson III, lognormal, and general extreme

### 2023-08-21 tidyindex.qmd
tweked paragrpah on linear projections in dimension reduction

@@
 
-There are also R packages supporting index calculation.
+There are also a few R packages supporting index calculation.
 The  `SPEI` package  [@spei] computes two drought indexes.
@@
 
-From reviewing this literature, and in the process of developing methods for making it easier to work with multivariate spatio- temporal data, it seems possible to think about indexes in a more organised, cohesive and standard manner.
-Actually, it seems that the area could benefit from a *tidy* approach.
+From reviewing this literature, and in the process of developing methods for making it easier to work with multivariate spatio-temporal data, it seems possible to think about indexes in a more organised, cohesive and standard manner.
+Actually, the area could benefit from a *tidy* approach.
 
@@
 For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) [@SPEI] calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
-This is the only step that differs from the Standardized Precipitation Index (SPI).
+This is the only step that differs from the Standardized Precipitation Index (SPI), and can be considered to be a dimension reduction using a particular linear combination.
 
-Common statistical methods for reducing the variable dimension include linear combinations, Principal Component Analysis (PCA), and linear projections.
-While all the three methods can lead in the same result due to their shared linear approach in combining variables in an additive structure: $\sum_{j = 1}^{p}\lambda_{j}x_j(s;t)$ where $\lambda$ is the weight or principal component, they offer different interpretation of the weight.
+Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) (XXX-ref) or linear discriminant analysis (XXX-ref), preparing contrasts to test particular elements in analysis of variance, or hand-crafted by a content-area expert.
+Linear combinations also form the basis for visualizing multivariate data, in methods such as tours (XXX-ref).
+<!--While all the three methods can lead in the same result due to their shared linear approach in combining variables in an additive structure: $\sum_{j = 1}^{p}\lambda_{j}x_j(s;t)$ where $\lambda$ is the weight or principal component, they offer different interpretation of the weight.
 Linear combination typically use equal contribution or use expert suggested weights for simplicity.
-Both PCA and linear projections determine their weights (principal components) through optimizing a function of the data.
-In PCA, this function is the variance while linear projections allow any function of the data to serve as the objective function in the optimization.
+Both PCA and linear projections determine their weights through optimizing a function of the data.
+In PCA, this function is the variance while linear projections allow any function of the data to serve as the objective function in the optimization.-->
 

### 2023-08-21 tidyindex.tex
tweked paragrpah on linear projections in dimension reduction

--- 2023-08-20 tidyindex.tex
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-There are also R packages supporting index calculation. The
+There are also a few R packages supporting index calculation. The
 \texttt{SPEI} package (Vicente-Serrano, Beguería, and López-Moreno 2010)
@@
 From reviewing this literature, and in the process of developing methods
-for making it easier to work with multivariate spatio- temporal data, it
+for making it easier to work with multivariate spatio-temporal data, it
 seems possible to think about indexes in a more organised, cohesive and
-standard manner. Actually, it seems that the area could benefit from a
-\emph{tidy} approach.
+standard manner. Actually, the area could benefit from a \emph{tidy}
+approach.
 
@@
 This is the only step that differs from the Standardized Precipitation
-Index (SPI).
+Index (SPI), and can be considered to be a dimension reduction using a
+particular linear combination.
+
+Linear combinations of variables are commonly used to reduce the
+dimension in statistical methodology, and chosen using a method like
+principal component analysis (PCA) (XXX-ref) or linear discriminant
+analysis (XXX-ref), preparing contrasts to test particular elements in
+analysis of variance, or hand-crafted by a content-area expert. Linear
+combinations also form the basis for visualizing multivariate data, in
+methods such as tours (XXX-ref).
 

### 2023-08-21 tidyindex.qmd
XXX indicates citations needed

@@
 
-The rest of the paper is structured as follows: @sec-tidy reviews the tidy framework in R and how index construction can benefit from such a framework.
+The rest of the paper is structured as follows.
+@sec-idx-dev provides background about the development of indexes.
+@sec-tidy reviews the tidy framework in R and how index construction can benefit from such a framework.
 The details of the pipeline modules are presented in @sec-pipeline.
 @sec-software explains the design of the `tidyindex` package that implements the modules.
-Examples are given in @sec-examples to illustrate the use cases of the pipeline.
+Examples are given in @sec-examples to illustrate three use cases of the pipeline.
 
@@
 
-# Background to index development
+# Background to index development {#sec-idx-dev}
 
@@
 
-## Does a minor change in variable weights cause a tornado?
+## Does a puff of change in variable weights cause a tornado in ranks?
 
@@
 The specific kind of tour that moves between the original and nearby projections is known as a "local tour".
+XXX-ref to tourr package, and also tour theory paper (Buja https://doi.org/10.1016/S0169-7161(04)24014-7 )
 

### 2023-08-21 tidyindex.tex
XXX indicates citations needed

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, breakable, interior hidden, frame hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-The rest of the paper is structured as follows: Section~\ref{sec-tidy}
-reviews the tidy framework in R and how index construction can benefit
-from such a framework. The details of the pipeline modules are presented
-in Section~\ref{sec-pipeline}. Section~\ref{sec-software} explains the
-design of the \texttt{tidyindex} package that implements the modules.
-Examples are given in Section~\ref{sec-examples} to illustrate the use
-cases of the pipeline.
-
-\hypertarget{background-to-index-development}{%
-\section{Background to index
-development}\label{background-to-index-development}}
+The rest of the paper is structured as follows.
+Section~\ref{sec-idx-dev} provides background about the development of
+indexes. Section~\ref{sec-tidy} reviews the tidy framework in R and how
+index construction can benefit from such a framework. The details of the
+pipeline modules are presented in Section~\ref{sec-pipeline}.
+Section~\ref{sec-software} explains the design of the \texttt{tidyindex}
+package that implements the modules. Examples are given in
+Section~\ref{sec-examples} to illustrate three use cases of the
+pipeline.
+
+\hypertarget{sec-idx-dev}{%
+\section{Background to index development}\label{sec-idx-dev}}
 
@@
 
-\hypertarget{does-a-minor-change-in-variable-weights-cause-a-tornado}{%
-\subsection{Does a minor change in variable weights cause a
-tornado?}\label{does-a-minor-change-in-variable-weights-cause-a-tornado}}
+\hypertarget{does-a-puff-of-change-in-variable-weights-cause-a-tornado-in-ranks}{%
+\subsection{Does a puff of change in variable weights cause a tornado in
+ranks?}\label{does-a-puff-of-change-in-variable-weights-cause-a-tornado-in-ranks}}
 
@@
 The specific kind of tour that moves between the original and nearby
-projections is known as a ``local tour''.
+projections is known as a ``local tour''. XXX-ref to tourr package, and
+also tour theory paper (Buja
+https://doi.org/10.1016/S0169-7161(04)24014-7 )
 

### 2023-08-21 tidyindex.qmd
table 2 and first paragraph of bootstrap example

@@
 The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023].
-The index, compiled from 14 variables expressed as female-to-male ratios, first aggregates these variables into four dimensions through linear combinations.
-These dimensions are then combined through another linear combination to form the index.
-@fig-pp-gggi illustrates this pipeline construction by applying the dimension reduction module twice on the data to generate the index.
-@tbl-gggi-weights presents the variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction steps.
-In the table, the variable weights are computed as the inverse of the standard deviation of each variable, scaled to sum to 1.
-These weights ensure that a one percentage point change in the standard deviation of each variable contributes equally to the index.
+The index, defined on 14 variables measuring female-to-male ratios, first aggregates these variables into four dimensions using the linear combination (`V-weight` in @tbl-gggi-weights).
+The weights are the inverse of the standard deviation of each variable, scaled to sum to 1, thus ensuring equal relative contribution of each variable to each of the four new variables.
+These new variables are then combined through another linear combination (`D-weight` in @tbl-gggi-weights) to form the final index value.
+@fig-pp-gggi illustrates that the pipeline is constructed by applying the dimension reduction module twice on the data.
 Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weights.
@@
 ```{r tbl-gggi-weights}
-#| tbl-cap: Weights used to compute the Global Gender Gap Index
+#| tbl-cap: "Weights for the two applications of dimension reduction to compute the Global Gender Gap Index.
+V-weight is used to compute four new variables from the original 14.
+These are then equally combined to get the final index value."
 #| tbl-cap-location: bottom
+library(kableExtra)
 gggi_weights %>%  
@@
             str_to_sentence()) %>% 
-  select(c(1,2, 5:7)) %>% 
-  mutate(dimension = c(rep("Economy", 5), rep("Education" ,4), 
-                       rep("Health", 2), rep("Politics", 3))) %>% 
+  select(c(1, 5, 2, 6)) %>% 
+  mutate(dimension = c("Economy", rep("", 4), 
+                       "Education", rep("" , 3), 
+                       "Health", "", 
+                       "Politics", rep("", 2)),
+         dim_weight = c(0.25, rep("", 4),
+                        0.25, rep("", 3),
+                        0.25, "",
+                        0.25, rep("", 2))) %>% 
   knitr::kable(
     digits = 3, 
-    col.names = c("Variable", "Dimension", "V-weight",
-        "D-weight", "Weight")
-    ) 
+    col.names = c("Variable", "V-weight", "Dimension", 
+        "D-weight"),
+        linesep = c("", "", "", "", "\\hline",
+                    "", "", "", "\\hline",
+                    "", "\\hline",
+                    "", "", "\\hline"),
+      align = c("l", "r", "l", "r")
+    ) %>%
+  kable_styling(full_width = FALSE) %>%
+  row_spec(0, bold = TRUE) 
 ```
@@
 
-Index uncertainties may arise from various stages of the pipeline, including from different parameterisation choices, as illustrated from @sec-example1, or from the statistical procedures in the pipeline.
-Given the lack of direct instruments for measurement, true values for indexes can never be obtained to compare the index against.
-Nevertheless, an index can be provided with confidence interval to report it accuracy.
-In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate the calculation of confidence interval for the Standardized Precipitation Index (SPI).
+Errors in measurement, variability and sampling error, may arise at various stages of the pipeline calculation, including from different parameterisation choices, as illustrated from @sec-example1, or from the statistical summarization procedures applied in the pipeline.
+Although it may note be possible to perfectly measure these errors, it is important that they are recognised and estimated for an index, especially to provide potential confidence intervals.
+In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate one possibility to compute a confidence interval for the Standardized Precipitation Index (SPI).
 Bootstrapping is used to account for the sampling uncertainty in the distribution fit step of the index pipeline and to assess its impact on the SPI series.

### 2023-08-21 tidyindex.tex
table 2 and first paragraph of bootstrap example

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \usepackage{orcidlink}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, breakable, interior hidden, frame hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 Opportunity, Educational Attainment, Health and Survival, and Political
-Empowerment (World Economic Forum 2023). The index, compiled from 14
-variables expressed as female-to-male ratios, first aggregates these
-variables into four dimensions through linear combinations. These
-dimensions are then combined through another linear combination to form
-the index. Figure~\ref{fig-pp-gggi} illustrates this pipeline
-construction by applying the dimension reduction module twice on the
-data to generate the index. Table~\ref{tbl-gggi-weights} presents the
-variable weights (\texttt{V-weight}) and dimension weights
-(\texttt{D-weight}) used in the two dimension reduction steps. In the
-table, the variable weights are computed as the inverse of the standard
-deviation of each variable, scaled to sum to 1. These weights ensure
-that a one percentage point change in the standard deviation of each
-variable contributes equally to the index. Dimension weights are equal
-across the four dimensions and the last column, \texttt{weight},
-multiples the variable and dimension weights to produce a single set of
-weights.
+Empowerment (World Economic Forum 2023). The index, defined on 14
+variables measuring female-to-male ratios, first aggregates these
+variables into four dimensions using the linear combination
+(\texttt{V-weight} in Table~\ref{tbl-gggi-weights}). The weights are the
+inverse of the standard deviation of each variable, scaled to sum to 1,
+thus ensuring equal relative contribution of each variable to each of
+the four new variables. These new variables are then combined through
+another linear combination (\texttt{D-weight} in
+Table~\ref{tbl-gggi-weights}) to form the final index value.
+Figure~\ref{fig-pp-gggi} illustrates that the pipeline is constructed by
+applying the dimension reduction module twice on the data. Dimension
+weights are equal across the four dimensions and the last column,
+\texttt{weight}, multiples the variable and dimension weights to produce
+a single set of weights.
 
@@
 \hypertarget{tbl-gggi-weights}{}
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.5455}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1299}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1169}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0909}}@{}}
-\caption{\label{tbl-gggi-weights}Weights used to compute the Global
-Gender Gap Index}\tabularnewline
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Variable
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Dimension
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-V-weight
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-D-weight
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Weight
-\end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Variable
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Dimension
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-V-weight
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-D-weight
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-Weight
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Labour force participation & Economy & 0.199 & 0.25 & 0.050 \\
-Wage equality for similar work & Economy & 0.310 & 0.25 & 0.078 \\
-Estimated earned income & Economy & 0.221 & 0.25 & 0.055 \\
-Legislators senior officials and managers & Economy & 0.149 & 0.25 &
-0.037 \\
-Professional and technical workers & Economy & 0.121 & 0.25 & 0.030 \\
-Literacy rate & Education & 0.191 & 0.25 & 0.048 \\
-Enrolment in primary education & Education & 0.459 & 0.25 & 0.115 \\
-Enrolment in secondary education & Education & 0.230 & 0.25 & 0.058 \\
-Enrolment in tertiary education & Education & 0.121 & 0.25 & 0.030 \\
-Sex ratio at birth & Health & 0.693 & 0.25 & 0.173 \\
-Healthy life expectancy & Health & 0.307 & 0.25 & 0.077 \\
-Women in parliament & Politics & 0.310 & 0.25 & 0.078 \\
-Women in ministerial positions & Politics & 0.247 & 0.25 & 0.062 \\
-Years with female head of state & Politics & 0.443 & 0.25 & 0.111 \\
-\end{longtable}
+\begin{table}
+\caption{\label{tbl-gggi-weights}Weights for the two applications of dimension reduction to compute the
+Global Gender Gap Index. V-weight is used to compute four new variables
+from the original 14. These are then equally combined to get the final
+index value. }\tabularnewline
+
+\centering
+\begin{tabular}{l|r|l|r}
+\hline
+\textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight}\\
+\hline
+Labour force participation & 0.199 & Economy & 0.25\\
+Wage equality for similar work & 0.310 &  & \\
+Estimated earned income & 0.221 &  & \\
+Legislators senior officials and managers & 0.149 &  & \\
+Professional and technical workers & 0.121 &  & \\
+\hline
+Literacy rate & 0.191 & Education & 0.25\\
+Enrolment in primary education & 0.459 &  & \\
+Enrolment in secondary education & 0.230 &  & \\
+Enrolment in tertiary education & 0.121 &  & \\
+\hline
+Sex ratio at birth & 0.693 & Health & 0.25\\
+Healthy life expectancy & 0.307 &  & \\
+\hline
+Women in parliament & 0.310 & Politics & 0.25\\
+Women in ministerial positions & 0.247 &  & \\
+Years with female head of state & 0.443 &  & \\
+\hline
+\end{tabular}
+\end{table}
 
@@
 
-Index uncertainties may arise from various stages of the pipeline,
-including from different parameterisation choices, as illustrated from
-Section~\ref{sec-example1}, or from the statistical procedures in the
-pipeline. Given the lack of direct instruments for measurement, true
-values for indexes can never be obtained to compare the index against.
-Nevertheless, an index can be provided with confidence interval to
-report it accuracy. In this example, the Texas post office station
-highlighted in Figure~\ref{fig-compute-temporal} is used to illustrate
-the calculation of confidence interval for the Standardized
-Precipitation Index (SPI). Bootstrapping is used to account for the
-sampling uncertainty in the distribution fit step of the index pipeline
-and to assess its impact on the SPI series.
+Errors in measurement, variability and sampling error, may arise at
+various stages of the pipeline calculation, including from different
+parameterisation choices, as illustrated from
+Section~\ref{sec-example1}, or from the statistical summarization
+procedures applied in the pipeline. Although it may note be possible to
+perfectly measure these errors, it is important that they are recognised
+and estimated for an index, especially to provide potential confidence
+intervals. In this example, the Texas post office station highlighted in
+Figure~\ref{fig-compute-temporal} is used to illustrate one possibility
+to compute a confidence interval for the Standardized Precipitation
+Index (SPI). Bootstrapping is used to account for the sampling
+uncertainty in the distribution fit step of the index pipeline and to
+assess its impact on the SPI series.
 

### 2023-08-22 tidyindex.qmd
add reference to dimension reduction + misc

@@
 
-Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) (XXX-ref) or linear discriminant analysis (XXX-ref), preparing contrasts to test particular elements in analysis of variance, or hand-crafted by a content-area expert.
-Linear combinations also form the basis for visualizing multivariate data, in methods such as tours (XXX-ref).
+Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) or linear discriminant analysis [@james2013introduction; hastie2009elements], preparing contrasts to test particular elements in analysis of variance, or hand-crafted by a content-area expert.
+Linear combinations also form the basis for visualizing multivariate data, in methods such as tours [@wickham_tourr_2011].
+
 <!--While all the three methods can lead in the same result due to their shared linear approach in combining variables in an additive structure: $\sum_{j = 1}^{p}\lambda_{j}x_j(s;t)$ where $\lambda$ is the weight or principal component, they offer different interpretation of the weight.
@@
   knitr::kable(
-    digits = 3, 
+    digits = 3, booktabs = TRUE, 
     col.names = c("Variable", "V-weight", "Dimension", 
@@
     ) %>%
-  kable_styling(full_width = FALSE) %>%
+  kable_styling(full_width = FALSE, ) %>%
   row_spec(0, bold = TRUE)
@@
 Errors in measurement, variability and sampling error, may arise at various stages of the pipeline calculation, including from different parameterisation choices, as illustrated from @sec-example1, or from the statistical summarization procedures applied in the pipeline.
-Although it may note be possible to perfectly measure these errors, it is important that they are recognised and estimated for an index, especially to provide potential confidence intervals.
+Although it may not be possible to perfectly measure these errors, it is important that they are recognised and estimated for an index, especially to provide potential confidence intervals.
 In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate one possibility to compute a confidence interval for the Standardized Precipitation Index (SPI).

### 2023-08-22 tidyindex.tex
add reference to dimension reduction + misc

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 dimension in statistical methodology, and chosen using a method like
-principal component analysis (PCA) (XXX-ref) or linear discriminant
-analysis (XXX-ref), preparing contrasts to test particular elements in
-analysis of variance, or hand-crafted by a content-area expert. Linear
-combinations also form the basis for visualizing multivariate data, in
-methods such as tours (XXX-ref).
+principal component analysis (PCA) or linear discriminant analysis
+{[}James et al. (2013); hastie2009elements{]}, preparing contrasts to
+test particular elements in analysis of variance, or hand-crafted by a
+content-area expert. Linear combinations also form the basis for
+visualizing multivariate data, in methods such as tours (Wickham et al.
+2011).
 
@@
 \centering
-\begin{tabular}{l|r|l|r}
-\hline
+\begin{tabular}{lr>{}l>{}r}
+\toprule
 \textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight}\\
-\hline
+\midrule
 Labour force participation & 0.199 & Economy & 0.25\\
@@
 Years with female head of state & 0.443 &  & \\
-\hline
+\bottomrule
 \end{tabular}
@@
 Section~\ref{sec-example1}, or from the statistical summarization
-procedures applied in the pipeline. Although it may note be possible to
+procedures applied in the pipeline. Although it may not be possible to
 perfectly measure these errors, it is important that they are recognised
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-james2013introduction}{}}%
+James, Gareth, Daniela Witten, Trevor Hastie, Robert Tibshirani, et al.
+2013. \emph{An Introduction to Statistical Learning}. Vol. 112.
+Springer.
+
 \leavevmode\vadjust pre{\hypertarget{ref-jones_vulnerability_2007}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_tourr_2011}{}}%
+Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
+{``Tourr: {An} {R} {Package} for {Exploring} {Multivariate} {Data} with
+{Projections}.''} \emph{Journal of Statistical Software} 40 (2).
+\url{https://doi.org/10.18637/jss.v040.i02}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%

### 2023-08-22 tidyindex.qmd
replace the reference with  the original articles

@@
 
-Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) or linear discriminant analysis [@james2013introduction; hastie2009elements], preparing contrasts to test particular elements in analysis of variance, or hand-crafted by a content-area expert.
+Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) [@hotelling1933analysis] or linear discriminant analysis [@fisher1936use], preparing contrasts to test particular elements in analysis of variance, or hand-crafted by a content-area expert.
 Linear combinations also form the basis for visualizing multivariate data, in methods such as tours [@wickham_tourr_2011].

### 2023-08-22 tidyindex.tex
replace the reference with  the original articles

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, frame hidden, sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 dimension in statistical methodology, and chosen using a method like
-principal component analysis (PCA) or linear discriminant analysis
-{[}James et al. (2013); hastie2009elements{]}, preparing contrasts to
-test particular elements in analysis of variance, or hand-crafted by a
+principal component analysis (PCA) (Hotelling 1933) or linear
+discriminant analysis (Fisher 1936), preparing contrasts to test
+particular elements in analysis of variance, or hand-crafted by a
 content-area expert. Linear combinations also form the basis for
@@
 \centering
-\begin{tabular}{lr>{}l>{}r}
+\begin{tabular}{lrlr}
 \toprule
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-fisher1936use}{}}%
+Fisher, Ronald A. 1936. {``The Use of Multiple Measurements in Taxonomic
+Problems.''} \emph{Annals of Eugenics} 7 (2): 179--88.
+
 \leavevmode\vadjust pre{\hypertarget{ref-fundiversity}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-james2013introduction}{}}%
-James, Gareth, Daniela Witten, Trevor Hastie, Robert Tibshirani, et al.
-2013. \emph{An Introduction to Statistical Learning}. Vol. 112.
-Springer.
+\leavevmode\vadjust pre{\hypertarget{ref-hotelling1933analysis}{}}%
+Hotelling, Harold. 1933. {``Analysis of a Complex of Statistical
+Variables into Principal Components.''} \emph{Journal of Educational
+Psychology} 24 (6): 417.
 

### 2023-08-22 tidyindex.qmd
working on the gggi index section

@@
 The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023].
-The index, defined on 14 variables measuring female-to-male ratios, first aggregates these variables into four dimensions using the linear combination (`V-weight` in @tbl-gggi-weights).
+The index, defined on 14 variables measuring female-to-male ratios, first aggregates these variables into four dimensions (using the linear combination given by `V-weight` in @tbl-gggi-weights).
 The weights are the inverse of the standard deviation of each variable, scaled to sum to 1, thus ensuring equal relative contribution of each variable to each of the four new variables.
@@
 @fig-pp-gggi illustrates that the pipeline is constructed by applying the dimension reduction module twice on the data.
-Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weights.
+<!--Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weights.--> The data for GGGI does not needs to be transformed or scaled so these steps are not included, but they might still need to be used for other similar indexes.
 
@@
 
-The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi` and @tbl-gggi-weights as `gggi_weights`.
-The index can be reproduced with the package as:
+The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in the `tidyindex` package as `gggi` with @tbl-gggi-weights as `gggi_weights`.
+The index can be reproduced with:
 
@@
 
-After initialzing the `gggi` object and attaching the `gggi_weights` as meta-data, a single linear combination step within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata.
+After initialzing the `gggi` object and attaching the `gggi_weights` as meta-data, a single linear combination within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata.
 While computing the index from the original 14 variables, it remains unclear how the missing values are handled within the index, which impacts 68 out of the total 146 countries.

### 2023-08-22 tidyindex.tex
working on the gggi index section

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, frame hidden, sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 variables measuring female-to-male ratios, first aggregates these
-variables into four dimensions using the linear combination
-(\texttt{V-weight} in Table~\ref{tbl-gggi-weights}). The weights are the
+variables into four dimensions (using the linear combination given by
+\texttt{V-weight} in Table~\ref{tbl-gggi-weights}). The weights are the
 inverse of the standard deviation of each variable, scaled to sum to 1,
@@
 Figure~\ref{fig-pp-gggi} illustrates that the pipeline is constructed by
-applying the dimension reduction module twice on the data. Dimension
-weights are equal across the four dimensions and the last column,
-\texttt{weight}, multiples the variable and dimension weights to produce
-a single set of weights.
+applying the dimension reduction module twice on the data. The data for
+GGGI does not needs to be transformed or scaled so these steps are not
+included, but they might still need to be used for other similar
+indexes.
 
@@
 The 2023 GGGI data is available from the Global Gender Gap Report 2023
-in the country's economy profile and can be accessed in R via the
-\texttt{tidyindex} package as \texttt{gggi} and
+in the country's economy profile and can be accessed in the
+\texttt{tidyindex} package as \texttt{gggi} with
 Table~\ref{tbl-gggi-weights} as \texttt{gggi\_weights}. The index can be
-reproduced with the package as:
+reproduced with:
 
@@
 After initialzing the \texttt{gggi} object and attaching the
-\texttt{gggi\_weights} as meta-data, a single linear combination step
-within the dimension reduction module is applied to the 14 variables
-(from column \texttt{labour\_force\_participation} to
+\texttt{gggi\_weights} as meta-data, a single linear combination within
+the dimension reduction module is applied to the 14 variables (from
+column \texttt{labour\_force\_participation} to
 \texttt{years\_with\_female\_head\_of\_state}), using the weight

### 2023-08-22 tidyindex.qmd
remove the notation  of q in dimension reduction

@@
 | 3.4   | Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
-| 3.5   | Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q$      |
+| 3.5   | Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p$      |
 | 3.6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $P_j(s;t) \quad P(.) \in [0, 1]$                   |
@@
 
-Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$, as the input.
-It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < q$, as the output.
+Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$, or a subset of variables in $\mathbf{x}(s;t)$, as the input.
+It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output.
 The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods.

### 2023-08-22 tidyindex.tex
remove the notation  of q in dimension reduction

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
-& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q\) \\
+& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
 3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
@@
 Dimension reduction takes the multivariate information
-\(\mathbf{x}(s;t)\), where
-\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\), as the input. It
-summarises the high-dimensional information into a lower-dimension
-representation \(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\)
-and \(d < q\), as the output. The transformation can be based on
-domain-specific knowledge, originating from theories describing the
-underlying physical processes, or guided by statistical methods. For
-example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
-(Beguería and Vicente-Serrano 2017) calculates the difference \(D\)
-between precipitation (\(P\)) and potential evapotranspiration
-(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
-This is the only step that differs from the Standardized Precipitation
-Index (SPI), and can be considered to be a dimension reduction using a
-particular linear combination.
+\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\), or a
+subset of variables in \(\mathbf{x}(s;t)\), as the input. It summarises
+the high-dimensional information into a lower-dimension representation
+\(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\) and \(d < p\),
+as the output. The transformation can be based on domain-specific
+knowledge, originating from theories describing the underlying physical
+processes, or guided by statistical methods. For example, the
+Standardized Precipitation-Evapotranspiration Index (SPEI) (Beguería and
+Vicente-Serrano 2017) calculates the difference \(D\) between
+precipitation (\(P\)) and potential evapotranspiration (\(\text{PET}\)),
+using a water balance model (\(D = P - \text{PET}\)). This is the only
+step that differs from the Standardized Precipitation Index (SPI), and
+can be considered to be a dimension reduction using a particular linear
+combination.
 

### 2023-08-22 tidyindex.qmd
table 2 details

@@
             str_to_sentence()) %>% 
-  select(c(1, 5, 2, 6)) %>% 
+  select(c(1, 5, 2, 6, 7)) %>% 
   mutate(dimension = c("Economy", rep("", 4), 
@@
     col.names = c("Variable", "V-weight", "Dimension", 
-        "D-weight"),
+        "D-weight", "weight"),
         linesep = c("", "", "", "", "\\hline",
@@
                     "", "", "\\hline"),
-      align = c("l", "r", "l", "r")
+      align = c("l", "r", "l", "r", "r")
     ) %>%

### 2023-08-22 tidyindex.tex
table 2 details

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, frame hidden, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \centering
-\begin{tabular}{lrlr}
+\begin{tabular}{lrlrr}
 \toprule
-\textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight}\\
+\textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight} & \textbf{weight}\\
 \midrule
-Labour force participation & 0.199 & Economy & 0.25\\
-Wage equality for similar work & 0.310 &  & \\
-Estimated earned income & 0.221 &  & \\
-Legislators senior officials and managers & 0.149 &  & \\
-Professional and technical workers & 0.121 &  & \\
+Labour force participation & 0.199 & Economy & 0.25 & 0.050\\
+Wage equality for similar work & 0.310 &  &  & 0.078\\
+Estimated earned income & 0.221 &  &  & 0.055\\
+Legislators senior officials and managers & 0.149 &  &  & 0.037\\
+Professional and technical workers & 0.121 &  &  & 0.030\\
 \hline
-Literacy rate & 0.191 & Education & 0.25\\
-Enrolment in primary education & 0.459 &  & \\
-Enrolment in secondary education & 0.230 &  & \\
-Enrolment in tertiary education & 0.121 &  & \\
+Literacy rate & 0.191 & Education & 0.25 & 0.048\\
+Enrolment in primary education & 0.459 &  &  & 0.115\\
+Enrolment in secondary education & 0.230 &  &  & 0.058\\
+Enrolment in tertiary education & 0.121 &  &  & 0.030\\
 \hline
-Sex ratio at birth & 0.693 & Health & 0.25\\
-Healthy life expectancy & 0.307 &  & \\
+Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
+Healthy life expectancy & 0.307 &  &  & 0.077\\
 \hline
-Women in parliament & 0.310 & Politics & 0.25\\
-Women in ministerial positions & 0.247 &  & \\
-Years with female head of state & 0.443 &  & \\
+Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
+Women in ministerial positions & 0.247 &  &  & 0.062\\
+Years with female head of state & 0.443 &  &  & 0.111\\
 \bottomrule

### 2023-08-22 tidyindex.qmd
fix table margin

--- 2023-08-22 tidyindex.qmd
@@
 #| tbl-cap-location: bottom
+# table cap location is not working: https://github.com/quarto-dev/quarto-cli/issues/6417
 library(kableExtra)
@@
     digits = 3, booktabs = TRUE, 
-    col.names = c("Variable", "V-weight", "Dimension", 
-        "D-weight"),
+    col.names = c("Variable", "V-weight", "Dimension", "D-weight", "weight"),
     linesep = c("", "", "", "", "\\hline",
@@
                 "", "", "\\hline"),
-      align = c("l", "r", "l", "r")
+      align = c("l", "r", "l", "r", "r")
     ) %>%
-  kable_styling(full_width = FALSE, ) %>%
-  row_spec(0, bold = TRUE)
+  kable_styling(font_size = 9) %>%
+  row_spec(0, bold = TRUE) %>% 
+  column_spec(1, width = "18em") %>% 
+  column_spec(3, width = "4em") %>% 
+  column_spec(c(2, 4, 5), width = "5em")
 ```

### 2023-08-22 tidyindex.tex
fix table margin

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, frame hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\centering
-\begin{tabular}{lrlrr}
+\centering\begingroup\fontsize{9}{11}\selectfont
+
+\begin{tabular}{>{\raggedright\arraybackslash}p{18em}>{\raggedleft\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{5em}>{\raggedleft\arraybackslash}p{5em}}
 \toprule
@@
 \end{tabular}
+\endgroup{}
 \end{table}

### 2023-08-22 tidyindex.qmd
rework on the writing of GGGI  example

@@
 
-A linear combination can also be interpreted as a linear projection of multivariate information with a weight vector.
-Projecting data from higher to lower dimensions unavoidably leads to information loss and the weights used require careful examination to understand their effects on the index and implications for interpretation and decision-making.
-By making slight adjustments to the weight vector, we can observe how index values and country rankings change.
-For illustration, we select a set of countries including:
+Frame 12 in @fig-idx-tour presents the original index values when equal weights are applied across the four dimensions.
+Leading the rankings, Nordic countries (Iceland, Norway, and Finland) and New Zealand show a good progression on reaching gender parity, with over 80% of the gender gap closed.
+In the middle range, an index value over 0.7 is achieved for Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia.
+The last group of countries (Pakistan, Iran, Algeria, Chad), with GGGI < 0.6, have a wider gender gap to closed and Afghanistan lags further behind with GGGI = 0.4.
 
-1)  Top-ranked countries with GGGI \> 0.85,
-2)  Countries ranked between 57 and 62, with GGGI values from 0.72 to 0.73, and
-3)  Low-ranked countries with GGGI \< 0.6.
+To observe the effect of a minor weight change on the index, we slightly vary the politics weight (while maintain equal weights among other dimensions) to generate an animation of the movement of index values in response to the changing weights.
+This visualization technique, which presents a sequence of data projections, is referred to as a "tour" and the specific kind of tour used here to move between nearby projections is known as a "local tour" [@buja2005computational; @wickham_tourr_2011].
 
-We slightly vary the weight of the politics dimension from the original 0.25 while keeping the weights constant for the other three dimensions.
-This process creates an animation showing the movement of index values in response to changing weights.
-This visualization technique, which presents a sequence of data projections, is referred to as a "tour".
-The specific kind of tour that moves between the original and nearby projections is known as a "local tour".
-XXX-ref to tourr package, and also tour theory paper (Buja https://doi.org/10.1016/S0169-7161(04)24014-7 )
+In @fig-idx-tour, six frames have been chosen from the local tour to explore the role of the politics dimension to the index.
+When the weight of politics is reduced (Frame 1 and 6 vs.
+Frame 12), the difference in GGGI values between the Nordic and other countries narrows, suggesting a smaller variation among countries in achieving gender parity.
+Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs.
+Frame 12), nearly all the countries experience a decrease in GGGI values.
+A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight, which leads to an almost similar index value to the Nordics when the politics dimension is assigned a weight of 0.52 in Frame 29.
 
-In @fig-idx-tour, six frames have been chosen from the animation available at https://vimeo.com/847874016.
-When the weight of politics is reduced (Frame 1 and 6 vs.
-Frame 12), the difference in GGGI values between the top Nordic countries and mid- or low-ranked countries narrows, suggesting a smaller variation among countries in achieving gender parity.
-Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs.
-Frame 12), nearly all the countries in the three categories experience a decrease in GGGI values.
-A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight.
-This leads to its index value being almost similar to those of the top-ranked Nordic countries when the politics dimension is assigned a weight of 0.52 in Frame 29.
-
-In GGGI, the index value has a direct interpretation as the percentage of the gender gap that has been closed.
-When countries exhibit closer or consistent decreasing index values, it can be directly interpreted as a reflection of a country's progress toward gender parity.
 Ideally, an index should be robust against minor changes in its construction components.
-This example provides researchers with means to observe changes in the index resulting from variations in the weights used in linear combinations.
-This approach can be applied broadly to different sets of weights of interest, extending beyond the change in a single dimension illustrated by the example.
+Unfortunately, this is not the case with GGGI, which raises concerns about the index's quality.
+For other indexes that use linear combination to reduce variable dimensions, the demonstrated local tour can be applied similarly to assess the index robustness through observing how index values respond to perturbation of the weights.
 

### 2023-08-22 tidyindex.tex
rework on the writing of GGGI  example

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 Examples of using a pipeline approach for data analysis can be traced
-back to the interactive graphics literature, including Buja et al.
+back to the interactive graphics literature, including A. Buja et al.
 (1988); Sutherland et al. (2000); Xie, Hofmann, and Cheng (2014);
@@
 
-A linear combination can also be interpreted as a linear projection of
-multivariate information with a weight vector. Projecting data from
-higher to lower dimensions unavoidably leads to information loss and the
-weights used require careful examination to understand their effects on
-the index and implications for interpretation and decision-making. By
-making slight adjustments to the weight vector, we can observe how index
-values and country rankings change. For illustration, we select a set of
-countries including:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
-\tightlist
-\item
-  Top-ranked countries with GGGI \textgreater{} 0.85,
-\item
-  Countries ranked between 57 and 62, with GGGI values from 0.72 to
-  0.73, and
-\item
-  Low-ranked countries with GGGI \textless{} 0.6.
-\end{enumerate}
-
-We slightly vary the weight of the politics dimension from the original
-0.25 while keeping the weights constant for the other three dimensions.
-This process creates an animation showing the movement of index values
-in response to changing weights. This visualization technique, which
-presents a sequence of data projections, is referred to as a ``tour''.
-The specific kind of tour that moves between the original and nearby
-projections is known as a ``local tour''. XXX-ref to tourr package, and
-also tour theory paper (Buja
-https://doi.org/10.1016/S0169-7161(04)24014-7 )
-
-In Figure~\ref{fig-idx-tour}, six frames have been chosen from the
-animation available at https://vimeo.com/847874016. When the weight of
-politics is reduced (Frame 1 and 6 vs.~Frame 12), the difference in GGGI
-values between the top Nordic countries and mid- or low-ranked countries
+Frame 12 in Figure~\ref{fig-idx-tour} presents the original index values
+when equal weights are applied across the four dimensions. Leading the
+rankings, Nordic countries (Iceland, Norway, and Finland) and New
+Zealand show a good progression on reaching gender parity, with over
+80\% of the gender gap closed. In the middle range, an index value over
+0.7 is achieved for Brazil, Panama, Poland, Bangladesh, Kazakhstan,
+Armenia, and Slovakia. The last group of countries (Pakistan, Iran,
+Algeria, Chad), with GGGI \textless{} 0.6, have a wider gender gap to
+closed and Afghanistan lags further behind with GGGI = 0.4.
+
+To observe the effect of a minor weight change on the index, we slightly
+vary the politics weight (while maintain equal weights among other
+dimensions) to generate an animation of the movement of index values in
+response to the changing weights. This visualization technique, which
+presents a sequence of data projections, is referred to as a ``tour''
+and the specific kind of tour used here to move between nearby
+projections is known as a ``local tour'' (Andreas Buja et al. 2005;
+Wickham et al. 2011).
+
+In Figure~\ref{fig-idx-tour}, six frames have been chosen from the local
+tour to explore the role of the politics dimension to the index. When
+the weight of politics is reduced (Frame 1 and 6 vs.~Frame 12), the
+difference in GGGI values between the Nordic and other countries
 narrows, suggesting a smaller variation among countries in achieving
 gender parity. Conversely, when the weight of politics increases (Frame
-18, 24, and 29 vs.~Frame 12), nearly all the countries in the three
-categories experience a decrease in GGGI values. A noticeable exception
-to this trend is Bangladesh, where its index value moves in the same
-direction as the politics weight. This leads to its index value being
-almost similar to those of the top-ranked Nordic countries when the
-politics dimension is assigned a weight of 0.52 in Frame 29.
-
-In GGGI, the index value has a direct interpretation as the percentage
-of the gender gap that has been closed. When countries exhibit closer or
-consistent decreasing index values, it can be directly interpreted as a
-reflection of a country's progress toward gender parity. Ideally, an
-index should be robust against minor changes in its construction
-components. This example provides researchers with means to observe
-changes in the index resulting from variations in the weights used in
-linear combinations. This approach can be applied broadly to different
-sets of weights of interest, extending beyond the change in a single
-dimension illustrated by the example.
+18, 24, and 29 vs.~Frame 12), nearly all the countries experience a
+decrease in GGGI values. A noticeable exception to this trend is
+Bangladesh, where its index value moves in the same direction as the
+politics weight, which leads to an almost similar index value to the
+Nordics when the politics dimension is assigned a weight of 0.52 in
+Frame 29.
+
+Ideally, an index should be robust against minor changes in its
+construction components. Unfortunately, this is not the case with GGGI,
+which raises concerns about the index's quality. For other indexes that
+use linear combination to reduce variable dimensions, the demonstrated
+local tour can be applied similarly to assess the index robustness
+through observing how index values respond to perturbation of the
+weights.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-buja2005computational}{}}%
+Buja, Andreas, Dianne Cook, Daniel Asimov, and Catherine Hurley. 2005.
+{``Computational Methods for High-Dimensional Rotations in Data
+Visualization.''} \emph{Handbook of Statistics} 24: 391--413.
+\url{https://doi.org/10.1016/S0169-7161(04)24014-7}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-fisher1936use}{}}%

### 2023-08-22 tidyindex.qmd
small changes suggested by Nicolas

@@
 The construction process is often not straightforward, and decisions need to be made, such as the selection of variables to be included, which might depend on data availability and the statistical definition of the index to be used, among others.
-For instance, the indexes constructed from a linear combination of variables need to decide on the weight assigned to each variable.
+For instance, the indexes constructed from a linear combination of variables require to decide the weight assigned to each variable.
 Some indexes have a spatial and/or temporal component, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes.
@@
 We call this process of breaking the index construction into different steps the index pipeline.
-Such decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes and calculating index uncertainty.
+Such decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes,  calculating index uncertainty, and assessing index robustness.
 
@@
 
-Examples of using a pipeline approach for data analysis can be traced back to the interactive graphics literature, including @buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009.
+Examples of using a pipeline approach for data analysis can be traced back to the interactive graphics literature, including @buja_elements_1988; @sutherland_orca_2000; @wickham_plumbing_2009; @xie_reactive_2014.
 @wickham_plumbing_2009 argue that whether made explicit or not, a pipeline has to be presented in every graphics program, and making them explicit is beneficial for understanding the implementation and comparing between different graphic systems.
@@
 
-Adoption of this pipeline approach would provide uniformity to the field of index development, research, and application.
+Adoption of this pipeline approach would provide uniformity to the field of index development, research, and application to improve comparability, reporducibility, and communication
 
@@
 | 3.8   | Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
-| 3.9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                      |
+| 3.9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$                                      |
 
@@
 When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality.
-For example, in Human Development Index (HDI), a logarithmic transformation is applied to the variable Gross National Income per capita (GNI), to reduce its impact on HDI, particularly for countries with high GNI values.
+For example, in the Human Development Index (HDI), a logarithmic transformation is applied to the variable Gross National Income per capita (GNI), to reduce its impact on HDI, particularly for countries with high GNI values.
 
@@
 It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$.
-In Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.
+In the Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.
 
@@
        `z-score` = (x - mean(x)/ sd(x)), 
-       qudratic = x^2,
+       quadratic = x^2,
        log = log(x),
@@
 latex_df <- tibble(
-  var = c("origin", "z-score", "qudratic", "log", "square root",
+  var = c("origin", "z-score", "quadratic", "log", "square root",
       "cubic root","minmax", "boxcox", "centering"),
@@
       "origin", "centering", "minmax","z-score", "boxcox", "cubic root",
-      "log", "qudratic", "square root"))), scales = "free") + 
+      "log", "quadratic", "square root"))), scales = "free") + 
   scale_color_brewer(palette = "Dark2", name = "module") + 
@@
 
-Simplification takes a continuous variable $x_j(s;t)$ and categorises it into a discrete set $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$ through a piecewise function,
+Simplification takes a continuous variable $x_j(s;t)$ and categorises it into a discrete set $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$ through a piecewise constant function,
 
@@
 @fig-compute-spatial and @fig-compute-temporal exemplify two possibilities.
-@fig-compute-spatial presents the spatial distribution of SPI during two periods: October 2010 to March 2011 for the 2010/11 Queensland flood and October 2019 to March 2020 for the 2019 Australia drought, which contributes to the notorious 2019/20 bushfire.
+@fig-compute-spatial presents the spatial distribution of SPI during two periods: October 2010 to March 2011 for the 2010/11 Queensland flood and October 2019 to March 2020 for the 2019 Australia drought, which contributes to the notorious 2019/20 bushfire season.
 @fig-compute-temporal displays the sensitivity of the SPEI series at the Texas post office to different time scales and fitted distributions.
@@
 The bootstrap procedure samples the aggregated precipitation (`.agg`) for 100 iterations (`.n_boot = 100`) and then fits the gamma distribution.
-The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `agument()`.
+The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `augment()`.
 
@@
   augment(.var = .agg)
-
 ```
@@
 From the start of 2019 to 2020,  the majority of the confidence intervals lie below the extreme drought line (SPI = -2), suggesting a high level of certainty that the Texas post office is suffering from a drastic drought.
-The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantifying the drought and flood severity in extreme events.
+The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantify the drought and flood severity in extreme events.
 

### 2023-08-22 tidyindex.tex
small changes suggested by Nicolas

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 
@@
 definition of the index to be used, among others. For instance, the
-indexes constructed from a linear combination of variables need to
-decide on the weight assigned to each variable. Some indexes have a
-spatial and/or temporal component, and variables can be aggregated to
-different spatial resolutions and temporal scales, leading to various
-indexes for different monitoring purposes. Hence, all these decisions
-can result in different index values and have different practical
-implications.
+indexes constructed from a linear combination of variables require to
+decide the weight assigned to each variable. Some indexes have a spatial
+and/or temporal component, and variables can be aggregated to different
+spatial resolutions and temporal scales, leading to various indexes for
+different monitoring purposes. Hence, all these decisions can result in
+different index values and have different practical implications.
 
@@
 means to standardize index construction via a pipeline and offers
-benefits for comparing among indexes and calculating index uncertainty.
+benefits for comparing among indexes, calculating index uncertainty, and
+assessing index robustness.
 
@@
 back to the interactive graphics literature, including A. Buja et al.
-(1988); Sutherland et al. (2000); Xie, Hofmann, and Cheng (2014);
-Wickham et al. (2009). Wickham et al. (2009) argue that whether made
-explicit or not, a pipeline has to be presented in every graphics
-program, and making them explicit is beneficial for understanding the
-implementation and comparing between different graphic systems. While
-this comment is made in the context of interactive graphics programs, it
-is also applicable generally to any data analysis workflow. More
-recently, the tidyverse suite (Wickham et al. 2019) takes the pipeline
-approach for general-purpose data wrangling and has gained popularity
-within the R community. The pipeline-style code can be directly read as
-a series of operations applied successively on tidy data objects,
-offering a method to document the data wrangling process with all the
-computational details for reproducibility.
+(1988); Sutherland et al. (2000); Wickham et al. (2009); Xie, Hofmann,
+and Cheng (2014). Wickham et al. (2009) argue that whether made explicit
+or not, a pipeline has to be presented in every graphics program, and
+making them explicit is beneficial for understanding the implementation
+and comparing between different graphic systems. While this comment is
+made in the context of interactive graphics programs, it is also
+applicable generally to any data analysis workflow. More recently, the
+tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
+general-purpose data wrangling and has gained popularity within the R
+community. The pipeline-style code can be directly read as a series of
+operations applied successively on tidy data objects, offering a method
+to document the data wrangling process with all the computational
+details for reproducibility.
 
@@
 Adoption of this pipeline approach would provide uniformity to the field
-of index development, research, and application.
+of index development, research, and application to improve
+comparability, reporducibility, and communication
 
@@
 3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
-\(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
+\(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\) \\
 \bottomrule()
@@
 distribution, transformations such as log, square root, or cubic root
-can adjust the distribution towards normality. For example, in Human
+can adjust the distribution towards normality. For example, in the Human
 Development Index (HDI), a logarithmic transformation is applied to the
@@
 standardization, and min-max standardization and can be expressed as
-\([x_{j}(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
-three dimensions (health, education, and economy) are converted into the
-same scale (0-1) using min-max standardization.
+\([x_{j}(s;t) - \alpha]/\gamma\). In the Human Development Index (HDI),
+the three dimensions (health, education, and economy) are converted into
+the same scale (0-1) using min-max standardization.
 
@@
 Simplification takes a continuous variable \(x_j(s;t)\) and categorises
-it into a discrete set \(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\)
-through a piecewise function,
+it into a discrete set \(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\)
+through a piecewise constant function,
 
@@
 2020 for the 2019 Australia drought, which contributes to the notorious
-2019/20 bushfire. Figure~\ref{fig-compute-temporal} displays the
+2019/20 bushfire season. Figure~\ref{fig-compute-temporal} displays the
 sensitivity of the SPEI series at the Texas post office to different
@@
 transformed into normal densities in the normalizing step with
-\texttt{agument()}.
+\texttt{augment()}.
 
@@
 parameters estimated from the bootstrap samples and its difficulty to
-accurately quantifying the drought and flood severity in extreme events.
+accurately quantify the drought and flood severity in extreme events.
 

### 2023-08-22 tidyindex.qmd
dimension reduction module explanation expanded

@@
 
-Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) [@hotelling1933analysis] or linear discriminant analysis [@fisher1936use], preparing contrasts to test particular elements in analysis of variance, or hand-crafted by a content-area expert.
+Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) [@hotelling1933analysis] or linear discriminant analysis [@fisher1936use], preparing contrasts to test particular elements in analysis of variance (XXX-ref), or hand-crafted by a content-area expert.
 Linear combinations also form the basis for visualizing multivariate data, in methods such as tours [@wickham_tourr_2011].
+This dimension reduction method can accommodate linear combinations as provided by any method, and hence is linear by design.
+The transformation module provides variable-wise non-linear transformation.
 

### 2023-08-22 tidyindex.tex
dimension reduction module explanation expanded

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 \(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\) \\
-\bottomrule()
 \end{longtable}
@@
 discriminant analysis (Fisher 1936), preparing contrasts to test
-particular elements in analysis of variance, or hand-crafted by a
-content-area expert. Linear combinations also form the basis for
+particular elements in analysis of variance (XXX-ref), or hand-crafted
+by a content-area expert. Linear combinations also form the basis for
 visualizing multivariate data, in methods such as tours (Wickham et al.
-2011).
+2011). This dimension reduction method can accommodate linear
+combinations as provided by any method, and hence is linear by design.
+The transformation module provides variable-wise non-linear
+transformation.
 

### 2023-08-22 tidyindex.tex
tweaking table location

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
-\caption{\label{tbl-notation}An notation overview of the input,
+\caption{\label{tbl-notation}Summary of the notation for input,
 operation, and output of each pipeline module.}\tabularnewline

### 2023-08-22 tidyindex.qmd
tweaking table location
table 2 width fixed

@@
 
-: An notation overview of the input, operation, and output of each pipeline module. {#tbl-notation}
+: Summary of the notation for input, operation, and output of each pipeline module. {#tbl-notation}
 
@@
 The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023].
-The index, defined on 14 variables measuring female-to-male ratios, first aggregates these variables into four dimensions (using the linear combination given by `V-weight` in @tbl-gggi-weights).
+The index, defined on 14 variables measuring female-to-male ratios, first aggregates these variables into four dimensions (using the linear combination given by `V-wgt` in @tbl-gggi-weights).
 The weights are the inverse of the standard deviation of each variable, scaled to sum to 1, thus ensuring equal relative contribution of each variable to each of the four new variables.
-These new variables are then combined through another linear combination (`D-weight` in @tbl-gggi-weights) to form the final index value.
+These new variables are then combined through another linear combination (`D-wgt` in @tbl-gggi-weights) to form the final index value.
 @fig-pp-gggi illustrates that the pipeline is constructed by applying the dimension reduction module twice on the data.
@@
 #| tbl-cap: "Weights for the two applications of dimension reduction to compute the Global Gender Gap Index.
-V-weight is used to compute four new variables from the original 14.
+V-wgt is used to compute four new variables from the original 14.
 These are then equally combined to get the final index value."
@@
     digits = 3, booktabs = TRUE, 
-    col.names = c("Variable", "V-weight", "Dimension", "D-weight", "weight"),
+    col.names = c("Variable", "V-wgt", "Dimension", "D-wgt", "wgt"),
     linesep = c("", "", "", "", "\\hline",
@@
     ) %>%
-  kable_styling(font_size = 9) %>%
+  kable_styling(font_size = 11) %>%
   row_spec(0, bold = TRUE) %>% 
-  column_spec(1, width = "18em") %>% 
-  column_spec(3, width = "4em") %>% 
-  column_spec(c(2, 4, 5), width = "5em")
+  column_spec(1, width = "17.5em") %>% 
+  column_spec(3, width = "3.5em") %>% 
+  column_spec(c(2, 4), width = "4em") %>% 
+  column_spec(5, width = "3em")
 ```
@@
 
-After initialzing the `gggi` object and attaching the `gggi_weights` as meta-data, a single linear combination within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `weight` column of the attached metadata.
+After initializing the `gggi` object and attaching the `gggi_weights` as meta-data, a single linear combination within the dimension reduction module is applied to the 14 variables (from column `labour_force_participation` to `years_with_female_head_of_state`), using the weight specified in the `wgt` column of the attached metadata.
 While computing the index from the original 14 variables, it remains unclear how the missing values are handled within the index, which impacts 68 out of the total 146 countries.

### 2023-08-22 tidyindex.tex
table 2 width fixed

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, sharp corners, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 variables into four dimensions (using the linear combination given by
-\texttt{V-weight} in Table~\ref{tbl-gggi-weights}). The weights are the
+\texttt{V-wgt} in Table~\ref{tbl-gggi-weights}). The weights are the
 inverse of the standard deviation of each variable, scaled to sum to 1,
@@
 the four new variables. These new variables are then combined through
-another linear combination (\texttt{D-weight} in
+another linear combination (\texttt{D-wgt} in
 Table~\ref{tbl-gggi-weights}) to form the final index value.
@@
 \caption{\label{tbl-gggi-weights}Weights for the two applications of dimension reduction to compute the
-Global Gender Gap Index. V-weight is used to compute four new variables
+Global Gender Gap Index. V-wgt is used to compute four new variables
 from the original 14. These are then equally combined to get the final
@@
 
-\centering\begingroup\fontsize{9}{11}\selectfont
-
-\begin{tabular}{>{\raggedright\arraybackslash}p{18em}>{\raggedleft\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{5em}>{\raggedleft\arraybackslash}p{5em}}
+\centering\begingroup\fontsize{11}{13}\selectfont
+
+\begin{tabular}{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
 \toprule
-\textbf{Variable} & \textbf{V-weight} & \textbf{Dimension} & \textbf{D-weight} & \textbf{weight}\\
+\textbf{Variable} & \textbf{V-wgt} & \textbf{Dimension} & \textbf{D-wgt} & \textbf{wgt}\\
 \midrule
@@
 
-After initialzing the \texttt{gggi} object and attaching the
+After initializing the \texttt{gggi} object and attaching the
 \texttt{gggi\_weights} as meta-data, a single linear combination within
@@
 \texttt{years\_with\_female\_head\_of\_state}), using the weight
-specified in the \texttt{weight} column of the attached metadata. While
+specified in the \texttt{wgt} column of the attached metadata. While
 computing the index from the original 14 variables, it remains unclear

### 2023-08-23 tidyindex.qmd
GGGI section revised

@@
 
-Frame 12 in @fig-idx-tour presents the original index values when equal weights are applied across the four dimensions.
-Leading the rankings, Nordic countries (Iceland, Norway, and Finland) and New Zealand show a good progression on reaching gender parity, with over 80% of the gender gap closed.
-In the middle range, an index value over 0.7 is achieved for Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia.
-The last group of countries (Pakistan, Iran, Algeria, Chad), with GGGI < 0.6, have a wider gender gap to closed and Afghanistan lags further behind with GGGI = 0.4.
+@fig-idx-tour illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries.
+Frame 12 shows the dot plot of the original index values sorted from highest to lowest.
+Leading the rankings are the Nordic countries (Iceland, Norway, and Finland) and New Zealand.
+The index values are between 0 and 1, and indicate proportional difference between men and women, with a value of 0.8 indicating women are 80% of the way to equality of these measures.
+There is a gap in values between these countries and the middle group (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, Slovakia), and another big drop to the next group (Pakistan, Iran, Algeria, Chad).
+Afghanistan lags much further behind.
 
-To observe the effect of a minor weight change on the index, we slightly vary the politics weight (while maintain equal weights among other dimensions) to generate an animation of the movement of index values in response to the changing weights.
-This visualization technique, which presents a sequence of data projections, is referred to as a "tour" and the specific kind of tour used here to move between nearby projections is known as a "local tour" [@buja2005computational; @wickham_tourr_2011].
+To make a simple illustration of sensitivity analysis, we slightly vary the weight for politics, between 0.07 and 0.52, while maintain equal weights among other dimensions.
+This can be viewed as an animation to examine change in relative index values as a response to the changing weights.
+This visualization technique, which presents a sequence of data projections, is referred to as a "tour" and the specific kind of tour used here to move between nearby projections is known as a "radial tour" (see @buja2005computational, @wickham_tourr_2011 and @radial-tour for more details].
 
-In @fig-idx-tour, six frames have been chosen from the local tour to explore the role of the politics dimension to the index.
+Frames 1 and 6 show linear combinations where politics contributes less than the original.
+It is interesting to note that the gap between the nordic countries and the middle countries dissipates, indicating that this component was one reason for the relatively higher GGGI values of these countries.
+Also interesting is the large drop in value for Bangladesh.
+
+Frames 18, 24, 29 show linear combinations where politics contributes more than the original.
+The most notable feature is that Bangladesh retains it's high index value whereas the other middle group countries decline, indicating that the politics score is a major component for Bangladesh's index value.
+
+<!--In @fig-idx-tour, six frames have been chosen from the local tour to explore the role of the politics dimension to the index.
 When the weight of politics is reduced (Frame 1 and 6 vs.
@@
 Frame 12), nearly all the countries experience a decrease in GGGI values.
-A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight, which leads to an almost similar index value to the Nordics when the politics dimension is assigned a weight of 0.52 in Frame 29.
+A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight, which leads to an almost similar index value to the Nordics when the politics dimension is assigned a weight of 0.52 in Frame 29.-->
 
 Ideally, an index should be robust against minor changes in its construction components.
-Unfortunately, this is not the case with GGGI, which raises concerns about the index's quality.
-For other indexes that use linear combination to reduce variable dimensions, the demonstrated local tour can be applied similarly to assess the index robustness through observing how index values respond to perturbation of the weights.
+This is not the case with GGGI, where small changes to one component lead to fairly large change in the index.
+The modular pipeline framework for computing the index makes it easy to conduct this type of sensitivity analysis, where one or more components are perturbed and the index re-calculated.
+One aspect of the GGGI not well-described in the Global Gender Gap Report is handling of missing values that are present in the initial variables for many countries, something that is common for this type of data.
+This could also be made more transparent with the dimension reduction module, by specifying an imputation method or providing warnings about missing values.
+
+<!--For other indexes that use linear combination to reduce variable dimensions, the demonstrated local tour can be applied similarly to assess the index robustness through observing how index values respond to perturbation of the weights.-->
 

### 2023-08-23 tidyindex.tex
GGGI section revised

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, sharp corners, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-Frame 12 in Figure~\ref{fig-idx-tour} presents the original index values
-when equal weights are applied across the four dimensions. Leading the
-rankings, Nordic countries (Iceland, Norway, and Finland) and New
-Zealand show a good progression on reaching gender parity, with over
-80\% of the gender gap closed. In the middle range, an index value over
-0.7 is achieved for Brazil, Panama, Poland, Bangladesh, Kazakhstan,
-Armenia, and Slovakia. The last group of countries (Pakistan, Iran,
-Algeria, Chad), with GGGI \textless{} 0.6, have a wider gender gap to
-closed and Afghanistan lags further behind with GGGI = 0.4.
-
-To observe the effect of a minor weight change on the index, we slightly
-vary the politics weight (while maintain equal weights among other
-dimensions) to generate an animation of the movement of index values in
-response to the changing weights. This visualization technique, which
-presents a sequence of data projections, is referred to as a ``tour''
-and the specific kind of tour used here to move between nearby
-projections is known as a ``local tour'' (Andreas Buja et al. 2005;
-Wickham et al. 2011).
-
-In Figure~\ref{fig-idx-tour}, six frames have been chosen from the local
-tour to explore the role of the politics dimension to the index. When
-the weight of politics is reduced (Frame 1 and 6 vs.~Frame 12), the
-difference in GGGI values between the Nordic and other countries
-narrows, suggesting a smaller variation among countries in achieving
-gender parity. Conversely, when the weight of politics increases (Frame
-18, 24, and 29 vs.~Frame 12), nearly all the countries experience a
-decrease in GGGI values. A noticeable exception to this trend is
-Bangladesh, where its index value moves in the same direction as the
-politics weight, which leads to an almost similar index value to the
-Nordics when the politics dimension is assigned a weight of 0.52 in
-Frame 29.
+Figure~\ref{fig-idx-tour} illustrates doing sensitivity analysis for
+GGGI, for a subset of 16 countries. Frame 12 shows the dot plot of the
+original index values sorted from highest to lowest. Leading the
+rankings are the Nordic countries (Iceland, Norway, and Finland) and New
+Zealand. The index values are between 0 and 1, and indicate proportional
+difference between men and women, with a value of 0.8 indicating women
+are 80\% of the way to equality of these measures. There is a gap in
+values between these countries and the middle group (Brazil, Panama,
+Poland, Bangladesh, Kazakhstan, Armenia, Slovakia), and another big drop
+to the next group (Pakistan, Iran, Algeria, Chad). Afghanistan lags much
+further behind.
+
+To make a simple illustration of sensitivity analysis, we slightly vary
+the weight for politics, between 0.07 and 0.52, while maintain equal
+weights among other dimensions. This can be viewed as an animation to
+examine change in relative index values as a response to the changing
+weights. This visualization technique, which presents a sequence of data
+projections, is referred to as a ``tour'' and the specific kind of tour
+used here to move between nearby projections is known as a ``radial
+tour'' (see Andreas Buja et al. (2005), Wickham et al. (2011) and Laa et
+al. (2023) for more details{]}.
+
+Frames 1 and 6 show linear combinations where politics contributes less
+than the original. It is interesting to note that the gap between the
+nordic countries and the middle countries dissipates, indicating that
+this component was one reason for the relatively higher GGGI values of
+these countries. Also interesting is the large drop in value for
+Bangladesh.
+
+Frames 18, 24, 29 show linear combinations where politics contributes
+more than the original. The most notable feature is that Bangladesh
+retains it's high index value whereas the other middle group countries
+decline, indicating that the politics score is a major component for
+Bangladesh's index value.
 
 Ideally, an index should be robust against minor changes in its
-construction components. Unfortunately, this is not the case with GGGI,
-which raises concerns about the index's quality. For other indexes that
-use linear combination to reduce variable dimensions, the demonstrated
-local tour can be applied similarly to assess the index robustness
-through observing how index values respond to perturbation of the
-weights.
+construction components. This is not the case with GGGI, where small
+changes to one component lead to fairly large change in the index. The
+modular pipeline framework for computing the index makes it easy to
+conduct this type of sensitivity analysis, where one or more components
+are perturbed and the index re-calculated. One aspect of the GGGI not
+well-described in the Global Gender Gap Report is handling of missing
+values that are present in the initial variables for many countries,
+something that is common for this type of data. This could also be made
+more transparent with the dimension reduction module, by specifying an
+imputation method or providing warnings about missing values.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-radial-tour}{}}%
+Laa, Ursula, Alex Aumann, Dianne Cook, and German Valencia. 2023. {``New
+and Simplified Manual Controls for Projection and Slice Tours, with
+Application to Exploring Classification Boundaries in High
+Dimensions.''} \emph{Journal of Computational and Graphical Statistics}
+0 (0): 1--8. \url{https://doi.org/10.1080/10618600.2023.2206459}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-laimighofer_how_2022}{}}%

### 2023-08-23 tidyindex.qmd
caption for fig 7

@@
 ```{r fig-idx-tour, fig.align='center', fig.height=5} 
-#| fig-cap: "Six frames selected to explore how varying the weights of the politics dimension changes the index values and country rankings in Global Gender Gap Index (GGGI).
-The top panel shows the GGGI value against the country, ranked by its original index value in Frame 12.
-The bottom panel displays the weight used to produce the index values in the top panel, with each frame corresponds to a set of weights.
-Countries selected includes 1) top-ranked countries with GGGI > 0.85, 2) countries ranked between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked countries with GGGI < 0.6.
-Compared to Frame 12 where equal weigts are used for the four dimension, a reduced weight in politics (Frame 1 and 6) shows narrower gaps between top and mid- or lower- ranked countries, while an increase in the politics weight (Frame 18, 24, and 29) leads to a systematic decrease of GGGI values across all the countries, except for Bangladesh.
-Full animation is available at https://vimeo.com/847874016."
+#| fig-cap: "Exploring the sensitivity of the GGGI, by varying the politics component's contribution, for a subset of countries.
+Each panel shows a dotplot of the index values, computed for the linear combination represented by the segment plots below.
+Frame 12 shows the actual GGGI values, and countries are sorted from highest to lowest on this.
+Frames 1 and 6 show the GGGI if the politics component is reduced.
+Frames 18, 24, 29 show the GGGI when the politics component is increased.
+The most notable feature is that Bangladesh's GGGI drops substantially when politics is removed, indicating that this component plays a large role in it's relatively high value.
+Also, politics plays a substantial role in the GGGI's for the top ranked countries, because each of them drops, to the state of being similar to the middle ranked countries when the politics component's contribution is reduced.
+The animation can be viewed at https://vimeo.com/847874016."
 #frames <- c("001", "013", "037", "061", "085", "109")
@@
 
-Errors in measurement, variability and sampling error, may arise at various stages of the pipeline calculation, including from different parameterisation choices, as illustrated from @sec-example1, or from the statistical summarization procedures applied in the pipeline.
-Although it may not be possible to perfectly measure these errors, it is important that they are recognised and estimated for an index, especially to provide potential confidence intervals.
+Errors in measurement, variability and sampling error, may arise at various stages of the pipeline calculation, including from different parameterization choices, as illustrated from @sec-example1, or from the statistical summarization procedures applied in the pipeline.
+Although it may not be possible to perfectly measure these errors, it is important that they are recognised and estimated for an index, so that it is possible to compute confidence intervals.
 In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate one possibility to compute a confidence interval for the Standardized Precipitation Index (SPI).

### 2023-08-23 tidyindex.tex
caption for fig 7

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\caption{\label{fig-idx-tour}Six frames selected to explore how varying
-the weights of the politics dimension changes the index values and
-country rankings in Global Gender Gap Index (GGGI). The top panel shows
-the GGGI value against the country, ranked by its original index value
-in Frame 12. The bottom panel displays the weight used to produce the
-index values in the top panel, with each frame corresponds to a set of
-weights. Countries selected includes 1) top-ranked countries with GGGI
-\textgreater{} 0.85, 2) countries ranked between 57 and 62 with GGGI
-from 0.72 to 0.73, and 3) low-ranked countries with GGGI \textless{}
-0.6. Compared to Frame 12 where equal weigts are used for the four
-dimension, a reduced weight in politics (Frame 1 and 6) shows narrower
-gaps between top and mid- or lower- ranked countries, while an increase
-in the politics weight (Frame 18, 24, and 29) leads to a systematic
-decrease of GGGI values across all the countries, except for Bangladesh.
-Full animation is available at https://vimeo.com/847874016.}
+\caption{\label{fig-idx-tour}Exploring the sensitivity of the GGGI, by
+varying the politics component's contribution, for a subset of
+countries. Each panel shows a dotplot of the index values, computed for
+the linear combination represented by the segment plots below. Frame 12
+shows the actual GGGI values, and countries are sorted from highest to
+lowest on this. Frames 1 and 6 show the GGGI if the politics component
+is reduced. Frames 18, 24, 29 show the GGGI when the politics component
+is increased. The most notable feature is that Bangladesh's GGGI drops
+substantially when politics is removed, indicating that this component
+plays a large role in it's relatively high value. Also, politics plays a
+substantial role in the GGGI's for the top ranked countries, because
+each of them drops, to the state of being similar to the middle ranked
+countries when the politics component's contribution is reduced. The
+animation can be viewed at https://vimeo.com/847874016.}
 
@@
 various stages of the pipeline calculation, including from different
-parameterisation choices, as illustrated from
+parameterization choices, as illustrated from
 Section~\ref{sec-example1}, or from the statistical summarization
@@
 perfectly measure these errors, it is important that they are recognised
-and estimated for an index, especially to provide potential confidence
+and estimated for an index, so that it is possible to compute confidence
 intervals. In this example, the Texas post office station highlighted in

### 2023-08-23 tidyindex.qmd
remake the gggi plot

@@
   filter(between(index, 0.72, 0.729)  | index < 0.6 | index >= 0.85) %>%
-  arrange(index) 
+  arrange(index) %>% 
+  mutate(country = ifelse(country == "Iran (Islamic Republic of)", "Iran", country))
 colnames(dt) <- c(colnames(dt)[1:4], "economy", "education", "health", "politics")
@@
   dt[,5:8], planned2_tour(b), dev = "png",
-  display_idx(half_range = 2,cex = 4, label_cex = 3.5, col = "black", 
-              panel_height_ratio = c(4,1),
-              axis_bar_lwd = c(rep(3, 3), 7),
+  display_idx(half_range = 2,cex = 6, label_cex = 4, col = "black", 
+              panel_height_ratio = c(4,1), label_x_pos = 0.1, 
+              axis_bar_lwd = c(rep(7, 3), 9), label_col = "grey70",
               axis_bar_col = c(rep("#000000", 3), "red"),
-              axis_bar_label_cex = c(rep(2.5, 3), 3),
+              axis_bar_label_cex = 5,
               axis_bar_label_col = c(rep("#000000", 3), "red"), 
-              axis_label_cex = 3, frame_x_pos = 0.16,  frame_cex = 4,
-              axis_var_cex = 3.6, 
+              axis_label_cex_upper = 3, axis_label_cex_lower = 3,
+              axis_var_cex = 5, 
               label = dt$country, abb_vars = FALSE),
-  width = 900, height = 1400, apf = 1/20, frames = 120,
+  width = 1200, height = 2000, apf = 1/20, frames = 120,
   here::here("figures/idx-tour/idx-tour-%03d.png")
 )
+
+library(magick)
+path <- list.files(here::here("figures/idx-tour"), full.names = TRUE)
+frame_num <- str_extract(path, "-[0-9]+") %>% as.numeric() %>% -.
+walk2(path, frame_num, ~{
+  img <- image_read(.x)
+  img <- image_annotate(img, glue::glue("Frame {.y}"), size = 60, location = "+920+10")
+  image_write(img, path = .x)
+})
+
+# img <- image_read(here::here("figures/idx-tour/idx-tour-015.png"))
+# img <- image_annotate(img, "Frame 10", size = 40, location = "+712+10")
+# image_write(img, path = here::here("figures/idx-tour/idx-tour-014.png"))
 
@@
   gif_file = here::here("figures/idx-tour.gif"),
-  delay = 0.15, width = 900, height = 1400)
+  delay = 0.15, width = 1200, height = 2000)
 
@@
 
-```{r fig-idx-tour, fig.align='center', fig.height=5} 
+```{r fig-idx-tour, fig.align='center', fig.height=6, fig.width=6}   
 #| fig-cap: "Exploring the sensitivity of the GGGI, by varying the politics component's contribution, for a subset of countries.
@@
 The animation can be viewed at https://vimeo.com/847874016."
-#frames <- c("001", "013", "037", "061", "085", "109")
 frames <- c("001", "006", "012", "018", "024", "029") 
@@
 In SPI, the distribution fit step fits the gamma distribution to the aggregated precipitation series separately for each month.
-This results in  31 or 32 points, from January 1990 to April 2022, for estimating each set of distribution parameters.
+This results in  32 or 33 points, from January 1990 to April 2022, for estimating each set of distribution parameters.
 To account for this sampling uncertainty with these samples, bootstrapping is used to generate replicates of the aggregated series.

### 2023-08-23 tidyindex.tex
remake the gggi plot

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, enhanced, breakable, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 \(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\) \\
+\bottomrule()
 \end{longtable}
@@
 aggregated precipitation series separately for each month. This results
-in 31 or 32 points, from January 1990 to April 2022, for estimating each
+in 32 or 33 points, from January 1990 to April 2022, for estimating each
 set of distribution parameters. To account for this sampling uncertainty

### 2023-08-23 tidyindex.qmd
various small tweaks

@@
 
-Adoption of this pipeline approach would provide uniformity to the field of index development, research, and application to improve comparability, reporducibility, and communication
+Adoption of this pipeline approach would provide uniformity to the field of index development, research, and application to improve comparability, reproducibility, and communication.
 
@@
 
-Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) [@hotelling1933analysis] or linear discriminant analysis [@fisher1936use], preparing contrasts to test particular elements in analysis of variance (XXX-ref), or hand-crafted by a content-area expert.
+Linear combinations of variables are commonly used to reduce the dimension in statistical methodology, and chosen using a method like principal component analysis (PCA) [@hotelling1933analysis] or linear discriminant analysis [@fisher1936use], preparing contrasts to test particular elements in analysis of variance [@fisher1970statistical], or hand-crafted by a content-area expert.
 Linear combinations also form the basis for visualizing multivariate data, in methods such as tours [@wickham_tourr_2011].
@@
 
-(more changes) The accompanied package, tidyindex, is not intended to offer an exhaustive implementation for all indexes across every domains.
+The accompanied package, tidyindex, is not intended to offer an exhaustive implementation for all indexes across every domains.
 Instead, it provides a realization of the pipeline framework proposed in the paper.
@@
 The output contains the original data, index values (`.index`), parameters used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`pet`, `.agg`, and `.fitted`).
-This data can be visualized to investigate the spatio-temporal distribution of the drought/flood events, as well as the response of index values to different time scales and distribution parameters at specific single locations.
+This data can be visualized to investigate the spatio-temporal distribution of the drought or flood events, as well as the response of index values to different time scales and distribution parameters at specific single locations.
 @fig-compute-spatial and @fig-compute-temporal exemplify two possibilities.
@@
 The index values are between 0 and 1, and indicate proportional difference between men and women, with a value of 0.8 indicating women are 80% of the way to equality of these measures.
-There is a gap in values between these countries and the middle group (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, Slovakia), and another big drop to the next group (Pakistan, Iran, Algeria, Chad).
+There is a gap in values between these countries and the middle group (Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and another big drop to the next group (Pakistan, Iran, Algeria, and Chad).
 Afghanistan lags much further behind.
-
-To make a simple illustration of sensitivity analysis, we slightly vary the weight for politics, between 0.07 and 0.52, while maintain equal weights among other dimensions.
-This can be viewed as an animation to examine change in relative index values as a response to the changing weights.
-This visualization technique, which presents a sequence of data projections, is referred to as a "tour" and the specific kind of tour used here to move between nearby projections is known as a "radial tour" (see @buja2005computational, @wickham_tourr_2011 and @radial-tour for more details].
-
-Frames 1 and 6 show linear combinations where politics contributes less than the original.
-It is interesting to note that the gap between the nordic countries and the middle countries dissipates, indicating that this component was one reason for the relatively higher GGGI values of these countries.
-Also interesting is the large drop in value for Bangladesh.
-
-Frames 18, 24, 29 show linear combinations where politics contributes more than the original.
-The most notable feature is that Bangladesh retains it's high index value whereas the other middle group countries decline, indicating that the politics score is a major component for Bangladesh's index value.
-
-<!--In @fig-idx-tour, six frames have been chosen from the local tour to explore the role of the politics dimension to the index.
-When the weight of politics is reduced (Frame 1 and 6 vs.
-Frame 12), the difference in GGGI values between the Nordic and other countries narrows, suggesting a smaller variation among countries in achieving gender parity.
-Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs.
-Frame 12), nearly all the countries experience a decrease in GGGI values.
-A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight, which leads to an almost similar index value to the Nordics when the politics dimension is assigned a weight of 0.52 in Frame 29.-->
-
-Ideally, an index should be robust against minor changes in its construction components.
-This is not the case with GGGI, where small changes to one component lead to fairly large change in the index.
-The modular pipeline framework for computing the index makes it easy to conduct this type of sensitivity analysis, where one or more components are perturbed and the index re-calculated.
-One aspect of the GGGI not well-described in the Global Gender Gap Report is handling of missing values that are present in the initial variables for many countries, something that is common for this type of data.
-This could also be made more transparent with the dimension reduction module, by specifying an imputation method or providing warnings about missing values.
-
-<!--For other indexes that use linear combination to reduce variable dimensions, the demonstrated local tour can be applied similarly to assess the index robustness through observing how index values respond to perturbation of the weights.-->
 
@@
 
+To make a simple illustration of sensitivity analysis, we slightly vary the weight for politics, between 0.07 and 0.52, while maintain equal weights among other dimensions.
+This can be viewed as an animation to examine change in relative index values as a response to the changing weights.
+This visualization technique, which presents a sequence of data projections, is referred to as a "tour" and the specific kind of tour used here to move between nearby projections is known as a "radial tour" (see @buja2005computational, @wickham_tourr_2011 and @RJ-2020-027 for more details).
+
+Frames 1 and 6 show linear combinations where politics contributes less than the original.
+It is interesting to note that the gap between the Nordic countries and the middle countries dissipates, indicating that this component was one reason for the relatively higher GGGI values of these countries.
+Also interesting is the large drop in value for Bangladesh.
+
+Frames 18, 24, 29 show linear combinations where politics contributes more than the original.
+The most notable feature is that Bangladesh retains it's high index value whereas the other middle group countries decline, indicating that the politics score is a major component for Bangladesh's index value.
+
+<!--In @fig-idx-tour, six frames have been chosen from the local tour to explore the role of the politics dimension to the index.
+When the weight of politics is reduced (Frame 1 and 6 vs.
+Frame 12), the difference in GGGI values between the Nordic and other countries narrows, suggesting a smaller variation among countries in achieving gender parity.
+Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs.
+Frame 12), nearly all the countries experience a decrease in GGGI values.
+A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight, which leads to an almost similar index value to the Nordics when the politics dimension is assigned a weight of 0.52 in Frame 29.-->
+
+Ideally, an index should be robust against minor changes in its construction components.
+This is not the case with GGGI, where small changes to one component lead to fairly large change in the index.
+The modular pipeline framework for computing the index makes it easy to conduct this type of sensitivity analysis, where one or more components are perturbed and the index re-calculated.
+One aspect of the GGGI not well-described in the Global Gender Gap Report is handling of missing values that are present in the initial variables for many countries, something that is common for this type of data.
+This could also be made more transparent with the dimension reduction module, by specifying an imputation method or providing warnings about missing values.
+
 ## Decoding uncertainty through the wisdom of the crowd
@@
 #| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI-24) for the Texas post office station, in Queensland, Australia.
-A bootstrap sample of 100 is taken from the aggregated pricipitation series to estimate gamma parameters and to calculate the index.
+A bootstrap sample of 100 is taken from the aggregated precipitation series to estimate gamma parameters and to calculate the index.
 The dashed line at SPI = -2 represents an extreme drought as defined by the SPI.
@@
 The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
-The source code for reproducing this paper can be found at: https://github.com/huizezhang-sherry/paper-tidyindex.
+The source code for reproducing this paper can be found at: <https://github.com/huizezhang-sherry/paper-tidyindex>.
 

### 2023-08-23 tidyindex.tex
various small tweaks

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, enhanced, breakable, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, frame hidden, breakable, boxrule=0pt, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 of index development, research, and application to improve
-comparability, reporducibility, and communication
+comparability, reproducibility, and communication.
 
@@
 principal component analysis (PCA) (Hotelling 1933) or linear
-discriminant analysis (Fisher 1936), preparing contrasts to test
-particular elements in analysis of variance (XXX-ref), or hand-crafted
-by a content-area expert. Linear combinations also form the basis for
-visualizing multivariate data, in methods such as tours (Wickham et al.
-2011). This dimension reduction method can accommodate linear
-combinations as provided by any method, and hence is linear by design.
-The transformation module provides variable-wise non-linear
-transformation.
+discriminant analysis (Ronald A. Fisher 1936), preparing contrasts to
+test particular elements in analysis of variance (Ronald Aylmer Fisher
+1970), or hand-crafted by a content-area expert. Linear combinations
+also form the basis for visualizing multivariate data, in methods such
+as tours (Wickham et al. 2011). This dimension reduction method can
+accommodate linear combinations as provided by any method, and hence is
+linear by design. The transformation module provides variable-wise
+non-linear transformation.
 
@@
 
-(more changes) The accompanied package, tidyindex, is not intended to
-offer an exhaustive implementation for all indexes across every domains.
-Instead, it provides a realization of the pipeline framework proposed in
-the paper. When adopting the pipeline approach to construct indexes,
+The accompanied package, tidyindex, is not intended to offer an
+exhaustive implementation for all indexes across every domains. Instead,
+it provides a realization of the pipeline framework proposed in the
+paper. When adopting the pipeline approach to construct indexes,
 analysts may consider developing software that can be readily deployed
@@
 \texttt{.fitted}). This data can be visualized to investigate the
-spatio-temporal distribution of the drought/flood events, as well as the
-response of index values to different time scales and distribution
+spatio-temporal distribution of the drought or flood events, as well as
+the response of index values to different time scales and distribution
 parameters at specific single locations.
@@
 values between these countries and the middle group (Brazil, Panama,
-Poland, Bangladesh, Kazakhstan, Armenia, Slovakia), and another big drop
-to the next group (Pakistan, Iran, Algeria, Chad). Afghanistan lags much
-further behind.
-
-To make a simple illustration of sensitivity analysis, we slightly vary
-the weight for politics, between 0.07 and 0.52, while maintain equal
-weights among other dimensions. This can be viewed as an animation to
-examine change in relative index values as a response to the changing
-weights. This visualization technique, which presents a sequence of data
-projections, is referred to as a ``tour'' and the specific kind of tour
-used here to move between nearby projections is known as a ``radial
-tour'' (see Andreas Buja et al. (2005), Wickham et al. (2011) and Laa et
-al. (2023) for more details{]}.
-
-Frames 1 and 6 show linear combinations where politics contributes less
-than the original. It is interesting to note that the gap between the
-nordic countries and the middle countries dissipates, indicating that
-this component was one reason for the relatively higher GGGI values of
-these countries. Also interesting is the large drop in value for
-Bangladesh.
-
-Frames 18, 24, 29 show linear combinations where politics contributes
-more than the original. The most notable feature is that Bangladesh
-retains it's high index value whereas the other middle group countries
-decline, indicating that the politics score is a major component for
-Bangladesh's index value.
-
-Ideally, an index should be robust against minor changes in its
-construction components. This is not the case with GGGI, where small
-changes to one component lead to fairly large change in the index. The
-modular pipeline framework for computing the index makes it easy to
-conduct this type of sensitivity analysis, where one or more components
-are perturbed and the index re-calculated. One aspect of the GGGI not
-well-described in the Global Gender Gap Report is handling of missing
-values that are present in the initial variables for many countries,
-something that is common for this type of data. This could also be made
-more transparent with the dimension reduction module, by specifying an
-imputation method or providing warnings about missing values.
+Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and another big
+drop to the next group (Pakistan, Iran, Algeria, and Chad). Afghanistan
+lags much further behind.
 
@@
 
+To make a simple illustration of sensitivity analysis, we slightly vary
+the weight for politics, between 0.07 and 0.52, while maintain equal
+weights among other dimensions. This can be viewed as an animation to
+examine change in relative index values as a response to the changing
+weights. This visualization technique, which presents a sequence of data
+projections, is referred to as a ``tour'' and the specific kind of tour
+used here to move between nearby projections is known as a ``radial
+tour'' (see Andreas Buja et al. (2005), Wickham et al. (2011) and
+Spyrison and Cook (2020) for more details).
+
+Frames 1 and 6 show linear combinations where politics contributes less
+than the original. It is interesting to note that the gap between the
+Nordic countries and the middle countries dissipates, indicating that
+this component was one reason for the relatively higher GGGI values of
+these countries. Also interesting is the large drop in value for
+Bangladesh.
+
+Frames 18, 24, 29 show linear combinations where politics contributes
+more than the original. The most notable feature is that Bangladesh
+retains it's high index value whereas the other middle group countries
+decline, indicating that the politics score is a major component for
+Bangladesh's index value.
+
+Ideally, an index should be robust against minor changes in its
+construction components. This is not the case with GGGI, where small
+changes to one component lead to fairly large change in the index. The
+modular pipeline framework for computing the index makes it easy to
+conduct this type of sensitivity analysis, where one or more components
+are perturbed and the index re-calculated. One aspect of the GGGI not
+well-described in the Global Gender Gap Report is handling of missing
+values that are present in the initial variables for many countries,
+something that is common for this type of data. This could also be made
+more transparent with the dimension reduction module, by specifying an
+imputation method or providing warnings about missing values.
+
 \hypertarget{decoding-uncertainty-through-the-wisdom-of-the-crowd}{%
@@
 station, in Queensland, Australia. A bootstrap sample of 100 is taken
-from the aggregated pricipitation series to estimate gamma parameters
+from the aggregated precipitation series to estimate gamma parameters
 and to calculate the index. The dashed line at SPI = -2 represents an
@@
 for reproducing this paper can be found at:
-https://github.com/huizezhang-sherry/paper-tidyindex.
+\url{https://github.com/huizezhang-sherry/paper-tidyindex}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-fisher1970statistical}{}}%
+Fisher, Ronald Aylmer. 1970. {``Statistical Methods for Research
+Workers.''} In \emph{Breakthroughs in Statistics: Methodology and
+Distribution}, 66--70. Springer.
+\url{https://doi.org/10.1007/978-1-4612-4380-9_6}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-fundiversity}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-radial-tour}{}}%
-Laa, Ursula, Alex Aumann, Dianne Cook, and German Valencia. 2023. {``New
-and Simplified Manual Controls for Projection and Slice Tours, with
-Application to Exploring Classification Boundaries in High
-Dimensions.''} \emph{Journal of Computational and Graphical Statistics}
-0 (0): 1--8. \url{https://doi.org/10.1080/10618600.2023.2206459}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-laimighofer_how_2022}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-RJ-2020-027}{}}%
+Spyrison, Nicholas, and Dianne Cook. 2020. {``The r Journal: Spinifex:
+An r Package for Creating a Manual Tour of Low-Dimensional Projections
+of Multivariate Data.''} \emph{The R Journal} 12: 243--57.
+\url{https://doi.org/10.32614/RJ-2020-027}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-sutherland_orca_2000}{}}%

### 2023-08-23 tidyindex.qmd
table 1 fixed

@@
 
+```{r fig-pipeline-steps}
+#| fig-align: center
+#| fig-cap: "Diagram of pipeline modules for index construction. The highlighted path illustrates one possible construction using the dimension reduction and simplification module."
+#| out-height: 90%
+#| out-width: 100%
+knitr::include_graphics(here::here("figures/pipeline-overall.png"))
+```
+
 Now, we introduce the notation used for describing pipeline modules. Consider a multivariate spatio-temporal process,
@@
 
-| Section | Module                  | Input                                                                   | Operation                           | Output                                                                        |
-|-------|---------------|---------|---------------|--------------------------|
-| 3.1   | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
-| 3.2   | Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
-| 3.3   | Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
-| 3.4   | Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
-| 3.5   | Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p$      |
-| 3.6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $P_j(s;t) \quad P(.) \in [0, 1]$                   |
-| 3.7   | Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
-| 3.8   | Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
-| 3.9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$                                      |
+| Module                  | Input                                                                   | Operation                           | Output                                                                        |
+|---------------|---------|---------------|--------------------------|
+| Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
+| Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
+| Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
+| Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
+| Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p$      |
+| Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $P_j(s;t) \quad P(.) \in [0, 1]$                   |
+| Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
+| Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
+| Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$                                      |
 
 : Summary of the notation for input, operation, and output of each pipeline module. {#tbl-notation}
-
-```{r fig-pipeline-steps}
-#| fig-align: center
-#| fig-cap: "Diagram of pipeline modules for index construction. The highlighted path illustrates one possible construction using the dimension reduction and simplification module."
-#| out-height: 90%
-#| out-width: 100%
-knitr::include_graphics(here::here("figures/pipeline-overall.png"))
-```
 

### 2023-08-23 tidyindex.tex
table 1 fixed

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, frame hidden, breakable, boxrule=0pt, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-overall.png}
+
+}
+
+\caption{\label{fig-pipeline-steps}Diagram of pipeline modules for index
+construction. The highlighted path illustrates one possible construction
+using the dimension reduction and simplification module.}
+
+\end{figure}
+
 Now, we introduce the notation used for describing pipeline modules.
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0972}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2083}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3611}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2308}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1385}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2308}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4000}}@{}}
 \caption{\label{tbl-notation}Summary of the notation for input,
 operation, and output of each pipeline module.}\tabularnewline
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
-Section
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
 Module
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endfirsthead
-\toprule()
+\toprule\noalign{}
 \begin{minipage}[b]{\linewidth}\raggedright
-Section
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
 Module
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
-3.1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
+\bottomrule\noalign{}
+\endlastfoot
+Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
 \(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
-3.2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
+Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
 \(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
-3.3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
+Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
 \(x^{\text{Trans}}_j(s;t)\) \\
-3.4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
+Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
 \(x^{\text{Scale}}_j(s;t)\) \\
-3.5 & Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\)
-& \(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
-3.6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
+Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\) &
+\(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
+Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
 \(P_j(s;t) \quad P(.) \in [0, 1]\) \\
-3.7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
+Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
 \(z_j(s; t)\) \\
-3.8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
-3.9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
+Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
+Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
 \(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\) \\
-\bottomrule()
 \end{longtable}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-overall.png}
-
-}
-
-\caption{\label{fig-pipeline-steps}Diagram of pipeline modules for index
-construction. The highlighted path illustrates one possible construction
-using the dimension reduction and simplification module.}
-
-\end{figure}
 

### 2023-08-23 tidyindex.tex
scaling v transformation explained

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 
@@
 
+Although the scaling might be considered to be a transformation, we have
+elected to make it a separate module because it is neater. Scaling
+simply changes the numbers in the data not the shape of a variable.
+Transformation will most likely change the shape, and is usually
+non-linear.
+
 \begin{figure}

### 2023-08-23 tidyindex.qmd
scaling v transformation explained
finished with the additional justifications for separate modules

@@
 
+Although the scaling might be considered to be a transformation, we have elected to make it a separate module because it is neater.
+Scaling simply changes the numbers in the data not the shape of a variable.
+Transformation will most likely change the shape, and is usually non-linear.
+
 ```{r fig-scale-var-trans-compare}
@@
 Different fitting methods and different goodness of fit tests may be used to compare the distribution choice on the index value.
+This could be considered to be a variable transformation because it is usually conducted separately for each variable.
+However, very occasionally a fit is conducted on two or more variables simulataneously.
+For this reaosn, and because it usually is applied later in the pipeline it is neater to make this a separate module.
 
@@
 Benchmark sets a value $b_j(s,t)$ for comparing against the original variable $x_j(s;t)$.
-This benchmark can be a fixed value consistently across space and time or determined by the data through the function $u[x_j(s;t)]$.
+This benchmark can be a fixed value consistently across space and time, perhaps extracted from expert knowledge, or determined by the data through the function $u[x_j(s;t)]$.
 Once a benchmark is set, observations can be highlighted for adjustments in other modules or can serve as targets for monitoring and planning.

### 2023-08-23 tidyindex.tex
finished with the additional justifications for separate modules

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, breakable]}{\end{tcolorbox}}\fi
 
@@
 fitting methods and different goodness of fit tests may be used to
-compare the distribution choice on the index value.
+compare the distribution choice on the index value. This could be
+considered to be a variable transformation because it is usually
+conducted separately for each variable. However, very occasionally a fit
+is conducted on two or more variables simulataneously. For this reaosn,
+and because it usually is applied later in the pipeline it is neater to
+make this a separate module.
 
@@
 variable \(x_j(s;t)\). This benchmark can be a fixed value consistently
-across space and time or determined by the data through the function
-\(u[x_j(s;t)]\). Once a benchmark is set, observations can be
-highlighted for adjustments in other modules or can serve as targets for
-monitoring and planning.
+across space and time, perhaps extracted from expert knowledge, or
+determined by the data through the function \(u[x_j(s;t)]\). Once a
+benchmark is set, observations can be highlighted for adjustments in
+other modules or can serve as targets for monitoring and planning.
 

### 2023-08-23 tidyindex.qmd
typo

@@
 | Module                  | Input                                                                   | Operation                           | Output                                                                        |
-|---------------|---------|---------------|--------------------------|
+|--------------|---------|---------------|--------------------------|
 | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
@@
 This could be considered to be a variable transformation because it is usually conducted separately for each variable.
-However, very occasionally a fit is conducted on two or more variables simulataneously.
-For this reaosn, and because it usually is applied later in the pipeline it is neater to make this a separate module.
+However, very occasionally a fit is conducted on two or more variables simultaneously.
+For this reason, and because it usually is applied later in the pipeline it is neater to make this a separate module.
 
@@
 Benchmark sets a value $b_j(s,t)$ for comparing against the original variable $x_j(s;t)$.
-This benchmark can be a fixed value consistently across space and time, perhaps extracted from expert knowledge, or determined by the data through the function $u[x_j(s;t)]$.
+This benchmark can be a fixed value consistently across space and time, perhaps extracted from expert knowledge, perhaps extracted from expert knowledge, or determined by the data through the function $u[x_j(s;t)]$.
 Once a benchmark is set, observations can be highlighted for adjustments in other modules or can serve as targets for monitoring and planning.

### 2023-08-23 tidyindex.tex
typo

--- 2023-08-23 tidyindex.tex
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2308}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1385}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2308}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4000}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2188}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1406}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2344}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4062}}@{}}
 \caption{\label{tbl-notation}Summary of the notation for input,
 operation, and output of each pipeline module.}\tabularnewline
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
@@
 fitting methods and different goodness of fit tests may be used to
-compare the distribution choice on the index value.
+compare the distribution choice on the index value. This could be
+considered to be a variable transformation because it is usually
+conducted separately for each variable. However, very occasionally a fit
+is conducted on two or more variables simultaneously. For this reason,
+and because it usually is applied later in the pipeline it is neater to
+make this a separate module.
 
@@
 variable \(x_j(s;t)\). This benchmark can be a fixed value consistently
-across space and time or determined by the data through the function
-\(u[x_j(s;t)]\). Once a benchmark is set, observations can be
+across space and time, perhaps extracted from expert knowledge, perhaps
+extracted from expert knowledge, or determined by the data through the
+function \(u[x_j(s;t)]\). Once a benchmark is set, observations can be
 highlighted for adjustments in other modules or can serve as targets for

### 2023-11-03 tidyindex.qmd
update as per the thesis

@@
   DATA |>
-    aggregate(...) |>
-    dist_fit(...) |> 
-    augment(...)
+    temporal_aggregate(...) |>
+    distribution_fit(...)|>
+    normalise(...)
 }
@@
 idx <- queensland %>%
-  init(id = id, time = ym) %>%
+  mutate(month = lubridate::month(ym)) |>
+  init(id = id, time = ym, group = month) |> 
   compute_indexes(
     spei = idx_spei(
-      .pet_method = "thornthwaite", .tavg = tavg, .lat = lat, 
-      .scale = .scale, .dist = c(gev(), loglogistic())),
+      .tavg = tavg, .lat = lat, 
+      .scale = .scale, .dist = list(dist_gev(), dist_glo())),
     spi = idx_spi(.scale = .scale)
   )
+
 ```
@@
     data = idx %>%
-      filter(.idx == "spi", .scale == 12) %>%
+      filter(.scale == 12) %>%
       filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
                ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )),
-    aes(x = long, y = lat, color = .index)) +
+    aes(x = long, y = lat, color = .value)) +
   colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
@@
   filter(name == "TEXAS POST OFFICE", .idx == "spei") %>%
-  mutate(.dist = ifelse(.dist == "pargev", "GEV", "Log Logistic")) %>%
-  rename(scale = .scale) %>% 
-  ggplot(aes(x = ym, y = .index, color = .dist, group = .dist)) +
+  ggplot(aes(x = ym, y = .value, color = .dist, group = .dist)) +
   geom_line() +
   theme_benchmark() +
-  facet_wrap(vars(scale), labeller = label_both, ncol = 1) +
+  facet_wrap(vars(.scale), labeller = label_both, ncol = 1) +
   scale_x_yearmonth(breaks = "4 year", date_labels = "%Y") +
@@
 The bootstrap procedure samples the aggregated precipitation (`.agg`) for 100 iterations (`.n_boot = 100`) and then fits the gamma distribution.
-The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `augment()`.
+The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `normalise()`.
 
 ```{r echo = TRUE, eval = FALSE}
-DATA %>% 
-  aggregate(.var = prcp, .scale = 24) %>% 
-  dist_fit(.dist = gamma(), .var = .agg, .n_boot = 100) %>% 
-  augment(.var = .agg)
+DATA |> 
+  temporal_aggregate(.agg = temporal_rolling_window(prcp, scale = 24)) |> 
+  distribution_fit(.fit = dist_gamma(var = ".agg", method = "lmoms",
+                                     .n_boot = 100)) |>
+  normalise(.index = norm_quantile(.fit))
 ```
@@
   filter(name == "TEXAS POST OFFICE") %>% 
-  init(id = id, time = ym) %>% 
-  aggregate(.var = prcp, .scale = 24) %>% 
-  dist_fit(.dist = gamma(), .var = .agg, .n_boot = 100) %>% 
-  augment(.var = .agg)
+  mutate(month = lubridate::month(ym)) |>
+  init(id = id, time = ym, group = month) |> 
+  temporal_aggregate(.agg = temporal_rolling_window(prcp, scale = 24)) |> 
+  distribution_fit(.fit = dist_gamma(var = ".agg", method = "lmoms", .n_boot = 100)) |>
+  tidyindex::normalise(.index = norm_quantile(.fit))
 
@@
     q10 = quantile(.index, 0.1),
+    q50 = quantile(.index, 0.5),
     q90 = quantile(.index, 0.9),
@@
 
-all <- idx %>% 
-  filter(name == "TEXAS POST OFFICE", .idx == "spi", .scale == 24) %>% 
-  select(ym, .index) %>% 
-  left_join(res_band, by = "ym")
-
-loc <- make_yearmonth(2022, 12)
-
 # the sampling uncertainty is quite large here
@@
 # each month fit has only 30 points
-all %>% 
+res_band %>% 
   ggplot(aes(x = ym)) + 
@@
   geom_ribbon(aes(ymin = q10, ymax = q90), fill = "steelblue4", alpha = 0.5) + 
-  geom_line(aes(y = .index)) + 
+  geom_line(aes(y = q50)) + 
   scale_x_yearmonth(name = "Year", date_break = "2 years", date_label = "%Y") +

### 2023-11-03 tidyindex.tex
update as per the thesis

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, enhanced, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 \end{minipage} \\
-\midrule()
+\midrule\noalign{}
 \endhead
+\bottomrule\noalign{}
+\endlastfoot
 Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
@@
 \NormalTok{  DATA }\SpecialCharTok{|\textgreater{}}
-    \FunctionTok{aggregate}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
-    \FunctionTok{dist\_fit}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} 
-    \FunctionTok{augment}\NormalTok{(...)}
+    \FunctionTok{temporal\_aggregate}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
+    \FunctionTok{distribution\_fit}\NormalTok{(...)}\SpecialCharTok{|\textgreater{}}
+    \FunctionTok{normalise}\NormalTok{(...)}
 \NormalTok{\}}
@@
 \NormalTok{idx }\OtherTok{\textless{}{-}}\NormalTok{ queensland }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ id, }\AttributeTok{time =}\NormalTok{ ym) }\SpecialCharTok{\%\textgreater{}\%}
+  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{month =}\NormalTok{ lubridate}\SpecialCharTok{::}\FunctionTok{month}\NormalTok{(ym)) }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ id, }\AttributeTok{time =}\NormalTok{ ym, }\AttributeTok{group =}\NormalTok{ month) }\SpecialCharTok{|\textgreater{}} 
   \FunctionTok{compute\_indexes}\NormalTok{(}
     \AttributeTok{spei =} \FunctionTok{idx\_spei}\NormalTok{(}
-      \AttributeTok{.pet\_method =} \StringTok{"thornthwaite"}\NormalTok{, }\AttributeTok{.tavg =}\NormalTok{ tavg, }\AttributeTok{.lat =}\NormalTok{ lat, }
-      \AttributeTok{.scale =}\NormalTok{ .scale, }\AttributeTok{.dist =} \FunctionTok{c}\NormalTok{(}\FunctionTok{gev}\NormalTok{(), }\FunctionTok{loglogistic}\NormalTok{())),}
+      \AttributeTok{.tavg =}\NormalTok{ tavg, }\AttributeTok{.lat =}\NormalTok{ lat, }
+      \AttributeTok{.scale =}\NormalTok{ .scale, }\AttributeTok{.dist =} \FunctionTok{list}\NormalTok{(}\FunctionTok{dist\_gev}\NormalTok{(), }\FunctionTok{dist\_glo}\NormalTok{())),}
     \AttributeTok{spi =} \FunctionTok{idx\_spi}\NormalTok{(}\AttributeTok{.scale =}\NormalTok{ .scale)}
@@
 \begin{verbatim}
-# A tibble: 128,586 x 19
-   .idx  .period id             ym  prcp  tmax  tmin  tavg  long   lat
-   <chr>   <dbl> <chr>       <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
- 1 spei        6 ASN0002~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5
- 2 spei        6 ASN0002~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5
- 3 spei        6 ASN0002~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7
- 4 spei        6 ASN0002~ 1990 Jun     0  23.0  11.8  17.4  139. -20.7
- 5 spei        6 ASN0003~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9
- 6 spei        6 ASN0003~ 1990 Jun   794  25.8  18.1  21.9  146. -16.9
- 7 spei        6 ASN0003~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1
- 8 spei        6 ASN0003~ 1990 Jun   504  23.0  13.8  18.4  145. -17.1
- 9 spei        6 ASN0003~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6
-10 spei        6 ASN0003~ 1990 Jun  1970  23.9  16.4  20.2  146. -17.6
-# i 128,576 more rows
-# i 9 more variables: name <chr>, pet <dbl>, diff <dbl>,
-#   .scale <dbl>, .agg <dbl>, .method <chr>, .fitted <dbl>,
-#   .dist <chr>, .index <dbl>
+# A tibble: 128,576 x 14
+   .idx  .dist id               ym  prcp  tmax  tmin  tavg  long   lat
+   <chr> <chr> <chr>         <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
+ 1 spei  gev   ASN000290~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5
+ 2 spei  gev   ASN000290~ 1990 Jul   102  31.2  17.2  24.2  142. -15.5
+ 3 spei  gev   ASN000290~ 1990 Aug     0  31.3  13.1  22.2  142. -15.5
+ 4 spei  gev   ASN000290~ 1990 Sep     0  32.8  16.3  24.5  142. -15.5
+ 5 spei  gev   ASN000290~ 1990 Oct     0  36.8  21.5  29.2  142. -15.5
+ 6 spei  gev   ASN000290~ 1990 Nov   278  36.3  24.8  30.6  142. -15.5
+ 7 spei  gev   ASN000290~ 1990 Dec  1869  34.4  24.5  29.4  142. -15.5
+ 8 spei  gev   ASN000290~ 1990 Dec  1869  34.4  24.5  29.4  142. -15.5
+ 9 spei  gev   ASN000290~ 1991 Jan  5088  31.2  24.4  27.8  142. -15.5
+10 spei  gev   ASN000290~ 1991 Jan  5088  31.2  24.4  27.8  142. -15.5
+# i 128,566 more rows
+# i 4 more variables: name <chr>, month <dbl>, .scale <dbl>,
+#   .value <dbl>
 \end{verbatim}
@@
 transformed into normal densities in the normalizing step with
-\texttt{augment()}.
+\texttt{normalise()}.
 
@@
 \begin{Highlighting}[]
-\NormalTok{DATA }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{aggregate}\NormalTok{(}\AttributeTok{.var =}\NormalTok{ prcp, }\AttributeTok{.scale =} \DecValTok{24}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
-  \FunctionTok{dist\_fit}\NormalTok{(}\AttributeTok{.dist =} \FunctionTok{gamma}\NormalTok{(), }\AttributeTok{.var =}\NormalTok{ .agg, }\AttributeTok{.n\_boot =} \DecValTok{100}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{augment}\NormalTok{(}\AttributeTok{.var =}\NormalTok{ .agg)}
+\NormalTok{DATA }\SpecialCharTok{|\textgreater{}} 
+  \FunctionTok{temporal\_aggregate}\NormalTok{(}\AttributeTok{.agg =} \FunctionTok{temporal\_rolling\_window}\NormalTok{(prcp, }\AttributeTok{scale =} \DecValTok{24}\NormalTok{)) }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{distribution\_fit}\NormalTok{(}\AttributeTok{.fit =} \FunctionTok{dist\_gamma}\NormalTok{(}\AttributeTok{var =} \StringTok{".agg"}\NormalTok{, }\AttributeTok{method =} \StringTok{"lmoms"}\NormalTok{,}
+                                     \AttributeTok{.n\_boot =} \DecValTok{100}\NormalTok{)) }\SpecialCharTok{|\textgreater{}}
+  \FunctionTok{normalise}\NormalTok{(}\AttributeTok{.index =} \FunctionTok{norm\_quantile}\NormalTok{(.fit))}
 \end{Highlighting}

### 2023-12-03 tidyindex.qmd
Update tidyindex.qmd
minor fixes

@@
 Benchmark sets a value $b_j(s,t)$ for comparing against the original variable $x_j(s;t)$.
-This benchmark can be a fixed value consistently across space and time, perhaps extracted from expert knowledge, perhaps extracted from expert knowledge, or determined by the data through the function $u[x_j(s;t)]$.
+This benchmark can be a fixed value consistently across space and time, perhaps extracted from expert knowledge or determined by the data through the function $u[x_j(s;t)]$.
 Once a benchmark is set, observations can be highlighted for adjustments in other modules or can serve as targets for monitoring and planning.
@@
 #| fig-cap: "Index pipeline for two drought indexes: the Standardized Precipitation Index (SPI) and the Standardized Precipitation-Evapotranspiration Index (SPEI).
-Both indexes share similar construction steps with SPEI having two steps additional steps (variable transformation and dimension reduction) to convert temperature into evapotranspiration and combine it with the precipitation series.
+Both indexes share similar construction steps with SPEI having two additional steps (variable transformation and dimension reduction) to convert temperature into evapotranspiration and combine it with the precipitation series.
 "
@@
 @fig-spei illustrates the pipeline steps of the two indexes.
-The two indexes are similar with the distinct that SPEI involves two additional steps -- variable transformation and dimension reduction -- prior to temporal processing.
+The two indexes are similar with the difference that SPEI involves two additional steps -- variable transformation and dimension reduction -- prior to temporal processing.
 As introduced in @sec-software, wrapper functions are available for both indexes as `idx_spi()` and `idx_spei()`, which allows for the specification of different time scales and distributions for fitting the aggregated series.
@@
 
-The output contains the original data, index values (`.index`), parameters used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`pet`, `.agg`, and `.fitted`).
+The output contains the original data, index values (`.index`), parameters used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`.pet`, `.agg`, and `.fitted`).
 This data can be visualized to investigate the spatio-temporal distribution of the drought or flood events, as well as the response of index values to different time scales and distribution parameters at specific single locations.
@@
 A larger time scale gives a smoother index series, while also taking longer to recover from an extreme situation as seen in the  2019/20 drought period.
-The SPEI values from the two distribution fit mostly agree, while GEV can result in more extreme values, i.e.
+The SPEI values from the two distributional fits mostly agree, while GEV can result in more extreme values, i.e.
 in 1998 and 2020.'
@@
 Frames 18, 24, 29 show the GGGI when the politics component is increased.
-The most notable feature is that Bangladesh's GGGI drops substantially when politics is removed, indicating that this component plays a large role in it's relatively high value.
+The most notable feature is that Bangladesh's GGGI drops substantially when politics is removed, indicating that this component plays a large role in its relatively high value.
 Also, politics plays a substantial role in the GGGI's for the top ranked countries, because each of them drops, to the state of being similar to the middle ranked countries when the politics component's contribution is reduced.
@@
 The paper introduces a data pipeline comprising nine modules designed for the construction and analysis of indexes within the tidy framework.
-The pipeline offers a modular workflow to allow compute index with different parameterizations, to test minor changes to the original index definition, and to quantify uncertainties.
+The pipeline offers a modular workflow to compute indexes with different parameterizations, to test minor changes to the original index definition, and to quantify uncertainties.
 The framework proposed in the paper is universal to index across diverse domains.
-Examples are provided, including the drought indexes (SPI and SPEI) and Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions, to illustrate how slight adjustment of linear combination weights impact the index, and to calculate confidence intervals on the index.
+Examples are provided, including the drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions, to illustrate how slight adjustment of linear combination weights impact the index, and to calculate confidence intervals on the index.
 

### 2023-12-05 tidyindex.tex
compile

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, enhanced, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 variable \(x_j(s;t)\). This benchmark can be a fixed value consistently
-across space and time, perhaps extracted from expert knowledge, perhaps
-extracted from expert knowledge, or determined by the data through the
-function \(u[x_j(s;t)]\). Once a benchmark is set, observations can be
-highlighted for adjustments in other modules or can serve as targets for
-monitoring and planning.
+across space and time, perhaps extracted from expert knowledge or
+determined by the data through the function \(u[x_j(s;t)]\). Once a
+benchmark is set, observations can be highlighted for adjustments in
+other modules or can serve as targets for monitoring and planning.
 
@@
 Precipitation-Evapotranspiration Index (SPEI). Both indexes share
-similar construction steps with SPEI having two steps additional steps
+similar construction steps with SPEI having two additional steps
 (variable transformation and dimension reduction) to convert temperature
@@
 Figure~\ref{fig-spei} illustrates the pipeline steps of the two indexes.
-The two indexes are similar with the distinct that SPEI involves two
+The two indexes are similar with the difference that SPEI involves two
 additional steps -- variable transformation and dimension reduction --
@@
 parameters used (\texttt{.scale}, \texttt{.method}, and \texttt{.dist}),
-and all the intermediate variables (\texttt{pet}, \texttt{.agg}, and
+and all the intermediate variables (\texttt{.pet}, \texttt{.agg}, and
 \texttt{.fitted}). This data can be visualized to investigate the
@@
 from an extreme situation as seen in the 2019/20 drought period. The
-SPEI values from the two distribution fit mostly agree, while GEV can
+SPEI values from the two distributional fits mostly agree, while GEV can
 result in more extreme values, i.e.~in 1998 and 2020.}
@@
 substantially when politics is removed, indicating that this component
-plays a large role in it's relatively high value. Also, politics plays a
+plays a large role in its relatively high value. Also, politics plays a
 substantial role in the GGGI's for the top ranked countries, because
@@
 for the construction and analysis of indexes within the tidy framework.
-The pipeline offers a modular workflow to allow compute index with
-different parameterizations, to test minor changes to the original index
+The pipeline offers a modular workflow to compute indexes with different
+parameterizations, to test minor changes to the original index
 definition, and to quantify uncertainties. The framework proposed in the
 paper is universal to index across diverse domains. Examples are
-provided, including the drought indexes (SPI and SPEI) and Global Gender
-Gap Index (GGGI), to demonstrate the index calculation with different
-time scales and distributions, to illustrate how slight adjustment of
-linear combination weights impact the index, and to calculate confidence
-intervals on the index.
+provided, including the drought indexes (SPI and SPEI) and the Global
+Gender Gap Index (GGGI), to demonstrate the index calculation with
+different time scales and distributions, to illustrate how slight
+adjustment of linear combination weights impact the index, and to
+calculate confidence intervals on the index.
 
@@
 Becker, William, Giulio Caperna, Maria Del Sorbo, Hedvig Norlen, Eleni
-Papadimitriou, and Michaela Saisana. 2022. {``COINr: An r Package for
+Papadimitriou, and Michaela Saisana. 2022. {``COINr: An {R} Package for
 Developing Composite Indicators.''} \emph{Journal of Open Source
@@
 \leavevmode\vadjust pre{\hypertarget{ref-RJ-2020-027}{}}%
-Spyrison, Nicholas, and Dianne Cook. 2020. {``The r Journal: Spinifex:
-An r Package for Creating a Manual Tour of Low-Dimensional Projections
-of Multivariate Data.''} \emph{The R Journal} 12: 243--57.
+Spyrison, Nicholas, and Dianne Cook. 2020. {``Spinifex: An {R} Package
+for Creating a Manual Tour of Low-Dimensional Projections of
+Multivariate Data.''} \emph{The R Journal} 12: 243--57.
 \url{https://doi.org/10.32614/RJ-2020-027}.

### 2024-01-11 cover-letter/cover-letter.Rmd
modify Nicolas' affiliation

---
from: H. Sherry Zhang
fromqualifications: Ph.D.
fromWWW: https://huizezhangsh.netlify.app/
fromemail: huizezhangsh@gmail.com
fromsignature: sig.png
toname: Editors
toaddress: Journal of Computational and Graphical Statistics
closing: Sincerely
spacing: 1.4
output: monash::letter
---

\bigskip
Please consider our manuscript "A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data" for publication in the Journal of Computational and Graphical Statistics.
\bigskip

The paper describes a new framework for composing indexes from multivariate spatio-temporal data. The work arose since every index is developed in its own unique way, by different researchers or organisations, and often indexes designed for the same purpose cannot easily be compared. The framework described in this paper helps to standardise the construction of indexes so they can be compared, evaluated, tested, and allow for incorporating confidence interval calculations. The framework is accompanied by an R package, already on CRAN, `tidyindex`. The paper should be of interest to index developer and index analysts to think about how to construct indexes in a more systematic way.

\bigskip
Thank you for the consideration of this manuscript. We believe that it is a good fit for the Journal of Computational and Graphical Statistics.

\bigskip

### 2024-01-11 tidyindex.qmd
modify Nicolas' affiliation

@@
     keep-tex: true  
+    linestretch: 2
 pdf-engine: pdflatex
@@
       - name: BNU-HKBU United International College
-        department: Department of Mathematical Sciences
+        department: Guangdong Provincial Key Laboratory of Interdisciplinary Research and Application for Data Science
         address: 2000 Jintong Road, Tangjiawan, Zhuhai
@@
 This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship.
+Nicolas Langrené acknowledges the partial support of the Guangdong Provincial Key Laboratory IRADS (2022B1212010006, R0400001-22) and the UIC Start-up Research Fund UICR0700041-22.
 The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].

### 2024-01-11 tidyindex.tex
modify Nicolas' affiliation

@@
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 Statistics, University of Natural Resources and Life
-Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
-Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
-Guangdong, China
+Sciences, Vienna, Austria\\ \textsuperscript{3} Guangdong Provincial Key
+Laboratory of Interdisciplinary Research and Application for Data
+Science, BNU-HKBU United International College, Zhuhai, Guangdong, China
 \begin{abstract}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, enhanced]}{\end{tcolorbox}}\fi
-
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, enhanced, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+
+\setstretch{2}
 \hypertarget{introduction}{%
@@
 This work is funded by a Commonwealth Scientific and Industrial Research
-Organisation (CSIRO) Data61 Scholarship. The article is created using
-Quarto (Allaire et al. 2022) in R (R Core Team 2021). The source code
-for reproducing this paper can be found at:
+Organisation (CSIRO) Data61 Scholarship. Nicolas Langrené acknowledges
+the partial support of the Guangdong Provincial Key Laboratory IRADS
+(2022B1212010006, R0400001-22) and the UIC Start-up Research Fund
+UICR0700041-22. The article is created using Quarto (Allaire et al.
+2022) in R (R Core Team 2021). The source code for reproducing this
+paper can be found at:
 \url{https://github.com/huizezhang-sherry/paper-tidyindex}.

### 2024-01-16 tidyindex.qmd
modify patricia's affiliation

@@
         address: 29 Ancora Imparo Way, Clayton
+        city: Melbourne, Victoria
+        country: Australia
+      - name: University of Melbourne
+        department: School of Mathematics and Statistics
+        address: 813 Swanston Street, Parkville VIC 3052, Australia
         city: Melbourne, Victoria

### 2024-01-16 tidyindex.tex
modify patricia's affiliation

@@
 Langrené$\textsuperscript{3}$~\orcidlink{0000-0001-7601-4618}, Patricia
-Menéndez$\textsuperscript{1}$~\orcidlink{0000-0003-0701-6315}}
+Menéndez$\textsuperscript{1,4}$~\orcidlink{0000-0003-0701-6315}}
 
@@
 Laboratory of Interdisciplinary Research and Application for Data
-Science, BNU-HKBU United International College, Zhuhai, Guangdong, China
+Science, BNU-HKBU United International College, Zhuhai,
+Guangdong, China\\ \textsuperscript{4} School of Mathematics and
+Statistics, University of Melbourne, Melbourne, Victoria, Australia
 \begin{abstract}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, breakable, enhanced, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 

### 2024-03-19 review/response.qmd
add response to reviewer

---
title: "Response to Reviewers JCGS-23-428"
subtitle: "A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data"
author: "H. Sherry Zhang, Dianne Cook, Ursula Laa, Nicolas Langrené, Patricia Menéndez"
date: "2024-03-31"
format: pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Below is a point-by-point explanation of how we have changed the paper. The reviewer comments are in black and our responses are in red. Reviewers' comments are in black and **\textcolor{purple}{our responses are in red}**.

# Reviewer: 1

I commend the authors for their work in the tidy index library.  Certainly, a user friendly and unified approach to indices is useful across many fields.  My main concern about the paper is that it doesn't seem to fit within the aims and scope of JCGS.  A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software.  This paper, while certainly useful, doesn't necessarily propose new statistical methods but rather proposes a great tidy framework for handling index calculation.  As such, for JCGS, I think the authors need to highlight the statistical contributions more than the library.  However, I don't think the authors statistical contributions (even if highlighted) are sufficiently novel to warrant publication in JCGS.  As written, I think the paper is more suited for publication in, say, Journal of Statistical Software than JCGS.

**\textcolor{purple}{we have highlighted the methodological contributions this framework makes in the conclusion.}**. methodology contribution: allow confidence interval to be calculated in the pipeline, doing sensitivity analysis throughout the pipeline.

# Reviewer: 2

This paper proposes new pipeline work for index development, including index generation, validation, and sensitivity analysis. The authors develop a new R package, "tidyindex" including all functions in the pipeline to produce a new index. This package will be beneficial for developing new indices. The central part of this paper is the development of the R package and the explanation of its usage. However, this package needs to be fixed.

The authors explain their functions briefly in Chapter 5 and use examples with R code in Chapter 6. However, the code in Chapter 6 does not work. Also, the demo in the R package (CRAN version 0.1.0) did not work (see attached R-output.html). Please check the package, including demos and examples in this paper. It would be great if all R codes for the example in Chapter 6 were provided, including generating plots.

## minor comments

- In Table 1, it would be great if there were a line between items. Because the name of the module is in two lines, it is not easy to match the other information (input, operation, and output) **\textcolor{purple}{We rearrange the table layout and the module name fits into one line now}**.

 - On page 19, " The output contains the original data, index values (.index), parameters used (.scale,
.method, and .dist), and all the intermediate variables (.pet, .agg, and .fitted)." However, the output has no .index, .pet, .agg, and .fitted. Please check. **\textcolor{purple}{Fixed. Some columns are shown at the bottom as "i 8 more variables: lat <dbl>, name <chr>, .pet <dbl>, .diff <dbl>, .scale <chr>, .agg <dbl>, .fit <dbl>, .index <dbl>" due to the number of columns to print and the page width.}**.

- Figure 7 shows the sequence of screenshots from the animation to explore the sensitivity of the GGGI. It would be helpful to understand "what Frame 12 is"  if the authors should explain this between " Figure 7 illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries" and "Frame 12 shows the dot plot of the original index values sorted from highest to lowest", **\textcolor{purple}{Added.}**

- Figure 8 on page 27 also shows the relatively wide confidence interval around 2004. It would be great if the author mentioned it, too. **\textcolor{purple}{Added.}**

- reference for "SPEI package" on page 4 should be (Beguería and Vicente-Serrano 2017),
and reference for "the Standardized Precipitation-Evapotranspiration Index (SPEI)" on page 12 should be (Vicente-Serrano, Beguería, and López-Moreno 2010) **\textcolor{purple}{Fixed}**.

### 2024-03-19 tidyindex.qmd
add response to reviewer

@@
 There are also a few R packages supporting index calculation.
-The  `SPEI` package  [@spei] computes two drought indexes.
+The  `SPEI` package  [@SPEI] computes two drought indexes.
 The `gpindex` package [@gpindex] computes price indexes, and the `fundiversity` package [@fundiversity] computes functional diversity indexes for ecological study.
@@
 | Module                  | Input                                                                   | Operation                           | Output                                                                        |
-|--------------|---------|---------------|--------------------------|
+|-----------------------|---------|---------------|--------------------------|
 | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
@@
 The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods.
-For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) [@SPEI] calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
+For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) [@spei] calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$).
 This is the only step that differs from the Standardized Precipitation Index (SPI), and can be considered to be a dimension reduction using a particular linear combination.
@@
                ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )),
-    aes(x = long, y = lat, color = .value)) +
+    aes(x = long, y = lat, color = .index)) +
   colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
@@
   filter(name == "TEXAS POST OFFICE", .idx == "spei") %>%
-  ggplot(aes(x = ym, y = .value, color = .dist, group = .dist)) +
+  ggplot(aes(x = ym, y = .index, color = .dist, group = .dist)) +
   geom_line() +
@@
 
+library(patchwork)
 (p1 | p2)  + 
@@
 @fig-idx-tour illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries.
-Frame 12 shows the dot plot of the original index values sorted from highest to lowest.
+It presents 6 frames from the animation that varies the weights of four dimensions (economy, education, health, and politics) in the index with Frame 12 showing the dot plot of the original index values sorted from highest to lowest.
 Leading the rankings are the Nordic countries (Iceland, Norway, and Finland) and New Zealand.
@@
 From the start of 2019 to 2020, the majority of the confidence intervals lie below the extreme drought line (SPI = -2), suggesting a high level of certainty that the Texas post office is suffering from a drastic drought.
-The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantify the drought and flood severity in extreme events.
+Also close to the extreme drought line is the period 2003-2004, which corresponds to the millennium drought.
+These relatively wide confidence intervals, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggest a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantify the drought and flood severity in extreme events.
 

### 2024-03-19 tidyindex.tex
add response to reviewer

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 There are also a few R packages supporting index calculation. The
-\texttt{SPEI} package (Vicente-Serrano, Beguería, and López-Moreno 2010)
-computes two drought indexes. The \texttt{gpindex} package (Martin 2023)
-computes price indexes, and the \texttt{fundiversity} package (Grenié
-and Gruson 2023) computes functional diversity indexes for ecological
-study. The package \texttt{COINr} (Becker et al. 2022) is more
-ambitious, making a start on following the broader guidelines in the
-OECD handbook to construct, analyze, and visualize composite indexes.
+\texttt{SPEI} package (Beguería and Vicente-Serrano 2017) computes two
+drought indexes. The \texttt{gpindex} package (Martin 2023) computes
+price indexes, and the \texttt{fundiversity} package (Grenié and Gruson
+2023) computes functional diversity indexes for ecological study. The
+package \texttt{COINr} (Becker et al. 2022) is more ambitious, making a
+start on following the broader guidelines in the OECD handbook to
+construct, analyze, and visualize composite indexes.
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2188}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1406}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2344}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4062}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3151}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1233}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2055}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3562}}@{}}
 \caption{\label{tbl-notation}Summary of the notation for input,
@@
 processes, or guided by statistical methods. For example, the
-Standardized Precipitation-Evapotranspiration Index (SPEI) (Beguería and
-Vicente-Serrano 2017) calculates the difference \(D\) between
-precipitation (\(P\)) and potential evapotranspiration (\(\text{PET}\)),
-using a water balance model (\(D = P - \text{PET}\)). This is the only
-step that differs from the Standardized Precipitation Index (SPI), and
-can be considered to be a dimension reduction using a particular linear
-combination.
+Standardized Precipitation-Evapotranspiration Index (SPEI)
+(Vicente-Serrano, Beguería, and López-Moreno 2010) calculates the
+difference \(D\) between precipitation (\(P\)) and potential
+evapotranspiration (\(\text{PET}\)), using a water balance model
+(\(D = P - \text{PET}\)). This is the only step that differs from the
+Standardized Precipitation Index (SPI), and can be considered to be a
+dimension reduction using a particular linear combination.
 
@@
 \begin{verbatim}
-# A tibble: 128,576 x 14
-   .idx  .dist id               ym  prcp  tmax  tmin  tavg  long   lat
-   <chr> <chr> <chr>         <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
- 1 spei  gev   ASN000290~ 1990 Jun   170  29.7  16.2  23.0  142. -15.5
- 2 spei  gev   ASN000290~ 1990 Jul   102  31.2  17.2  24.2  142. -15.5
- 3 spei  gev   ASN000290~ 1990 Aug     0  31.3  13.1  22.2  142. -15.5
- 4 spei  gev   ASN000290~ 1990 Sep     0  32.8  16.3  24.5  142. -15.5
- 5 spei  gev   ASN000290~ 1990 Oct     0  36.8  21.5  29.2  142. -15.5
- 6 spei  gev   ASN000290~ 1990 Nov   278  36.3  24.8  30.6  142. -15.5
- 7 spei  gev   ASN000290~ 1990 Dec  1869  34.4  24.5  29.4  142. -15.5
- 8 spei  gev   ASN000290~ 1990 Dec  1869  34.4  24.5  29.4  142. -15.5
- 9 spei  gev   ASN000290~ 1991 Jan  5088  31.2  24.4  27.8  142. -15.5
-10 spei  gev   ASN000290~ 1991 Jan  5088  31.2  24.4  27.8  142. -15.5
+# A tibble: 128,576 x 18
+   .idx  .dist id         month       ym  prcp  tmax  tmin  tavg  long
+   <chr> <chr> <chr>      <dbl>    <mth> <dbl> <dbl> <dbl> <dbl> <dbl>
+ 1 spei  gev   ASN000290~     6 1990 Jun   170  29.7  16.2  23.0  142.
+ 2 spei  gev   ASN000290~     7 1990 Jul   102  31.2  17.2  24.2  142.
+ 3 spei  gev   ASN000290~     8 1990 Aug     0  31.3  13.1  22.2  142.
+ 4 spei  gev   ASN000290~     9 1990 Sep     0  32.8  16.3  24.5  142.
+ 5 spei  gev   ASN000290~    10 1990 Oct     0  36.8  21.5  29.2  142.
+ 6 spei  gev   ASN000290~    11 1990 Nov   278  36.3  24.8  30.6  142.
+ 7 spei  gev   ASN000290~    12 1990 Dec  1869  34.4  24.5  29.4  142.
+ 8 spei  gev   ASN000290~    12 1990 Dec  1869  34.4  24.5  29.4  142.
+ 9 spei  gev   ASN000290~     1 1991 Jan  5088  31.2  24.4  27.8  142.
+10 spei  gev   ASN000290~     1 1991 Jan  5088  31.2  24.4  27.8  142.
 # i 128,566 more rows
-# i 4 more variables: name <chr>, month <dbl>, .scale <dbl>,
-#   .value <dbl>
+# i 8 more variables: lat <dbl>, name <chr>, .pet <dbl>, .diff <dbl>,
+#   .scale <chr>, .agg <dbl>, .fit <dbl>, .index <dbl>
 \end{verbatim}
@@
 Figure~\ref{fig-idx-tour} illustrates doing sensitivity analysis for
-GGGI, for a subset of 16 countries. Frame 12 shows the dot plot of the
-original index values sorted from highest to lowest. Leading the
-rankings are the Nordic countries (Iceland, Norway, and Finland) and New
-Zealand. The index values are between 0 and 1, and indicate proportional
-difference between men and women, with a value of 0.8 indicating women
-are 80\% of the way to equality of these measures. There is a gap in
-values between these countries and the middle group (Brazil, Panama,
-Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia), and another big
-drop to the next group (Pakistan, Iran, Algeria, and Chad). Afghanistan
-lags much further behind.
+GGGI, for a subset of 16 countries. It presents 6 frames from the
+animation that varies the weights of four dimensions (economy,
+education, health, and politics) in the index with Frame 12 showing the
+dot plot of the original index values sorted from highest to lowest.
+Leading the rankings are the Nordic countries (Iceland, Norway, and
+Finland) and New Zealand. The index values are between 0 and 1, and
+indicate proportional difference between men and women, with a value of
+0.8 indicating women are 80\% of the way to equality of these measures.
+There is a gap in values between these countries and the middle group
+(Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia),
+and another big drop to the next group (Pakistan, Iran, Algeria, and
+Chad). Afghanistan lags much further behind.
 
@@
 suggesting a high level of certainty that the Texas post office is
-suffering from a drastic drought. The relatively wide confidence
-interval, as well as during the excessive precipitation events in
-1996-1998 and 1999-2000, suggests a high variation of the gamma
-parameters estimated from the bootstrap samples and its difficulty to
-accurately quantify the drought and flood severity in extreme events.
+suffering from a drastic drought. Also close to the extreme drought line
+is the period 2003-2004, which corresponds to the millennium drought.
+These relatively wide confidence intervals, as well as during the
+excessive precipitation events in 1996-1998 and 1999-2000, suggest a
+high variation of the gamma parameters estimated from the bootstrap
+samples and its difficulty to accurately quantify the drought and flood
+severity in extreme events.
 

### 2024-03-29 review/response.qmd
address reviewer 1 through emphasizing mothodology contribution in the conclusion

@@
 
-**\textcolor{purple}{we have highlighted the methodological contributions this framework makes in the conclusion.}**.
-methodology contribution: allow confidence interval to be calculated in the pipeline, doing sensitivity analysis throughout the pipeline.
+**\textcolor{purple}{we have highlighted the methodology contributions this framework makes in the conclusion: While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives.
+The pipeline proposed in the paper allows confidence interval to be calculated in the pipeline and sensitivity analysis to be incorporated in the index construction process.}**
 

### 2024-03-29 tidyindex.qmd
address reviewer 1 through emphasizing mothodology contribution in the conclusion

@@
     affiliations:
+      - name: University of Texas at Austin
+        department: Department of Statistics and Data Science
+        address: Welch 5.216, 105E 24th Street, Stop G3800
+        city: Austin, Texas
+        country: USA
       - name: Monash University
@@
     orcid: 0000-0002-7122-1463
-    email: huize.zhang@monash.edu
+    email: huize.zhang@austin.utexas.edu
   - name: Dianne Cook
@@
 knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
-options(width = 70)
+options(width = 70, pillow.print_max = 5, pillar.print_min = 5)
 library(tidyverse)
@@
 
-Now, we introduce the notation used for describing pipeline modules. Consider a multivariate spatio-temporal process,
-
-```{=tex}
-\begin{equation}
-\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
-\end{equation}
-```
+Now, we introduce the notation used for describing pipeline modules.
+Consider a multivariate spatio-temporal process, $$\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n $$
 where:
@@
 Simplification takes a continuous variable $x_j(s;t)$ and categorises it into a discrete set $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$ through a piecewise constant function,
-
 ```{=tex}
@@
 ```{r eval = FALSE, echo = TRUE}
-DATA |> 
-  module1(...) |>
-  module2(...) |>
-  module3(...) |>
-  ...
+DATA |> module1(...) |> module2(...) |> module3(...) |> ...
 ```
@@
 idx_spi <- function(...){
-  DATA |>
-    temporal_aggregate(...) |>
-    distribution_fit(...)|>
-    normalise(...)
+  DATA |> temporal_aggregate(...) |> distribution_fit(...)|> normalise(...)
 }
@@
 
-The paper introduces a data pipeline comprising nine modules designed for the construction and analysis of indexes within the tidy framework.
-The pipeline offers a modular workflow to compute indexes with different parameterizations, to test minor changes to the original index definition, and to quantify uncertainties.
-The framework proposed in the paper is universal to index across diverse domains.
-Examples are provided, including the drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions, to illustrate how slight adjustment of linear combination weights impact the index, and to calculate confidence intervals on the index.
+The paper introduces a data pipeline for constructing and analyzing indexes within the tidy framework.
+The pipeline comprises nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
+The proposed pipeline is implemented as an R package, `tidyindex`, available on CRAN.
+
+One of the contributions of our work is its emphasis on aspects often overlooked in index construction: uncertainty quantification and sensitivity analysis.
+While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives.
+The pipeline proposed in the paper allows for confidence interval calculation and sensitivity analysis to be incorporated in the index construction process, for indexes across various domains.
+These is illustrated through the three examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).
+
+One future development of the work is to extend the tidyindex framework to support index calculation on commonly used domain-specific data formats, for example, NetCDF for climate indexes.
+
+ <!-- Given the decisions analysts need to make when combining multivariate data into indexes, the proposed pipeline enables the evaluation of how the specific choice can affect the index, as well as how the index may appear under alternative options.
+Furthermore, uncertainty calculation can also flow through the pipeline, providing the index with confidence measures.
+-->
 

### 2024-03-29 tidyindex.tex
address reviewer 1 through emphasizing mothodology contribution in the conclusion

@@
 \author{H. Sherry
-Zhang$\textsuperscript{1}$~\orcidlink{0000-0002-7122-1463}, Dianne
-Cook$\textsuperscript{1}$~\orcidlink{0000-0002-3813-7155}, Ursula
-Laa$\textsuperscript{2}$~\orcidlink{0000-0002-0249-6439}, Nicolas
-Langrené$\textsuperscript{3}$~\orcidlink{0000-0001-7601-4618}, Patricia
-Menéndez$\textsuperscript{1,4}$~\orcidlink{0000-0003-0701-6315}}
+Zhang$\textsuperscript{1,2}$~\orcidlink{0000-0002-7122-1463}, Dianne
+Cook$\textsuperscript{2}$~\orcidlink{0000-0002-3813-7155}, Ursula
+Laa$\textsuperscript{3}$~\orcidlink{0000-0002-0249-6439}, Nicolas
+Langrené$\textsuperscript{4}$~\orcidlink{0000-0001-7601-4618}, Patricia
+Menéndez$\textsuperscript{2,5}$~\orcidlink{0000-0003-0701-6315}}
 
 \thanks{CONTACT: H. Sherry
-Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
+Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. }
 \begin{document}
@@
 \maketitle
-\textsuperscript{1} Department of Econometrics and Business
+\textsuperscript{1} Department of Statistics and Data
+Science, University of Texas at Austin, Austin,
+Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business
 Statistics, Monash University, Melbourne,
-Victoria, Australia\\ \textsuperscript{2} Institute of
+Victoria, Australia\\ \textsuperscript{3} Institute of
 Statistics, University of Natural Resources and Life
-Sciences, Vienna, Austria\\ \textsuperscript{3} Guangdong Provincial Key
+Sciences, Vienna, Austria\\ \textsuperscript{4} Guangdong Provincial Key
 Laboratory of Interdisciplinary Research and Application for Data
 Science, BNU-HKBU United International College, Zhuhai,
-Guangdong, China\\ \textsuperscript{4} School of Mathematics and
+Guangdong, China\\ \textsuperscript{5} School of Mathematics and
 Statistics, University of Melbourne, Melbourne, Victoria, Australia
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 Consider a multivariate spatio-temporal process,
-
-\begin{equation}
-\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
-\end{equation}
-
+\[\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n \]
 where:
@@
 \begin{Highlighting}[]
-\NormalTok{DATA }\SpecialCharTok{|\textgreater{}} 
-  \FunctionTok{module1}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
-  \FunctionTok{module2}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
-  \FunctionTok{module3}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
-\NormalTok{  ...}
+\NormalTok{DATA }\SpecialCharTok{|\textgreater{}} \FunctionTok{module1}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} \FunctionTok{module2}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} \FunctionTok{module3}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}\NormalTok{ ...}
 \end{Highlighting}
@@
 \NormalTok{idx\_spi }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(...)\{}
-\NormalTok{  DATA }\SpecialCharTok{|\textgreater{}}
-    \FunctionTok{temporal\_aggregate}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}}
-    \FunctionTok{distribution\_fit}\NormalTok{(...)}\SpecialCharTok{|\textgreater{}}
-    \FunctionTok{normalise}\NormalTok{(...)}
+\NormalTok{  DATA }\SpecialCharTok{|\textgreater{}} \FunctionTok{temporal\_aggregate}\NormalTok{(...) }\SpecialCharTok{|\textgreater{}} \FunctionTok{distribution\_fit}\NormalTok{(...)}\SpecialCharTok{|\textgreater{}} \FunctionTok{normalise}\NormalTok{(...)}
 \NormalTok{\}}
@@
 # A tibble: 128,576 x 18
-   .idx  .dist id         month       ym  prcp  tmax  tmin  tavg  long
-   <chr> <chr> <chr>      <dbl>    <mth> <dbl> <dbl> <dbl> <dbl> <dbl>
- 1 spei  gev   ASN000290~     6 1990 Jun   170  29.7  16.2  23.0  142.
- 2 spei  gev   ASN000290~     7 1990 Jul   102  31.2  17.2  24.2  142.
- 3 spei  gev   ASN000290~     8 1990 Aug     0  31.3  13.1  22.2  142.
- 4 spei  gev   ASN000290~     9 1990 Sep     0  32.8  16.3  24.5  142.
- 5 spei  gev   ASN000290~    10 1990 Oct     0  36.8  21.5  29.2  142.
- 6 spei  gev   ASN000290~    11 1990 Nov   278  36.3  24.8  30.6  142.
- 7 spei  gev   ASN000290~    12 1990 Dec  1869  34.4  24.5  29.4  142.
- 8 spei  gev   ASN000290~    12 1990 Dec  1869  34.4  24.5  29.4  142.
- 9 spei  gev   ASN000290~     1 1991 Jan  5088  31.2  24.4  27.8  142.
-10 spei  gev   ASN000290~     1 1991 Jan  5088  31.2  24.4  27.8  142.
-# i 128,566 more rows
-# i 8 more variables: lat <dbl>, name <chr>, .pet <dbl>, .diff <dbl>,
+  .idx  .dist id    month       ym  prcp  tmax  tmin  tavg  long   lat
+  <chr> <chr> <chr> <dbl>    <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
+1 spei  gev   ASN0~     6 1990 Jun   170  29.7  16.2  23.0  142. -15.5
+2 spei  gev   ASN0~     7 1990 Jul   102  31.2  17.2  24.2  142. -15.5
+3 spei  gev   ASN0~     8 1990 Aug     0  31.3  13.1  22.2  142. -15.5
+4 spei  gev   ASN0~     9 1990 Sep     0  32.8  16.3  24.5  142. -15.5
+5 spei  gev   ASN0~    10 1990 Oct     0  36.8  21.5  29.2  142. -15.5
+# i 128,571 more rows
+# i 7 more variables: name <chr>, .pet <dbl>, .diff <dbl>,
 #   .scale <chr>, .agg <dbl>, .fit <dbl>, .index <dbl>
@@
 
-The paper introduces a data pipeline comprising nine modules designed
-for the construction and analysis of indexes within the tidy framework.
-The pipeline offers a modular workflow to compute indexes with different
-parameterizations, to test minor changes to the original index
-definition, and to quantify uncertainties. The framework proposed in the
-paper is universal to index across diverse domains. Examples are
-provided, including the drought indexes (SPI and SPEI) and the Global
-Gender Gap Index (GGGI), to demonstrate the index calculation with
-different time scales and distributions, to illustrate how slight
-adjustment of linear combination weights impact the index, and to
-calculate confidence intervals on the index.
+The paper introduces a data pipeline for constructing and analyzing
+indexes within the tidy framework. The pipeline comprises nine modules
+including temporal and spatial aggregation, variable transformation and
+combination, distribution fitting, benchmark setting, and index
+communication. The proposed pipeline is implemented as an R package,
+\texttt{tidyindex}, available on CRAN.
+
+One of the contributions of our work is its emphasis on aspects often
+overlooked in index construction: uncertainty quantification and
+sensitivity analysis. While indexes are widely constructed and used to
+inform decision-making, such as on resource allocation, little effort is
+made to justify choices made during indexes construction and to
+experiment with alternatives. The pipeline proposed in the paper allows
+for confidence interval calculation and sensitivity analysis to be
+incorporated in the index construction process, for indexes across
+various domains. These is illustrated through the three examples on
+drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).
+
+One future development of the work is to extend the tidyindex framework
+to support index calculation on commonly used domain-specific data
+formats, for example, NetCDF for climate indexes.
 

### 2024-03-29 tidyindex.qmd
include R file and the output

@@
   init(id = country) %>%
-  add_meta(gggi_weights, var_col = variable) %>% 
+  add_paras(gggi_weights, by = "variable") %>% 
   dimension_reduction(

### 2024-03-29 tidyindex.R

@@ -1,0 +1,360 @@
+## ----setup, include=FALSE-----------------------------------------------
+knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
+options(width = 70, pillow.print_max = 5, pillar.print_min = 5)
+library(tidyverse)
+library(latex2exp)
+library(patchwork)
+library(tidyindex)
+library(lmomco)
+library(lubridate)
+library(SPEI)
+library(tsibble)
+library(GGally)
+library(patchwork)
+library(tourr)
+
+
+## ----fig-pipeline-steps-------------------------------------------------
+#| fig-align: center
+#| fig-cap: "Diagram of pipeline modules for index construction. The highlighted path illustrates one possible construction using the dimension reduction and simplification module."
+#| out-height: 90%
+#| out-width: 100%
+knitr::include_graphics(here::here("figures/pipeline-overall.png"))
+
+
+## ----fig-scale-var-trans-compare----------------------------------------
+#| fig-cap: "Comparison of the module scaling (green) and variable transformation (orange). While both modules change the variable range, scaling maintains the same distributional shape, which is not the case with variable transformation. "
+#| fig-height: 6
+#| fig-width: 8
+set.seed(123)
+x <- rgamma(n = 1000, shape = 1, scale = 2)
+
+
+trans_df <- tibble(origin = x, 
+       `z-score` = (x - mean(x)/ sd(x)), 
+       quadratic = x^2,
+       log = log(x),
+       `square root` = sqrt(x),
+       `cubic root` = x^(1/3),
+       minmax = (x - min(x))/ (max(x) - min(x)),
+       boxcox = (x^0.5 - 1)/0.5,
+       centering = x - mean(x)
+       ) %>% 
+  mutate(id = row_number(),
+         ) %>% 
+  pivot_longer(-id, names_to = "var", values_to = "value") %>% 
+  mutate(category = ifelse(
+    var %in% c("origin", "z-score", "minmax", "centering"),
+    "Scaling", "Variable Transformation")
+    )
+
+latex_df <- tibble(
+  var = c("origin", "z-score", "quadratic", "log", "square root",
+      "cubic root","minmax", "boxcox", "centering"),
+  latex = c(
+           r"($x$)",
+           r"($\frac{x - \bar{x}}{sd(x)}$)",
+           r"($x^2$)",
+           r"($log(x)$)",
+           r"($\sqrt{x}$)",
[…]

### 2024-04-11 review/response.qmd
expanded the response to review

@@
 
+# Editor and Associate Editor
+
+Two referees, an Associate Editor (AE), and I have reviewed your paper.
+The two referees hold significantly different views on the paper.
+As you will see, the first referee's assessment is rather negative, whereas the second is more positive.
+As recommended by the AE, it is essential to incorporate or highlight your methodological contributions in the paper to address the concerns raised by the first referee.
+Therefore, I would like to extend an opportunity for revision.
+However, the decision for the revision will depend on whether you can adequately address the reviewers' concerns.
+
+**\textcolor{purple}{To be addressed}**
+
 # Reviewer: 1
@@
 
-**\textcolor{purple}{we have highlighted the methodology contributions this framework makes in the conclusion: While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives.
+**\textcolor{purple}{We have highlighted the methodology contributions this framework makes in the conclusion: While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives.
 The pipeline proposed in the paper allows confidence interval to be calculated in the pipeline and sensitivity analysis to be incorporated in the index construction process.}**

### 2024-04-11 tidyindex.tex
expanded the response to review

@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \usepackage{orcidlink}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
 \setstretch{2}
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 
-\hypertarget{sec-idx-dev}{%
-\section{Background to index development}\label{sec-idx-dev}}
+\section{Background to index development}\label{sec-idx-dev}
 
@@
 
-\hypertarget{sec-tidy}{%
-\section{Tidy framework}\label{sec-tidy}}
+\section{Tidy framework}\label{sec-tidy}
 
@@
 
-\hypertarget{sec-pipeline}{%
-\section{Details of the index pipeline}\label{sec-pipeline}}
+\section{Details of the index pipeline}\label{sec-pipeline}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-overall.png}
+\centering{
+
+\includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-overall.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
@@
 
-\hypertarget{tbl-notation}{}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3562}}@{}}
-\caption{\label{tbl-notation}Summary of the notation for input,
-operation, and output of each pipeline module.}\tabularnewline
+\caption{Summary of the notation for input, operation, and output of
+each pipeline module.}\label{tbl-notation}\tabularnewline
 \toprule\noalign{}
@@
 
-\hypertarget{temporal-processing}{%
-\subsection{Temporal processing}\label{temporal-processing}}
+\subsection{Temporal processing}\label{temporal-processing}
 
@@
 
-\hypertarget{spatial-processing}{%
-\subsection{Spatial processing}\label{spatial-processing}}
+\subsection{Spatial processing}\label{spatial-processing}
 
@@
 
-\hypertarget{variable-transformation}{%
-\subsection{Variable transformation}\label{variable-transformation}}
+\subsection{Variable transformation}\label{variable-transformation}
 
@@
 
-\hypertarget{scaling}{%
-\subsection{Scaling}\label{scaling}}
+\subsection{Scaling}\label{scaling}
 
@@
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
+\centering{
+
+\includegraphics{tidyindex_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
 
@@
 
-\end{figure}
-
-\hypertarget{dimension-reduction}{%
-\subsection{Dimension reduction}\label{dimension-reduction}}
+\end{figure}%
+
+\subsection{Dimension reduction}\label{dimension-reduction}
 
@@
 
-\hypertarget{distribution-fit}{%
-\subsection{Distribution fit}\label{distribution-fit}}
+\subsection{Distribution fit}\label{distribution-fit}
 
@@
 
-\hypertarget{normalising}{%
-\subsection{Normalising}\label{normalising}}
+\subsection{Normalising}\label{normalising}
 
@@
 
-\hypertarget{benchmarking}{%
-\subsection{Benchmarking}\label{benchmarking}}
+\subsection{Benchmarking}\label{benchmarking}
 
@@
 
-\hypertarget{simplification}{%
-\subsection{Simplification}\label{simplification}}
+\subsection{Simplification}\label{simplification}
 
@@
 
-\hypertarget{sec-software}{%
-\section{Software design}\label{sec-software}}
+\section{Software design}\label{sec-software}
 
@@
 
-\hypertarget{sec-examples}{%
-\section{Examples}\label{sec-examples}}
+\section{Examples}\label{sec-examples}
 
@@
 
-\hypertarget{sec-example1}{%
 \subsection{Every distribution, every scale, every index all at
-once}\label{sec-example1}}
+once}\label{sec-example1}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-spei.png}
+\centering{
+
+\includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-spei.png}
 
@@
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/fig-compute-spatial-1.pdf}
+\centering{
+
+\includegraphics{tidyindex_files/figure-pdf/fig-compute-spatial-1.pdf}
 
@@
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/fig-compute-temporal-1.pdf}
+\centering{
+
+\includegraphics{tidyindex_files/figure-pdf/fig-compute-temporal-1.pdf}
 
@@
 
-\end{figure}
-
-\hypertarget{does-a-puff-of-change-in-variable-weights-cause-a-tornado-in-ranks}{%
+\end{figure}%
+
 \subsection{Does a puff of change in variable weights cause a tornado in
-ranks?}\label{does-a-puff-of-change-in-variable-weights-cause-a-tornado-in-ranks}}
+ranks?}\label{does-a-puff-of-change-in-variable-weights-cause-a-tornado-in-ranks}
 
@@
 
-{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-gggi.png}
+\centering{
+
+\includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-gggi.png}
 
@@
 
-\end{figure}
-
-\hypertarget{tbl-gggi-weights}{}
+\end{figure}%
+
 \begin{table}
-\caption{\label{tbl-gggi-weights}Weights for the two applications of dimension reduction to compute the
-Global Gender Gap Index. V-wgt is used to compute four new variables
-from the original 14. These are then equally combined to get the final
-index value. }\tabularnewline
+
+\centering{
 
@@
 \endgroup{}
-\end{table}
+
+}
+
+\caption{\label{tbl-gggi-weights}Weights for the two applications of
+dimension reduction to compute the Global Gender Gap Index. V-wgt is
+used to compute four new variables from the original 14. These are then
+equally combined to get the final index value.}
+
+\end{table}%
 
@@
   \FunctionTok{init}\NormalTok{(}\AttributeTok{id =}\NormalTok{ country) }\SpecialCharTok{\%\textgreater{}\%}
-  \FunctionTok{add\_meta}\NormalTok{(gggi\_weights, }\AttributeTok{var\_col =}\NormalTok{ variable) }\SpecialCharTok{\%\textgreater{}\%} 
+  \FunctionTok{add\_paras}\NormalTok{(gggi\_weights, }\AttributeTok{by =} \StringTok{"variable"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
   \FunctionTok{dimension\_reduction}\NormalTok{(}
@@
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/fig-idx-tour-1.pdf}
+\centering{
+
+\includegraphics{tidyindex_files/figure-pdf/fig-idx-tour-1.pdf}
 
@@
 
-\hypertarget{decoding-uncertainty-through-the-wisdom-of-the-crowd}{%
 \subsection{Decoding uncertainty through the wisdom of the
-crowd}\label{decoding-uncertainty-through-the-wisdom-of-the-crowd}}
+crowd}\label{decoding-uncertainty-through-the-wisdom-of-the-crowd}
 
@@
 
-{\centering \includegraphics{tidyindex_files/figure-pdf/fig-conf-interval-1.pdf}
+\centering{
+
+\includegraphics{tidyindex_files/figure-pdf/fig-conf-interval-1.pdf}
 
@@
 
-\end{figure}
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+\end{figure}%
+
+\section{Conclusion}\label{conclusion}
 
@@
 
-\hypertarget{acknowledgement}{%
-\section{Acknowledgement}\label{acknowledgement}}
+\section{Acknowledgement}\label{acknowledgement}
 
@@
 
-\hypertarget{reference}{%
-\section*{Reference}\label{reference}}
+\section*{Reference}\label{reference}
 \addcontentsline{toc}{section}{Reference}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alahacoon_comprehensive_2022}{}}%
+\bibitem[\citeproctext]{ref-alahacoon_comprehensive_2022}
 Alahacoon, Niranga, and Mahesh Edirisinghe. 2022. {``A Comprehensive
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Allaire_Quarto_2022}{}}%
+\bibitem[\citeproctext]{ref-Allaire_Quarto_2022}
 Allaire, J. J., Charles Teague, Carlos Scheidegger, Yihui Xie, and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-COINr}{}}%
+\bibitem[\citeproctext]{ref-COINr}
 Becker, William, Giulio Caperna, Maria Del Sorbo, Hedvig Norlen, Eleni
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-SPEI}{}}%
+\bibitem[\citeproctext]{ref-SPEI}
 Beguería, Santiago, and Sergio M. Vicente-Serrano. 2017. \emph{SPEI:
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-buja_elements_1988}{}}%
+\bibitem[\citeproctext]{ref-buja_elements_1988}
 Buja, A, D Asimov, C Hurley, and JA McDonald. 1988. {``Elements of a
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-buja2005computational}{}}%
+\bibitem[\citeproctext]{ref-buja2005computational}
 Buja, Andreas, Dianne Cook, Daniel Asimov, and Catherine Hurley. 2005.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fisher1936use}{}}%
+\bibitem[\citeproctext]{ref-fisher1936use}
 Fisher, Ronald A. 1936. {``The Use of Multiple Measurements in Taxonomic
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fisher1970statistical}{}}%
+\bibitem[\citeproctext]{ref-fisher1970statistical}
 Fisher, Ronald Aylmer. 1970. {``Statistical Methods for Research
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fundiversity}{}}%
+\bibitem[\citeproctext]{ref-fundiversity}
 Grenié, Matthias, and Hugo Gruson. 2023. \emph{{fundiversity}: Easy
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hao_drought_2015}{}}%
+\bibitem[\citeproctext]{ref-hao_drought_2015}
 Hao, Zengchao, and Vijay P. Singh. 2015. {``Drought Characterization
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hotelling1933analysis}{}}%
+\bibitem[\citeproctext]{ref-hotelling1933analysis}
 Hotelling, Harold. 1933. {``Analysis of a Complex of Statistical
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-jones_vulnerability_2007}{}}%
+\bibitem[\citeproctext]{ref-jones_vulnerability_2007}
 Jones, Brenda, and Jean Andrey. 2007. {``Vulnerability Index
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tidymodels}{}}%
+\bibitem[\citeproctext]{ref-tidymodels}
 Kuhn, Max, and Hadley Wickham. 2020. \emph{Tidymodels: A Collection of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-laimighofer_how_2022}{}}%
+\bibitem[\citeproctext]{ref-laimighofer_how_2022}
 Laimighofer, Johannes, and Gregor Laaha. 2022. {``How Standard Are
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gpindex}{}}%
+\bibitem[\citeproctext]{ref-gpindex}
 Martin, Steve. 2023. \emph{Gpindex: Generalized Price and Quantity
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mckee1993relationship}{}}%
+\bibitem[\citeproctext]{ref-mckee1993relationship}
 McKee, Thomas B, Nolan J Doesken, John Kleist, et al. 1993. {``The
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-oecd_handbook_2008}{}}%
+\bibitem[\citeproctext]{ref-oecd_handbook_2008}
 OECD, European Union, and Joint Research Centre - European Commission.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
+\bibitem[\citeproctext]{ref-R}
 R Core Team. 2021. \emph{R: A Language and Environment for Statistical
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-uncertainty}{}}%
+\bibitem[\citeproctext]{ref-uncertainty}
 Saisana, M., A. Saltelli, and S. Tarantola. 2005. {``{Uncertainty and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-RJ-2020-027}{}}%
+\bibitem[\citeproctext]{ref-RJ-2020-027}
 Spyrison, Nicholas, and Dianne Cook. 2020. {``Spinifex: An {R} Package
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sutherland_orca_2000}{}}%
+\bibitem[\citeproctext]{ref-sutherland_orca_2000}
 Sutherland, Peter, Anthony Rossini, Thomas Lumley, Nicholas Lewin-Koh,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-svoboda2016handbook}{}}%
+\bibitem[\citeproctext]{ref-svoboda2016handbook}
 Svoboda, Mark, Brian Fuchs, et al. 2016. {``Handbook of Drought
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tate_social_2012}{}}%
+\bibitem[\citeproctext]{ref-tate_social_2012}
 Tate, Eric. 2012. {``Social Vulnerability Indices: A Comparative
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tate_uncertainty_2013}{}}%
+\bibitem[\citeproctext]{ref-tate_uncertainty_2013}
 ---------. 2013. {``Uncertainty {Analysis} for a {Social}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%
+\bibitem[\citeproctext]{ref-spei}
 Vicente-Serrano, Sergio M., Santiago Beguería, and Juan I. López-Moreno.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%
+\bibitem[\citeproctext]{ref-wickham_tidy_2014}
 Wickham, Hadley. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wickham_welcome_2019}{}}%
+\bibitem[\citeproctext]{ref-wickham_welcome_2019}
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wickham_tourr_2011}{}}%
+\bibitem[\citeproctext]{ref-wickham_tourr_2011}
 Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%
+\bibitem[\citeproctext]{ref-wickham_plumbing_2009}
 Wickham, Hadley, Michael Lawrence, Dianne Cook, Andreas Buja, Heike
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-WEF2023}{}}%
+\bibitem[\citeproctext]{ref-WEF2023}
 World Economic Forum. 2023. {``{The Global Gender Gap Report 2023}.''}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-xie_reactive_2014}{}}%
+\bibitem[\citeproctext]{ref-xie_reactive_2014}
 Xie, Yihui, Heike Hofmann, and Xiaoyue Cheng. 2014. {``Reactive
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-zargar2011review}{}}%
+\bibitem[\citeproctext]{ref-zargar2011review}
 Zargar, Amin, Rehan Sadiq, Bahman Naser, and Faisal I Khan. 2011. {``A

### 2024-04-11 tidyindex.qmd
Updates in abstract

@@
   Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making.
-While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their values and variation in values.
+While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their values and the variability therein.
 Here we discuss a modular data pipeline recommendation to assemble indexes.
@@
 One can compute indexes with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess indexes' robustness.
-The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on countries' ranking; and how to calculate bootstrap confidence intervals for the Standardized Precipitation Index (SPI).
+The paper presents three examples to illustrate the usage of the pipeline framework: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on countries' ranking; and how to calculate bootstrap confidence intervals for the Standardized Precipitation Index (SPI).
 The methods are supported by a new R package, called `tidyindex`.

### 2024-04-11 tidyindex.tex
Updates in abstract

--- 2024-03-29 tidyindex.tex
@@
 index behavior in different data conditions, and to determine how their
-structure affects their values and variation in values. Here we discuss
-a modular data pipeline recommendation to assemble indexes. It is
-universally applicable to index computation and allows investigation of
-index behavior as part of the development procedure. One can compute
+structure affects their values and the variability therein. Here we
+discuss a modular data pipeline recommendation to assemble indexes. It
+is universally applicable to index computation and allows investigation
+of index behavior as part of the development procedure. One can compute
 indexes with different parameter choices, adjust steps in the index
@@
 indexes' robustness. The paper presents three examples to illustrate the
-pipeline framework usage: comparison of two different indexes designed
-to monitor the spatio-temporal distribution of drought in Queensland,
-Australia; the effect of dimension reduction choices on the Global
-Gender Gap Index (GGGI) on countries' ranking; and how to calculate
-bootstrap confidence intervals for the Standardized Precipitation Index
-(SPI). The methods are supported by a new R package, called
-\texttt{tidyindex}.
+usage of the pipeline framework: comparison of two different indexes
+designed to monitor the spatio-temporal distribution of drought in
+Queensland, Australia; the effect of dimension reduction choices on the
+Global Gender Gap Index (GGGI) on countries' ranking; and how to
+calculate bootstrap confidence intervals for the Standardized
+Precipitation Index (SPI). The methods are supported by a new R package,
+called \texttt{tidyindex}.
 \end{abstract}
@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
+\begingroup\fontsize{11}{13}\selectfont
+
 \hypertarget{tbl-gggi-weights}{}
-\begin{table}
+\begin{longtable}[t]{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
 \caption{\label{tbl-gggi-weights}Weights for the two applications of dimension reduction to compute the
@@
 
-\centering\begingroup\fontsize{11}{13}\selectfont
-
-\begin{tabular}{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
 \toprule
@@
 Professional and technical workers & 0.121 &  &  & 0.030\\
-\hline
+\addlinespace
 Literacy rate & 0.191 & Education & 0.25 & 0.048\\
@@
 Enrolment in tertiary education & 0.121 &  &  & 0.030\\
-\hline
 Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
+\addlinespace
 Healthy life expectancy & 0.307 &  &  & 0.077\\
-\hline
 Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
@@
 \bottomrule
-\end{tabular}
+\end{longtable}
 \endgroup{}
-\end{table}
 

### 2024-04-11 tidyindex.tex
Minor change in section 1

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 definition of the index to be used, among others. For instance, the
-indexes constructed from a linear combination of variables require to
-decide the weight assigned to each variable. Some indexes have a spatial
-and/or temporal component, and variables can be aggregated to different
-spatial resolutions and temporal scales, leading to various indexes for
-different monitoring purposes. Hence, all these decisions can result in
-different index values and have different practical implications.
+indexes constructed from a linear combination of variables require a
+decision on the weight assigned to each variable. Some indexes have a
+spatial and/or temporal component, and variables can be aggregated to
+different spatial resolutions and temporal scales, leading to various
+indexes for different monitoring purposes. Hence, all these decisions
+can result in different index values and have different practical
+implications.
 

### 2024-04-12 tidyindex.qmd
Minor change in section 1
Add on reproducibility sec 3

@@
 The construction process is often not straightforward, and decisions need to be made, such as the selection of variables to be included, which might depend on data availability and the statistical definition of the index to be used, among others.
-For instance, the indexes constructed from a linear combination of variables require to decide the weight assigned to each variable.
+For instance, the indexes constructed from a linear combination of variables require a decision on the weight assigned to each variable.
 Some indexes have a spatial and/or temporal component, and variables can be aggregated to different spatial resolutions and temporal scales, leading to various indexes for different monitoring purposes.
@@
 
--   makes the computation more transparent, and thus more easily debugged.
+-   makes the computation more transparent, and thus more easily debugged, facilitating reproducibility.
 -   allows for rapidly processing new data to check how different features, like outliers, might affect the index value.

### 2024-04-12 tidyindex.tex
Add on reproducibility sec 3

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 \item
-  makes the computation more transparent, and thus more easily debugged.
+  makes the computation more transparent, and thus more easily debugged,
+  facilitating reproducibility.
 \item

### 2024-04-12 tidyindex.tex
Fix typo intro sec 4

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, interior hidden, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, interior hidden, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
 
@@
 varying process for a fixed \(t\). An overview of the notation for
-pipeline input, operation, and output is present in
+pipeline input, operation, and output is presented in
 Table~\ref{tbl-notation}.

### 2024-04-12 tidyindex.qmd
Fix typo intro sec 4
Changes sect 4

@@
 While $x_{tj}(s)$ denotes the spatial varying process for a fixed $t$.
-An overview of the notation for pipeline input, operation, and output is present in @tbl-notation.
+An overview of the notation for pipeline input, operation, and output is presented in @tbl-notation.
 
@@
 The spatial processing module takes a single variable with a fixed temporal dimension, $x_{tj}(s)$, as input.
-This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]$.
+This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]$ via a function $g$.
 The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions.
@@
 
-Variable transformation takes the input of a single variable $x_j(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^{\text{Trans}}_{j}(s;t)$.
+Variable transformation takes the input of a single variable $x_j(s;t)$ and reshapes its distribution using the function $T$ to produce $x^{\text{Trans}}_{j}(s;t)$.
 When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality.
@@
 Unlike variable transformation, scaling maintains the distributional shape of the variable.
-It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$.
+It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$ where $\alpha$ and $\gamma$ are constants .
 In the Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.
@@
 
-Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$, or a subset of variables in $\mathbf{x}(s;t)$, as the input.
-It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output.
+Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^m \times \mathbb{R}^n$, or a subset of variables in $\mathbf{x}(s;t)$, as the input.
+It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$, as the output.
 The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods.

### 2024-04-12 tidyindex.tex
Changes sect 4

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, interior hidden, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 dimension \(s^\prime \in D_{s^\prime}\) through
-\(x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]\). The change of spatial
-dimension allows for the alignment of variables collected from different
-measurements, such as in-situ stations and satellite imagery, or
-originating from different resolutions. This also includes the
-aggregation of variables into different levels, such as city, state, and
-country scales.
+\(x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]\) via a function \(g\).
+The change of spatial dimension allows for the alignment of variables
+collected from different measurements, such as in-situ stations and
+satellite imagery, or originating from different resolutions. This also
+includes the aggregation of variables into different levels, such as
+city, state, and country scales.
 
@@
 Variable transformation takes the input of a single variable
-\(x_j(s;t)\) and reshapes its distribution using the function \(T[.]\)
-to produce \(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
+\(x_j(s;t)\) and reshapes its distribution using the function \(T\) to
+produce \(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
 distribution, transformations such as log, square root, or cubic root
@@
 standardization, and min-max standardization and can be expressed as
-\([x_{j}(s;t) - \alpha]/\gamma\). In the Human Development Index (HDI),
-the three dimensions (health, education, and economy) are converted into
-the same scale (0-1) using min-max standardization.
+\([x_{j}(s;t) - \alpha]/\gamma\) where \(\alpha\) and \(\gamma\) are
+constants . In the Human Development Index (HDI), the three dimensions
+(health, education, and economy) are converted into the same scale (0-1)
+using min-max standardization.
 
@@
 Dimension reduction takes the multivariate information
-\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\), or a
-subset of variables in \(\mathbf{x}(s;t)\), as the input. It summarises
-the high-dimensional information into a lower-dimension representation
-\(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\) and \(d < p\),
-as the output. The transformation can be based on domain-specific
-knowledge, originating from theories describing the underlying physical
-processes, or guided by statistical methods. For example, the
-Standardized Precipitation-Evapotranspiration Index (SPEI)
+\(\mathbf{x}(s;t)\), where
+\(\mathbf{x} \subseteq \mathbb{R}^m \times \mathbb{R}^n\), or a subset
+of variables in \(\mathbf{x}(s;t)\), as the input. It summarises the
+high-dimensional information into a lower-dimension representation
+\(\mathbf{y}(s;t)\), where \(\mathbf{y} \subseteq \mathbb{R}^d\) and
+\(d < p\), as the output. The transformation can be based on
+domain-specific knowledge, originating from theories describing the
+underlying physical processes, or guided by statistical methods. For
+example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
 (Vicente-Serrano, Beguería, and López-Moreno 2010) calculates the

### 2024-04-12 tidyindex.qmd
Updates to section 4.5

@@
 
-Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^m \times \mathbb{R}^n$, or a subset of variables in $\mathbf{x}(s;t)$, as the input.
+Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$, or a subset of variables $x_i(s;t)$ in  $\mathbf{x}(s;t)$, as the input.
 It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$, as the output.

### 2024-04-12 tidyindex.tex
Updates to section 4.5

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 Dimension reduction takes the multivariate information
-\(\mathbf{x}(s;t)\), where
-\(\mathbf{x} \subseteq \mathbb{R}^m \times \mathbb{R}^n\), or a subset
-of variables in \(\mathbf{x}(s;t)\), as the input. It summarises the
-high-dimensional information into a lower-dimension representation
-\(\mathbf{y}(s;t)\), where \(\mathbf{y} \subseteq \mathbb{R}^d\) and
-\(d < p\), as the output. The transformation can be based on
-domain-specific knowledge, originating from theories describing the
-underlying physical processes, or guided by statistical methods. For
-example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
-(Vicente-Serrano, Beguería, and López-Moreno 2010) calculates the
-difference \(D\) between precipitation (\(P\)) and potential
-evapotranspiration (\(\text{PET}\)), using a water balance model
-(\(D = P - \text{PET}\)). This is the only step that differs from the
-Standardized Precipitation Index (SPI), and can be considered to be a
-dimension reduction using a particular linear combination.
+\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\), or a
+subset of variables \(x_i(s;t)\) in \(\mathbf{x}(s;t)\), as the input.
+It summarises the high-dimensional information into a lower-dimension
+representation \(\mathbf{y}(s;t)\), where
+\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\), as the output. The
+transformation can be based on domain-specific knowledge, originating
+from theories describing the underlying physical processes, or guided by
+statistical methods. For example, the Standardized
+Precipitation-Evapotranspiration Index (SPEI) (Vicente-Serrano,
+Beguería, and López-Moreno 2010) calculates the difference \(D\) between
+precipitation (\(P\)) and potential evapotranspiration (\(\text{PET}\)),
+using a water balance model (\(D = P - \text{PET}\)). This is the only
+step that differs from the Standardized Precipitation Index (SPI), and
+can be considered to be a dimension reduction using a particular linear
+combination.
 

### 2024-04-12 tidyindex.qmd
Update conclusions

@@
 
-The paper introduces a data pipeline for constructing and analyzing indexes within the tidy framework.
+The paper introduces a novel data pipeline for constructing and analyzing indexes within the tidy framework.
 The pipeline comprises nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
@@
 
-One of the contributions of our work is its emphasis on aspects often overlooked in index construction: uncertainty quantification and sensitivity analysis.
-While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives.
+One of the contributions of our work is its emphasis on aspects often overlooked in index construction: uncertainty quantification and sensitivity.
 The pipeline proposed in the paper allows for confidence interval calculation and sensitivity analysis to be incorporated in the index construction process, for indexes across various domains.
 These is illustrated through the three examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).
+While indexes are widely constructed and used to inform decision-making, such as in resource allocation and policy making worldwide, little effort is made to justify the choices made during index construction or to experiment with alternatives while providing means for reproducibility.
+This work highlights the importance of carefully constructing statistical indexes and introduces novel computational methods for studying and calculating statistical indexes and their uncertainties.
 

### 2024-04-12 tidyindex.tex
Update conclusions

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-The paper introduces a data pipeline for constructing and analyzing
-indexes within the tidy framework. The pipeline comprises nine modules
-including temporal and spatial aggregation, variable transformation and
-combination, distribution fitting, benchmark setting, and index
-communication. The proposed pipeline is implemented as an R package,
-\texttt{tidyindex}, available on CRAN.
+The paper introduces a novel data pipeline for constructing and
+analyzing indexes within the tidy framework. The pipeline comprises nine
+modules including temporal and spatial aggregation, variable
+transformation and combination, distribution fitting, benchmark setting,
+and index communication. The proposed pipeline is implemented as an R
+package, \texttt{tidyindex}, available on CRAN.
 
@@
 overlooked in index construction: uncertainty quantification and
-sensitivity analysis. While indexes are widely constructed and used to
-inform decision-making, such as on resource allocation, little effort is
-made to justify choices made during indexes construction and to
-experiment with alternatives. The pipeline proposed in the paper allows
-for confidence interval calculation and sensitivity analysis to be
-incorporated in the index construction process, for indexes across
-various domains. These is illustrated through the three examples on
-drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).
+sensitivity. The pipeline proposed in the paper allows for confidence
+interval calculation and sensitivity analysis to be incorporated in the
+index construction process, for indexes across various domains. These is
+illustrated through the three examples on drought indexes (SPI and SPEI)
+and the Global Gender Gap Index (GGGI). While indexes are widely
+constructed and used to inform decision-making, such as in resource
+allocation and policy making worldwide, little effort is made to justify
+the choices made during index construction or to experiment with
+alternatives while providing means for reproducibility. This work
+highlights the importance of carefully constructing statistical indexes
+and introduces novel computational methods for studying and calculating
+statistical indexes and their uncertainties.
 

### 2024-04-15 review/response.qmd
expand the response to reviewer

@@
 
+**\textcolor{purple}{Fixed.
+The scripts in the demo folder contains some legacy code from the past and we have cleaned them up, as well as cross-checked teh code in the paper.
+The new CRAN version???[TODO].
+We select the most relevant code to present in the paper given the length of the manuscript.
+However, all the code used to reproduce the paper is available in the GitHub repo, which can be found from the Acknowledgement section.}**
+
 ## minor comments
@@
 **\textcolor{purple}{Fixed.
-Some columns are shown at the bottom as "i 8 more variables: lat <dbl>, name <chr>, .pet <dbl>, .diff <dbl>, .scale <chr>, .agg <dbl>, .fit <dbl>, .index <dbl>" due to the number of columns to print and the page width.}**.
+We use the dplyr::glimpse() function to print all the columns of the output.}**.
 
 - Figure 7 shows the sequence of screenshots from the animation to explore the sensitivity of the GGGI.
-It would be helpful to understand "what Frame 12 is"  if the authors should explain this between " Figure 7 illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries" and "Frame 12 shows the dot plot of the original index values sorted from highest to lowest", **\textcolor{purple}{Added.}**
+It would be helpful to understand "what Frame 12 is"  if the authors should explain this between " Figure 7 illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries" and "Frame 12 shows the dot plot of the original index values sorted from highest to lowest", **\textcolor{purple}{We have added more text to describe Frame 12 being from an animation where the weight on one dimension, politics, is gradually increased.
+}**
 
-- Figure 8 on page 27 also shows the relatively wide confidence interval around 2004. It would be great if the author mentioned it, too. **\textcolor{purple}{Added.}**
+- Figure 8 on page 27 also shows the relatively wide confidence interval around 2004.
+It would be great if the author mentioned it, too.
+**\textcolor{purple}{We point out the wider confidence interval around 2004 corresponds to the millennium drought in Australia.}**
 

### 2024-04-15 tidyindex.qmd
expand the response to reviewer

@@
 
-```{r}
-idx
+We use the `dplyr::glimpse()` function to inspect the `idx` object created:
+
+```{r echo = TRUE}
+dplyr::glimpse(idx)
 ```
@@
 @fig-idx-tour illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries.
-It presents 6 frames from the animation that varies the weights of four dimensions (economy, education, health, and politics) in the index with Frame 12 showing the dot plot of the original index values sorted from highest to lowest.
-Leading the rankings are the Nordic countries (Iceland, Norway, and Finland) and New Zealand.
+It presents 6 frames selected from the animation where the weight on the politics dimension is gradually increased, where other dimensions (economy, education, health) decrease correspondingly.
+Frame 12 present the original index where all the four dimensions receive equal weight.
+The index values are sorted from highest to lowest, with the Nordic countries (Iceland, Norway, and Finland) and New Zealand leading the rankings.
 The index values are between 0 and 1, and indicate proportional difference between men and women, with a value of 0.8 indicating women are 80% of the way to equality of these measures.

### 2024-04-15 tidyindex.tex
expand the response to reviewer

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
+We use the \texttt{dplyr::glimpse()} function to inspect the
+\texttt{idx} object created:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{dplyr}\SpecialCharTok{::}\FunctionTok{glimpse}\NormalTok{(idx)}
+\end{Highlighting}
+\end{Shaded}
+
 \begin{verbatim}
-# A tibble: 128,576 x 18
-  .idx  .dist id    month       ym  prcp  tmax  tmin  tavg  long   lat
-  <chr> <chr> <chr> <dbl>    <mth> <dbl> <dbl> <dbl> <dbl> <dbl> <dbl>
-1 spei  gev   ASN0~     6 1990 Jun   170  29.7  16.2  23.0  142. -15.5
-2 spei  gev   ASN0~     7 1990 Jul   102  31.2  17.2  24.2  142. -15.5
-3 spei  gev   ASN0~     8 1990 Aug     0  31.3  13.1  22.2  142. -15.5
-4 spei  gev   ASN0~     9 1990 Sep     0  32.8  16.3  24.5  142. -15.5
-5 spei  gev   ASN0~    10 1990 Oct     0  36.8  21.5  29.2  142. -15.5
-# i 128,571 more rows
-# i 7 more variables: name <chr>, .pet <dbl>, .diff <dbl>,
-#   .scale <chr>, .agg <dbl>, .fit <dbl>, .index <dbl>
+Rows: 128,576
+Columns: 18
+$ .idx   <chr> "spei", "spei", "spei", "spei", "spei", "spei", "spei~
+$ .dist  <chr> "gev", "gev", "gev", "gev", "gev", "gev", "gev", "gev~
+$ id     <chr> "ASN00029038", "ASN00029038", "ASN00029038", "ASN0002~
+$ month  <dbl> 6, 7, 8, 9, 10, 11, 12, 12, 1, 1, 2, 2, 3, 3, 4, 4, 5~
+$ ym     <mth> 1990 Jun, 1990 Jul, 1990 Aug, 1990 Sep, 1990 Oct, 199~
+$ prcp   <dbl> 170, 102, 0, 0, 0, 278, 1869, 1869, 5088, 5088, 8484,~
+$ tmax   <dbl> 29.65357, 31.20323, 31.32581, 32.80870, 36.80357, 36.~
+$ tmin   <dbl> 16.25000, 17.15161, 13.11613, 16.25714, 21.49655, 24.~
+$ tavg   <dbl> 22.95179, 24.17742, 22.22097, 24.53292, 29.15006, 30.~
+$ long   <dbl> 141.7483, 141.7483, 141.7483, 141.7483, 141.7483, 141~
+$ lat    <dbl> -15.4818, -15.4818, -15.4818, -15.4818, -15.4818, -15~
+$ name   <chr> "KOWANYAMA AIRPORT", "KOWANYAMA AIRPORT", "KOWANYAMA ~
+$ .pet   <dbl> 67.46933, 86.64868, 63.27450, 94.93572, 204.63793, 24~
+$ .diff  <dbl> 102.53067, 15.35132, -63.27450, -94.93572, -204.63793~
+$ .scale <chr> "6", "6", "6", "6", "6", "6", "6", "12", "6", "12", "~
+$ .agg   <dbl> 4263.7863, 2819.8773, 2529.0243, 578.8843, -117.6571,~
+$ .fit   <dbl> 0.02902164, 0.10512807, 0.57680687, 0.83297600, 0.818~
+$ .index <dbl> -1.89537090, -1.25286143, 0.19373133, 0.96599235, 0.9~
 \end{verbatim}
@@
 Figure~\ref{fig-idx-tour} illustrates doing sensitivity analysis for
-GGGI, for a subset of 16 countries. It presents 6 frames from the
-animation that varies the weights of four dimensions (economy,
-education, health, and politics) in the index with Frame 12 showing the
-dot plot of the original index values sorted from highest to lowest.
-Leading the rankings are the Nordic countries (Iceland, Norway, and
-Finland) and New Zealand. The index values are between 0 and 1, and
-indicate proportional difference between men and women, with a value of
-0.8 indicating women are 80\% of the way to equality of these measures.
-There is a gap in values between these countries and the middle group
-(Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia),
-and another big drop to the next group (Pakistan, Iran, Algeria, and
-Chad). Afghanistan lags much further behind.
+GGGI, for a subset of 16 countries. It presents 6 frames selected from
+the animation where the weight on the politics dimension is gradually
+increased, where other dimensions (economy, education, health) decrease
+correspondingly. Frame 12 present the original index where all the four
+dimensions receive equal weight. The index values are sorted from
+highest to lowest, with the Nordic countries (Iceland, Norway, and
+Finland) and New Zealand leading the rankings. The index values are
+between 0 and 1, and indicate proportional difference between men and
+women, with a value of 0.8 indicating women are 80\% of the way to
+equality of these measures. There is a gap in values between these
+countries and the middle group (Brazil, Panama, Poland, Bangladesh,
+Kazakhstan, Armenia, and Slovakia), and another big drop to the next
+group (Pakistan, Iran, Algeria, and Chad). Afghanistan lags much further
+behind.
 

### 2024-04-16 review/response.qmd
tidied intro and conclusions

--- 2024-04-11 review/response.qmd
@@
 
+**\textcolor{purple}{Fixed.
+We were expecting that readers worked from the vignettes folder.
+The problem was that the demo folder had some irrelevant code.
+We've removed these now.
+We have also re-organised the code for the paper as follows.}**
+
+- **\textcolor{purple}{The main file tidyindex.R, reproduces the content in the paper from saved data objects, available in the data folder.}** 
+- **\textcolor{purple}{New scripts have been provided in the scripts folder which completely generate each of the examples.
+There is a README.md, that instructs the reader how to set their environment up to reproduce all of the analyses.}**
+- **\textcolor{purple}{The GitHub repo is listed in the Acknowledgements section, and any further updates needed for future versions to run will be made there.}**
+
 ## minor comments
 
-- In Table 1, it would be great if there were a line between items.
-Because the name of the module is in two lines, it is not easy to match the other information (input, operation, and output) **\textcolor{purple}{We rearrange the table layout and the module name fits into one line now}**.
+- In Table 1, it would be great if there were a line between items. Because the name of the module is in two lines, it is not easy to match the other information (input, operation, and output)
+
+**\textcolor{purple}{Fixed. The table layout has been modified accordingly.}**.
 
  - On page 19, " The output contains the original data, index values (.index), parameters used (.scale,
-.method, and .dist), and all the intermediate variables (.pet, .agg, and .fitted)." However, the output has no .index, .pet, .agg, and .fitted.
-Please check.
-**\textcolor{purple}{Fixed.
-Some columns are shown at the bottom as "i 8 more variables: lat <dbl>, name <chr>, .pet <dbl>, .diff <dbl>, .scale <chr>, .agg <dbl>, .fit <dbl>, .index <dbl>" due to the number of columns to print and the page width.}**.
+.method, and .dist), and all the intermediate variables (.pet, .agg, and .fitted)." However, the output has no .index, .pet, .agg, and .fitted. Please check.
+
+**\textcolor{purple}{Fixed. The output has been written in a way that all variables are visible now (using dplyr::glimpse()).}**.
 
 - Figure 7 shows the sequence of screenshots from the animation to explore the sensitivity of the GGGI.
-It would be helpful to understand "what Frame 12 is"  if the authors should explain this between " Figure 7 illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries" and "Frame 12 shows the dot plot of the original index values sorted from highest to lowest", **\textcolor{purple}{Added.}**
+It would be helpful to understand "what Frame 12 is"  if the authors should explain this between " Figure 7 illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries" and "Frame 12 shows the dot plot of the original index values sorted from highest to lowest",
 
-- Figure 8 on page 27 also shows the relatively wide confidence interval around 2004. It would be great if the author mentioned it, too. **\textcolor{purple}{Added.}**
+**\textcolor{purple}{Frame 12 is described in more detail now, and additional explanation of the information being communicated by the figure is provided.}**
+
+- Figure 8 on page 27 also shows the relatively wide confidence interval around 2004. It would be great if the author mentioned it, too.
+
+**\textcolor{purple}{Added. This wider confidence interval is interesting and corresponds to the millennium drought in Australia.}**
 
 - reference for "SPEI package" on page 4 should be (Beguería and Vicente-Serrano 2017),
-and reference for "the Standardized Precipitation-Evapotranspiration Index (SPEI)" on page 12 should be (Vicente-Serrano, Beguería, and López-Moreno 2010) **\textcolor{purple}{Fixed}**.
+and reference for "the Standardized Precipitation-Evapotranspiration Index (SPEI)" on page 12 should be (Vicente-Serrano, Beguería, and López-Moreno 2010)
+
+**\textcolor{purple}{Fixed}**.

### 2024-04-16 tidyindex.qmd
tidied intro and conclusions

@@
 
-<!-- - Define what is an index, what is not -->
-
 <!-- - What is the challenges with current index construction -->
@@
 In social sciences, indexes are used to monitor human development, gender equity, or university quality.
+
+<!-- what is the gap - why is this statistics -->
 The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared.
+This violates *tidy principles* [@tidyverse] which makes it difficult to investigate the nature of calculations through a principled data pipeline.
+It also violates statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric [@???Spiegelhalter, someone else].
 
+<!-- - Define what is an index, what is not -->
 To construct an index, experts typically start by defining a concept of interest that requires measurement.
@@
 
-To be able to test different decision choices systematically for an index, the index needs to be broken down into its fundamental building blocks to analyze the contribution and effect of each component.
-We call this process of breaking the index construction into different steps the index pipeline.
+To be able to test different decision choices for an index, systematically and statistically, the index needs to be broken down into its fundamental building blocks to analyze the contribution and effect of each component.
+We call this process the *index pipeline*, which are the steps of the data analysis pipeline for the index construction.
 Such decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes, calculating index uncertainty, and assessing index robustness.
@@
 
-The paper introduces a novel data pipeline for constructing and analyzing indexes within the tidy framework.
-The pipeline comprises nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
-The proposed pipeline is implemented as an R package, `tidyindex`, available on CRAN.
+The paper introduces a tidy data pipeline for constructing and analyzing indexes.
+It has nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
 
-One of the contributions of our work is its emphasis on aspects often overlooked in index construction: uncertainty quantification and sensitivity.
-The pipeline proposed in the paper allows for confidence interval calculation and sensitivity analysis to be incorporated in the index construction process, for indexes across various domains.
-These is illustrated through the three examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).
+This work addresses aspects often overlooked in index construction: uncertainty quantification and sensitivity.
+The pipeline supports confidence interval calculation and sensitivity analysis to be incorporated in the index construction process.
+It should encourage better statistical practice wherever indexes are used as critical tools in natural and social sciences.
+
+The pipeline approach is illustrated with examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).
 While indexes are widely constructed and used to inform decision-making, such as in resource allocation and policy making worldwide, little effort is made to justify the choices made during index construction or to experiment with alternatives while providing means for reproducibility.
-This work highlights the importance of carefully constructing statistical indexes and introduces novel computational methods for studying and calculating statistical indexes and their uncertainties.
+This work highlights the importance of carefully constructing statistical indexes and introduces new computational methods for studying and calculating statistical indexes and their uncertainties.
 
-One future development of the work is to extend the tidyindex framework to support index calculation on commonly used domain-specific data formats, for example, NetCDF for climate indexes.
+There are many potential directions for the development of the work.
+Computationally, the tidyindex framework could be extended to support other commonly used data formats, for example, NetCDF for climate indexes.
+Conceptually, extending the examples to re-express additional common-practice indexes in the pipeline structure will help broader adoption, and further test that the framework can indeed accommodate any and all possible indexes.
 
@@
 
+The steps of this pipeline are available in the R package, `tidyindex`, available on CRAN.
+The source code for reproducing the work reported in this paper are in Supplementary Materials and can be found at: <https://github.com/huizezhang-sherry/paper-tidyindex>.
+
 This work is funded by a Commonwealth Scientific and Industrial Research Organisation (CSIRO) Data61 Scholarship.
@@
 The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
-The source code for reproducing this paper can be found at: <https://github.com/huizezhang-sherry/paper-tidyindex>.
 

### 2024-04-16 tidyindex.tex
tidied intro and conclusions

@@
 \end{keywords}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi
 
 \setstretch{2}
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 social sciences, indexes are used to monitor human development, gender
-equity, or university quality. The problem is that every index is
-developed in its own unique way, by different researchers or
-organizations, and often indexes designed for the same purpose cannot
-easily be compared.
+equity, or university quality.
+
+The problem is that every index is developed in its own unique way, by
+different researchers or organizations, and often indexes designed for
+the same purpose cannot easily be compared. This violates \emph{tidy
+principles} (Wickham et al. 2019a) which makes it difficult to
+investigate the nature of calculations through a principled data
+pipeline. It also violates statistical principles, where quantifying and
+understanding uncertainty are essential to deciding on a best measure or
+metric {[}@???Spiegelhalter, someone else{]}.
 
@@
 
-To be able to test different decision choices systematically for an
-index, the index needs to be broken down into its fundamental building
-blocks to analyze the contribution and effect of each component. We call
-this process of breaking the index construction into different steps the
-index pipeline. Such decomposition of index components provides the
-means to standardize index construction via a pipeline and offers
-benefits for comparing among indexes, calculating index uncertainty, and
-assessing index robustness.
+To be able to test different decision choices for an index,
+systematically and statistically, the index needs to be broken down into
+its fundamental building blocks to analyze the contribution and effect
+of each component. We call this process the \emph{index pipeline}, which
+are the steps of the data analysis pipeline for the index construction.
+Such decomposition of index components provides the means to standardize
+index construction via a pipeline and offers benefits for comparing
+among indexes, calculating index uncertainty, and assessing index
+robustness.
 
@@
 applicable generally to any data analysis workflow. More recently, the
-tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
+tidyverse suite (Wickham et al. 2019b) takes the pipeline approach for
 general-purpose data wrangling and has gained popularity within the R
@@
 
-\end{figure}
-
-\begingroup\fontsize{11}{13}\selectfont
-
-\hypertarget{tbl-gggi-weights}{}
-\begin{longtable}[t]{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
-\caption{\label{tbl-gggi-weights}Weights for the two applications of dimension reduction to compute the
-Global Gender Gap Index. V-wgt is used to compute four new variables
-from the original 14. These are then equally combined to get the final
-index value. }\tabularnewline
-
+\end{figure}%
+
+\begin{table}
+
+\centering{
+
+\centering\begingroup\fontsize{11}{13}\selectfont
+
+\begin{tabular}{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
 \toprule
@@
 Professional and technical workers & 0.121 &  &  & 0.030\\
-\addlinespace
+\hline
 Literacy rate & 0.191 & Education & 0.25 & 0.048\\
@@
 Enrolment in tertiary education & 0.121 &  &  & 0.030\\
+\hline
 Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
-\addlinespace
 Healthy life expectancy & 0.307 &  &  & 0.077\\
+\hline
 Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
@@
 \bottomrule
-\end{longtable}
+\end{tabular}
 \endgroup{}
+
+}
+
+\caption{\label{tbl-gggi-weights}Weights for the two applications of
+dimension reduction to compute the Global Gender Gap Index. V-wgt is
+used to compute four new variables from the original 14. These are then
+equally combined to get the final index value.}
+
+\end{table}%
 
@@
 
-\end{figure}
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-The paper introduces a novel data pipeline for constructing and
-analyzing indexes within the tidy framework. The pipeline comprises nine
-modules including temporal and spatial aggregation, variable
-transformation and combination, distribution fitting, benchmark setting,
-and index communication. The proposed pipeline is implemented as an R
-package, \texttt{tidyindex}, available on CRAN.
-
-One of the contributions of our work is its emphasis on aspects often
-overlooked in index construction: uncertainty quantification and
-sensitivity. The pipeline proposed in the paper allows for confidence
-interval calculation and sensitivity analysis to be incorporated in the
-index construction process, for indexes across various domains. These is
-illustrated through the three examples on drought indexes (SPI and SPEI)
-and the Global Gender Gap Index (GGGI). While indexes are widely
-constructed and used to inform decision-making, such as in resource
-allocation and policy making worldwide, little effort is made to justify
-the choices made during index construction or to experiment with
-alternatives while providing means for reproducibility. This work
+\end{figure}%
+
+\section{Conclusion}\label{conclusion}
+
+The paper introduces a tidy data pipeline for constructing and analyzing
+indexes. It has nine modules including temporal and spatial aggregation,
+variable transformation and combination, distribution fitting, benchmark
+setting, and index communication.
+
+This work addresses aspects often overlooked in index construction:
+uncertainty quantification and sensitivity. The pipeline supports
+confidence interval calculation and sensitivity analysis to be
+incorporated in the index construction process. It should encourage
+better statistical practice wherever indexes are used as critical tools
+in natural and social sciences.
+
+The pipeline approach is illustrated with examples on drought indexes
+(SPI and SPEI) and the Global Gender Gap Index (GGGI). While indexes are
+widely constructed and used to inform decision-making, such as in
+resource allocation and policy making worldwide, little effort is made
+to justify the choices made during index construction or to experiment
+with alternatives while providing means for reproducibility. This work
 highlights the importance of carefully constructing statistical indexes
-and introduces novel computational methods for studying and calculating
+and introduces new computational methods for studying and calculating
 statistical indexes and their uncertainties.
 
-One future development of the work is to extend the tidyindex framework
-to support index calculation on commonly used domain-specific data
-formats, for example, NetCDF for climate indexes.
-
-\hypertarget{acknowledgement}{%
-\section{Acknowledgement}\label{acknowledgement}}
+There are many potential directions for the development of the work.
+Computationally, the tidyindex framework could be extended to support
+other commonly used data formats, for example, NetCDF for climate
+indexes. Conceptually, extending the examples to re-express additional
+common-practice indexes in the pipeline structure will help broader
+adoption, and further test that the framework can indeed accommodate any
+and all possible indexes.
+
+\section{Acknowledgement}\label{acknowledgement}
+
+The steps of this pipeline are available in the R package,
+\texttt{tidyindex}, available on CRAN. The source code for reproducing
+the work reported in this paper are in Supplementary Materials and can
+be found at: \url{https://github.com/huizezhang-sherry/paper-tidyindex}.
 
@@
 UICR0700041-22. The article is created using Quarto (Allaire et al.
-2022) in R (R Core Team 2021). The source code for reproducing this
-paper can be found at:
-\url{https://github.com/huizezhang-sherry/paper-tidyindex}.
-
-\hypertarget{reference}{%
-\section*{Reference}\label{reference}}
+2022) in R (R Core Team 2021).
+
+\section*{Reference}\label{reference}
 \addcontentsline{toc}{section}{Reference}
 
-\hypertarget{refs}{}
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alahacoon_comprehensive_2022}{}}%
+\bibitem[\citeproctext]{ref-alahacoon_comprehensive_2022}
 Alahacoon, Niranga, and Mahesh Edirisinghe. 2022. {``A Comprehensive
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wickham_welcome_2019}{}}%
+\bibitem[\citeproctext]{ref-tidyverse}
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
-D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
-{``Welcome to the {Tidyverse}.''} \emph{Journal of Open Source Software}
+D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019a.
+{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-wickham_tourr_2011}{}}%
+\bibitem[\citeproctext]{ref-wickham_welcome_2019}
+---------, et al. 2019b. {``Welcome to the {Tidyverse}.''} \emph{Journal
+of Open Source Software} 4 (43): 1686.
+\url{https://doi.org/10.21105/joss.01686}.
+
+\bibitem[\citeproctext]{ref-wickham_tourr_2011}
 Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.

### 2024-04-22 review/response.qmd
Minor fixes in the response

@@
 
-Below is a point-by-point explanation of how we have changed the paper.
-The reviewer comments are in black and our responses are in red.
-Reviewers' comments are in black and **\textcolor{purple}{our responses are in red}**.
+Below is a point-by-point explanation of how we have changed the paper. Reviewers' comments are in black and **\textcolor{purple}{our responses are in red}**.
 
@@
 
-**\textcolor{purple}{Fixed. The table layout has been modified accordingly.}**.
+**\textcolor{purple}{Fixed. The table layout has been modified accordingly.}**
 
@@
 
-**\textcolor{purple}{Fixed. The output has been written in a way that all variables are visible now (using dplyr::glimpse()).}**.
+**\textcolor{purple}{Fixed. The output has been written in a way that all variables are visible now (using dplyr::glimpse()).}**
 

### 2024-04-22 tidyindex.tex
include more reproducible materials

@@
 
-\begin{table}
-
-\centering{
-
-\centering\begingroup\fontsize{11}{13}\selectfont
-
-\begin{tabular}{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
+\begingroup\fontsize{11}{13}\selectfont
+
+\begin{longtable}[t]{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
+
 \toprule
@@
 \bottomrule
-\end{tabular}
-\endgroup{}
-
-}
 
@@
 
-\end{table}%
+\tabularnewline
+
+\end{longtable}
+
+\endgroup{}
 

### 2024-04-22 scripts/example6-1.R

@@ -1,0 +1,67 @@
+library(dplyr)
+library(tidyindex)
+library(tsibble)
+library(patchwork)
+library(ozmaps)
+library(rmapshaper)
+library(colorspace)
+load(here::here("data/idx.rda"))
+
+dplyr::glimpse(idx)
+queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
+  rmapshaper::ms_simplify(keep = 0.02)
+queensland_map %>%
+  ggplot() +
+  geom_sf(fill = "transparent", color = "grey50", linewidth = 0.4) +
+  geom_point(
+    data = idx %>%
+      filter(.scale == 12) %>%
+      filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
+               ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )),
+    aes(x = long, y = lat, color = .index)) +
+  colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
+  theme_void() +
+  facet_wrap(vars(ym), ncol = 6)
+
+texas <- idx %>%
+  filter(name == "TEXAS POST OFFICE", .idx == "spei") %>% distinct(long, lat)
+p1 <- queensland_map %>%
+  ggplot() +
+  geom_sf(fill = "transparent") +
+  geom_point(data = queensland %>%
+               distinct(id, long, lat, name),
+             aes(x = long, y = lat)) +
+  geom_point(data = texas, aes(x = long, y = lat),
+             color = "orange", shape = 18, fill = "orange", size = 4) +  
+  theme_void()
+
+
+p2 <- idx %>%
+  filter(name == "TEXAS POST OFFICE", .idx == "spei") %>%
+  ggplot(aes(x = ym, y = .index, color = .dist, group = .dist)) +
+  geom_line() +
+  theme_benchmark() +
+  facet_wrap(vars(.scale), labeller = label_both, ncol = 1) +
+  scale_x_yearmonth(breaks = "4 year", date_labels = "%Y") +
+  scale_color_brewer(palette = "Dark2", name = "Distribution") + 
+  xlab("Year") + 
+  ylab("Index")
+
+(p1 | p2)  + 
+  patchwork::plot_annotation(tag_levels = "a") + 
+  patchwork::plot_layout(guides = "collect", widths = c(1, 4))  &
+  theme(legend.position = "bottom") 
+
+
[…]

### 2024-04-22 scripts/example6-2.R

@@ -1,0 +1,65 @@
+library(tidyindex)
+library(dplyr)
+library(patchwork)
+library(tourr)
+gggi %>% 
+  init(id = country) %>%
+  add_paras(gggi_weights, by = "variable") %>% 
+  dimension_reduction(
+    index_new = aggregate_linear(
+      ~labour_force_participation:years_with_female_head_of_state,
+      weight = weight)) 
+
+frames <- c("001", "006", "012", "018", "024", "029") 
+ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
+rl <- lapply(ani, png::readPNG)
+gl <-  lapply(rl, grid::rasterGrob)
+wrap_plots(gl, ncol = 3)
+
+
+## code to generate the animation frames: 
+dt <- tidyindex::gggi %>% 
+  dplyr::select(country: political_empowerment) %>%
+  filter(between(index, 0.72, 0.729)  | index < 0.6 | index >= 0.85) %>%
+  arrange(index) %>% 
+  mutate(country = ifelse(country == "Iran (Islamic Republic of)", "Iran", country))
+colnames(dt) <- c(colnames(dt)[1:4], "economy", "education", "health", "politics")
+
+w_proj2idx <- function(w){w/sum(w)}
+w_idx2proj <- function(w){w/(sqrt(sum(w^2))) %>% matrix(nrow = length(w))}
+find_next <- function(matrix, angle = 0.4){
+  start <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1)
+  tourr:::step_angle(tourr:::geodesic_info(start, matrix), angle = angle)
+}
+b <- array(dim = c(4, 1, 3))
+b[,,1] <- matrix(c(0.3, 0.3, 0.3, 0.1), ncol = 1) %>% find_next()
+b[,,2] <- matrix(c(0.25, 0.25, 0.25, 0.25), ncol = 1) %>% w_idx2proj()
+b[,,3] <- matrix(c(0.1, 0.1, 0.1, 0.7), ncol= 1) %>% find_next(angle = 0.6)
+
+render(
+  dt[,5:8], planned2_tour(b), dev = "png",
+  display_idx(half_range = 2,cex = 6, label_cex = 4, col = "black", 
+              panel_height_ratio = c(4,1), label_x_pos = 0.1, 
+              axis_bar_lwd = c(rep(7, 3), 9), label_col = "grey70",
+              axis_bar_col = c(rep("#000000", 3), "red"),
[…]

### 2024-04-22 scripts/example6-3.R

@@ -1,0 +1,41 @@
+library(tidyindex)
+library(dplyr)
+library(ggplot2)
+library(lmomco)
+library(tsibble)
+res <- queensland %>% 
+  filter(name == "TEXAS POST OFFICE") %>% 
+  mutate(month = lubridate::month(ym)) |>
+  init(id = id, time = ym, group = month) |> 
+  temporal_aggregate(.agg = temporal_rolling_window(prcp, scale = 24)) |> 
+  distribution_fit(.fit = dist_gamma(var = ".agg", method = "lmoms", .n_boot = 100)) |>
+  tidyindex::normalise(.index = norm_quantile(.fit))
+
+res_band <- res$data %>% 
+  group_by(ym, id, name) %>% 
+  summarise(
+    q025 = quantile(.index, 0.025),
+    q10 = quantile(.index, 0.1),
+    q50 = quantile(.index, 0.5),
+    q90 = quantile(.index, 0.9),
+    q975 = quantile(.index, 0.975)
+  ) %>% 
+  ungroup()
+
+
+# the sampling uncertainty is quite large here
+# and this is mainly due to we only have 30 year worth of data
+# each month fit has only 30 points
+res_band %>% 
+  ggplot(aes(x = ym)) + 
+  geom_hline(yintercept = -2, linetype = "dashed", color = "grey50") + 
+  #geom_text(aes(x = loc, y = -2.7, label = "SPI = -2"), color = "grey50") + 
+  geom_ribbon(aes(ymin = q025, ymax = q975), fill = "steelblue1", alpha = 0.5) + 
+  geom_ribbon(aes(ymin = q10, ymax = q90), fill = "steelblue4", alpha = 0.5) + 
+  geom_line(aes(y = q50)) + 
+  scale_x_yearmonth(name = "Year", date_break = "2 years", date_label = "%Y") +
+  theme_bw() + 
+  facet_wrap(vars(name), ncol = 1) + 
+  theme(panel.grid = element_blank(), 
+        legend.position = "bottom") + 
+  ylab("SPI")

### 2024-04-23 tidyindex.qmd
add references and connect sentences in intro

@@
 The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared.
-This violates *tidy principles* [@tidyverse] which makes it difficult to investigate the nature of calculations through a principled data pipeline.
-It also violates statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric [@???Spiegelhalter, someone else].
+This echoes an issue highlighted by @donoho_50_2017, where varying data analysis workflows can resulted in different, or even reversed, conclusions when the identical data is analyzed in medicine research.
+In Greater Data Science (GDS) 6: Science About Data Science, Donoho envisions future research on structuring unified workflow to address methodological variation across studies in data science.
+It also violates statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric [@efron_bootstrap_1979].
 

### 2024-04-23 tidyindex.tex
add references and connect sentences in intro

@@
 different researchers or organizations, and often indexes designed for
-the same purpose cannot easily be compared. This violates \emph{tidy
-principles} (Wickham et al. 2019a) which makes it difficult to
-investigate the nature of calculations through a principled data
-pipeline. It also violates statistical principles, where quantifying and
-understanding uncertainty are essential to deciding on a best measure or
-metric {[}@???Spiegelhalter, someone else{]}.
+the same purpose cannot easily be compared. This echoes an issue
+highlighted by Donoho (2017), where varying data analysis workflows can
+resulted in different, or even reversed, conclusions when the identical
+data is analyzed in medicine research. In Greater Data Science (GDS) 6:
+Science About Data Science, Donoho envisions future research on
+structuring unified workflow to address methodological variation across
+studies in data science. It also violates statistical principles, where
+quantifying and understanding uncertainty are essential to deciding on a
+best measure or metric (Efron 1979).
 
@@
 applicable generally to any data analysis workflow. More recently, the
-tidyverse suite (Wickham et al. 2019b) takes the pipeline approach for
+tidyverse suite (Wickham et al. 2019) takes the pipeline approach for
 general-purpose data wrangling and has gained popularity within the R
@@
 
+\bibitem[\citeproctext]{ref-donoho_50_2017}
+Donoho, David. 2017. {``50 {Years} of {Data} {Science}.''} \emph{Journal
+of Computational and Graphical Statistics} 26 (4): 745--66.
+\url{https://doi.org/10.1080/10618600.2017.1384734}.
+
+\bibitem[\citeproctext]{ref-efron_bootstrap_1979}
+Efron, B. 1979. {``Bootstrap {Methods}: {Another} {Look} at the
+{Jackknife}.''} \emph{The Annals of Statistics} 7 (1): 1--26.
+\url{https://doi.org/10.1214/aos/1176344552}.
+
 \bibitem[\citeproctext]{ref-fisher1936use}
@@
 
-\bibitem[\citeproctext]{ref-tidyverse}
+\bibitem[\citeproctext]{ref-wickham_welcome_2019}
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
-D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019a.
-{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software}
+D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
+{``Welcome to the {Tidyverse}.''} \emph{Journal of Open Source Software}
 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
-
-\bibitem[\citeproctext]{ref-wickham_welcome_2019}
----------, et al. 2019b. {``Welcome to the {Tidyverse}.''} \emph{Journal
-of Open Source Software} 4 (43): 1686.
-\url{https://doi.org/10.21105/joss.01686}.
 

### 2024-04-23 review/response.qmd
modified response

@@
 
-**\textcolor{purple}{To be addressed}**
+**\textcolor{purple}{We have made changes to the introduction and to the conclusions being more specific about the methodological contributions of this work.}**
 
@@
 
-**\textcolor{purple}{We have highlighted the methodology contributions this framework makes in the conclusion: While indexes are widely constructed and used to inform decision-making, such as on resource allocation, little effort is made to justify choices made during indexes construction and to experiment with alternatives.
-The pipeline proposed in the paper allows confidence interval to be calculated in the pipeline and sensitivity analysis to be incorporated in the index construction process.}**
+**\textcolor{purple}{We are quite disappointed to see the language in this review: "A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." If the reviewer had read the journal scope they will find "The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area, also.
+We have added language to the introduction and to the conclusions that hopefully helps see these contributions.}**
 

### 2024-04-23 review/response.qmd
changes to intro and conclusions to address the big criticism

@@
 
-**\textcolor{purple}{We are quite disappointed to see the language in this review: "A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." If the reviewer had read the journal scope they will find "The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area, also.
+**\textcolor{purple}{We are quite disappointed to see the language in this review: ``A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." Reading the journal scope would have informed them that ``The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area.
 We have added language to the introduction and to the conclusions that hopefully helps see these contributions.}**

### 2024-04-23 tidyindex.qmd
changes to intro and conclusions to address the big criticism

--- 2024-04-16 tidyindex.qmd
@@
 The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared.
-This violates *tidy principles* [@tidyverse] which makes it difficult to investigate the nature of calculations through a principled data pipeline.
-It also violates statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric [@???Spiegelhalter, someone else].
+This echoes an issue raised in @donoho_50_2017, that different data analysts might arrive at different conclusions despite using the same data.
+This is especially pertinent to index use which affects important decisions such as in natural disaster prevention, economic interventions, resource allocation or human development.
+It is primarily due to a lack of standards in data analysis workflow.
+@donoho_50_2017 called for research on structuring unified workflow to address methodological variation across studies in data science.
+Current practices also violate statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric, for example by incorporating bootstrap confidence intervals [@efron_bootstrap_1979].
+There has been considerable research in tidying up routine data analyses [XXX tidyr, dplyr, tidymodels, ..., tsibble, cubble] that turn ideas into software quickly and faithfully, as envisioned by @chambers.
+Index development and use needs tidying.
+<!--In Greater Data Science (GDS) 6: Science About Data Science, Donoho envisions future research on structuring unified workflow to address methodological variation across studies in data science.
+-->.
 
@@
 It has nine modules including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
+This addresses statistical principles absent from current index definitions: uncertainty quantification and sensitivity.
+The tidyindex framework should encourage better statistical practice wherever indexes are used as critical tools in natural and social sciences.
 
-This work addresses aspects often overlooked in index construction: uncertainty quantification and sensitivity.
-The pipeline supports confidence interval calculation and sensitivity analysis to be incorporated in the index construction process.
-It should encourage better statistical practice wherever indexes are used as critical tools in natural and social sciences.
-
-The pipeline approach is illustrated with examples on drought indexes (SPI and SPEI) and the Global Gender Gap Index (GGGI).
-While indexes are widely constructed and used to inform decision-making, such as in resource allocation and policy making worldwide, little effort is made to justify the choices made during index construction or to experiment with alternatives while providing means for reproducibility.
-This work highlights the importance of carefully constructing statistical indexes and introduces new computational methods for studying and calculating statistical indexes and their uncertainties.
+Several examples are shown illustrating usage.
+For the drought indexes (SPI and SPEI) we showed how multiple indexes can be computed with a range of parameter choices, and compared across space and time.
+We showed how bootstrap confidence intervals can be readily computed and plotted to assess uncertainty about the index values, and how it may be used to make better decisions on drought declarations.
+The Global Gender Gap Index (GGGI) was used to illustrate how choices in dimension reduction can radically affect index values and country rankings.
+This also illustrates how the pipeline feeds nicely into advanced interactive graphics.
 
 There are many potential directions for the development of the work.
-Computationally, the tidyindex framework could be extended to support other commonly used data formats, for example, NetCDF for climate indexes.
+Computationally, the tidyindex framework could be extended to support other data formats, like NetCDF for climate indexes.
 Conceptually, extending the examples to re-express additional common-practice indexes in the pipeline structure will help broader adoption, and further test that the framework can indeed accommodate any and all possible indexes.

### 2024-04-23 tidyindex.tex
changes to intro and conclusions to address the big criticism

--- 2024-04-16 tidyindex.tex
@@
 different researchers or organizations, and often indexes designed for
-the same purpose cannot easily be compared. This violates \emph{tidy
-principles} (Wickham et al. 2019a) which makes it difficult to
-investigate the nature of calculations through a principled data
-pipeline. It also violates statistical principles, where quantifying and
-understanding uncertainty are essential to deciding on a best measure or
-metric {[}@???Spiegelhalter, someone else{]}.
+the same purpose cannot easily be compared. This echoes an issue raised
+in Donoho (2017), that different data analysts might arrive at different
+conclusions despite using the same data. This is especially pertinent to
+index use which affects important decisions such as in natural disaster
+prevention, economic interventions, resource allocation or human
+development. It is primarily due to a lack of standards in data analysis
+workflow. Donoho (2017) called for research on structuring unified
+workflow to address methodological variation across studies in data
+science. Current practices also violate statistical principles, where
+quantifying and understanding uncertainty are essential to deciding on a
+best measure or metric, for example by incorporating bootstrap
+confidence intervals (Efron 1979). There has been considerable research
+in tidying up routine data analyses {[}XXX tidyr, dplyr, tidymodels,
+\ldots, tsibble, cubble{]} that turn ideas into software quickly and
+faithfully, as envisioned by (\textbf{chambers?}). Index development and
+use needs tidying. .
 
@@
 variable transformation and combination, distribution fitting, benchmark
-setting, and index communication.
-
-This work addresses aspects often overlooked in index construction:
-uncertainty quantification and sensitivity. The pipeline supports
-confidence interval calculation and sensitivity analysis to be
-incorporated in the index construction process. It should encourage
-better statistical practice wherever indexes are used as critical tools
-in natural and social sciences.
-
-The pipeline approach is illustrated with examples on drought indexes
-(SPI and SPEI) and the Global Gender Gap Index (GGGI). While indexes are
-widely constructed and used to inform decision-making, such as in
-resource allocation and policy making worldwide, little effort is made
-to justify the choices made during index construction or to experiment
-with alternatives while providing means for reproducibility. This work
-highlights the importance of carefully constructing statistical indexes
-and introduces new computational methods for studying and calculating
-statistical indexes and their uncertainties.
+setting, and index communication. This addresses statistical principles
+absent from current index definitions: uncertainty quantification and
+sensitivity. The tidyindex framework should encourage better statistical
+practice wherever indexes are used as critical tools in natural and
+social sciences.
+
+Several examples are shown illustrating usage. For the drought indexes
+(SPI and SPEI) we showed how multiple indexes can be computed with a
+range of parameter choices, and compared across space and time. We
+showed how bootstrap confidence intervals can be readily computed and
+plotted to assess uncertainty about the index values, and how it may be
+used to make better decisions on drought declarations. The Global Gender
+Gap Index (GGGI) was used to illustrate how choices in dimension
+reduction can radically affect index values and country rankings. This
+also illustrates how the pipeline feeds nicely into advanced interactive
+graphics.
 
@@
 Computationally, the tidyindex framework could be extended to support
-other commonly used data formats, for example, NetCDF for climate
-indexes. Conceptually, extending the examples to re-express additional
-common-practice indexes in the pipeline structure will help broader
-adoption, and further test that the framework can indeed accommodate any
-and all possible indexes.
+other data formats, like NetCDF for climate indexes. Conceptually,
+extending the examples to re-express additional common-practice indexes
+in the pipeline structure will help broader adoption, and further test
+that the framework can indeed accommodate any and all possible indexes.
 

### 2024-04-23 tidyindex.qmd
quote

@@
 Current practices also violate statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric, for example by incorporating bootstrap confidence intervals [@efron_bootstrap_1979].
-There has been considerable research in tidying up routine data analyses [XXX tidyr, dplyr, tidymodels, ..., tsibble, cubble] that turn ideas into software quickly and faithfully, as envisioned by @chambers.
+There has been considerable research in tidying up routine data analyses [XXX tidyr, dplyr, tidymodels, ..., tsibble, cubble] that "turn ideas into software quickly and faithfully", as envisioned by @chambers.
 Index development and use needs tidying.

### 2024-04-23 review/response.qmd
Update first paragraph
Update answer to reviewers

@@
 ```
-
-Below is a point-by-point explanation of how we have changed the paper. Reviewers' comments are in black and **\textcolor{purple}{our responses are in red}**.
+We would like to thank the reviewers and editor for taking the time to review our paper and provide constructive comments.
+Below is a point-by-point explanation of how we have changed the paper.
+Reviewers' comments are in black and **\textcolor{purple}{our responses are in red}**.
 
@@
 
-**\textcolor{purple}{We have made changes to the introduction and to the conclusions being more specific about the methodological contributions of this work.}**
+**\textcolor{purple}{We have made changes to the introduction and to the conclusions being more specific about the methodological and computational contributions of this work.}**
 
@@
 
-**\textcolor{purple}{Fixed.
-We were expecting that readers worked from the vignettes folder.
-The problem was that the demo folder had some irrelevant code.
-We've removed these now.
-We have also re-organised the code for the paper as follows.}**
+**\textcolor{purple}{ This issue is now fixed in the Demo folder (the vignette folder worked fine) and we have also re-organised the code for the paper as follows:}**
 
-- **\textcolor{purple}{The main file tidyindex.R, reproduces the content in the paper from saved data objects, available in the data folder.}** 
-- **\textcolor{purple}{New scripts have been provided in the scripts folder which completely generate each of the examples.
-There is a README.md, that instructs the reader how to set their environment up to reproduce all of the analyses.}**
+- **\textcolor{purple}{The main file tidyindex.R, can be used to reproduce the paper from saved data objects, available in the data folder.}** 
+- **\textcolor{purple}{New scripts can be found in the 'scripts' folder, which generate each of the examples.
+The README.md file has been updated and it contains instructions on how to set up the R environment to reproduce all of the analyses in the paper.}**
 - **\textcolor{purple}{The GitHub repo is listed in the Acknowledgements section, and any further updates needed for future versions to run will be made there.}**
@@
 
-**\textcolor{purple}{Fixed. The table layout has been modified accordingly.}**
+**\textcolor{purple}{The table layout has been modified accordingly.}**
 
@@
 
-**\textcolor{purple}{Fixed. The output has been written in a way that all variables are visible now (using dplyr::glimpse()).}**
+**\textcolor{purple}{ The output is now written in a way that all variables are visible (availble from dplyr::glimpse()).}**
 
@@
 
-**\textcolor{purple}{Frame 12 is described in more detail now, and additional explanation of the information being communicated by the figure is provided.}**
+**\textcolor{purple}{Frame 12 is now described in more detail, with additional explanations provided regarding the information conveyed by the figure.}**
 
@@
 
-**\textcolor{purple}{Added. This wider confidence interval is interesting and corresponds to the millennium drought in Australia.}**
+**\textcolor{purple}{The following has been added in the text: This wider confidence interval is interesting and corresponds to the millennium drought in Australia.}**
 
@@
 
-**\textcolor{purple}{Fixed}**.
+**\textcolor{purple}{The reference has been fixed}**.

### 2024-04-24 review/response.qmd
Update response to rev

@@
 
-**\textcolor{purple}{We are quite disappointed to see the language in this review: ``A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." Reading the journal scope would have informed them that ``The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area.
+**\textcolor{purple}{It is a bit disappointed to see the following comment ``A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." We disagree with that statement based on the journal scope:  "The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area.
 We have added language to the introduction and to the conclusions that hopefully helps see these contributions.}**

### 2024-04-24 review/response.qmd
Update typo in response

@@
 
-**\textcolor{purple}{It is a bit disappointed to see the following comment ``A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." We disagree with that statement based on the journal scope:  "The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area.
+**\textcolor{purple}{It is a bit disappointing to see the following comment ``A perusal of recent JCGS papers shows that a novel statistical component is needed for publication in addition to the software." We disagree with that statement based on the journal scope:  "The Journal of Computational and Graphical Statistics (JCGS) presents the very latest techniques on improving and extending the use of computational and graphical methods in statistics and data analysis." The contributions that this paper makes are in computational statistics for an especially prevalent data analysis activity, that also provides potential for making better graphics in this area.
 We have added language to the introduction and to the conclusions that hopefully helps see these contributions.}**

### 2024-04-24 tidyindex.qmd
add missing references

--- 2024-04-23 tidyindex.qmd
@@
     linestretch: 2
+    fontsize: 12pt
 pdf-engine: pdflatex
@@
 Current practices also violate statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric, for example by incorporating bootstrap confidence intervals [@efron_bootstrap_1979].
-There has been considerable research in tidying up routine data analyses [XXX tidyr, dplyr, tidymodels, ..., tsibble, cubble] that turn ideas into software quickly and faithfully, as envisioned by @chambers.
+There has been considerable research in tidying up routine data analyses [@wickham_tidy_2014; @wickham_split-apply-combine_2011; @tidymodels; @wang2020; @zhang2024cubble] that "turn ideas into software quickly and faithfully", as envisioned by @chambers1998programming.
 Index development and use needs tidying.
 <!--In Greater Data Science (GDS) 6: Science About Data Science, Donoho envisions future research on structuring unified workflow to address methodological variation across studies in data science.
--->.
+-->
 
@@
 
-# Reference
+# References

### 2024-04-24 tidyindex.tex
quote
add missing references

@@
 \documentclass[
+  12pt,
 ]{interact}
@@
 confidence intervals (Efron 1979). There has been considerable research
-in tidying up routine data analyses {[}XXX tidyr, dplyr, tidymodels,
-\ldots, tsibble, cubble{]} that turn ideas into software quickly and
-faithfully, as envisioned by (\textbf{chambers?}). Index development and
-use needs tidying. .
+in tidying up routine data analyses (Wickham 2014, 2011; Kuhn and Silge
+2022; Wang, Cook, and Hyndman 2020; Zhang et al. 2024) that ``turn ideas
+into software quickly and faithfully'', as envisioned by Chambers
+(1998). Index development and use needs tidying.
 
@@
 a noticeable example of which is \texttt{tidymodels} for building
-machine learning models (Kuhn and Wickham 2020). To create a tidy
-workflow tailored to a specific domain, developers first need to
-identify the fundamental building blocks to create a workflow. These
-components are then implemented as modules, which can be combined to
-form the pipeline. For example, in supervised machine learning models,
-steps such as data splitting, model training, and model evaluation are
-commonly used in most workflow. In the \texttt{tidymodels}, these steps
-are correspondingly implemented as package \texttt{rsample},
+machine learning models (Kuhn and Silge 2022). To create a tidy workflow
+tailored to a specific domain, developers first need to identify the
+fundamental building blocks to create a workflow. These components are
+then implemented as modules, which can be combined to form the pipeline.
+For example, in supervised machine learning models, steps such as data
+splitting, model training, and model evaluation are commonly used in
+most workflow. In the \texttt{tidymodels}, these steps are
+correspondingly implemented as package \texttt{rsample},
 \texttt{parsnip}, and \texttt{yardstick}, agnostic to the specific model
@@
 
-\section*{Reference}\label{reference}
-\addcontentsline{toc}{section}{Reference}
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 
@@
 
+\bibitem[\citeproctext]{ref-chambers1998programming}
+Chambers, John M. 1998. \emph{Programming with Data: A Guide to the s
+Language}. Berlin, Heidelberg: Springer-Verlag.
+
 \bibitem[\citeproctext]{ref-donoho_50_2017}
@@
 \bibitem[\citeproctext]{ref-tidymodels}
-Kuhn, Max, and Hadley Wickham. 2020. \emph{Tidymodels: A Collection of
-Packages for Modeling and Machine Learning Using Tidyverse Principles.}
-\url{https://www.tidymodels.org}.
+Kuhn, Max, and Julia Silge. 2022. \emph{Tidy Modeling with r}. "
+O'Reilly Media, Inc.".
 
@@
 
+\bibitem[\citeproctext]{ref-wang2020}
+Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. {``A New Tidy Data
+Structure to Support Exploration and Modeling of Temporal Data.''}
+\emph{Journal of Computational and Graphical Statistics} 29 (3):
+466--78. \url{https://doi.org/10.1080/10618600.2019.1695624}.
+
+\bibitem[\citeproctext]{ref-wickham_split-apply-combine_2011}
+Wickham, Hadley. 2011. {``The {Split}-{Apply}-{Combine} {Strategy} for
+{Data} {Analysis}.''} \emph{Journal of Statistical Software} 40 (April):
+1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+
 \bibitem[\citeproctext]{ref-wickham_tidy_2014}
-Wickham, Hadley. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
+---------. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
 Software} 59 (September): 1--23.
@@
 
+\bibitem[\citeproctext]{ref-zhang2024cubble}
+Zhang, H. Sherry, Dianne Cook, Ursula Laa, Nicolas Langrené, and
+Patricia Menéndez. 2024. {``Cubble: An r Package for Organizing and
+Wrangling Multivariate Spatio-Temporal Data.''}
+\url{https://arxiv.org/abs/2205.00259}.
+
 \end{CSLReferences}

### 2024-04-24 tidyindex.qmd
fix institution as requested by Patricia

@@
       - name: University of Texas at Austin
-        department: Department of Statistics and Data Science
+        department: Department of Statistics and Data Sciences
         address: Welch 5.216, 105E 24th Street, Stop G3800
@@
     affiliation: 
-      - name: Monash University
-        department: Department of Econometrics and Business Statistics
-        address: 29 Ancora Imparo Way, Clayton
-        city: Melbourne, Victoria
-        country: Australia
       - name: University of Melbourne

### 2024-04-24 tidyindex.tex
fix institution as requested by Patricia

@@
 Langrené$\textsuperscript{4}$~\orcidlink{0000-0001-7601-4618}, Patricia
-Menéndez$\textsuperscript{2,5}$~\orcidlink{0000-0003-0701-6315}}
+Menéndez$\textsuperscript{5}$~\orcidlink{0000-0003-0701-6315}}
 
@@
 \textsuperscript{1} Department of Statistics and Data
-Science, University of Texas at Austin, Austin,
+Sciences, University of Texas at Austin, Austin,
 Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business

### 2024-04-24 scripts/example6-1.R
making sure the scripts run

@@ -5,6 +5,8 @@ library(patchwork)
 library(ozmaps)
 library(rmapshaper)
 library(colorspace)
+library(ggplot2)
+library(lubridate)
 load(here::here("data/idx.rda"))
 
 dplyr::glimpse(idx)

### 2024-04-24 tidyindex.qmd
Small changes and fixes

@@
 It is primarily due to a lack of standards in data analysis workflow.
-@donoho_50_2017 called for research on structuring unified workflow to address methodological variation across studies in data science.
+@donoho_50_2017 called for research on structuring a unified workflow to address methodological variation across studies in data science.
 Current practices also violate statistical principles, where quantifying and understanding uncertainty are essential to deciding on a best measure or metric, for example by incorporating bootstrap confidence intervals [@efron_bootstrap_1979].
@@
 To be able to test different decision choices for an index, systematically and statistically, the index needs to be broken down into its fundamental building blocks to analyze the contribution and effect of each component.
-We call this process the *index pipeline*, which are the steps of the data analysis pipeline for the index construction.
-Such decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes, calculating index uncertainty, and assessing index robustness.
+We call this process the *index pipeline*, which are the steps of the data analysis pipeline for index construction.
+Such a decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes, calculating index uncertainty, and assessing index robustness.
+In addition it provides an unambiguous documentation for the index definition, facilitating reproducibility of results.
 
@@
 For example, in supervised machine learning models, steps such as data splitting, model training, and model evaluation are commonly used in most workflow.
-In the `tidymodels`, these steps are correspondingly implemented as package `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen.
+In the `tidymodels`, these steps are correspondingly implemented as packages `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen.
 The uniform interface in tidymodels frees analysts from recalling model-specific syntax for performing the same operation across different models, increasing the efficiency to work with different models simultaneously.
@@
 -   allows for rapidly processing new data to check how different features, like outliers, might affect the index value.
--   provides the capacity to measure uncertainty by computing confidence intervals from multiple samples as generated by bootstrapping to original data.
+-   provides the capacity to measure uncertainty by computing confidence intervals from multiple samples as generated by bootstrapping the original data.
 -   enables systematic comparison of surrogate indexes designed to measure the same phenomenon.
@@
 Unlike variable transformation, scaling maintains the distributional shape of the variable.
-It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$ where $\alpha$ and $\gamma$ are constants .
+It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$ where $\alpha$ and $\gamma$ are constants.
 In the Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.
@@
 
-The accompanied package, tidyindex, is not intended to offer an exhaustive implementation for all indexes across every domains.
+The tidyindex package is not intended to offer an exhaustive implementation for all indexes across all domains.
 Instead, it provides a realization of the pipeline framework proposed in the paper.
@@
 This example showcases the basic calculation of indexes with different parameter specifications within the pipeline framework.
-The dataset used in this example is available in the `tidyindex` package as `queensland` and blow prints the first few rows of the data:
+The dataset used in this example is available in the `tidyindex` package as `queensland` and below we show the first few rows of the data:
 
@@
 @fig-idx-tour illustrates doing sensitivity analysis for GGGI, for a subset of 16 countries.
-It presents 6 frames selected from the animation where the weight on the politics dimension is gradually increased, where other dimensions (economy, education, health) decrease correspondingly.
+It presents 6 frames selected from an animation where the weight on the politics dimension is gradually increased, while other dimensions (economy, education, health) decrease correspondingly.
 Frame 12 present the original index where all the four dimensions receive equal weight.
@@
 
-To make a simple illustration of sensitivity analysis, we slightly vary the weight for politics, between 0.07 and 0.52, while maintain equal weights among other dimensions.
+To make a simple illustration of sensitivity analysis, we slightly vary the weight for politics, between 0.07 and 0.52, while maintaining equal weights among other dimensions.
 This can be viewed as an animation to examine change in relative index values as a response to the changing weights.

### 2024-04-24 tidyindex.tex
Small changes and fixes

@@
 development. It is primarily due to a lack of standards in data analysis
-workflow. Donoho (2017) called for research on structuring unified
+workflow. Donoho (2017) called for research on structuring a unified
 workflow to address methodological variation across studies in data
@@
 of each component. We call this process the \emph{index pipeline}, which
-are the steps of the data analysis pipeline for the index construction.
-Such decomposition of index components provides the means to standardize
+are the steps of the data analysis pipeline for index construction. Such
+a decomposition of index components provides the means to standardize
 index construction via a pipeline and offers benefits for comparing
 among indexes, calculating index uncertainty, and assessing index
-robustness.
+robustness. In addition it provides an unambiguous documentation for the
+index definition, facilitating reproducibility of results.
 
@@
 most workflow. In the \texttt{tidymodels}, these steps are
-correspondingly implemented as package \texttt{rsample},
+correspondingly implemented as packages \texttt{rsample},
 \texttt{parsnip}, and \texttt{yardstick}, agnostic to the specific model
@@
   provides the capacity to measure uncertainty by computing confidence
-  intervals from multiple samples as generated by bootstrapping to
+  intervals from multiple samples as generated by bootstrapping the
   original data.
@@
 \([x_{j}(s;t) - \alpha]/\gamma\) where \(\alpha\) and \(\gamma\) are
-constants . In the Human Development Index (HDI), the three dimensions
+constants. In the Human Development Index (HDI), the three dimensions
 (health, education, and economy) are converted into the same scale (0-1)
@@
 
-The accompanied package, tidyindex, is not intended to offer an
-exhaustive implementation for all indexes across every domains. Instead,
-it provides a realization of the pipeline framework proposed in the
-paper. When adopting the pipeline approach to construct indexes,
-analysts may consider developing software that can be readily deployed
-in the cloud for production purposes.
+The tidyindex package is not intended to offer an exhaustive
+implementation for all indexes across all domains. Instead, it provides
+a realization of the pipeline framework proposed in the paper. When
+adopting the pipeline approach to construct indexes, analysts may
+consider developing software that can be readily deployed in the cloud
+for production purposes.
 
@@
 dataset used in this example is available in the \texttt{tidyindex}
-package as \texttt{queensland} and blow prints the first few rows of the
-data:
+package as \texttt{queensland} and below we show the first few rows of
+the data:
 
@@
 2 ASN00029038 1990 Feb   416  35.2  23.2  29.2  142. -15.5 KOWANYAMA ~
-3 ASN00029038 1990 Mar  2026  32.5  23.6  28.0  142. -15.5 KOWANYAMA ~
+3 ASN00029038 1990 Mär  2026  32.5  23.6  28.0  142. -15.5 KOWANYAMA ~
 4 ASN00029038 1990 Apr   597  32.9  17.7  25.3  142. -15.5 KOWANYAMA ~
-5 ASN00029038 1990 May   244  31.8  20.1  25.9  142. -15.5 KOWANYAMA ~
+5 ASN00029038 1990 Mai   244  31.8  20.1  25.9  142. -15.5 KOWANYAMA ~
 \end{verbatim}
@@
 $ month  <dbl> 6, 7, 8, 9, 10, 11, 12, 12, 1, 1, 2, 2, 3, 3, 4, 4, 5~
-$ ym     <mth> 1990 Jun, 1990 Jul, 1990 Aug, 1990 Sep, 1990 Oct, 199~
+$ ym     <mth> 1990 Jun, 1990 Jul, 1990 Aug, 1990 Sep, 1990 Okt, 199~
 $ prcp   <dbl> 170, 102, 0, 0, 0, 278, 1869, 1869, 5088, 5088, 8484,~
@@
 GGGI, for a subset of 16 countries. It presents 6 frames selected from
-the animation where the weight on the politics dimension is gradually
-increased, where other dimensions (economy, education, health) decrease
+an animation where the weight on the politics dimension is gradually
+increased, while other dimensions (economy, education, health) decrease
 correspondingly. Frame 12 present the original index where all the four
@@
 To make a simple illustration of sensitivity analysis, we slightly vary
-the weight for politics, between 0.07 and 0.52, while maintain equal
+the weight for politics, between 0.07 and 0.52, while maintaining equal
 weights among other dimensions. This can be viewed as an animation to

### 2024-04-25 tidyindex.qmd
further tidying of intro

@@
 We call this process the *index pipeline*, which are the steps of the data analysis pipeline for index construction.
-Such a decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes, calculating index uncertainty, and assessing index robustness.
-In addition it provides an unambiguous documentation for the index definition, facilitating reproducibility of results.
+Such a decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing versions of indexes, calculating index uncertainty, and assessing index robustness.
+It also provides clear recipes for the index definition, facilitating reproducibility of results.
 
-In this work, we provide statistical and computational methods for developing a data pipeline framework to construct and customize indexes using data.
-The proposed pipeline comprises various modules, including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
-Given the decisions analysts need to make when combining multivariate data into indexes, the proposed pipeline enables the evaluation of how the specific choice can affect the index, as well as how the index may appear under alternative options.
-Furthermore, uncertainty calculation can also flow through the pipeline, providing the index with confidence measures.
+Here we detail the statistical and computational methods for developing a data pipeline framework to construct and customize indexes using data.
+The pipeline comprises various modules, including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication.
+When combining multivariate data into indexes, the pipeline enables the evaluation of how any particular combination can affect the index.
+Uncertainty calculation can also flow through the pipeline to provide an index with confidence intervals.
+The pipeline also fits neatly into current tidy data workflows and data visualisation.
 

### 2024-04-25 tidyindex.tex
further tidying of intro

--- 2024-04-24 tidyindex.tex
@@
 of each component. We call this process the \emph{index pipeline}, which
-are the steps of the data analysis pipeline for the index construction.
-Such decomposition of index components provides the means to standardize
+are the steps of the data analysis pipeline for index construction. Such
+a decomposition of index components provides the means to standardize
 index construction via a pipeline and offers benefits for comparing
-among indexes, calculating index uncertainty, and assessing index
-robustness.
-
-In this work, we provide statistical and computational methods for
-developing a data pipeline framework to construct and customize indexes
-using data. The proposed pipeline comprises various modules, including
-temporal and spatial aggregation, variable transformation and
-combination, distribution fitting, benchmark setting, and index
-communication. Given the decisions analysts need to make when combining
-multivariate data into indexes, the proposed pipeline enables the
-evaluation of how the specific choice can affect the index, as well as
-how the index may appear under alternative options. Furthermore,
-uncertainty calculation can also flow through the pipeline, providing
-the index with confidence measures.
+versions of indexes, calculating index uncertainty, and assessing index
+robustness. It also provides clear recipes for the index definition,
+facilitating reproducibility of results.
+
+Here we detail the statistical and computational methods for developing
+a data pipeline framework to construct and customize indexes using data.
+The pipeline comprises various modules, including temporal and spatial
+aggregation, variable transformation and combination, distribution
+fitting, benchmark setting, and index communication. When combining
+multivariate data into indexes, the pipeline enables the evaluation of
+how any particular combination can affect the index. Uncertainty
+calculation can also flow through the pipeline to provide an index with
+confidence intervals. The pipeline also fits neatly into current tidy
+data workflows and data visualisation.
 
@@
 
-\begingroup\fontsize{11}{13}\selectfont
-
-\begin{longtable}[t]{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
-
+\begin{table}
+
+\centering{
+
+\centering\begingroup\fontsize{11}{13}\selectfont
+
+\begin{tabular}{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
 \toprule
@@
 \bottomrule
+\end{tabular}
+\endgroup{}
+
+}
 
@@
 
-\tabularnewline
-
-\end{longtable}
-
-\endgroup{}
+\end{table}%
 

### 2024-04-26 tidyindex.qmd
fit into 33 pages

@@
 
--   $x_j(s, t)$ represents a variable of interest for example precipitation, $j = 1, \cdots, p$ and
+-   $x_j(s, t)$ represents a variable of interest for example precipitation, $j = 1, \cdots, p$,
 
@@
 
-## Scaling
-
-Unlike variable transformation, scaling maintains the distributional shape of the variable.
-It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$ where $\alpha$ and $\gamma$ are constants.
-In the Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.
-
-Although the scaling might be considered to be a transformation, we have elected to make it a separate module because it is neater.
-Scaling simply changes the numbers in the data not the shape of a variable.
-Transformation will most likely change the shape, and is usually non-linear.
-
 ```{r fig-scale-var-trans-compare}
@@
 "
-#| fig-height: 6
+#| fig-height: 5.5
 #| fig-width: 8
@@
 ```
+
+## Scaling
+
+Unlike variable transformation, scaling maintains the distributional shape of the variable.
+It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$ where $\alpha$ and $\gamma$ are constants.
+In the Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.
+
+Although the scaling might be considered to be a transformation, we have elected to make it a separate module because it is neater.
+@fig-scale-var-trans-compare shows that scaling simply changes the numbers in the data not the shape of a variable, while transformation will most likely change the shape, and is usually non-linear.
 
@@
 
-```{r echo = TRUE}
+```{r echo = FALSE}
 dplyr::glimpse(idx)
 ```
-
-The output contains the original data, index values (`.index`), parameters used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`.pet`, `.agg`, and `.fitted`).
-This data can be visualized to investigate the spatio-temporal distribution of the drought or flood events, as well as the response of index values to different time scales and distribution parameters at specific single locations.
-@fig-compute-spatial and @fig-compute-temporal exemplify two possibilities.
-@fig-compute-spatial presents the spatial distribution of SPI during two periods: October 2010 to March 2011 for the 2010/11 Queensland flood and October 2019 to March 2020 for the 2019 Australia drought, which contributes to the notorious 2019/20 bushfire season.
-@fig-compute-temporal displays the sensitivity of the SPEI series at the Texas post office to different time scales and fitted distributions.
-Larger time scales produce a smoother index across time, however, all time scales indicate an extreme drought (corresponding to -2 in SPEI) in 2020, confirming the severity of the drought across different time horizons.
-Moreover, the chosen distribution has less influence on the index, with general extreme value distribution tending to produce more extreme outcomes than log-logistic distribution for the extreme events (index > 2 or <-2).
 
@@
 #| fig-width: 8
-#| fig-height: 5
+#| fig-height: 4.5
 texas <- idx %>%
@@
 ```
+
+The output contains the original data, index values (`.index`), parameters used (`.scale`, `.method`, and `.dist`), and all the intermediate variables (`.pet`, `.agg`, and `.fitted`).
+This data can be visualized to investigate the spatio-temporal distribution of the drought or flood events, as well as the response of index values to different time scales and distribution parameters at specific single locations.
+@fig-compute-spatial and @fig-compute-temporal exemplify two possibilities.
+@fig-compute-spatial presents the spatial distribution of SPI during two periods: October 2010 to March 2011 for the 2010/11 Queensland flood and October 2019 to March 2020 for the 2019 Australia drought, which contributes to the notorious 2019/20 bushfire season.
+@fig-compute-temporal displays the sensitivity of the SPEI series at the Texas post office to different time scales and fitted distributions.
+Larger time scales produce a smoother index across time, however, all time scales indicate an extreme drought (corresponding to -2 in SPEI) in 2020, confirming the severity of the drought across different time horizons.
+Moreover, the chosen distribution has less influence on the index, with general extreme value distribution tending to produce more extreme outcomes than log-logistic distribution for the extreme events (index > 2 or <-2).
 
@@
 Also interesting is the large drop in value for Bangladesh.
-
 Frames 18, 24, 29 show linear combinations where politics contributes more than the original.

### 2024-04-26 tidyindex.tex
fit into 33 pages

@@
   \(x_j(s, t)\) represents a variable of interest for example
-  precipitation, \(j = 1, \cdots, p\) and
+  precipitation, \(j = 1, \cdots, p\),
 \item
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics{tidyindex_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
+
+}
+
+\caption{\label{fig-scale-var-trans-compare}Comparison of the module
+scaling (green) and variable transformation (orange). While both modules
+change the variable range, scaling maintains the same distributional
+shape, which is not the case with variable transformation.}
+
+\end{figure}%
+
 \subsection{Scaling}\label{scaling}
@@
 Although the scaling might be considered to be a transformation, we have
-elected to make it a separate module because it is neater. Scaling
-simply changes the numbers in the data not the shape of a variable.
-Transformation will most likely change the shape, and is usually
+elected to make it a separate module because it is neater.
+Figure~\ref{fig-scale-var-trans-compare} shows that scaling simply
+changes the numbers in the data not the shape of a variable, while
+transformation will most likely change the shape, and is usually
 non-linear.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{tidyindex_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
-
-}
-
-\caption{\label{fig-scale-var-trans-compare}Comparison of the module
-scaling (green) and variable transformation (orange). While both modules
-change the variable range, scaling maintains the same distributional
-shape, which is not the case with variable transformation.}
-
-\end{figure}%
 
@@
 \texttt{idx} object created:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{dplyr}\SpecialCharTok{::}\FunctionTok{glimpse}\NormalTok{(idx)}
-\end{Highlighting}
-\end{Shaded}
 
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics{tidyindex_files/figure-pdf/fig-compute-spatial-1.pdf}
+
+}
+
+\caption{\label{fig-compute-spatial}Spatial distribution of Standardized
+Precipitation Index (SPI-12) in Queensland, Australia during two major
+flood and drought events: 2010/11 and 2019/20. The map shows a
+continuous wet period during the 2010/11 flood period and a mitigated
+drought situation, after its worst in 2019 December and 2020 January,
+likely due to the increased rainfall in February from the meteorological
+record.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{tidyindex_files/figure-pdf/fig-compute-temporal-1.pdf}
+
+}
+
+\caption{\label{fig-compute-temporal}Time series plot of Standardized
+Precipitation-Evapotranspiration Index (SPEI) at the Texas post office
+station (highlighted by a diamond shape in panel a). The SPEI is
+calculated at four time scales (6, 12, 24, and 36 months) and fitted
+with two distributions (Log Logistic and GEV). The dashed line at -2
+represents the class ``extreme drought'' by the SPEI. A larger time
+scale gives a smoother index series, while also taking longer to recover
+from an extreme situation as seen in the 2019/20 drought period. The
+SPEI values from the two distributional fits mostly agree, while GEV can
+result in more extreme values, i.e.~in 1998 and 2020.}
+
+\end{figure}%
+
 The output contains the original data, index values (\texttt{.index}),
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics{tidyindex_files/figure-pdf/fig-compute-spatial-1.pdf}
-
-}
-
-\caption{\label{fig-compute-spatial}Spatial distribution of Standardized
-Precipitation Index (SPI-12) in Queensland, Australia during two major
-flood and drought events: 2010/11 and 2019/20. The map shows a
-continuous wet period during the 2010/11 flood period and a mitigated
-drought situation, after its worst in 2019 December and 2020 January,
-likely due to the increased rainfall in February from the meteorological
-record.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{tidyindex_files/figure-pdf/fig-compute-temporal-1.pdf}
-
-}
-
-\caption{\label{fig-compute-temporal}Time series plot of Standardized
-Precipitation-Evapotranspiration Index (SPEI) at the Texas post office
-station (highlighted by a diamond shape in panel a). The SPEI is
-calculated at four time scales (6, 12, 24, and 36 months) and fitted
-with two distributions (Log Logistic and GEV). The dashed line at -2
-represents the class ``extreme drought'' by the SPEI. A larger time
-scale gives a smoother index series, while also taking longer to recover
-from an extreme situation as seen in the 2019/20 drought period. The
-SPEI values from the two distributional fits mostly agree, while GEV can
-result in more extreme values, i.e.~in 1998 and 2020.}
-
-\end{figure}%
-
 \subsection{Does a puff of change in variable weights cause a tornado in
@@
 these countries. Also interesting is the large drop in value for
-Bangladesh.
-
-Frames 18, 24, 29 show linear combinations where politics contributes
-more than the original. The most notable feature is that Bangladesh
-retains it's high index value whereas the other middle group countries
-decline, indicating that the politics score is a major component for
-Bangladesh's index value.
+Bangladesh. Frames 18, 24, 29 show linear combinations where politics
+contributes more than the original. The most notable feature is that
+Bangladesh retains it's high index value whereas the other middle group
+countries decline, indicating that the politics score is a major
+component for Bangladesh's index value.
 

### 2024-04-28 tidyindex.qmd
minor polishing
mostly small typos

@@
 
-Since the success of tidyverse, more packages have been developed to analyze data using the tidy framework for domains specific applications, a noticeable example of which is `tidymodels` for building machine learning models [@tidymodels].
+Since the success of tidyverse, more packages have been developed to analyze data using the tidy framework for domain-specific applications, a noticeable example of which is `tidymodels` for building machine learning models [@tidymodels].
 To create a tidy workflow tailored to a specific domain, developers first need to identify the fundamental building blocks to create a workflow.
 These components are then implemented as modules, which can be combined to form the pipeline.
-For example, in supervised machine learning models, steps such as data splitting, model training, and model evaluation are commonly used in most workflow.
+For example, in supervised machine learning models, steps such as data splitting, model training, and model evaluation are commonly used in most workflows.
 In the `tidymodels`, these steps are correspondingly implemented as packages `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen.
@@
 
-Adoption of this pipeline approach would provide uniformity to the field of index development, research, and application to improve comparability, reproducibility, and communication.
+The adoption of this pipeline approach would provide uniformity to the field of index development, research, and application to improve comparability, reproducibility, and communication.
 
@@
 #| fig-align: center
-#| fig-cap: "Diagram of pipeline modules for index construction. The highlighted path illustrates one possible construction using the dimension reduction and simplification module."
+#| fig-cap: "Diagram of pipeline modules for index construction. The highlighted path illustrates one possible construction using the dimension reduction and simplification modules."
 #| out-height: 90%
@@
 In what follows when geographic or temporal components of the $x_j(s,t)$ process are fixed we will be using suffix notation.
-For example, $x_{sj}(t)$ represents the data for a fixed location $s$ as a function of time $t$.
-While $x_{tj}(s)$ denotes the spatial varying process for a fixed $t$.
+For example, $x_{sj}(t)$ represents the data for a fixed location $s$ as a function of time $t$, while $x_{tj}(s)$ denotes the spatial varying process for a fixed $t$.
 An overview of the notation for pipeline input, operation, and output is presented in @tbl-notation.
@@
 The temporal processing module takes as input argument a single variable $x_{sj}(t)$ at location $s$ as a function of time.
-In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation.
+In this step, the original time series can be transformed or summarized into a new one via time aggregation.
+The transformation is represented by the function $f$, $x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation.
 An example of temporal processing done in the computation of the Standardized Precipitation Index (SPI) [@mckee1993relationship], consists of summing the monthly precipitation series over a rolling time window of size $k$.
@@
 ```{r fig-scale-var-trans-compare}
-#| fig-cap: "Comparison of the module scaling (green) and variable transformation (orange).
+#| fig-cap: "Comparison of the scaling (green) and variable transformation (orange) modules.
 While both modules change the variable range, scaling maintains the same distributional shape, which is not the case with variable transformation.
@@
 Although the scaling might be considered to be a transformation, we have elected to make it a separate module because it is neater.
-@fig-scale-var-trans-compare shows that scaling simply changes the numbers in the data not the shape of a variable, while transformation will most likely change the shape, and is usually non-linear.
+@fig-scale-var-trans-compare shows that scaling simply changes the numbers in the data but not the shape of a variable, while transformation will most likely change the shape, as it is usually non-linear.
 
@@
 \begin{cases}
-a_0 & C_1 \leq x^i(s; t) < C_0 \\
-a_1 & C_2 \leq x^i(s; t) < C_1 \\
-a_2 & C_3 \leq x^i(s; t) < C_2 \\
+a_0, & C_1 \leq x^i(s; t) < C_0 \\
+a_1, & C_2 \leq x^i(s; t) < C_1 \\
+a_2, & C_3 \leq x^i(s; t) < C_2 \\
 \cdots \\
-a_z & C_z \leq x^i(s; t)
+a_z, & C_z \leq x^i(s; t)
 \end{cases}
@@
 It presents 6 frames selected from an animation where the weight on the politics dimension is gradually increased, while other dimensions (economy, education, health) decrease correspondingly.
-Frame 12 present the original index where all the four dimensions receive equal weight.
+Frame 12 presents the original index where all the four dimensions receive equal weight.
 The index values are sorted from highest to lowest, with the Nordic countries (Iceland, Norway, and Finland) and New Zealand leading the rankings.
@@
 Frames 18, 24, 29 show linear combinations where politics contributes more than the original.
-The most notable feature is that Bangladesh retains it's high index value whereas the other middle group countries decline, indicating that the politics score is a major component for Bangladesh's index value.
+The most notable feature is that Bangladesh retains its high index value whereas the other middle group countries decline, indicating that the politics score is a major component for Bangladesh's index value.
 
@@
 This is not the case with GGGI, where small changes to one component lead to fairly large change in the index.
-The modular pipeline framework for computing the index makes it easy to conduct this type of sensitivity analysis, where one or more components are perturbed and the index re-calculated.
-One aspect of the GGGI not well-described in the Global Gender Gap Report is handling of missing values that are present in the initial variables for many countries, something that is common for this type of data.
+The modular pipeline framework for computing the index makes it easy to conduct this type of sensitivity analysis, where one or more components are perturbed and the index recalculated.
+One aspect of the GGGI not well-described in the Global Gender Gap Report is the handling of missing values that are present in the initial variables for many countries, something that is common for this type of data.
 This could also be made more transparent with the dimension reduction module, by specifying an imputation method or providing warnings about missing values.
@@
 The confidence interval can then be calculated using the quantile method from the bootstrap samples.
-@fig-conf-interval presents the 80% and 95% confidence interval for the Texas post office station, in Queensland, Australia.
+@fig-conf-interval presents the 80% and 95% confidence intervals for the Texas post office station, in Queensland, Australia.
 From the start of 2019 to 2020, the majority of the confidence intervals lie below the extreme drought line (SPI = -2), suggesting a high level of certainty that the Texas post office is suffering from a drastic drought.
@@
 #| fig-height: 2
-#| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI-24) for the Texas post office station, in Queensland, Australia.
+#| fig-cap: "80% and 95% confidence intervals of the Standardized Precipitation Index (SPI-24) for the Texas post office station, in Queensland, Australia.
 A bootstrap sample of 100 is taken from the aggregated precipitation series to estimate gamma parameters and to calculate the index.

### 2024-04-28 tidyindex.tex
minor polishing
mostly small typos

@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
   \fi
   % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \usepackage{orcidlink}
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
 \makeatletter
@@
 \makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
 \makeatother
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
 \makeatother
@@
 \fi
-\usepackage{bookmark}
-
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \end{keywords}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners, frame hidden, enhanced]}{\end{tcolorbox}}\fi
 
 \setstretch{2}
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-\section{Background to index development}\label{sec-idx-dev}
+\hypertarget{sec-idx-dev}{%
+\section{Background to index development}\label{sec-idx-dev}}
 
@@
 
-\section{Tidy framework}\label{sec-tidy}
+\hypertarget{sec-tidy}{%
+\section{Tidy framework}\label{sec-tidy}}
 
@@
 Since the success of tidyverse, more packages have been developed to
-analyze data using the tidy framework for domains specific applications,
+analyze data using the tidy framework for domain-specific applications,
 a noticeable example of which is \texttt{tidymodels} for building
@@
 splitting, model training, and model evaluation are commonly used in
-most workflow. In the \texttt{tidymodels}, these steps are
+most workflows. In the \texttt{tidymodels}, these steps are
 correspondingly implemented as packages \texttt{rsample},
@@
 
-Adoption of this pipeline approach would provide uniformity to the field
-of index development, research, and application to improve
+The adoption of this pipeline approach would provide uniformity to the
+field of index development, research, and application to improve
 comparability, reproducibility, and communication.
 
-\section{Details of the index pipeline}\label{sec-pipeline}
+\hypertarget{sec-pipeline}{%
+\section{Details of the index pipeline}\label{sec-pipeline}}
 
@@
 
-\centering{
-
-\includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-overall.png}
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-overall.png}
 
@@
 construction. The highlighted path illustrates one possible construction
-using the dimension reduction and simplification module.}
-
-\end{figure}%
+using the dimension reduction and simplification modules.}
+
+\end{figure}
 
@@
 example, \(x_{sj}(t)\) represents the data for a fixed location \(s\) as
-a function of time \(t\). While \(x_{tj}(s)\) denotes the spatial
+a function of time \(t\), while \(x_{tj}(s)\) denotes the spatial
 varying process for a fixed \(t\). An overview of the notation for
@@
 
+\hypertarget{tbl-notation}{}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3562}}@{}}
-\caption{Summary of the notation for input, operation, and output of
-each pipeline module.}\label{tbl-notation}\tabularnewline
+\caption{\label{tbl-notation}Summary of the notation for input,
+operation, and output of each pipeline module.}\tabularnewline
 \toprule\noalign{}
@@
 
-\subsection{Temporal processing}\label{temporal-processing}
+\hypertarget{temporal-processing}{%
+\subsection{Temporal processing}\label{temporal-processing}}
 
 The temporal processing module takes as input argument a single variable
-\(x_{sj}(t)\) at location \(s\) as a function of time. In this step the
+\(x_{sj}(t)\) at location \(s\) as a function of time. In this step, the
 original time series can be transformed or summarized into a new one via
-time aggregation, the transformation is represented by the function
+time aggregation. The transformation is represented by the function
 \(f\), \(x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]\) where
@@
 
-\subsection{Spatial processing}\label{spatial-processing}
+\hypertarget{spatial-processing}{%
+\subsection{Spatial processing}\label{spatial-processing}}
 
@@
 
-\subsection{Variable transformation}\label{variable-transformation}
+\hypertarget{variable-transformation}{%
+\subsection{Variable transformation}\label{variable-transformation}}
 
@@
 
-\centering{
-
-\includegraphics{tidyindex_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}
 
@@
 
-\caption{\label{fig-scale-var-trans-compare}Comparison of the module
-scaling (green) and variable transformation (orange). While both modules
+\caption{\label{fig-scale-var-trans-compare}Comparison of the scaling
+(green) and variable transformation (orange) modules. While both modules
 change the variable range, scaling maintains the same distributional
@@
 
-\end{figure}%
-
-\subsection{Scaling}\label{scaling}
+\end{figure}
+
+\hypertarget{scaling}{%
+\subsection{Scaling}\label{scaling}}
 
@@
 Figure~\ref{fig-scale-var-trans-compare} shows that scaling simply
-changes the numbers in the data not the shape of a variable, while
-transformation will most likely change the shape, and is usually
+changes the numbers in the data but not the shape of a variable, while
+transformation will most likely change the shape, as it is usually
 non-linear.
 
-\subsection{Dimension reduction}\label{dimension-reduction}
+\hypertarget{dimension-reduction}{%
+\subsection{Dimension reduction}\label{dimension-reduction}}
 
@@
 
-\subsection{Distribution fit}\label{distribution-fit}
+\hypertarget{distribution-fit}{%
+\subsection{Distribution fit}\label{distribution-fit}}
 
@@
 
-\subsection{Normalising}\label{normalising}
+\hypertarget{normalising}{%
+\subsection{Normalising}\label{normalising}}
 
@@
 
-\subsection{Benchmarking}\label{benchmarking}
+\hypertarget{benchmarking}{%
+\subsection{Benchmarking}\label{benchmarking}}
 
@@
 
-\subsection{Simplification}\label{simplification}
+\hypertarget{simplification}{%
+\subsection{Simplification}\label{simplification}}
 
@@
 
-\section{Software design}\label{sec-software}
+\hypertarget{sec-software}{%
+\section{Software design}\label{sec-software}}
 
@@
 
-\section{Examples}\label{sec-examples}
+\hypertarget{sec-examples}{%
+\section{Examples}\label{sec-examples}}
 
@@
 
+\hypertarget{sec-example1}{%
 \subsection{Every distribution, every scale, every index all at
-once}\label{sec-example1}
+once}\label{sec-example1}}
 
@@
 
-\centering{
-
-\includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-spei.png}
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-spei.png}
 
@@
 
-\end{figure}%
+\end{figure}
 
@@
 
-\centering{
-
-\includegraphics{tidyindex_files/figure-pdf/fig-compute-spatial-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-compute-spatial-1.pdf}
 
@@
 
-\centering{
-
-\includegraphics{tidyindex_files/figure-pdf/fig-compute-temporal-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-compute-temporal-1.pdf}
 
@@
 
+\hypertarget{does-a-puff-of-change-in-variable-weights-cause-a-tornado-in-ranks}{%
 \subsection{Does a puff of change in variable weights cause a tornado in
-ranks?}\label{does-a-puff-of-change-in-variable-weights-cause-a-tornado-in-ranks}
+ranks?}\label{does-a-puff-of-change-in-variable-weights-cause-a-tornado-in-ranks}}
 
@@
 
-\centering{
-
-\includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-gggi.png}
+{\centering \includegraphics[width=1\textwidth,height=0.9\textheight]{figures/pipeline-gggi.png}
 
@@
 
+\hypertarget{tbl-gggi-weights}{}
 \begin{longtable}[t]{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
+\caption{\label{tbl-gggi-weights}Weights for the two applications of dimension reduction to compute the
+Global Gender Gap Index. V-wgt is used to compute four new variables
+from the original 14. These are then equally combined to get the final
+index value. }\tabularnewline
 
@@
 \bottomrule
-
-\caption{\label{tbl-gggi-weights}Weights for the two applications of
-dimension reduction to compute the Global Gender Gap Index. V-wgt is
-used to compute four new variables from the original 14. These are then
-equally combined to get the final index value.}
-
-\tabularnewline
-
 \end{longtable}
-
 \endgroup{}
@@
 increased, while other dimensions (economy, education, health) decrease
-correspondingly. Frame 12 present the original index where all the four
+correspondingly. Frame 12 presents the original index where all the four
 dimensions receive equal weight. The index values are sorted from
@@
 
-\centering{
-
-\includegraphics{tidyindex_files/figure-pdf/fig-idx-tour-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-idx-tour-1.pdf}
 
@@
 contributes more than the original. The most notable feature is that
-Bangladesh retains it's high index value whereas the other middle group
+Bangladesh retains its high index value whereas the other middle group
 countries decline, indicating that the politics score is a major
@@
 conduct this type of sensitivity analysis, where one or more components
-are perturbed and the index re-calculated. One aspect of the GGGI not
-well-described in the Global Gender Gap Report is handling of missing
-values that are present in the initial variables for many countries,
-something that is common for this type of data. This could also be made
-more transparent with the dimension reduction module, by specifying an
-imputation method or providing warnings about missing values.
-
+are perturbed and the index recalculated. One aspect of the GGGI not
+well-described in the Global Gender Gap Report is the handling of
+missing values that are present in the initial variables for many
+countries, something that is common for this type of data. This could
+also be made more transparent with the dimension reduction module, by
+specifying an imputation method or providing warnings about missing
+values.
+
+\hypertarget{decoding-uncertainty-through-the-wisdom-of-the-crowd}{%
 \subsection{Decoding uncertainty through the wisdom of the
-crowd}\label{decoding-uncertainty-through-the-wisdom-of-the-crowd}
+crowd}\label{decoding-uncertainty-through-the-wisdom-of-the-crowd}}
 
@@
 from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
-80\% and 95\% confidence interval for the Texas post office station, in
+80\% and 95\% confidence intervals for the Texas post office station, in
 Queensland, Australia. From the start of 2019 to 2020, the majority of
@@
 
-\centering{
-
-\includegraphics{tidyindex_files/figure-pdf/fig-conf-interval-1.pdf}
+{\centering \includegraphics{tidyindex_files/figure-pdf/fig-conf-interval-1.pdf}
 
@@
 
-\caption{\label{fig-conf-interval}80\% and 95\% confidence interval of
+\caption{\label{fig-conf-interval}80\% and 95\% confidence intervals of
 the Standardized Precipitation Index (SPI-24) for the Texas post office
@@
 
-\end{figure}%
-
-\section{Conclusion}\label{conclusion}
+\end{figure}
+
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
@@
 
-\section{Acknowledgement}\label{acknowledgement}
+\hypertarget{acknowledgement}{%
+\section{Acknowledgement}\label{acknowledgement}}
 
@@
 
-\section*{References}\label{references}
+\hypertarget{references}{%
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}
 
-\phantomsection\label{refs}
+\hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-alahacoon_comprehensive_2022}
+\leavevmode\vadjust pre{\hypertarget{ref-alahacoon_comprehensive_2022}{}}%
 Alahacoon, Niranga, and Mahesh Edirisinghe. 2022. {``A Comprehensive
@@
 
-\bibitem[\citeproctext]{ref-Allaire_Quarto_2022}
+\leavevmode\vadjust pre{\hypertarget{ref-Allaire_Quarto_2022}{}}%
 Allaire, J. J., Charles Teague, Carlos Scheidegger, Yihui Xie, and
@@
 
-\bibitem[\citeproctext]{ref-COINr}
+\leavevmode\vadjust pre{\hypertarget{ref-COINr}{}}%
 Becker, William, Giulio Caperna, Maria Del Sorbo, Hedvig Norlen, Eleni
@@
 
-\bibitem[\citeproctext]{ref-SPEI}
+\leavevmode\vadjust pre{\hypertarget{ref-SPEI}{}}%
 Beguería, Santiago, and Sergio M. Vicente-Serrano. 2017. \emph{SPEI:
@@
 
-\bibitem[\citeproctext]{ref-buja_elements_1988}
+\leavevmode\vadjust pre{\hypertarget{ref-buja_elements_1988}{}}%
 Buja, A, D Asimov, C Hurley, and JA McDonald. 1988. {``Elements of a
@@
 
-\bibitem[\citeproctext]{ref-buja2005computational}
+\leavevmode\vadjust pre{\hypertarget{ref-buja2005computational}{}}%
 Buja, Andreas, Dianne Cook, Daniel Asimov, and Catherine Hurley. 2005.
@@
 
-\bibitem[\citeproctext]{ref-chambers1998programming}
-Chambers, John M. 1998. \emph{Programming with Data: A Guide to the s
+\leavevmode\vadjust pre{\hypertarget{ref-chambers1998programming}{}}%
+Chambers, John M. 1998. \emph{Programming with Data: A Guide to the {S}
 Language}. Berlin, Heidelberg: Springer-Verlag.
 
-\bibitem[\citeproctext]{ref-donoho_50_2017}
+\leavevmode\vadjust pre{\hypertarget{ref-donoho_50_2017}{}}%
 Donoho, David. 2017. {``50 {Years} of {Data} {Science}.''} \emph{Journal
@@
 
-\bibitem[\citeproctext]{ref-efron_bootstrap_1979}
+\leavevmode\vadjust pre{\hypertarget{ref-efron_bootstrap_1979}{}}%
 Efron, B. 1979. {``Bootstrap {Methods}: {Another} {Look} at the
@@
 
-\bibitem[\citeproctext]{ref-fisher1936use}
+\leavevmode\vadjust pre{\hypertarget{ref-fisher1936use}{}}%
 Fisher, Ronald A. 1936. {``The Use of Multiple Measurements in Taxonomic
@@
 
-\bibitem[\citeproctext]{ref-fisher1970statistical}
+\leavevmode\vadjust pre{\hypertarget{ref-fisher1970statistical}{}}%
 Fisher, Ronald Aylmer. 1970. {``Statistical Methods for Research
@@
 
-\bibitem[\citeproctext]{ref-fundiversity}
+\leavevmode\vadjust pre{\hypertarget{ref-fundiversity}{}}%
 Grenié, Matthias, and Hugo Gruson. 2023. \emph{{fundiversity}: Easy
@@
 
-\bibitem[\citeproctext]{ref-hao_drought_2015}
+\leavevmode\vadjust pre{\hypertarget{ref-hao_drought_2015}{}}%
 Hao, Zengchao, and Vijay P. Singh. 2015. {``Drought Characterization
@@
 
-\bibitem[\citeproctext]{ref-hotelling1933analysis}
+\leavevmode\vadjust pre{\hypertarget{ref-hotelling1933analysis}{}}%
 Hotelling, Harold. 1933. {``Analysis of a Complex of Statistical
@@
 
-\bibitem[\citeproctext]{ref-jones_vulnerability_2007}
+\leavevmode\vadjust pre{\hypertarget{ref-jones_vulnerability_2007}{}}%
 Jones, Brenda, and Jean Andrey. 2007. {``Vulnerability Index
@@
 
-\bibitem[\citeproctext]{ref-tidymodels}
-Kuhn, Max, and Julia Silge. 2022. \emph{Tidy Modeling with r}. "
+\leavevmode\vadjust pre{\hypertarget{ref-tidymodels}{}}%
+Kuhn, Max, and Julia Silge. 2022. \emph{Tidy Modeling with {R}}. "
 O'Reilly Media, Inc.".
 
-\bibitem[\citeproctext]{ref-laimighofer_how_2022}
+\leavevmode\vadjust pre{\hypertarget{ref-laimighofer_how_2022}{}}%
 Laimighofer, Johannes, and Gregor Laaha. 2022. {``How Standard Are
@@
 
-\bibitem[\citeproctext]{ref-gpindex}
+\leavevmode\vadjust pre{\hypertarget{ref-gpindex}{}}%
 Martin, Steve. 2023. \emph{Gpindex: Generalized Price and Quantity
@@
 
-\bibitem[\citeproctext]{ref-mckee1993relationship}
+\leavevmode\vadjust pre{\hypertarget{ref-mckee1993relationship}{}}%
 McKee, Thomas B, Nolan J Doesken, John Kleist, et al. 1993. {``The
@@
 
-\bibitem[\citeproctext]{ref-oecd_handbook_2008}
+\leavevmode\vadjust pre{\hypertarget{ref-oecd_handbook_2008}{}}%
 OECD, European Union, and Joint Research Centre - European Commission.
@@
 
-\bibitem[\citeproctext]{ref-R}
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
 R Core Team. 2021. \emph{R: A Language and Environment for Statistical
@@
 
-\bibitem[\citeproctext]{ref-uncertainty}
+\leavevmode\vadjust pre{\hypertarget{ref-uncertainty}{}}%
 Saisana, M., A. Saltelli, and S. Tarantola. 2005. {``{Uncertainty and
@@
 
-\bibitem[\citeproctext]{ref-RJ-2020-027}
+\leavevmode\vadjust pre{\hypertarget{ref-RJ-2020-027}{}}%
 Spyrison, Nicholas, and Dianne Cook. 2020. {``Spinifex: An {R} Package
@@
 
-\bibitem[\citeproctext]{ref-sutherland_orca_2000}
+\leavevmode\vadjust pre{\hypertarget{ref-sutherland_orca_2000}{}}%
 Sutherland, Peter, Anthony Rossini, Thomas Lumley, Nicholas Lewin-Koh,
@@
 
-\bibitem[\citeproctext]{ref-svoboda2016handbook}
+\leavevmode\vadjust pre{\hypertarget{ref-svoboda2016handbook}{}}%
 Svoboda, Mark, Brian Fuchs, et al. 2016. {``Handbook of Drought
@@
 
-\bibitem[\citeproctext]{ref-tate_social_2012}
+\leavevmode\vadjust pre{\hypertarget{ref-tate_social_2012}{}}%
 Tate, Eric. 2012. {``Social Vulnerability Indices: A Comparative
@@
 
-\bibitem[\citeproctext]{ref-tate_uncertainty_2013}
+\leavevmode\vadjust pre{\hypertarget{ref-tate_uncertainty_2013}{}}%
 ---------. 2013. {``Uncertainty {Analysis} for a {Social}
@@
 
-\bibitem[\citeproctext]{ref-spei}
+\leavevmode\vadjust pre{\hypertarget{ref-spei}{}}%
 Vicente-Serrano, Sergio M., Santiago Beguería, and Juan I. López-Moreno.
@@
 
-\bibitem[\citeproctext]{ref-wang2020}
+\leavevmode\vadjust pre{\hypertarget{ref-wang2020}{}}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. {``A New Tidy Data
@@
 
-\bibitem[\citeproctext]{ref-wickham_split-apply-combine_2011}
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_split-apply-combine_2011}{}}%
 Wickham, Hadley. 2011. {``The {Split}-{Apply}-{Combine} {Strategy} for
@@
 
-\bibitem[\citeproctext]{ref-wickham_tidy_2014}
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%
 ---------. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
@@
 
-\bibitem[\citeproctext]{ref-wickham_welcome_2019}
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_welcome_2019}{}}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
@@
 
-\bibitem[\citeproctext]{ref-wickham_tourr_2011}
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_tourr_2011}{}}%
 Wickham, Hadley, Dianne Cook, Heike Hofmann, and Andreas Buja. 2011.
@@
 
-\bibitem[\citeproctext]{ref-wickham_plumbing_2009}
+\leavevmode\vadjust pre{\hypertarget{ref-wickham_plumbing_2009}{}}%
 Wickham, Hadley, Michael Lawrence, Dianne Cook, Andreas Buja, Heike
@@
 
-\bibitem[\citeproctext]{ref-WEF2023}
+\leavevmode\vadjust pre{\hypertarget{ref-WEF2023}{}}%
 World Economic Forum. 2023. {``{The Global Gender Gap Report 2023}.''}
@@
 
-\bibitem[\citeproctext]{ref-xie_reactive_2014}
+\leavevmode\vadjust pre{\hypertarget{ref-xie_reactive_2014}{}}%
 Xie, Yihui, Heike Hofmann, and Xiaoyue Cheng. 2014. {``Reactive
@@
 
-\bibitem[\citeproctext]{ref-zargar2011review}
+\leavevmode\vadjust pre{\hypertarget{ref-zargar2011review}{}}%
 Zargar, Amin, Rehan Sadiq, Bahman Naser, and Faisal I Khan. 2011. {``A
@@
 
-\bibitem[\citeproctext]{ref-zhang2024cubble}
+\leavevmode\vadjust pre{\hypertarget{ref-zhang2024cubble}{}}%
 Zhang, H. Sherry, Dianne Cook, Ursula Laa, Nicolas Langrené, and
-Patricia Menéndez. 2024. {``Cubble: An r Package for Organizing and
+Patricia Menéndez. 2024. {``Cubble: An {R} Package for Organizing and
 Wrangling Multivariate Spatio-Temporal Data.''}

### 2024-05-07 tidyindex.qmd
small fixes

@@
     digits = 3, booktabs = TRUE, 
-    col.names = c("Variable", "V-wgt", "Dimension", "D-wgt", "wgt"),
+    col.names = c("Variable", "V-wgt", "Dimension", "D-wgt", "weight"),
     linesep = c("", "", "", "", "\\hline",

### 2024-05-07 tidyindex.tex
small fixes

--- 2024-04-26 tidyindex.tex
@@
 in tidying up routine data analyses (Wickham 2014, 2011; Kuhn and Silge
-2022; Wang, Cook, and Hyndman 2020; Zhang et al. 2024) that ``turn ideas
-into software quickly and faithfully'', as envisioned by Chambers
+2022; Wang, Cook, and Hyndman 2020; Zhang et al. to appear) that ``turn
+ideas into software quickly and faithfully'', as envisioned by Chambers
 (1998). Index development and use needs tidying.
@@
 
-Adoption of this pipeline approach would provide uniformity to the field
-of index development, research, and application to improve
+The adoption of this pipeline approach would provide uniformity to the
+field of index development, research, and application to improve
 comparability, reproducibility, and communication.
@@
 construction. The highlighted path illustrates one possible construction
-using the dimension reduction and simplification module.}
+using the dimension reduction and simplification modules.}
 
@@
 The temporal processing module takes as input argument a single variable
-\(x_{sj}(t)\) at location \(s\) as a function of time. In this step the
+\(x_{sj}(t)\) at location \(s\) as a function of time. In this step, the
 original time series can be transformed or summarized into a new one via
-time aggregation, the transformation is represented by the function
+time aggregation. The transformation is represented by the function
 \(f\), \(x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]\) where
@@
 Figure~\ref{fig-scale-var-trans-compare} shows that scaling simply
-changes the numbers in the data not the shape of a variable, while
-transformation will most likely change the shape, and is usually
+changes the numbers in the data but not the shape of a variable, while
+transformation will most likely change the shape, as it is usually
 non-linear.
@@
 \toprule
-\textbf{Variable} & \textbf{V-wgt} & \textbf{Dimension} & \textbf{D-wgt} & \textbf{wgt}\\
+\textbf{Variable} & \textbf{V-wgt} & \textbf{Dimension} & \textbf{D-wgt} & \textbf{weight}\\
 \midrule
@@
 conduct this type of sensitivity analysis, where one or more components
-are perturbed and the index re-calculated. One aspect of the GGGI not
-well-described in the Global Gender Gap Report is handling of missing
-values that are present in the initial variables for many countries,
-something that is common for this type of data. This could also be made
-more transparent with the dimension reduction module, by specifying an
-imputation method or providing warnings about missing values.
+are perturbed and the index recalculated. One aspect of the GGGI not
+well-described in the Global Gender Gap Report is the handling of
+missing values that are present in the initial variables for many
+countries, something that is common for this type of data. This could
+also be made more transparent with the dimension reduction module, by
+specifying an imputation method or providing warnings about missing
+values.
 
@@
 \bibitem[\citeproctext]{ref-chambers1998programming}
-Chambers, John M. 1998. \emph{Programming with Data: A Guide to the s
+Chambers, John M. 1998. \emph{Programming with Data: A Guide to the {S}
 Language}. Berlin, Heidelberg: Springer-Verlag.
@@
 \bibitem[\citeproctext]{ref-tidymodels}
-Kuhn, Max, and Julia Silge. 2022. \emph{Tidy Modeling with r}. "
+Kuhn, Max, and Julia Silge. 2022. \emph{Tidy Modeling with {R}}. "
 O'Reilly Media, Inc.".
@@
 Zhang, H. Sherry, Dianne Cook, Ursula Laa, Nicolas Langrené, and
-Patricia Menéndez. 2024. {``Cubble: An r Package for Organizing and
-Wrangling Multivariate Spatio-Temporal Data.''}
-\url{https://arxiv.org/abs/2205.00259}.
+Patricia Menéndez. to appear. {``Cubble: An {R} Package for Organizing
+and Wrangling Multivariate Spatio-Temporal Data.''} \emph{Journal of
+Statistical Software}, to appear.
 

### 2024-05-07 tidyindex.qmd
small changes

@@
 The methods are supported by a new R package, called `tidyindex`.
+Supplemental materials for the article are available online.
 keywords:
@@
 Although it may not be possible to perfectly measure these errors, it is important that they are recognised and estimated for an index, so that it is possible to compute confidence intervals.
-In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate one possibility to compute a confidence interval for the Standardized Precipitation Index (SPI).
+In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate one possibility to compute a confidence interval for SPI.
 Bootstrapping is used to account for the sampling uncertainty in the distribution fit step of the index pipeline and to assess its impact on the SPI series.
@@
 
+# Supplementary Materials
+
+The supplementary materials include the full scripts for the three examples in Section 6 (`/scripts` folder), a saved data for Example 6.1 (`/data` folder), and a `README.md` file containing the install instructions for running the scripts.
+
 # References

### 2024-05-07 tidyindex.tex
small changes

@@
 Precipitation Index (SPI). The methods are supported by a new R package,
-called \texttt{tidyindex}.
+called \texttt{tidyindex}. Supplemental materials for the article are
+available online.
 \end{abstract}
@@
 Figure~\ref{fig-compute-temporal} is used to illustrate one possibility
-to compute a confidence interval for the Standardized Precipitation
-Index (SPI). Bootstrapping is used to account for the sampling
-uncertainty in the distribution fit step of the index pipeline and to
-assess its impact on the SPI series.
+to compute a confidence interval for SPI. Bootstrapping is used to
+account for the sampling uncertainty in the distribution fit step of the
+index pipeline and to assess its impact on the SPI series.
 
@@
 
+\section{Supplementary Materials}\label{supplementary-materials}
+
+The supplementary materials include the full scripts for the three
+examples in Section 6 (\texttt{/scripts} folder), a saved data for
+Example 6.1 (\texttt{/data} folder), and a \texttt{README.md} file
+containing the install instructions for running the scripts.
+
 \section*{References}\label{references}
